%
% see tex.stackexchange.com/a/600/151276
\def\changemargin#1#2{\list{}{\rightmargin#2\leftmargin#1}\item[]}
\let\endchangemargin=\endlist

\documentclass[openany,12pt,english]{book}
\usepackage[utf8]{inputenc}
\usepackage{babel}
\usepackage{microtype}
\usepackage{csquotes}

\usepackage[
  paperwidth=8.5in,
  paperheight=11in,
  top=1in,
  outer=1in,
  bottom=1.25in,
  inner=1.5in
]{geometry}

\title{Solomon}
\author{}
\date{}

\newenvironment{para}{\par\pretolerance=100\tolerance=200\setlength{\emergencystretch}{0.6em}\relax}{\par}

\begin{document}

\maketitle

% Copyright page
{
\thispagestyle{empty}
\vspace*{\fill}
\begin{changemargin}{.8in}{.8in}
\begin{center}
\textcopyright{}~2021 Michael Hendricks

\vspace{0.5\baselineskip}

Permission is granted to use, copy, modify, and distribute
this work for any purpose with or without fee.

\vspace{0.5\baselineskip}

This work is provided as-is without any express or implied
warranty. In no event is the author liable for any damages
arising from this work.
\end{center}
\end{changemargin}
}
\pagenumbering{arabic}
\clearpage{}

% Table of Contents
\tableofcontents
\clearpage{}

\markboth{}{}
\twocolumn[\begin{changemargin}{.6in}{.6in}
  \begin{center}
    \itshape
    An a\-bridg\-ment of the proverbs at\-trib\-ut\-ed to Solomon.
    Verse references such
    as~\textsuperscript{3:9}\thinspace{}indicate
    the original source of the text, but
    don't imply a full quotation.
  \end{center}
\end{changemargin}]{}


\vspace{0pt plus 8\baselineskip{}}\section*{My Son}\addcontentsline{toc}{section}{My Son}
\begin{para}
    \textsuperscript{3:1}\thinspace{}My son, don't for\-get my teach\-ing, but let your heart keep my commandments,
    \textsuperscript{2}\thinspace{}for they will bring you man\-y days, years of life, and peace.
\end{para}

\begin{para}
    \textsuperscript{3:3}\thinspace{}Don't let kind\-ness and truth leave you. Tie them a\-round your neck. Write them on the tab\-let of your heart.
    \textsuperscript{4}\thinspace{}Then you will find fa\-vor and high re\-gard in the sight of God and man.
\end{para}

\begin{para}
    \textsuperscript{3:5}\thinspace{}Trust in \textsc{God} with all your heart, and don't lean on your own un\-der\-stand\-ing.
    \textsuperscript{6}\thinspace{}In all your ways ac\-knowl\-edge him, and he will make your paths straight.
    \textsuperscript{7}\thinspace{}Don't be wise in your own eyes. Fear \textsc{God}, and turn a\-way from evil.
    \textsuperscript{8}\thinspace{}It will be heal\-ing to your bod\-y, and re\-fresh\-ment to your bones.
\end{para}

\begin{para}
    \textsuperscript{3:9}\thinspace{}Hon\-or \textsc{God} with your wealth, and with the first\-fruits of all your pro\-duce,
    \textsuperscript{10}\thinspace{}then your barns will be filled with plen\-ty, and your vats will over\-flow with new wine.
\end{para}

\begin{para}
    \textsuperscript{3:11}\thinspace{}My son, don't re\-ject the dis\-ci\-pline of \textsc{God} or loathe his re\-proof;
    \textsuperscript{12}\thinspace{}for \textsc{God} reproves the one he loves, just as a fa\-ther reproves the son in whom he delights.
\end{para}

\begin{para}
    \textsuperscript{3:13}\thinspace{}Hap\-py is the man who finds wis\-dom, and the man who gains un\-der\-stand\-ing.
    \textsuperscript{14}\thinspace{}For she is more prof\-it\-a\-ble than sil\-ver, and her gain is bet\-ter than fi\-ne gold.
    \textsuperscript{15}\thinspace{}She is more pre\-cious than jewels, and noth\-ing you de\-sire compares with her.
    \textsuperscript{16}\thinspace{}Long life is in her right hand. In her left hand are rich\-es and hon\-or.
    \textsuperscript{17}\thinspace{}Her ways are pleas\-ant, and all her paths are peace\-ful.
    \textsuperscript{18}\thinspace{}She is a tree of life to those who take hold of her. Those who hold her fast are hap\-py.
\end{para}

\begin{para}
    \textsuperscript{3:19}\thinspace{}\textsc{God} founded the earth by wis\-dom. He established the heavens by un\-der\-stand\-ing.
    \textsuperscript{20}\thinspace{}By his knowl\-edge the depths broke o\-pen, and the clouds drip with dew.
\end{para}

\begin{para}
    \textsuperscript{3:21}\thinspace{}My son, don't lose sight of them. Keep sound wis\-dom and dis\-cre\-tion.
    \textsuperscript{22}\thinspace{}They will be life for your soul, and a\-dorn\-ment for your neck.
    \textsuperscript{23}\thinspace{}Then you will walk on your way se\-cure\-ly, and your foot won't stum\-ble.
    \textsuperscript{24}\thinspace{}When you lie down, you won't be a\-fraid. When you lie down, your sleep will be sweet.
\end{para}

\begin{para}
    \textsuperscript{3:25}\thinspace{}Don't fear sud\-den dis\-as\-ter or the ru\-in of the wick\-ed when it co\-mes,
    \textsuperscript{26}\thinspace{}for \textsc{God} will be your con\-fi\-dence, and will keep your foot from be\-ing caught.
\end{para}

\begin{para}
    \textsuperscript{3:27}\thinspace{}Don't with\-hold good from those who de\-serve it, when it is in your pow\-er to do it.
    \textsuperscript{28}\thinspace{}Don't say to your neigh\-bor, “Go, and come back. I'll give it to\-mor\-row,” when you have it with you.
    \textsuperscript{29}\thinspace{}Don't plan evil a\-gainst your neigh\-bor, who lives trustingly be\-side you.
\end{para}

\begin{para}
    \textsuperscript{3:30}\thinspace{}Don't con\-tend with a man with\-out cause, when he has done you no harm.
    \textsuperscript{31}\thinspace{}Don't en\-vy a vi\-o\-lent man or choose any of his ways,
    \textsuperscript{32}\thinspace{}for the de\-vi\-ous are an a\-bom\-i\-na\-tion to \textsc{God}, but he is a friend to the up\-right.
\end{para}

\begin{para}
    \textsuperscript{3:33}\thinspace{}The curse of \textsc{God} is on the house of the wick\-ed, but he blesses the dwell\-ing of the right\-eous.
    \textsuperscript{34}\thinspace{}He mocks the mock\-ers, but shows fa\-vor to the hum\-ble.
    \textsuperscript{35}\thinspace{}The wise will in\-her\-it hon\-or, but fools get shame.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{4:1}\thinspace{}Lis\-ten, sons, to a father's in\-struc\-tion, and pay at\-ten\-tion so that you may gain un\-der\-stand\-ing.
    \textsuperscript{2}\thinspace{}For I give you sound in\-struc\-tion. Don't a\-ban\-don my teach\-ing.
\end{para}

\begin{para}
    \textsuperscript{4:3}\thinspace{}When I was a son to my fa\-ther, ten\-der and the on\-ly son in my mother's sight,
    \textsuperscript{4}\thinspace{}he taught me, and said to me: “Let your heart hold on to my words. Keep my commandments and live.
    \textsuperscript{5}\thinspace{}Get wis\-dom. Get un\-der\-stand\-ing. Don't for\-get or turn a\-way from the words of my mouth.
    \textsuperscript{6}\thinspace{}Don't for\-sake her, and she will guard you. Love her, and she will watch over you.
    \textsuperscript{7}\thinspace{}The be\-gin\-ning of wis\-dom is this: Get wis\-dom, and what\-ev\-er else you get, get un\-der\-stand\-ing.
    \textsuperscript{8}\thinspace{}Prize her, and she will ex\-alt you. If you em\-brace her, she will hon\-or you.
    \textsuperscript{9}\thinspace{}She will place on your head a gar\-land of grace. She will pres\-ent you with a crown of beau\-ty.”
\end{para}

\begin{para}
    \textsuperscript{4:10}\thinspace{}Lis\-ten, my son, and ac\-cept my sayings, and the years of your life will be man\-y.
    \textsuperscript{11}\thinspace{}I have taught you in the way of wis\-dom. I have led you in up\-right paths.
    \textsuperscript{12}\thinspace{}When you walk, your steps won't be hindered. When you run, you won't stum\-ble.
    \textsuperscript{13}\thinspace{}Hold on to in\-struc\-tion. Don't let go. Guard her, for she is your life.
\end{para}

\begin{para}
    \textsuperscript{4:14}\thinspace{}Don't en\-ter the path of the wick\-ed, and don't walk in the way of evil men.
    \textsuperscript{15}\thinspace{}A\-void it. Don't trav\-el on it. Turn a\-way from it, and pass on.
    \textsuperscript{16}\thinspace{}For they can\-not sleep un\-less they do evil. They are robbed of sleep un\-less they make some\-one stum\-ble.
    \textsuperscript{17}\thinspace{}For they eat the bread of wick\-ed\-ness, and drink the wine of vi\-o\-lence.
    \textsuperscript{18}\thinspace{}But the path of the right\-eous is like the light of dawn, which shines brighter and brighter un\-til mid\-day.
    \textsuperscript{19}\thinspace{}The way of the wick\-ed is like dark\-ness. They don't know what makes them stum\-ble.
\end{para}

\begin{para}
    \textsuperscript{4:20}\thinspace{}My son, pay at\-ten\-tion to my words. Lis\-ten care\-ful\-ly to my sayings.
    \textsuperscript{21}\thinspace{}Don't lose sight of them. Keep them with\-in your heart.
    \textsuperscript{22}\thinspace{}For they are life to those who find them, and health to their whole bod\-y.
\end{para}

\begin{para}
    \textsuperscript{4:23}\thinspace{}Watch over your heart with all dil\-i\-gence, for the springs of life flow from it.
    \textsuperscript{24}\thinspace{}Put a\-way from you de\-ceit\-ful speech, and put de\-vi\-ous talk far from you.
\end{para}

\begin{para}
    \textsuperscript{4:25}\thinspace{}Let your eyes look di\-rect\-ly a\-head and your gaze straight in front of you.
    \textsuperscript{26}\thinspace{}Care\-ful\-ly con\-sid\-er the path of your feet, then all your ways will be sure.
    \textsuperscript{27}\thinspace{}Don't turn to the right or to the left. Keep your foot from evil.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{5:1}\thinspace{}My son, pay at\-ten\-tion to my wis\-dom. Lis\-ten care\-ful\-ly to my un\-der\-stand\-ing,
    \textsuperscript{2}\thinspace{}that you may main\-tain dis\-cre\-tion, and that your lips may guard knowl\-edge.
\end{para}

\begin{para}
    \textsuperscript{5:3}\thinspace{}The lips of a for\-bid\-den wom\-an drip hon\-ey, and her speech is smooth\-er than oil,
    \textsuperscript{4}\thinspace{}but in the end she's as bit\-ter as worm\-wood, and as sharp as a double-edged sword.
    \textsuperscript{5}\thinspace{}Her feet go down to death. Her steps lead straight to hell.
    \textsuperscript{6}\thinspace{}She does\-n't pon\-der the path of life. Her ways are un\-sta\-ble, and she does\-n't know it.
\end{para}

\begin{para}
    \textsuperscript{5:7}\thinspace{}Now then, my sons, lis\-ten to me, and don't de\-part from the words of my mouth.
    \textsuperscript{8}\thinspace{}Keep your way far from her. Don't go near the door of her house
    \textsuperscript{9}\thinspace{}lest you give your vig\-or to others and your years to some\-one cru\-el,
    \textsuperscript{10}\thinspace{}lest strangers feast on your wealth, your labors en\-rich the house of a for\-eign\-er,
    \textsuperscript{11}\thinspace{}and at the end of your life you groan when your flesh and bod\-y are con\-sumed.
    \textsuperscript{12}\thinspace{}You will say, “How I hat\-ed dis\-ci\-pline! How my heart de\-spised re\-proof!
    \textsuperscript{13}\thinspace{}I did\-n't o\-bey my teachers or lis\-ten to my instructors!
    \textsuperscript{14}\thinspace{}I am al\-most in ut\-ter ru\-in in the midst of the as\-sem\-bly and con\-gre\-ga\-tion.”
\end{para}

\begin{para}
    \textsuperscript{5:15}\thinspace{}Drink wa\-ter from your own cis\-tern, and fresh wa\-ter from your own well.
    \textsuperscript{16}\thinspace{}Should your springs be scattered a\-broad, streams of wa\-ter in the streets?
    \textsuperscript{17}\thinspace{}Let them be yours a\-lone, and not for strangers with you.
    \textsuperscript{18}\thinspace{}Let your foun\-tain be bless\-ed, and re\-joice in the wife of your youth,
    \textsuperscript{19}\thinspace{}as a lov\-ing deer and a grace\-ful doe. Let her breasts sat\-is\-fy you al\-ways, and be ex\-hil\-a\-rat\-ed al\-ways with her love.
\end{para}

\begin{para}
    \textsuperscript{5:20}\thinspace{}Why, my son, should you be ex\-hil\-a\-rat\-ed with a for\-bid\-den wom\-an, or em\-brace the bos\-om of a strang\-er?
    \textsuperscript{21}\thinspace{}For a man's ways are be\-fore the eyes of \textsc{God}, and he considers all his paths.
    \textsuperscript{22}\thinspace{}The iniquities of the wick\-ed en\-trap him. He is held in the cords of his sin.
    \textsuperscript{23}\thinspace{}He dies for lack of dis\-ci\-pline, and be\-cause of his great fol\-ly he will go a\-stray.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{6:1}\thinspace{}My son, if you have put up se\-cu\-ri\-ty for your neigh\-bor or have giv\-en a pledge for a strang\-er,
    \textsuperscript{2}\thinspace{}if you have been snared or caught with the words of your mouth,
    \textsuperscript{3}\thinspace{}then do this, my son, and free your\-self. You have fall\-en in\-to your neighbor's hands. Go, hum\-ble your\-self, and plead with your neigh\-bor.
    \textsuperscript{4}\thinspace{}Give your eyes no sleep and your eyelids no slum\-ber.
    \textsuperscript{5}\thinspace{}Save your\-self like a ga\-zelle from the hunter's hand, and like a bird from the fowler's hand.
\end{para}

\begin{para}
    \textsuperscript{6:6}\thinspace{}Go to the ant, you i\-dler. Ob\-serve its ways, and be wise.
    \textsuperscript{7}\thinspace{}With\-out chief, of\-fic\-er, or rul\-er,
    \textsuperscript{8}\thinspace{}it prepares its provisions in sum\-mer, and gathers its food dur\-ing har\-vest.
\end{para}

\begin{para}
    \textsuperscript{6:9}\thinspace{}How long will you lie down, you i\-dler? When will you get up from your sleep?
    \textsuperscript{10}\thinspace{}A lit\-tle sleep, a lit\-tle slum\-ber, a lit\-tle folding of the hands to rest,
    \textsuperscript{11}\thinspace{}and pov\-er\-ty will come up\-on you like a rob\-ber, and want like an armed man.
\end{para}

\begin{para}
    \textsuperscript{6:12}\thinspace{}A worth\-less per\-son, a wick\-ed man, goes a\-round speak\-ing per\-verse\-ly.
    \textsuperscript{14}\thinspace{}With per\-ver\-si\-ty in his heart, he devises evil con\-tin\-u\-al\-ly, and sows dis\-cord.
    \textsuperscript{15}\thinspace{}There\-fore ca\-lam\-i\-ty will strike him sud\-den\-ly. In\-stant\-ly he will be bro\-ken be\-yond heal\-ing.
\end{para}

\begin{para}
    \textsuperscript{6:16}\thinspace{}Sev\-en things are an a\-bom\-i\-na\-tion to \textsc{God}:
    \textsuperscript{17}\thinspace{}haugh\-ty eyes, a ly\-ing tongue, hands that shed in\-no\-cent blood,
    \textsuperscript{18}\thinspace{}a heart that devises wick\-ed plans, feet that run swift\-ly to evil,
    \textsuperscript{19}\thinspace{}a false wit\-ness who utters lies, and one who sows dis\-cord a\-mong brothers.
\end{para}

\begin{para}
    \textsuperscript{6:20}\thinspace{}My son, keep your father's com\-mand, and don't for\-sake your mother's teach\-ing.
    \textsuperscript{21}\thinspace{}Al\-ways bind them on your heart. Tie them a\-round your neck.
    \textsuperscript{22}\thinspace{}When you walk, they will guide you. When you sleep, they will watch over you. When you wake up, they will talk to you.
    \textsuperscript{23}\thinspace{}For this com\-mand is a lamp, this teach\-ing is a light, and cor\-rec\-tive dis\-ci\-pline is the way to life,
    \textsuperscript{24}\thinspace{}keep\-ing you from the evil wom\-an, from the flattering tongue of an a\-dul\-ter\-ess.
\end{para}

\begin{para}
    \textsuperscript{6:25}\thinspace{}Don't de\-sire her beau\-ty in your heart or let her cap\-ti\-vate you with her eyes.
    \textsuperscript{26}\thinspace{}For a prostitute's fee is on\-ly a loaf of bread, but a mar\-ried wom\-an hunts down a pre\-cious life.
    \textsuperscript{27}\thinspace{}Can a man em\-brace fire with\-out burn\-ing his clothes?
    \textsuperscript{28}\thinspace{}Or can he walk on hot coals with\-out scorching his feet?
    \textsuperscript{29}\thinspace{}So is he who goes in to his neighbor's wife. None who touches her will go un\-pun\-ished.
\end{para}

\begin{para}
    \textsuperscript{6:30}\thinspace{}Peo\-ple don't de\-spise a thief if he steals to sat\-is\-fy his hun\-ger,
    \textsuperscript{31}\thinspace{}but if caught, he must re\-pay sev\-en\-fold. He will give eve\-ry\-thing in his house.
\end{para}

\begin{para}
    \textsuperscript{6:32}\thinspace{}He who commits a\-dul\-ter\-y lacks sense, and destroys him\-self.
    \textsuperscript{33}\thinspace{}He will get wounds and dis\-hon\-or, and his dis\-grace won't be e\-rased.
    \textsuperscript{34}\thinspace{}For jeal\-ous\-y enrages a hus\-band, and he will show no mer\-cy in the day of venge\-ance.
    \textsuperscript{35}\thinspace{}He won't ac\-cept com\-pen\-sa\-tion. He will re\-fuse, though you give man\-y gifts.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Relationships}\addcontentsline{toc}{section}{Relationships}
\begin{para}
    \textsuperscript{7:10}\thinspace{}A wom\-an met him, dressed as a pros\-ti\-tute, and cun\-ning of heart.
    \textsuperscript{13}\thinspace{}She seizes him, kisses him, and bra\-zen\-ly says to him,
    \textsuperscript{14}\thinspace{}“I had to of\-fer peace offerings. To\-day I've paid my vows.
    \textsuperscript{16}\thinspace{}I've spread my couch with coverings, with col\-ored linens from Egypt.
    \textsuperscript{18}\thinspace{}Come, let's drink our fill of love un\-til morn\-ing. Let's de\-light our\-selves with caresses.”
    \textsuperscript{21}\thinspace{}She seduces him with her per\-sua\-sive speech. With her flattering lips she lures him.
    \textsuperscript{22}\thinspace{}He follows her at once, like an ox go\-ing to the slaugh\-ter, like a deer step\-ping in\-to a trap.
    \textsuperscript{25}\thinspace{}Don't let your heart turn a\-side to her ways. Don't stray in\-to her paths.
    \textsuperscript{26}\thinspace{}For she has cast down man\-y victims, and all her slain are nu\-mer\-ous.
\end{para}

\begin{para}
    \textsuperscript{11:29}\thinspace{}He who troubles his own house will in\-her\-it the wind.
\end{para}

\begin{para}
    \textsuperscript{12:4}\thinspace{}An ex\-cel\-lent wife is her husband's crown, but she who brings shame is like rottenness in his bones.
\end{para}

\begin{para}
    \textsuperscript{13:20}\thinspace{}He who walks with wise men will be\-come wise, but the com\-pan\-ion of fools will suf\-fer harm.
\end{para}

\begin{para}
    \textsuperscript{13:24}\thinspace{}He who spares the rod hates his son, but he who loves him disciplines him dil\-i\-gent\-ly.
\end{para}

\begin{para}
    \textsuperscript{14:7}\thinspace{}Leave the pres\-ence of a fool, for you won't find knowl\-edge in his speech.
\end{para}

\begin{para}
    \textsuperscript{14:28}\thinspace{}A large pop\-u\-la\-tion is a king's glo\-ry, but with\-out peo\-ple a prince is ruined.
\end{para}

\begin{para}
    \textsuperscript{15:17}\thinspace{}Bet\-ter is a serv\-ing of vegetables where there is love, than a fattened ox served with ha\-tred.
\end{para}

\begin{para}
    \textsuperscript{16:7}\thinspace{}When a man's ways please \textsc{God}, he makes e\-ven his enemies to be at peace with him.
\end{para}

\begin{para}
    \textsuperscript{17:6}\thinspace{}Grand\-chil\-dren are the crown of old men, and fathers are the glo\-ry of their sons.
\end{para}

\begin{para}
    \textsuperscript{17:9}\thinspace{}He who conceals an of\-fense seeks love, but he who repeats a mat\-ter sep\-a\-rates close friends.
\end{para}

\begin{para}
    \textsuperscript{17:12}\thinspace{}Let a man meet a bear robbed of her cubs, rath\-er than a fool in his fol\-ly.
\end{para}

\begin{para}
    \textsuperscript{17:21}\thinspace{}He who sires a fool does so to his own sor\-row, and the fa\-ther of a fool has no joy.
    \textsuperscript{25}\thinspace{}A fool\-ish son is a grief to his fa\-ther, and bit\-ter\-ness to her who bore him.
\end{para}

\begin{para}
    \textsuperscript{18:1}\thinspace{}He who isolates him\-self seeks his own de\-sire, and rebels a\-gainst all sound wis\-dom.
\end{para}

\begin{para}
    \textsuperscript{18:22}\thinspace{}He who finds a wife finds a good thing, and obtains fa\-vor from \textsc{God}.
\end{para}

\begin{para}
    \textsuperscript{18:24}\thinspace{}He who has man\-y friends co\-mes to ru\-in, but there is a friend who stays clos\-er than a broth\-er.
\end{para}

\begin{para}
    \textsuperscript{19:14}\thinspace{}A house and wealth are in\-her\-it\-ed from fathers, but a pru\-dent wife is from \textsc{God}.
\end{para}

\begin{para}
    \textsuperscript{20:2}\thinspace{}A king's ter\-ror is like the lion's roar; he who provokes him to an\-ger forfeits his own life.
\end{para}

\begin{para}
    \textsuperscript{20:19}\thinspace{}A gos\-sip reveals secrets, there\-fore don't as\-so\-ci\-ate with him who talks too much.
\end{para}

\begin{para}
    \textsuperscript{22:6}\thinspace{}Train up a child in the way he should go; e\-ven when he is old he won't de\-part from it.
\end{para}

\begin{para}
    \textsuperscript{25:17}\thinspace{}Sel\-dom set foot in your neighbor's house, lest he grow wea\-ry of you, and hate you.
\end{para}

\begin{para}
    \textsuperscript{29:3}\thinspace{}He who loves wis\-dom brings joy to his fa\-ther, but a com\-pan\-ion of prostitutes wastes his wealth.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Instruction}\addcontentsline{toc}{section}{Instruction}
\begin{para}
    \textsuperscript{8:1}\thinspace{}Does\-n't wis\-dom call, and un\-der\-stand\-ing raise her voice?
    \textsuperscript{4}\thinspace{}“To you, O men, I call, and my voice is to the sons of men.
    \textsuperscript{5}\thinspace{}O na\-ive ones, learn pru\-dence. O fools, learn sense.
    \textsuperscript{10}\thinspace{}Take my in\-struc\-tion in\-stead of sil\-ver, and knowl\-edge rath\-er than choice gold.
    \textsuperscript{11}\thinspace{}For wis\-dom is bet\-ter than jewels, and all de\-sir\-a\-ble things can\-not com\-pare to it.
    \textsuperscript{22}\thinspace{}\textsc{God} pos\-sessed me at the be\-gin\-ning of his way, be\-fore his works of old.
    \textsuperscript{27}\thinspace{}When he established the heavens, I was there.
    \textsuperscript{32}\thinspace{}Now, my sons, lis\-ten to me, for bless\-ed are those who keep my ways.
    \textsuperscript{33}\thinspace{}Lis\-ten to in\-struc\-tion and be wise. Don't ne\-glect it.
    \textsuperscript{35}\thinspace{}For he who finds me finds life, and will ob\-tain fa\-vor of \textsc{God}.”
\end{para}

\begin{para}
    \textsuperscript{9:8}\thinspace{}Don't re\-prove a scoff\-er, or he will hate you. Re\-prove a wise man, and he will love you.
    \textsuperscript{9}\thinspace{}In\-struct a wise man, and he will be wis\-er still. Teach a right\-eous man, and he will in\-crease in learn\-ing.
    \textsuperscript{10}\thinspace{}The fear of \textsc{God} is the be\-gin\-ning of wis\-dom.
\end{para}

\begin{para}
    \textsuperscript{10:17}\thinspace{}He who heeds in\-struc\-tion is on the path of life, but he who ignores re\-proof goes a\-stray.
\end{para}

\begin{para}
    \textsuperscript{12:15}\thinspace{}The fool's way is right in his own eyes, but a wise man listens to coun\-sel.
\end{para}

\begin{para}
    \textsuperscript{13:14}\thinspace{}The teach\-ing of the wise is a foun\-tain of life, turn\-ing one a\-way from the snares of death.
\end{para}

\begin{para}
    \textsuperscript{13:18}\thinspace{}Pov\-er\-ty and shame will come to him who ignores in\-struc\-tion, but he who heeds re\-proof will be honored.
\end{para}

\begin{para}
    \textsuperscript{17:10}\thinspace{}A re\-buke goes deeper in\-to one who has un\-der\-stand\-ing than a hun\-dred lashes in\-to a fool.
\end{para}

\begin{para}
    \textsuperscript{19:27}\thinspace{}Cease listening to in\-struc\-tion, my son, and you will stray from the words of knowl\-edge.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Wise and Foolish}\addcontentsline{toc}{section}{Wise and Foolish}
\begin{para}
    \textsuperscript{10:1}\thinspace{}A wise son makes a fa\-ther glad, but a fool\-ish son brings grief to his moth\-er.
\end{para}

\begin{para}
    \textsuperscript{10:23}\thinspace{}Do\-ing wick\-ed\-ness is like sport to a fool, but wis\-dom is pleas\-ure to a man of un\-der\-stand\-ing.
\end{para}

\begin{para}
    \textsuperscript{11:29}\thinspace{}The fool will be serv\-ant to the wise of heart.
\end{para}

\begin{para}
    \textsuperscript{12:16}\thinspace{}A fool's dis\-pleas\-ure is known at once, but a pru\-dent man ignores an in\-sult.
\end{para}

\begin{para}
    \textsuperscript{14:15}\thinspace{}The na\-ive believes eve\-ry\-thing, but the pru\-dent considers his steps.
    \textsuperscript{16}\thinspace{}A wise man is cau\-tious and turns a\-way from evil, but a fool is reck\-less and care\-less.
\end{para}

\begin{para}
    \textsuperscript{17:2}\thinspace{}A pru\-dent serv\-ant will rule over a shame\-ful son, and will share the in\-her\-it\-ance a\-mong the brothers.
\end{para}

\begin{para}
    \textsuperscript{22:3}\thinspace{}The pru\-dent sees dan\-ger and hides him\-self, but the na\-ive go on and suf\-fer for it.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Righteous and Wicked}\addcontentsline{toc}{section}{Righteous and Wicked}
\begin{para}
    \textsuperscript{10:2}\thinspace{}Ill-gotten gains do not prof\-it, but right\-eous\-ness delivers from death.
    \textsuperscript{3}\thinspace{}\textsc{God} will not let the right\-eous go hun\-gry, but he thwarts the crav\-ing of the wick\-ed.
\end{para}

\begin{para}
    \textsuperscript{10:7}\thinspace{}The mem\-o\-ry of the right\-eous is a bless\-ing, but the name of the wick\-ed will rot.
\end{para}

\begin{para}
    \textsuperscript{10:20}\thinspace{}The tongue of the right\-eous is like choice sil\-ver; the heart of the wick\-ed is of lit\-tle worth.
    \textsuperscript{21}\thinspace{}The lips of the right\-eous feed man\-y, but fools die for lack of sense.
\end{para}

\begin{para}
    \textsuperscript{10:27}\thinspace{}The fear of \textsc{God} prolongs life, but the years of the wick\-ed will be shortened.
\end{para}

\begin{para}
    \textsuperscript{11:5}\thinspace{}The right\-eous\-ness of the blame\-less will di\-rect his way, but the wick\-ed will fall by his own wick\-ed\-ness.
    \textsuperscript{6}\thinspace{}The right\-eous\-ness of the up\-right will res\-cue them, but the treach\-er\-ous will be trapped by their own desires.
    \textsuperscript{8}\thinspace{}The right\-eous is delivered from trou\-ble, but the wick\-ed takes his place.
\end{para}

\begin{para}
    \textsuperscript{11:18}\thinspace{}The wick\-ed earns de\-cep\-tive wages, but he who sows right\-eous\-ness gets a sure re\-ward.
\end{para}

\begin{para}
    \textsuperscript{11:23}\thinspace{}The de\-sire of the right\-eous turns out well, but the ex\-pec\-ta\-tion of the wick\-ed is wrath.
\end{para}

\begin{para}
    \textsuperscript{11:31}\thinspace{}If the right\-eous will be rewarded on earth, how much more the wick\-ed and the sin\-ner.
\end{para}

\begin{para}
    \textsuperscript{12:3}\thinspace{}A man will not be established by wick\-ed\-ness, but the root of the right\-eous will not be moved.
    \textsuperscript{7}\thinspace{}The wick\-ed are overthrown and are no more, but the house of the right\-eous will stand.
\end{para}

\begin{para}
    \textsuperscript{12:26}\thinspace{}The right\-eous is a guide to his neigh\-bor, but the way of the wick\-ed leads them a\-stray.
\end{para}

\begin{para}
    \textsuperscript{13:21}\thinspace{}Dis\-as\-ter pursues sinners, but the right\-eous will be rewarded with good.
\end{para}

\begin{para}
    \textsuperscript{13:25}\thinspace{}The right\-eous has e\-nough to sat\-is\-fy his ap\-pe\-tite, but the stom\-ach of the wick\-ed is emp\-ty.
\end{para}

\begin{para}
    \textsuperscript{14:19}\thinspace{}The evil bow down be\-fore the good, and the wick\-ed at the gates of the right\-eous.
\end{para}

\begin{para}
    \textsuperscript{14:22}\thinspace{}Won't they who de\-vise evil go a\-stray? But those who de\-vise good will find kind\-ness and truth.
\end{para}

\begin{para}
    \textsuperscript{14:34}\thinspace{}Right\-eous\-ness exalts a na\-tion, but sin is a dis\-grace to any peo\-ple.
\end{para}

\begin{para}
    \textsuperscript{15:8}\thinspace{}The sac\-ri\-fice of the wick\-ed is an a\-bom\-i\-na\-tion to \textsc{God}, but the pray\-er of the up\-right is his de\-light.
\end{para}

\begin{para}
    \textsuperscript{21:15}\thinspace{}When jus\-tice is done, it is a joy to the right\-eous, but a ter\-ror to evildoers.
\end{para}

\begin{para}
    \textsuperscript{28:1}\thinspace{}The wick\-ed flee when no one pursues, but the right\-eous are as bold as a li\-on.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Diligence}\addcontentsline{toc}{section}{Diligence}
\begin{para}
    \textsuperscript{10:4}\thinspace{}A slack hand causes pov\-er\-ty, but a dil\-i\-gent hand makes rich.
    \textsuperscript{5}\thinspace{}A son who gathers in the sum\-mer is pru\-dent, but one who sleeps dur\-ing har\-vest is shame\-ful.
\end{para}

\begin{para}
    \textsuperscript{10:26}\thinspace{}Like vin\-e\-gar to the teeth and smoke to the eyes, so is the i\-dler to those who send him.
\end{para}

\begin{para}
    \textsuperscript{12:24}\thinspace{}The hand of the dil\-i\-gent will rule, but the sloth\-ful will be put to forced la\-bor.
\end{para}

\begin{para}
    \textsuperscript{13:4}\thinspace{}The i\-dler craves, yet gets noth\-ing, but the dil\-i\-gent is ful\-ly sat\-is\-fied.
\end{para}

\begin{para}
    \textsuperscript{14:23}\thinspace{}There is prof\-it in all la\-bor, but mer\-e talk leads on\-ly to pov\-er\-ty.
\end{para}

\begin{para}
    \textsuperscript{16:26}\thinspace{}A worker's ap\-pe\-tite works for him, for his hun\-ger urges him on.
\end{para}

\begin{para}
    \textsuperscript{18:9}\thinspace{}Who\-ev\-er is slack in his work is broth\-er to him who destroys.
\end{para}

\begin{para}
    \textsuperscript{19:15}\thinspace{}La\-zi\-ness brings on deep sleep, and an i\-dle man will suf\-fer hun\-ger.
\end{para}

\begin{para}
    \textsuperscript{19:24}\thinspace{}The i\-dler bur\-ies his hand in the dish, but will not e\-ven bring it back to his mouth.
\end{para}

\begin{para}
    \textsuperscript{20:4}\thinspace{}The i\-dler does\-n't plow in sea\-son, so he looks dur\-ing har\-vest, and has noth\-ing.
\end{para}

\begin{para}
    \textsuperscript{20:13}\thinspace{}Don't love sleep, lest you come to pov\-er\-ty. O\-pen your eyes, and you will have plen\-ty of food.
\end{para}

\begin{para}
    \textsuperscript{22:13}\thinspace{}The i\-dler says, “There's a li\-on out\-side. I'll be killed in the streets.”
\end{para}

\begin{para}
    \textsuperscript{26:14}\thinspace{}As a door turns on its hinges, so does the i\-dler on his bed.
\end{para}

\begin{para}
    \textsuperscript{26:16}\thinspace{}The i\-dler is wis\-er in his own eyes than sev\-en men who can an\-swer sen\-si\-bly.
\end{para}

\begin{para}
    \textsuperscript{27:23}\thinspace{}Know well the con\-di\-tion of your flocks, and pay at\-ten\-tion to your herds;
    \textsuperscript{24}\thinspace{}for rich\-es aren't for\-ev\-er and a crown does\-n't en\-dure to all generations.
\end{para}

\begin{para}
    \textsuperscript{28:19}\thinspace{}He who works his land will have plen\-ty of food, but he who follows worth\-less pursuits will have plen\-ty of pov\-er\-ty.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Integrity}\addcontentsline{toc}{section}{Integrity}
\begin{para}
    \textsuperscript{10:9}\thinspace{}He who walks in in\-teg\-ri\-ty walks se\-cure\-ly, but he who perverts his ways will be found out.
\end{para}

\begin{para}
    \textsuperscript{11:3}\thinspace{}The in\-teg\-ri\-ty of the up\-right will guide them, but the per\-ver\-si\-ty of the treach\-er\-ous will de\-stroy them.
\end{para}

\begin{para}
    \textsuperscript{20:7}\thinspace{}A right\-eous man who walks in his in\-teg\-ri\-ty; bless\-ed are his chil\-dren af\-ter him.
\end{para}

\begin{para}
    \textsuperscript{21:3}\thinspace{}To do right\-eous\-ness and jus\-tice is more ac\-cept\-a\-ble to \textsc{God} than sac\-ri\-fice.
    \textsuperscript{27}\thinspace{}The sac\-ri\-fice of the wick\-ed is an a\-bom\-i\-na\-tion; how much more when he brings it with evil in\-tent.
\end{para}

\begin{para}
    \textsuperscript{28:6}\thinspace{}Bet\-ter is the poor who walks in his in\-teg\-ri\-ty, than the rich who is crook\-ed in his ways.
\end{para}

\begin{para}
    \textsuperscript{28:18}\thinspace{}He who walks in in\-teg\-ri\-ty will be delivered, but he who is crook\-ed will sud\-den\-ly fall.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Contention}\addcontentsline{toc}{section}{Contention}
\begin{para}
    \textsuperscript{10:12}\thinspace{}Ha\-tred stirs up con\-flict, but love co\-vers all offenses.
\end{para}

\begin{para}
    \textsuperscript{15:18}\thinspace{}A hot-tempered man stirs up con\-flict, but the slow to an\-ger calms con\-ten\-tion.
\end{para}

\begin{para}
    \textsuperscript{17:14}\thinspace{}Starting a con\-flict is like re\-leas\-ing a flood, so a\-ban\-don the dis\-pute be\-fore it breaks out.
\end{para}

\begin{para}
    \textsuperscript{18:18}\thinspace{}Cast\-ing the lot ends quarrels, and decides be\-tween pow\-er\-ful opponents.
\end{para}

\begin{para}
    \textsuperscript{18:19}\thinspace{}An offended broth\-er is harder to win than a strong cit\-y, and contentions are like the bars of a cas\-tle.
\end{para}

\begin{para}
    \textsuperscript{20:3}\thinspace{}Stay\-ing a\-way from strife is an hon\-or for a man, but any fool will quar\-rel.
\end{para}

\begin{para}
    \textsuperscript{21:9}\thinspace{}It's bet\-ter to live in a cor\-ner of the roof than in a house shared with a con\-ten\-tious wom\-an.
    \textsuperscript{19}\thinspace{}It's bet\-ter to live in a des\-ert than with a con\-ten\-tious and complaining wom\-an.
\end{para}

\begin{para}
    \textsuperscript{22:10}\thinspace{}Drive out the mocker, and con\-ten\-tion will de\-part, and quarrels and dis\-hon\-or will cease.
\end{para}

\begin{para}
    \textsuperscript{26:20}\thinspace{}With\-out wood the fire goes out, and with\-out a gos\-sip con\-ten\-tion dies down.
    \textsuperscript{21}\thinspace{}Like char\-coal to hot embers and wood to fire, so is a con\-ten\-tious man for kin\-dling strife.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Understanding}\addcontentsline{toc}{section}{Understanding}
\begin{para}
    \textsuperscript{10:13}\thinspace{}Wis\-dom is found on the lips of the dis\-cern\-ing, but a rod is for the back of him who lacks sense.
\end{para}

\begin{para}
    \textsuperscript{13:16}\thinspace{}Eve\-ry pru\-dent man works with knowl\-edge, but a fool flaunts his fol\-ly.
\end{para}

\begin{para}
    \textsuperscript{14:12}\thinspace{}There is a way that seems right to a man, but its end is the way to death.
\end{para}

\begin{para}
    \textsuperscript{16:16}\thinspace{}How much bet\-ter to get wis\-dom than gold. To get un\-der\-stand\-ing is pref\-er\-a\-ble to sil\-ver.
\end{para}

\begin{para}
    \textsuperscript{18:13}\thinspace{}He who answers be\-fore he listens: it is fol\-ly and shame to him.
\end{para}

\begin{para}
    \textsuperscript{18:15}\thinspace{}The mind of the pru\-dent acquires knowl\-edge, and the ear of the wise seeks knowl\-edge.
\end{para}

\begin{para}
    \textsuperscript{19:2}\thinspace{}En\-thu\-si\-asm with\-out knowl\-edge is not good, and has\-ty feet go a\-stray.
\end{para}

\begin{para}
    \textsuperscript{19:20}\thinspace{}Lis\-ten to coun\-sel and re\-ceive in\-struc\-tion so that you may be wise the rest of your days.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Speech}\addcontentsline{toc}{section}{Speech}
\begin{para}
    \textsuperscript{10:18}\thinspace{}He who conceals ha\-tred has ly\-ing lips, and he who spreads slan\-der is a fool.
\end{para}

\begin{para}
    \textsuperscript{10:19}\thinspace{}When there are man\-y words, trans\-gres\-sion is un\-a\-void\-a\-ble, but he who restrains his lips is wise.
\end{para}

\begin{para}
    \textsuperscript{11:12}\thinspace{}He who belittles his neigh\-bor lacks sense, but a man of un\-der\-stand\-ing re\-mains si\-lent.
    \textsuperscript{13}\thinspace{}A gos\-sip goes a\-bout revealing secrets, but he who is trust\-wor\-thy conceals a mat\-ter.
\end{para}

\begin{para}
    \textsuperscript{11:22}\thinspace{}A beau\-ti\-ful wom\-an with\-out dis\-cre\-tion is like a gold ring in a pig's snout.
\end{para}

\begin{para}
    \textsuperscript{12:13}\thinspace{}An evil man is en\-snared by the trans\-gres\-sion of his lips, but the right\-eous will es\-cape from trou\-ble.
\end{para}

\begin{para}
    \textsuperscript{12:18}\thinspace{}There is one who speaks rash\-ly like thrusts of a sword, but the tongue of the wise brings heal\-ing.
\end{para}

\begin{para}
    \textsuperscript{12:23}\thinspace{}A pru\-dent man conceals knowl\-edge, but a fool's heart proclaims fol\-ly.
\end{para}

\begin{para}
    \textsuperscript{13:3}\thinspace{}He who guards his mouth preserves his life, but he who opens wide his lips co\-mes to ru\-in.
\end{para}

\begin{para}
    \textsuperscript{14:3}\thinspace{}In the fool's mouth is a rod for his back, but the lips of the wise will pro\-tect them.
\end{para}

\begin{para}
    \textsuperscript{15:1}\thinspace{}A gen\-tle an\-swer turns a\-way wrath, but a harsh word stirs up an\-ger.
    \textsuperscript{4}\thinspace{}A sooth\-ing tongue is a tree of life, but a per\-verse tongue breaks the spir\-it.
\end{para}

\begin{para}
    \textsuperscript{15:23}\thinspace{}A man finds joy in giv\-ing an apt an\-swer; and a time\-ly word, how good it is!
\end{para}

\begin{para}
    \textsuperscript{15:28}\thinspace{}The heart of the right\-eous ponders how to an\-swer, but the mouth of the wick\-ed blurts out evil things.
\end{para}

\begin{para}
    \textsuperscript{16:24}\thinspace{}Pleas\-ant words are a hon\-ey\-comb: sweet to the soul, and health to the bod\-y.
\end{para}

\begin{para}
    \textsuperscript{16:28}\thinspace{}A per\-verse man spreads strife, and a gos\-sip sep\-a\-rates close friends.
\end{para}

\begin{para}
    \textsuperscript{17:27}\thinspace{}He who restrains his words has knowl\-edge, and he who has a cool spir\-it is a man of un\-der\-stand\-ing.
    \textsuperscript{28}\thinspace{}E\-ven a fool is con\-sid\-ered wise when he keeps si\-lent. When he closes his lips, he is con\-sid\-ered pru\-dent.
\end{para}

\begin{para}
    \textsuperscript{18:2}\thinspace{}A fool does not de\-light in un\-der\-stand\-ing, but on\-ly in air\-ing his o\-pin\-ion.
    \textsuperscript{6}\thinspace{}His lips bring strife, and his mouth invites a beat\-ing.
    \textsuperscript{7}\thinspace{}His mouth is his ru\-in, and his lips are the snare of his soul.
\end{para}

\begin{para}
    \textsuperscript{20:15}\thinspace{}There is gold, and an a\-bun\-dance of jewels, but lips of knowl\-edge are a rare treas\-ure.
\end{para}

\begin{para}
    \textsuperscript{21:23}\thinspace{}He who guards his mouth and tongue keeps him\-self out of trou\-ble.
\end{para}

\begin{para}
    \textsuperscript{22:11}\thinspace{}He who loves pu\-ri\-ty of heart, and speaks gra\-cious\-ly will have the king as a friend.
\end{para}

\begin{para}
    \textsuperscript{26:18}\thinspace{}Like a mad\-man who throws firebrands, ar\-rows, and death,
    \textsuperscript{19}\thinspace{}is the man who deceives his neigh\-bor, and says, “I was on\-ly jok\-ing!”
\end{para}

\begin{para}
    \textsuperscript{27:14}\thinspace{}He who blesses his neigh\-bor with a loud voice ear\-ly in the morn\-ing will be counted as curs\-ing.
\end{para}

\begin{para}
    \textsuperscript{29:20}\thinspace{}Do you see a man who is has\-ty in his words? There is more hope for a fool than for him.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Honesty}\addcontentsline{toc}{section}{Honesty}
\begin{para}
    \textsuperscript{11:1}\thinspace{}Dis\-hon\-est scales are an a\-bom\-i\-na\-tion to \textsc{God}, but an ac\-cu\-rate weight is his de\-light.
\end{para}

\begin{para}
    \textsuperscript{12:22}\thinspace{}Ly\-ing lips are an a\-bom\-i\-na\-tion to \textsc{God}, but those who act faith\-ful\-ly are his de\-light.
\end{para}

\begin{para}
    \textsuperscript{16:11}\thinspace{}A just bal\-ance and scales be\-long to \textsc{God}, all the weights in the bag are his con\-cern.
\end{para}

\begin{para}
    \textsuperscript{17:15}\thinspace{}Ac\-quit\-ting the guilt\-y, and condemning the right\-eous are both a\-like an a\-bom\-i\-na\-tion to \textsc{God}.
\end{para}

\begin{para}
    \textsuperscript{17:23}\thinspace{}A wick\-ed man takes a bribe in se\-cret to per\-vert the ways of jus\-tice.
\end{para}

\begin{para}
    \textsuperscript{18:5}\thinspace{}Be\-ing par\-tial to the wick\-ed is not good, nei\-ther is de\-priv\-ing the right\-eous of jus\-tice.
\end{para}

\begin{para}
    \textsuperscript{19:5}\thinspace{}A false wit\-ness will not go un\-pun\-ished, and he who speaks lies will not es\-cape.
\end{para}

\begin{para}
    \textsuperscript{20:17}\thinspace{}Bread gained by de\-ceit is sweet to a man, but af\-ter\-ward his mouth will be full of grav\-el.
\end{para}

\begin{para}
    \textsuperscript{20:23}\thinspace{}Differing weights are an a\-bom\-i\-na\-tion to \textsc{God}, and a false scale is not good.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Humility}\addcontentsline{toc}{section}{Humility}
\begin{para}
    \textsuperscript{11:2}\thinspace{}When pride co\-mes, then co\-mes dis\-grace, but with the hum\-ble is wis\-dom.
\end{para}

\begin{para}
    \textsuperscript{12:9}\thinspace{}Bet\-ter is he who is light\-ly esteemed, and has a serv\-ant, than he that hon\-ors him\-self, and lacks bread.
\end{para}

\begin{para}
    \textsuperscript{15:25}\thinspace{}\textsc{God} will tear down the house of the proud, but will pro\-tect the borders of the wid\-ow.
\end{para}

\begin{para}
    \textsuperscript{16:5}\thinspace{}Eve\-ry\-one who is proud in heart is an a\-bom\-i\-na\-tion to \textsc{God}; be as\-sured, he will not go un\-pun\-ished.
\end{para}

\begin{para}
    \textsuperscript{16:18}\thinspace{}Pride co\-mes be\-fore de\-struc\-tion, and an ar\-ro\-gant spir\-it be\-fore a fall.
\end{para}

\begin{para}
    \textsuperscript{16:19}\thinspace{}It is bet\-ter to be hum\-ble in spir\-it a\-mong the low\-ly, than to di\-vide the spoil with the proud.
\end{para}

\begin{para}
    \textsuperscript{18:12}\thinspace{}Be\-fore de\-struc\-tion a man's heart is haugh\-ty, but hu\-mil\-i\-ty co\-mes be\-fore hon\-or.
\end{para}

\begin{para}
    \textsuperscript{21:30}\thinspace{}No wis\-dom, no un\-der\-stand\-ing, and no coun\-sel will pre\-vail a\-gainst \textsc{God}.
\end{para}

\begin{para}
    \textsuperscript{22:4}\thinspace{}The re\-ward for hu\-mil\-i\-ty and fear of \textsc{God} is rich\-es, hon\-or, and life.
\end{para}

\begin{para}
    \textsuperscript{26:12}\thinspace{}Do you see a man wise in his own eyes? There is more hope for a fool than for him.
\end{para}

\begin{para}
    \textsuperscript{27:1}\thinspace{}Don't boast a\-bout to\-mor\-row, for you don't know what a day may bring.
\end{para}

\begin{para}
    \textsuperscript{27:2}\thinspace{}Let an\-oth\-er praise you, and not your own mouth; a strang\-er, and not your own lips.
\end{para}

\begin{para}
    \textsuperscript{29:23}\thinspace{}A man's pride will bring him low, but a hum\-ble spir\-it will ob\-tain hon\-or.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Counsel}\addcontentsline{toc}{section}{Counsel}
\begin{para}
    \textsuperscript{11:14}\thinspace{}With\-out guid\-ance the peo\-ple fall, but with man\-y counselors there is safe\-ty.
\end{para}

\begin{para}
    \textsuperscript{14:27}\thinspace{}The fear of \textsc{God} is a foun\-tain of life, turn\-ing one a\-way from the snares of death.
\end{para}

\begin{para}
    \textsuperscript{15:22}\thinspace{}With\-out coun\-sel plans fail, but with man\-y advisers they suc\-ceed.
\end{para}

\begin{para}
    \textsuperscript{16:33}\thinspace{}The lot is cast in\-to the lap, but its eve\-ry de\-ci\-sion is from \textsc{God}.
\end{para}

\begin{para}
    \textsuperscript{20:18}\thinspace{}Es\-tab\-lish plans by coun\-sel, and make war with wise guid\-ance.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Finance}\addcontentsline{toc}{section}{Finance}
\begin{para}
    \textsuperscript{11:15}\thinspace{}He who is guar\-an\-tor for a strang\-er will sure\-ly suf\-fer for it, but he who hates such agreements is se\-cure.
\end{para}

\begin{para}
    \textsuperscript{11:28}\thinspace{}He who trusts in his rich\-es will fall, but the right\-eous will flour\-ish like a green leaf.
\end{para}

\begin{para}
    \textsuperscript{13:7}\thinspace{}There is one who pretends to be rich, yet has noth\-ing. An\-oth\-er pretends to be poor, yet has great wealth.
\end{para}

\begin{para}
    \textsuperscript{13:11}\thinspace{}Wealth obtained by fraud will dwin\-dle, but he who gathers by la\-bor will in\-crease it.
\end{para}

\begin{para}
    \textsuperscript{13:22}\thinspace{}A good man leaves an in\-her\-it\-ance to his children's chil\-dren, but the sinner's wealth is stored up for the right\-eous.
\end{para}

\begin{para}
    \textsuperscript{14:4}\thinspace{}Where there are no ox\-en, the man\-ger is clean, but a\-bun\-dant har\-vest co\-mes through an ox's strength.
\end{para}

\begin{para}
    \textsuperscript{14:20}\thinspace{}The poor is hat\-ed e\-ven by his neigh\-bor, but there are man\-y who love the rich.
\end{para}

\begin{para}
    \textsuperscript{15:6}\thinspace{}There is great wealth in the house of the right\-eous, but trou\-ble accompanies the in\-come of the wick\-ed.
\end{para}

\begin{para}
    \textsuperscript{15:27}\thinspace{}He who profits il\-lic\-it\-ly troubles his own house, but he who hates bribes will live.
\end{para}

\begin{para}
    \textsuperscript{16:8}\thinspace{}Bet\-ter a lit\-tle with right\-eous\-ness than great in\-come with in\-jus\-tice.
\end{para}

\begin{para}
    \textsuperscript{19:4}\thinspace{}Wealth attracts man\-y friends, but a poor man is sep\-a\-rat\-ed from his friend.
    \textsuperscript{6}\thinspace{}Man\-y seek a gen\-er\-ous man's fa\-vor, and eve\-ry\-one is a friend to him who gives gifts.
    \textsuperscript{7}\thinspace{}All a poor man's brothers hate him. How much more do his friends a\-void him! He pursues them with words, but they are gone.
\end{para}

\begin{para}
    \textsuperscript{21:5}\thinspace{}The plans of the dil\-i\-gent lead sure\-ly to a\-bun\-dance, but eve\-ry\-one who is has\-ty co\-mes sure\-ly to pov\-er\-ty.
\end{para}

\begin{para}
    \textsuperscript{21:6}\thinspace{}Ac\-quir\-ing treasures by a ly\-ing tongue is a fleet\-ing va\-por and the pur\-suit of death.
\end{para}

\begin{para}
    \textsuperscript{21:17}\thinspace{}He who loves pleas\-ure will be\-come poor. He who loves wine and oil will not get rich.
    \textsuperscript{20}\thinspace{}There is pre\-cious treas\-ure and oil in a wise man's dwell\-ing, but a fool consumes them.
\end{para}

\begin{para}
    \textsuperscript{22:1}\thinspace{}A good name is to be cho\-sen over great rich\-es, and fa\-vor over sil\-ver and gold.
\end{para}

\begin{para}
    \textsuperscript{22:2}\thinspace{}The rich and the poor have this in com\-mon: \textsc{God} is the mak\-er of them all.
\end{para}

\begin{para}
    \textsuperscript{22:7}\thinspace{}The rich rules over the poor, and the bor\-row\-er is a slave to the lend\-er.
\end{para}

\begin{para}
    \textsuperscript{26:10}\thinspace{}Like an arch\-er who wounds eve\-ry\-one, so is he who hires a fool or who hires those pass\-ing by.
\end{para}

\begin{para}
    \textsuperscript{28:22}\thinspace{}A stin\-gy man hastens af\-ter wealth, and does\-n't know that pov\-er\-ty will come up\-on him.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Kindness}\addcontentsline{toc}{section}{Kindness}
\begin{para}
    \textsuperscript{11:17}\thinspace{}A kind man benefits him\-self, but a cru\-el man harms him\-self.
\end{para}

\begin{para}
    \textsuperscript{12:10}\thinspace{}A right\-eous man has re\-gard for the life of his an\-i\-mal, but e\-ven the mer\-cy of the wick\-ed is cru\-el.
\end{para}

\begin{para}
    \textsuperscript{14:21}\thinspace{}He who despises his neigh\-bor sins, but he who shows kind\-ness to the poor is hap\-py.
    \textsuperscript{31}\thinspace{}He who oppresses the poor insults his mak\-er, but he who is kind to the needy hon\-ors him.
\end{para}

\begin{para}
    \textsuperscript{17:13}\thinspace{}If any\-one returns evil for good, evil will nev\-er leave his house.
\end{para}

\begin{para}
    \textsuperscript{21:21}\thinspace{}He who pursues right\-eous\-ness and kind\-ness will find life, right\-eous\-ness, and hon\-or.
\end{para}

\begin{para}
    \textsuperscript{26:27}\thinspace{}He who digs a pit will fall in\-to it, and he who rolls a stone, it will come back on him.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Generosity}\addcontentsline{toc}{section}{Generosity}
\begin{para}
    \textsuperscript{11:24}\thinspace{}One gives free\-ly, yet increases all the more, and an\-oth\-er withholds what is due, but co\-mes on\-ly to pov\-er\-ty.
    \textsuperscript{25}\thinspace{}The gen\-er\-ous man will pros\-per, and he who waters will al\-so be watered him\-self.
    \textsuperscript{26}\thinspace{}The peo\-ple will curse him who withholds grain, but a bless\-ing is on the head of him who sells it.
\end{para}

\begin{para}
    \textsuperscript{19:17}\thinspace{}Kind\-ness to the poor is a loan to \textsc{God}, and he will re\-pay his good deed.
\end{para}

\begin{para}
    \textsuperscript{20:25}\thinspace{}It is a snare to say rash\-ly, “It is ho\-ly,” and on\-ly make in\-quir\-y af\-ter the vows.
\end{para}

\begin{para}
    \textsuperscript{21:13}\thinspace{}He who shuts his ear to the cry of the poor will al\-so call out and not be answered.
\end{para}

\begin{para}
    \textsuperscript{22:9}\thinspace{}He who is gen\-er\-ous will be bless\-ed, for he shares his bread with the poor.
    \textsuperscript{16}\thinspace{}He who oppresses the poor to in\-crease his wealth, or who gives to the rich, will on\-ly come to pov\-er\-ty.
\end{para}

\begin{para}
    \textsuperscript{25:21}\thinspace{}If your en\-e\-my is hun\-gry, give him bread to eat, and if he is thirst\-y, give him wa\-ter to drink.
\end{para}

\begin{para}
    \textsuperscript{27:7}\thinspace{}He who is full loathes hon\-ey, but to the hun\-gry any bit\-ter thing is sweet.
\end{para}

\begin{para}
    \textsuperscript{28:27}\thinspace{}He who gives to the poor will nev\-er want, but he who hides his eyes will have man\-y curs\-es.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Emotion}\addcontentsline{toc}{section}{Emotion}
\begin{para}
    \textsuperscript{12:25}\thinspace{}Anx\-i\-e\-ty in a man's heart weighs it down, but a good word makes it glad.
\end{para}

\begin{para}
    \textsuperscript{13:12}\thinspace{}Hope de\-ferred makes the heart sick, but de\-sire ful\-filled is a tree of life.
\end{para}

\begin{para}
    \textsuperscript{14:13}\thinspace{}E\-ven in laugh\-ter the heart may ache, and joy may end in grief.
\end{para}

\begin{para}
    \textsuperscript{14:17}\thinspace{}A man of quick tem\-per acts fool\-ish\-ly, and a man of evil schemes is hat\-ed.
    \textsuperscript{29}\thinspace{}He who is slow to an\-ger has great un\-der\-stand\-ing, but he who has a quick tem\-per promotes fol\-ly.
\end{para}

\begin{para}
    \textsuperscript{14:30}\thinspace{}A tran\-quil heart is life to the bod\-y, but en\-vy makes the bones rot.
\end{para}

\begin{para}
    \textsuperscript{15:13}\thinspace{}A glad heart makes a cheer\-ful coun\-te\-nance, but a sorrowful heart breaks the spir\-it.
\end{para}

\begin{para}
    \textsuperscript{15:15}\thinspace{}All the days of the afflicted are bad, but a cheer\-ful heart has a con\-tin\-u\-al feast.
\end{para}

\begin{para}
    \textsuperscript{16:14}\thinspace{}A king's wrath is a mes\-sen\-ger of death, but a wise man appeases it.
\end{para}

\begin{para}
    \textsuperscript{16:32}\thinspace{}He who is slow to an\-ger is bet\-ter than the might\-y, and he who rules his spir\-it, than he who captures a cit\-y.
\end{para}

\begin{para}
    \textsuperscript{17:5}\thinspace{}Who\-ev\-er mocks the poor insults his mak\-er, and he who rejoices at ca\-lam\-i\-ty will not go un\-pun\-ished.
\end{para}

\begin{para}
    \textsuperscript{17:22}\thinspace{}A joy\-ful heart is good med\-i\-cine, but a bro\-ken spir\-it dries up the bones.
\end{para}

\begin{para}
    \textsuperscript{18:14}\thinspace{}A man's spir\-it can en\-dure sick\-ness, but who can bear a bro\-ken spir\-it?
\end{para}

\begin{para}
    \textsuperscript{19:3}\thinspace{}A man's fool\-ish\-ness ruins his way, yet his heart rages a\-gainst \textsc{God}.
\end{para}

\begin{para}
    \textsuperscript{19:11}\thinspace{}Good sense makes a man slow to an\-ger, and it is his glo\-ry to over\-look an of\-fense.
\end{para}

\begin{para}
    \textsuperscript{19:19}\thinspace{}A man of great an\-ger will pay the pen\-al\-ty, for if you res\-cue him, you will have to do it a\-gain.
\end{para}

\begin{para}
    \textsuperscript{20:1}\thinspace{}Wine is a mocker, strong drink a brawl\-er, and who\-ev\-er is led a\-stray by them is not wise.
\end{para}

\begin{para}
    \textsuperscript{20:22}\thinspace{}Don't say, “I will re\-pay evil”. Wait for \textsc{God}, and he will save you.
\end{para}

\begin{para}
    \textsuperscript{29:11}\thinspace{}A fool gives full vent to his an\-ger, but a wise man holds it back.
\end{para}


\end{document}

% Copyright 2021 Michael Hendricks
%
% Permission is granted to use, copy, modify, and distribute
% this work for any purpose with or without fee.
%
% This work is provided as-is without any express or implied
% warranty. In no event is the author liable for any damages
% arising from this work.
