%
% see tex.stackexchange.com/a/600/151276
\def\changemargin#1#2{\list{}{\rightmargin#2\leftmargin#1}\item[]}
\let\endchangemargin=\endlist

\documentclass[openany,12pt,english]{book}
\input{lib/packages}

\usepackage[
  paperwidth=5.875in,
  paperheight=8.25in,
  inner=1in,     % samples: 0.75 0.75 0.875 1 1
  outer=0.725in, % samples: 0.5 0.5 0.625 0.725 0.75
  top=0.75in,
  bottom=1in
]{geometry}


\title{Di\-gests and Mes\-sages}
\author{}
\date{}

\newenvironment{para}{\par\pretolerance=100\tolerance=200\setlength{\emergencystretch}{0.6em}\relax}{\par}

\begin{document}

\maketitle

% Copyright page
{
\thispagestyle{empty}
\vspace*{\fill}
\begin{changemargin}{.5in}{.5in}
\begin{center}
\textcopyright{}~2021 Michael Hendricks

\vspace{0.5\baselineskip}

Permission is granted to use, copy, modify, and distribute
this work for any purpose with or without fee.

\vspace{0.5\baselineskip}

This work is provided as-is without any express or implied
warranty. In no event is the author liable for any damages
arising from this work.
\end{center}
\end{changemargin}
}
\pagenumbering{arabic}
\clearpage{}

% Table of Contents
\tableofcontents
\clearpage{}

\markboth{}{}


\chapter*{Introduction}
\markboth{Introduction}{Introduction}
\addcontentsline{toc}{chapter}{Introduction}
\begin{para}
    \textsuperscript{1}\thinspace{}God loves eve\-ry\-one and wants each of us to find joy and sat\-is\-fac\-tion.
    \textsuperscript{2}\thinspace{}To help us to\-wards that goal, God gave us rea\-son, con\-science, and the a\-bil\-i\-ty to choose.
    \textsuperscript{3}\thinspace{}Our re\-spon\-si\-bil\-i\-ty is to fol\-low, sin\-cere\-ly, the dictates of our con\-science, guid\-ed by rea\-son, while re\-spect\-ing the e\-qual priv\-i\-lege of others to do the same.
\end{para}

\begin{para}
    \textsuperscript{4}\thinspace{}We're ac\-count\-a\-ble to God for our choices.
    \textsuperscript{5}\thinspace{}Each of us can have a per\-son\-al re\-la\-tion\-ship with God.
    \textsuperscript{6}\thinspace{}God communicates with each of us as needed,
    \textsuperscript{7}\thinspace{}and sends messengers to dif\-fer\-ent times, places, and cultures to guide us. Stud\-y\-ing the messages that are in\-tend\-ed for us can en\-light\-en our un\-der\-stand\-ing.
\end{para}

\begin{para}
    \textsuperscript{8}\thinspace{}God knows the vast\-ness of hu\-man di\-ver\-si\-ty. Just as a skilled op\-tom\-e\-trist prescribes eye\-glass\-es to com\-pen\-sate for a patient's shortcomings, God guides us to a pre\-scrip\-tion that's op\-ti\-mized for our spir\-it\-u\-al growth. Our pre\-scrip\-tion may dif\-fer from, or co\-in\-cide with, that of our peers.
    \textsuperscript{9}\thinspace{}Be\-cause of this di\-ver\-si\-ty, a com\-mu\-ni\-ty of believers can strength\-en and in\-spire one an\-oth\-er.
\end{para}

\begin{para}
    \textsuperscript{10}\thinspace{}Details a\-bout God's na\-ture and methods are large\-ly a dis\-trac\-tion from the prop\-er fo\-cus of our lives: a life\-long mi\-gra\-tion to\-wards joy through our per\-son\-al choices.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{11}\thinspace{}While striving to live the principles de\-scribed a\-bove, I felt guid\-ed to prayer\-ful\-ly un\-der\-take the actions be\-low. The results are for my own ben\-e\-fit and may not ben\-e\-fit others in the same way.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{12}\thinspace{}Cre\-ate a prioritized list of es\-sen\-tial commandments to help me fo\-cus on what matters most and to re\-solve conflicts that may a\-rise be\-tween those commandments. This be\-came the Di\-gest of Priorities.
\end{para}

\begin{para}
    \textsuperscript{13}\thinspace{}Write pairs of es\-sen\-tial vir\-tues whose opposing demands I must bal\-ance in my own life. This be\-came the Di\-gest of Vir\-tues.
\end{para}

\begin{para}
    \textsuperscript{14}\thinspace{}Try dif\-fer\-ent methods of pray\-er, listening care\-ful\-ly to un\-der\-stand which are most ef\-fec\-tive at strengthening my re\-la\-tion\-ship with God. Cod\-i\-fy those practices and ob\-serve them faith\-ful\-ly. This be\-came the Di\-gest of Pray\-er.
\end{para}

\begin{para}
    \textsuperscript{15}\thinspace{}Rec\-ord short stories that God has used to clar\-i\-fy my un\-der\-stand\-ing of cer\-tain principles. This be\-came the Di\-gest of Parables.
\end{para}

\begin{para}
    \textsuperscript{16}\thinspace{}Stud\-y the teachings of those who are wide\-ly con\-sid\-ered to be messengers of God. A\-bridge those messages by selecting the most prac\-ti\-cal and im\-por\-tant teachings for guid\-ing my life. This be\-came the Messages. I re\-fer to them as The Mes\-sage of Mo\-ses, The Mes\-sage of Jesus, etc.
\end{para}

\chapter*{Digests}
\markboth{Digests}{Digests}
\addcontentsline{toc}{chapter}{Digests}
\vspace{0pt plus 8\baselineskip{}}\section*{Priorities}
\addcontentsline{toc}{section}{Priorities}
\markright{Priorities}
\begin{para}
    \textsuperscript{0}\thinspace{}You have rea\-son and con\-science to gov\-ern your choices. The fol\-low\-ing priorities are a guide to keep con\-science with\-in prop\-er bounds. They are list\-ed by de\-creas\-ing im\-por\-tance. Try to ob\-serve all the priorities. If two must con\-flict, pre\-fer the more im\-por\-tant one:
\end{para}

\begin{para}
    \textsuperscript{1}\thinspace{}Com\-pul\-sion may on\-ly be used to stop an im\-me\-di\-ate threat to life, prop\-er\-ty, or free will.
    \textsuperscript{2}\thinspace{}Choose what you will have at the ex\-pense of what you won't.
    \textsuperscript{3}\thinspace{}Give no im\-por\-tance to that which has none.
    \textsuperscript{4}\thinspace{}Tend to your own af\-fairs.
\end{para}

\begin{para}
    \textsuperscript{5}\thinspace{}Make promises spar\-ing\-ly.
    \textsuperscript{6}\thinspace{}Keep your word.
    \textsuperscript{7}\thinspace{}Make a\-mends.
    \textsuperscript{8}\thinspace{}Be grate\-ful.
\end{para}

\begin{para}
    \textsuperscript{9}\thinspace{}Take care of your\-self.
    \textsuperscript{10}\thinspace{}Be spar\-ing and cau\-tious with behaviors that might be ad\-dic\-tive.
\end{para}

\begin{para}
    \textsuperscript{11}\thinspace{}Re\-spect others.
    \textsuperscript{12}\thinspace{}Speak kind\-ly and qui\-et\-ly.
    \textsuperscript{13}\thinspace{}Main\-tain a few strong relationships.
    \textsuperscript{14}\thinspace{}Main\-tain sta\-ble, long-term relationships with your spouses.
    \textsuperscript{15}\thinspace{}Ob\-tain con\-sent from your spouses for all in\-ti\-mate re\-la\-tions.
    \textsuperscript{16}\thinspace{}Have chil\-dren with your spouses.
\end{para}

\begin{para}
    \textsuperscript{17}\thinspace{}Spend some time each day qui\-et and a\-lone.
    \textsuperscript{18}\thinspace{}Pray dai\-ly.
    \textsuperscript{19}\thinspace{}Stud\-y the Digests and Messages dai\-ly, bal\-anc\-ing lit\-er\-al and fig\-ur\-a\-tive in\-ter\-pre\-ta\-tion.
\end{para}

\begin{para}
    \textsuperscript{20}\thinspace{}A\-void con\-ten\-tion.
    \textsuperscript{21}\thinspace{}For\-give wrongs, re\-al and perceived.
    \textsuperscript{22}\thinspace{}Main\-tain po\-lit\-i\-cal neu\-tral\-i\-ty.
\end{para}

\begin{para}
    \textsuperscript{23}\thinspace{}Ac\-quire use\-ful knowl\-edge, then act on it.
    \textsuperscript{24}\thinspace{}En\-gage skill\-ful\-ly in an oc\-cu\-pa\-tion, trade, or craft.
    \textsuperscript{25}\thinspace{}Make an hon\-est prof\-it.
    \textsuperscript{26}\thinspace{}In\-cur debts rare\-ly and spar\-ing\-ly, and pay them back quick\-ly.
    \textsuperscript{27}\thinspace{}Ten per\-cent of your an\-nu\-al expenditures should be donations.
    \textsuperscript{28}\thinspace{}Keep a mod\-est re\-serve of use\-ful as\-sets.
\end{para}

\begin{para}
    \textsuperscript{29}\thinspace{}Smile.
    \textsuperscript{30}\thinspace{}Serve a\-non\-y\-mous\-ly, with\-out re\-mu\-ner\-a\-tion.
\end{para}

\begin{para}
    \textsuperscript{31}\thinspace{}Main\-tain your bod\-y in strength, speed, flex\-i\-bil\-i\-ty, and bal\-ance.
    \textsuperscript{32}\thinspace{}De\-vote one day in sev\-en to rejuvenation and spir\-it\-u\-al growth.
    \textsuperscript{33}\thinspace{}Im\-prove your sur\-round\-ings.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Virtues}
\addcontentsline{toc}{section}{Virtues}
\markright{Virtues}
\begin{para}
    \textsuperscript{0}\thinspace{}Eve\-ry vir\-tue tak\-en to ex\-treme becomes vi\-ce. The vir\-tues be\-low are list\-ed in pairs. Bal\-ance the two vir\-tues in a pair. If you're un\-sure how to bal\-ance them, err on the side of the first. The pairs are list\-ed al\-pha\-bet\-i\-cal\-ly with no rel\-a\-tive im\-por\-tance in\-tend\-ed be\-tween them:
\end{para}

\begin{para}
    \textsuperscript{1}\thinspace{}Chal\-lenge, com\-fort.
    \textsuperscript{2}\thinspace{}Com\-pan\-ion\-ship, sol\-i\-tude.
    \textsuperscript{3}\thinspace{}Con\-tent\-ment, am\-bi\-tion.
    \textsuperscript{4}\thinspace{}De\-tach\-ment, at\-tach\-ment.
    \textsuperscript{5}\thinspace{}Flex\-i\-ble, te\-na\-cious.
    \textsuperscript{6}\thinspace{}Gen\-er\-ous, fru\-gal.
    \textsuperscript{7}\thinspace{}Grad\-u\-al, swift.
    \textsuperscript{8}\thinspace{}Ho\-lis\-tic, at\-om\-is\-tic.
    \textsuperscript{9}\thinspace{}Hum\-ble, con\-fi\-dent.
    \textsuperscript{10}\thinspace{}In\-de\-pend\-ence, con\-form\-i\-ty.
    \textsuperscript{11}\thinspace{}In\-tel\-lect, e\-mo\-tion.
    \textsuperscript{12}\thinspace{}Mer\-cy, jus\-tice.
    \textsuperscript{13}\thinspace{}O\-pen, guard\-ed.
    \textsuperscript{14}\thinspace{}Pre\-pare, im\-pro\-vise.
    \textsuperscript{15}\thinspace{}Si\-lence, com\-mu\-ni\-ca\-tion.
    \textsuperscript{16}\thinspace{}Sim\-ple, a\-bun\-dant.
    \textsuperscript{17}\thinspace{}So\-lem\-ni\-ty, hu\-mor.
    \textsuperscript{18}\thinspace{}Sta\-ble, dy\-nam\-ic.
    \textsuperscript{19}\thinspace{}Tact\-ful, hon\-est.
    \textsuperscript{20}\thinspace{}Tra\-di\-tion, in\-no\-va\-tion.
    \textsuperscript{21}\thinspace{}Tran\-quil\-i\-ty, ex\-cite\-ment.
    \textsuperscript{22}\thinspace{}Work, re\-pose.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Prayer}
\addcontentsline{toc}{section}{Prayer}
\markright{Prayer}
\begin{para}
    \textsuperscript{1:1}\thinspace{}Of\-fer in\-for\-mal pray\-er at least three times dur\-ing your wak\-ing hours.
\end{para}

\begin{para}
    \textsuperscript{1:2}\thinspace{}Po\-si\-tion your\-self how\-ev\-er is mean\-ing\-ful, com\-fort\-a\-ble, and ap\-pro\-pri\-ate to your circumstances.
    \textsuperscript{3}\thinspace{}Pray si\-lent\-ly or a\-loud.
    \textsuperscript{4}\thinspace{}Be\-gin by say\-ing, “God in heav\-en.”
    \textsuperscript{5}\thinspace{}In your own words, ex\-press grat\-i\-tude, ac\-knowl\-edge your faults, and re\-quest guid\-ance and as\-sis\-tance.
    \textsuperscript{6}\thinspace{}Pause to lis\-ten.
\end{para}

\begin{para}
    \textsuperscript{1:7}\thinspace{}Con\-clude by say\-ing, “A\-men.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{2:1}\thinspace{}Of\-fer for\-mal pray\-er once or twice dur\-ing your wak\-ing hours,
    \textsuperscript{2}\thinspace{}with at least one performed out\-side.
\end{para}

\begin{para}
    \textsuperscript{2:3}\thinspace{}When praying out\-side, con\-sid\-er turn\-ing your back to\-wards the ap\-prox\-i\-mate lo\-ca\-tion of the sun.
    \textsuperscript{4}\thinspace{}Pray a\-loud if pos\-si\-ble.
    \textsuperscript{5}\thinspace{}Make your best ef\-fort to per\-form each pos\-ture of pray\-er in turn:
\end{para}

\begin{para}
    \textsuperscript{2:6}\thinspace{}Grat\-i\-tude: Stand, face up\-ward, eyes closed, hands clasped, un\-til tran\-quil.
    \textsuperscript{7}\thinspace{}Re\-cite, “God in heav\-en, I thank you for life, air to breathe, qui\-et places, and the a\-bil\-i\-ty to choose;
    \textsuperscript{8}\thinspace{}for challenges, work, e\-mo\-tion, and the gift of rea\-son.”
    \textsuperscript{9}\thinspace{}In your own words, ex\-press heart\-felt grat\-i\-tude to God.
\end{para}

\begin{para}
    \textsuperscript{2:10}\thinspace{}Con\-fes\-sion: Bow, face down\-ward, eyes closed, hands on knees.
    \textsuperscript{11}\thinspace{}Re\-cite, “God in heav\-en, I strive to keep the Priorities, to bal\-ance the Vir\-tues, and to learn from my failures when I don't.
    \textsuperscript{12}\thinspace{}I want to be more faith\-ful.
    \textsuperscript{13}\thinspace{}I fall short in these ways.”
    \textsuperscript{14}\thinspace{}In your own words, con\-fess to God your failures, weaknesses, and ways you need to im\-prove.
\end{para}

\begin{para}
    \textsuperscript{2:15}\thinspace{}Pe\-ti\-tion: Squat, face for\-ward, eyes closed, arms out\-stretched, hands cupped to form a bowl.
    \textsuperscript{16}\thinspace{}Re\-cite, “God in heav\-en, you have giv\-en me a beau\-ti\-ful world to live in, messengers to teach me, and guid\-ance when I ask.
    \textsuperscript{17}\thinspace{}Hear my pray\-er.”
    \textsuperscript{18}\thinspace{}In your own words, re\-quest God's guid\-ance.
\end{para}

\begin{para}
    \textsuperscript{2:19}\thinspace{}Re\-cep\-tion: Stand, face down\-ward, eyes closed, hands clasped.
    \textsuperscript{20}\thinspace{}Re\-cite, “God in heav\-en, you let the sun rise on good and evil peo\-ple a\-like.
    \textsuperscript{21}\thinspace{}You send guid\-ance through a gen\-tle thought, a soft feel\-ing, or a qui\-et voice.
    \textsuperscript{22}\thinspace{}I'm listening.” Lis\-ten for God's guid\-ance.
\end{para}

\begin{para}
    \textsuperscript{2:23}\thinspace{}Ac\-tion: Bow, face down\-ward, eyes closed, hands on knees.
    \textsuperscript{24}\thinspace{}Re\-cite, “God in heav\-en, the great\-est gift brings no prof\-it when left un\-used.
    \textsuperscript{25}\thinspace{}I thank you for giv\-ing me guid\-ance.
    \textsuperscript{26}\thinspace{}I plan to act.”
    \textsuperscript{27}\thinspace{}In your own words, tell God how you will act on the guid\-ance you have re\-ceived.
\end{para}

\begin{para}
    \textsuperscript{2:28}\thinspace{}Con\-clude by say\-ing, “A\-men.”
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Parables}
\addcontentsline{toc}{section}{Parables}
\markright{Parables}
\begin{para}
    \textsuperscript{1:1}\thinspace{}A young child asks his moth\-er, “How is a ba\-by made?”
    \textsuperscript{2}\thinspace{}Know\-ing the child's ru\-di\-men\-ta\-ry un\-der\-stand\-ing, she responds,
    \textsuperscript{3}\thinspace{}“The fa\-ther places a seed in the mother's bel\-ly, and it grows in\-to a ba\-by.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{2:1}\thinspace{}A king had two vaults filled with treas\-ure, each protected by its own lock.
    \textsuperscript{2}\thinspace{}The king said that who\-ev\-er opened a vault could have the treas\-ure in\-side.
    \textsuperscript{3}\thinspace{}Two men sought the treas\-ure.
\end{para}

\begin{para}
    \textsuperscript{2:4}\thinspace{}The first man turned the dials of the lock.
    \textsuperscript{5}\thinspace{}He listened care\-ful\-ly as the mech\-an\-ism moved in\-side.
    \textsuperscript{6}\thinspace{}He no\-ticed patterns of sound and vi\-bra\-tion with\-in the dials.
    \textsuperscript{7}\thinspace{}Af\-ter man\-y days of per\-sis\-tence and stud\-y, he discovered the com\-bi\-na\-tion and unlocked the vault.
\end{para}

\begin{para}
    \textsuperscript{2:8}\thinspace{}The sec\-ond man offered to be the king's serv\-ant.
    \textsuperscript{9}\thinspace{}He listened care\-ful\-ly to the king's wishes, and ex\-e\-cut\-ed them faith\-ful\-ly.
    \textsuperscript{10}\thinspace{}He no\-ticed patterns a\-mong the king's requests and an\-tic\-i\-pat\-ed his wishes.
    \textsuperscript{11}\thinspace{}Af\-ter man\-y days of per\-sis\-tence and serv\-ice, the sec\-ond man asked the king for the com\-bi\-na\-tion and re\-ceived it.
\end{para}

\begin{para}
    \textsuperscript{2:13}\thinspace{}Both men re\-ceived treas\-ure.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{3:1}\thinspace{}There were two lakes.
    \textsuperscript{2}\thinspace{}The first lake said to him\-self, “I will re\-tain eve\-ry drop of wa\-ter that flows in\-to me so that I can be great.”
    \textsuperscript{3}\thinspace{}His waters be\-came salt\-y and pu\-trid.
    \textsuperscript{4}\thinspace{}His fish died and the plants a\-long his shore wilted.
    \textsuperscript{5}\thinspace{}Animals and travelers avoided him.
\end{para}

\begin{para}
    \textsuperscript{3:6}\thinspace{}The sec\-ond lake said to him\-self, “I want to pros\-per all life on the moun\-tain be\-low me. I will re\-tain a mod\-est pool for my\-self then al\-low each drop to pass down the riv\-er to those who might use it.”
    \textsuperscript{7}\thinspace{}His waters be\-came clean and sweet.
    \textsuperscript{8}\thinspace{}His fish grew large and the plants a\-long his shore prospered.
    \textsuperscript{9}\thinspace{}Animals and travelers came from a\-far to en\-joy his splen\-dor.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{4:1}\thinspace{}Rain, fresh\-ly fall\-en from a cloud, is clean and good to drink.
    \textsuperscript{2}\thinspace{}As it flows down mountains and a\-cross plains, it becomes soiled and bit\-ter.
    \textsuperscript{3}\thinspace{}The sun must ex\-tract pure wa\-ter from the sea a\-gain, leav\-ing impurities be\-hind, to pro\-duce fresh rain once more.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{5:1}\thinspace{}A horse, a dol\-phin, and an ea\-gle each needs dif\-fer\-ent nour\-ish\-ment,
    \textsuperscript{2}\thinspace{}and yet each grows to full stat\-ure and maj\-es\-ty.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{6:1}\thinspace{}A child asks his moth\-er, “Please get me a cup from the top shelf.”
    \textsuperscript{2}\thinspace{}The moth\-er reaches the cup for him.
    \textsuperscript{3}\thinspace{}A grown man asks his moth\-er, “Please get me a cup from the top shelf.”
    \textsuperscript{4}\thinspace{}She responds, “You can reach it your\-self.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{7:1}\thinspace{}A man mixed one meas\-ure of sug\-ar and the juice from one lem\-on in\-to 2,000 meas\-ures of wa\-ter,
    \textsuperscript{2}\thinspace{}then served it to his guests as lem\-on\-ade.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{8:1}\thinspace{}A man brought eggs home from the mar\-ket.
    \textsuperscript{2}\thinspace{}Up\-on ar\-riv\-ing, he an\-nounced, “I purchased one doz\-en eggs.”
    \textsuperscript{3}\thinspace{}His wife said, “No you did\-n't. You bought twelve eggs.”
    \textsuperscript{4}\thinspace{}A son said, “You're both wrong. There are clear\-ly three rows of four eggs each.”
    \textsuperscript{5}\thinspace{}A daugh\-ter dis\-a\-greed, “Not at all. There are four columns of three eggs each.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{9:1}\thinspace{}A man de\-cid\-ed to climb a moun\-tain.
    \textsuperscript{2}\thinspace{}He fixed his gaze per\-ma\-nent\-ly on the sum\-mit and be\-gan hik\-ing.
    \textsuperscript{3}\thinspace{}He tripped o\-ver un\-seen rocks at his feet,
    \textsuperscript{4}\thinspace{}and was un\-a\-ble to reach the moun\-tain top.
\end{para}

\begin{para}
    \textsuperscript{9:5}\thinspace{}An\-oth\-er man de\-cid\-ed to climb the same moun\-tain.
    \textsuperscript{6}\thinspace{}He gazed per\-ma\-nent\-ly at his feet, to a\-void the errors of the first man, and be\-gan hik\-ing.
    \textsuperscript{7}\thinspace{}With\-out trip\-ping once, he wandered off in\-to a val\-ley,
    \textsuperscript{8}\thinspace{}and was un\-a\-ble to reach the moun\-tain top.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{10:1}\thinspace{}A man no\-ticed jig\-saw puz\-zle pieces rest\-ing on a ta\-ble.
    \textsuperscript{2}\thinspace{}He wanted to as\-sem\-ble the puz\-zle but pieces were mis\-sing, and he did\-n't know how the puz\-zle should look when fin\-ished.
\end{para}

\begin{para}
    \textsuperscript{10:3}\thinspace{}From time to time he lo\-cat\-ed mis\-sing pieces, but there were still too few to guess how the puz\-zle should look.
    \textsuperscript{4}\thinspace{}One day he found a sketch of the fin\-ished puz\-zle.
    \textsuperscript{5}\thinspace{}He as\-sem\-bled the pieces he had, on\-ly to re\-al\-ize that he was still mis\-sing some.
    \textsuperscript{6}\thinspace{}He con\-tin\-ued search\-ing for mis\-sing pieces, plac\-ing them where they belonged.
    \textsuperscript{7}\thinspace{}In time, he found all the pieces, and as\-sem\-bled the puz\-zle in\-to its beau\-ti\-ful, com\-plete form.
\end{para}

\chapter*{Moses}
\markboth{Moses}{Moses}
\addcontentsline{toc}{chapter}{Messages}
\addcontentsline{toc}{section}{Moses}\begin{changemargin}{.6in}{.6in}
  \begin{center}
    \itshape
    An a\-bridg\-ment of the Book of Deuteronomy.
    Verse references such
    as~\textsuperscript{5:6}\thinspace{}indicate
    the original source of the text, but
    don't imply a full quotation.
  \end{center}
\end{changemargin}
\begin{para}
    \textsuperscript{5:1}\thinspace{}Mo\-ses summoned all Israel and said to them, Hear the statutes and ordinances that I speak to\-day. Learn them and ob\-serve them care\-ful\-ly.
    \textsuperscript{2}\thinspace{}\textsc{God} made a cov\-e\-nant with us.
    \textsuperscript{3}\thinspace{}\textsc{God} did not make this cov\-e\-nant with our fathers.
    \textsuperscript{4}\thinspace{}\textsc{God} spoke with you on the moun\-tain, say\-ing,
\end{para}

\begin{para}
    \textsuperscript{5:6}\thinspace{}I am \textsc{God}, who brought you out of the land of Egypt, out of the house of slav\-er\-y.
    \textsuperscript{7}\thinspace{}Don't have oth\-er gods be\-fore me.
    \textsuperscript{8}\thinspace{}Don't make an i\-dol for your\-self in the form of any\-thing in the heavens or on the earth or in the waters.
    \textsuperscript{9}\thinspace{}Don't wor\-ship them or serve them, for I am a jeal\-ous god, visiting the in\-iq\-ui\-ty of the fathers on the chil\-dren to the third and fourth gen\-er\-a\-tion,
    \textsuperscript{10}\thinspace{}but show\-ing love to thousands who love me and keep my commandments.
\end{para}

\begin{para}
    \textsuperscript{5:11}\thinspace{}Don't mis\-use the name of \textsc{God}.
\end{para}

\begin{para}
    \textsuperscript{5:12}\thinspace{}Ob\-serve the sabbath day by keep\-ing it ho\-ly.
    \textsuperscript{13}\thinspace{}La\-bor six days and do all your work,
    \textsuperscript{14}\thinspace{}but the sev\-enth day is a sabbath to \textsc{God}. On it you shall not do any work: nei\-ther you, your chil\-dren, your serv\-ant, your live\-stock, nor any for\-eign\-er liv\-ing a\-mong you, that they may rest as you do.
    \textsuperscript{15}\thinspace{}Re\-mem\-ber that you were slaves in Egypt, and that \textsc{God} brought you out. There\-fore, \textsc{God} commanded you to keep the sabbath day.
\end{para}

\begin{para}
    \textsuperscript{5:16}\thinspace{}Hon\-or your parents so that you may live long and that it may go well with you in the land that \textsc{God} is giv\-ing you.
\end{para}

\begin{para}
    \textsuperscript{5:17}\thinspace{}Don't mur\-der.
    \textsuperscript{18}\thinspace{}Don't com\-mit a\-dul\-ter\-y.
    \textsuperscript{19}\thinspace{}Don't steal.
    \textsuperscript{20}\thinspace{}Don't give false tes\-ti\-mo\-ny a\-gainst your neigh\-bor.
    \textsuperscript{21}\thinspace{}Don't cov\-et your neighbor's wife, house, field, servants, live\-stock, or any\-thing that belongs to your neigh\-bor.
\end{para}

\begin{para}
    \textsuperscript{5:32}\thinspace{}Be care\-ful to do as \textsc{God} has commanded you. Don't turn a\-side to the right or to the left.
    \textsuperscript{33}\thinspace{}Walk on the path which \textsc{God} has commanded you, that you may live, pros\-per, and pro\-long your days in the land that you will pos\-sess.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{6:5}\thinspace{}Love \textsc{God} with all your heart, soul, and strength.
    \textsuperscript{6}\thinspace{}These words that I com\-mand you to\-day shall be on your heart.
\end{para}

\begin{para}
    \textsuperscript{6:10}\thinspace{}When \textsc{God} brings you in\-to the land that he swore to your fathers Abraham, Isaac, and Jacob that he would give you --- a land with large and splen\-did cit\-ies,
    \textsuperscript{11}\thinspace{}hous\-es filled with eve\-ry good thing, cisterns, vineyards, and ol\-ive trees --- when you eat and are sat\-is\-fied,
    \textsuperscript{12}\thinspace{}then be care\-ful not to for\-get \textsc{God}.
\end{para}

\begin{para}
    \textsuperscript{6:14}\thinspace{}Don't fol\-low oth\-er gods, the gods of the peo\-ple a\-round you,
    \textsuperscript{15}\thinspace{}lest the an\-ger of \textsc{God} be kindled a\-gainst you, and de\-stroy you from the face of the earth.
\end{para}

\begin{para}
    \textsuperscript{6:20}\thinspace{}When your son asks you, “What is the mean\-ing of the testimonies, the statutes, and the rules that \textsc{God} has commanded you?”
    \textsuperscript{21}\thinspace{}tell him, “We were slaves to Pharaoh in Egypt, but \textsc{God} brought us out of Egypt.
    \textsuperscript{23}\thinspace{}He brought us out from there in or\-der to give us the land that he swore to our fathers.
    \textsuperscript{24}\thinspace{}So \textsc{God} commanded us to ob\-serve all these statutes and to fear \textsc{God} for our good and our sur\-viv\-al.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{7:1}\thinspace{}When \textsc{God} brings you in\-to the land, and clears a\-way man\-y nations be\-fore you,
    \textsuperscript{3}\thinspace{}don't in\-ter\-mar\-ry with them.
    \textsuperscript{4}\thinspace{}They will turn your chil\-dren a\-way from me, to wor\-ship oth\-er gods.
\end{para}

\begin{para}
    \textsuperscript{7:7}\thinspace{}\textsc{God} did\-n't choose you be\-cause you were more nu\-mer\-ous than oth\-er peo\-ple,
    \textsuperscript{8}\thinspace{}but be\-cause \textsc{God} loves you and is keep\-ing the oath that he swore to your fathers.
\end{para}

\begin{para}
    \textsuperscript{7:12}\thinspace{}If you lis\-ten to and care\-ful\-ly keep these ordinances, \textsc{God} will keep the cov\-e\-nant of love with you that he swore to your fathers.
    \textsuperscript{13}\thinspace{}He will love you, bless you, and mul\-ti\-ply you. He will bless the fruit of your womb, the crops of your land, and the in\-crease of your herd.
\end{para}

\begin{para}
    \textsuperscript{7:17}\thinspace{}If you say in your heart, “These nations are greater than I. How can I dis\-pos\-sess them?”
    \textsuperscript{18}\thinspace{}Don't be a\-fraid of them. Re\-mem\-ber what \textsc{God} did to Pharaoh and to all Egypt.
    \textsuperscript{21}\thinspace{}Don't dread them, for \textsc{God} is a\-mong you, a great and awe\-some god.
    \textsuperscript{22}\thinspace{}\textsc{God} will clear a\-way these nations be\-fore you lit\-tle by lit\-tle. You will not e\-lim\-i\-nate them all at once, lest the wild animals be\-come too nu\-mer\-ous for you.
    \textsuperscript{25}\thinspace{}Burn the carved images of their gods. Don't cov\-et the sil\-ver or gold that is on them, nor take it for your\-self, lest you be en\-snared by it.
    \textsuperscript{26}\thinspace{}Don't bring an a\-bom\-i\-na\-tion in\-to your house, lest you be set a\-part for de\-struc\-tion. In\-stead, utterly de\-test it and ab\-hor it.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{8:2}\thinspace{}Re\-mem\-ber how \textsc{God} led you these for\-ty years in the wil\-der\-ness, to hum\-ble you, and to test what was in your heart, wheth\-er you would keep his commandments.
    \textsuperscript{3}\thinspace{}He humbled you, let you go hun\-gry, then fed you with man\-na, which nei\-ther you nor your fathers had known, to teach you that man does not live by bread a\-lone, but by eve\-ry word that co\-mes from the mouth of \textsc{God}.
    \textsuperscript{5}\thinspace{}Know in your heart that \textsc{God} disciplines you just as a man disciplines his son.
    \textsuperscript{6}\thinspace{}There\-fore keep the commandments of \textsc{God} by walk\-ing in his ways and by fearing him.
    \textsuperscript{10}\thinspace{}When you've eat\-en and are sat\-is\-fied, then bless \textsc{God} for the good land he has giv\-en you.
\end{para}

\begin{para}
    \textsuperscript{8:11}\thinspace{}Be care\-ful not to for\-get \textsc{God}, by fail\-ing to keep his commandments, his ordinances, and his statutes.
    \textsuperscript{12}\thinspace{}Oth\-er\-wise, once you've eat\-en and are sat\-is\-fied, and have built good hous\-es to live in,
    \textsuperscript{13}\thinspace{}and when your herds mul\-ti\-ply, and your sil\-ver and gold mul\-ti\-ply, and eve\-ry\-thing you have multiplies,
    \textsuperscript{14}\thinspace{}then your heart will be\-come proud, and you will for\-get \textsc{God},
    \textsuperscript{17}\thinspace{}and you may say in your heart, “My pow\-er and the strength of my hand has made me this wealth.”
    \textsuperscript{18}\thinspace{}But re\-mem\-ber that \textsc{God} gives you pow\-er to make wealth in or\-der to con\-firm his cov\-e\-nant.
    \textsuperscript{19}\thinspace{}If you for\-get \textsc{God}, fol\-low oth\-er gods, serve them, and wor\-ship them, I tes\-ti\-fy a\-gainst you to\-day that you will sure\-ly per\-ish.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{9:4}\thinspace{}Once \textsc{God} has driv\-en the peo\-ple out be\-fore you, don't say in your heart, “\textsc{God} brought me in to pos\-sess this land be\-cause of my right\-eous\-ness.” In\-stead, \textsc{God} drives them out be\-cause of their wick\-ed\-ness.
    \textsuperscript{6}\thinspace{}Un\-der\-stand that \textsc{God} is\-n't giv\-ing you this good land be\-cause of your right\-eous\-ness, for you are a stub\-born peo\-ple.
    \textsuperscript{7}\thinspace{}Don't for\-get how you pro\-voked \textsc{God} in the wil\-der\-ness. From the day you de\-part\-ed Egypt un\-til you ar\-rived in this place, you have been re\-bel\-lious a\-gainst \textsc{God}.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{10:17}\thinspace{}\textsc{God} is not par\-tial and takes no bribe.
    \textsuperscript{18}\thinspace{}He executes jus\-tice for orphans and widows, and loves the for\-eign\-er by giv\-ing him food and cloth\-ing.
    \textsuperscript{19}\thinspace{}There\-fore, love the for\-eign\-er, for you were foreigners in Egypt.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{11:2}\thinspace{}Re\-mem\-ber that your chil\-dren did\-n't ex\-pe\-ri\-ence the dis\-ci\-pline of \textsc{God}; his works which he did in Egypt;
    \textsuperscript{4}\thinspace{}what he did to the ar\-my of Egypt; how he made the wa\-ter of the Red sea en\-gulf them as they pur\-sued you;
    \textsuperscript{5}\thinspace{}and what he did to you in the wil\-der\-ness, un\-til you reached this place;
    \textsuperscript{7}\thinspace{}but your eyes have seen eve\-ry great work \textsc{God} has done.
\end{para}

\begin{para}
    \textsuperscript{11:16}\thinspace{}Be care\-ful not to be en\-ticed to turn a\-side and serve oth\-er gods and wor\-ship them.
    \textsuperscript{18}\thinspace{}There\-fore im\-print my words on your heart and on your soul, bind them as a sign on your hand, and wear them on your foreheads.
    \textsuperscript{19}\thinspace{}Teach them to your chil\-dren, speak of them when you sit in your house, when you walk a\-long the road, when you lie down, and when you get up.
    \textsuperscript{20}\thinspace{}Write them on the doorposts of your house and on your gates,
    \textsuperscript{21}\thinspace{}that your days, and the days of your chil\-dren, may be mul\-ti\-plied as long as the heavens are a\-bove the earth.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{12:4}\thinspace{}Don't wor\-ship \textsc{God} in the same way as the nations you dis\-pos\-sess.
    \textsuperscript{5}\thinspace{}In\-stead, seek the place where \textsc{God} chooses to es\-tab\-lish his name. Go there
    \textsuperscript{6}\thinspace{}and bring your burnt offerings, your sacrifices, your tithes, your contributions, your vow offerings, your free\-will offerings, and the firstborn of your live\-stock.
    \textsuperscript{7}\thinspace{}There, in the pres\-ence of \textsc{God}, you and your families will eat and re\-joice in eve\-ry\-thing you un\-der\-take, be\-cause \textsc{God} has bless\-ed you.
\end{para}

\begin{para}
    \textsuperscript{12:20}\thinspace{}When \textsc{God} enlarges your ter\-ri\-to\-ry as he has prom\-ised you, and you say, “I want to eat meat,” you may eat meat when\-ev\-er you want.
\end{para}

\begin{para}
    \textsuperscript{12:29}\thinspace{}When \textsc{God} cuts off the nations be\-fore you, and you live in their land,
    \textsuperscript{30}\thinspace{}be care\-ful not to be en\-snared by their ways,
    \textsuperscript{31}\thinspace{}for they have done eve\-ry de\-test\-a\-ble act for their gods; e\-ven burn\-ing their chil\-dren as sacrifices.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{13:1}\thinspace{}If a proph\-et arises a\-mong you, and gives you a sign or a won\-der,
    \textsuperscript{2}\thinspace{}and it co\-mes true, but he says, “Let us fol\-low oth\-er gods and wor\-ship them,”
    \textsuperscript{3}\thinspace{}don't lis\-ten to that proph\-et. For \textsc{God} is testing you to find out wheth\-er you love \textsc{God} with all your heart and with all your soul.
\end{para}

\begin{para}
    \textsuperscript{13:6}\thinspace{}If your broth\-er, child, wife, or friend se\-cret\-ly entices you, say\-ing, “Let us wor\-ship oth\-er gods,”
    \textsuperscript{8}\thinspace{}don't yield to him and don't lis\-ten to him.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{14:22}\thinspace{}Set a\-side a tenth of the pro\-duce that is harvested from your fields each year.
    \textsuperscript{23}\thinspace{}Eat the tithe of your grain, of your wine, and of your oil, and the firstborn of your live\-stock be\-fore \textsc{God}, in the place where he chooses to es\-tab\-lish his name, so that you may learn to fear \textsc{God} al\-ways.
\end{para}

\begin{para}
    \textsuperscript{14:24}\thinspace{}If the dis\-tance is so great that you can't car\-ry it,
    \textsuperscript{25}\thinspace{}then turn it in\-to mon\-ey and take the mon\-ey to the place where \textsc{God} chooses to es\-tab\-lish his name.
    \textsuperscript{26}\thinspace{}Spend the mon\-ey for what\-ev\-er you want, for ox\-en, or sheep, or wine, or strong drink. Eat it be\-fore \textsc{God} and re\-joice, you and your house\-hold.
    \textsuperscript{27}\thinspace{}Don't ne\-glect the Levite who is in your town, for he has no por\-tion or in\-her\-it\-ance with you.
\end{para}

\begin{para}
    \textsuperscript{14:28}\thinspace{}At the end of eve\-ry third year, bring all the tithe of that year's pro\-duce and store it in your town.
    \textsuperscript{29}\thinspace{}The Levite, the for\-eign\-er, the fa\-ther\-less, and the wid\-ow in your town shall come and eat and be sat\-is\-fied, so that \textsc{God} may bless you in all your work.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{15:1}\thinspace{}At the end of eve\-ry sev\-enth year, can\-cel debts.
    \textsuperscript{2}\thinspace{}Eve\-ry cred\-i\-tor must can\-cel the loans he made to his neigh\-bor. He must not de\-mand pay\-ment from his neigh\-bor.
    \textsuperscript{3}\thinspace{}You may re\-quire pay\-ment from a for\-eign\-er, but you must for\-give what\-ev\-er your broth\-er owes you.
    \textsuperscript{6}\thinspace{}For \textsc{God} will bless you, as he has prom\-ised. You will lend to man\-y nations, but you will not bor\-row. You will rule man\-y nations, but none will rule you.
\end{para}

\begin{para}
    \textsuperscript{15:7}\thinspace{}If there is a poor man a\-mong you, in any of your towns, don't hard\-en your heart or shut your hand to\-wards him.
    \textsuperscript{8}\thinspace{}In\-stead, o\-pen your hand to him and lend him suf\-fi\-cient for his need, what\-ev\-er it may be.
\end{para}

\begin{para}
    \textsuperscript{15:9}\thinspace{}Be care\-ful not to have this wick\-ed thought: “The sev\-enth year, the year of re\-lease, is near,” and you look grudg\-ing\-ly on your poor broth\-er and give him noth\-ing. He will cry to \textsc{God} a\-gainst you, and you will be guilt\-y.
    \textsuperscript{10}\thinspace{}Give gen\-er\-ous\-ly to him with\-out a grudg\-ing heart. Be\-cause of this \textsc{God} will bless you in eve\-ry\-thing you do.
    \textsuperscript{11}\thinspace{}There will al\-ways be poor peo\-ple in the land. There\-fore free\-ly o\-pen your hand to your broth\-er, to the poor, and to the needy.
\end{para}

\begin{para}
    \textsuperscript{15:12}\thinspace{}If a fel\-low Hebrew is sold to you, and serves you for six years, free him in the sev\-enth year.
    \textsuperscript{13}\thinspace{}When you free him, don't send him a\-way empty-handed.
    \textsuperscript{14}\thinspace{}Fur\-nish him lib\-er\-al\-ly from your flock, from your threshing floor, and from your wine\-press. Give to him as \textsc{God} has bless\-ed you.
    \textsuperscript{15}\thinspace{}Re\-mem\-ber that you were a slave in Egypt, and \textsc{God} redeemed you. That is why I am giv\-ing you this com\-mand.
    \textsuperscript{18}\thinspace{}Don't con\-sid\-er it a hard\-ship when you set him free, for he served you six years at half the cost of a hired serv\-ant. \textsc{God} will bless you in all you do.
\end{para}

\begin{para}
    \textsuperscript{15:19}\thinspace{}Ded\-i\-cate to \textsc{God} eve\-ry firstborn male of your live\-stock. Do no work with the firstborn of your herd, nor shear the firstborn of your flock.
    \textsuperscript{20}\thinspace{}Each year you and your fam\-i\-ly are to eat it be\-fore \textsc{God} in the place \textsc{God} chooses.
    \textsuperscript{21}\thinspace{}If it has any de\-fect, is lame or blind, don't sac\-ri\-fice it to \textsc{God}.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{16:1}\thinspace{}Ob\-serve the month of Aviv and cel\-e\-brate the Passover to \textsc{God}, for in the month of Aviv \textsc{God} brought you out of Egypt by night.
    \textsuperscript{3}\thinspace{}For sev\-en days you are to eat un\-leav\-ened bread---be\-cause you left Egypt in haste---so that you may re\-mem\-ber the day you left Egypt for the rest of your life.
    \textsuperscript{4}\thinspace{}Let no yeast be found in your ter\-ri\-to\-ry for sev\-en days.
    \textsuperscript{6}\thinspace{}Sac\-ri\-fice the Passover at sun\-set, at the time you came out of Egypt.
    \textsuperscript{8}\thinspace{}On the sev\-enth day hold a sol\-emn as\-sem\-bly to \textsc{God}, and do no work.
\end{para}

\begin{para}
    \textsuperscript{16:9}\thinspace{}Sev\-en weeks af\-ter you start the grain har\-vest,
    \textsuperscript{10}\thinspace{}cel\-e\-brate the Feast of Weeks to \textsc{God} by giv\-ing a free\-will of\-fer\-ing in pro\-por\-tion to how \textsc{God} has bless\-ed you.
    \textsuperscript{11}\thinspace{}Re\-joice be\-fore \textsc{God}; you, your chil\-dren, your servants, and the Levites in your town; the for\-eign\-er, the or\-phan, and the wid\-ow, who are a\-mong you.
\end{para}

\begin{para}
    \textsuperscript{16:13}\thinspace{}Cel\-e\-brate the Feast of Tabernacles for sev\-en days af\-ter gath\-er\-ing the pro\-duce from your threshing floor and wine\-press.
    \textsuperscript{14}\thinspace{}Re\-joice dur\-ing your feast; you, your chil\-dren, your servants, and the Levites; the strang\-er, the or\-phan, and the wid\-ow who are a\-mong you.
    \textsuperscript{15}\thinspace{}Cel\-e\-brate the feast to \textsc{God} for sev\-en days, be\-cause \textsc{God} will bless you in all your pro\-duce and in all the work of your hands, so that you will be al\-to\-geth\-er joy\-ful.
\end{para}

\begin{para}
    \textsuperscript{16:16}\thinspace{}All males must ap\-pear three times a year be\-fore \textsc{God} where he chooses: at the Feast of Un\-leav\-ened Bread, the Feast of Weeks, and the Feast of Tabernacles. None shall ap\-pear be\-fore \textsc{God} empty-handed.
    \textsuperscript{17}\thinspace{}Each of you must bring a gift in pro\-por\-tion to how \textsc{God} has bless\-ed you.
\end{para}

\begin{para}
    \textsuperscript{16:18}\thinspace{}Ap\-point judges and officers for your tribes in all your towns. They shall judge the peo\-ple right\-eous\-ly.
    \textsuperscript{19}\thinspace{}Don't per\-vert jus\-tice or be par\-tial. Don't take a bribe, for it blinds the wise and twists the words of the right\-eous.
    \textsuperscript{20}\thinspace{}Pur\-sue jus\-tice, and on\-ly jus\-tice.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{17:1}\thinspace{}Don't sac\-ri\-fice to \textsc{God} an ox or a sheep with any blem\-ish or de\-fect, for that is de\-test\-a\-ble to \textsc{God}.
\end{para}

\begin{para}
    \textsuperscript{17:2}\thinspace{}If any\-one in any of your towns
    \textsuperscript{3}\thinspace{}has wor\-shiped oth\-er gods,
    \textsuperscript{4}\thinspace{}and if you hear a\-bout it, then in\-ves\-ti\-gate dil\-i\-gent\-ly. If it is true and cer\-tain,
    \textsuperscript{5}\thinspace{}bring him to your cit\-y gates and stone him to death.
    \textsuperscript{6}\thinspace{}He shall be put to death on the tes\-ti\-mo\-ny of two or three witnesses, but not on the tes\-ti\-mo\-ny of one wit\-ness.
    \textsuperscript{7}\thinspace{}The hands of the witnesses shall be first a\-gainst him to put him to death, and af\-ter\-ward the hands of all the peo\-ple.
\end{para}

\begin{para}
    \textsuperscript{17:8}\thinspace{}If a case is too dif\-fi\-cult for you to judge, wheth\-er hom\-i\-cide, law\-suit, or as\-sault, then go to the place \textsc{God} chooses,
    \textsuperscript{9}\thinspace{}and ask the priests or the judge who is in of\-fice, and they will give you the ver\-dict.
    \textsuperscript{10}\thinspace{}You must a\-bide by the ver\-dict they give you. Be care\-ful to do eve\-ry\-thing they tell you.
\end{para}

\begin{para}
    \textsuperscript{17:14}\thinspace{}When you en\-ter the land \textsc{God} is giv\-ing you, and you pos\-sess it, and live in it, and say, “I will set a king o\-ver me like all the nations a\-round me,”
    \textsuperscript{15}\thinspace{}ap\-point a king whom \textsc{God} chooses, one from a\-mong your brothers. Don't ap\-point a for\-eign\-er who is not your broth\-er.
    \textsuperscript{16}\thinspace{}He must not ac\-quire man\-y hors\-es for him\-self or send the peo\-ple to Egypt to ac\-quire man\-y hors\-es. For \textsc{God} has told you, “Nev\-er re\-turn that way a\-gain.”
    \textsuperscript{17}\thinspace{}He must not ac\-quire man\-y wives for him\-self, lest his heart go a\-stray. He must not ac\-quire ex\-ces\-sive sil\-ver and gold for him\-self.
\end{para}

\begin{para}
    \textsuperscript{17:18}\thinspace{}When he sits on the throne, he must write for him\-self a cop\-y of this law, in the pres\-ence of the Levites.
    \textsuperscript{19}\thinspace{}He must read from it all the days of his life, so he may learn to fear \textsc{God} by ob\-serv\-ing this law and these statutes,
    \textsuperscript{20}\thinspace{}lest his heart be lifted up a\-bove his countrymen or he turn a\-side from the com\-mand\-ment.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{18:9}\thinspace{}When you en\-ter the land which \textsc{God} is giv\-ing you, don't im\-i\-tate the de\-test\-a\-ble practices of those nations.
    \textsuperscript{10}\thinspace{}No one a\-mong you is to sac\-ri\-fice his child in the fire, prac\-tice div\-i\-na\-tion, tell fortunes, prac\-tice sor\-cer\-y,
    \textsuperscript{11}\thinspace{}cast spells, con\-sult a me\-di\-um or spir\-it\-ist, or in\-quire of the dead.
\end{para}

\begin{para}
    \textsuperscript{18:15}\thinspace{}\textsc{God} will raise up for you a proph\-et like me from a\-mong your countrymen. Lis\-ten to him.
    \textsuperscript{18}\thinspace{}I will put my words in his mouth, and he will tell them eve\-ry\-thing I com\-mand him.
    \textsuperscript{19}\thinspace{}I will hold ac\-count\-a\-ble those who do not lis\-ten to my words that he speaks in my name.
\end{para}

\begin{para}
    \textsuperscript{18:21}\thinspace{}You may ask your\-self, “How can we know when a mes\-sage is not from \textsc{God}?”
    \textsuperscript{22}\thinspace{}When a proph\-et speaks in \textsc{God}'s name and it does not come true, then \textsc{God} has not spo\-ken that mes\-sage. The proph\-et spoke pre\-sump\-tu\-ous\-ly. Don't be a\-fraid of him.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{19:2}\thinspace{}Set a\-part three cit\-ies for your\-selves in the mid\-dle of the land that \textsc{God} is giv\-ing you.
    \textsuperscript{3}\thinspace{}Di\-vide the land in\-to three parts, so that any man\-slay\-er can flee to these cit\-ies.
    \textsuperscript{4}\thinspace{}A man\-slay\-er is one who kills his neigh\-bor ac\-ci\-den\-tal\-ly with\-out pre\-vi\-ous hos\-til\-i\-ty.
    \textsuperscript{5}\thinspace{}He may flee to one of these cit\-ies and live,
    \textsuperscript{6}\thinspace{}lest the a\-veng\-er of blood pur\-sue the man\-slay\-er in an\-ger, and kill him, though he did not de\-serve death.
\end{para}

\begin{para}
    \textsuperscript{19:14}\thinspace{}Don't move your neighbor's bound\-a\-ry mark, established by your predecessors.
\end{para}

\begin{para}
    \textsuperscript{19:15}\thinspace{}One wit\-ness is in\-suf\-fi\-cient to es\-tab\-lish wrong\-do\-ing by any man. A charge must be established on the ev\-i\-dence of two or three witnesses.
    \textsuperscript{16}\thinspace{}If a ma\-li\-cious wit\-ness accuses some\-one of wrong\-do\-ing,
    \textsuperscript{17}\thinspace{}then both peo\-ple in the dis\-pute must ap\-pear be\-fore \textsc{God}, be\-fore the priests and the judges.
    \textsuperscript{18}\thinspace{}The judges must in\-ves\-ti\-gate thor\-ough\-ly, and if the wit\-ness has ac\-cused his broth\-er false\-ly,
    \textsuperscript{19}\thinspace{}then you must do to him as he in\-tend\-ed to do to his broth\-er.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{20:1}\thinspace{}When you go to war a\-gainst your enemies and see hors\-es, chariots, and an ar\-my larg\-er than yours, don't be a\-fraid of them, for \textsc{God}, who brought you out of Egypt, is with you.
    \textsuperscript{2}\thinspace{}When you ap\-proach the bat\-tle, the priest shall come for\-ward and speak to the ar\-my,
    \textsuperscript{3}\thinspace{}say\-ing, “Hear, O Israel, to\-day you ap\-proach bat\-tle a\-gainst your enemies. Don't be faint\-heart\-ed. Don't fear or pan\-ic or be ter\-ri\-fied of them,
    \textsuperscript{4}\thinspace{}for \textsc{God} goes with you to fight a\-gainst your enemies, to give you vic\-to\-ry.”
    \textsuperscript{5}\thinspace{}The officers shall speak to the ar\-my, say\-ing, “Has any man built a new house and not ded\-i\-cat\-ed it?
    \textsuperscript{6}\thinspace{}Has any man planted a vine\-yard and not yet enjoyed its fruit?
    \textsuperscript{7}\thinspace{}Has any man be\-come en\-gaged to a wom\-an and not yet mar\-ried her?
    \textsuperscript{8}\thinspace{}Is any man a\-fraid or faint\-heart\-ed? Let them re\-turn home.”
\end{para}

\begin{para}
    \textsuperscript{20:19}\thinspace{}When you be\-siege a cit\-y for a long time, don't de\-stroy its trees. Are the trees hu\-man that you should be\-siege them?
    \textsuperscript{20}\thinspace{}You may on\-ly cut down trees that you know to pro\-duce no food. Use them to con\-struct siegeworks a\-gainst the cit\-y un\-til it falls.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{21:15}\thinspace{}If a man has two wives, one loved and the oth\-er un\-loved, and both bear him sons, and the firstborn belongs to the un\-loved,
    \textsuperscript{16}\thinspace{}then, when his sons in\-her\-it what he has, he may not fa\-vor the son of the loved wife o\-ver the son of the un\-loved wife.
    \textsuperscript{17}\thinspace{}He must ac\-knowl\-edge the firstborn, the son of the un\-loved wife, by giv\-ing him a dou\-ble por\-tion of all he has. The right of the firstborn is his.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{22:1}\thinspace{}If you see your brother's an\-i\-mal straying, don't ig\-nore it. Bring it back to him.
    \textsuperscript{2}\thinspace{}If he lives far a\-way or you don't know who he is, bring it home and care for it un\-til your broth\-er looks for it, then re\-turn it to him.
    \textsuperscript{3}\thinspace{}Do the same with any\-thing else that your broth\-er loses and you find. Don't ig\-nore it.
    \textsuperscript{4}\thinspace{}Don't ig\-nore your brother's don\-key or ox when it collapses on the road. Help him lift it up a\-gain.
\end{para}

\begin{para}
    \textsuperscript{22:5}\thinspace{}A wom\-an shall not wear men's clothes, nei\-ther shall a man wear women's clothes, for \textsc{God} detests those who do it.
\end{para}

\begin{para}
    \textsuperscript{22:6}\thinspace{}If you come a\-cross a bird's nest with chicks or eggs, and the moth\-er is sit\-ting on them, don't take the moth\-er with the young.
    \textsuperscript{7}\thinspace{}You may take the young but let the moth\-er go so that you may pros\-per and have a long life.
\end{para}

\begin{para}
    \textsuperscript{22:8}\thinspace{}When you build a new house, con\-struct a rail\-ing a\-round your roof, lest you bring bloodguilt on your house if any\-one falls from it.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{23:9}\thinspace{}When you are encamped a\-gainst your enemies,
    \textsuperscript{12}\thinspace{}des\-ig\-nate a place out\-side the camp where you can go to re\-lieve your\-self.
    \textsuperscript{13}\thinspace{}Keep a dig\-ging tool in your e\-quip\-ment so that when you re\-lieve your\-self you can dig a hole and cov\-er your ex\-cre\-ment.
    \textsuperscript{14}\thinspace{}For \textsc{God} walks through your camp to pro\-tect you and to de\-feat your enemies. Your camp must be ho\-ly lest he see some\-thing in\-de\-cent a\-mong you and turn a\-way from you.
\end{para}

\begin{para}
    \textsuperscript{23:15}\thinspace{}Don't re\-turn a slave to his mas\-ter af\-ter he has es\-caped to you.
    \textsuperscript{16}\thinspace{}Let him live a\-mong you wher\-ev\-er he chooses. Don't mis\-treat him.
\end{para}

\begin{para}
    \textsuperscript{23:17}\thinspace{}No man or wom\-an of Israel shall be a shrine pros\-ti\-tute.
    \textsuperscript{18}\thinspace{}Don't bring the earn\-ings of a pros\-ti\-tute, fe\-male or male, in\-to the house of \textsc{God} to pay any vow, be\-cause both are de\-test\-a\-ble to \textsc{God}.
\end{para}

\begin{para}
    \textsuperscript{23:19}\thinspace{}Don't charge your broth\-er in\-ter\-est.
    \textsuperscript{20}\thinspace{}You may charge in\-ter\-est to a for\-eign\-er, but not to your broth\-er, so that \textsc{God} may bless you in eve\-ry\-thing you un\-der\-take.
\end{para}

\begin{para}
    \textsuperscript{23:21}\thinspace{}When you make a vow to \textsc{God}, don't de\-lay ful\-fil\-ling it, for \textsc{God} will sure\-ly re\-quire it of you, and you will be guilt\-y of sin.
    \textsuperscript{22}\thinspace{}But it's not a sin to re\-frain from mak\-ing a vow.
    \textsuperscript{23}\thinspace{}Be care\-ful to do what you say, be\-cause you made your prom\-ise free\-ly to \textsc{God} with your own mouth.
\end{para}

\begin{para}
    \textsuperscript{23:24}\thinspace{}If you en\-ter your neighbor's vine\-yard, you may eat your fill of grapes, but don't put any in your bas\-ket.
    \textsuperscript{25}\thinspace{}If you en\-ter your neighbor's stand\-ing grain, you may pluck grain with your hand, but don't har\-vest it with a sick\-le.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{24:1}\thinspace{}If a man marries a wom\-an, but she displeases him be\-cause he finds some in\-de\-cen\-cy in her, then let him write her a di\-vorce cer\-tif\-i\-cate, and send her out of his house.
    \textsuperscript{2}\thinspace{}When she leaves his house, she may mar\-ry an\-oth\-er man.
\end{para}

\begin{para}
    \textsuperscript{24:5}\thinspace{}When a man takes a new wife, he shall not go to war or be li\-a\-ble for any pub\-lic du\-ty for one year. He is free to stay home, so that he can bring joy to his new wife.
\end{para}

\begin{para}
    \textsuperscript{24:6}\thinspace{}Don't take a mill or an up\-per mill\-stone in pledge, for that would be tak\-ing a life in pledge.
\end{para}

\begin{para}
    \textsuperscript{24:10}\thinspace{}If you lend any\-thing to your neigh\-bor, don't en\-ter his house to col\-lect his pledge.
    \textsuperscript{11}\thinspace{}Re\-main out\-side and let your neigh\-bor bring the pledge out to you.
    \textsuperscript{12}\thinspace{}If the man is poor, don't keep his pledge o\-ver\-night.
    \textsuperscript{13}\thinspace{}Re\-turn it to him when the sun sets, so that he may sleep in his cloak and bless you.
\end{para}

\begin{para}
    \textsuperscript{24:14}\thinspace{}Don't op\-press a hired work\-er who is poor and needy, wheth\-er he is a broth\-er or a for\-eign\-er liv\-ing in your land.
    \textsuperscript{15}\thinspace{}Give him his wages each day be\-fore the sun sets, for he is poor and depends on them; lest he cry to \textsc{God} a\-gainst you, and you be\-come guilt\-y of sin.
\end{para}

\begin{para}
    \textsuperscript{24:16}\thinspace{}Fathers shall not be put to death for their chil\-dren, nor chil\-dren put to death for their fathers. Each one shall be put to death for his own sin.
\end{para}

\begin{para}
    \textsuperscript{24:17}\thinspace{}Don't de\-ny jus\-tice to a for\-eign\-er or an or\-phan, and don't take a widow's cloak as a pledge.
    \textsuperscript{18}\thinspace{}Re\-mem\-ber that you were a slave in Egypt, and that \textsc{God} redeemed you from there.
\end{para}

\begin{para}
    \textsuperscript{24:19}\thinspace{}When you reap your har\-vest and for\-get a sheaf in the field, don't go back to get it. Leave it for the for\-eign\-er, the or\-phan, and the wid\-ow, so that \textsc{God} may bless you in all the work of your hands.
    \textsuperscript{20}\thinspace{}When you beat the olives from your trees, don't go o\-ver the branches a\-gain.
    \textsuperscript{21}\thinspace{}When you gath\-er the grapes from your vine\-yard, don't go o\-ver the vines a\-gain. What re\-mains is for the for\-eign\-er, the or\-phan, and the wid\-ow.
    \textsuperscript{22}\thinspace{}Re\-mem\-ber that you were a slave in Egypt.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{25:4}\thinspace{}Don't muz\-zle the ox while it treads out the grain.
\end{para}

\begin{para}
    \textsuperscript{25:13}\thinspace{}Don't car\-ry differing weights in your bag, one heav\-y and one light.
    \textsuperscript{14}\thinspace{}Don't have differing meas\-ures in your house, one large and one small.
    \textsuperscript{15}\thinspace{}Use ac\-cu\-rate and hon\-est weights and meas\-ures.
    \textsuperscript{16}\thinspace{}For \textsc{God} detests those who are dis\-hon\-est in such things.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{26:2}\thinspace{}Put in\-to a bas\-ket some of the first fruit of the har\-vest.
    \textsuperscript{3}\thinspace{}Go to the priest.
    \textsuperscript{4}\thinspace{}The priest shall take the bas\-ket from your hand and set it down be\-fore the al\-tar of \textsc{God}.
    \textsuperscript{5}\thinspace{}Say be\-fore \textsc{God}, “My fa\-ther was a wan\-der\-ing Aramean. He went down to Egypt with a few peo\-ple, sojourned there, and be\-came a great, might\-y, and pop\-u\-lous na\-tion.
    \textsuperscript{6}\thinspace{}The Egyptians mistreated us, hu\-mil\-i\-at\-ed us, and im\-posed hard la\-bor on us.
    \textsuperscript{7}\thinspace{}Then we cried to \textsc{God}, the god of our fathers, and \textsc{God} heard our voice and saw our af\-flic\-tion, our toil, and our op\-pres\-sion,
    \textsuperscript{8}\thinspace{}\textsc{God} brought us out of Egypt with a might\-y hand, an out\-stretched arm, great deeds of ter\-ror, signs, and wonders.
    \textsuperscript{9}\thinspace{}He brought us to this place and gave us this land, a land flow\-ing with milk and hon\-ey.
    \textsuperscript{10}\thinspace{}I bring the first fruit of the ground, which you, O \textsc{God}, have giv\-en me.” Set it be\-fore \textsc{God} and wor\-ship be\-fore \textsc{God}.
    \textsuperscript{11}\thinspace{}You, the Levite, and the for\-eign\-er shall re\-joice in all the good things which \textsc{God} has giv\-en to you and your house\-hold.
\end{para}

\begin{para}
    \textsuperscript{26:12}\thinspace{}When you have fin\-ished pay\-ing all the tithes of your pro\-duce in the third year, give it to the Levite, the for\-eign\-er, the or\-phan, and the wid\-ow, so that they may eat in your towns and be sat\-is\-fied.
    \textsuperscript{13}\thinspace{}Say be\-fore \textsc{God}, “I have tak\-en the sa\-cred por\-tion out of my house. I have giv\-en it to the Levite, the for\-eign\-er, the or\-phan, and the wid\-ow, ac\-cord\-ing to all the commandments you gave me. I have not transgressed or for\-got\-ten your commandments.
    \textsuperscript{14}\thinspace{}I have not eat\-en any of the tithe while in mourn\-ing, or re\-moved any of it while un\-clean, or offered any of it to the dead. I have obeyed \textsc{God}. I have done all you commanded me.
    \textsuperscript{15}\thinspace{}Look down from heav\-en, your ho\-ly hab\-i\-ta\-tion, and bless your peo\-ple Israel and the land you have giv\-en us, as you swore to our fathers, a land flow\-ing with milk and hon\-ey.”
\end{para}

\chapter*{David}
\markboth{David}{David}
\addcontentsline{toc}{section}{David}\begin{changemargin}{.6in}{.6in}
  \begin{center}
    \itshape
    A se\-lec\-tion of proverbs tak\-en from the Book of Psalms in the Hebrew Bible.
  \end{center}
\end{changemargin}
\begin{para}
    \textsuperscript{2:11}\thinspace{}Serve \textsc{God} with fear and re\-joice with trem\-bling.
\end{para}

\begin{para}
    \textsuperscript{4:4}\thinspace{}In your an\-ger, do not sin. Pon\-der in your heart on your bed and be still.
    \textsuperscript{5}\thinspace{}Of\-fer the sacrifices of right\-eous\-ness and put your trust in \textsc{God}.
\end{para}

\begin{para}
    \textsuperscript{12:8}\thinspace{}The wick\-ed walk on eve\-ry side when vile\-ness is ex\-alt\-ed a\-mong the chil\-dren of men.
\end{para}

\begin{para}
    \textsuperscript{15:1}\thinspace{}\textsc{God}, who may so\-journ in your tent? Who may dwell on your ho\-ly hill?
    \textsuperscript{2}\thinspace{}He who walks with in\-teg\-ri\-ty and does what is right and speaks truth in his heart;
    \textsuperscript{3}\thinspace{}who does not slan\-der with his tongue and does no evil to his neigh\-bor and does not take up a re\-proach a\-gainst his friend;
    \textsuperscript{4}\thinspace{}who despises a vile per\-son but hon\-ors those who fear \textsc{God}; who keeps a prom\-ise e\-ven when it hurts and does not change;
    \textsuperscript{5}\thinspace{}who does not loan out his mon\-ey with in\-ter\-est and does not take a bribe a\-gainst the in\-no\-cent. He who does these things shall nev\-er be moved.
\end{para}

\begin{para}
    \textsuperscript{18:25}\thinspace{}\textsc{God} shows mer\-cy to the mer\-ci\-ful.
\end{para}

\begin{para}
    \textsuperscript{18:30}\thinspace{}The word of \textsc{God} proves true. He is a shield to all who take ref\-uge in him.
\end{para}

\begin{para}
    \textsuperscript{19:1}\thinspace{}The heavens de\-clare the glo\-ry of God and the skies show his hand\-i\-work.
\end{para}

\begin{para}
    \textsuperscript{19:7}\thinspace{}The law of \textsc{God} is per\-fect, re\-viv\-ing the soul. The tes\-ti\-mo\-ny of \textsc{God} is trust\-wor\-thy, mak\-ing wise the sim\-ple.
    \textsuperscript{8}\thinspace{}The precepts of \textsc{God} are right, re\-joic\-ing the heart. The com\-mand\-ment of \textsc{God} is pure, enlightening the eyes.
    \textsuperscript{9}\thinspace{}The fear of \textsc{God} is clean, en\-dur\-ing for\-ev\-er. The laws of \textsc{God} are true and right\-eous al\-to\-geth\-er.
\end{para}

\begin{para}
    \textsuperscript{27:14}\thinspace{}Wait for \textsc{God}. Be strong and let your heart take cour\-age. Wait for \textsc{God}.
\end{para}

\begin{para}
    \textsuperscript{30:5}\thinspace{}Weep\-ing may last for a night, but joy co\-mes in the morn\-ing.
\end{para}

\begin{para}
    \textsuperscript{32:10}\thinspace{}Man\-y sorrows come to the wick\-ed, but lov\-ing-kind\-ness surrounds him who trusts in \textsc{God}.
\end{para}

\begin{para}
    \textsuperscript{33:6}\thinspace{}By the word of \textsc{God} the heavens were made and all their host by the breath of his mouth.
\end{para}

\begin{para}
    \textsuperscript{34:13}\thinspace{}Keep your tongue from evil and your lips from speak\-ing de\-ceit.
    \textsuperscript{14}\thinspace{}De\-part from evil and do good. Seek peace and pur\-sue it.
\end{para}

\begin{para}
    \textsuperscript{37:1}\thinspace{}Don't wor\-ry a\-bout evildoers, and don't en\-vy wrongdoers.
    \textsuperscript{7}\thinspace{}Be still be\-fore \textsc{God} and wait pa\-tient\-ly for him. Don't wor\-ry a\-bout him who prospers or a\-bout him who carries out wick\-ed schemes.
    \textsuperscript{8}\thinspace{}Cease from an\-ger and for\-sake wrath. Don't wor\-ry: it on\-ly leads to evil\-do\-ing.
\end{para}

\begin{para}
    \textsuperscript{37:16}\thinspace{}A lit\-tle that the right\-eous has is bet\-ter than the a\-bun\-dance of man\-y wick\-ed.
\end{para}

\begin{para}
    \textsuperscript{37:21}\thinspace{}The wick\-ed borrows and does not pay back, but the right\-eous is gen\-er\-ous and gives.
\end{para}

\begin{para}
    \textsuperscript{39:1}\thinspace{}I will guard my ways so that I may not sin with my tongue. I will muz\-zle my mouth while the wick\-ed are in my pres\-ence.
\end{para}

\begin{para}
    \textsuperscript{41:1}\thinspace{}Bless\-ed is he who considers the poor. \textsc{God} will de\-liv\-er him in the day of trou\-ble.
\end{para}

\begin{para}
    \textsuperscript{46:1}\thinspace{}God is our ref\-uge and strength, a ver\-y pres\-ent help in trou\-ble.
    \textsuperscript{10}\thinspace{}Be still and know that I am God.
\end{para}

\begin{para}
    \textsuperscript{49:16}\thinspace{}Do not be a\-fraid when one is made rich, when the splen\-dor of his house increases.
    \textsuperscript{17}\thinspace{}For when he dies he will car\-ry noth\-ing a\-way. His glo\-ry will not de\-scend af\-ter him.
\end{para}

\begin{para}
    \textsuperscript{50:14}\thinspace{}Of\-fer to God a sac\-ri\-fice of thanks\-giv\-ing, and pay your vows to the Most High,
    \textsuperscript{15}\thinspace{}and call on me in the day of trou\-ble. I will de\-liv\-er you and you will glo\-ri\-fy me.
\end{para}

\begin{para}
    \textsuperscript{55:22}\thinspace{}Cast your bur\-den on \textsc{God} and he will sus\-tain you. He will nev\-er al\-low the right\-eous to be moved.
\end{para}

\begin{para}
    \textsuperscript{82:3}\thinspace{}De\-fend the poor and fa\-ther\-less. Do jus\-tice to the afflicted and des\-ti\-tute.
    \textsuperscript{4}\thinspace{}Res\-cue the poor and needy. De\-liv\-er them from the hand of the wick\-ed.
\end{para}

\begin{para}
    \textsuperscript{92:12}\thinspace{}The right\-eous will flour\-ish like the palm tree. He will grow like a ce\-dar in Lebanon.
\end{para}

\begin{para}
    \textsuperscript{96:2}\thinspace{}Sing to \textsc{God}, bless his name. Pro\-claim his sal\-va\-tion from day to day.
    \textsuperscript{3}\thinspace{}De\-clare his glo\-ry a\-mong the nations, his mar\-vel\-lous deeds a\-mong all the peoples.
\end{para}

\begin{para}
    \textsuperscript{100:1}\thinspace{}Make a joy\-ful noise to \textsc{God}, all the earth.
    \textsuperscript{2}\thinspace{}Serve \textsc{God} with glad\-ness. Come be\-fore him with sing\-ing.
\end{para}

\begin{para}
    \textsuperscript{102:17}\thinspace{}\textsc{God} regards the pray\-er of the des\-ti\-tute and does not des\-pise their pray\-er.
\end{para}

\begin{para}
    \textsuperscript{103:8}\thinspace{}\textsc{God} is mer\-ci\-ful and gra\-cious, slow to an\-ger, abounding in lov\-ing-kind\-ness.
\end{para}

\begin{para}
    \textsuperscript{105:1}\thinspace{}Give thanks to \textsc{God}. Call on his name. Make known his deeds a\-mong the peoples.
    \textsuperscript{2}\thinspace{}Sing to him. Sing praises to him. Talk a\-bout all his mar\-vel\-ous deeds.
    \textsuperscript{3}\thinspace{}Glo\-ry in his ho\-ly name. Let the hearts of those who seek \textsc{God} re\-joice.
    \textsuperscript{4}\thinspace{}Seek \textsc{God} and his strength. Seek his face con\-tin\-u\-al\-ly.
    \textsuperscript{5}\thinspace{}Re\-mem\-ber the wonders that he has done, his miracles and the judgments he has pro\-nounced.
\end{para}

\begin{para}
    \textsuperscript{106:1}\thinspace{}Praise \textsc{God}. Give thanks to \textsc{God}, for he is good; for his lov\-ing-kind\-ness endures for\-ev\-er.
\end{para}

\begin{para}
    \textsuperscript{111:10}\thinspace{}The fear of \textsc{God} is the be\-gin\-ning of wis\-dom. Those who o\-bey his commandments have good un\-der\-stand\-ing.
\end{para}

\begin{para}
    \textsuperscript{112:1}\thinspace{}Bless\-ed is the man who fears \textsc{God}, who delights great\-ly in his commandments.
    \textsuperscript{4}\thinspace{}For the up\-right, light arises in the dark\-ness. He is gra\-cious, mer\-ci\-ful and right\-eous.
    \textsuperscript{5}\thinspace{}It is well with the man who deals gen\-er\-ous\-ly and lends, who conducts his af\-fairs with jus\-tice.
\end{para}

\begin{para}
    \textsuperscript{118:8}\thinspace{}It is bet\-ter to take ref\-uge in \textsc{God} than to put con\-fi\-dence in man.
    \textsuperscript{9}\thinspace{}It is bet\-ter to take ref\-uge in \textsc{God} than to put con\-fi\-dence in prin\-ces.
\end{para}

\begin{para}
    \textsuperscript{122:6}\thinspace{}Pray for the peace of Je\-ru\-sa\-lem.
\end{para}

\begin{para}
    \textsuperscript{127:1}\thinspace{}Ex\-cept \textsc{God} build the house, they la\-bor in vain who build it. Ex\-cept \textsc{God} keep the cit\-y, the watch\-man stays a\-wake in vain.
    \textsuperscript{2}\thinspace{}It is vain for you to rise up ear\-ly, to stay up late, to eat the bread of toil---for he gives sleep to his be\-lov\-ed.
\end{para}

\begin{para}
    \textsuperscript{127:3}\thinspace{}Chil\-dren are a her\-it\-age from \textsc{God} and the fruit of the womb is a re\-ward.
\end{para}

\begin{para}
    \textsuperscript{133:1}\thinspace{}How good and pleas\-ant it is for brothers to live to\-geth\-er in u\-ni\-ty.
\end{para}

\begin{para}
    \textsuperscript{146:3}\thinspace{}Do not put your trust in prin\-ces or in the son of man, who can\-not save.
\end{para}

\chapter*{Solomon}
\markboth{Solomon}{Solomon}
\addcontentsline{toc}{section}{Solomon}\begin{changemargin}{.6in}{.6in}
  \begin{center}
    \itshape
    An a\-bridg\-ment of the proverbs at\-trib\-ut\-ed to Solomon.
  \end{center}
\end{changemargin}
\vspace{0pt plus 8\baselineskip{}}\section*{My Son}
\markright{My Son}
\begin{para}
    \textsuperscript{3:1}\thinspace{}My son, don't for\-get my teach\-ing, but let your heart keep my commandments,
    \textsuperscript{2}\thinspace{}for they will bring you man\-y days, years of life, and peace.
\end{para}

\begin{para}
    \textsuperscript{3:3}\thinspace{}Don't let kind\-ness and truth leave you. Tie them a\-round your neck. Write them on the tab\-let of your heart.
    \textsuperscript{4}\thinspace{}Then you will find fa\-vor and high re\-gard in the sight of God and man.
\end{para}

\begin{para}
    \textsuperscript{3:5}\thinspace{}Trust in \textsc{God} with all your heart, and don't lean on your own un\-der\-stand\-ing.
    \textsuperscript{6}\thinspace{}In all your ways ac\-knowl\-edge him, and he will make your paths straight.
    \textsuperscript{7}\thinspace{}Don't be wise in your own eyes. Fear \textsc{God}, and turn a\-way from evil.
    \textsuperscript{8}\thinspace{}It will be heal\-ing to your bod\-y, and re\-fresh\-ment to your bones.
\end{para}

\begin{para}
    \textsuperscript{3:9}\thinspace{}Hon\-or \textsc{God} with your wealth, and with the first\-fruits of all your pro\-duce,
    \textsuperscript{10}\thinspace{}then your barns will be filled with plen\-ty, and your vats will o\-ver\-flow with new wine.
\end{para}

\begin{para}
    \textsuperscript{3:11}\thinspace{}My son, don't re\-ject the dis\-ci\-pline of \textsc{God} or loathe his re\-proof;
    \textsuperscript{12}\thinspace{}for \textsc{God} reproves the one he loves, just as a fa\-ther reproves the son in whom he delights.
\end{para}

\begin{para}
    \textsuperscript{3:13}\thinspace{}Hap\-py is the man who finds wis\-dom, and the man who gains un\-der\-stand\-ing.
    \textsuperscript{14}\thinspace{}For she is more prof\-it\-a\-ble than sil\-ver, and her gain is bet\-ter than fi\-ne gold.
    \textsuperscript{15}\thinspace{}She is more pre\-cious than jewels, and noth\-ing you de\-sire compares with her.
    \textsuperscript{16}\thinspace{}Long life is in her right hand. In her left hand are rich\-es and hon\-or.
    \textsuperscript{17}\thinspace{}Her ways are pleas\-ant, and all her paths are peace\-ful.
    \textsuperscript{18}\thinspace{}She is a tree of life to those who take hold of her. Those who hold her fast are hap\-py.
\end{para}

\begin{para}
    \textsuperscript{3:19}\thinspace{}\textsc{God} founded the earth by wis\-dom. He established the heavens by un\-der\-stand\-ing.
    \textsuperscript{20}\thinspace{}By his knowl\-edge the depths broke o\-pen, and the clouds drip with dew.
\end{para}

\begin{para}
    \textsuperscript{3:21}\thinspace{}My son, don't lose sight of them. Keep sound wis\-dom and dis\-cre\-tion.
    \textsuperscript{22}\thinspace{}They will be life for your soul, and a\-dorn\-ment for your neck.
    \textsuperscript{23}\thinspace{}Then you will walk on your way se\-cure\-ly, and your foot won't stum\-ble.
    \textsuperscript{24}\thinspace{}When you lie down, you won't be a\-fraid. When you lie down, your sleep will be sweet.
\end{para}

\begin{para}
    \textsuperscript{3:25}\thinspace{}Don't fear sud\-den dis\-as\-ter or the ru\-in of the wick\-ed when it co\-mes,
    \textsuperscript{26}\thinspace{}for \textsc{God} will be your con\-fi\-dence, and will keep your foot from be\-ing caught.
\end{para}

\begin{para}
    \textsuperscript{3:27}\thinspace{}Don't with\-hold good from those who de\-serve it, when it is in your pow\-er to do it.
    \textsuperscript{28}\thinspace{}Don't say to your neigh\-bor, “Go, and come back. I'll give it to\-mor\-row,” when you have it with you.
    \textsuperscript{29}\thinspace{}Don't plan evil a\-gainst your neigh\-bor, who lives trustingly be\-side you.
\end{para}

\begin{para}
    \textsuperscript{3:30}\thinspace{}Don't con\-tend with a man with\-out cause, when he has done you no harm.
    \textsuperscript{31}\thinspace{}Don't en\-vy a vi\-o\-lent man or choose any of his ways,
    \textsuperscript{32}\thinspace{}for the de\-vi\-ous are an a\-bom\-i\-na\-tion to \textsc{God}, but he is a friend to the up\-right.
\end{para}

\begin{para}
    \textsuperscript{3:33}\thinspace{}The curse of \textsc{God} is on the house of the wick\-ed, but he blesses the dwell\-ing of the right\-eous.
    \textsuperscript{34}\thinspace{}He mocks the mock\-ers, but shows fa\-vor to the hum\-ble.
    \textsuperscript{35}\thinspace{}The wise will in\-her\-it hon\-or, but fools get shame.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{4:1}\thinspace{}Lis\-ten, sons, to a father's in\-struc\-tion, and pay at\-ten\-tion so that you may gain un\-der\-stand\-ing.
    \textsuperscript{2}\thinspace{}For I give you sound in\-struc\-tion. Don't a\-ban\-don my teach\-ing.
\end{para}

\begin{para}
    \textsuperscript{4:3}\thinspace{}When I was a son to my fa\-ther, ten\-der and the on\-ly son in my mother's sight,
    \textsuperscript{4}\thinspace{}he taught me, and said to me: “Let your heart hold on to my words. Keep my commandments and live.
    \textsuperscript{5}\thinspace{}Get wis\-dom. Get un\-der\-stand\-ing. Don't for\-get or turn a\-way from the words of my mouth.
    \textsuperscript{6}\thinspace{}Don't for\-sake her, and she will guard you. Love her, and she will watch o\-ver you.
    \textsuperscript{7}\thinspace{}The be\-gin\-ning of wis\-dom is this: Get wis\-dom, and what\-ev\-er else you get, get un\-der\-stand\-ing.
    \textsuperscript{8}\thinspace{}Prize her, and she will ex\-alt you. If you em\-brace her, she will hon\-or you.
    \textsuperscript{9}\thinspace{}She will place on your head a gar\-land of grace. She will pres\-ent you with a crown of beau\-ty.”
\end{para}

\begin{para}
    \textsuperscript{4:10}\thinspace{}Lis\-ten, my son, and ac\-cept my sayings, and the years of your life will be man\-y.
    \textsuperscript{11}\thinspace{}I have taught you in the way of wis\-dom. I have led you in up\-right paths.
    \textsuperscript{12}\thinspace{}When you walk, your steps won't be hindered. When you run, you won't stum\-ble.
    \textsuperscript{13}\thinspace{}Hold on to in\-struc\-tion. Don't let go. Guard her, for she is your life.
\end{para}

\begin{para}
    \textsuperscript{4:14}\thinspace{}Don't en\-ter the path of the wick\-ed, and don't walk in the way of evil men.
    \textsuperscript{15}\thinspace{}A\-void it. Don't trav\-el on it. Turn a\-way from it, and pass on.
    \textsuperscript{16}\thinspace{}For they can\-not sleep un\-less they do evil. They are robbed of sleep un\-less they make some\-one stum\-ble.
    \textsuperscript{17}\thinspace{}For they eat the bread of wick\-ed\-ness, and drink the wine of vi\-o\-lence.
    \textsuperscript{18}\thinspace{}But the path of the right\-eous is like the light of dawn, which shines brighter and brighter un\-til mid\-day.
    \textsuperscript{19}\thinspace{}The way of the wick\-ed is like dark\-ness. They don't know what makes them stum\-ble.
\end{para}

\begin{para}
    \textsuperscript{4:20}\thinspace{}My son, pay at\-ten\-tion to my words. Lis\-ten care\-ful\-ly to my sayings.
    \textsuperscript{21}\thinspace{}Don't lose sight of them. Keep them with\-in your heart.
    \textsuperscript{22}\thinspace{}For they are life to those who find them, and health to their whole bod\-y.
\end{para}

\begin{para}
    \textsuperscript{4:23}\thinspace{}Watch o\-ver your heart with all dil\-i\-gence, for the springs of life flow from it.
    \textsuperscript{24}\thinspace{}Put a\-way from you de\-ceit\-ful speech, and put de\-vi\-ous talk far from you.
\end{para}

\begin{para}
    \textsuperscript{4:25}\thinspace{}Let your eyes look di\-rect\-ly a\-head and your gaze straight in front of you.
    \textsuperscript{26}\thinspace{}Care\-ful\-ly con\-sid\-er the path of your feet, then all your ways will be sure.
    \textsuperscript{27}\thinspace{}Don't turn to the right or to the left. Keep your foot from evil.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{5:1}\thinspace{}My son, pay at\-ten\-tion to my wis\-dom. Lis\-ten care\-ful\-ly to my un\-der\-stand\-ing,
    \textsuperscript{2}\thinspace{}that you may main\-tain dis\-cre\-tion, and that your lips may guard knowl\-edge.
\end{para}

\begin{para}
    \textsuperscript{5:3}\thinspace{}The lips of a for\-bid\-den wom\-an drip hon\-ey, and her speech is smooth\-er than oil,
    \textsuperscript{4}\thinspace{}but in the end she's as bit\-ter as worm\-wood, and as sharp as a double-edged sword.
    \textsuperscript{5}\thinspace{}Her feet go down to death. Her steps lead straight to hell.
    \textsuperscript{6}\thinspace{}She does\-n't pon\-der the path of life. Her ways are un\-sta\-ble, and she does\-n't know it.
\end{para}

\begin{para}
    \textsuperscript{5:7}\thinspace{}Now then, my sons, lis\-ten to me, and don't de\-part from the words of my mouth.
    \textsuperscript{8}\thinspace{}Keep your way far from her. Don't go near the door of her house
    \textsuperscript{9}\thinspace{}lest you give your vig\-or to others and your years to some\-one cru\-el,
    \textsuperscript{10}\thinspace{}lest strangers feast on your wealth, your labors en\-rich the house of a for\-eign\-er,
    \textsuperscript{11}\thinspace{}and at the end of your life you groan when your flesh and bod\-y are con\-sumed.
    \textsuperscript{12}\thinspace{}You will say, “How I hat\-ed dis\-ci\-pline! How my heart de\-spised re\-proof!
    \textsuperscript{13}\thinspace{}I did\-n't o\-bey my teachers or lis\-ten to my instructors!
    \textsuperscript{14}\thinspace{}I am al\-most in ut\-ter ru\-in in the midst of the as\-sem\-bly and con\-gre\-ga\-tion.”
\end{para}

\begin{para}
    \textsuperscript{5:15}\thinspace{}Drink wa\-ter from your own cis\-tern, and fresh wa\-ter from your own well.
    \textsuperscript{16}\thinspace{}Should your springs be scattered a\-broad, streams of wa\-ter in the streets?
    \textsuperscript{17}\thinspace{}Let them be yours a\-lone, and not for strangers with you.
    \textsuperscript{18}\thinspace{}Let your foun\-tain be bless\-ed, and re\-joice in the wife of your youth,
    \textsuperscript{19}\thinspace{}as a lov\-ing deer and a grace\-ful doe. Let her breasts sat\-is\-fy you al\-ways, and be ex\-hil\-a\-rat\-ed al\-ways with her love.
\end{para}

\begin{para}
    \textsuperscript{5:20}\thinspace{}Why, my son, should you be ex\-hil\-a\-rat\-ed with a for\-bid\-den wom\-an, or em\-brace the bos\-om of a strang\-er?
    \textsuperscript{21}\thinspace{}For a man's ways are be\-fore the eyes of \textsc{God}, and he considers all his paths.
    \textsuperscript{22}\thinspace{}The iniquities of the wick\-ed en\-trap him. He is held in the cords of his sin.
    \textsuperscript{23}\thinspace{}He dies for lack of dis\-ci\-pline, and be\-cause of his great fol\-ly he will go a\-stray.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{6:1}\thinspace{}My son, if you have put up se\-cu\-ri\-ty for your neigh\-bor or have giv\-en a pledge for a strang\-er,
    \textsuperscript{2}\thinspace{}if you have been snared or caught with the words of your mouth,
    \textsuperscript{3}\thinspace{}then do this, my son, and free your\-self. You have fall\-en in\-to your neighbor's hands. Go, hum\-ble your\-self, and plead with your neigh\-bor.
    \textsuperscript{4}\thinspace{}Give your eyes no sleep and your eyelids no slum\-ber.
    \textsuperscript{5}\thinspace{}Save your\-self like a ga\-zelle from the hunter's hand, and like a bird from the fowler's hand.
\end{para}

\begin{para}
    \textsuperscript{6:6}\thinspace{}Go to the ant, you i\-dler. Ob\-serve its ways, and be wise.
    \textsuperscript{7}\thinspace{}With\-out chief, of\-fic\-er, or rul\-er,
    \textsuperscript{8}\thinspace{}it prepares its provisions in sum\-mer, and gathers its food dur\-ing har\-vest.
\end{para}

\begin{para}
    \textsuperscript{6:9}\thinspace{}How long will you lie down, you i\-dler? When will you get up from your sleep?
    \textsuperscript{10}\thinspace{}A lit\-tle sleep, a lit\-tle slum\-ber, a lit\-tle folding of the hands to rest,
    \textsuperscript{11}\thinspace{}and pov\-er\-ty will come up\-on you like a rob\-ber, and want like an armed man.
\end{para}

\begin{para}
    \textsuperscript{6:12}\thinspace{}A worth\-less per\-son, a wick\-ed man, goes a\-round speak\-ing per\-verse\-ly.
    \textsuperscript{14}\thinspace{}With per\-ver\-si\-ty in his heart, he devises evil con\-tin\-u\-al\-ly, and sows dis\-cord.
    \textsuperscript{15}\thinspace{}There\-fore ca\-lam\-i\-ty will strike him sud\-den\-ly. In\-stant\-ly he will be bro\-ken be\-yond heal\-ing.
\end{para}

\begin{para}
    \textsuperscript{6:16}\thinspace{}Sev\-en things are an a\-bom\-i\-na\-tion to \textsc{God}:
    \textsuperscript{17}\thinspace{}haugh\-ty eyes, a ly\-ing tongue, hands that shed in\-no\-cent blood,
    \textsuperscript{18}\thinspace{}a heart that devises wick\-ed plans, feet that run swift\-ly to evil,
    \textsuperscript{19}\thinspace{}a false wit\-ness who utters lies, and one who sows dis\-cord a\-mong brothers.
\end{para}

\begin{para}
    \textsuperscript{6:20}\thinspace{}My son, keep your father's com\-mand, and don't for\-sake your mother's teach\-ing.
    \textsuperscript{21}\thinspace{}Al\-ways bind them on your heart. Tie them a\-round your neck.
    \textsuperscript{22}\thinspace{}When you walk, they will guide you. When you sleep, they will watch o\-ver you. When you wake up, they will talk to you.
    \textsuperscript{23}\thinspace{}For this com\-mand is a lamp, this teach\-ing is a light, and cor\-rec\-tive dis\-ci\-pline is the way to life,
    \textsuperscript{24}\thinspace{}keep\-ing you from the evil wom\-an, from the flattering tongue of an a\-dul\-ter\-ess.
\end{para}

\begin{para}
    \textsuperscript{6:25}\thinspace{}Don't de\-sire her beau\-ty in your heart or let her cap\-ti\-vate you with her eyes.
    \textsuperscript{26}\thinspace{}For a prostitute's fee is on\-ly a loaf of bread, but a mar\-ried wom\-an hunts down a pre\-cious life.
    \textsuperscript{27}\thinspace{}Can a man em\-brace fire with\-out burn\-ing his clothes?
    \textsuperscript{28}\thinspace{}Or can he walk on hot coals with\-out scorching his feet?
    \textsuperscript{29}\thinspace{}So is he who goes in to his neighbor's wife. None who touches her will go un\-pun\-ished.
\end{para}

\begin{para}
    \textsuperscript{6:30}\thinspace{}Peo\-ple don't des\-pise a thief if he steals to sat\-is\-fy his hun\-ger,
    \textsuperscript{31}\thinspace{}but if caught, he must re\-pay sev\-en\-fold. He will give eve\-ry\-thing in his house.
\end{para}

\begin{para}
    \textsuperscript{6:32}\thinspace{}He who commits a\-dul\-ter\-y lacks sense, and destroys him\-self.
    \textsuperscript{33}\thinspace{}He will get wounds and dis\-hon\-or, and his dis\-grace won't be e\-rased.
    \textsuperscript{34}\thinspace{}For jeal\-ous\-y enrages a hus\-band, and he will show no mer\-cy in the day of venge\-ance.
    \textsuperscript{35}\thinspace{}He won't ac\-cept com\-pen\-sa\-tion. He will ref\-use, though you give man\-y gifts.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Relationships}
\markright{Relationships}
\begin{para}
    \textsuperscript{7:10}\thinspace{}A wom\-an met him, dressed as a pros\-ti\-tute, and cun\-ning of heart.
    \textsuperscript{13}\thinspace{}She seizes him, kisses him, and bra\-zen\-ly says to him,
    \textsuperscript{14}\thinspace{}“I had to of\-fer peace offerings. To\-day I've paid my vows.
    \textsuperscript{16}\thinspace{}I've spread my couch with coverings, with col\-ored linens from Egypt.
    \textsuperscript{18}\thinspace{}Come, let's drink our fill of love un\-til morn\-ing. Let's de\-light our\-selves with caresses.”
    \textsuperscript{21}\thinspace{}She seduces him with her per\-sua\-sive speech. With her flattering lips she lures him.
    \textsuperscript{22}\thinspace{}He follows her at once, like an ox go\-ing to the slaugh\-ter, like a deer step\-ping in\-to a trap.
    \textsuperscript{25}\thinspace{}Don't let your heart turn a\-side to her ways. Don't stray in\-to her paths.
    \textsuperscript{26}\thinspace{}For she has cast down man\-y victims, and all her slain are nu\-mer\-ous.
\end{para}

\begin{para}
    \textsuperscript{11:29}\thinspace{}He who troubles his own house will in\-her\-it the wind.
\end{para}

\begin{para}
    \textsuperscript{12:4}\thinspace{}An ex\-cel\-lent wife is her husband's crown, but she who brings shame is like rottenness in his bones.
\end{para}

\begin{para}
    \textsuperscript{13:20}\thinspace{}He who walks with wise men will be\-come wise, but the com\-pan\-ion of fools will suf\-fer harm.
\end{para}

\begin{para}
    \textsuperscript{13:24}\thinspace{}He who spares the rod hates his son, but he who loves him disciplines him dil\-i\-gent\-ly.
\end{para}

\begin{para}
    \textsuperscript{14:7}\thinspace{}Leave the pres\-ence of a fool, for you won't find knowl\-edge in his speech.
\end{para}

\begin{para}
    \textsuperscript{14:28}\thinspace{}A large pop\-u\-la\-tion is a king's glo\-ry, but with\-out peo\-ple a prince is ruined.
\end{para}

\begin{para}
    \textsuperscript{15:17}\thinspace{}Bet\-ter is a serv\-ing of vegetables where there is love, than a fattened ox served with ha\-tred.
\end{para}

\begin{para}
    \textsuperscript{16:7}\thinspace{}When a man's ways please \textsc{God}, he makes e\-ven his enemies to be at peace with him.
\end{para}

\begin{para}
    \textsuperscript{17:6}\thinspace{}Grand\-chil\-dren are the crown of old men, and fathers are the glo\-ry of their sons.
\end{para}

\begin{para}
    \textsuperscript{17:9}\thinspace{}He who conceals an of\-fense seeks love, but he who repeats a mat\-ter sep\-a\-rates close friends.
\end{para}

\begin{para}
    \textsuperscript{17:12}\thinspace{}Let a man meet a bear robbed of her cubs, rath\-er than a fool in his fol\-ly.
\end{para}

\begin{para}
    \textsuperscript{17:21}\thinspace{}He who sires a fool does so to his own sor\-row, and the fa\-ther of a fool has no joy.
    \textsuperscript{25}\thinspace{}A fool\-ish son is a grief to his fa\-ther, and bit\-ter\-ness to her who bore him.
\end{para}

\begin{para}
    \textsuperscript{18:1}\thinspace{}He who isolates him\-self seeks his own de\-sire, and rebels a\-gainst all sound wis\-dom.
\end{para}

\begin{para}
    \textsuperscript{18:22}\thinspace{}He who finds a wife finds a good thing, and obtains fa\-vor from \textsc{God}.
\end{para}

\begin{para}
    \textsuperscript{18:24}\thinspace{}He who has man\-y friends co\-mes to ru\-in, but there is a friend who stays clos\-er than a broth\-er.
\end{para}

\begin{para}
    \textsuperscript{19:14}\thinspace{}A house and wealth are in\-her\-it\-ed from fathers, but a pru\-dent wife is from \textsc{God}.
\end{para}

\begin{para}
    \textsuperscript{20:2}\thinspace{}A king's ter\-ror is like the lion's roar; he who provokes him to an\-ger forfeits his own life.
\end{para}

\begin{para}
    \textsuperscript{20:19}\thinspace{}A gos\-sip reveals secrets, there\-fore don't as\-so\-ci\-ate with him who talks too much.
\end{para}

\begin{para}
    \textsuperscript{22:6}\thinspace{}Train up a child in the way he should go; e\-ven when he is old he won't de\-part from it.
\end{para}

\begin{para}
    \textsuperscript{25:17}\thinspace{}Sel\-dom set foot in your neighbor's house, lest he grow wea\-ry of you, and hate you.
\end{para}

\begin{para}
    \textsuperscript{29:3}\thinspace{}He who loves wis\-dom brings joy to his fa\-ther, but a com\-pan\-ion of prostitutes wastes his wealth.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Instruction}
\markright{Instruction}
\begin{para}
    \textsuperscript{8:1}\thinspace{}Does\-n't wis\-dom call, and un\-der\-stand\-ing raise her voice?
    \textsuperscript{4}\thinspace{}“To you, O men, I call, and my voice is to the sons of men.
    \textsuperscript{5}\thinspace{}O na\-ive ones, learn pru\-dence. O fools, learn sense.
    \textsuperscript{10}\thinspace{}Take my in\-struc\-tion in\-stead of sil\-ver, and knowl\-edge rath\-er than choice gold.
    \textsuperscript{11}\thinspace{}For wis\-dom is bet\-ter than jewels, and all de\-sir\-a\-ble things can\-not com\-pare to it.
    \textsuperscript{22}\thinspace{}\textsc{God} pos\-sessed me at the be\-gin\-ning of his way, be\-fore his works of old.
    \textsuperscript{27}\thinspace{}When he established the heavens, I was there.
    \textsuperscript{32}\thinspace{}Now, my sons, lis\-ten to me, for bless\-ed are those who keep my ways.
    \textsuperscript{33}\thinspace{}Lis\-ten to in\-struc\-tion and be wise. Don't ne\-glect it.
    \textsuperscript{35}\thinspace{}For he who finds me finds life, and will ob\-tain fa\-vor of \textsc{God}.”
\end{para}

\begin{para}
    \textsuperscript{9:8}\thinspace{}Don't re\-prove a scoff\-er, or he will hate you. Re\-prove a wise man, and he will love you.
    \textsuperscript{9}\thinspace{}In\-struct a wise man, and he will be wis\-er still. Teach a right\-eous man, and he will in\-crease in learn\-ing.
    \textsuperscript{10}\thinspace{}The fear of \textsc{God} is the be\-gin\-ning of wis\-dom.
\end{para}

\begin{para}
    \textsuperscript{10:17}\thinspace{}He who heeds in\-struc\-tion is on the path of life, but he who ignores re\-proof goes a\-stray.
\end{para}

\begin{para}
    \textsuperscript{12:15}\thinspace{}The fool's way is right in his own eyes, but a wise man listens to coun\-sel.
\end{para}

\begin{para}
    \textsuperscript{13:14}\thinspace{}The teach\-ing of the wise is a foun\-tain of life, turn\-ing one a\-way from the snares of death.
\end{para}

\begin{para}
    \textsuperscript{13:18}\thinspace{}Pov\-er\-ty and shame will come to him who ignores in\-struc\-tion, but he who heeds re\-proof will be honored.
\end{para}

\begin{para}
    \textsuperscript{17:10}\thinspace{}A re\-buke goes deeper in\-to one who has un\-der\-stand\-ing than a hun\-dred lashes in\-to a fool.
\end{para}

\begin{para}
    \textsuperscript{19:27}\thinspace{}Cease listening to in\-struc\-tion, my son, and you will stray from the words of knowl\-edge.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Wise and Foolish}
\markright{Wise and Foolish}
\begin{para}
    \textsuperscript{10:1}\thinspace{}A wise son makes a fa\-ther glad, but a fool\-ish son brings grief to his moth\-er.
\end{para}

\begin{para}
    \textsuperscript{10:23}\thinspace{}Do\-ing wick\-ed\-ness is like sport to a fool, but wis\-dom is pleas\-ure to a man of un\-der\-stand\-ing.
\end{para}

\begin{para}
    \textsuperscript{11:29}\thinspace{}The fool will be serv\-ant to the wise of heart.
\end{para}

\begin{para}
    \textsuperscript{12:16}\thinspace{}A fool's dis\-pleas\-ure is known at once, but a pru\-dent man ignores an in\-sult.
\end{para}

\begin{para}
    \textsuperscript{14:15}\thinspace{}The na\-ive believes eve\-ry\-thing, but the pru\-dent considers his steps.
    \textsuperscript{16}\thinspace{}A wise man is cau\-tious and turns a\-way from evil, but a fool is reck\-less and care\-less.
\end{para}

\begin{para}
    \textsuperscript{17:2}\thinspace{}A pru\-dent serv\-ant will rule o\-ver a shame\-ful son, and will share the in\-her\-it\-ance a\-mong the brothers.
\end{para}

\begin{para}
    \textsuperscript{22:3}\thinspace{}The pru\-dent sees dan\-ger and hides him\-self, but the na\-ive go on and suf\-fer for it.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Righteous and Wicked}
\markright{Righteous and Wicked}
\begin{para}
    \textsuperscript{10:2}\thinspace{}Ill-gotten gains do not prof\-it, but right\-eous\-ness delivers from death.
    \textsuperscript{3}\thinspace{}\textsc{God} will not let the right\-eous go hun\-gry, but he thwarts the crav\-ing of the wick\-ed.
\end{para}

\begin{para}
    \textsuperscript{10:7}\thinspace{}The mem\-o\-ry of the right\-eous is a bless\-ing, but the name of the wick\-ed will rot.
\end{para}

\begin{para}
    \textsuperscript{10:20}\thinspace{}The tongue of the right\-eous is like choice sil\-ver; the heart of the wick\-ed is of lit\-tle worth.
    \textsuperscript{21}\thinspace{}The lips of the right\-eous feed man\-y, but fools die for lack of sense.
\end{para}

\begin{para}
    \textsuperscript{10:27}\thinspace{}The fear of \textsc{God} prolongs life, but the years of the wick\-ed will be shortened.
\end{para}

\begin{para}
    \textsuperscript{11:5}\thinspace{}The right\-eous\-ness of the blame\-less will di\-rect his way, but the wick\-ed will fall by his own wick\-ed\-ness.
    \textsuperscript{6}\thinspace{}The right\-eous\-ness of the up\-right will res\-cue them, but the treach\-er\-ous will be trapped by their own desires.
    \textsuperscript{8}\thinspace{}The right\-eous is delivered from trou\-ble, but the wick\-ed takes his place.
\end{para}

\begin{para}
    \textsuperscript{11:18}\thinspace{}The wick\-ed earns de\-cep\-tive wages, but he who sows right\-eous\-ness gets a sure re\-ward.
\end{para}

\begin{para}
    \textsuperscript{11:23}\thinspace{}The de\-sire of the right\-eous turns out well, but the ex\-pec\-ta\-tion of the wick\-ed is wrath.
\end{para}

\begin{para}
    \textsuperscript{11:31}\thinspace{}If the right\-eous will be rewarded on earth, how much more the wick\-ed and the sin\-ner.
\end{para}

\begin{para}
    \textsuperscript{12:3}\thinspace{}A man will not be established by wick\-ed\-ness, but the root of the right\-eous will not be moved.
    \textsuperscript{7}\thinspace{}The wick\-ed are overthrown and are no more, but the house of the right\-eous will stand.
\end{para}

\begin{para}
    \textsuperscript{12:26}\thinspace{}The right\-eous is a guide to his neigh\-bor, but the way of the wick\-ed leads them a\-stray.
\end{para}

\begin{para}
    \textsuperscript{13:21}\thinspace{}Dis\-as\-ter pursues sinners, but the right\-eous will be rewarded with good.
\end{para}

\begin{para}
    \textsuperscript{13:25}\thinspace{}The right\-eous has e\-nough to sat\-is\-fy his ap\-pe\-tite, but the stom\-ach of the wick\-ed is emp\-ty.
\end{para}

\begin{para}
    \textsuperscript{14:19}\thinspace{}The evil bow down be\-fore the good, and the wick\-ed at the gates of the right\-eous.
\end{para}

\begin{para}
    \textsuperscript{14:22}\thinspace{}Won't they who de\-vise evil go a\-stray? But those who de\-vise good will find kind\-ness and truth.
\end{para}

\begin{para}
    \textsuperscript{14:34}\thinspace{}Right\-eous\-ness exalts a na\-tion, but sin is a dis\-grace to any peo\-ple.
\end{para}

\begin{para}
    \textsuperscript{15:8}\thinspace{}The sac\-ri\-fice of the wick\-ed is an a\-bom\-i\-na\-tion to \textsc{God}, but the pray\-er of the up\-right is his de\-light.
\end{para}

\begin{para}
    \textsuperscript{21:15}\thinspace{}When jus\-tice is done, it is a joy to the right\-eous, but a ter\-ror to evildoers.
\end{para}

\begin{para}
    \textsuperscript{28:1}\thinspace{}The wick\-ed flee when no one pursues, but the right\-eous are as bold as a li\-on.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Diligence}
\markright{Diligence}
\begin{para}
    \textsuperscript{10:4}\thinspace{}A slack hand causes pov\-er\-ty, but a dil\-i\-gent hand makes rich.
    \textsuperscript{5}\thinspace{}A son who gathers in the sum\-mer is pru\-dent, but one who sleeps dur\-ing har\-vest is shame\-ful.
\end{para}

\begin{para}
    \textsuperscript{10:26}\thinspace{}Like vin\-e\-gar to the teeth and smoke to the eyes, so is the i\-dler to those who send him.
\end{para}

\begin{para}
    \textsuperscript{12:24}\thinspace{}The hand of the dil\-i\-gent will rule, but the sloth\-ful will be put to forced la\-bor.
\end{para}

\begin{para}
    \textsuperscript{13:4}\thinspace{}The i\-dler craves, yet gets noth\-ing, but the dil\-i\-gent is ful\-ly sat\-is\-fied.
\end{para}

\begin{para}
    \textsuperscript{14:23}\thinspace{}There is prof\-it in all la\-bor, but mer\-e talk leads on\-ly to pov\-er\-ty.
\end{para}

\begin{para}
    \textsuperscript{16:26}\thinspace{}A worker's ap\-pe\-tite works for him, for his hun\-ger urges him on.
\end{para}

\begin{para}
    \textsuperscript{18:9}\thinspace{}Who\-ev\-er is slack in his work is broth\-er to him who destroys.
\end{para}

\begin{para}
    \textsuperscript{19:15}\thinspace{}La\-zi\-ness brings on deep sleep, and an i\-dle man will suf\-fer hun\-ger.
\end{para}

\begin{para}
    \textsuperscript{19:24}\thinspace{}The i\-dler bur\-ies his hand in the dish, but will not e\-ven bring it back to his mouth.
\end{para}

\begin{para}
    \textsuperscript{20:4}\thinspace{}The i\-dler does\-n't plow in sea\-son, so he looks dur\-ing har\-vest, and has noth\-ing.
\end{para}

\begin{para}
    \textsuperscript{20:13}\thinspace{}Don't love sleep, lest you come to pov\-er\-ty. O\-pen your eyes, and you will have plen\-ty of food.
\end{para}

\begin{para}
    \textsuperscript{22:13}\thinspace{}The i\-dler says, “There's a li\-on out\-side. I'll be killed in the streets.”
\end{para}

\begin{para}
    \textsuperscript{26:14}\thinspace{}As a door turns on its hinges, so does the i\-dler on his bed.
\end{para}

\begin{para}
    \textsuperscript{26:16}\thinspace{}The i\-dler is wis\-er in his own eyes than sev\-en men who can an\-swer sen\-si\-bly.
\end{para}

\begin{para}
    \textsuperscript{27:23}\thinspace{}Know well the con\-di\-tion of your flocks, and pay at\-ten\-tion to your herds;
    \textsuperscript{24}\thinspace{}for rich\-es aren't for\-ev\-er and a crown does\-n't en\-dure to all generations.
\end{para}

\begin{para}
    \textsuperscript{28:19}\thinspace{}He who works his land will have plen\-ty of food, but he who follows worth\-less pursuits will have plen\-ty of pov\-er\-ty.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Integrity}
\markright{Integrity}
\begin{para}
    \textsuperscript{10:9}\thinspace{}He who walks in in\-teg\-ri\-ty walks se\-cure\-ly, but he who perverts his ways will be found out.
\end{para}

\begin{para}
    \textsuperscript{11:3}\thinspace{}The in\-teg\-ri\-ty of the up\-right will guide them, but the per\-ver\-si\-ty of the treach\-er\-ous will de\-stroy them.
\end{para}

\begin{para}
    \textsuperscript{20:7}\thinspace{}A right\-eous man who walks in his in\-teg\-ri\-ty; bless\-ed are his chil\-dren af\-ter him.
\end{para}

\begin{para}
    \textsuperscript{21:3}\thinspace{}To do right\-eous\-ness and jus\-tice is more ac\-cept\-a\-ble to \textsc{God} than sac\-ri\-fice.
    \textsuperscript{27}\thinspace{}The sac\-ri\-fice of the wick\-ed is an a\-bom\-i\-na\-tion; how much more when he brings it with evil in\-tent.
\end{para}

\begin{para}
    \textsuperscript{28:6}\thinspace{}Bet\-ter is the poor who walks in his in\-teg\-ri\-ty, than the rich who is crook\-ed in his ways.
\end{para}

\begin{para}
    \textsuperscript{28:18}\thinspace{}He who walks in in\-teg\-ri\-ty will be delivered, but he who is crook\-ed will sud\-den\-ly fall.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Contention}
\markright{Contention}
\begin{para}
    \textsuperscript{10:12}\thinspace{}Ha\-tred stirs up con\-flict, but love co\-vers all offenses.
\end{para}

\begin{para}
    \textsuperscript{15:18}\thinspace{}A hot-tempered man stirs up con\-flict, but the slow to an\-ger calms con\-ten\-tion.
\end{para}

\begin{para}
    \textsuperscript{17:14}\thinspace{}Starting a con\-flict is like re\-leas\-ing a flood, so a\-ban\-don the dis\-pute be\-fore it breaks out.
\end{para}

\begin{para}
    \textsuperscript{18:18}\thinspace{}Cast\-ing the lot ends quarrels, and decides be\-tween pow\-er\-ful opponents.
\end{para}

\begin{para}
    \textsuperscript{18:19}\thinspace{}An offended broth\-er is harder to win than a strong cit\-y, and contentions are like the bars of a cas\-tle.
\end{para}

\begin{para}
    \textsuperscript{20:3}\thinspace{}Stay\-ing a\-way from strife is an hon\-or for a man, but any fool will quar\-rel.
\end{para}

\begin{para}
    \textsuperscript{21:9}\thinspace{}It's bet\-ter to live in a cor\-ner of the roof than in a house shared with a con\-ten\-tious wom\-an.
    \textsuperscript{19}\thinspace{}It's bet\-ter to live in a des\-ert than with a con\-ten\-tious and complaining wom\-an.
\end{para}

\begin{para}
    \textsuperscript{22:10}\thinspace{}Drive out the mocker, and con\-ten\-tion will de\-part, and quarrels and dis\-hon\-or will cease.
\end{para}

\begin{para}
    \textsuperscript{26:20}\thinspace{}With\-out wood the fire goes out, and with\-out a gos\-sip con\-ten\-tion dies down.
    \textsuperscript{21}\thinspace{}Like char\-coal to hot embers and wood to fire, so is a con\-ten\-tious man for kin\-dling strife.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Understanding}
\markright{Understanding}
\begin{para}
    \textsuperscript{10:13}\thinspace{}Wis\-dom is found on the lips of the dis\-cern\-ing, but a rod is for the back of him who lacks sense.
\end{para}

\begin{para}
    \textsuperscript{13:16}\thinspace{}Eve\-ry pru\-dent man works with knowl\-edge, but a fool flaunts his fol\-ly.
\end{para}

\begin{para}
    \textsuperscript{14:12}\thinspace{}There is a way that seems right to a man, but its end is the way to death.
\end{para}

\begin{para}
    \textsuperscript{16:16}\thinspace{}How much bet\-ter to get wis\-dom than gold. To get un\-der\-stand\-ing is pref\-er\-a\-ble to sil\-ver.
\end{para}

\begin{para}
    \textsuperscript{18:13}\thinspace{}He who answers be\-fore he listens: it is fol\-ly and shame to him.
\end{para}

\begin{para}
    \textsuperscript{18:15}\thinspace{}The mind of the pru\-dent acquires knowl\-edge, and the ear of the wise seeks knowl\-edge.
\end{para}

\begin{para}
    \textsuperscript{19:2}\thinspace{}En\-thu\-si\-asm with\-out knowl\-edge is not good, and hast\-y feet go a\-stray.
\end{para}

\begin{para}
    \textsuperscript{19:20}\thinspace{}Lis\-ten to coun\-sel and re\-ceive in\-struc\-tion so that you may be wise the rest of your days.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Speech}
\markright{Speech}
\begin{para}
    \textsuperscript{10:18}\thinspace{}He who conceals ha\-tred has ly\-ing lips, and he who spreads slan\-der is a fool.
\end{para}

\begin{para}
    \textsuperscript{10:19}\thinspace{}When there are man\-y words, trans\-gres\-sion is un\-a\-void\-a\-ble, but he who restrains his lips is wise.
\end{para}

\begin{para}
    \textsuperscript{11:12}\thinspace{}He who belittles his neigh\-bor lacks sense, but a man of un\-der\-stand\-ing re\-mains si\-lent.
    \textsuperscript{13}\thinspace{}A gos\-sip goes a\-bout revealing secrets, but he who is trust\-wor\-thy conceals a mat\-ter.
\end{para}

\begin{para}
    \textsuperscript{11:22}\thinspace{}A beau\-ti\-ful wom\-an with\-out dis\-cre\-tion is like a gold ring in a pig's snout.
\end{para}

\begin{para}
    \textsuperscript{12:13}\thinspace{}An evil man is en\-snared by the trans\-gres\-sion of his lips, but the right\-eous will es\-cape from trou\-ble.
\end{para}

\begin{para}
    \textsuperscript{12:18}\thinspace{}There is one who speaks rash\-ly like thrusts of a sword, but the tongue of the wise brings heal\-ing.
\end{para}

\begin{para}
    \textsuperscript{12:23}\thinspace{}A pru\-dent man conceals knowl\-edge, but a fool's heart proclaims fol\-ly.
\end{para}

\begin{para}
    \textsuperscript{13:3}\thinspace{}He who guards his mouth preserves his life, but he who opens wide his lips co\-mes to ru\-in.
\end{para}

\begin{para}
    \textsuperscript{14:3}\thinspace{}In the fool's mouth is a rod for his back, but the lips of the wise will pro\-tect them.
\end{para}

\begin{para}
    \textsuperscript{15:1}\thinspace{}A gen\-tle an\-swer turns a\-way wrath, but a harsh word stirs up an\-ger.
    \textsuperscript{4}\thinspace{}A sooth\-ing tongue is a tree of life, but a per\-verse tongue breaks the spir\-it.
\end{para}

\begin{para}
    \textsuperscript{15:23}\thinspace{}A man finds joy in giv\-ing an apt an\-swer; and a time\-ly word, how good it is!
\end{para}

\begin{para}
    \textsuperscript{15:28}\thinspace{}The heart of the right\-eous ponders how to an\-swer, but the mouth of the wick\-ed blurts out evil things.
\end{para}

\begin{para}
    \textsuperscript{16:24}\thinspace{}Pleas\-ant words are a hon\-ey\-comb: sweet to the soul, and health to the bod\-y.
\end{para}

\begin{para}
    \textsuperscript{16:28}\thinspace{}A per\-verse man spreads strife, and a gos\-sip sep\-a\-rates close friends.
\end{para}

\begin{para}
    \textsuperscript{17:27}\thinspace{}He who restrains his words has knowl\-edge, and he who has a cool spir\-it is a man of un\-der\-stand\-ing.
    \textsuperscript{28}\thinspace{}E\-ven a fool is con\-sid\-ered wise when he keeps si\-lent. When he closes his lips, he is con\-sid\-ered pru\-dent.
\end{para}

\begin{para}
    \textsuperscript{18:2}\thinspace{}A fool does not de\-light in un\-der\-stand\-ing, but on\-ly in air\-ing his o\-pin\-ion.
    \textsuperscript{6}\thinspace{}His lips bring strife, and his mouth invites a beat\-ing.
    \textsuperscript{7}\thinspace{}His mouth is his ru\-in, and his lips are the snare of his soul.
\end{para}

\begin{para}
    \textsuperscript{20:15}\thinspace{}There is gold, and an a\-bun\-dance of jewels, but lips of knowl\-edge are a rare treas\-ure.
\end{para}

\begin{para}
    \textsuperscript{21:23}\thinspace{}He who guards his mouth and tongue keeps him\-self out of trou\-ble.
\end{para}

\begin{para}
    \textsuperscript{22:11}\thinspace{}He who loves pu\-ri\-ty of heart, and speaks gra\-cious\-ly will have the king as a friend.
\end{para}

\begin{para}
    \textsuperscript{26:18}\thinspace{}Like a mad\-man who throws firebrands, ar\-rows, and death,
    \textsuperscript{19}\thinspace{}is the man who deceives his neigh\-bor, and says, “I was on\-ly jok\-ing!”
\end{para}

\begin{para}
    \textsuperscript{27:14}\thinspace{}He who blesses his neigh\-bor with a loud voice ear\-ly in the morn\-ing will be counted as curs\-ing.
\end{para}

\begin{para}
    \textsuperscript{29:20}\thinspace{}Do you see a man who is hast\-y in his words? There is more hope for a fool than for him.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Honesty}
\markright{Honesty}
\begin{para}
    \textsuperscript{11:1}\thinspace{}Dis\-hon\-est scales are an a\-bom\-i\-na\-tion to \textsc{God}, but an ac\-cu\-rate weight is his de\-light.
\end{para}

\begin{para}
    \textsuperscript{12:22}\thinspace{}Ly\-ing lips are an a\-bom\-i\-na\-tion to \textsc{God}, but those who act faith\-ful\-ly are his de\-light.
\end{para}

\begin{para}
    \textsuperscript{16:11}\thinspace{}A just bal\-ance and scales be\-long to \textsc{God}, all the weights in the bag are his con\-cern.
\end{para}

\begin{para}
    \textsuperscript{17:15}\thinspace{}Ac\-quit\-ting the guilt\-y, and condemning the right\-eous are both a\-like an a\-bom\-i\-na\-tion to \textsc{God}.
\end{para}

\begin{para}
    \textsuperscript{17:23}\thinspace{}A wick\-ed man takes a bribe in se\-cret to per\-vert the ways of jus\-tice.
\end{para}

\begin{para}
    \textsuperscript{18:5}\thinspace{}Be\-ing par\-tial to the wick\-ed is not good, nei\-ther is de\-priv\-ing the right\-eous of jus\-tice.
\end{para}

\begin{para}
    \textsuperscript{19:5}\thinspace{}A false wit\-ness will not go un\-pun\-ished, and he who speaks lies will not es\-cape.
\end{para}

\begin{para}
    \textsuperscript{20:17}\thinspace{}Bread gained by de\-ceit is sweet to a man, but af\-ter\-ward his mouth will be full of grav\-el.
\end{para}

\begin{para}
    \textsuperscript{20:23}\thinspace{}Differing weights are an a\-bom\-i\-na\-tion to \textsc{God}, and a false scale is not good.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Humility}
\markright{Humility}
\begin{para}
    \textsuperscript{11:2}\thinspace{}When pride co\-mes, then co\-mes dis\-grace, but with the hum\-ble is wis\-dom.
\end{para}

\begin{para}
    \textsuperscript{12:9}\thinspace{}Bet\-ter is he who is light\-ly esteemed, and has a serv\-ant, than he that hon\-ors him\-self, and lacks bread.
\end{para}

\begin{para}
    \textsuperscript{15:25}\thinspace{}\textsc{God} will tear down the house of the proud, but will pro\-tect the borders of the wid\-ow.
\end{para}

\begin{para}
    \textsuperscript{16:5}\thinspace{}Eve\-ry\-one who is proud in heart is an a\-bom\-i\-na\-tion to \textsc{God}; be as\-sured, he will not go un\-pun\-ished.
\end{para}

\begin{para}
    \textsuperscript{16:18}\thinspace{}Pride co\-mes be\-fore de\-struc\-tion, and an ar\-ro\-gant spir\-it be\-fore a fall.
\end{para}

\begin{para}
    \textsuperscript{16:19}\thinspace{}It is bet\-ter to be hum\-ble in spir\-it a\-mong the low\-ly, than to di\-vide the spoil with the proud.
\end{para}

\begin{para}
    \textsuperscript{18:12}\thinspace{}Be\-fore de\-struc\-tion a man's heart is haugh\-ty, but hu\-mil\-i\-ty co\-mes be\-fore hon\-or.
\end{para}

\begin{para}
    \textsuperscript{21:30}\thinspace{}No wis\-dom, no un\-der\-stand\-ing, and no coun\-sel will pre\-vail a\-gainst \textsc{God}.
\end{para}

\begin{para}
    \textsuperscript{22:4}\thinspace{}The re\-ward for hu\-mil\-i\-ty and fear of \textsc{God} is rich\-es, hon\-or, and life.
\end{para}

\begin{para}
    \textsuperscript{26:12}\thinspace{}Do you see a man wise in his own eyes? There is more hope for a fool than for him.
\end{para}

\begin{para}
    \textsuperscript{27:1}\thinspace{}Don't boast a\-bout to\-mor\-row, for you don't know what a day may bring.
\end{para}

\begin{para}
    \textsuperscript{27:2}\thinspace{}Let an\-oth\-er praise you, and not your own mouth; a strang\-er, and not your own lips.
\end{para}

\begin{para}
    \textsuperscript{29:23}\thinspace{}A man's pride will bring him low, but a hum\-ble spir\-it will ob\-tain hon\-or.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Counsel}
\markright{Counsel}
\begin{para}
    \textsuperscript{11:14}\thinspace{}With\-out guid\-ance the peo\-ple fall, but with man\-y counselors there is safe\-ty.
\end{para}

\begin{para}
    \textsuperscript{14:27}\thinspace{}The fear of \textsc{God} is a foun\-tain of life, turn\-ing one a\-way from the snares of death.
\end{para}

\begin{para}
    \textsuperscript{15:22}\thinspace{}With\-out coun\-sel plans fail, but with man\-y advisers they suc\-ceed.
\end{para}

\begin{para}
    \textsuperscript{16:33}\thinspace{}The lot is cast in\-to the lap, but its eve\-ry de\-ci\-sion is from \textsc{God}.
\end{para}

\begin{para}
    \textsuperscript{20:18}\thinspace{}Es\-tab\-lish plans by coun\-sel, and make war with wise guid\-ance.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Finance}
\markright{Finance}
\begin{para}
    \textsuperscript{11:15}\thinspace{}He who is guar\-an\-tor for a strang\-er will sure\-ly suf\-fer for it, but he who hates such agreements is se\-cure.
\end{para}

\begin{para}
    \textsuperscript{11:28}\thinspace{}He who trusts in his rich\-es will fall, but the right\-eous will flour\-ish like a green leaf.
\end{para}

\begin{para}
    \textsuperscript{13:7}\thinspace{}There is one who pretends to be rich, yet has noth\-ing. An\-oth\-er pretends to be poor, yet has great wealth.
\end{para}

\begin{para}
    \textsuperscript{13:11}\thinspace{}Wealth obtained by fraud will dwin\-dle, but he who gathers by la\-bor will in\-crease it.
\end{para}

\begin{para}
    \textsuperscript{13:22}\thinspace{}A good man leaves an in\-her\-it\-ance to his children's chil\-dren, but the sinner's wealth is stored up for the right\-eous.
\end{para}

\begin{para}
    \textsuperscript{14:4}\thinspace{}Where there are no ox\-en, the man\-ger is clean, but a\-bun\-dant har\-vest co\-mes through an ox's strength.
\end{para}

\begin{para}
    \textsuperscript{14:20}\thinspace{}The poor is hat\-ed e\-ven by his neigh\-bor, but there are man\-y who love the rich.
\end{para}

\begin{para}
    \textsuperscript{15:6}\thinspace{}There is great wealth in the house of the right\-eous, but trou\-ble accompanies the in\-come of the wick\-ed.
\end{para}

\begin{para}
    \textsuperscript{15:27}\thinspace{}He who profits il\-lic\-it\-ly troubles his own house, but he who hates bribes will live.
\end{para}

\begin{para}
    \textsuperscript{16:8}\thinspace{}Bet\-ter a lit\-tle with right\-eous\-ness than great in\-come with in\-jus\-tice.
\end{para}

\begin{para}
    \textsuperscript{19:4}\thinspace{}Wealth attracts man\-y friends, but a poor man is sep\-a\-rat\-ed from his friend.
    \textsuperscript{6}\thinspace{}Man\-y seek a gen\-er\-ous man's fa\-vor, and eve\-ry\-one is a friend to him who gives gifts.
    \textsuperscript{7}\thinspace{}All a poor man's brothers hate him. How much more do his friends a\-void him! He pursues them with words, but they are gone.
\end{para}

\begin{para}
    \textsuperscript{21:5}\thinspace{}The plans of the dil\-i\-gent lead sure\-ly to a\-bun\-dance, but eve\-ry\-one who is hast\-y co\-mes sure\-ly to pov\-er\-ty.
\end{para}

\begin{para}
    \textsuperscript{21:6}\thinspace{}Ac\-quir\-ing treasures by a ly\-ing tongue is a fleet\-ing va\-por and the pur\-suit of death.
\end{para}

\begin{para}
    \textsuperscript{21:17}\thinspace{}He who loves pleas\-ure will be\-come poor. He who loves wine and oil will not get rich.
    \textsuperscript{20}\thinspace{}There is pre\-cious treas\-ure and oil in a wise man's dwell\-ing, but a fool consumes them.
\end{para}

\begin{para}
    \textsuperscript{22:1}\thinspace{}A good name is to be cho\-sen o\-ver great rich\-es, and fa\-vor o\-ver sil\-ver and gold.
\end{para}

\begin{para}
    \textsuperscript{22:2}\thinspace{}The rich and the poor have this in com\-mon: \textsc{God} is the mak\-er of them all.
\end{para}

\begin{para}
    \textsuperscript{22:7}\thinspace{}The rich rules o\-ver the poor, and the bor\-row\-er is a slave to the lend\-er.
\end{para}

\begin{para}
    \textsuperscript{26:10}\thinspace{}Like an arch\-er who wounds eve\-ry\-one, so is he who hires a fool or who hires those pass\-ing by.
\end{para}

\begin{para}
    \textsuperscript{28:22}\thinspace{}A sting\-y man hastens af\-ter wealth, and does\-n't know that pov\-er\-ty will come up\-on him.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Kindness}
\markright{Kindness}
\begin{para}
    \textsuperscript{11:17}\thinspace{}A kind man benefits him\-self, but a cru\-el man harms him\-self.
\end{para}

\begin{para}
    \textsuperscript{12:10}\thinspace{}A right\-eous man has re\-gard for the life of his an\-i\-mal, but e\-ven the mer\-cy of the wick\-ed is cru\-el.
\end{para}

\begin{para}
    \textsuperscript{14:21}\thinspace{}He who despises his neigh\-bor sins, but he who shows kind\-ness to the poor is hap\-py.
    \textsuperscript{31}\thinspace{}He who oppresses the poor insults his mak\-er, but he who is kind to the needy hon\-ors him.
\end{para}

\begin{para}
    \textsuperscript{17:13}\thinspace{}If any\-one returns evil for good, evil will nev\-er leave his house.
\end{para}

\begin{para}
    \textsuperscript{21:21}\thinspace{}He who pursues right\-eous\-ness and kind\-ness will find life, right\-eous\-ness, and hon\-or.
\end{para}

\begin{para}
    \textsuperscript{26:27}\thinspace{}He who digs a pit will fall in\-to it, and he who rolls a stone, it will come back on him.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Generosity}
\markright{Generosity}
\begin{para}
    \textsuperscript{11:24}\thinspace{}One gives free\-ly, yet increases all the more, and an\-oth\-er withholds what is due, but co\-mes on\-ly to pov\-er\-ty.
    \textsuperscript{25}\thinspace{}The gen\-er\-ous man will pros\-per, and he who waters will al\-so be watered him\-self.
    \textsuperscript{26}\thinspace{}The peo\-ple will curse him who withholds grain, but a bless\-ing is on the head of him who sells it.
\end{para}

\begin{para}
    \textsuperscript{19:17}\thinspace{}Kind\-ness to the poor is a loan to \textsc{God}, and he will re\-pay his good deed.
\end{para}

\begin{para}
    \textsuperscript{20:25}\thinspace{}It is a snare to say rash\-ly, “It is ho\-ly,” and on\-ly make in\-quir\-y af\-ter the vows.
\end{para}

\begin{para}
    \textsuperscript{21:13}\thinspace{}He who shuts his ear to the cry of the poor will al\-so call out and not be answered.
\end{para}

\begin{para}
    \textsuperscript{22:9}\thinspace{}He who is gen\-er\-ous will be bless\-ed, for he shares his bread with the poor.
    \textsuperscript{16}\thinspace{}He who oppresses the poor to in\-crease his wealth, or who gives to the rich, will on\-ly come to pov\-er\-ty.
\end{para}

\begin{para}
    \textsuperscript{25:21}\thinspace{}If your en\-e\-my is hun\-gry, give him bread to eat, and if he is thirst\-y, give him wa\-ter to drink.
\end{para}

\begin{para}
    \textsuperscript{27:7}\thinspace{}He who is full loathes hon\-ey, but to the hun\-gry any bit\-ter thing is sweet.
\end{para}

\begin{para}
    \textsuperscript{28:27}\thinspace{}He who gives to the poor will nev\-er want, but he who hides his eyes will have man\-y curs\-es.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Emotion}
\markright{Emotion}
\begin{para}
    \textsuperscript{12:25}\thinspace{}Anx\-i\-e\-ty in a man's heart weighs it down, but a good word makes it glad.
\end{para}

\begin{para}
    \textsuperscript{13:12}\thinspace{}Hope de\-ferred makes the heart sick, but de\-sire ful\-filled is a tree of life.
\end{para}

\begin{para}
    \textsuperscript{14:13}\thinspace{}E\-ven in laugh\-ter the heart may ache, and joy may end in grief.
\end{para}

\begin{para}
    \textsuperscript{14:17}\thinspace{}A man of quick tem\-per acts fool\-ish\-ly, and a man of evil schemes is hat\-ed.
    \textsuperscript{29}\thinspace{}He who is slow to an\-ger has great un\-der\-stand\-ing, but he who has a quick tem\-per promotes fol\-ly.
\end{para}

\begin{para}
    \textsuperscript{14:30}\thinspace{}A tran\-quil heart is life to the bod\-y, but en\-vy makes the bones rot.
\end{para}

\begin{para}
    \textsuperscript{15:13}\thinspace{}A glad heart makes a cheer\-ful coun\-te\-nance, but a sorrowful heart breaks the spir\-it.
\end{para}

\begin{para}
    \textsuperscript{15:15}\thinspace{}All the days of the afflicted are bad, but a cheer\-ful heart has a con\-tin\-u\-al feast.
\end{para}

\begin{para}
    \textsuperscript{16:14}\thinspace{}A king's wrath is a mes\-sen\-ger of death, but a wise man appeases it.
\end{para}

\begin{para}
    \textsuperscript{16:32}\thinspace{}He who is slow to an\-ger is bet\-ter than the might\-y, and he who rules his spir\-it, than he who captures a cit\-y.
\end{para}

\begin{para}
    \textsuperscript{17:5}\thinspace{}Who\-ev\-er mocks the poor insults his mak\-er, and he who rejoices at ca\-lam\-i\-ty will not go un\-pun\-ished.
\end{para}

\begin{para}
    \textsuperscript{17:22}\thinspace{}A joy\-ful heart is good med\-i\-cine, but a bro\-ken spir\-it dries up the bones.
\end{para}

\begin{para}
    \textsuperscript{18:14}\thinspace{}A man's spir\-it can en\-dure sick\-ness, but who can bear a bro\-ken spir\-it?
\end{para}

\begin{para}
    \textsuperscript{19:3}\thinspace{}A man's fool\-ish\-ness ruins his way, yet his heart rages a\-gainst \textsc{God}.
\end{para}

\begin{para}
    \textsuperscript{19:11}\thinspace{}Good sense makes a man slow to an\-ger, and it is his glo\-ry to o\-ver\-look an of\-fense.
\end{para}

\begin{para}
    \textsuperscript{19:19}\thinspace{}A man of great an\-ger will pay the pen\-al\-ty, for if you res\-cue him, you will have to do it a\-gain.
\end{para}

\begin{para}
    \textsuperscript{20:1}\thinspace{}Wine is a mocker, strong drink a brawl\-er, and who\-ev\-er is led a\-stray by them is not wise.
\end{para}

\begin{para}
    \textsuperscript{20:22}\thinspace{}Don't say, “I will re\-pay evil”. Wait for \textsc{God}, and he will save you.
\end{para}

\begin{para}
    \textsuperscript{29:11}\thinspace{}A fool gives full vent to his an\-ger, but a wise man holds it back.
\end{para}

\chapter*{Confucius}
\markboth{Confucius}{Confucius}
\addcontentsline{toc}{section}{Confucius}\begin{changemargin}{.6in}{.6in}
  \begin{center}
    \itshape
    An a\-bridg\-ment of the An\-a\-lects.
  \end{center}
\end{changemargin}
\begin{para}
    \textsuperscript{1:1}\thinspace{}Is\-n't it a pleas\-ure to learn some\-thing, then to ap\-ply it at the right time? Is\-n't it a joy to have friends come from a\-far? Is\-n't he a su\-pe\-ri\-or man who takes no of\-fense when his mer\-its are ig\-nored?
\end{para}

\begin{para}
    \textsuperscript{1:3}\thinspace{}Cun\-ning words and an in\-gra\-ti\-at\-ing ap\-pear\-ance rare\-ly ac\-com\-pa\-ny be\-nev\-o\-lence.
\end{para}

\begin{para}
    \textsuperscript{1:5}\thinspace{}The Way to guide a state of a thou\-sand chariots: be at\-ten\-tive to busi\-ness and trust\-wor\-thy; be fru\-gal in expenditures and love your fel\-low men; on\-ly mo\-bi\-lize the com\-mon peo\-ple at the prop\-er time of year.
\end{para}

\begin{para}
    \textsuperscript{1:6}\thinspace{}A young man should re\-spect his parents at home, re\-spect his elders a\-broad, say lit\-tle but be trust\-wor\-thy, love the peo\-ple at large but be in\-ti\-mate with the be\-nev\-o\-lent. If he still has en\-er\-gy to spare, let him stud\-y lit\-er\-a\-ture.
\end{para}

\begin{para}
    \textsuperscript{1:10}\thinspace{}Ziqin asked Zigong, “When Con\-fu\-cius arrives in a coun\-try, he al\-ways learns a\-bout its gov\-ern\-ment. Does he seek out this in\-for\-ma\-tion or is it giv\-en to him?” Zigong replied, “Con\-fu\-cius obtains it by be\-ing cor\-dial, up\-right, re\-spect\-ful, fru\-gal, and def\-er\-en\-tial. He seeks it dif\-fer\-ent\-ly than oth\-er men do.”
\end{para}

\begin{para}
    \textsuperscript{1:14}\thinspace{}A su\-pe\-ri\-or man seeks nei\-ther a full stom\-ach nor com\-fort\-a\-ble lodg\-ings. He is dil\-i\-gent in his actions but cau\-tious in his speech. He associates with those who know the Way in or\-der to be put right. Such a man may be said to love learn\-ing.
\end{para}

\begin{para}
    \textsuperscript{1:16}\thinspace{}Don't wor\-ry that men don't rec\-og\-nize your mer\-its; wor\-ry that you don't rec\-og\-nize theirs.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{2:1}\thinspace{}He who governs by vir\-tue is like the North Star: it stays in its place while all the oth\-er stars pay hom\-age to it.
\end{para}

\begin{para}
    \textsuperscript{2:3}\thinspace{}If you lead the peo\-ple with edicts and keep them or\-der\-ly with punishments, they will e\-vade the pun\-ish\-ment and have no sense of shame. If you lead them with vir\-tue and keep them or\-der\-ly with pro\-pri\-e\-ty, they will have a sense of shame and or\-der them\-selves.
\end{para}

\begin{para}
    \textsuperscript{2:6}\thinspace{}Meng Wubo asked a\-bout fil\-ial pi\-e\-ty. Con\-fu\-cius said, “Give your parents noth\-ing to wor\-ry a\-bout oth\-er than ill\-ness.”
\end{para}

\begin{para}
    \textsuperscript{2:7}\thinspace{}Ziyou asked a\-bout fil\-ial pi\-e\-ty. Con\-fu\-cius said, “Now\-a\-days fil\-ial pi\-e\-ty on\-ly means be\-ing a\-ble to feed your parents. But e\-ven dogs and hors\-es re\-ceive this much care. With\-out re\-spect, what's the dif\-fer\-ence?”
\end{para}

\begin{para}
    \textsuperscript{2:8}\thinspace{}Zixia asked a\-bout fil\-ial pi\-e\-ty. Con\-fu\-cius said, “It's the ex\-pres\-sion on your face that's dif\-fi\-cult to man\-age. It's not fil\-ial pi\-e\-ty mere\-ly for the young to shoul\-der the bur\-den when there is work to do, and to de\-fer to their elders when there is wine and food a\-vail\-a\-ble.”
\end{para}

\begin{para}
    \textsuperscript{2:10}\thinspace{}Look at the means a man employs. Ob\-serve his motives. Ex\-am\-ine what gives him com\-fort. How can a man con\-ceal his char\-ac\-ter?
\end{para}

\begin{para}
    \textsuperscript{2:11}\thinspace{}If a man learns the new by reviewing the old, he is wor\-thy to be a teach\-er.
\end{para}

\begin{para}
    \textsuperscript{2:13}\thinspace{}A su\-pe\-ri\-or man acts be\-fore he speaks, and then speaks ac\-cord\-ing to his actions.
    \textsuperscript{14}\thinspace{}A su\-pe\-ri\-or man is in\-clu\-sive and not par\-ti\-san, while an in\-fe\-ri\-or man is the op\-po\-site.
\end{para}

\begin{para}
    \textsuperscript{2:15}\thinspace{}Learn\-ing with\-out think\-ing is a waste. Think\-ing with\-out learn\-ing is dan\-ger\-ous.
\end{para}

\begin{para}
    \textsuperscript{2:17}\thinspace{}Knowl\-edge is to know when you know some\-thing, and to know when you don't.
\end{para}

\begin{para}
    \textsuperscript{2:18}\thinspace{}Zizhang was stud\-y\-ing in or\-der to get an of\-fi\-cial po\-si\-tion. Con\-fu\-cius said, “If you lis\-ten much, put a\-side what is doubt\-ful, and speak cau\-tious\-ly a\-bout the rest, you will make few mistakes. If you ob\-serve much, put a\-side what is per\-i\-lous, and act cau\-tious\-ly on the rest, you will have few regrets. Few mistakes in speech, few regrets in ac\-tion: an of\-fi\-cial po\-si\-tion will fol\-low.”
\end{para}

\begin{para}
    \textsuperscript{2:19}\thinspace{}If you raise up the straight and place them o\-ver the crook\-ed, the peo\-ple will fol\-low you. If you raise up the crook\-ed and place them o\-ver the straight, the peo\-ple will not fol\-low you.
\end{para}

\begin{para}
    \textsuperscript{2:20}\thinspace{}Ji Kangzi asked, “How can I en\-cour\-age the peo\-ple to be re\-spect\-ful, loy\-al, and dil\-i\-gent?” Con\-fu\-cius said, “Ap\-proach them with dig\-ni\-ty then they will be re\-spect\-ful. Be fil\-ial and kind then they will be loy\-al. Raise up the good and in\-struct the in\-com\-pe\-tent then they will be dil\-i\-gent.”
\end{para}

\begin{para}
    \textsuperscript{2:21}\thinspace{}Some\-one asked Con\-fu\-cius, “Why don't you en\-gage in gov\-ern\-ment?” Con\-fu\-cius said, “The Book of Documents says, ‘Just by be\-ing fil\-ial and be\-ing friend\-ly to your brothers you will have an ef\-fect on gov\-ern\-ment.’ Since this is al\-so en\-gag\-ing in gov\-ern\-ment, why do I need to ‘en\-gage in gov\-ern\-ment’?”
\end{para}

\begin{para}
    \textsuperscript{2:22}\thinspace{}I don't know what to do with an un\-trust\-wor\-thy man. How can you pull a wag\-on that's mis\-sing its yoke pin?
\end{para}

\begin{para}
    \textsuperscript{2:24}\thinspace{}To see what is right and not to do it is a lack of cour\-age.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{3:3}\thinspace{}If a man is not be\-nev\-o\-lent, what can he do with pro\-pri\-e\-ty?
\end{para}

\begin{para}
    \textsuperscript{3:4}\thinspace{}In rit\-u\-al, pre\-fer fru\-gal\-i\-ty to ex\-trav\-a\-gance. In funerals, pre\-fer grief to for\-mal\-i\-ty.
\end{para}

\begin{para}
    \textsuperscript{3:12}\thinspace{}If I don't par\-tic\-i\-pate in a sac\-ri\-fice, it's as if I have not sacrificed at all.
\end{para}

\begin{para}
    \textsuperscript{3:15}\thinspace{}When Con\-fu\-cius entered the Grand Tem\-ple, he asked questions a\-bout eve\-ry\-thing. Some\-one remarked, “Who says that Con\-fu\-cius knows a\-bout rit\-u\-al? When visiting the Grand Tem\-ple, he asked questions a\-bout eve\-ry\-thing.” Con\-fu\-cius, hear\-ing this, said, “This is the rit\-u\-al.”
\end{para}

\begin{para}
    \textsuperscript{3:21}\thinspace{}Don't ex\-plain a\-way what is al\-read\-y done. Don't ar\-gue a\-gainst what is al\-read\-y ac\-com\-plished. Don't as\-sign blame for what is al\-read\-y past.
\end{para}

\begin{para}
    \textsuperscript{3:26}\thinspace{}Hold\-ing a high po\-si\-tion with\-out tol\-er\-ance, per\-form\-ing rit\-u\-al with\-out rev\-er\-ence, joining a fu\-ner\-al with\-out grief: how can I look at such things?
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{4:1}\thinspace{}It is be\-nev\-o\-lence that makes a neigh\-bor\-hood beau\-ti\-ful. How can a man be con\-sid\-ered wise if he chooses not to re\-side a\-mong be\-nev\-o\-lence?
\end{para}

\begin{para}
    \textsuperscript{4:2}\thinspace{}Those who lack be\-nev\-o\-lence can en\-dure nei\-ther ad\-ver\-si\-ty nor en\-joy\-ment for long. The be\-nev\-o\-lent are con\-tent with be\-nev\-o\-lence. The wise cul\-ti\-vate be\-nev\-o\-lence for its ad\-van\-tage.
\end{para}

\begin{para}
    \textsuperscript{4:4}\thinspace{}If you set your heart on be\-nev\-o\-lence, you will be free from evil.
\end{para}

\begin{para}
    \textsuperscript{4:5}\thinspace{}All men de\-sire wealth and hon\-or, but if they are obtained by not fol\-low\-ing the Way, they should not be kept. All men hate pov\-er\-ty and dis\-grace, but if they can't be avoided by fol\-low\-ing the Way, they should not be avoided. If a su\-pe\-ri\-or man abandons be\-nev\-o\-lence, how is he wor\-thy of that name? A su\-pe\-ri\-or man nev\-er leaves be\-nev\-o\-lence e\-ven for the space of a sin\-gle meal. In moments of haste, he acts ac\-cord\-ing to it. In times of dif\-fi\-cul\-ty, he acts ac\-cord\-ing to it.
\end{para}

\begin{para}
    \textsuperscript{4:6}\thinspace{}One who re\-al\-ly loves be\-nev\-o\-lence puts noth\-ing a\-bove it. One who re\-al\-ly hates what is not be\-nev\-o\-lent would not al\-low it to ap\-proach him. Has any\-one de\-vot\-ed all his strength to be\-nev\-o\-lence for a sin\-gle day? I have\-n't seen any\-one who lacked the strength to do so.
\end{para}

\begin{para}
    \textsuperscript{4:7}\thinspace{}Peo\-ple make errors ac\-cord\-ing to the type of per\-son they are. Ob\-serve their errors and you can know their de\-gree of be\-nev\-o\-lence.
\end{para}

\begin{para}
    \textsuperscript{4:9}\thinspace{}A gen\-tle\-man who sets his heart on the Way, but is a\-shamed of poor clothes and poor food, is not worth en\-gag\-ing in dis\-cus\-sion.
\end{para}

\begin{para}
    \textsuperscript{4:10}\thinspace{}When a su\-pe\-ri\-or man deals with the world he is not in\-var\-i\-a\-bly for or a\-gainst any\-thing. He sides with what is right.
\end{para}

\begin{para}
    \textsuperscript{4:11}\thinspace{}A su\-pe\-ri\-or man cherishes vir\-tue; an in\-fe\-ri\-or man cherishes land. A su\-pe\-ri\-or man cherishes jus\-tice; an in\-fe\-ri\-or man cherishes fa\-vors.
\end{para}

\begin{para}
    \textsuperscript{4:12}\thinspace{}One who acts out of per\-son\-al prof\-it incurs much re\-sent\-ment.
\end{para}

\begin{para}
    \textsuperscript{4:14}\thinspace{}Don't wor\-ry that you have no po\-si\-tion; wor\-ry a\-bout what it takes to mer\-it one. Don't wor\-ry that no one knows you; seek to be worth know\-ing.
\end{para}

\begin{para}
    \textsuperscript{4:16}\thinspace{}A su\-pe\-ri\-or man understands what is right. An in\-fe\-ri\-or man understands per\-son\-al prof\-it.
\end{para}

\begin{para}
    \textsuperscript{4:17}\thinspace{}When you see a wor\-thy man, think of be\-com\-ing his e\-qual; when you see an un\-wor\-thy man, look with\-in and ex\-am\-ine your\-self.
\end{para}

\begin{para}
    \textsuperscript{4:18}\thinspace{}In serv\-ing your parents, you may ad\-mon\-ish them gen\-tly; if you see that they don't fol\-low your ad\-vice, re\-main re\-spect\-ful, don't be\-come dis\-o\-be\-di\-ent, and work with\-out complaining.
\end{para}

\begin{para}
    \textsuperscript{4:22}\thinspace{}The ancients were re\-luc\-tant to speak be\-cause they would have been a\-shamed if their actions did\-n't live up to their words.
\end{para}

\begin{para}
    \textsuperscript{4:23}\thinspace{}One rare\-ly miss\-es the mark be\-cause of self-control.
\end{para}

\begin{para}
    \textsuperscript{4:24}\thinspace{}A su\-pe\-ri\-or man wants to be slow in speech but quick in ac\-tion.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{5:5}\thinspace{}What is the use of el\-o\-quence? Those who give a quick re\-tort of\-ten in\-cur the ha\-tred of others.
\end{para}

\begin{para}
    \textsuperscript{5:10}\thinspace{}Zai Yu was sleep\-ing dur\-ing the day. Con\-fu\-cius said, “Rot\-ten wood can\-not be carved, and a wall of ma\-nure can\-not be trow\-eled. What is the use of scolding Zai Yu? I used to lis\-ten to what peo\-ple said and trust them to act ac\-cord\-ing\-ly. Now I lis\-ten to what peo\-ple say and watch what they do. It is be\-cause of Zai Yu that I have changed.”
\end{para}

\begin{para}
    \textsuperscript{5:11}\thinspace{}I have\-n't seen a tru\-ly stead\-fast man. How can a man who is full of desires be stead\-fast?
\end{para}

\begin{para}
    \textsuperscript{5:15}\thinspace{}Kong the Re\-fined was men\-tal\-ly quick, fond of learn\-ing, and not a\-shamed to ask questions of his inferiors. That's why he was called “Re\-fined”.
\end{para}

\begin{para}
    \textsuperscript{5:16}\thinspace{}Zi Chan had four characteristics of the Way of a su\-pe\-ri\-or man: he was re\-spect\-ful in his pri\-vate con\-duct, re\-spect\-ful in serv\-ing his superiors, gen\-er\-ous in car\-ing for the peo\-ple, and just in employing the peo\-ple.
\end{para}

\begin{para}
    \textsuperscript{5:17}\thinspace{}Yang Ping Zhong ex\-celled in his re\-la\-tions with others. E\-ven af\-ter long ac\-quaint\-ance he con\-tin\-ued to treat them with re\-spect.
\end{para}

\begin{para}
    \textsuperscript{5:20}\thinspace{}Ji Wen Zi al\-ways thought three times be\-fore act\-ing. Hear\-ing this, Con\-fu\-cius said, “Twice is e\-nough.”
\end{para}

\begin{para}
    \textsuperscript{5:23}\thinspace{}Bo Yi and Shu Qi did not re\-mem\-ber old grievances, so there was lit\-tle re\-sent\-ment a\-gainst them.
\end{para}

\begin{para}
    \textsuperscript{5:25}\thinspace{}Glib talk, an in\-gra\-ti\-at\-ing ex\-pres\-sion, and ob\-se\-qui\-ous con\-duct: I am a\-shamed of such be\-hav\-ior. Act\-ing friend\-ly to\-wards some\-one while concealing your re\-sent\-ment for them: I am a\-shamed of such be\-hav\-ior.
\end{para}

\begin{para}
    \textsuperscript{5:26}\thinspace{}My heart's de\-sire is to bring peace to the old, to share trust with my friends, and to cher\-ish the young.
\end{para}

\begin{para}
    \textsuperscript{5:27}\thinspace{}I have not yet met any\-one who can see his own faults and in\-ward\-ly ac\-cuse him\-self.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{6:3}\thinspace{}Yanhui loved learn\-ing. He nev\-er took his an\-ger out on others, and nev\-er made the same mis\-take twice.
    \textsuperscript{11}\thinspace{}How ad\-mi\-ra\-ble he was! He lived on a bowlful of rice to eat, a la\-dle\-ful of wa\-ter to drink, and a hov\-el for shel\-ter. Others could not have en\-dured such mis\-er\-y, yet Yanhui's joy was un\-changed.
\end{para}

\begin{para}
    \textsuperscript{6:12}\thinspace{}Ran Qiu said, “It's not that I don't en\-joy your Way, but I don't have the strength to fol\-low it.” Con\-fu\-cius said, “Those who don't have the strength col\-lapse a\-long the way, but you have drawn a line be\-fore starting.”
\end{para}

\begin{para}
    \textsuperscript{6:15}\thinspace{}Meng Zhifan was not boast\-ful. He once held the rear dur\-ing a re\-treat. As they reached the cit\-y gate, he whipped his horse, say\-ing, “It's not that I dared to be last, on\-ly that my horse would\-n't go.”
\end{para}

\begin{para}
    \textsuperscript{6:18}\thinspace{}When nat\-u\-ral dis\-po\-si\-tion prevails o\-ver re\-fine\-ment, you get a sav\-age. When re\-fine\-ment prevails o\-ver nat\-u\-ral dis\-po\-si\-tion, you get a pe\-dan\-tic clerk. When nat\-u\-ral dis\-po\-si\-tion and re\-fine\-ment are prop\-er\-ly blend\-ed, you get a su\-pe\-ri\-or man.
\end{para}

\begin{para}
    \textsuperscript{6:28}\thinspace{}If I have done wrong, may Heav\-en de\-test and a\-ban\-don me!
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{7:2}\thinspace{}It's nat\-u\-ral for me to store up knowl\-edge si\-lent\-ly, to learn with\-out sa\-ti\-e\-ty, and to teach others with\-out wea\-ri\-ness.
\end{para}

\begin{para}
    \textsuperscript{7:3}\thinspace{}My worries are fail\-ure to cul\-ti\-vate vir\-tue, fail\-ure to ex\-plore what I learn, in\-a\-bil\-i\-ty to fol\-low what is right when I hear it, and in\-a\-bil\-i\-ty to change what is not good.
\end{para}

\begin{para}
    \textsuperscript{7:6}\thinspace{}Set your heart on the Way. Base your\-self in vir\-tue. Re\-ly on be\-nev\-o\-lence. En\-joy the arts.
\end{para}

\begin{para}
    \textsuperscript{7:7}\thinspace{}I have nev\-er de\-nied in\-struc\-tion to any\-one, e\-ven if his own resources could on\-ly af\-ford a bun\-dle of dried meat for tu\-i\-tion.
\end{para}

\begin{para}
    \textsuperscript{7:8}\thinspace{}I on\-ly en\-light\-en those who are driv\-en by ea\-ger\-ness to learn and who are anx\-ious\-ly try\-ing to ex\-plain their ideas. If I lift up one cor\-ner and they can\-not come back with the oth\-er three, I don't re\-peat my\-self.
\end{para}

\begin{para}
    \textsuperscript{7:9}\thinspace{}When Con\-fu\-cius was eat\-ing in the pres\-ence of a mourn\-er, he nev\-er ate his fill.
\end{para}

\begin{para}
    \textsuperscript{7:11}\thinspace{}Few have the a\-bil\-i\-ty to act when employed and to hide when dismissed. If I were com\-mand\-ing the Three Armies, who would I take with me? A man who would wres\-tle a ti\-ger with bare hands or cross a riv\-er on foot, dy\-ing with\-out re\-gret---I would not take that man. It must be some\-one who approaches a task with ap\-pre\-hen\-sion, likes to make plans, and can ex\-e\-cute them suc\-cess\-ful\-ly.
\end{para}

\begin{para}
    \textsuperscript{7:12}\thinspace{}If wealth can be sought de\-cent\-ly, I will do so, e\-ven if I have to have a low\-ly job. If it can\-not be sought de\-cent\-ly, I will fol\-low what I love.
\end{para}

\begin{para}
    \textsuperscript{7:15}\thinspace{}Bo Yi and Shu Qi sought be\-nev\-o\-lence and got it. Why should they have complaints?
\end{para}

\begin{para}
    \textsuperscript{7:16}\thinspace{}Coarse rice to eat, wa\-ter to drink, and your bent arm for a pil\-low---there is still joy in the midst of these things. Wealth and hon\-or obtained by un\-right\-eous means are like fleet\-ing clouds to me.
\end{para}

\begin{para}
    \textsuperscript{7:17}\thinspace{}Give me a few more years; if I can stud\-y un\-til fif\-ty, I may be free from great mistakes.
\end{para}

\begin{para}
    \textsuperscript{7:19}\thinspace{}I'm the kind of man who is so ea\-ger for knowl\-edge that he forgets to eat, who is so joy\-ful that he forgets to wor\-ry, and who is not a\-ware that old age is com\-ing.
\end{para}

\begin{para}
    \textsuperscript{7:20}\thinspace{}I was not born with knowl\-edge, but I love an\-tiq\-ui\-ty and ear\-nest\-ly seek it.
\end{para}

\begin{para}
    \textsuperscript{7:22}\thinspace{}When walk\-ing with any two peo\-ple, they must have some\-thing to teach me. I se\-lect their strengths and fol\-low them, their weaknesses and cor\-rect my\-self.
\end{para}

\begin{para}
    \textsuperscript{7:23}\thinspace{}Heav\-en gave birth to the vir\-tue in me. What can an\-gry rulers do to me?
\end{para}

\begin{para}
    \textsuperscript{7:24}\thinspace{}My friends, I hide noth\-ing from you. What\-ev\-er I do, I share with you. That is who I am.
\end{para}

\begin{para}
    \textsuperscript{7:26}\thinspace{}I can't hope to meet a sage, but I would be con\-tent to meet a su\-pe\-ri\-or man. I can't hope to meet a tru\-ly good man, but I would be con\-tent to meet a stead\-fast man. It is hard to be stead\-fast when noth\-ing pretends to be some\-thing, emp\-ti\-ness pretends to be full\-ness, and pov\-er\-ty pretends to be com\-fort.
\end{para}

\begin{para}
    \textsuperscript{7:27}\thinspace{}Con\-fu\-cius fished with a line, but did\-n't use a net. He hunt\-ed, but did\-n't shoot roosting birds.
\end{para}

\begin{para}
    \textsuperscript{7:28}\thinspace{}There may be those who can act with\-out knowl\-edge, but I am not one of them. I hear much, se\-lect what is good, and fol\-low it. I see much, and re\-mem\-ber it.
\end{para}

\begin{para}
    \textsuperscript{7:29}\thinspace{}To ap\-prove of a man's vis\-it does\-n't mean ap\-prov\-ing of what he does af\-ter he leaves. When a man purifies him\-self to ap\-proach me, I ap\-prove of his pu\-ri\-ty with\-out en\-dors\-ing his past or his fu\-ture.
\end{para}

\begin{para}
    \textsuperscript{7:30}\thinspace{}Is be\-nev\-o\-lence far a\-way? As soon as I want be\-nev\-o\-lence, it is at hand.
\end{para}

\begin{para}
    \textsuperscript{7:31}\thinspace{}I'm for\-tu\-nate. When I make a mis\-take, others al\-ways no\-tice it.
\end{para}

\begin{para}
    \textsuperscript{7:34}\thinspace{}How would I dare claim to be be\-nev\-o\-lent or a sage? Rath\-er, it may be said that I strive for them with\-out flag\-ging, and teach others with\-out wea\-ri\-ness.
\end{para}

\begin{para}
    \textsuperscript{7:36}\thinspace{}Ex\-trav\-a\-gance leads to ar\-ro\-gance; fru\-gal\-i\-ty leads to stin\-gi\-ness. It's bet\-ter to be sting\-y than ar\-ro\-gant.
\end{para}

\begin{para}
    \textsuperscript{7:37}\thinspace{}A su\-pe\-ri\-or man is eas\-y\-go\-ing and un\-per\-turbed. An in\-fe\-ri\-or man is al\-ways tense and anx\-ious.
\end{para}

\begin{para}
    \textsuperscript{7:38}\thinspace{}Con\-fu\-cius was warm yet stern; au\-thor\-i\-ta\-tive but not fierce; re\-spect\-ful yet at ease.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{8:1}\thinspace{}Tai Bo can be said to have had the highest vir\-tue. Three times he de\-clined his claim to the king\-dom.
\end{para}

\begin{para}
    \textsuperscript{8:2}\thinspace{}Cour\-te\-sy with\-out pro\-pri\-e\-ty is tire\-some. Cau\-tion with\-out pro\-pri\-e\-ty is ti\-mid\-i\-ty. Brav\-er\-y with\-out pro\-pri\-e\-ty is cha\-os. Can\-dor with\-out pro\-pri\-e\-ty is rude\-ness.
\end{para}

\begin{para}
    \textsuperscript{8:9}\thinspace{}The peo\-ple can be made to fol\-low a path, but they can't be made to un\-der\-stand it.
\end{para}

\begin{para}
    \textsuperscript{8:10}\thinspace{}A man who loves brav\-er\-y and hates pov\-er\-ty will pro\-ceed to un\-ru\-ly be\-hav\-ior. A man with\-out be\-nev\-o\-lence who is de\-spised by others too much will pro\-ceed to un\-ru\-ly be\-hav\-ior.
\end{para}

\begin{para}
    \textsuperscript{8:11}\thinspace{}If a man had talents as ad\-mi\-ra\-ble as the Duke of Zhou, but was ar\-ro\-gant and sting\-y, the rest would not be worth no\-tic\-ing.
\end{para}

\begin{para}
    \textsuperscript{8:13}\thinspace{}Have sin\-cere faith, love learn\-ing, and de\-fend the good Way un\-to death. Don't en\-ter a state that is in per\-il, and don't stay in one that is in cha\-os. Show your\-self when the Way prevails in the world, but hide when it does not. When the Way prevails in the state, it is shame\-ful to be poor and hum\-ble. When the Way does not pre\-vail in the state, it is shame\-ful to be rich and no\-ble.
\end{para}

\begin{para}
    \textsuperscript{8:14}\thinspace{}Don't plan the policies of a po\-si\-tion that you don't hold.
\end{para}

\begin{para}
    \textsuperscript{8:21}\thinspace{}I can find no fault in Yu. His food and drink were fru\-gal, yet his offerings to the spirits and gods showed ut\-ter de\-vo\-tion. His clothes were coarse, yet his rit\-u\-al cap and robes were mag\-nif\-i\-cent. His dwell\-ing was hum\-ble, yet he exhausted his strength on ir\-ri\-ga\-tion ditches and canals.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{9:3}\thinspace{}The rules of rit\-u\-al pre\-scribe a hemp cap. Now\-a\-days it is made of silk. That is more fru\-gal. I fol\-low the com\-mon prac\-tice. The rules of rit\-u\-al pre\-scribe bow\-ing be\-low the steps. Now\-a\-days peo\-ple bow af\-ter as\-cend\-ing. That is ar\-ro\-gant. Al\-though it goes a\-gainst the com\-mon prac\-tice, I bow be\-low the steps.
\end{para}

\begin{para}
    \textsuperscript{9:4}\thinspace{}Con\-fu\-cius for\-bade four things: don't act on spec\-u\-la\-tion, don't de\-mand cer\-tain\-ty, don't be stub\-born, and don't be e\-go\-tis\-ti\-cal.
\end{para}

\begin{para}
    \textsuperscript{9:5}\thinspace{}When Con\-fu\-cius was in dan\-ger in Kuang, he said, “Since King Wen died, does\-n't his cul\-ture re\-side in me? If Heav\-en does\-n't want this cul\-ture destroyed, what can the peo\-ple of Kuang do to me?”
\end{para}

\begin{para}
    \textsuperscript{9:6}\thinspace{}I was poor when I was young, so I learn\-ed man\-y me\-ni\-al skills. Must a su\-pe\-ri\-or man be skilled in man\-y things? No, not man\-y.
\end{para}

\begin{para}
    \textsuperscript{9:8}\thinspace{}Do I have knowl\-edge? No, I do not. A rus\-tic asked me a ques\-tion and my mind went blank. I ham\-mered at both ends of the ques\-tion un\-til I got eve\-ry\-thing out of it.
\end{para}

\begin{para}
    \textsuperscript{9:13}\thinspace{}Zigong asked, “If you had a beau\-ti\-ful piece of jade here, would you hide it safe\-ly in a box or try to get a good price and sell it?” Con\-fu\-cius said, “Sell it! All I'm waiting for is the right of\-fer.”
\end{para}

\begin{para}
    \textsuperscript{9:16}\thinspace{}I have no dif\-fi\-cul\-ty serv\-ing high officials when a\-broad, serv\-ing my elders at home, giv\-ing my best efforts in fu\-ner\-al arrangements, and avoiding trou\-ble from wine.
\end{para}

\begin{para}
    \textsuperscript{9:18}\thinspace{}I have nev\-er seen a man who loves vir\-tue as much as he loves beau\-ty in wom\-en.
\end{para}

\begin{para}
    \textsuperscript{9:19}\thinspace{}It's like build\-ing a mound: if I stop be\-fore the last bas\-ket of earth, it re\-mains un\-fin\-ished. It's like fill\-ing a ditch: al\-though I've dumped in on\-ly a sin\-gle bas\-ket of earth, if I con\-tin\-ue then I'm mak\-ing prog\-ress.
\end{para}

\begin{para}
    \textsuperscript{9:23}\thinspace{}The young\-er gen\-er\-a\-tion should be held in awe. How do we know that it will not e\-qual the pres\-ent one? But if a man reaches for\-ty or fif\-ty years old with\-out dis\-tin\-guish\-ing him\-self, he no long\-er deserves awe.
\end{para}

\begin{para}
    \textsuperscript{9:24}\thinspace{}I can do noth\-ing with the man who enjoys po\-lite words with\-out unfolding their pur\-pose, or who agrees with ex\-em\-pla\-ry words with\-out reforming him\-self.
\end{para}

\begin{para}
    \textsuperscript{9:25}\thinspace{}Put loy\-al\-ty and trust\-wor\-thi\-ness a\-bove eve\-ry\-thing else. On\-ly have friends who are your mor\-al equals. When you make a mis\-take, don't be a\-fraid to cor\-rect it.
\end{para}

\begin{para}
    \textsuperscript{9:30}\thinspace{}You may be a\-ble to stud\-y to\-geth\-er, and yet be un\-a\-ble to pur\-sue the Way to\-geth\-er. You may be a\-ble to pur\-sue the Way to\-geth\-er, and yet be un\-a\-ble to take a stand to\-geth\-er. You may be a\-ble to take a stand to\-geth\-er, and yet be un\-a\-ble to share au\-thor\-i\-ty to\-geth\-er.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{10:1}\thinspace{}In his vil\-lage, Con\-fu\-cius was most def\-er\-en\-tial, as if he were un\-a\-ble to speak. In the an\-ces\-tral tem\-ple or at court, he was el\-o\-quent but cau\-tious.
\end{para}

\begin{para}
    \textsuperscript{10:8}\thinspace{}E\-ven if there was plen\-ty of meat, he would not eat more meat than rice. On\-ly with wine did he have no set lim\-it, but he nev\-er drank to the point of con\-fu\-sion.
\end{para}

\begin{para}
    \textsuperscript{10:10}\thinspace{}He did\-n't con\-verse while eat\-ing, or talk in bed.
    \textsuperscript{11}\thinspace{}E\-ven with a sim\-ple meal of coarse rice and veg\-e\-ta\-ble broth, he made an of\-fer\-ing with so\-lem\-ni\-ty.
    \textsuperscript{12}\thinspace{}He would not sit un\-less his mat was in prop\-er po\-si\-tion.
\end{para}

\begin{para}
    \textsuperscript{10:17}\thinspace{}When the stables burned, Con\-fu\-cius returned from court and asked, “Was any\-one hurt?” He did\-n't ask a\-bout the hors\-es.
\end{para}

\begin{para}
    \textsuperscript{10:22}\thinspace{}When a friend died who had no fam\-i\-ly to make arrangements, he said, “Let the fu\-ner\-al be at my home.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{11:4}\thinspace{}Yan Hui is no help to me. Eve\-ry\-thing I say pleases him.
\end{para}

\begin{para}
    \textsuperscript{11:12}\thinspace{}If you aren't yet a\-ble to serve peo\-ple, how would you be a\-ble to serve the spirits. If you don't yet un\-der\-stand life, how could you un\-der\-stand death?
\end{para}

\begin{para}
    \textsuperscript{11:14}\thinspace{}The peo\-ple of Lu were re\-build\-ing the Long Treas\-ur\-y. Min Ziqian said, “Why not re\-store the old struc\-ture? Why must it be changed com\-plete\-ly?” Con\-fu\-cius said, “This man ei\-ther says noth\-ing, or his words hit the mark.”
\end{para}

\begin{para}
    \textsuperscript{11:16}\thinspace{}Go\-ing too far is as bad as fall\-ing short.
\end{para}

\begin{para}
    \textsuperscript{11:21}\thinspace{}If a man's opinions are sound, is he a su\-pe\-ri\-or man or is it on\-ly a sol\-emn ap\-pear\-ance?
\end{para}

\begin{para}
    \textsuperscript{11:22}\thinspace{}Zilu asked, “Should I im\-me\-di\-ate\-ly put in\-to prac\-tice what I learn?” Con\-fu\-cius said, “You have a fa\-ther and eld\-er brothers to con\-sult first.” Ranyou asked the same ques\-tion. Con\-fu\-cius said, “You should im\-me\-di\-ate\-ly put in\-to prac\-tice what you learn.” Gongxi Hua said, “I'm con\-fused. May I ask for an ex\-pla\-na\-tion?” Con\-fu\-cius said, “Ranyou holds back, so I urged him for\-ward. Zilu has en\-er\-gy for two, so I held him back.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{12:1}\thinspace{}Yan Hui asked a\-bout be\-nev\-o\-lence. Con\-fu\-cius said, “Be\-nev\-o\-lence is o\-ver\-com\-ing one\-self and returning to pro\-pri\-e\-ty. Be\-nev\-o\-lence co\-mes from one\-self. How could it come from others? Don't look at any\-thing im\-prop\-er. Don't lis\-ten to any\-thing im\-prop\-er. Don't say any\-thing im\-prop\-er. Don't do any\-thing im\-prop\-er.”
\end{para}

\begin{para}
    \textsuperscript{12:2}\thinspace{}Zhonggong asked a\-bout be\-nev\-o\-lence. Con\-fu\-cius said, “When a\-broad, be\-have as if you were re\-ceiv\-ing an im\-por\-tant guest. When employing the peo\-ple, be\-have as if you were assisting at an im\-por\-tant sac\-ri\-fice. Don't im\-pose on others what you don't want for your\-self.”
\end{para}

\begin{para}
    \textsuperscript{12:3}\thinspace{}Si\-ma Niu asked a\-bout be\-nev\-o\-lence. Con\-fu\-cius said, “A be\-nev\-o\-lent per\-son is re\-luc\-tant to speak. When do\-ing it is dif\-fi\-cult, how can you not be re\-luc\-tant to speak a\-bout it.”
\end{para}

\begin{para}
    \textsuperscript{12:4}\thinspace{}Si\-ma Niu asked a\-bout the su\-pe\-ri\-or man. Con\-fu\-cius said, “A su\-pe\-ri\-or man is free from wor\-ry and fear. If in\-tro\-spec\-tion finds noth\-ing to be a\-shamed of, why should he have wor\-ry or fear?”
\end{para}

\begin{para}
    \textsuperscript{12:6}\thinspace{}Zizhang asked a\-bout dis\-cern\-ment. Con\-fu\-cius said, “He who is not in\-flu\-enced by slan\-der that saturates and by accusations that spread like a rash, can be called dis\-cern\-ing. He can al\-so be called far\-sight\-ed.”
\end{para}

\begin{para}
    \textsuperscript{12:7}\thinspace{}Zigong asked a\-bout gov\-ern\-ment. Con\-fu\-cius said, “Suf\-fi\-cient food, suf\-fi\-cient arms, and the trust of the peo\-ple. If I had to give up one of these three, I would give up the arms. If I had to give up one of the remaining two, I would give up the food. Death has been with us since an\-cient times, but with\-out trust the peo\-ple can\-not stand.”
\end{para}

\begin{para}
    \textsuperscript{12:10}\thinspace{}Zizhang asked a\-bout ac\-cu\-mu\-lat\-ing vir\-tue. Con\-fu\-cius said, “Make these your guid\-ing principles: loy\-al\-ty, trust\-wor\-thi\-ness, and fol\-low\-ing what is right. This is the way to ac\-cu\-mu\-late vir\-tue.”
\end{para}

\begin{para}
    \textsuperscript{12:13}\thinspace{}In hear\-ing lawsuits, I am the same as any\-one. What we need is to make it so that there are no lawsuits.
\end{para}

\begin{para}
    \textsuperscript{12:14}\thinspace{}Zizhang asked a\-bout gov\-ern\-ment. Con\-fu\-cius said, “Pon\-der it with\-out tir\-ing, and car\-ry it out with loy\-al\-ty.”
\end{para}

\begin{para}
    \textsuperscript{12:15}\thinspace{}Stud\-y cul\-ture broad\-ly, and re\-strain your\-self with pro\-pri\-e\-ty. You will not stray from what is right.
\end{para}

\begin{para}
    \textsuperscript{12:16}\thinspace{}The su\-pe\-ri\-or man perfects the good in peo\-ple, he does not per\-fect the bad. The in\-fe\-ri\-or man does the op\-po\-site.
\end{para}

\begin{para}
    \textsuperscript{12:19}\thinspace{}Ji Kangzi asked Con\-fu\-cius a\-bout gov\-ern\-ment: “What if I were to kill those who don't fol\-low the Way in or\-der to help those who do?” Con\-fu\-cius said, “In administering your gov\-ern\-ment, what use is kill\-ing? If you de\-sire what is good, the peo\-ple will be good. The vir\-tue of the su\-pe\-ri\-or man is like the wind, and the vir\-tue of the in\-fe\-ri\-or man is like the grass. When the wind blows, the grass is sure to bend.”
\end{para}

\begin{para}
    \textsuperscript{12:20}\thinspace{}Zizhang asked, “When can a gen\-tle\-man be called dis\-tin\-guished? By this I mean that he is sure to be known in his state and in his clan.” Con\-fu\-cius said, “That is fame, not dis\-tinc\-tion. A dis\-tin\-guished man has up\-right char\-ac\-ter, loves what is right, examines what peo\-ple say and observes their fa\-cial expressions, and is mind\-ful of def\-er\-ence to others. Such a man will be dis\-tin\-guished in his state and in his clan. As for fame, it is e\-nough to be whol\-ly con\-fi\-dent a\-bout as\-sum\-ing the ap\-pear\-ance of be\-nev\-o\-lence while act\-ing to the con\-tra\-ry. Such a man is sure to be known in his state and in his clan.”
\end{para}

\begin{para}
    \textsuperscript{12:21}\thinspace{}Fan Chi asked, “How do I ac\-cu\-mu\-late vir\-tue, re\-form wick\-ed\-ness, and rec\-og\-nize con\-fu\-sion?” Con\-fu\-cius said, “Ex\-cel\-lent ques\-tion! Put the ef\-fort be\-fore the re\-ward: is that not ac\-cu\-mu\-lat\-ing vir\-tue? At\-tack your own evil rath\-er than the evil of others: is that not reforming wick\-ed\-ness? In a mo\-ment of an\-ger, to for\-get your own safe\-ty and that of your fam\-i\-ly: is that not con\-fu\-sion?”
\end{para}

\begin{para}
    \textsuperscript{12:23}\thinspace{}Zigong asked how to treat friends. Con\-fu\-cius said, “Loy\-al\-ly ad\-vise them, and skill\-ful\-ly show them the Way. If that does\-n't work, then stop. Don't dis\-grace your\-self.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{13:1}\thinspace{}Zilu asked a\-bout gov\-ern\-ment. Con\-fu\-cius said, “Set an ex\-am\-ple for them, and en\-cour\-age them. Be tire\-less.”
\end{para}

\begin{para}
    \textsuperscript{13:2}\thinspace{}Zhonggong asked a\-bout gov\-ern\-ment. Con\-fu\-cius said, “Set an ex\-am\-ple for your officers. Grant par\-don for mi\-nor offenses. Pro\-mote men of vir\-tue and tal\-ent.”
\end{para}

\begin{para}
    \textsuperscript{13:3}\thinspace{}A su\-pe\-ri\-or man re\-mains si\-lent a\-bout things he does\-n't un\-der\-stand. He permits no care\-less\-ness in his speech.
\end{para}

\begin{para}
    \textsuperscript{13:5}\thinspace{}Con\-sid\-er a man who can re\-cite the three hun\-dred Poems. If he fails when you en\-trust him with gov\-ern\-ance, and can\-not act on his own in\-i\-ti\-a\-tive, what use is all his learn\-ing?
\end{para}

\begin{para}
    \textsuperscript{13:6}\thinspace{}If a man is up\-right in his per\-son\-al con\-duct, others will fol\-low with\-out him giv\-ing orders. If a man is not up\-right in his per\-son\-al con\-duct, e\-ven if he gives orders, others won't o\-bey.
\end{para}

\begin{para}
    \textsuperscript{13:12}\thinspace{}E\-ven with a true king, it would still take a gen\-er\-a\-tion for be\-nev\-o\-lence to pre\-vail.
\end{para}

\begin{para}
    \textsuperscript{13:13}\thinspace{}If a man can cor\-rect him\-self, what dif\-fi\-cul\-ty will he have in governing? If he can't cor\-rect him\-self, how can he cor\-rect others?
\end{para}

\begin{para}
    \textsuperscript{13:15}\thinspace{}A sin\-gle say\-ing can\-not ru\-in a coun\-try, but there is a say\-ing that, “The on\-ly pleas\-ure of rul\-ing is that no one contradicts you.” If a rul\-er is good and no one contradicts him, that's fi\-ne. But if a rul\-er is not good and no one contradicts him, then is\-n't this al\-most a say\-ing that could ru\-in a coun\-try?
\end{para}

\begin{para}
    \textsuperscript{13:16}\thinspace{}The Gov\-er\-nor of She asked a\-bout gov\-ern\-ment. Con\-fu\-cius said, “Peo\-ple near\-by are pleased, and peo\-ple far a\-way come.”
\end{para}

\begin{para}
    \textsuperscript{13:17}\thinspace{}Zixia asked a\-bout gov\-ern\-ment. Con\-fu\-cius said, “Don't try to rush things. Don't look for small advantages. If you try to rush things, you won't reach your goal. If you look for small advantages, you won't ac\-com\-plish great things.”
\end{para}

\begin{para}
    \textsuperscript{13:19}\thinspace{}Be cour\-te\-ous at home. Be re\-spect\-ful in han\-dling pub\-lic af\-fairs. Be loy\-al in deal\-ing with others. E\-ven a\-mong barbarians, you can't dis\-card these.
\end{para}

\begin{para}
    \textsuperscript{13:21}\thinspace{}If I can't find mod\-er\-ate men to as\-so\-ci\-ate with, I must find the reck\-less and the cau\-tious. The reck\-less ad\-vance to\-wards the goal. The cau\-tious have lines they won't cross.
\end{para}

\begin{para}
    \textsuperscript{13:23}\thinspace{}A su\-pe\-ri\-or man seeks har\-mo\-ny with\-out con\-form\-i\-ty. An in\-fe\-ri\-or man seeks con\-form\-i\-ty with\-out har\-mo\-ny.
\end{para}

\begin{para}
    \textsuperscript{13:24}\thinspace{}It's not e\-nough for all the peo\-ple to love you. It's not e\-nough for all the peo\-ple to hate you. It's bet\-ter if the good peo\-ple love you and the bad peo\-ple hate you.
\end{para}

\begin{para}
    \textsuperscript{13:25}\thinspace{}If you try to please a su\-pe\-ri\-or man with\-out fol\-low\-ing the Way, he will not be pleased; but when employing others, he on\-ly assigns tasks ac\-cord\-ing to their ca\-pac\-i\-ty. If you try to please an in\-fe\-ri\-or man with\-out fol\-low\-ing the Way, he will be pleased; but when employing others, he expects them to be good at eve\-ry\-thing.
\end{para}

\begin{para}
    \textsuperscript{13:26}\thinspace{}A su\-pe\-ri\-or man has a dig\-ni\-fied ease with\-out be\-ing ar\-ro\-gant. An in\-fe\-ri\-or man is ar\-ro\-gant with\-out a dig\-ni\-fied ease.
\end{para}

\begin{para}
    \textsuperscript{13:27}\thinspace{}Firm, res\-o\-lute, sim\-ple, slow to speak: these are close to be\-nev\-o\-lence.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{14:1}\thinspace{}To serve a state in which the Way prevails, then to con\-tin\-ue serv\-ing it af\-ter it has lost the Way. That is shame\-ful.
\end{para}

\begin{para}
    \textsuperscript{14:2}\thinspace{}A gen\-tle\-man who is at\-tached to ma\-te\-ri\-al comforts is not wor\-thy of be\-ing called a gen\-tle\-man.
\end{para}

\begin{para}
    \textsuperscript{14:4}\thinspace{}A vir\-tu\-ous man will cer\-tain\-ly have some\-thing to say, but a man with some\-thing to say is not nec\-es\-sar\-i\-ly vir\-tu\-ous. A be\-nev\-o\-lent man will cer\-tain\-ly be brave, but a brave man is not nec\-es\-sar\-i\-ly be\-nev\-o\-lent.
\end{para}

\begin{para}
    \textsuperscript{14:7}\thinspace{}Can you love some\-one with\-out mak\-ing them work hard? Can you be loy\-al to some\-one with\-out instructing them?
\end{para}

\begin{para}
    \textsuperscript{14:10}\thinspace{}To be poor with\-out re\-sent\-ment is dif\-fi\-cult; to be rich with\-out ar\-ro\-gance is eas\-y.
\end{para}

\begin{para}
    \textsuperscript{14:12}\thinspace{}A man who is wise, free from desires, brave, ac\-com\-plished, and al\-so re\-fined through rit\-u\-al and mu\-sic, may be called a com\-plete man. But, does a com\-plete man need all this now\-a\-days? If he sees prof\-it and thinks of what is right; if he sees dan\-ger and is read\-y to give his life; if he endures long hard\-ship and does\-n't for\-get the words he lives by: he may be called a com\-plete man.
\end{para}

\begin{para}
    \textsuperscript{14:20}\thinspace{}When your words are not hum\-ble, it's dif\-fi\-cult to live up to them.
\end{para}

\begin{para}
    \textsuperscript{14:22}\thinspace{}Zilu asked how to serve a rul\-er. Con\-fu\-cius said, “Don't use de\-ceit when you op\-pose him.”
\end{para}

\begin{para}
    \textsuperscript{14:27}\thinspace{}A su\-pe\-ri\-or man is a\-shamed if his words sur\-pass his deeds.
\end{para}

\begin{para}
    \textsuperscript{14:28}\thinspace{}The Way of a su\-pe\-ri\-or man has three things that I can't find in my\-self: the be\-nev\-o\-lent are not anx\-ious; the wise are not con\-fused; the brave are not a\-fraid.
\end{para}

\begin{para}
    \textsuperscript{14:29}\thinspace{}Zigong was judg\-ing oth\-er peo\-ple. Con\-fu\-cius said, “Zigong must be of such su\-pe\-ri\-or char\-ac\-ter him\-self that he has lei\-sure for this. I don't.”
\end{para}

\begin{para}
    \textsuperscript{14:30}\thinspace{}Don't wor\-ry that men don't rec\-og\-nize you; wor\-ry a\-bout your own lack of a\-bil\-i\-ty.
\end{para}

\begin{para}
    \textsuperscript{14:31}\thinspace{}To be the first to be a\-ware of de\-cep\-tion or bad faith, with\-out an\-tic\-i\-pat\-ing or suspecting it: is that not be\-ing wor\-thy?
\end{para}

\begin{para}
    \textsuperscript{14:34}\thinspace{}Some\-one asked, “What do you think of the say\-ing, ‘Re\-pay re\-sent\-ment with kind\-ness’?” Con\-fu\-cius said, “Then how will you re\-pay kind\-ness? Re\-pay re\-sent\-ment with uprightness. Re\-pay kind\-ness with kind\-ness.”
\end{para}

\begin{para}
    \textsuperscript{14:35}\thinspace{}No one understands me! I don't com\-plain a\-gainst Heav\-en, nor do I blame men. I stud\-y here be\-low, and I pen\-e\-trate up a\-bove. Who understands me? Heav\-en does!
\end{para}

\begin{para}
    \textsuperscript{14:36}\thinspace{}Gongbo Liao had slandered Zilu. Zifu Jingbo said, “I have e\-nough pow\-er to have Gongbo Liao's corpse dis\-played in the mar\-ket\-place.” Con\-fu\-cius said, “If circumstances al\-low, the Way will pre\-vail; if not, the Way will be a\-ban\-doned. What ef\-fect does Gongbo Liao have on these circumstances?”
\end{para}

\begin{para}
    \textsuperscript{14:37}\thinspace{}The highest vir\-tue is to a\-void the world; next, to a\-void cer\-tain places; next, to a\-void cer\-tain attitudes; next, to a\-void cer\-tain words.
\end{para}

\begin{para}
    \textsuperscript{14:41}\thinspace{}When those a\-bove love pro\-pri\-e\-ty, the com\-mon peo\-ple are eas\-y to lead.
\end{para}

\begin{para}
    \textsuperscript{14:44}\thinspace{}Con\-fu\-cius employed a boy to car\-ry messages. Some\-one asked, “Is he mak\-ing any prog\-ress?” Con\-fu\-cius said, “He's not seek\-ing prog\-ress. He wants quick results.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{15:2}\thinspace{}A su\-pe\-ri\-or man is stead\-fast in the face of ad\-ver\-si\-ty, while an in\-fe\-ri\-or man is engulfed by it.
\end{para}

\begin{para}
    \textsuperscript{15:5}\thinspace{}Can't Shun be said to have a\-chieved or\-der with\-out forc\-ing? What did he do? He just sat reverently on the throne, and faced due south.
\end{para}

\begin{para}
    \textsuperscript{15:6}\thinspace{}If your words are loy\-al and trust\-wor\-thy, and your con\-duct is sin\-cere and re\-spect\-ful, then your con\-duct will be prop\-er e\-ven a\-mong barbarians. If your words are not loy\-al and trust\-wor\-thy, and your con\-duct is not sin\-cere and re\-spect\-ful, how can your con\-duct be prop\-er, e\-ven in your own neigh\-bor\-hood?
\end{para}

\begin{para}
    \textsuperscript{15:8}\thinspace{}If you fail to speak to a man who is ca\-pa\-ble of un\-der\-stand\-ing, you waste the man. If you speak to a man who is in\-ca\-pa\-ble of un\-der\-stand\-ing, you waste your words. The wise waste nei\-ther men nor words.
\end{para}

\begin{para}
    \textsuperscript{15:9}\thinspace{}A be\-nev\-o\-lent man will not in\-jure his be\-nev\-o\-lence in or\-der to save his life, but he might sac\-ri\-fice his life in or\-der to a\-chieve be\-nev\-o\-lence.
\end{para}

\begin{para}
    \textsuperscript{15:10}\thinspace{}In what\-ev\-er coun\-try you're liv\-ing, serve its most vir\-tu\-ous officers, and be\-friend its most be\-nev\-o\-lent gentelmen. This is how to prac\-tice be\-nev\-o\-lence.
\end{para}

\begin{para}
    \textsuperscript{15:12}\thinspace{}If a man does\-n't con\-cern him\-self with what's far a\-way, he will have sorrows near at hand.
\end{para}

\begin{para}
    \textsuperscript{15:15}\thinspace{}If you de\-mand much from your\-self and lit\-tle from others, you will keep re\-sent\-ment at a dis\-tance.
\end{para}

\begin{para}
    \textsuperscript{15:18}\thinspace{}A su\-pe\-ri\-or man takes what is right as his ba\-sis. He practices it with pro\-pri\-e\-ty, demonstrates it with mod\-es\-ty, and completes it with good faith.
\end{para}

\begin{para}
    \textsuperscript{15:20}\thinspace{}A su\-pe\-ri\-or man hates the thought of leav\-ing the world with\-out mak\-ing a name for him\-self.
    \textsuperscript{21}\thinspace{}A su\-pe\-ri\-or man makes demands on him\-self; an in\-fe\-ri\-or man makes demands on others.
    \textsuperscript{22}\thinspace{}A su\-pe\-ri\-or man is con\-fi\-dent with\-out be\-ing con\-ten\-tious; gathers with others but does\-n't form cliques.
    \textsuperscript{23}\thinspace{}A su\-pe\-ri\-or man does\-n't pro\-mote a man be\-cause of his words, and does\-n't re\-ject the words be\-cause of the man.
\end{para}

\begin{para}
    \textsuperscript{15:27}\thinspace{}Clev\-er words dis\-rupt vir\-tue. Im\-pa\-tience in small matters disrupts great plans.
\end{para}

\begin{para}
    \textsuperscript{15:28}\thinspace{} When eve\-ry\-one loves a man, you must in\-ves\-ti\-gate. When eve\-ry\-one hates a man, you must in\-ves\-ti\-gate.
\end{para}

\begin{para}
    \textsuperscript{15:30}\thinspace{}To have a fault and not cor\-rect it is a fault in\-deed.
\end{para}

\begin{para}
    \textsuperscript{15:31}\thinspace{}I once spent a whole day with\-out eat\-ing and a whole night with\-out sleep\-ing in or\-der to think. I got noth\-ing out of it. It would have been bet\-ter to stud\-y.
\end{para}

\begin{para}
    \textsuperscript{15:32}\thinspace{}A su\-pe\-ri\-or man makes his plans a\-round the Way, not a\-round food. A su\-pe\-ri\-or man worries a\-bout the Way, not a\-bout pov\-er\-ty.
\end{para}

\begin{para}
    \textsuperscript{15:36}\thinspace{}In the pur\-suit of be\-nev\-o\-lence, don't yield e\-ven to your teach\-er.
\end{para}

\begin{para}
    \textsuperscript{15:37}\thinspace{}A su\-pe\-ri\-or man is prin\-ci\-pled, but not rig\-id.
\end{para}

\begin{para}
    \textsuperscript{15:38}\thinspace{}In serv\-ing your rul\-er, de\-vo\-tion to du\-ty co\-mes be\-fore any thought of com\-pen\-sa\-tion.
    \textsuperscript{39}\thinspace{}In teach\-ing, there is no such thing as so\-cial class.
\end{para}

\begin{para}
    \textsuperscript{15:40}\thinspace{}Those who fol\-low a dif\-fer\-ent Way can't make plans to\-geth\-er.
\end{para}

\begin{para}
    \textsuperscript{15:41}\thinspace{}In expressing your\-self, get the point a\-cross then stop.
\end{para}

\chapter*{Laozi}
\markboth{Laozi}{Laozi}
\addcontentsline{toc}{section}{Laozi}\begin{changemargin}{.6in}{.6in}
  \begin{center}
    \itshape
    An a\-bridg\-ment of the Daodejing.
  \end{center}
\end{changemargin}
\begin{para}
    \textsuperscript{1:1}\thinspace{}The Way that can be put in\-to words is not the con\-stant Way.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{2:1}\thinspace{}As soon as eve\-ry\-one in the world knows beau\-ty as beau\-ty, there is al\-read\-y ug\-li\-ness.
    \textsuperscript{2}\thinspace{}As soon as eve\-ry\-one knows good as good, there is bad.
\end{para}

\begin{para}
    \textsuperscript{2:3}\thinspace{}Be\-ing and non\-be\-ing pro\-duce each oth\-er;
    \textsuperscript{4}\thinspace{}dif\-fi\-cult and eas\-y com\-plete each oth\-er;
    \textsuperscript{5}\thinspace{}long and short shape each oth\-er;
    \textsuperscript{6}\thinspace{}high and low fill each oth\-er;
    \textsuperscript{7}\thinspace{}note and voice har\-mo\-nize with each oth\-er;
    \textsuperscript{8}\thinspace{}front and back fol\-low each oth\-er:
    \textsuperscript{9}\thinspace{}these are con\-stant.
\end{para}

\begin{para}
    \textsuperscript{2:10}\thinspace{}There\-fore the sage dwells in the work of nonforced ac\-tion,
    \textsuperscript{11}\thinspace{}and practices word\-less teach\-ing.
    \textsuperscript{12}\thinspace{}All things a\-rise, but he does\-n't start them;
    \textsuperscript{13}\thinspace{}he acts on their be\-half, but does\-n't make them de\-pend\-ent;
    \textsuperscript{14}\thinspace{}he succeeds, but does\-n't dwell on it.
    \textsuperscript{15}\thinspace{}It's on\-ly be\-cause he does\-n't dwell on it, that it does\-n't leave.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{3:1}\thinspace{}By not el\-e\-vat\-ing the wor\-thy, you keep the peo\-ple from con\-ten\-tion.
    \textsuperscript{2}\thinspace{}By not valuing rare treasures, you keep the peo\-ple from steal\-ing.
    \textsuperscript{3}\thinspace{}By not displaying what is de\-sir\-a\-ble, you keep the peo\-ple from con\-fu\-sion.
\end{para}

\begin{para}
    \textsuperscript{3:4}\thinspace{}There\-fore the sage governs by clear\-ing the people's minds, but fill\-ing their bellies; weakening their ambitions, but strengthening their bones.
\end{para}

\begin{para}
    \textsuperscript{3:5}\thinspace{}He con\-stant\-ly keeps the peo\-ple with\-out cun\-ning and with\-out desires
    \textsuperscript{6}\thinspace{}so that the intellectuals won't dare to med\-dle.
    \textsuperscript{7}\thinspace{}Act with\-out forc\-ing, then eve\-ry\-thing is governed prop\-er\-ly.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{7:5}\thinspace{}The sage puts him\-self in the back\-ground yet finds him\-self in the fore\-ground;
    \textsuperscript{6}\thinspace{}puts him\-self out of mind yet finds him\-self cared for.
    \textsuperscript{7}\thinspace{}Is\-n't it be\-cause he's self\-less that
    \textsuperscript{8}\thinspace{}he's a\-ble to a\-chieve his self-interest?
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{8:1}\thinspace{}The highest good is like wa\-ter.
    \textsuperscript{2}\thinspace{}Water's good\-ness is that it benefits all things,
    \textsuperscript{3}\thinspace{}yet it strives to dwell in the places that the crowd hates,
    \textsuperscript{4}\thinspace{}there\-fore it co\-mes close to the Way.
\end{para}

\begin{para}
    \textsuperscript{8:6}\thinspace{}In a dwell\-ing, its good\-ness is lo\-ca\-tion;
    \textsuperscript{7}\thinspace{}in a mind, depth;
    \textsuperscript{8}\thinspace{}in giv\-ing, be\-ing like Heav\-en;
    \textsuperscript{9}\thinspace{}in speak\-ing, sin\-cer\-i\-ty;
    \textsuperscript{10}\thinspace{}in gov\-ern\-ment, or\-der;
    \textsuperscript{11}\thinspace{}in han\-dling af\-fairs, a\-bil\-i\-ty;
    \textsuperscript{12}\thinspace{}in ac\-tion, time\-li\-ness.
\end{para}

\begin{para}
    \textsuperscript{8:13}\thinspace{}Be\-cause there is no con\-ten\-tion, there is no blame.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{9:1}\thinspace{}If you hold and o\-ver\-fill a cup, it's not as good as stop\-ping short.
    \textsuperscript{2}\thinspace{}If you ham\-mer and oversharpen a blade, you can't main\-tain it for long.
    \textsuperscript{3}\thinspace{}If gold and jade fill your rooms, no one is a\-ble to pro\-tect it.
    \textsuperscript{4}\thinspace{}If sta\-tus and wealth lead to ar\-ro\-gance, you bring dis\-as\-ter up\-on your\-self.
\end{para}

\begin{para}
    \textsuperscript{9:5}\thinspace{}When the work is done, with\-draw; this is the Way of Heav\-en.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{11:1}\thinspace{}Thir\-ty spokes come to\-geth\-er at one hub,
    \textsuperscript{2}\thinspace{}but it's the emp\-ti\-ness in\-side the hub that makes it use\-ful to the cart.
    \textsuperscript{3}\thinspace{}We form clay in\-to a pot,
    \textsuperscript{4}\thinspace{}but it's the emp\-ti\-ness in\-side that makes the clay pot use\-ful.
    \textsuperscript{5}\thinspace{}We chis\-el out doors and windows to make a room,
    \textsuperscript{6}\thinspace{}but it's the emp\-ti\-ness in\-side that makes the room use\-ful.
\end{para}

\begin{para}
    \textsuperscript{11:7}\thinspace{}There\-fore, what is pres\-ent brings ben\-e\-fit,
    \textsuperscript{8}\thinspace{}but what is ab\-sent makes it use\-ful.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{15:1}\thinspace{}In an\-tiq\-ui\-ty, those who were good at the Way
    \textsuperscript{2}\thinspace{}were sub\-tle, won\-der\-ful, mys\-te\-ri\-ous, pen\-e\-trat\-ing, and too deep to know.
    \textsuperscript{3}\thinspace{}Be\-cause they can't be known, we're forced to de\-scribe their ap\-pear\-ance:
    \textsuperscript{4}\thinspace{}Hes\-i\-tant, as if cross\-ing a riv\-er in win\-ter.
    \textsuperscript{5}\thinspace{}A\-lert, as if fearing the neighbors on all sides.
    \textsuperscript{6}\thinspace{}Sol\-emn, like a visiting guest.
    \textsuperscript{7}\thinspace{}Yield\-ing, like ice a\-bout to melt.
    \textsuperscript{8}\thinspace{}Gen\-u\-ine, like un\-carved wood.
    \textsuperscript{9}\thinspace{}O\-paque, like mud\-dy wa\-ter.
    \textsuperscript{10}\thinspace{}Vast, like a val\-ley.
\end{para}

\begin{para}
    \textsuperscript{15:11}\thinspace{}If mud\-dy wa\-ter is still, it grad\-u\-al\-ly becomes clear.
    \textsuperscript{12}\thinspace{}If some\-thing at rest is moved, it grad\-u\-al\-ly co\-mes to life.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{17:1}\thinspace{}With the ver\-y best rulers, those be\-low on\-ly know that the rulers ex\-ist,
    \textsuperscript{2}\thinspace{}the next best are the rulers they love and praise,
    \textsuperscript{3}\thinspace{}next are the rulers they fear,
    \textsuperscript{4}\thinspace{}and the worst are the rulers they des\-pise.
\end{para}

\begin{para}
    \textsuperscript{17:7}\thinspace{}When the task is ac\-com\-plished and the work is done, all the com\-mon peo\-ple say, “We did it our\-selves.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{18:1}\thinspace{}It's when the great Way is a\-ban\-doned that be\-nev\-o\-lence and right\-eous\-ness ap\-pear.
    \textsuperscript{2}\thinspace{}It's when learn\-ing and clev\-er\-ness e\-merge that the great hy\-poc\-ri\-sy appears.
    \textsuperscript{3}\thinspace{}It's when fam\-i\-ly relationships are not har\-mo\-ni\-ous that fil\-ial pi\-e\-ty and pa\-ren\-tal com\-pas\-sion ap\-pear.
    \textsuperscript{4}\thinspace{}It's when the coun\-try is con\-fused and cha\-ot\-ic that up\-right officials ap\-pear.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{19:5}\thinspace{}Dem\-on\-strate plain\-ness, em\-brace sim\-plic\-i\-ty, have lit\-tle self\-ish\-ness and few desires.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{22:1}\thinspace{}Bent o\-ver then pre\-served whole;
    \textsuperscript{2}\thinspace{}crook\-ed then straight;
    \textsuperscript{3}\thinspace{}hol\-low then filled;
    \textsuperscript{4}\thinspace{}worn out then renewed;
    \textsuperscript{5}\thinspace{}hav\-ing lit\-tle then sat\-is\-fied;
    \textsuperscript{6}\thinspace{}hav\-ing much then con\-fused.
    \textsuperscript{7}\thinspace{}There\-fore the sage holds to the one and becomes shep\-herd to the world.
\end{para}

\begin{para}
    \textsuperscript{22:8}\thinspace{}He does\-n't pro\-mote him\-self, so he's ac\-knowl\-edged.
    \textsuperscript{9}\thinspace{}He does\-n't dis\-play him\-self, so he shines bright\-ly.
    \textsuperscript{10}\thinspace{}He does\-n't brag a\-bout him\-self, so he has mer\-it.
    \textsuperscript{11}\thinspace{}He's not boast\-ful, so he can en\-dure.
    \textsuperscript{12}\thinspace{}It's on\-ly be\-cause he does\-n't con\-tend that no one can con\-tend with him.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{23:1}\thinspace{}It's nat\-u\-ral to speak rare\-ly.
    \textsuperscript{2}\thinspace{}A vi\-o\-lent wind does\-n't last all morn\-ing.
    \textsuperscript{3}\thinspace{}A tor\-ren\-tial rain does\-n't last all day.
    \textsuperscript{5}\thinspace{}If heav\-en and earth can't sus\-tain it for long, how much less can man?
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{24:1}\thinspace{}He who blows hard can't stand.
    \textsuperscript{2}\thinspace{}He who promotes him\-self is\-n't ac\-knowl\-edged.
    \textsuperscript{3}\thinspace{}He who displays him\-self does\-n't shine bright\-ly.
    \textsuperscript{5}\thinspace{}He who brags a\-bout him\-self has no mer\-it.
    \textsuperscript{6}\thinspace{}He who boasts a\-bout him\-self won't en\-dure.
    \textsuperscript{8}\thinspace{}These are so de\-spised that e\-ven those who want things don't a\-bide them.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{27:1}\thinspace{}A skill\-ful trav\-el\-er leaves no ruts or tracks.
    \textsuperscript{2}\thinspace{}A skill\-ful speak\-er makes no slips or gaffes.
    \textsuperscript{3}\thinspace{}A skill\-ful count\-er needs no tallies or papers.
\end{para}

\begin{para}
    \textsuperscript{27:6}\thinspace{}The sage is al\-ways skill\-ful at sav\-ing peo\-ple, and nev\-er abandons any\-one;
    \textsuperscript{7}\thinspace{}with things, he nev\-er abandons val\-u\-a\-ble prop\-er\-ty.
\end{para}

\begin{para}
    \textsuperscript{27:9}\thinspace{}Skill\-ful peo\-ple are the teachers of skill\-ful peo\-ple;
    \textsuperscript{10}\thinspace{}un\-skilled peo\-ple are the raw ma\-te\-ri\-als of skill\-ful peo\-ple.
    \textsuperscript{11}\thinspace{}He who does\-n't val\-ue his teach\-er and cher\-ish his raw ma\-te\-ri\-als is great\-ly con\-fused, no mat\-ter how clev\-er he is.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{29:1}\thinspace{}Those who want to take the world and rule it,
    \textsuperscript{2}\thinspace{}I see that they won't suc\-ceed.
    \textsuperscript{3}\thinspace{}The world is a sa\-cred ves\-sel,
    \textsuperscript{4}\thinspace{}and it can't be ruled.
    \textsuperscript{5}\thinspace{}Those who rule it ru\-in it.
    \textsuperscript{6}\thinspace{}Those who hold it lose it.
\end{para}

\begin{para}
    \textsuperscript{29:7}\thinspace{}It's the na\-ture of things that some lead and some fol\-low,
    \textsuperscript{8}\thinspace{}some blow hot and some blow cold,
    \textsuperscript{9}\thinspace{}some are strong and some are weak,
    \textsuperscript{10}\thinspace{}some cul\-ti\-vate and some de\-te\-ri\-o\-rate.
    \textsuperscript{11}\thinspace{}There\-fore the sage avoids extremes, ar\-ro\-gance, and ex\-trav\-a\-gance.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{30:1}\thinspace{}Those with the Way who as\-sist the rul\-er don't use the force of arms to con\-quer the world;
    \textsuperscript{2}\thinspace{}such methods would on\-ly back\-fire.
    \textsuperscript{3}\thinspace{}Wher\-ev\-er armies camp, brambles and thorns will grow.
\end{para}

\begin{para}
    \textsuperscript{30:5}\thinspace{}A skill\-ful lead\-er achieves results then stops,
    \textsuperscript{6}\thinspace{}and does\-n't use the oc\-ca\-sion to sieze pow\-er.
    \textsuperscript{7}\thinspace{}He achieves results with\-out ar\-ro\-gance.
    \textsuperscript{8}\thinspace{}He achieves results with\-out prais\-ing his deeds.
    \textsuperscript{9}\thinspace{}He achieves results with\-out boasting.
    \textsuperscript{10}\thinspace{}He achieves results on\-ly as a last re\-sort.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{31:1}\thinspace{}The mil\-i\-tar\-y is an om\-i\-nous in\-stru\-ment.
    \textsuperscript{2}\thinspace{}It's so de\-spised that e\-ven those who want things don't a\-bide it.
\end{para}

\begin{para}
    \textsuperscript{31:5}\thinspace{}The mil\-i\-tar\-y is an om\-i\-nous in\-stru\-ment,
    \textsuperscript{6}\thinspace{}so on\-ly use it as a last re\-sort;
    \textsuperscript{7}\thinspace{}with calm de\-tach\-ment is best.
    \textsuperscript{8}\thinspace{}Don't glo\-ri\-fy it;
    \textsuperscript{9}\thinspace{}to glo\-ri\-fy it would be to de\-light in kill\-ing peo\-ple.
    \textsuperscript{10}\thinspace{}He who delights in kill\-ing peo\-ple can't a\-chieve his ambitions in the world.
    \textsuperscript{16}\thinspace{}When man\-y peo\-ple have been killed,
    \textsuperscript{17}\thinspace{}weep for them with sor\-row.
    \textsuperscript{18}\thinspace{}When the war has been won, treat it like a fu\-ner\-al cer\-e\-mo\-ny.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{33:1}\thinspace{}He who knows others is in\-tel\-li\-gent;
    \textsuperscript{2}\thinspace{}he who knows him\-self is enlightened.
    \textsuperscript{3}\thinspace{}He who overcomes others is force\-ful;
    \textsuperscript{4}\thinspace{}he who overcomes him\-self is pow\-er\-ful.
    \textsuperscript{5}\thinspace{}He who knows con\-tent\-ment is rich.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{34:1}\thinspace{}The Way floats and drifts;
    \textsuperscript{2}\thinspace{}it can go left or right.
    \textsuperscript{4}\thinspace{}When the task is ac\-com\-plished and the work is done, it claims no cred\-it.
\end{para}

\begin{para}
    \textsuperscript{34:5}\thinspace{}All things turn to it
    \textsuperscript{6}\thinspace{}and yet it does\-n't act as their mas\-ter,
    \textsuperscript{8}\thinspace{}so it can be called a small thing.
\end{para}

\begin{para}
    \textsuperscript{34:9}\thinspace{}All things turn to it
    \textsuperscript{10}\thinspace{}and yet it does\-n't act as their mas\-ter,
    \textsuperscript{11}\thinspace{}so it can be called a great thing.
\end{para}

\begin{para}
    \textsuperscript{34:12}\thinspace{}That's why the sage is a\-ble to ac\-com\-plish great things:
    \textsuperscript{13}\thinspace{}he does\-n't try to do great things, there\-fore he accomplishes great things.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{35:3}\thinspace{}Com\-ing to you, the world suffers no harm, but finds safe\-ty and peace.
\end{para}

\begin{para}
    \textsuperscript{35:4}\thinspace{}Pass\-ing travelers will stop for mu\-sic and food.
    \textsuperscript{5}\thinspace{}But if the Way were de\-scribed,
    \textsuperscript{6}\thinspace{}it would seem in\-sip\-id and with\-out fla\-vor.
    \textsuperscript{7}\thinspace{}When you look at it, it's not suf\-fi\-cient to be seen;
    \textsuperscript{8}\thinspace{}when you lis\-ten to it, it's not suf\-fi\-cient to be heard;
    \textsuperscript{9}\thinspace{}yet when you use it, it can't be exhausted.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{36:1}\thinspace{}If you want to shrink some\-thing,
    \textsuperscript{2}\thinspace{}it must first be ex\-pand\-ed;
    \textsuperscript{3}\thinspace{}if you want to weak\-en some\-thing,
    \textsuperscript{4}\thinspace{}it must first be strengthened;
    \textsuperscript{5}\thinspace{}if you want to leave some\-thing,
    \textsuperscript{6}\thinspace{}it must first be joined;
    \textsuperscript{7}\thinspace{}if you want to take some\-thing a\-way,
    \textsuperscript{8}\thinspace{}it must first be giv\-en.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{38:1}\thinspace{}A man of su\-pe\-ri\-or vir\-tue does\-n't cling to vir\-tue,
    \textsuperscript{2}\thinspace{}there\-fore he has true vir\-tue.
    \textsuperscript{5}\thinspace{}A man of su\-pe\-ri\-or vir\-tue acts with\-out force,
    \textsuperscript{6}\thinspace{}and with\-out an a\-gen\-da.
\end{para}

\begin{para}
    \textsuperscript{38:16}\thinspace{}There\-fore, af\-ter los\-ing the Way there is vir\-tue;
    \textsuperscript{17}\thinspace{}af\-ter los\-ing vir\-tue there is be\-nev\-o\-lence;
    \textsuperscript{18}\thinspace{}af\-ter los\-ing be\-nev\-o\-lence there is right\-eous\-ness;
    \textsuperscript{19}\thinspace{}af\-ter los\-ing right\-eous\-ness there is pro\-pri\-e\-ty.
\end{para}

\begin{para}
    \textsuperscript{38:20}\thinspace{}As for pro\-pri\-e\-ty, it's a thin ve\-neer of loy\-al\-ty and sin\-cer\-i\-ty,
    \textsuperscript{21}\thinspace{}and the be\-gin\-ning of dis\-or\-der.
\end{para}

\begin{para}
    \textsuperscript{38:24}\thinspace{}There\-fore a tru\-ly great man dwells in the thick sub\-stance,
    \textsuperscript{25}\thinspace{}not in the thin ve\-neer;
    \textsuperscript{26}\thinspace{}in the fruit, not in the flow\-er.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{43:1}\thinspace{}The softest, most pli\-a\-ble thing in the world
    \textsuperscript{2}\thinspace{}runs rough\-shod o\-ver the hardest, most sol\-id thing in the world.
    \textsuperscript{3}\thinspace{}That which has no sub\-stance enters in\-to that which has no openings.
    \textsuperscript{4}\thinspace{}That's how I know the benefits of nonforced ac\-tion.
\end{para}

\begin{para}
    \textsuperscript{43:5}\thinspace{}Teachings that use no words,
    \textsuperscript{6}\thinspace{}and the benefits of nonforced ac\-tion:
    \textsuperscript{7}\thinspace{}few in the world can un\-der\-stand them.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{44:4}\thinspace{}Ex\-ces\-sive at\-tach\-ment co\-mes at great cost,
    \textsuperscript{5}\thinspace{}and great ac\-cu\-mu\-la\-tion leads to heav\-y loss.
    \textsuperscript{6}\thinspace{}There\-fore, know\-ing con\-tent\-ment avoids dis\-grace,
    \textsuperscript{7}\thinspace{}and know\-ing when to stop avoids dan\-ger;
    \textsuperscript{8}\thinspace{}thus you can en\-dure for a long time.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{46:1}\thinspace{}When the world has the Way,
    \textsuperscript{2}\thinspace{}fi\-ne hors\-es are rel\-e\-gat\-ed to fer\-ti\-liz\-ing the fields.
    \textsuperscript{3}\thinspace{}When the world lacks the Way,
    \textsuperscript{4}\thinspace{}warhorses are bred just out\-side the cit\-y walls.
\end{para}

\begin{para}
    \textsuperscript{46:5}\thinspace{}There is no greater crime than de\-sire.
    \textsuperscript{6}\thinspace{}There is no greater dis\-as\-ter than be\-ing dis\-con\-tent.
    \textsuperscript{7}\thinspace{}There is no greater fault than want\-ing more.
    \textsuperscript{8}\thinspace{}There\-fore the sat\-is\-fac\-tion that co\-mes from be\-ing con\-tent is last\-ing sat\-is\-fac\-tion.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{47:1}\thinspace{}You don't have to go out the door to know the world.
    \textsuperscript{2}\thinspace{}You don't have to peek out the win\-dow to know the Way of Heav\-en.
    \textsuperscript{3}\thinspace{}The far\-ther you go, the less you know.
\end{para}

\begin{para}
    \textsuperscript{47:4}\thinspace{}There\-fore the sage does\-n't trav\-el but knows;
    \textsuperscript{5}\thinspace{}does\-n't look, but discerns;
    \textsuperscript{6}\thinspace{}and does\-n't do, but succeeds.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{48:1}\thinspace{}He who pursues learn\-ing gains some\-thing each day.
    \textsuperscript{2}\thinspace{}He who hears the Way loses some\-thing each day.
    \textsuperscript{3}\thinspace{}He loses, and loses a\-gain, un\-til he acts with\-out force.
    \textsuperscript{4}\thinspace{}He acts with\-out force and yet noth\-ing is left un\-done.
\end{para}

\begin{para}
    \textsuperscript{48:5}\thinspace{}If you want to rule the world, con\-stant\-ly a\-void in\-ter\-fer\-ing with it.
    \textsuperscript{6}\thinspace{}If you in\-ter\-fere, you're un\-fit to rule the world.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{49:1}\thinspace{}The sage con\-stant\-ly has no mind of his own.
    \textsuperscript{2}\thinspace{}He takes the mind of the peo\-ple as his mind.
\end{para}

\begin{para}
    \textsuperscript{49:3}\thinspace{}He's good to good peo\-ple
    \textsuperscript{4}\thinspace{}and al\-so good to bad peo\-ple,
    \textsuperscript{5}\thinspace{}thus he gains in good\-ness.
    \textsuperscript{6}\thinspace{}He trusts trust\-wor\-thy peo\-ple
    \textsuperscript{7}\thinspace{}and al\-so trusts un\-trust\-wor\-thy peo\-ple,
    \textsuperscript{8}\thinspace{}thus he gains in trust.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{53:1}\thinspace{}If I had the least bit of knowl\-edge,
    \textsuperscript{2}\thinspace{}I would walk on the great Way
    \textsuperscript{3}\thinspace{}and my on\-ly fear would be straying from it.
    \textsuperscript{4}\thinspace{}The great Way is ver\-y much lev\-el,
    \textsuperscript{5}\thinspace{}but peo\-ple ver\-y much pre\-fer moun\-tain trails.
\end{para}

\begin{para}
    \textsuperscript{53:6}\thinspace{}The court is full of splen\-dor,
    \textsuperscript{7}\thinspace{}but the fields are full of weeds,
    \textsuperscript{8}\thinspace{}and the granaries are full of noth\-ing;
    \textsuperscript{9}\thinspace{}their clothes are embroidered and col\-or\-ful,
    \textsuperscript{10}\thinspace{}their belts have sharp swords,
    \textsuperscript{11}\thinspace{}they're stuffed with food,
    \textsuperscript{12}\thinspace{}and have mon\-ey and possessions in ex\-cess:
    \textsuperscript{13}\thinspace{}this is called rob\-ber\-y,
    \textsuperscript{14}\thinspace{}and rob\-ber\-y is cer\-tain\-ly not the Way.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{56:1}\thinspace{}Those who know it, don't talk a\-bout it;
    \textsuperscript{2}\thinspace{}those who talk a\-bout it, don't know it.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{57:1}\thinspace{}Gov\-ern the state with jus\-tice,
    \textsuperscript{2}\thinspace{}de\-ploy the mil\-i\-tar\-y with sur\-prise,
    \textsuperscript{3}\thinspace{}and take the world with\-out in\-ter\-fer\-ence.
    \textsuperscript{4}\thinspace{}How do I know this? From the fol\-low\-ing:
\end{para}

\begin{para}
    \textsuperscript{57:5}\thinspace{}The world has man\-y prohibitions and taboos,
    \textsuperscript{6}\thinspace{}and yet the peo\-ple are full of pov\-er\-ty.
    \textsuperscript{7}\thinspace{}The peo\-ple have man\-y ben\-e\-fi\-cial tools,
    \textsuperscript{8}\thinspace{}and yet the state and fam\-i\-ly are in\-creas\-ing\-ly faint.
    \textsuperscript{9}\thinspace{}Peo\-ple have man\-y clev\-er tricks,
    \textsuperscript{10}\thinspace{}and yet strange things in\-creas\-ing\-ly hap\-pen.
    \textsuperscript{11}\thinspace{}The laws are in\-creas\-ing\-ly pub\-li\-cized,
    \textsuperscript{12}\thinspace{}and yet there are man\-y thieves and robbers.
\end{para}

\begin{para}
    \textsuperscript{57:13}\thinspace{}There\-fore, in the words of the sage:
    \textsuperscript{14}\thinspace{}“I act with\-out force, and the peo\-ple trans\-form them\-selves.
    \textsuperscript{15}\thinspace{}I en\-joy qui\-et, and the peo\-ple cor\-rect them\-selves.
    \textsuperscript{16}\thinspace{}I don't intefere, and the peo\-ple pros\-per them\-selves.
    \textsuperscript{17}\thinspace{}I have no desires, and the peo\-ple sim\-pli\-fy them\-selves.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{58:5}\thinspace{}Good for\-tune perch\-es on mis\-for\-tune;
    \textsuperscript{6}\thinspace{}mis\-for\-tune lurks be\-neath good for\-tune.
    \textsuperscript{7}\thinspace{}Who knows where it will end?
\end{para}

\begin{para}
    \textsuperscript{58:8}\thinspace{}Noth\-ing is ob\-vi\-ous.
    \textsuperscript{9}\thinspace{}Ob\-vi\-ous a\-gain becomes sur\-pris\-ing;
    \textsuperscript{10}\thinspace{}good a\-gain becomes mon\-strous.
    \textsuperscript{11}\thinspace{}The peo\-ple have been con\-fused for a long time.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{59:1}\thinspace{}In governing the peo\-ple and serv\-ing Heav\-en,
    \textsuperscript{2}\thinspace{}there is noth\-ing like fru\-gal\-i\-ty.
    \textsuperscript{3}\thinspace{}On\-ly with fru\-gal\-i\-ty can you sub\-mit ear\-ly.
    \textsuperscript{4}\thinspace{}Sub\-mit\-ting ear\-ly is called re\-peat\-ed\-ly ac\-cu\-mu\-lat\-ing vir\-tue.
    \textsuperscript{5}\thinspace{}If you re\-peat\-ed\-ly ac\-cu\-mu\-late vir\-tue, then there is noth\-ing you can't o\-ver\-come.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{62:2}\thinspace{}The Way is the treas\-ure of a good man,
    \textsuperscript{3}\thinspace{}and the shel\-ter of a bad man.
\end{para}

\begin{para}
    \textsuperscript{62:4}\thinspace{}Beau\-ti\-ful words can buy hon\-or,
    \textsuperscript{5}\thinspace{}and good deeds can raise you a\-bove others.
\end{para}

\begin{para}
    \textsuperscript{62:6}\thinspace{}E\-ven if a man is bad, how can you a\-ban\-don him?
\end{para}

\begin{para}
    \textsuperscript{62:7}\thinspace{}At the emperor's cor\-o\-na\-tion,
    \textsuperscript{9}\thinspace{}of\-fer\-ing jade discs brought in by a team of hors\-es
    \textsuperscript{10}\thinspace{}is not as good as sit\-ting still and of\-fer\-ing this Way.
\end{para}

\begin{para}
    \textsuperscript{62:11}\thinspace{}Why was this Way so val\-u\-a\-ble to the ancients?
    \textsuperscript{12}\thinspace{}Is\-n't it said, “Those who seek, with this will ob\-tain,
    \textsuperscript{13}\thinspace{}and those who have guilt, with this will es\-cape”?
    \textsuperscript{14}\thinspace{}There\-fore it's the most val\-u\-a\-ble thing in the world.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{63:1}\thinspace{}Act with\-out forc\-ing,
    \textsuperscript{2}\thinspace{}serve with\-out in\-ter\-fer\-ing,
    \textsuperscript{3}\thinspace{}sa\-vor the fla\-vor\-less,
    \textsuperscript{4}\thinspace{}con\-sid\-er the small as great and the few as man\-y,
    \textsuperscript{5}\thinspace{}re\-spond to re\-sent\-ment with vir\-tue.
\end{para}

\begin{para}
    \textsuperscript{63:6}\thinspace{}Plan for some\-thing dif\-fi\-cult while it's still eas\-y.
    \textsuperscript{7}\thinspace{}Deal with some\-thing great while it's still small.
    \textsuperscript{8}\thinspace{}The dif\-fi\-cult things of the world start off eas\-y.
    \textsuperscript{9}\thinspace{}The great things of the world start off small.
    \textsuperscript{10}\thinspace{}There\-fore, the sage nev\-er does great things,
    \textsuperscript{11}\thinspace{}and thus he is a\-ble to ac\-com\-plish great things.
\end{para}

\begin{para}
    \textsuperscript{63:12}\thinspace{}He who makes promises light\-ly is sure to have lit\-tle trust,
    \textsuperscript{13}\thinspace{}and he who treats man\-y things as eas\-y is sure to have man\-y difficul\-ties.
    \textsuperscript{14}\thinspace{}There\-fore, the sage treats things as dif\-fi\-cult,
    \textsuperscript{15}\thinspace{}and thus, in the end, noth\-ing is dif\-fi\-cult.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{64:1}\thinspace{}It's eas\-y to hold what is at rest.
    \textsuperscript{2}\thinspace{}It's eas\-y to plan for what has\-n't be\-gun.
    \textsuperscript{3}\thinspace{}It's eas\-y to break what is brit\-tle.
    \textsuperscript{4}\thinspace{}It's eas\-y to scat\-ter what is ti\-ny.
    \textsuperscript{5}\thinspace{}Deal with it be\-fore it happens;
    \textsuperscript{6}\thinspace{}man\-age it be\-fore it's cha\-ot\-ic.
\end{para}

\begin{para}
    \textsuperscript{64:7}\thinspace{}A tree that's thick e\-nough to em\-brace rises from a ti\-ny shoot.
    \textsuperscript{8}\thinspace{}A ter\-race that's nine stories tall rises from a heap of earth.
    \textsuperscript{9}\thinspace{}A wall that's three hun\-dred yards high begins at the ground be\-neath your feet.
\end{para}

\begin{para}
    \textsuperscript{64:10}\thinspace{}Those who rule it ru\-in it.
    \textsuperscript{11}\thinspace{}Those who hold it lose it.
    \textsuperscript{12}\thinspace{}There\-fore, the sage rules noth\-ing and thus ruins noth\-ing.
    \textsuperscript{13}\thinspace{}He holds on to noth\-ing and thus loses noth\-ing.
\end{para}

\begin{para}
    \textsuperscript{64:14}\thinspace{}Peo\-ple u\-su\-al\-ly fail in their endeavors on the verge of suc\-cess.
    \textsuperscript{15}\thinspace{}There\-fore we say, “If you're as care\-ful at the end as you are at the be\-gin\-ning,
    \textsuperscript{16}\thinspace{}you'll have no fail\-ure.”
\end{para}

\begin{para}
    \textsuperscript{64:17}\thinspace{}The sage desires not to de\-sire,
    \textsuperscript{18}\thinspace{}and does\-n't val\-ue goods that are hard to ob\-tain;
    \textsuperscript{19}\thinspace{}he learns not to stud\-y,
    \textsuperscript{20}\thinspace{}and returns to what the mul\-ti\-tude has passed o\-ver.
    \textsuperscript{21}\thinspace{}He could help all things to be nat\-u\-ral,
    \textsuperscript{22}\thinspace{}but dares not do it.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{66:1}\thinspace{}The rivers and lakes are a\-ble to be the kings of the valleys
    \textsuperscript{2}\thinspace{}be\-cause of their skill at be\-ing low.
    \textsuperscript{4}\thinspace{}There\-fore, want\-ing to be a\-bove the peo\-ple, the sage must speak to them hum\-bly from be\-low;
    \textsuperscript{5}\thinspace{}want\-ing to be be\-fore the peo\-ple, he must put him\-self be\-hind them.
    \textsuperscript{6}\thinspace{}There\-fore, he dwells a\-bove them yet the peo\-ple don't feel that he's a bur\-den.
    \textsuperscript{7}\thinspace{}He dwells be\-fore them yet the peo\-ple don't feel that he's an ob\-struc\-tion.
    \textsuperscript{8}\thinspace{}Eve\-ry\-one in the world delights in pro\-mot\-ing him and nev\-er tires of him.
\end{para}

\begin{para}
    \textsuperscript{66:9}\thinspace{}It's be\-cause he does\-n't con\-tend
    \textsuperscript{10}\thinspace{}that no one in the world can con\-tend with him.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{67:4}\thinspace{}I con\-stant\-ly have three treasures which I hold and treas\-ure:
    \textsuperscript{5}\thinspace{}the first is com\-pas\-sion,
    \textsuperscript{6}\thinspace{}the sec\-ond is fru\-gal\-i\-ty,
    \textsuperscript{7}\thinspace{}the third is not dar\-ing to be pre\-em\-i\-nent in the world.
    \textsuperscript{8}\thinspace{}It's be\-cause of com\-pas\-sion that I can be cou\-ra\-geous;
    \textsuperscript{9}\thinspace{}it's be\-cause of fru\-gal\-i\-ty that I can be gen\-er\-ous;
    \textsuperscript{10}\thinspace{}it's be\-cause of not dar\-ing to be pre\-em\-i\-nent in the world that I can be head of all things.
\end{para}

\begin{para}
    \textsuperscript{67:11}\thinspace{}Now, abandoning com\-pas\-sion to be cou\-ra\-geous,
    \textsuperscript{12}\thinspace{}abandoning fru\-gal\-i\-ty to be gen\-er\-ous,
    \textsuperscript{13}\thinspace{}abandoning the rear to be in front:
    \textsuperscript{14}\thinspace{}this is courting death.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{68:1}\thinspace{}A skill\-ful war\-ri\-or does\-n't show his might.
    \textsuperscript{2}\thinspace{}A skill\-ful fight\-er does\-n't get an\-gry.
    \textsuperscript{3}\thinspace{}A skill\-ful vic\-tor does\-n't en\-gage his en\-e\-my.
    \textsuperscript{4}\thinspace{}A skill\-ful em\-ploy\-er places him\-self be\-low others.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{73:1}\thinspace{}Those who are brave in dar\-ing will die;
    \textsuperscript{2}\thinspace{}those who are brave in cau\-tion will live.
    \textsuperscript{3}\thinspace{}In these two cases, one is ben\-e\-fi\-cial and the oth\-er is harm\-ful.
    \textsuperscript{4}\thinspace{}Who understands why Heav\-en hates what it does?
\end{para}

\begin{para}
    \textsuperscript{73:6}\thinspace{}The Way of Heav\-en does\-n't fight yet wins skill\-ful\-ly,
    \textsuperscript{7}\thinspace{}does\-n't speak yet responds skill\-ful\-ly,
    \textsuperscript{8}\thinspace{}is\-n't summoned yet co\-mes on its own,
    \textsuperscript{9}\thinspace{}is at ease yet plans skill\-ful\-ly.
\end{para}

\begin{para}
    \textsuperscript{73:10}\thinspace{}The net of Heav\-en is vast; it has a coarse mesh yet noth\-ing slips through.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{75:1}\thinspace{}The peo\-ple starve be\-cause those a\-bove con\-sume too much in taxes.
    \textsuperscript{2}\thinspace{}The peo\-ple reb\-el be\-cause those a\-bove in\-ter\-fere.
    \textsuperscript{3}\thinspace{}The peo\-ple scoff at death be\-cause those a\-bove de\-mand so much from life.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{76:1}\thinspace{}While liv\-ing, a man is sup\-ple and soft,
    \textsuperscript{2}\thinspace{}but once dead, he is hard and rig\-id.
    \textsuperscript{3}\thinspace{}While liv\-ing, all things, grasses, and trees are sup\-ple and crisp,
    \textsuperscript{4}\thinspace{}but once dead, they are withered and dry.
\end{para}

\begin{para}
    \textsuperscript{76:5}\thinspace{}There\-fore it is said: those who are hard and rig\-id are the followers of death;
    \textsuperscript{6}\thinspace{}those who are sup\-ple and soft are the followers of life.
    \textsuperscript{7}\thinspace{}There\-fore, a rig\-id ar\-my won't win,
    \textsuperscript{8}\thinspace{}and a rig\-id tree will snap.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{77:1}\thinspace{}The Way of Heav\-en is like an arch\-er draw\-ing his bow.
    \textsuperscript{2}\thinspace{}If it's too high, he presses down; if it's too low, he lifts up;
    \textsuperscript{3}\thinspace{}if there is ex\-cess, he reduces it;
    \textsuperscript{4}\thinspace{}if there is not e\-nough, he adds to it.
    \textsuperscript{5}\thinspace{}The Way of Heav\-en is al\-so to re\-duce what has ex\-cess and add to what does\-n't have e\-nough.
\end{para}

\begin{para}
    \textsuperscript{77:6}\thinspace{}The Way of man is dif\-fer\-ent:
    \textsuperscript{7}\thinspace{}he reduces what does\-n't have e\-nough and offers it to that which has ex\-cess.
\end{para}

\begin{para}
    \textsuperscript{77:8}\thinspace{}Who can have ex\-cess and of\-fer it to Heav\-en?
    \textsuperscript{9}\thinspace{}On\-ly he who has the Way.
\end{para}

\begin{para}
    \textsuperscript{77:10}\thinspace{}There\-fore, the sage takes actions with\-out claiming the results;
    \textsuperscript{11}\thinspace{}he succeeds with\-out dwell\-ing on it;
    \textsuperscript{12}\thinspace{}sim\-i\-lar\-ly, he does\-n't de\-sire to make a dis\-play of his wor\-thi\-ness.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{79:1}\thinspace{}When har\-mo\-ny is made af\-ter great re\-sent\-ment,
    \textsuperscript{2}\thinspace{}there is al\-ways some re\-sent\-ment remaining.
    \textsuperscript{3}\thinspace{}How can this be con\-sid\-ered good?
    \textsuperscript{4}\thinspace{}There\-fore, the sage keeps his part of the con\-tract, but makes no demands on the oth\-er par\-ty.
\end{para}

\begin{para}
    \textsuperscript{79:7}\thinspace{}The Way of Heav\-en has no favorites,
    \textsuperscript{8}\thinspace{}it's con\-stant\-ly on the side of the good man.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{80:1}\thinspace{}Let there be a small state with few peo\-ple.
    \textsuperscript{2}\thinspace{}Let them have pow\-er\-ful tools and not use them.
    \textsuperscript{3}\thinspace{}Let the peo\-ple take death se\-ri\-ous\-ly and not mi\-grate far a\-way.
    \textsuperscript{4}\thinspace{}There are boats and vehicles but no rea\-son to ride them.
    \textsuperscript{5}\thinspace{}There are ar\-mor and weapons but no rea\-son to pa\-rade them.
    \textsuperscript{6}\thinspace{}Let the peo\-ple keep their records with knot\-ted string a\-gain.
    \textsuperscript{7}\thinspace{}Their food is sweet,
    \textsuperscript{8}\thinspace{}their clothes are beau\-ti\-ful,
    \textsuperscript{9}\thinspace{}their homes are se\-cure,
    \textsuperscript{10}\thinspace{}and their cus\-toms are en\-joy\-a\-ble.
\end{para}

\begin{para}
    \textsuperscript{80:11}\thinspace{}Neighboring states can see each oth\-er,
    \textsuperscript{12}\thinspace{}they can hear each other's dogs and roosters,
    \textsuperscript{13}\thinspace{}yet the peo\-ple grow old and die
    \textsuperscript{14}\thinspace{}with\-out any deal\-ings be\-tween each oth\-er.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{81:1}\thinspace{}Cred\-i\-ble words aren't el\-o\-quent;
    \textsuperscript{2}\thinspace{}el\-o\-quent words aren't cred\-i\-ble.
    \textsuperscript{3}\thinspace{}The skill\-ful aren't polymaths;
    \textsuperscript{4}\thinspace{}those who are polymaths aren't skill\-ful.
    \textsuperscript{5}\thinspace{}Those who know aren't learn\-ed;
    \textsuperscript{6}\thinspace{}the learn\-ed don't know.
\end{para}

\begin{para}
    \textsuperscript{81:7}\thinspace{}The sage accumulates noth\-ing.
    \textsuperscript{8}\thinspace{}The more he does for others, the greater his possessions.
    \textsuperscript{9}\thinspace{}The more he gives to others, the greater his a\-bun\-dance.
\end{para}

\begin{para}
    \textsuperscript{81:10}\thinspace{}There\-fore, the Way of Heav\-en is to ben\-e\-fit with\-out harming.
    \textsuperscript{11}\thinspace{}The Way of the sage is to act with\-out contending.
\end{para}

\chapter*{Buddha}
\markboth{Buddha}{Buddha}
\addcontentsline{toc}{section}{Buddha}\begin{changemargin}{.6in}{.6in}
  \begin{center}
    \itshape
    A col\-lec\-tion of sayings of the Bud\-dha tak\-en from the Dham\-ma\-pa\-da.
  \end{center}
\end{changemargin}
\vspace{0pt plus 8\baselineskip{}}\section*{Pairs}
\markright{Pairs}
\begin{para}
    \textsuperscript{1}\thinspace{}Our life is shaped by our mind: it is ruled by our mind, it is made by our mind. If one speaks or acts with an evil mind, suf\-fer\-ing follows him, as the wheel follows the foot of the ox that draws the cart.
    \textsuperscript{2}\thinspace{}Our life is shaped by our mind: it is ruled by our mind, it is made by our mind. If one speaks or acts with a pure mind, hap\-pi\-ness follows him, like a shad\-ow that nev\-er leaves him.
\end{para}

\begin{para}
    \textsuperscript{3}\thinspace{}“He insulted me, hit me, defeated me, robbed me.” In those who har\-bor such thoughts, ha\-tred will nev\-er cease.
    \textsuperscript{4}\thinspace{}“He insulted me, hit me, defeated me, robbed me.” In those who do not har\-bor such thoughts, ha\-tred will cease.
    \textsuperscript{5}\thinspace{}For ha\-tred nev\-er ceases by ha\-tred. Ha\-tred ceases by love. This is an e\-ter\-nal law.
    \textsuperscript{6}\thinspace{}Some do not re\-al\-ize that we all must die, but those who do re\-al\-ize it cease their quarrels at once.
\end{para}

\begin{para}
    \textsuperscript{7}\thinspace{}Just as the wind throws down a weak tree, so will the temp\-ter o\-ver\-throw him who fo\-cus\-es on\-ly on pleasures, his sen\-ses un\-con\-trolled, im\-mod\-er\-ate in his food, i\-dle, and weak.
    \textsuperscript{8}\thinspace{}Just as the wind can\-not throw down a rock\-y moun\-tain, so the temp\-ter will not o\-ver\-throw him who does not fo\-cus on pleasures, his sen\-ses well con\-trolled, mod\-er\-ate in his food, faith\-ful and strong.
\end{para}

\begin{para}
    \textsuperscript{9}\thinspace{}He who puts on the monk's yel\-low robe with\-out hav\-ing cleansed him\-self from sin, who disregards self-control and truth, is un\-wor\-thy of the robe.
    \textsuperscript{10}\thinspace{}But he who has cleansed him\-self from sin, is well grounded in vir\-tues, and regards self-control and truth, is in\-deed wor\-thy of the robe.
\end{para}

\begin{para}
    \textsuperscript{11}\thinspace{}They who mis\-take the triv\-i\-al for the es\-sen\-tial, and the es\-sen\-tial for the triv\-i\-al, nev\-er ar\-rive at the es\-sen\-tial, but fol\-low vain desires.
    \textsuperscript{12}\thinspace{}They who know the es\-sen\-tial as es\-sen\-tial, and the triv\-i\-al as triv\-i\-al, ar\-rive at the es\-sen\-tial, and fol\-low true desires.
\end{para}

\begin{para}
    \textsuperscript{13}\thinspace{}Just as rain seeps through a poor\-ly tha\-tched house, pas\-sion seeps through an undeveloped mind.
    \textsuperscript{14}\thinspace{}Just as rain can\-not seep through a well tha\-tched house, pas\-sion can\-not seep through a well de\-vel\-oped mind.
\end{para}

\begin{para}
    \textsuperscript{15}\thinspace{}The evil\-do\-er grieves in this world, and in the next. He grieves in both. He grieves and suffers when he sees the evil of his own work.
    \textsuperscript{16}\thinspace{}The vir\-tu\-ous man delights in this world, and in the next. He delights in both. He delights and rejoices when he sees the pu\-ri\-ty of his own work.
\end{para}

\begin{para}
    \textsuperscript{17}\thinspace{}The evil\-do\-er suffers in this world, and in the next. He suffers in both. He suffers when he thinks of the evil he has done. He suffers more when go\-ing to the next world.
    \textsuperscript{18}\thinspace{}The vir\-tu\-ous man is hap\-py in this world, and in the next. He is hap\-py in both. He is hap\-py when he thinks of the good he has done. He is hap\-pi\-er still when go\-ing to the next world.
\end{para}

\begin{para}
    \textsuperscript{19}\thinspace{}The thought\-less man, if he recites man\-y scriptures, but does not do what they say, has no share in the blessings of the ho\-ly life, but is like a cow\-herd counting the cows of others.
    \textsuperscript{20}\thinspace{}He who recites on\-ly a few scriptures, but does what they say, hav\-ing for\-sak\-en lust and ha\-tred and de\-lu\-sion, possesses true knowl\-edge and se\-ren\-i\-ty of mind, car\-ing for noth\-ing in this world or that to come, he shares in the blessings of the ho\-ly life.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Vigilance}
\markright{Vigilance}
\begin{para}
    \textsuperscript{21}\thinspace{}Vig\-i\-lance is the path to sal\-va\-tion. Care\-less\-ness is the path to death. Those who are vig\-i\-lant do not die. Those who are care\-less are as if dead al\-read\-y.
    \textsuperscript{22}\thinspace{}The wise, hav\-ing un\-der\-stood this clear\-ly, de\-light in vig\-i\-lance, and re\-joice in the knowl\-edge of the no\-ble ones.
    \textsuperscript{23}\thinspace{}These wise peo\-ple, meditative, stead\-fast\-ly per\-se\-ver\-ing, at\-tain to sal\-va\-tion, the highest hap\-pi\-ness.
\end{para}

\begin{para}
    \textsuperscript{24}\thinspace{}If a vig\-i\-lant per\-son is en\-er\-get\-ic, if he is mind\-ful, if his deeds are pure, if he acts with con\-sid\-er\-a\-tion, if he has self-control, and is right\-eous, then his glo\-ry will in\-crease.
    \textsuperscript{25}\thinspace{}By ef\-fort, by vig\-i\-lance, by dis\-ci\-pline and self-control, the wise man may make for him\-self an is\-land which no flood can o\-ver\-whelm.
\end{para}

\begin{para}
    \textsuperscript{26}\thinspace{}The fool\-ish and ig\-no\-rant in\-dulge in care\-less\-ness. The wise man keeps vig\-i\-lance as his great\-est treas\-ure.
    \textsuperscript{27}\thinspace{}Do not give way to care\-less\-ness or to the en\-joy\-ment of sen\-su\-al pleasures. He who is vig\-i\-lant and meditative, obtains am\-ple joy.
\end{para}

\begin{para}
    \textsuperscript{28}\thinspace{}When the wise man drives a\-way care\-less\-ness by vig\-i\-lance, he climbs the tow\-er of wis\-dom and looks down up\-on the fools. With\-out suf\-fer\-ing, he looks at the suf\-fer\-ing crowd, as one that stands on a moun\-tain looks at them that stand on the ground be\-low.
\end{para}

\begin{para}
    \textsuperscript{29}\thinspace{}Vig\-i\-lant a\-mong the care\-less, a\-wake a\-mong the sleep\-ing, the wise man ad\-vanc\-es like a fast horse, leav\-ing the weak be\-hind.
    \textsuperscript{30}\thinspace{}Vig\-i\-lance is al\-ways praised; care\-less\-ness is al\-ways de\-spised.
\end{para}

\begin{para}
    \textsuperscript{31}\thinspace{}The monk who delights in vig\-i\-lance, who looks with fear at care\-less\-ness, moves a\-bout like fire, burn\-ing all fetters, small and large.
    \textsuperscript{32}\thinspace{}The monk who delights in vig\-i\-lance, who looks with fear at care\-less\-ness, can\-not fall a\-way. He is close to sal\-va\-tion.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Mind}
\markright{Mind}
\begin{para}
    \textsuperscript{33}\thinspace{}Just as a fletch\-er straightens his ar\-row, a wise man straightens his trem\-bling and un\-stead\-y mind, which is dif\-fi\-cult to guard, dif\-fi\-cult to re\-strain.
\end{para}

\begin{para}
    \textsuperscript{34}\thinspace{}Just as a fish quivers when tak\-en from its home in the wa\-ter and thrown on land, our mind trem\-bles all o\-ver to es\-cape the in\-flu\-ence of the temp\-ter.
\end{para}

\begin{para}
    \textsuperscript{35}\thinspace{}It is good to tame the mind, which is dif\-fi\-cult to tame and flight\-y, rush\-ing wher\-ev\-er it wants. A tamed mind brings hap\-pi\-ness.
    \textsuperscript{36}\thinspace{}Let the wise man guard his mind, for it is sub\-tle and e\-lu\-sive, and it rushes wher\-ev\-er it wants. A guard\-ed mind brings hap\-pi\-ness.
\end{para}

\begin{para}
    \textsuperscript{37}\thinspace{}The mind wanders far, moves a\-bout a\-lone, is with\-out a bod\-y, and hides in the ca\-ve of the heart. Those who bri\-dle it are free from the bonds of the temp\-ter.
\end{para}

\begin{para}
    \textsuperscript{38}\thinspace{}If a man's mind is un\-stead\-y, if he does not know the true law, and if his faith is troubled, then his knowl\-edge will nev\-er be per\-fect.
    \textsuperscript{39}\thinspace{}If a man's mind is free from pas\-sion, if his mind is free from ill will, if he is un\-af\-fect\-ed by good and evil, then there is dan\-ger for him.
\end{para}

\begin{para}
    \textsuperscript{40}\thinspace{}Know\-ing that this bod\-y is frag\-ile like a clay pot, and securing this mind like a for\-tress, at\-tack the temp\-ter with the sword of knowl\-edge. Then guard what is won with\-out feel\-ing at\-tach\-ment to it.
    \textsuperscript{41}\thinspace{}Soon, this bod\-y will lie on the ground, discarded, with\-out con\-scious\-ness, like a use\-less log.
\end{para}

\begin{para}
    \textsuperscript{42}\thinspace{}What\-ev\-er an en\-e\-my might do to an en\-e\-my, or a foe to a foe, a wrong\-ly di\-rect\-ed mind can do to you e\-ven worse.
    \textsuperscript{43}\thinspace{}What\-ev\-er a moth\-er, a fa\-ther, or any oth\-er rel\-a\-tive might do for you, a well di\-rect\-ed mind can do for you e\-ven more.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Flowers}
\markright{Flowers}
\begin{para}
    \textsuperscript{44}\thinspace{}Who will o\-ver\-come this earth, and the world of the dead, and the world of the gods? Who will choose the well taught path of vir\-tue, as an ex\-pert flo\-rist chooses the right flow\-er?
    \textsuperscript{45}\thinspace{}A dis\-ci\-ple will o\-ver\-come this earth, and the world of the dead, and the world of the gods. A dis\-ci\-ple will choose the well taught path of vir\-tue, as an ex\-pert flo\-rist chooses the right flow\-er.
\end{para}

\begin{para}
    \textsuperscript{46}\thinspace{}Know\-ing that this bod\-y is like froth, and learn\-ing that it is in\-sub\-stan\-tial like a mi\-rage, you will re\-move the flower-tipped ar\-rows of the temp\-ter, and nev\-er see the king of death.
\end{para}

\begin{para}
    \textsuperscript{47}\thinspace{}As a flood sweeps a\-way a sleep\-ing vil\-lage, so death sweeps a\-way a man who is im\-mersed in gath\-er\-ing flowers, his mind dis\-tract\-ed.
    \textsuperscript{48}\thinspace{}Death subdues a man who is im\-mersed in gath\-er\-ing flowers, his mind dis\-tract\-ed, be\-fore he is sa\-ti\-at\-ed in his pleasures.
    \textsuperscript{49}\thinspace{}As a bee collects nec\-tar and departs with\-out dam\-ag\-ing the flow\-er or its col\-or or its scent, so should a sage col\-lect alms in a vil\-lage.
\end{para}

\begin{para}
    \textsuperscript{50}\thinspace{}Do not no\-tice the faults of others, their sins of com\-mis\-sion or o\-mis\-sion, but no\-tice your own mis\-con\-duct and neg\-li\-gence.
\end{para}

\begin{para}
    \textsuperscript{51}\thinspace{}Like a beau\-ti\-ful flow\-er, full of col\-or but with\-out scent, are fair but fruit\-less words of him who does not prac\-tice them.
    \textsuperscript{52}\thinspace{}But, like a beau\-ti\-ful flow\-er, full of col\-or and full of scent, are fair and fruit\-ful words of him who practices them.
\end{para}

\begin{para}
    \textsuperscript{53}\thinspace{}Just as man\-y wreaths can be made from a heap of flowers, so can man\-y good things be done by a mor\-tal.
\end{para}

\begin{para}
    \textsuperscript{54}\thinspace{}The scent of flowers does not trav\-el a\-gainst the wind, not e\-ven that of san\-dal\-wood, tagara or jas\-mine; but the rep\-u\-ta\-tion of the vir\-tu\-ous does trav\-el a\-gainst the wind. The rep\-u\-ta\-tion of the vir\-tu\-ous pervades eve\-ry place.
    \textsuperscript{55}\thinspace{}A\-mong the scent of san\-dal\-wood, tagara, lo\-tus, and jas\-mine, the scent of vir\-tue is un\-sur\-passed.
\end{para}

\begin{para}
    \textsuperscript{56}\thinspace{}The scent that co\-mes from tagara and san\-dal\-wood is faint. The scent of the vir\-tu\-ous is best, it rises up to the gods.
    \textsuperscript{57}\thinspace{}The temp\-ter nev\-er finds the path of the vir\-tu\-ous, who lives with vig\-i\-lance, and is freed through per\-fect knowl\-edge.
\end{para}

\begin{para}
    \textsuperscript{58}\thinspace{}On a gar\-bage heap by the roadside, a lo\-tus grows full of sweet per\-fume and de\-light.
    \textsuperscript{59}\thinspace{}Like\-wise, on a gar\-bage heap of blind mortals, a true dis\-ci\-ple of the Bud\-dha shines forth in wis\-dom.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{The Fool}
\markright{The Fool}
\begin{para}
    \textsuperscript{60}\thinspace{}The night is long to the sleep\-less; a mile is long to the tired; life is long to the fool\-ish who do not know the true law.
\end{para}

\begin{para}
    \textsuperscript{61}\thinspace{}If a trav\-el\-er does not meet a com\-pan\-ion who is his bet\-ter, or his e\-qual, let him firm\-ly keep to his sol\-i\-tar\-y jour\-ney. There is no com\-pan\-ion\-ship with a fool.
    \textsuperscript{62}\thinspace{}He worries, “I have sons, and I have wealth,” but he him\-self is not his own, how much less sons and wealth?
\end{para}

\begin{para}
    \textsuperscript{63}\thinspace{}A fool who knows his own fool\-ish\-ness, is wise to that ex\-tent. But a fool who thinks he is wise, is a fool in\-deed.
    \textsuperscript{64}\thinspace{}E\-ven if a fool associates with a wise man all his life, he will per\-ceive the truth no more than a spoon perceives the fla\-vor of the soup.
    \textsuperscript{65}\thinspace{}If an in\-tel\-li\-gent man associates with a wise man for on\-ly one min\-ute, he will soon per\-ceive the truth as the tongue perceives the taste of soup.
\end{para}

\begin{para}
    \textsuperscript{66}\thinspace{}Fools of lit\-tle un\-der\-stand\-ing are their own enemies, for they do evil deeds that bear bit\-ter fruit.
    \textsuperscript{67}\thinspace{}That deed is poor\-ly done if a man must re\-pent af\-ter\-ward, and its re\-ward is cry\-ing and a tear\-ful face.
    \textsuperscript{68}\thinspace{}That deed is well done if a man does not re\-pent af\-ter\-ward, and its re\-ward is de\-light and hap\-pi\-ness.
\end{para}

\begin{para}
    \textsuperscript{69}\thinspace{}As long as an evil deed does not bear fruit, the fool thinks it is like hon\-ey. But when it ripens, then the fool suffers grief.
    \textsuperscript{70}\thinspace{}E\-ven if a fool, month af\-ter month, eats his food like an as\-cet\-ic with the tip of a blade of grass, he is not worth e\-ven one six\-teenth of those who have un\-der\-stood the truth.
\end{para}

\begin{para}
    \textsuperscript{71}\thinspace{}An evil deed, like fresh\-ly drawn milk, does not sour im\-me\-di\-ate\-ly; it follows the fool, smoldering, like fire cov\-ered by ash\-es.
\end{para}

\begin{para}
    \textsuperscript{73}\thinspace{}The fool wishes for a false rep\-u\-ta\-tion, for prec\-e\-dence a\-mong monks, for au\-thor\-i\-ty in the monasteries, for hon\-or a\-mong oth\-er peo\-ple.
    \textsuperscript{74}\thinspace{}“May both laymen and monks think that I did it. Let them o\-bey me in eve\-ry\-thing, great and small.” Such is the mind of the fool, and thus his de\-sire and pride in\-crease.
\end{para}

\begin{para}
    \textsuperscript{75}\thinspace{}One road leads to wealth, an\-oth\-er road leads to sal\-va\-tion. If the monk, a dis\-ci\-ple of Bud\-dha, has learn\-ed this, he will not yearn for hon\-or, he will strive for sep\-a\-ra\-tion from the world.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{The Wise}
\markright{The Wise}
\begin{para}
    \textsuperscript{76}\thinspace{}If you find a wise man who points out faults, and who reproves, fol\-low that wise man as one would fol\-low a guide to hid\-den treas\-ure. It will be bet\-ter, not worse, for those who fol\-low him.
    \textsuperscript{77}\thinspace{}Let him ad\-mon\-ish, let him teach, let him re\-strain you from what is im\-prop\-er. The good will love him. The bad will hate him.
\end{para}

\begin{para}
    \textsuperscript{78}\thinspace{}Do not have evildoers as companions, do not have vile peo\-ple as friends. Have vir\-tu\-ous peo\-ple as companions, have the best as friends.
\end{para}

\begin{para}
    \textsuperscript{79}\thinspace{}He who drinks in the law lives hap\-pi\-ly with a se\-rene mind. The wise man rejoices al\-ways in the law, as preached by the no\-ble ones.
\end{para}

\begin{para}
    \textsuperscript{80}\thinspace{}Irrigators guide the wa\-ter. Fletchers straight\-en the ar\-row. Carpenters shape the wood. The wise con\-trol them\-selves.
\end{para}

\begin{para}
    \textsuperscript{81}\thinspace{}Just as a sol\-id rock is not shak\-en by the wind, the wise do not fal\-ter a\-midst blame or praise.
    \textsuperscript{82}\thinspace{}Af\-ter they have listened to the law, the wise be\-come se\-rene, like a deep, smooth, and still lake.
\end{para}

\begin{para}
    \textsuperscript{83}\thinspace{}The good keep walk\-ing what\-ev\-er happens. The good do not prat\-tle, long\-ing for pleas\-ure. Wheth\-er touched by hap\-pi\-ness or sor\-row, the wise nev\-er ap\-pear e\-lat\-ed or de\-pressed.
    \textsuperscript{84}\thinspace{}A man is good, wise, and vir\-tu\-ous if he does no wrong for his own sa\-ke or for the sa\-ke of others, if he wishes nei\-ther for wealth, a son, or a king\-dom, and if he does not wish for suc\-cess by un\-fair means.
\end{para}

\begin{para}
    \textsuperscript{85}\thinspace{}Few a\-mong men ar\-rive at the oth\-er shore (sal\-va\-tion). The rest run up and down on this shore.
    \textsuperscript{86}\thinspace{}But those who fol\-low the law, when it has been well taught to them, will ar\-rive at the oth\-er shore hav\-ing crossed the do\-min\-ion of death, how\-ev\-er dif\-fi\-cult to o\-ver\-come.
\end{para}

\begin{para}
    \textsuperscript{87}\thinspace{}Let a wise man leave the dark way, and fol\-low the bright path. Af\-ter go\-ing from his home to home\-less\-ness, let him yearn for de\-light in de\-tach\-ment, so dif\-fi\-cult to en\-joy.
    \textsuperscript{88}\thinspace{}Leav\-ing all pleasures be\-hind, and call\-ing noth\-ing his own, let the wise man cleanse him\-self of that which defiles the mind.
    \textsuperscript{89}\thinspace{}Those whose minds are well grounded in the factors of en\-light\-en\-ment, who with\-out cling\-ing to any\-thing, re\-joice in free\-dom from at\-tach\-ment, whose appetites have been conquered, and who are full of light, are free e\-ven in this world.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{The Saint}
\markright{The Saint}
\begin{para}
    \textsuperscript{90}\thinspace{}There is no suf\-fer\-ing for him who has fin\-ished his jour\-ney, and is with\-out sor\-row, and is free on all sides, and has thrown off all fetters.
\end{para}

\begin{para}
    \textsuperscript{91}\thinspace{}The mind\-ful strive dil\-i\-gent\-ly. They are not at\-tached to any home. Like swans who leave a lake, they leave their house and home.
\end{para}

\begin{para}
    \textsuperscript{92}\thinspace{}Like birds in the air, the path of the saint is hard to trace. They have no rich\-es, are wise re\-gard\-ing food, and their ob\-jec\-tive is emp\-ti\-ness and free\-dom.
    \textsuperscript{93}\thinspace{}Like birds in the air, his path is hard to trace. He whose appetites are stilled, who is not at\-tached to food, whose ob\-jec\-tive is emp\-ti\-ness and free\-dom.
\end{para}

\begin{para}
    \textsuperscript{94}\thinspace{}E\-ven the gods a\-dore him whose sen\-ses have been sub\-dued, like hors\-es well trained by a char\-i\-ot\-eer, who is free from pride, and free from appetites.
    \textsuperscript{95}\thinspace{}Like the earth, he is not pro\-voked. He is firm like a pil\-lar. He is like a pure lake with\-out mud. No new births are in store for him.
\end{para}

\begin{para}
    \textsuperscript{96}\thinspace{}His mind is calm, and calm are his words and deeds. By true knowl\-edge, he has obtained free\-dom, and thus be\-come tran\-quil.
\end{para}

\begin{para}
    \textsuperscript{98}\thinspace{}In a vil\-lage or in a for\-est, in a val\-ley or on a hill, wher\-ev\-er saints dwell, that place is de\-light\-ful.
    \textsuperscript{99}\thinspace{}De\-light\-ful are the forests in which the crowds find no pleas\-ure. Those with\-out pas\-sion will find de\-light there, for they do not seek pleas\-ure.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Thousands}
\markright{Thousands}
\begin{para}
    \textsuperscript{100}\thinspace{}Bet\-ter than a speech of a thou\-sand vain words, is one thought\-ful word, hear\-ing which brings peace.
    \textsuperscript{101}\thinspace{}Bet\-ter than a po\-em of a thou\-sand vain verses, is one thought\-ful verse, hear\-ing which brings peace.
    \textsuperscript{102}\thinspace{}Bet\-ter than re\-cit\-ing a hun\-dred vain verses, is re\-cit\-ing one verse of law, hear\-ing which brings peace.
\end{para}

\begin{para}
    \textsuperscript{103}\thinspace{}Greater than he who conquers a thou\-sand times a thou\-sand men in bat\-tle, is he who conquerors just one --- him\-self.
    \textsuperscript{104}\thinspace{}It is far bet\-ter to con\-quer one\-self than to con\-quer others.
    \textsuperscript{105}\thinspace{}For a man who has conquered him\-self and al\-ways lives with self-control, not e\-ven a god, an an\-gel, or a de\-mon with a god can change that vic\-to\-ry in\-to de\-feat.
\end{para}

\begin{para}
    \textsuperscript{106}\thinspace{}A man could sac\-ri\-fice by the thousands for month af\-ter month for a hun\-dred years, or he could pay hom\-age for one mo\-ment to a man liv\-ing in wis\-dom. Bet\-ter is that hom\-age than sacrifices for a cen\-tu\-ry.
    \textsuperscript{107}\thinspace{}A man could tend the sac\-ri\-fi\-cial fire in the for\-est for a hun\-dred years, or he could pay hom\-age for one mo\-ment to a man liv\-ing in wis\-dom. Bet\-ter is that hom\-age than sacrifices for a cen\-tu\-ry.
    \textsuperscript{108}\thinspace{}In this world, a man may make sacrifices and oblations for a whole year in or\-der to gain mer\-it, yet all of it is not worth a quar\-ter of the mer\-it gained by re\-ver\-ing the right\-eous.
\end{para}

\begin{para}
    \textsuperscript{109}\thinspace{}For him who al\-ways respects and hon\-ors the wise, four things will in\-crease: life, beau\-ty, hap\-pi\-ness, pow\-er.
\end{para}

\begin{para}
    \textsuperscript{110}\thinspace{}Bet\-ter than a hun\-dred years lived im\-mor\-al and un\-re\-strained, is one day lived vir\-tu\-ous and reflecting.
    \textsuperscript{111}\thinspace{}Bet\-ter than a hun\-dred years lived ig\-no\-rant and un\-re\-strained, is one day lived wise and reflecting.
    \textsuperscript{112}\thinspace{}Bet\-ter than a hun\-dred years i\-dle and weak, is one day lived en\-er\-get\-ic and firm.
    \textsuperscript{113}\thinspace{}Bet\-ter than a hun\-dred years lived with\-out see\-ing be\-gin\-ning or end, is one day lived see\-ing be\-gin\-ning and end.
    \textsuperscript{114}\thinspace{}Bet\-ter than a hun\-dred years lived with\-out see\-ing the death\-less state, is one day lived see\-ing the death\-less state.
    \textsuperscript{115}\thinspace{}Bet\-ter than a hun\-dred years lived with\-out see\-ing the highest law, is one day lived see\-ing the highest law.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Evil}
\markright{Evil}
\begin{para}
    \textsuperscript{116}\thinspace{}Has\-ten to do good. Keep your mind from evil. If a man is slow to do good, his mind delights in evil.
\end{para}

\begin{para}
    \textsuperscript{117}\thinspace{}If a man does what is evil, let him not do it a\-gain and a\-gain. Let him not de\-light in it, for ac\-cu\-mu\-lat\-ing evil brings pain.
    \textsuperscript{118}\thinspace{}If a man does what is good, let him do it a\-gain and a\-gain. Let him de\-light in it, for ac\-cu\-mu\-lat\-ing good brings joy.
\end{para}

\begin{para}
    \textsuperscript{119}\thinspace{}E\-ven an evil\-do\-er may be hap\-py be\-fore his evil deed ripens, but when it ripens, he suffers.
    \textsuperscript{120}\thinspace{}E\-ven a good man may suf\-fer be\-fore his good deed ripens, but when it ripens, he is hap\-py.
\end{para}

\begin{para}
    \textsuperscript{121}\thinspace{}Let no man think light\-ly of evil, say\-ing in his heart, “It will not a\-mount to much.” As a wa\-ter pot is filled by the fall\-ing of wa\-ter drops, so the fool becomes full of evil, e\-ven if he gathers it lit\-tle by lit\-tle.
    \textsuperscript{122}\thinspace{}Let no man think light\-ly of good, say\-ing in his heart, “It will not a\-mount to much.” As a wa\-ter pot is filled by the fall\-ing of wa\-ter drops, so the wise man becomes full of good, e\-ven if he gathers it lit\-tle by lit\-tle.
\end{para}

\begin{para}
    \textsuperscript{123}\thinspace{}Let a man a\-void evil deeds, as a wealth\-y mer\-chant with few companions avoids a dan\-ger\-ous road; as a man who loves life avoids poi\-son.
\end{para}

\begin{para}
    \textsuperscript{125}\thinspace{}If a man offends a harm\-less, pure, and in\-no\-cent per\-son, the evil falls back up\-on that fool, like dust thrown a\-gainst the wind.
\end{para}

\begin{para}
    \textsuperscript{126}\thinspace{}Some are born in the womb. Evildoers go to hell. The right\-eous go to heav\-en. Those with\-out any world\-ly desires at\-tain sal\-va\-tion.
\end{para}

\begin{para}
    \textsuperscript{127}\thinspace{}Not in the sky, not in the mid\-dle of the sea, not in a cleft of the mountains, is there a spot to be found where a man can es\-cape from the con\-se\-quenc\-es of his evil deed.
    \textsuperscript{128}\thinspace{}Not in the sky, not in the mid\-dle of the sea, not in a cleft of the mountains, is there a spot to be found where a man will not be o\-ver\-come by death.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Punishment}
\markright{Punishment}
\begin{para}
    \textsuperscript{129}\thinspace{}Eve\-ry\-one trem\-bles at pun\-ish\-ment, eve\-ry\-one fears death. Put\-ting your\-self in his place, do not kill or cause others to kill.
    \textsuperscript{130}\thinspace{}Eve\-ry\-one trem\-bles at pun\-ish\-ment, eve\-ry\-one loves his own life. Put\-ting your\-self in his place, do not kill or cause others to kill.
\end{para}

\begin{para}
    \textsuperscript{131}\thinspace{}He who, seek\-ing hap\-pi\-ness, punishes beings who al\-so de\-sire hap\-pi\-ness, will not find hap\-pi\-ness af\-ter death.
    \textsuperscript{132}\thinspace{}He who, seek\-ing hap\-pi\-ness, does not pun\-ish beings who al\-so de\-sire hap\-pi\-ness, will find hap\-pi\-ness af\-ter death.
\end{para}

\begin{para}
    \textsuperscript{133}\thinspace{}Speak harsh\-ly to no one, or those who are spo\-ken to will an\-swer you in the same way. Con\-ten\-tious speech hurts, and re\-tal\-i\-a\-tion will come to you.
    \textsuperscript{134}\thinspace{}If your mind is still, like a bro\-ken gong, you have attained sal\-va\-tion, for there's no con\-ten\-tion in you.
\end{para}

\begin{para}
    \textsuperscript{135}\thinspace{}As a cow\-herd drives cows to pas\-ture with a staff, so age\-ing and death drive the life of men.
    \textsuperscript{136}\thinspace{}A fool does not re\-al\-ize when he commits evil, but he suffers for those deeds, as if burned by fire.
\end{para}

\begin{para}
    \textsuperscript{137}\thinspace{}He who inflicts pain on the in\-no\-cent and un\-armed, will soon suf\-fer in one of these ten ways:
    \textsuperscript{138}\thinspace{}se\-vere pain, im\-pov\-er\-ish\-ment, bod\-i\-ly in\-ju\-ry, se\-ri\-ous ill\-ness, men\-tal ill\-ness,
    \textsuperscript{139}\thinspace{}trou\-ble from the gov\-ern\-ment, a fear\-ful ac\-cu\-sa\-tion, loss of relatives, loss of wealth,
    \textsuperscript{140}\thinspace{}or fire will de\-stroy his hous\-es. When his bod\-y is destroyed, the fool will go to hell.
\end{para}

\begin{para}
    \textsuperscript{141}\thinspace{}Nei\-ther na\-ked\-ness, nor mat\-ted hair, nor dirt, nor fasting, nor ly\-ing on the earth, nor rub\-bing with dust, nor sit\-ting mo\-tion\-less, can pu\-ri\-fy a mor\-tal who has not o\-ver\-come doubt.
    \textsuperscript{142}\thinspace{}He who exercises tran\-quil\-i\-ty, is qui\-et, sub\-dued, restrained, chaste, and has set a\-side vi\-o\-lence to\-wards all others, he in\-deed is a ho\-ly man, an as\-cet\-ic, a monk, e\-ven if dressed in fi\-ne ap\-par\-el.
\end{para}

\begin{para}
    \textsuperscript{143}\thinspace{}Who in this world is so restrained by con\-science that he needs no re\-proof, as a fi\-ne horse needs no whip?
    \textsuperscript{144}\thinspace{}Like a fi\-ne horse when touched by the whip, be ac\-tive and live\-ly. By faith, by vir\-tue, by ef\-fort, by med\-i\-ta\-tion, by dis\-cern\-ing the truth, by per\-fect knowl\-edge and con\-duct, by mind\-ful\-ness, you will leave this pain be\-hind.
\end{para}

\begin{para}
    \textsuperscript{145}\thinspace{}Irrigators guide the wa\-ter. Fletchers straight\-en the ar\-row. Carpenters shape the wood. Good peo\-ple fash\-ion them\-selves.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Old Age}
\markright{Old Age}
\begin{para}
    \textsuperscript{146}\thinspace{}Why is there laugh\-ter and joy when this world is al\-ways burn\-ing? Surrounded by dark\-ness, don't you look for a light?
\end{para}

\begin{para}
    \textsuperscript{147}\thinspace{}Look at this bod\-y --- a paint\-ed im\-age, cov\-ered with festering wounds, sick\-ly, full of man\-y thoughts --- nei\-ther last\-ing nor sta\-ble.
    \textsuperscript{148}\thinspace{}This bod\-y is worn out, full of sick\-ness, and frail. This heap of cor\-rup\-tion breaks to pieces, for life in\-deed ends with death.
    \textsuperscript{149}\thinspace{}These white bones are like gourds thrown a\-way in the au\-tumn. What pleas\-ure is there in see\-ing them?
\end{para}

\begin{para}
    \textsuperscript{150}\thinspace{}A cit\-y made of bones is cov\-ered with flesh and blood. Liv\-ing in it are old age and death, pride and de\-ceit.
    \textsuperscript{151}\thinspace{}The beau\-ti\-ful chariots of kings get worn out and so does the bod\-y. But the vir\-tue of good peo\-ple does not age: thus do the good say a\-mong them\-selves.
\end{para}

\begin{para}
    \textsuperscript{152}\thinspace{}A man who has learn\-ed lit\-tle grows old like an ox. His bod\-y grows, but his wis\-dom does not.
\end{para}

\begin{para}
    \textsuperscript{153}\thinspace{}I roamed through man\-y births, with\-out re\-ward and with\-out rest, looking for the build\-er of this house. Pain\-ful is birth a\-gain and a\-gain.
    \textsuperscript{154}\thinspace{}But now, build\-er of this house, I have seen you. You will not build this house a\-gain. All your rafters are bro\-ken, your ridge pole is dismantled. My mind has attained the ex\-tinc\-tion of all desires.
\end{para}

\begin{para}
    \textsuperscript{155}\thinspace{}Those who, in their youth, have not lived the ho\-ly life nor gained wealth, per\-ish like old herons in a lake with\-out fish.
    \textsuperscript{156}\thinspace{}Those who, in their youth, have not lived the ho\-ly life nor gained wealth, lie sighing o\-ver the past like ar\-rows misfired from a bow.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Self}
\markright{Self}
\begin{para}
    \textsuperscript{157}\thinspace{}If a man holds him\-self dear, let him watch him\-self care\-ful\-ly. Let a wise man keep vig\-il dur\-ing one of the three watches of the night.
\end{para}

\begin{para}
    \textsuperscript{158}\thinspace{}Let each man first di\-rect him\-self to what is prop\-er, then let him teach others. Thus a wise man will not be reproached.
\end{para}

\begin{para}
    \textsuperscript{159}\thinspace{}If a man makes him\-self as he teaches others to be, then, be\-ing him\-self well tamed, he may tame others. One's self is in\-deed dif\-fi\-cult to tame.
\end{para}

\begin{para}
    \textsuperscript{160}\thinspace{}A man is tru\-ly his own pro\-tec\-tor. Who else could be the pro\-tec\-tor? With self well con\-trolled, one finds a pro\-tec\-tor such as few can find.
\end{para}

\begin{para}
    \textsuperscript{161}\thinspace{}The evil a fool\-ish man does by him\-self, self-born, self-created, grinds him down, as a dia\-mond breaks a hard gem.
    \textsuperscript{162}\thinspace{}When o\-ver\-spread with great wick\-ed\-ness, like a tree cov\-ered by vines, a man brings him\-self down to where his en\-e\-my wishes.
    \textsuperscript{163}\thinspace{}Evil deeds, which harm our\-selves, are eas\-y to do. Good and ben\-e\-fi\-cial deeds are ver\-y dif\-fi\-cult to do.
\end{para}

\begin{para}
    \textsuperscript{164}\thinspace{}The fool who scorns the teachings of the wise, of the no\-ble, of the right\-eous, and follows false doc\-trine, is like the bam\-boo that dies af\-ter bear\-ing fruit.
\end{para}

\begin{para}
    \textsuperscript{165}\thinspace{}Evil is done by one\-self, by one\-self one is de\-filed. Evil is left un\-done by one\-self, by one\-self is one pu\-ri\-fied. Pu\-ri\-ty and im\-pu\-ri\-ty are one's own do\-ing. No one can pu\-ri\-fy an\-oth\-er.
\end{para}

\begin{para}
    \textsuperscript{166}\thinspace{}Let no man sac\-ri\-fice his own wel\-fare for that of an\-oth\-er, how\-ev\-er great. Let a man, af\-ter dis\-cern\-ing his own wel\-fare, be al\-ways at\-ten\-tive to it.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{The World}
\markright{The World}
\begin{para}
    \textsuperscript{167}\thinspace{}Do not fol\-low vul\-gar ways. Do not live with a\-ban\-don. Do not fol\-low false doc\-trine. Do not be a friend of the world.
\end{para}

\begin{para}
    \textsuperscript{168}\thinspace{}Wake up! Do not be i\-dle. Live a vir\-tu\-ous life. The vir\-tu\-ous sleeps with ease in this world and the next.
    \textsuperscript{169}\thinspace{}Live a vir\-tu\-ous life. Do not live a life of sin. The vir\-tu\-ous sleeps with ease in this world and the next.
\end{para}

\begin{para}
    \textsuperscript{170}\thinspace{}The king of death does not see him who looks up\-on the world as a bub\-ble, as a mi\-rage.
    \textsuperscript{171}\thinspace{}Come look at this world, dec\-o\-rat\-ed like a roy\-al char\-i\-ot. Fools are im\-mersed in it, but the wise do not cling to it.
\end{para}

\begin{para}
    \textsuperscript{172}\thinspace{}He who was reck\-less and then becomes mind\-ful, brightens up this world like the moon when freed from clouds.
    \textsuperscript{173}\thinspace{}He who replaces evil deeds with good deeds, brightens up this world, like the moon when freed from clouds.
\end{para}

\begin{para}
    \textsuperscript{174}\thinspace{}In this dark world, on\-ly a few can see. On\-ly a few, like birds es\-cap\-ing from the net, go to heav\-en.
    \textsuperscript{175}\thinspace{}Swans fly on the path of the sun, through the air by their won\-der\-ful pow\-er. The wise flee from the world af\-ter defeating the temp\-ter and his armies.
\end{para}

\begin{para}
    \textsuperscript{176}\thinspace{}A man who transgresses the one law, speaks lies, and scoffs at the world be\-yond, is ca\-pa\-ble of any evil.
\end{para}

\begin{para}
    \textsuperscript{177}\thinspace{}Misers do not go to the world of the gods; nei\-ther do fools praise gen\-er\-os\-i\-ty. A wise man rejoices in gen\-er\-os\-i\-ty, and through it becomes bless\-ed in the world be\-yond.
\end{para}

\begin{para}
    \textsuperscript{178}\thinspace{}The first step to\-wards ho\-li\-ness is bet\-ter than sov\-er\-eign\-ty o\-ver the earth, bet\-ter than go\-ing to heav\-en, and bet\-ter than be\-ing lord o\-ver all the worlds.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{The Awakened}
\markright{The Awakened}
\begin{para}
    \textsuperscript{181}\thinspace{}E\-ven the gods en\-vy those who are awakened, who are de\-vot\-ed to med\-i\-ta\-tion, who are wise, and who de\-light in the calm of re\-nun\-ci\-a\-tion.
\end{para}

\begin{para}
    \textsuperscript{182}\thinspace{}It is dif\-fi\-cult to ob\-tain a hu\-man birth, dif\-fi\-cult is the life of mortals. It is dif\-fi\-cult to ob\-tain the chance to hear the true law, dif\-fi\-cult is the birth of the awakened.
\end{para}

\begin{para}
    \textsuperscript{183}\thinspace{}To a\-void all evil, to cul\-ti\-vate good, and to pu\-ri\-fy one's mind, that is the teach\-ing of the awakened.
    \textsuperscript{184}\thinspace{}Pa\-tient en\-dur\-ance is the highest aus\-ter\-i\-ty. “Sal\-va\-tion is su\-preme”, say the awakened. He is no monk who injures others. He is no as\-cet\-ic who insults others.
    \textsuperscript{185}\thinspace{}Not to re\-vile, not to in\-jure, to live restrained un\-der the law, to be mod\-er\-ate in food, to live in sol\-i\-tude, and to dwell on the highest thoughts: this is the teach\-ing of the awakened.
\end{para}

\begin{para}
    \textsuperscript{186}\thinspace{}There is no way to sat\-is\-fy lusts, e\-ven with a show\-er of gold coins. He is wise who knows that lusts give lit\-tle sat\-is\-fac\-tion and much pain.
    \textsuperscript{187}\thinspace{}E\-ven in heav\-en\-ly pleasures the wise finds no de\-light. The dis\-ci\-ple who is ful\-ly awakened delights on\-ly in the de\-struc\-tion of crav\-ing.
\end{para}

\begin{para}
    \textsuperscript{188}\thinspace{}When threatened with dan\-ger, men go to man\-y a ref\-uge: to mountains and forests, to groves and shrines.
    \textsuperscript{189}\thinspace{}But that is not a se\-cure ref\-uge, that is not the best ref\-uge. A man is not delivered from all pain af\-ter go\-ing to that ref\-uge.
\end{para}

\begin{para}
    \textsuperscript{190}\thinspace{}He who takes ref\-uge with the law, the awakened, and their or\-der, sees with clear un\-der\-stand\-ing the four ho\-ly truths:
    \textsuperscript{191}\thinspace{}pain, the cause of pain, the de\-struc\-tion of pain, and the ho\-ly eight\-fold path that leads to the quieting of pain;
    \textsuperscript{192}\thinspace{}That is a se\-cure ref\-uge, that is the best ref\-uge. Af\-ter go\-ing to that ref\-uge, a man is delivered from all pain.
\end{para}

\begin{para}
    \textsuperscript{193}\thinspace{}An awakened man is hard to find. He is not born eve\-ry\-where. Wher\-ev\-er such a wise man is born, that clan prospers.
\end{para}

\begin{para}
    \textsuperscript{194}\thinspace{}Bless\-ed is the arising of awakened ones, bless\-ed is the teach\-ing of the true law, bless\-ed is har\-mo\-ny in the or\-der, bless\-ed is the de\-vo\-tion of those who are at peace.
\end{para}

\begin{para}
    \textsuperscript{195}\thinspace{}He who reveres those who de\-serve rev\-er\-ence --- the awakened ones or their disciples, who have o\-ver\-come all obstacles, sor\-row, and lam\-en\-ta\-tion,
    \textsuperscript{196}\thinspace{}who found de\-liv\-er\-ance and know no fear --- his mer\-it can nev\-er be meas\-ured.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Happiness}
\markright{Happiness}
\begin{para}
    \textsuperscript{197}\thinspace{}We live hap\-pi\-ly, not hat\-ing those who hate us. When a\-mong men who hate us, we dwell with\-out ha\-tred.
    \textsuperscript{198}\thinspace{}We live hap\-pi\-ly, with\-out ill\-ness a\-mong the ill. When a\-mong men who are ill, we dwell with\-out ill\-ness.
    \textsuperscript{199}\thinspace{}We live hap\-pi\-ly, with\-out greed a\-mong the greed\-y. When a\-mong men who are greed\-y, we dwell with\-out greed.
    \textsuperscript{200}\thinspace{}We live hap\-pi\-ly, we who own noth\-ing.
\end{para}

\begin{para}
    \textsuperscript{201}\thinspace{}Vic\-to\-ry breeds ha\-tred, for the conquered live in pain. The con\-tent\-ed is hap\-py, hav\-ing giv\-en up both vic\-to\-ry and de\-feat.
\end{para}

\begin{para}
    \textsuperscript{202}\thinspace{}There is no fire like lust, no loss like ha\-tred, no pain like this ex\-ist\-ence, and no hap\-pi\-ness higher than peace.
\end{para}

\begin{para}
    \textsuperscript{203}\thinspace{}Hun\-ger is the worst dis\-ease, this ex\-ist\-ence the great\-est pain. If one tru\-ly knows this, he realizes sal\-va\-tion, the highest hap\-pi\-ness.
    \textsuperscript{204}\thinspace{}Health is the great\-est gift, con\-tent\-ment the great\-est wealth. Trust is the best kins\-man, sal\-va\-tion the highest hap\-pi\-ness.
\end{para}

\begin{para}
    \textsuperscript{205}\thinspace{}He who has savored the taste of sol\-i\-tude and tran\-quil\-i\-ty is free from fear and evil, refreshed with the nour\-ish\-ment of the law.
\end{para}

\begin{para}
    \textsuperscript{206}\thinspace{}The sight of no\-ble ones is good. To live with them is al\-ways hap\-pi\-ness. If a man does not see fools, he will be tru\-ly hap\-py.
\end{para}

\begin{para}
    \textsuperscript{207}\thinspace{}He who walks with fools suffers for a long time. Com\-pa\-ny with fools, as with an en\-e\-my, is al\-ways pain\-ful. Com\-pa\-ny with the wise is a pleas\-ure, like re\-un\-ion with fam\-i\-ly.
    \textsuperscript{208}\thinspace{}There\-fore, fol\-low the wise, the in\-tel\-li\-gent, the learn\-ed, the stead\-fast, the du\-ti\-ful, the no\-ble. Fol\-low a good and dis\-cern\-ing man, as the moon follows the path of the stars.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Pleasure}
\markright{Pleasure}
\begin{para}
    \textsuperscript{209}\thinspace{}He who gives him\-self to van\-i\-ty, in\-stead of to med\-i\-ta\-tion, who forgets the goal and grasps at pleas\-ure, will come to en\-vy him who has exerted him\-self.
\end{para}

\begin{para}
    \textsuperscript{210}\thinspace{}Don't seek for what is pleas\-ant or un\-pleas\-ant, for it is pain not to see the pleas\-ant, and it is pain to see what is un\-pleas\-ant.
\end{para}

\begin{para}
    \textsuperscript{211}\thinspace{}There\-fore, let no man hold any\-thing dear, for loss of the dear is pain\-ful. Those who hold noth\-ing dear and hate noth\-ing have no fetters.
\end{para}

\begin{para}
    \textsuperscript{212}\thinspace{}From pleas\-ure co\-mes grief, from pleas\-ure co\-mes fear. He who is free from pleas\-ure knows nei\-ther grief nor fear.
\end{para}

\begin{para}
    \textsuperscript{213}\thinspace{}From af\-fec\-tion co\-mes grief, from af\-fec\-tion co\-mes fear. He who is free from af\-fec\-tion knows nei\-ther grief nor fear.
\end{para}

\begin{para}
    \textsuperscript{214}\thinspace{}From at\-tach\-ment co\-mes grief, from at\-tach\-ment co\-mes fear. He who is free from at\-tach\-ment knows nei\-ther grief nor fear.
\end{para}

\begin{para}
    \textsuperscript{215}\thinspace{}From lust co\-mes grief, from lust co\-mes fear. He who is free from lust knows nei\-ther grief nor fear.
\end{para}

\begin{para}
    \textsuperscript{216}\thinspace{}From crav\-ing co\-mes grief, from crav\-ing co\-mes fear. He who is free from crav\-ing knows nei\-ther grief nor fear.
\end{para}

\begin{para}
    \textsuperscript{217}\thinspace{}He who has vir\-tue and in\-sight, who is just, who speaks the truth, and does his own busi\-ness, him the world holds dear.
\end{para}

\begin{para}
    \textsuperscript{218}\thinspace{}He who is in\-tent up\-on sal\-va\-tion, who is sat\-is\-fied in his mind, and who is not at\-tached to pleas\-ure, is called “car\-ried up\-stream”.
\end{para}

\begin{para}
    \textsuperscript{219}\thinspace{}Fam\-i\-ly, friends, and well-wishers wel\-come a man, long ab\-sent, who returns safe\-ly from a\-far.
    \textsuperscript{220}\thinspace{}As fam\-i\-ly receives a friend on his re\-turn, so do his good works wel\-come him who has done good, and has gone from this world to the next.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Anger}
\markright{Anger}
\begin{para}
    \textsuperscript{221}\thinspace{}Let a man leave an\-ger, for\-sake pride, and o\-ver\-come all bond\-age. No suf\-fer\-ing befalls the man who is de\-tached from name and form, and who calls noth\-ing his own.
\end{para}

\begin{para}
    \textsuperscript{222}\thinspace{}I call him a mas\-ter char\-i\-ot\-eer who holds back ris\-ing an\-ger like a rac\-ing char\-i\-ot. Others mere\-ly hold the reins.
\end{para}

\begin{para}
    \textsuperscript{223}\thinspace{}O\-ver\-come an\-ger with love, evil with good, greed with gen\-er\-os\-i\-ty, and lies with truth.
    \textsuperscript{224}\thinspace{}Speak the truth, do not yield to an\-ger, and give, if asked, e\-ven if it's on\-ly a lit\-tle. By these three steps you will go near the gods.
\end{para}

\begin{para}
    \textsuperscript{225}\thinspace{}The sages who harm no one, and al\-ways con\-trol their bod\-y, go to the un\-change\-a\-ble state, where they suf\-fer no more.
    \textsuperscript{226}\thinspace{}Those who are al\-ways vig\-i\-lant, stud\-y\-ing day and night, in\-tent on sal\-va\-tion, their passions fade a\-way.
\end{para}

\begin{para}
    \textsuperscript{227}\thinspace{}This is an old say\-ing, not just of to\-day: “They blame him who is si\-lent, they blame him who speaks much, they blame him who speaks lit\-tle.” There is none on earth who escapes blame.
    \textsuperscript{228}\thinspace{}There nev\-er was, there nev\-er will be, nor is there now, a man who is al\-ways blamed or al\-ways praised.
\end{para}

\begin{para}
    \textsuperscript{229}\thinspace{}If the knowl\-edge\-a\-ble praise him, hav\-ing ob\-served him day af\-ter day to be with\-out blem\-ish, wise, and rich in knowl\-edge and vir\-tue,
    \textsuperscript{230}\thinspace{}like an in\-got of gold, who can blame him? E\-ven the gods praise him. E\-ven the Cre\-a\-tor praises him.
\end{para}

\begin{para}
    \textsuperscript{231}\thinspace{}With your bod\-y: be\-ware of an\-ger, be restrained, leave its sins, and with it prac\-tise vir\-tue.
    \textsuperscript{232}\thinspace{}With your tongue: be\-ware of an\-ger, be restrained, leave its sins, and prac\-tise vir\-tue.
    \textsuperscript{233}\thinspace{}With your mind: be\-ware of an\-ger, be restrained, leave its sins, and prac\-tise vir\-tue.
    \textsuperscript{234}\thinspace{}The wise who con\-trol their bod\-y, their tongue, and their mind, are well con\-trolled in\-deed.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Impurity}
\markright{Impurity}
\begin{para}
    \textsuperscript{235}\thinspace{}You are now like a withered leaf, the messengers of death stand near\-by. You stand at the door of de\-par\-ture, yet have no pro\-vi\-sion for your jour\-ney.
    \textsuperscript{236}\thinspace{}Make an is\-land for your\-self, work hard, be wise. When your impurities are blown a\-way, and you are free from guilt, you will en\-ter in\-to the heav\-en\-ly world of the e\-lect.
\end{para}

\begin{para}
    \textsuperscript{237}\thinspace{}Your life has come to an end, you are near death, there is no rest\-ing place for you on the road, yet you have no pro\-vi\-sion for your jour\-ney.
    \textsuperscript{238}\thinspace{}Make an is\-land for your\-self, work hard, be wise. When your impurities are blown a\-way, and you are free from guilt, you will not en\-ter a\-gain in\-to birth and de\-cay.
\end{para}

\begin{para}
    \textsuperscript{239}\thinspace{}Let a wise man re\-move his own impurities, as a smith removes impurities from sil\-ver one by one, lit\-tle by lit\-tle, and mo\-ment by mo\-ment.
    \textsuperscript{240}\thinspace{}Just as rust destroys the i\-ron from which it springs, so do a transgressor's deeds lead him to an evil des\-ti\-na\-tion.
\end{para}

\begin{para}
    \textsuperscript{241}\thinspace{}The im\-pu\-ri\-ty of prayers is not re\-cit\-ing them. The im\-pu\-ri\-ty of hous\-es is not repairing them. The im\-pu\-ri\-ty of beau\-ty is sloth. The im\-pu\-ri\-ty of a watch\-man is mind\-less\-ness.
\end{para}

\begin{para}
    \textsuperscript{242}\thinspace{}Sex\-u\-al mis\-con\-duct is the im\-pu\-ri\-ty of wom\-an, stin\-gi\-ness the im\-pu\-ri\-ty of a do\-nor. All evil ways are impurities, both in this world and in the next.
    \textsuperscript{243}\thinspace{}But there is an im\-pu\-ri\-ty worse than all others: ig\-no\-rance. O monks, throw off that im\-pu\-ri\-ty to be\-come pure.
\end{para}

\begin{para}
    \textsuperscript{244}\thinspace{}Life is eas\-y for a shame\-less man, who is cun\-ning as a crow, who makes mis\-chief, and is in\-sult\-ing, ar\-ro\-gant, and cor\-rupt.
    \textsuperscript{245}\thinspace{}But life is hard for a mod\-est man, who al\-ways seeks pu\-ri\-ty, is dis\-in\-ter\-est\-ed, qui\-et, spot\-less, and ob\-ser\-vant.
\end{para}

\begin{para}
    \textsuperscript{246}\thinspace{}He who destroys life, speaks lies, takes what is not giv\-en, goes to an\-oth\-er man's wife,
    \textsuperscript{247}\thinspace{}and gives him\-self to in\-tox\-i\-cat\-ing liq\-uor, he digs up his own root here in this world.
    \textsuperscript{248}\thinspace{}Know this, good man, that evil is hard to con\-trol. Do not let greed and vi\-ce bring you to prolonged grief.
\end{para}

\begin{para}
    \textsuperscript{249}\thinspace{}Peo\-ple give ac\-cord\-ing to their faith or con\-vic\-tion. If a man worries a\-bout the food and drink giv\-en to others, he will find no rest by day or night.
    \textsuperscript{250}\thinspace{}He in whom that feel\-ing is cut down, uprooted, and destroyed, finds rest by day and night.
\end{para}

\begin{para}
    \textsuperscript{251}\thinspace{}There is no fire like pas\-sion, no grip like ha\-tred, no snare like de\-lu\-sion, no riv\-er like crav\-ing.
\end{para}

\begin{para}
    \textsuperscript{252}\thinspace{}It is eas\-y to see the faults of others, but dif\-fi\-cult to see one's own. A man winnows the faults of others like chaff, but hides his own, like a cheat hides an un\-luck\-y throw of the dice.
    \textsuperscript{253}\thinspace{}If a man fo\-cus\-es on the errors of others, and is al\-ways in\-clined to find fault, his own faults will grow, and he is far from destroying them.
\end{para}

\begin{para}
    \textsuperscript{254}\thinspace{}The world delights in van\-i\-ty, but the awakened are free from it.
    \textsuperscript{255}\thinspace{}No crea\-ture is e\-ter\-nal, but the awakened are nev\-er shak\-en.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{The Just}
\markright{The Just}
\begin{para}
    \textsuperscript{256}\thinspace{}A man does not be\-come just by pass\-ing ar\-bi\-trar\-y judge\-ment. No, he is called just who weighs both right and wrong,
    \textsuperscript{257}\thinspace{}who judges im\-par\-tial\-ly and with\-out hur\-ry, in line with law, guarding the law, guard\-ed by the law, and who is in\-tel\-li\-gent.
\end{para}

\begin{para}
    \textsuperscript{258}\thinspace{}A man is not wise just be\-cause he talks a lot. He is called wise who is pa\-tient, and free from hos\-til\-i\-ty and fear.
\end{para}

\begin{para}
    \textsuperscript{259}\thinspace{}A man does not up\-hold the law just be\-cause he talks a lot. He who has learn\-ed on\-ly a lit\-tle of the law, yet comprehends it and nev\-er neglects it, he upholds the law.
\end{para}

\begin{para}
    \textsuperscript{260}\thinspace{}A man is not an eld\-er just be\-cause his hair is grey. He may be ad\-vanced in years, but is called “one grown old in vain.”
    \textsuperscript{261}\thinspace{}He in whom there is truth, vir\-tue, love, re\-straint, self con\-trol, who is free from im\-pu\-ri\-ty, and is wise, he is called an eld\-er.
\end{para}

\begin{para}
    \textsuperscript{262}\thinspace{}El\-o\-quence or phys\-i\-cal beau\-ty do not make an en\-vi\-ous, greed\-y, dis\-hon\-est man re\-spect\-a\-ble.
    \textsuperscript{263}\thinspace{}He is called re\-spect\-a\-ble when all these are cut down, uprooted, and destroyed, when he is free from ha\-tred, and is wise.
\end{para}

\begin{para}
    \textsuperscript{264}\thinspace{}An un\-dis\-ci\-plined man who speaks lies does not be\-come a monk just by shav\-ing his head. How can he be a monk who is held cap\-tive by de\-sire and greed?
    \textsuperscript{265}\thinspace{}He who quiets evil, both small and large, is called a monk, be\-cause he has o\-ver\-come all evil.
\end{para}

\begin{para}
    \textsuperscript{266}\thinspace{}A man is not a monk just be\-cause he asks for alms. He who adopts the whole law is a monk.
    \textsuperscript{267}\thinspace{}He who is a\-bove both good and evil, who is chaste, and who goes through the world with un\-der\-stand\-ing, he in\-deed is called a monk.
\end{para}

\begin{para}
    \textsuperscript{268}\thinspace{}A fool\-ish and ig\-no\-rant man does not be\-come a sage just by si\-lence. But the wise, as if hold\-ing the scales, chooses good and avoids evil.
    \textsuperscript{269}\thinspace{}For this rea\-son he is a sage. He who weighs both sides of this world is called a sage.
\end{para}

\begin{para}
    \textsuperscript{270}\thinspace{}A man does not be\-come no\-ble by harming liv\-ing creatures. He is called no\-ble for harming no liv\-ing thing.
\end{para}

\begin{para}
    \textsuperscript{271}\thinspace{}Not by rules and practices, not by much learn\-ing, not by great con\-cen\-tra\-tion, not by life in se\-clu\-sion,
    \textsuperscript{272}\thinspace{}not by think\-ing, “I en\-joy the hap\-pi\-ness of re\-nun\-ci\-a\-tion which no world\-ling can know”, should the monk be con\-fi\-dent with\-out attaining the ex\-tinc\-tion of desires.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{The Path}
\markright{The Path}
\begin{para}
    \textsuperscript{273}\thinspace{}Of paths, the Eight\-fold is best. Of truths, the Four No\-ble are best. Of vir\-tues, de\-tach\-ment is best. Of men, he who has eyes to see is best.
\end{para}

\begin{para}
    \textsuperscript{274}\thinspace{}This is the on\-ly path, there is no oth\-er, to pu\-ri\-fy vi\-sion. Fol\-low this path and it will be\-wil\-der the temp\-ter.
    \textsuperscript{275}\thinspace{}Fol\-low\-ing this path will end pain. Once I un\-der\-stood how to re\-move thorns in the flesh, I preached this path.
\end{para}

\begin{para}
    \textsuperscript{276}\thinspace{}You your\-self must make an ef\-fort. Buddhas on\-ly show the way. The thought\-ful who fol\-low the path are freed from the tempter's bonds.
\end{para}

\begin{para}
    \textsuperscript{277}\thinspace{}“All cre\-at\-ed things are tem\-po\-rar\-y.” He who sees and knows this turns a\-way from pain. This is the path to pu\-ri\-ty.
\end{para}

\begin{para}
    \textsuperscript{278}\thinspace{}“All cre\-at\-ed things are pain\-ful.” He who sees and knows this turns a\-way from pain. This is the path to pu\-ri\-ty.
\end{para}

\begin{para}
    \textsuperscript{280}\thinspace{}He who does not ex\-ert him\-self when it is time, who, though young and strong, is full of sloth, whose will and mind are weak, that la\-zy and i\-dle man will nev\-er find the path to knowl\-edge.
\end{para}

\begin{para}
    \textsuperscript{281}\thinspace{}Let a man watch his speech, be well restrained in mind, and nev\-er com\-mit wrong with his bod\-y. If a man purifies these three cours\-es of ac\-tion, he will a\-chieve the path taught by the wise.
\end{para}

\begin{para}
    \textsuperscript{282}\thinspace{}Zeal brings knowl\-edge. Lack of zeal loses knowl\-edge. Know\-ing these two paths of gain and loss, let a man con\-duct him\-self so that knowl\-edge may grow.
\end{para}

\begin{para}
    \textsuperscript{283}\thinspace{}Cut down the whole for\-est of de\-sire, not just one tree. Dan\-ger co\-mes from the for\-est. Once you have cut down this for\-est and its un\-der\-growth, monks, you will be free.
\end{para}

\begin{para}
    \textsuperscript{284}\thinspace{}As long as the smallest crav\-ing of man for wom\-an is not destroyed, his mind is in bond\-age, as the suck\-ing calf to its moth\-er.
    \textsuperscript{285}\thinspace{}Cut out the love of self, like an au\-tumn lo\-tus with your hand. Cher\-ish the road of peace, sal\-va\-tion, as shown by the awakened.
\end{para}

\begin{para}
    \textsuperscript{286}\thinspace{}“I will live here dur\-ing the rains, and o\-ver here in win\-ter and sum\-mer.” Thus thinks the fool, un\-a\-ware of obstructions.
\end{para}

\begin{para}
    \textsuperscript{287}\thinspace{}Death carries off the man, praised for his chil\-dren and flocks, his mind dis\-tract\-ed, as a flood carries off a sleep\-ing vil\-lage.
\end{para}

\begin{para}
    \textsuperscript{288}\thinspace{}For him whom death has seized, kinsmen are no pro\-tec\-tion. Nei\-ther sons, a fa\-ther, nor re\-la\-tions can give shel\-ter.
    \textsuperscript{289}\thinspace{}Know\-ing this, a wise and good man should quick\-ly clear the path lead\-ing to sal\-va\-tion.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Miscellaneous}
\markright{Miscellaneous}
\begin{para}
    \textsuperscript{290}\thinspace{}If by leav\-ing a small pleas\-ure one sees a great pleas\-ure, let a wise man leave the small pleas\-ure and look to the great.
\end{para}

\begin{para}
    \textsuperscript{291}\thinspace{}He who seeks pleas\-ure for him\-self by caus\-ing pain to others is entangled by the bonds of ha\-tred, and can\-not be free from ha\-tred.
\end{para}

\begin{para}
    \textsuperscript{292}\thinspace{}Desires al\-ways in\-crease for un\-ru\-ly, heed\-less peo\-ple who ne\-glect what should be done and do what should not be done.
    \textsuperscript{293}\thinspace{}But desires will come to an end for watch\-ful, wise peo\-ple who do not do what should not be done, who stead\-fast\-ly do what should be done, and whose whole watch\-ful\-ness is al\-ways di\-rect\-ed to their bod\-y.
\end{para}

\begin{para}
    \textsuperscript{296}\thinspace{}The disciples of Gotama are wide a\-wake and vig\-i\-lant, and their minds day and night are al\-ways set on the Awakened,
    \textsuperscript{297}\thinspace{}on the law,
    \textsuperscript{298}\thinspace{}on the church,
    \textsuperscript{299}\thinspace{}and on their bod\-y.
\end{para}

\begin{para}
    \textsuperscript{300}\thinspace{}The disciples of Gotama are wide a\-wake and vig\-i\-lant, and their minds day and night al\-ways de\-light in com\-pas\-sion,
    \textsuperscript{301}\thinspace{}and in med\-i\-ta\-tion.
\end{para}

\begin{para}
    \textsuperscript{302}\thinspace{}It is hard to be a monk. It is hard to en\-joy that life. House\-hold life is al\-so hard and pain\-ful. It is pain\-ful to dwell with unequals and the wan\-der\-ing monk is be\-set with pain. There\-fore let no man wan\-der aim\-less\-ly and he will not be be\-set with pain.
\end{para}

\begin{para}
    \textsuperscript{303}\thinspace{}A faith\-ful, vir\-tu\-ous, rep\-u\-ta\-ble, and wealth\-y man is respected wher\-ev\-er he goes.
    \textsuperscript{304}\thinspace{}Like the Himalayas, good peo\-ple shine from a\-far. Like ar\-rows shot by night, bad peo\-ple are un\-seen e\-ven when they are near.
\end{para}

\begin{para}
    \textsuperscript{305}\thinspace{}He who sits a\-lone, sleeps a\-lone, walks a\-lone, is dil\-i\-gent, and subdues him\-self a\-lone, will re\-joice in the sol\-i\-tude of the for\-est.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Hell}
\markright{Hell}
\begin{para}
    \textsuperscript{306}\thinspace{}He goes to hell, the one who asserts what did\-n't hap\-pen, and the one who did a thing but says, “I have not done it”. They are both evildoers, and are e\-qual in the next world.
\end{para}

\begin{para}
    \textsuperscript{307}\thinspace{}Man\-y men who wear the saf\-fron robe are ill-mannered and un\-re\-strained. Such evildoers by their evil deeds go to hell.
    \textsuperscript{308}\thinspace{}It would be bet\-ter to swal\-low an i\-ron ball, red-hot and a\-flame, than for an im\-mor\-al, un\-re\-strained man to eat the alms of the peo\-ple.
\end{para}

\begin{para}
    \textsuperscript{309}\thinspace{}A wreckless man who lies down with his neighbor's wife gains four things: a bad rep\-u\-ta\-tion, poor sleep, pun\-ish\-ment, and hell.
    \textsuperscript{310}\thinspace{}There is bad rep\-u\-ta\-tion, an evil des\-ti\-na\-tion, the brief pleas\-ure of the fright\-ened man and wom\-an, and the king imposes a heav\-y pun\-ish\-ment. There\-fore, let no man lie down with his neighbor's wife.
\end{para}

\begin{para}
    \textsuperscript{311}\thinspace{}Just as a blade of grass, if wrong\-ly han\-dled, cuts the hand, so does as\-cet\-i\-cism, if wrong\-ly prac\-tised, lead to hell.
\end{para}

\begin{para}
    \textsuperscript{312}\thinspace{}An act performed care\-less\-ly, a bro\-ken vow, and ques\-tion\-a\-ble cel\-i\-ba\-cy, none of these bring much re\-ward.
    \textsuperscript{313}\thinspace{}If any\-thing is to be done, then do it vig\-or\-ous\-ly. A care\-less pil\-grim on\-ly scatters the dust of his passions more wide\-ly.
\end{para}

\begin{para}
    \textsuperscript{314}\thinspace{}It is bet\-ter not to do an evil deed, for an evil deed torments a man af\-ter\-wards. It is bet\-ter to do a good deed, for hav\-ing done it, one is not tormented.
    \textsuperscript{315}\thinspace{}Like a fron\-tier fort with defences in\-side and out, so let a man guard him\-self. Do not let a mo\-ment pass by, for they who al\-low the right mo\-ment to pass, suf\-fer pain when consigned to hell.
\end{para}

\begin{para}
    \textsuperscript{316}\thinspace{}They who are a\-shamed of what is not shame\-ful, and are not a\-shamed of what is shame\-ful, such men, em\-brac\-ing false doctrines go to an evil des\-ti\-na\-tion.
\end{para}

\begin{para}
    \textsuperscript{317}\thinspace{}They who see dan\-ger where there is none, and see no dan\-ger where there is, such men, em\-brac\-ing false doctrines go to an evil des\-ti\-na\-tion.
\end{para}

\begin{para}
    \textsuperscript{318}\thinspace{}They who see wrong where there is none, and see no wrong where there is, such men, em\-brac\-ing false doctrines go to an evil des\-ti\-na\-tion.
\end{para}

\begin{para}
    \textsuperscript{319}\thinspace{}They who see wrong as wrong, and see right as right, such men, em\-brac\-ing true doc\-trine go to a good des\-ti\-na\-tion.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{The Elephant}
\markright{The Elephant}
\begin{para}
    \textsuperscript{320}\thinspace{}As an el\-e\-phant in bat\-tle endures an ar\-row shot from a bow, so will I en\-dure def\-a\-ma\-tion. Man\-y peo\-ple lack vir\-tue.
\end{para}

\begin{para}
    \textsuperscript{321}\thinspace{}They on\-ly lead a tamed el\-e\-phant in\-to crowds. The king on\-ly rides a tamed el\-e\-phant. The tamed man who endures def\-a\-ma\-tion is the best a\-mong men.
\end{para}

\begin{para}
    \textsuperscript{322}\thinspace{}Trained mules, thor\-ough\-bred hors\-es, and elephants with large tusks are ex\-cel\-lent, but he who tames him\-self is bet\-ter still.
    \textsuperscript{323}\thinspace{}For with these animals no man reaches the un\-trod\-den coun\-try, where a tamed man goes by tam\-ing him\-self.
\end{para}

\begin{para}
    \textsuperscript{324}\thinspace{}The el\-e\-phant called Dhanapalaka, deep in rut, is dif\-fi\-cult to hold. He does not eat a mor\-sel in cap\-tiv\-i\-ty, but longs for the el\-e\-phant grove.
\end{para}

\begin{para}
    \textsuperscript{325}\thinspace{}When a man is in\-ac\-tive and glut\-ton\-ous, sleep\-ing and roll\-ing a\-round in bed like an o\-ver\-fed hog, that fool is born a\-gain and a\-gain.
\end{para}

\begin{para}
    \textsuperscript{326}\thinspace{}This mind used to wan\-der a\-bout as it liked, wher\-ev\-er it wished, how\-ev\-er it pleased. Now I will hold it in check, as a rid\-er with a hook controls an el\-e\-phant in rut.
\end{para}

\begin{para}
    \textsuperscript{327}\thinspace{}De\-light in mind\-ful\-ness. Watch o\-ver your mind. Lift your\-self out of the evil way, as an el\-e\-phant lifts it\-self out of the mud.
\end{para}

\begin{para}
    \textsuperscript{328}\thinspace{}If you find a wise, pru\-dent, and vir\-tu\-ous friend, then, o\-ver\-com\-ing all dangers, walk with him joy\-ful\-ly and mind\-ful\-ly.
\end{para}

\begin{para}
    \textsuperscript{329}\thinspace{}If you find no friend who is wise, pru\-dent, and vir\-tu\-ous, then walk a\-lone like a king leav\-ing be\-hind his king\-dom, like an el\-e\-phant in the for\-est.
    \textsuperscript{330}\thinspace{}It is bet\-ter to live a\-lone. There is no com\-pan\-ion\-ship with a fool. Live a\-lone, do no evil, and be con\-tent, like an el\-e\-phant in the for\-est.
\end{para}

\begin{para}
    \textsuperscript{331}\thinspace{}These are good: friends when the need arises, con\-tent\-ment with what you have, mer\-it in the hour of death, giv\-ing up all grief,
    \textsuperscript{332}\thinspace{}serv\-ing your moth\-er, serv\-ing your fa\-ther, serv\-ing monks, serv\-ing ho\-ly men,
    \textsuperscript{333}\thinspace{}vir\-tue last\-ing in\-to old age, firm\-ly root\-ed faith, attaining dis\-cern\-ment, avoiding evil.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Craving}
\markright{Craving}
\begin{para}
    \textsuperscript{334}\thinspace{}The crav\-ing of a heed\-less man grows like a creep\-ing vine. He runs from life to life, like a mon\-key seek\-ing fruit in the for\-est.
\end{para}

\begin{para}
    \textsuperscript{335}\thinspace{}He who is o\-ver\-come by this vile and cling\-ing crav\-ing, his suf\-fer\-ing increases like grass af\-ter rain.
    \textsuperscript{336}\thinspace{}He who overcomes this vile crav\-ing, so dif\-fi\-cult to con\-quer, suf\-fer\-ing falls off him like wa\-ter drops from a lo\-tus leaf.
\end{para}

\begin{para}
    \textsuperscript{337}\thinspace{}To all as\-sem\-bled here I say, “Good luck. Dig up the root of crav\-ing, like one seek\-ing the sweet scented root of birana grass. Let not the temp\-ter crush you a\-gain and a\-gain, as the stream crushes a reed.”
    \textsuperscript{338}\thinspace{}A tree will grow a\-gain, e\-ven if cut down, as long as its root is safe and firm. Thus, un\-less la\-tent crav\-ing is destroyed, suf\-fer\-ing will re\-turn a\-gain and a\-gain.
\end{para}

\begin{para}
    \textsuperscript{339}\thinspace{}The thir\-ty six streams of crav\-ing flow to\-wards pleas\-ure. The mis\-guid\-ed man, in whom they flow strongly, is car\-ried a\-way.
    \textsuperscript{340}\thinspace{}The streams flow eve\-ry\-where. The creep\-ing vines of crav\-ing grow eve\-ry\-where. If you see the vine spring up, cut its root with knowl\-edge.
\end{para}

\begin{para}
    \textsuperscript{342}\thinspace{}Driv\-en by crav\-ing, men run a\-bout like a snared rab\-bit. Held by fetters and bonds, they suf\-fer a\-gain and a\-gain for a long time.
    \textsuperscript{343}\thinspace{}Driv\-en by crav\-ing, men run a\-bout like a snared rab\-bit. So a monk should drive out crav\-ing, and strive to be with\-out pas\-sion.
\end{para}

\begin{para}
    \textsuperscript{345}\thinspace{}The wise do not say that a fet\-ter made of i\-ron, wood, or hemp is strong. Far stronger is a long\-ing for jewels and ornaments, chil\-dren and wives.
    \textsuperscript{346}\thinspace{}The wise say that a strong fet\-ter drags one down, though seem\-ing loose, it is dif\-fi\-cult to re\-move. Af\-ter hav\-ing cut it, the wise are free from cares, and leave the world, desires, and pleasures be\-hind.
\end{para}

\begin{para}
    \textsuperscript{348}\thinspace{}Give up the past, the fu\-ture, and the pres\-ent, to cross to the oth\-er shore of ex\-ist\-ence. If your mind is com\-plete\-ly free, you will not en\-ter in\-to birth and de\-cay a\-gain.
\end{para}

\begin{para}
    \textsuperscript{349}\thinspace{}If a man is tossed a\-bout by doubts, full of pas\-sion, and fo\-cused on pleas\-ure, his crav\-ing grows more and more. He makes his fetters strong, in\-deed.
    \textsuperscript{350}\thinspace{}If a man delights in quieting doubts, is al\-ways mind\-ful, and meditates on the impurities, he will end crav\-ing and cut the fet\-ter of the temp\-ter.
\end{para}

\begin{para}
    \textsuperscript{351}\thinspace{}He who has reached the goal is with\-out fear, is with\-out crav\-ing, is with\-out blem\-ish, and has re\-moved all the thorns of life: this will be his last bod\-y.
\end{para}

\begin{para}
    \textsuperscript{354}\thinspace{}The gift of the law exceeds all gifts. The taste of the law exceeds all tastes. De\-light in the law exceeds all delights. The ex\-tinc\-tion of crav\-ing overcomes all suf\-fer\-ing.
\end{para}

\begin{para}
    \textsuperscript{355}\thinspace{}Rich\-es de\-stroy the fool\-ish, but not those who seek the be\-yond. The fool, by crav\-ing for rich\-es, destroys him\-self and others.
\end{para}

\begin{para}
    \textsuperscript{356}\thinspace{}Weeds spoil a field. Pas\-sion spoils a man. There\-fore, a gift giv\-en to those with\-out pas\-sion yields great fruit.
\end{para}

\begin{para}
    \textsuperscript{357}\thinspace{}Weeds spoil a field. Ha\-tred spoils a man. There\-fore, a gift giv\-en to those with\-out ha\-tred yields great fruit.
\end{para}

\begin{para}
    \textsuperscript{358}\thinspace{}Weeds spoil a field. De\-lu\-sion spoils a man. There\-fore, a gift giv\-en to those with\-out de\-lu\-sion yields great fruit.
\end{para}

\begin{para}
    \textsuperscript{359}\thinspace{}Weeds spoil a field. De\-sire spoils a man. There\-fore, a gift giv\-en to those with\-out de\-sire yields great fruit.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{The Monk}
\markright{The Monk}
\begin{para}
    \textsuperscript{360}\thinspace{}Good is re\-straint o\-ver the eye, o\-ver the ear, o\-ver the nose, and o\-ver the tongue.
    \textsuperscript{361}\thinspace{}Good is re\-straint in the bod\-y, in speech, in mind, in all things. A monk, restrained in all things, is freed from all suf\-fer\-ing.
\end{para}

\begin{para}
    \textsuperscript{362}\thinspace{}He who controls his hands, feet, and speech; who is well con\-trolled, delights in the in\-ward, is col\-lect\-ed, sol\-i\-tar\-y and con\-tent, him they call a monk.
\end{para}

\begin{para}
    \textsuperscript{363}\thinspace{}The monk who controls his tongue, speaks mod\-er\-ate\-ly and calm\-ly, and teaches the law and its mean\-ing, his speech is sweet.
\end{para}

\begin{para}
    \textsuperscript{364}\thinspace{}The monk who abides in the law, delights in the law, meditates on the law, and keeps the law in mind, he will nev\-er fall a\-way from the true law.
\end{para}

\begin{para}
    \textsuperscript{365}\thinspace{}Let him not des\-pise what he has re\-ceived, nor en\-vy others. A monk who envies others does not ob\-tain con\-cen\-tra\-tion.
\end{para}

\begin{para}
    \textsuperscript{366}\thinspace{}If a monk lives pure\-ly, is dil\-i\-gent, and does not des\-pise what he has re\-ceived, though he receives lit\-tle, then e\-ven the gods will praise him.
    \textsuperscript{368}\thinspace{}The monk who abides in kind\-ness, and is de\-vot\-ed to the teachings of Bud\-dha, will at\-tain peace, hap\-pi\-ness, and ces\-sa\-tion of nat\-u\-ral desires.
\end{para}

\begin{para}
    \textsuperscript{369}\thinspace{}Monk, bail out this boat. Once bailed, it will sail light\-ly. Hav\-ing cut out pas\-sion and ha\-tred, you will reach Sal\-va\-tion.
\end{para}

\begin{para}
    \textsuperscript{371}\thinspace{}Med\-i\-tate monk. Be mind\-ful. Do not di\-rect your mind to sen\-su\-al pleas\-ure. Heed\-less, do not swal\-low the flam\-ing i\-ron ball lest you cry out when burn\-ing, “This is suf\-fer\-ing.”
\end{para}

\begin{para}
    \textsuperscript{372}\thinspace{}There is no med\-i\-ta\-tion with\-out in\-sight. There is no in\-sight with\-out med\-i\-ta\-tion. He who has in\-sight and med\-i\-ta\-tion is near Sal\-va\-tion.
\end{para}

\begin{para}
    \textsuperscript{373}\thinspace{}A monk who has entered his emp\-ty house, whose mind is tran\-quil, and who sees the law clear\-ly, feels a de\-light more than hu\-man.
    \textsuperscript{374}\thinspace{}Once he has con\-sid\-ered the or\-i\-gin and de\-struc\-tion of the elements of the bod\-y, he finds hap\-pi\-ness and joy be\-long\-ing to those who know the im\-mor\-tal.
\end{para}

\begin{para}
    \textsuperscript{375}\thinspace{}This is the be\-gin\-ning for a wise monk: guarding the sen\-ses, con\-tent\-ment, re\-straint un\-der the law.
    \textsuperscript{376}\thinspace{}He should have friends who are no\-ble, en\-er\-get\-ic, and pure. He should be friend\-ly, and skilled in con\-duct. Then, hav\-ing much joy, he will put an end to suf\-fer\-ing.
\end{para}

\begin{para}
    \textsuperscript{377}\thinspace{}Shed pas\-sion and ha\-tred, O monks, as the jas\-mine plant sheds its withered flowers.
    \textsuperscript{378}\thinspace{}That monk is called qui\-et whose bod\-y and speech and mind are calm, who is col\-lect\-ed, and who has purged the baits of the world.
\end{para}

\begin{para}
    \textsuperscript{379}\thinspace{}Re\-prove your\-self and ex\-am\-ine your\-self by your\-self. By be\-ing vig\-i\-lant and at\-ten\-tive, you will live hap\-pi\-ly.
    \textsuperscript{380}\thinspace{}You are your own mas\-ter and ref\-uge. There\-fore, watch o\-ver your\-self as a mer\-chant watches o\-ver a good horse.
\end{para}

\begin{para}
    \textsuperscript{381}\thinspace{}The monk who is full of joy, and who has faith in the teachings of Bud\-dha will reach the qui\-et place, ces\-sa\-tion of nat\-u\-ral desires, and hap\-pi\-ness.
    \textsuperscript{382}\thinspace{}E\-ven a young monk who applies him\-self to the teachings of Bud\-dha brightens this world, like the moon set free from clouds.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{The Holy Man}
\markright{The Holy Man}
\begin{para}
    \textsuperscript{386}\thinspace{}He who is mind\-ful, blame\-less, settled, du\-ti\-ful, with\-out passions, and who has attained the highest goal, him I call a ho\-ly man.
\end{para}

\begin{para}
    \textsuperscript{387}\thinspace{}The sun shines by day, the moon shines by night, the war\-ri\-or shines in ar\-mor, the ho\-ly man shines in med\-i\-ta\-tion. But the Awakened One shines with splen\-dor day and night.
\end{para}

\begin{para}
    \textsuperscript{388}\thinspace{}A man is called ho\-ly for discarding evil, con\-tem\-pla\-tive for walk\-ing qui\-et\-ly, a pil\-grim for leav\-ing his own impurities.
\end{para}

\begin{para}
    \textsuperscript{389}\thinspace{}No one should at\-tack a ho\-ly man, but no ho\-ly man, if attacked, should give in to his an\-ger. Shame on him who attacks a ho\-ly man, but more shame to him who gives in to his an\-ger.
\end{para}

\begin{para}
    \textsuperscript{390}\thinspace{}Noth\-ing is bet\-ter for a ho\-ly man than to hold his mind back from the pleasures of life. When all de\-sire to in\-jure has vanished, pain ceases.
\end{para}

\begin{para}
    \textsuperscript{391}\thinspace{}I call him a ho\-ly man who does not of\-fend in bod\-y, word, or mind, and is restrained on these three points.
\end{para}

\begin{para}
    \textsuperscript{392}\thinspace{}Re\-vere that teach\-er who has taught you the law as taught by the Bud\-dha, just as a ho\-ly man\-y reveres the sac\-ri\-fi\-cial fire.
\end{para}

\begin{para}
    \textsuperscript{393}\thinspace{}A man does not be\-come ho\-ly by his mat\-ted hair, by his fam\-i\-ly, or by birth. He in whom there is truth and right\-eous\-ness, he is pure, he is a ho\-ly man.
    \textsuperscript{394}\thinspace{}What use is mat\-ted hair, O fool? What use is your coat of goat skins? The tan\-gle is with\-in you, but you comb the out\-side.
\end{para}

\begin{para}
    \textsuperscript{397}\thinspace{}I call him a ho\-ly man who has cut eve\-ry fet\-ter, is fear\-less, in\-de\-pend\-ent and unshackled.
\end{para}

\begin{para}
    \textsuperscript{399}\thinspace{}I call him a ho\-ly man who, with\-out an\-ger, endures in\-sult, im\-pris\-on\-ment, and beat\-ing, who has pa\-tience like the strength of an ar\-my.
\end{para}

\begin{para}
    \textsuperscript{400}\thinspace{}I call him a ho\-ly man who is free from an\-ger, du\-ti\-ful, vir\-tu\-ous, with\-out crav\-ing, sub\-dued, and who has re\-ceived his last bod\-y.
\end{para}

\begin{para}
    \textsuperscript{401}\thinspace{}I call him a ho\-ly man who does not cling to pleasures, like wa\-ter on a lo\-tus leaf or a mus\-tard seed on the tip of a nee\-dle.
    \textsuperscript{402}\thinspace{}I call him a ho\-ly man who, e\-ven here, discerns the end of his suf\-fer\-ing, has put down his bur\-den, and is unshackled.
\end{para}

\begin{para}
    \textsuperscript{403}\thinspace{}I call him a ho\-ly man who is wise, has deep un\-der\-stand\-ing, discerns the right and wrong path, and has reached the ul\-ti\-mate goal.
\end{para}

\begin{para}
    \textsuperscript{404}\thinspace{}I call him a ho\-ly man who keeps a\-loof from both laymen and as\-cet\-ic, who has no per\-ma\-nent home, and few desires.
\end{para}

\begin{para}
    \textsuperscript{405}\thinspace{}I call him a ho\-ly man who renounces vi\-o\-lence to\-wards all beings, wheth\-er weak or strong, and nei\-ther kills nor causes others to kill.
\end{para}

\begin{para}
    \textsuperscript{407}\thinspace{}I call him a ho\-ly man from whom pas\-sion and ha\-tred, pride and en\-vy have dropped like a mus\-tard seed from the tip of a nee\-dle.
\end{para}

\begin{para}
    \textsuperscript{408}\thinspace{}I call him a ho\-ly man who speaks true and in\-struc\-tive words, free from harsh\-ness, so that he offends no one.
    \textsuperscript{409}\thinspace{}I call him a ho\-ly man who on\-ly takes what is giv\-en to him, be it long or short, small or large, good or bad.
\end{para}

\begin{para}
    \textsuperscript{412}\thinspace{}I call him a ho\-ly man who, in this world, is a\-bove at\-tach\-ment to mer\-it and evil, is with\-out grief, stain\-less, and pure.
\end{para}

\begin{para}
    \textsuperscript{416}\thinspace{}I call him a ho\-ly man who, leav\-ing all crav\-ing, travels with\-out a home, and in whom all crav\-ing is ex\-tinct.
\end{para}

\begin{para}
    \textsuperscript{417}\thinspace{}I call him a ho\-ly man who, af\-ter leav\-ing all at\-tach\-ment to men, has ris\-en a\-bove all at\-tach\-ment to the gods, and is free from eve\-ry at\-tach\-ment.
\end{para}

\begin{para}
    \textsuperscript{421}\thinspace{}I call him a ho\-ly man who does not cling to past, pres\-ent, or fu\-ture, who has noth\-ing, who clings to noth\-ing.
\end{para}

\begin{para}
    \textsuperscript{422}\thinspace{}I call him a ho\-ly man, the fear\-less, the no\-ble, the he\-ro\-ic, the great sage, the con\-quer\-or, the pas\-sion\-less, the pure, the awakened.
\end{para}

\begin{para}
    \textsuperscript{423}\thinspace{}I call him a ho\-ly man who knows his form\-er lives, who sees heav\-en and hell, has reached the end of births, is per\-fect in knowl\-edge, a sage, and who has mastered all there is to mas\-ter.
\end{para}

\chapter*{Jesus}
\markboth{Jesus}{Jesus}
\addcontentsline{toc}{section}{Jesus}\begin{changemargin}{.6in}{.6in}
  \begin{center}
    \itshape
    An a\-bridg\-ment of the Gos\-pel of Mark fo\-cus\-ing on the Tri\-ple Tra\-di\-tion.
  \end{center}
\end{changemargin}
\begin{para}
    \textsuperscript{1:4}\thinspace{}John the Baptist appeared in the wil\-der\-ness, preaching a bap\-tism of re\-pent\-ance for the for\-give\-ness of sins.
    \textsuperscript{5}\thinspace{}All the peo\-ple were go\-ing out to him and were be\-ing bap\-tized by him in the riv\-er Jor\-dan, con\-fess\-ing their sins.
    \textsuperscript{6}\thinspace{}John was clothed with camel's hair, wore a leath\-er belt a\-round his waist, and ate locusts and wild hon\-ey.
\end{para}

\begin{para}
    \textsuperscript{1:7}\thinspace{}He preached, say\-ing, “Af\-ter me co\-mes he who is might\-i\-er than I. I'm not wor\-thy to stoop down and un\-tie the strap of his sandals.
    \textsuperscript{8}\thinspace{}I bap\-tized you with wa\-ter, but he'll bap\-tize you with the ho\-ly spir\-it.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{1:9}\thinspace{}In those days Jesus was bap\-tized by John in the Jor\-dan.
    \textsuperscript{10}\thinspace{}Just as he was com\-ing up out of the wa\-ter, he saw the heavens be\-ing torn o\-pen and the spir\-it descending on him like a dove.
    \textsuperscript{11}\thinspace{}And a voice came out of the heavens, “You are my be\-lov\-ed son; with you I am well pleased.”
\end{para}

\begin{para}
    \textsuperscript{1:12}\thinspace{}Im\-me\-di\-ate\-ly the spir\-it drove him in\-to the wil\-der\-ness.
    \textsuperscript{13}\thinspace{}He was in the wil\-der\-ness for\-ty days, be\-ing tempted by the temp\-ter. He was with the wild animals, and the angels ministered to him.
\end{para}

\begin{para}
    \textsuperscript{1:14}\thinspace{}Af\-ter John was arrested, Jesus came in\-to Gal\-i\-lee, preaching the gos\-pel of God,
    \textsuperscript{15}\thinspace{}and say\-ing, “The time is ful\-filled, and God's king\-dom is at hand. Re\-pent and be\-lieve in the gos\-pel.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{1:16}\thinspace{}Walk\-ing a\-long by the Sea of Gal\-i\-lee, he saw Simon and Andrew the broth\-er of Simon cast\-ing a net in\-to the sea, for they were fishermen.
    \textsuperscript{17}\thinspace{}Jesus said to them, “Come, fol\-low me, and I'll make you be\-come fishers of men.”
    \textsuperscript{18}\thinspace{}Im\-me\-di\-ate\-ly they left the nets, and followed him.
\end{para}

\begin{para}
    \textsuperscript{1:19}\thinspace{}Go\-ing on a lit\-tle far\-ther, he saw James the son of Zebedee, and John his broth\-er, who were in the boat mend\-ing their nets.
    \textsuperscript{20}\thinspace{}Im\-me\-di\-ate\-ly he called them, and they left their fa\-ther Zebedee in the boat with the hired servants, and followed him.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{1:29}\thinspace{}They entered the house of Simon and Andrew, with James and John.
\end{para}

\begin{para}
    \textsuperscript{1:30}\thinspace{}Now Simon's mother-in-law lay sick with a fe\-ver, and im\-me\-di\-ate\-ly they told him a\-bout her.
    \textsuperscript{31}\thinspace{}He came, took her by the hand, and helped her up. The fe\-ver left her, and she be\-gan to wait on them.
\end{para}

\begin{para}
    \textsuperscript{1:32}\thinspace{}That eve\-ning, af\-ter the sun had set, they brought to him all who were sick and pos\-sessed by demons.
    \textsuperscript{33}\thinspace{}The whole cit\-y gathered to\-geth\-er at the door.
    \textsuperscript{34}\thinspace{}He healed man\-y who were sick with var\-i\-ous diseases, and cast out man\-y demons. He would not al\-low the demons to speak, be\-cause they knew him.
    \textsuperscript{39}\thinspace{}He went in\-to their synagogues through\-out all Gal\-i\-lee, preaching and cast\-ing out demons.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{1:40}\thinspace{}A lep\-er came to him, and kneel\-ing said to him, “If you're will\-ing, you can make me clean.”
\end{para}

\begin{para}
    \textsuperscript{1:41}\thinspace{}Moved with com\-pas\-sion, he touched him, and said to him, “I am will\-ing. Be clean.”
    \textsuperscript{42}\thinspace{}Im\-me\-di\-ate\-ly the lep\-ro\-sy left him, and he was cleansed.
    \textsuperscript{43}\thinspace{}Jesus sent him a\-way at once and sternly warned him,
    \textsuperscript{44}\thinspace{}“Say noth\-ing to any\-one, but go show your\-self to the priest and of\-fer for your cleans\-ing what Mo\-ses commanded, as a tes\-ti\-mo\-ny to them.”
\end{para}

\begin{para}
    \textsuperscript{1:45}\thinspace{}But he went out and be\-gan to talk free\-ly a\-bout it, spreading the news, so that Jesus could no long\-er en\-ter a cit\-y pub\-lic\-ly, but stayed out in de\-sert\-ed places, and peo\-ple came to him from eve\-ry\-where.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{2:1}\thinspace{}When he returned to Capernaum some days lat\-er, peo\-ple heard that he was at home.
    \textsuperscript{2}\thinspace{}Man\-y were gathered to\-geth\-er, so that there was no more room for them, not e\-ven near the door. He preached the word to them.
    \textsuperscript{3}\thinspace{}They came, bring\-ing to him a par\-a\-lyt\-ic, car\-ried by four of them.
    \textsuperscript{4}\thinspace{}When they could not come near him be\-cause of the crowd, they re\-moved the roof a\-bove him, made an o\-pen\-ing, and let down the mat on which the par\-a\-lyt\-ic was ly\-ing.
    \textsuperscript{5}\thinspace{}When Jesus saw their faith, he said to the par\-a\-lyt\-ic, “Son, your sins are for\-giv\-en.”
\end{para}

\begin{para}
    \textsuperscript{2:6}\thinspace{}But some of the scribes were sit\-ting there, think\-ing to them\-selves,
    \textsuperscript{7}\thinspace{}“Why does he talk like that? He's blas\-phem\-ing. Who can for\-give sins but God a\-lone?”
\end{para}

\begin{para}
    \textsuperscript{2:8}\thinspace{}Im\-me\-di\-ate\-ly Jesus perceived in his spir\-it what they were think\-ing, and said to them, “Why do you ques\-tion these things in your hearts?
    \textsuperscript{9}\thinspace{}Which is eas\-i\-er, to say to the par\-a\-lyt\-ic, ‘Your sins are for\-giv\-en,’ or to say, ‘Rise, pick up your mat, and walk’?
    \textsuperscript{10}\thinspace{}But that you may know that the son of man has au\-thor\-i\-ty on earth to for\-give sins,” he said to the par\-a\-lyt\-ic,
    \textsuperscript{11}\thinspace{}“I tell you, Rise, pick up your mat, and go home.”
\end{para}

\begin{para}
    \textsuperscript{2:12}\thinspace{}He got up, im\-me\-di\-ate\-ly pick\-ed up the mat, and went out in front of them all, so that they were all a\-mazed, and glo\-ri\-fied God, say\-ing, “We've nev\-er seen any\-thing like this.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{2:13}\thinspace{}He went out a\-gain be\-side the sea, and all the crowd came to him, and he taught them.
    \textsuperscript{14}\thinspace{}As he passed by, he saw Levi the son of Alphaeus sit\-ting in the tax booth, and said to him, “Fol\-low me.” He got up and followed him.
\end{para}

\begin{para}
    \textsuperscript{2:15}\thinspace{}As he re\-clined at the ta\-ble in his house, man\-y tax collectors and sinners re\-clined with Jesus and his disciples, for there were man\-y who followed him.
    \textsuperscript{16}\thinspace{}The scribes of the Pharisees asked his disciples, “Why does he eat with tax collectors and sinners?”
\end{para}

\begin{para}
    \textsuperscript{2:17}\thinspace{}When Jesus heard it, he said to them, “Those who are health\-y don't need a doc\-tor, but those who are sick. I did\-n't come to call the right\-eous, but sinners.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{2:18}\thinspace{}Some peo\-ple came and asked him, “Why do John's disciples and the disciples of the Pharisees fast, but your disciples don't fast?”
\end{para}

\begin{para}
    \textsuperscript{2:19}\thinspace{}Jesus said to them, “Can wed\-ding guests fast while the groom is with them? As long as they have the groom with them, they can't fast.
    \textsuperscript{20}\thinspace{}The days will come, when the groom is tak\-en a\-way from them, and then they will fast in that day.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{2:23}\thinspace{}One Sabbath he was go\-ing through the grainfields, and his disciples be\-gan to pluck some heads of grain as they went.
    \textsuperscript{24}\thinspace{}The Pharisees said to him, “Look. Why are they do\-ing what's un\-law\-ful on the Sabbath?”
\end{para}

\begin{para}
    \textsuperscript{2:25}\thinspace{}He said to them, “Have\-n't you read what David did when he and his companions were in need and hun\-gry?
    \textsuperscript{26}\thinspace{}He entered the house of God and ate the con\-se\-crat\-ed bread, which is on\-ly law\-ful for the priests to eat, and he al\-so gave some to his companions.”
\end{para}

\begin{para}
    \textsuperscript{2:27}\thinspace{}He said to them, “The Sabbath was made for man, not man for the Sabbath.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{3:1}\thinspace{}He entered the syn\-a\-gogue a\-gain, and a man with a withered hand was there.
    \textsuperscript{2}\thinspace{}They watched Jesus to see if he would heal him on the Sabbath, so that they might ac\-cuse him.
\end{para}

\begin{para}
    \textsuperscript{3:3}\thinspace{}He said to the man with the withered hand, “Come for\-ward.”
    \textsuperscript{4}\thinspace{}He asked them, “Is it law\-ful on the Sabbath to do good or to do harm, to save life or to kill?” But they were si\-lent.
\end{para}

\begin{para}
    \textsuperscript{3:5}\thinspace{}He looked a\-round at them with an\-ger, grieved at the hard\-ness of their hearts, and said to the man, “Stretch out your hand.” He stretched it out, and his hand was re\-stored.
    \textsuperscript{6}\thinspace{}The Pharisees went out and im\-me\-di\-ate\-ly held coun\-sel with the Herodians a\-gainst him, how to de\-stroy him.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{3:7}\thinspace{}Jesus with\-drew with his disciples to the sea, and a large crowd followed him.
    \textsuperscript{8}\thinspace{}Hear\-ing the great things he did, a large crowd came to him.
\end{para}

\begin{para}
    \textsuperscript{3:9}\thinspace{}Be\-cause of the crowd, he told his disciples to have a boat wait for him, so they would\-n't crush him.
    \textsuperscript{10}\thinspace{}Since he had healed so man\-y, eve\-ry\-one who had diseases pressed a\-round him to tou\-ch him.
    \textsuperscript{11}\thinspace{}When\-ev\-er the un\-clean spirits saw him, they fell down in front of him, and cried out, “You are the son of God.”
    \textsuperscript{12}\thinspace{}And he sternly ordered them not to make him known.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{3:13}\thinspace{}He went up on the moun\-tain and summoned those he wanted, and they came to him.
    \textsuperscript{14}\thinspace{}He ap\-point\-ed twelve, whom he al\-so named apostles, to be with him, and to send them out to preach,
    \textsuperscript{15}\thinspace{}and to have au\-thor\-i\-ty to cast out demons.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{3:22}\thinspace{}The scribes said, “He casts out demons by the prince of demons.”
\end{para}

\begin{para}
    \textsuperscript{3:23}\thinspace{}He called them o\-ver to him, and said to them in parables, “How can the temp\-ter cast out the temp\-ter?
    \textsuperscript{24}\thinspace{}If a king\-dom is di\-vid\-ed a\-gainst it\-self, that king\-dom can\-not stand.
    \textsuperscript{25}\thinspace{}If a house is di\-vid\-ed a\-gainst it\-self, that house will not be a\-ble to stand.
    \textsuperscript{26}\thinspace{}If the temp\-ter has ris\-en up a\-gainst him\-self, and is di\-vid\-ed, he can\-not stand, but is fin\-ished.
    \textsuperscript{27}\thinspace{}No one can en\-ter a strong man's house, and plun\-der his goods, un\-less he first ties up the strong man, and then he can plun\-der his house.”
\end{para}

\begin{para}
    \textsuperscript{3:28}\thinspace{}“Tru\-ly I say to you, the sons of men will be for\-giv\-en all sins and what\-ev\-er blasphemies they speak,
    \textsuperscript{29}\thinspace{}but who\-ev\-er blasphemes a\-gainst the ho\-ly spir\-it will nev\-er be for\-giv\-en, but is guilt\-y of an e\-ter\-nal sin,”
    \textsuperscript{30}\thinspace{}be\-cause they said, “He has an un\-clean spir\-it.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{3:32}\thinspace{}A crowd was sit\-ting a\-round him, and said to him, “Your moth\-er and brothers are out\-side looking for you.”
\end{para}

\begin{para}
    \textsuperscript{3:33}\thinspace{}He answered them, “Who are my moth\-er and my brothers?”
    \textsuperscript{34}\thinspace{}Looking at those seated a\-round him, he said, “Here are my moth\-er and my brothers.
    \textsuperscript{35}\thinspace{}Who\-ev\-er does the will of God, he is my broth\-er and sis\-ter and moth\-er.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{4:1}\thinspace{}He be\-gan to teach by the sea a\-gain. A ver\-y great crowd gathered a\-round him, so he got in\-to a boat on the sea and sat down. The whole crowd was by the sea on the land.
    \textsuperscript{2}\thinspace{}He taught them man\-y things in parables, and in his teach\-ing said to them,
    \textsuperscript{3}\thinspace{}“A sower went out to sow.
    \textsuperscript{4}\thinspace{}As he sowed, some seed fell a\-long the path, and the birds came and devoured it.
    \textsuperscript{5}\thinspace{}Oth\-er seed fell on rock\-y ground, where it had lit\-tle soil. It sprang up quick\-ly, be\-cause it had no depth of soil.
    \textsuperscript{6}\thinspace{}When the sun rose, it was scorched. Be\-cause it had no root, it withered a\-way.
    \textsuperscript{7}\thinspace{}Oth\-er seed fell a\-mong thorns. The thorns grew up, and choked it, and it pro\-duced no grain.
    \textsuperscript{8}\thinspace{}Oth\-er seed fell on good soil. It grew up, in\-creased, and pro\-duced grain, bring\-ing forth thir\-ty, six\-ty, and a hun\-dred\-fold.
    \textsuperscript{9}\thinspace{}He who has ears to hear, let him hear.”
\end{para}

\begin{para}
    \textsuperscript{4:10}\thinspace{}When he was a\-lone, those a\-round him with the twelve asked him a\-bout the parables.
    \textsuperscript{14}\thinspace{}He said to them, “The sower sows the word.
    \textsuperscript{15}\thinspace{}These are the ones a\-long the path, where the word is sown. When they hear, im\-me\-di\-ate\-ly the temp\-ter co\-mes, and takes a\-way the word sown in them.
    \textsuperscript{16}\thinspace{}These are the ones sown on rock\-y ground. When they hear the word, they im\-me\-di\-ate\-ly re\-ceive it with joy,
    \textsuperscript{17}\thinspace{}but they have no root in them\-selves, and en\-dure brief\-ly. When trou\-ble or per\-se\-cu\-tion co\-mes be\-cause of the word, they im\-me\-di\-ate\-ly fall a\-way.
    \textsuperscript{18}\thinspace{}These are the ones sown a\-mong thorns. When they hear the word,
    \textsuperscript{19}\thinspace{}the worries of the world, the de\-ceit\-ful\-ness of rich\-es, and the de\-sire for oth\-er things en\-ter in and choke the word, and it becomes un\-fruit\-ful.
    \textsuperscript{20}\thinspace{}These are the ones sown on good soil. They hear the word, ac\-cept it, and bear fruit: thir\-ty, six\-ty, and a hun\-dred\-fold.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{4:21}\thinspace{}He said to them, “Is a lamp brought in to be put un\-der a bas\-ket, or un\-der a bed, and not on a stand?
    \textsuperscript{22}\thinspace{}Noth\-ing is hid\-den that will not be revealed, and noth\-ing is se\-cret that will not come to light.”
\end{para}

\begin{para}
    \textsuperscript{4:24}\thinspace{}He said to them, “Pay at\-ten\-tion to what you hear. With the meas\-ure you use, it will be meas\-ured to you, and e\-ven more will be added to you.
    \textsuperscript{25}\thinspace{}For who\-ev\-er has will be giv\-en more; and who\-ev\-er does\-n't have, e\-ven what he has will be tak\-en a\-way.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{4:30}\thinspace{}He said, “With what can we com\-pare God's king\-dom? What par\-a\-ble can we use to de\-scribe it?
    \textsuperscript{31}\thinspace{}It's like a mus\-tard seed. When sown up\-on the soil, it's smaller than all oth\-er seeds on the ground,
    \textsuperscript{32}\thinspace{}yet when it's sown, it grows up and becomes larg\-er than all the gar\-den plants, and puts out large branches, so the birds of the air can nest in its shade.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{4:36}\thinspace{}Leav\-ing the crowd, they took him with them since he was in the boat; and oth\-er boats were with him.
    \textsuperscript{37}\thinspace{}A great wind\-storm a\-rose, and the waves were break\-ing o\-ver the boat, so that the boat was al\-read\-y fill\-ing up.
\end{para}

\begin{para}
    \textsuperscript{4:38}\thinspace{}He was in the stern, sleep\-ing on the cush\-ion. They woke him up, and said to him, “Teach\-er, don't you care that we're go\-ing to die?”
\end{para}

\begin{para}
    \textsuperscript{4:39}\thinspace{}He woke up, re\-buked the wind, and said to the sea, “Peace. Be still.” The wind stopped, and there was a great calm.
    \textsuperscript{40}\thinspace{}He said to them, “Why are you a\-fraid? Do you still have no faith?”
\end{para}

\begin{para}
    \textsuperscript{4:41}\thinspace{}They were ter\-ri\-fied, and asked each oth\-er, “Who is this, that e\-ven the wind and the sea o\-bey him?”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{5:1}\thinspace{}They came to the oth\-er side of the sea.
    \textsuperscript{2}\thinspace{}As soon as Jesus got out of the boat, a man with an un\-clean spir\-it came out of the tombs and met him.
    \textsuperscript{3}\thinspace{}He lived in the tombs. No one could bind him any\-more, not e\-ven with a chain,
    \textsuperscript{4}\thinspace{}be\-cause he had of\-ten been bound with shackles and chains, but had torn the chains a\-part, and broke the shackles in pieces. No one was strong e\-nough to sub\-due him.
    \textsuperscript{5}\thinspace{}Night and day, in the tombs and in the mountains, he was al\-ways cry\-ing out, and cut\-ting him\-self with stones.
\end{para}

\begin{para}
    \textsuperscript{5:6}\thinspace{}When he saw Jesus from a dis\-tance, he ran and knelt down in front of him.
    \textsuperscript{7}\thinspace{}He cried out with a loud voice, “What do you want with me, Jesus, you son of the most high God? I beg you by God, not to tor\-ment me.”
    \textsuperscript{8}\thinspace{}For he had said to him, “Come out of the man, you un\-clean spir\-it.”
\end{para}

\begin{para}
    \textsuperscript{5:11}\thinspace{}There was a large herd of pigs on the hill\-side, feed\-ing.
    \textsuperscript{13}\thinspace{}So he gave the spirits per\-mis\-sion to en\-ter the pigs. The un\-clean spirits came out, and entered the pigs. The herd of a\-bout two thou\-sand rushed down the steep bank in\-to the sea and drowned.
\end{para}

\begin{para}
    \textsuperscript{5:14}\thinspace{}The herdsmen fled and told it in the cit\-y and the coun\-try\-side. Peo\-ple came to see what had happened.
    \textsuperscript{15}\thinspace{}They came to Jesus, and saw him who had been pos\-sessed with the demons, sit\-ting, clothed, and in his right mind; and they were a\-fraid.
    \textsuperscript{16}\thinspace{}Those who had seen it de\-scribed to them what had happened to the demon-possessed man and the pigs.
    \textsuperscript{17}\thinspace{}They be\-gan to beg Jesus to leave their re\-gion.
\end{para}

\begin{para}
    \textsuperscript{5:18}\thinspace{}As he was entering the boat, the man who had been pos\-sessed with demons begged to go with him.
    \textsuperscript{19}\thinspace{}He did not let him, but said to him, “Go home to your own peo\-ple, and tell them how much the Lord has done for you, and how he had mer\-cy on you.”
    \textsuperscript{20}\thinspace{}He went a\-way and be\-gan to pro\-claim how much Jesus had done for him. Eve\-ry\-one was a\-mazed.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{5:21}\thinspace{}When Jesus had crossed a\-gain by boat to the oth\-er side, a large crowd gathered a\-round him, and he was by the sea.
    \textsuperscript{22}\thinspace{}One of the syn\-a\-gogue leaders, named Jairus, came. See\-ing Jesus, he fell at his feet,
    \textsuperscript{23}\thinspace{}begging him ear\-nest\-ly, “My lit\-tle daugh\-ter is dy\-ing. Come and lay your hands on her, so that she will get well, and live.”
\end{para}

\begin{para}
    \textsuperscript{5:24}\thinspace{}He went with him. A large crowd followed him, and pressed a\-round him.
    \textsuperscript{25}\thinspace{}A wom\-an, bleed\-ing for twelve years,
    \textsuperscript{26}\thinspace{}had suffered much un\-der man\-y doctors, and had spent all she had, and was no bet\-ter, but rath\-er grew worse.
    \textsuperscript{27}\thinspace{}Hav\-ing heard a\-bout Jesus, she came up be\-hind him in the crowd, and touched his cloak.
    \textsuperscript{28}\thinspace{}For she said, “If I can just tou\-ch his clothes, I'll get well.”
    \textsuperscript{29}\thinspace{}Im\-me\-di\-ate\-ly, the flow of blood dried up. She felt in her bod\-y that she was healed of her af\-flic\-tion.
\end{para}

\begin{para}
    \textsuperscript{5:30}\thinspace{}Jesus re\-al\-ized at once that pow\-er had gone out from him, and turned a\-round in the crowd, and said, “Who touched my clothes?”
\end{para}

\begin{para}
    \textsuperscript{5:31}\thinspace{}His disciples said to him, “You see the crowd press\-ing a\-round you, and yet you ask, ‘Who touched me?’”
\end{para}

\begin{para}
    \textsuperscript{5:32}\thinspace{}He looked a\-round to see who had done it.
    \textsuperscript{33}\thinspace{}The wom\-an, fearing and trem\-bling, know\-ing what had happened to her, came and fell down in front of him, and told him the whole truth.
    \textsuperscript{34}\thinspace{}He said to her, “Daugh\-ter, your faith has made you well. Go in peace, and be healed of your af\-flic\-tion.”
\end{para}

\begin{para}
    \textsuperscript{5:35}\thinspace{}While he was still speak\-ing, peo\-ple came from Jairus's house and said, “Your daugh\-ter is dead. Why both\-er the teach\-er any\-more?”
\end{para}

\begin{para}
    \textsuperscript{5:36}\thinspace{}Jesus o\-ver\-heard what they said, and told Jairus, “Don't be a\-fraid, just be\-lieve.”
\end{para}

\begin{para}
    \textsuperscript{5:37}\thinspace{}He al\-lowed no one to fol\-low him ex\-cept Pe\-ter, James, and John.
    \textsuperscript{38}\thinspace{}When they came to Jairus's house, he saw a com\-mo\-tion, and peo\-ple weep\-ing and wailing loud\-ly.
    \textsuperscript{39}\thinspace{}He went in\-side, and asked them, “Why are you mak\-ing a com\-mo\-tion, and weep\-ing? The child is\-n't dead but a\-sleep.”
\end{para}

\begin{para}
    \textsuperscript{5:40}\thinspace{}They laughed at him, but he sent them all out\-side. He took the child's fa\-ther and moth\-er and those who were with him, and went in to where the child was.
    \textsuperscript{41}\thinspace{}Tak\-ing her by the hand, he said to her, “Lit\-tle girl, I say to you, get up.”
    \textsuperscript{42}\thinspace{}Im\-me\-di\-ate\-ly the girl got up and be\-gan walk\-ing. They were com\-plete\-ly a\-mazed.
    \textsuperscript{43}\thinspace{}He gave them strict orders that no one should know a\-bout this, and told them to give her some\-thing to eat.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{6:1}\thinspace{}He left there, and went to his home\-town, and his disciples followed him.
    \textsuperscript{2}\thinspace{}When the Sabbath came, he be\-gan to teach in the syn\-a\-gogue. Man\-y who heard him were astonished, say\-ing, “Where did this man get these things? What is this wis\-dom giv\-en to him? How are such miracles done by his hands?
    \textsuperscript{3}\thinspace{}Is\-n't this the car\-pen\-ter, the son of Mar\-y, and broth\-er of James, and Joses, and Ju\-das, and Simon? Aren't his sisters here with us?” And they took of\-fense at him.
\end{para}

\begin{para}
    \textsuperscript{6:4}\thinspace{}Jesus said to them, “A proph\-et is with\-out hon\-or on\-ly in his home\-town, a\-mong his relatives, and in his own house\-hold.”
    \textsuperscript{5}\thinspace{}He could do no mir\-a\-cle there, ex\-cept to lay his hands on a few sick peo\-ple, and heal them.
    \textsuperscript{6}\thinspace{}He was a\-mazed at their un\-be\-lief.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{6:6}\thinspace{}He went a\-round the villages teach\-ing.
    \textsuperscript{7}\thinspace{}He summoned the twelve, and be\-gan to send them out in pairs. He gave them au\-thor\-i\-ty o\-ver un\-clean spirits.
    \textsuperscript{8}\thinspace{}He instructed them to take noth\-ing for their jour\-ney ex\-cept a staff: no bread, no bag, no mon\-ey in their belts;
    \textsuperscript{9}\thinspace{}but to wear sandals and not a sec\-ond shirt.
\end{para}

\begin{para}
    \textsuperscript{6:12}\thinspace{}They went out, and preached that peo\-ple should re\-pent.
    \textsuperscript{13}\thinspace{}They cast out man\-y demons, anointed man\-y sick peo\-ple with oil, and healed them.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{6:14}\thinspace{}King Her\-od heard a\-bout this be\-cause Jesus's name had be\-come well known.
    \textsuperscript{16}\thinspace{}When Her\-od heard it, he said, “John, whom I beheaded, has ris\-en.”
    \textsuperscript{17}\thinspace{}For Her\-od had mar\-ried Herodias, his brother's wife. He had ordered John to be arrested and bound in pris\-on for her sa\-ke.
    \textsuperscript{18}\thinspace{}For John had been tell\-ing Her\-od, “It's not law\-ful for you to have your brother's wife.”
    \textsuperscript{19}\thinspace{}So Herodias had a grudge a\-gainst him, and wanted to kill him. But she could not,
    \textsuperscript{20}\thinspace{}be\-cause Her\-od feared John, and kept him safe, know\-ing that he was a right\-eous and ho\-ly man. When he heard him, he was great\-ly per\-plexed, yet he enjoyed listening to him.
\end{para}

\begin{para}
    \textsuperscript{6:21}\thinspace{}An op\-por\-tu\-ni\-ty came on Herod's birth\-day. He hosted a ban\-quet for his nobles, the mil\-i\-tar\-y commanders, and the lead\-ing men of Gal\-i\-lee.
    \textsuperscript{22}\thinspace{}When Herodias's daugh\-ter came in and danced, she pleased Her\-od and his guests. The king said to the girl, “Ask me for what\-ev\-er you want, and I'll give it to you.”
\end{para}

\begin{para}
    \textsuperscript{6:25}\thinspace{}She said, “I want you to give me John the Baptist's head on a plat\-ter right now.”
\end{para}

\begin{para}
    \textsuperscript{6:26}\thinspace{}The king was great\-ly dis\-tressed, but be\-cause of his vow and his guests he did not want to ref\-use her.
    \textsuperscript{27}\thinspace{}Im\-me\-di\-ate\-ly the king sent an ex\-e\-cu\-tion\-er who went and beheaded him in the pris\-on,
    \textsuperscript{28}\thinspace{}brought his head on a plat\-ter, and gave it to the girl, and the girl gave it to her moth\-er.
\end{para}

\begin{para}
    \textsuperscript{6:29}\thinspace{}When his disciples heard a\-bout it, they came and took his bod\-y, and laid it in a tomb.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{6:32}\thinspace{}They went a\-way in the boat by them\-selves to a de\-sert\-ed place.
    \textsuperscript{33}\thinspace{}Man\-y saw them go\-ing and rec\-og\-nized them, and ran on foot from all the towns and ar\-rived a\-head of them.
    \textsuperscript{34}\thinspace{}When he went a\-shore, he saw a large crowd, and had com\-pas\-sion on them, be\-cause they were like sheep with\-out a shep\-herd. He be\-gan to teach them man\-y things.
\end{para}

\begin{para}
    \textsuperscript{6:35}\thinspace{}When it grew late, his disciples came to him, and said, “This is a des\-o\-late place, and it's get\-ting late.
    \textsuperscript{36}\thinspace{}Send them a\-way to go in\-to the sur\-round\-ing coun\-try\-side and villages to buy them\-selves some\-thing to eat.”
\end{para}

\begin{para}
    \textsuperscript{6:37}\thinspace{}He answered, “You give them some\-thing to eat.” They asked him, “Should we go and buy two hun\-dred de\-nar\-i\-i worth of bread, and give it to them to eat?”
\end{para}

\begin{para}
    \textsuperscript{6:38}\thinspace{}He said, “How man\-y loaves do you have? Go and see.” Once they had found out, they said, “Five, and two fish.”
\end{para}

\begin{para}
    \textsuperscript{6:39}\thinspace{}He commanded them all to sit down in groups on the green grass.
    \textsuperscript{40}\thinspace{}They sat down in groups of hundreds and fifties.
\end{para}

\begin{para}
    \textsuperscript{6:41}\thinspace{}He took the five loaves and the two fish. Looking up to heav\-en, he bless\-ed and broke the loaves. He gave them to his disciples to set be\-fore the peo\-ple, and di\-vid\-ed the two fish a\-mong them all.
    \textsuperscript{42}\thinspace{}Eve\-ry\-one ate and was sat\-is\-fied.
    \textsuperscript{43}\thinspace{}They pick\-ed up twelve baskets full of bro\-ken pieces and of fish.
    \textsuperscript{44}\thinspace{}There were five thou\-sand men who ate the loaves.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{7:1}\thinspace{}The Pharisees, and some of the scribes who had come from Je\-ru\-sa\-lem, gathered a\-round him,
    \textsuperscript{2}\thinspace{}and saw some of his disciples eat\-ing their bread with un\-washed hands.
    \textsuperscript{3}\thinspace{}The Pharisees on\-ly eat if they have care\-ful\-ly washed their hands, keep\-ing the tra\-di\-tion of the elders.
    \textsuperscript{4}\thinspace{}When they come from the mar\-ket\-place, they on\-ly eat af\-ter they have washed. They keep man\-y oth\-er traditions such as the wash\-ing of cups, pitchers, cop\-per pots, and din\-ing couches.
\end{para}

\begin{para}
    \textsuperscript{7:5}\thinspace{}The Pharisees and the scribes asked him, “Why don't your disciples keep the tra\-di\-tion of the elders? In\-stead, they eat with un\-washed hands.”
\end{para}

\begin{para}
    \textsuperscript{7:6}\thinspace{}He replied, “Isaiah prophesied cor\-rect\-ly a\-bout you hypocrites, as it is writ\-ten: ‘This peo\-ple hon\-ors me with their lips, but their heart is far from me.
    \textsuperscript{7}\thinspace{}They wor\-ship me in vain, teach\-ing as doctrines the commandments of men.’
    \textsuperscript{8}\thinspace{}Neglecting the com\-mand\-ment of God, you hold on to the tra\-di\-tion of men.
    \textsuperscript{9}\thinspace{}You have a fi\-ne way of rejecting the com\-mand\-ment of God in or\-der to keep your tra\-di\-tion.
    \textsuperscript{13}\thinspace{}The tra\-di\-tion that you have hand\-ed down nullifies the word of God. You do man\-y things like that.”
\end{para}

\begin{para}
    \textsuperscript{7:14}\thinspace{}He summoned the crowd a\-gain, and said to them, “Eve\-ry\-one lis\-ten to me and un\-der\-stand:
    \textsuperscript{15}\thinspace{}Noth\-ing that enters a per\-son from out\-side can de\-file him, but the things that come out of a per\-son are what de\-file him.”
\end{para}

\begin{para}
    \textsuperscript{7:17}\thinspace{}When he entered the house a\-way from the crowd, his disciples asked him a\-bout the par\-a\-ble.
    \textsuperscript{18}\thinspace{}He said to them, “Are you al\-so with\-out un\-der\-stand\-ing? Do you not see that what\-ev\-er goes in\-to a per\-son from out\-side can\-not de\-file him,
    \textsuperscript{19}\thinspace{}be\-cause it does\-n't go in\-to his heart, but in\-to his stom\-ach and is e\-lim\-i\-nat\-ed?” Thus he de\-clared all foods clean.
\end{para}

\begin{para}
    \textsuperscript{7:20}\thinspace{}He said, “What co\-mes out of a per\-son is what defiles him.
    \textsuperscript{21}\thinspace{}For from with\-in, out of a person's heart, come evil thoughts, sex\-u\-al im\-mo\-ral\-i\-ty, theft, mur\-der, a\-dul\-ter\-y,
    \textsuperscript{22}\thinspace{}coveting, wick\-ed\-ness, de\-ceit, sen\-su\-al\-i\-ty, en\-vy, slan\-der, pride, fool\-ish\-ness.
    \textsuperscript{23}\thinspace{}All these evil things pro\-ceed from with\-in, and de\-file a per\-son.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{8:11}\thinspace{}The Pharisees came and be\-gan to ar\-gue with him, de\-mand\-ing of him a sign from heav\-en to test him.
    \textsuperscript{12}\thinspace{}He sighed deep\-ly in his spir\-it, and said, “Why does this gen\-er\-a\-tion de\-mand a sign? No sign will be giv\-en to this gen\-er\-a\-tion.”
    \textsuperscript{13}\thinspace{}He left them, got back in\-to the boat, and went to the oth\-er side.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{8:27}\thinspace{}Jesus and his disciples went to the villages of Caesarea Phi\-lip\-pi. On the way he asked his disciples, “Who do peo\-ple say that I am?”
\end{para}

\begin{para}
    \textsuperscript{8:28}\thinspace{}They told him, “John the Baptist; others say Elijah; but others, one of the prophets.”
\end{para}

\begin{para}
    \textsuperscript{8:29}\thinspace{}He asked them, “But who do you say that I am?” Pe\-ter answered, “You are the Christ.”
\end{para}

\begin{para}
    \textsuperscript{8:30}\thinspace{}He warned them to tell no one a\-bout him.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{8:31}\thinspace{}He be\-gan to teach them that the son of man must suf\-fer man\-y things, and be rejected by the elders, chief priests, and scribes, and be killed, and af\-ter three days rise a\-gain.
    \textsuperscript{32}\thinspace{}He spoke openly a\-bout this. Pe\-ter took him a\-side, and be\-gan to re\-buke him.
\end{para}

\begin{para}
    \textsuperscript{8:33}\thinspace{}Turn\-ing and see\-ing his disciples, he re\-buked Pe\-ter, and said, “Get be\-hind me, Temp\-ter. For you're not think\-ing a\-bout the things of God, but the things of man.”
\end{para}

\begin{para}
    \textsuperscript{8:34}\thinspace{}He summoned the crowd with his disciples, and said to them, “If any\-one wants to come af\-ter me, let him de\-ny him\-self, take up his cross, and fol\-low me.
    \textsuperscript{35}\thinspace{}For who\-ev\-er wants to save his life will lose it, but who\-ev\-er loses his life for my sa\-ke and the gospel's will save it.
    \textsuperscript{36}\thinspace{}What does it prof\-it a man to gain the whole world, and for\-feit his soul?
    \textsuperscript{37}\thinspace{}What can a man give in ex\-change for his soul?
    \textsuperscript{38}\thinspace{}Who\-ev\-er is a\-shamed of me and my words in this a\-dul\-ter\-ous and sin\-ful gen\-er\-a\-tion, the son of man will al\-so be a\-shamed of him when he co\-mes in his father's glo\-ry with the ho\-ly angels.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{9:2}\thinspace{}Af\-ter six days Jesus took Pe\-ter, James, and John with him, and led them up a high moun\-tain by them\-selves. He was transfigured in front of them,
    \textsuperscript{3}\thinspace{}and his clothes be\-came ra\-di\-ant, in\-tense\-ly white, as no one on earth could whit\-en them.
    \textsuperscript{4}\thinspace{}Elijah and Mo\-ses appeared to them, and they were talking with Jesus.
\end{para}

\begin{para}
    \textsuperscript{9:7}\thinspace{}A cloud overshadowed them, and a voice from the cloud said, “This is my be\-lov\-ed son. Lis\-ten to him.”
    \textsuperscript{8}\thinspace{}Sud\-den\-ly, looking a\-round, they no long\-er saw any\-one with them ex\-cept Jesus.
\end{para}

\begin{para}
    \textsuperscript{9:9}\thinspace{}As they were com\-ing down the moun\-tain, he ordered them to tell no one what they had seen, un\-til the son of man had ris\-en from the dead.
    \textsuperscript{10}\thinspace{}They kept it to them\-selves, discussing what it meant to rise from the dead.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{9:14}\thinspace{}When they came to the disciples, they saw a large crowd a\-round them, and scribes ar\-gu\-ing with them.
    \textsuperscript{15}\thinspace{}As soon as the crowd saw him, they were a\-mazed, and ran to greet him.
    \textsuperscript{16}\thinspace{}He asked them, “What are you ar\-gu\-ing with them a\-bout?”
\end{para}

\begin{para}
    \textsuperscript{9:17}\thinspace{}Some\-one in the crowd answered him, “Teach\-er, I brought my son to you. He has a spir\-it that makes him mute.
    \textsuperscript{18}\thinspace{}When\-ev\-er it seizes him, it throws him down. He foams at the mouth, grinds his teeth, and becomes rig\-id. I asked your disciples to cast it out, but they could\-n't.”
\end{para}

\begin{para}
    \textsuperscript{9:19}\thinspace{}He answered them, “You un\-be\-liev\-ing gen\-er\-a\-tion, how long will I be with you? How long will I put up with you? Bring him to me.”
\end{para}

\begin{para}
    \textsuperscript{9:20}\thinspace{}They brought the boy to him. When the spir\-it saw him, it im\-me\-di\-ate\-ly threw the boy in\-to a con\-vul\-sion. He fell to the ground, and rolled a\-round, foaming at the mouth.
\end{para}

\begin{para}
    \textsuperscript{9:21}\thinspace{}Jesus asked his fa\-ther, “How long has this been hap\-pen\-ing to him?” And he said, “From child\-hood.
    \textsuperscript{22}\thinspace{}It has of\-ten thrown him in\-to fire or wa\-ter to de\-stroy him. If you can do any\-thing, have com\-pas\-sion on us, and help us.”
\end{para}

\begin{para}
    \textsuperscript{9:23}\thinspace{}Jesus said to him, “‘If you can?’ Eve\-ry\-thing is pos\-si\-ble for one who believes.”
\end{para}

\begin{para}
    \textsuperscript{9:24}\thinspace{}Im\-me\-di\-ate\-ly the boy's fa\-ther cried out, “I do be\-lieve. Help my un\-be\-lief.”
\end{para}

\begin{para}
    \textsuperscript{9:25}\thinspace{}When Jesus saw that a crowd was rap\-id\-ly gath\-er\-ing, he re\-buked the un\-clean spir\-it, say\-ing to it, “You mute and deaf spir\-it, I com\-mand you: come out of him and nev\-er en\-ter him a\-gain.”
\end{para}

\begin{para}
    \textsuperscript{9:26}\thinspace{}Af\-ter cry\-ing out, and throwing him in\-to ter\-ri\-ble convulsions, it came out. The boy was like a corpse, so that most of them said, “He's dead.”
    \textsuperscript{27}\thinspace{}But Jesus took him by the hand, and lifted him up, and he got up.
\end{para}

\begin{para}
    \textsuperscript{9:28}\thinspace{}Af\-ter he entered the house, his disciples asked him pri\-vate\-ly, “Why could\-n't we cast it out?”
\end{para}

\begin{para}
    \textsuperscript{9:29}\thinspace{}He said to them, “This kind on\-ly co\-mes out by pray\-er.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{9:30}\thinspace{}They left that place, and passed through Gal\-i\-lee. He did not want any\-one to know,
    \textsuperscript{31}\thinspace{}be\-cause he was teach\-ing his disciples, say\-ing to them, “The son of man will be delivered in\-to the hands of men, and they will kill him. Af\-ter he's killed, he'll rise three days lat\-er.”
    \textsuperscript{32}\thinspace{}But they did not un\-der\-stand this state\-ment, and were a\-fraid to ask him.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{9:33}\thinspace{}They came to Capernaum. He asked them, “What were you discussing on the way?”
    \textsuperscript{34}\thinspace{}But they were si\-lent, be\-cause they had been ar\-gu\-ing a\-bout which of them was the great\-est.
    \textsuperscript{35}\thinspace{}He sat down, called the twelve, and said to them, “If any\-one wants to be first, he must be last of all, and serv\-ant of all.”
\end{para}

\begin{para}
    \textsuperscript{9:36}\thinspace{}He took a child, and had him stand a\-mong them. Tak\-ing him in his arms, he said to them,
    \textsuperscript{37}\thinspace{}“Who\-ev\-er welcomes one child like this in my name welcomes me. Who\-ev\-er welcomes me, does\-n't wel\-come me, but him that sent me.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{9:38}\thinspace{}John said to him, “Teach\-er, we saw some\-one cast\-ing out demons in your name, and we tried to stop him, be\-cause he did\-n't fol\-low us.”
\end{para}

\begin{para}
    \textsuperscript{9:39}\thinspace{}But Jesus said, “Don't stop him. No one can do a mir\-a\-cle in my name, then soon af\-ter\-ward speak evil of me.
    \textsuperscript{40}\thinspace{}Who\-ev\-er is not a\-gainst us is for us.
    \textsuperscript{41}\thinspace{}Tru\-ly I tell you, who\-ev\-er gives you a cup of wa\-ter to drink, be\-cause you are Christ's, will nev\-er lose his re\-ward.”
\end{para}

\begin{para}
    \textsuperscript{9:42}\thinspace{}“Who\-ev\-er causes one of these lit\-tle ones who be\-lieve in me to stum\-ble, it would be bet\-ter for him if a heav\-y mill\-stone were hung a\-round his neck, and he were thrown in\-to the sea.
    \textsuperscript{43}\thinspace{}If your hand causes you to stum\-ble, cut it off. It's bet\-ter for you to en\-ter life crippled than to go to hell with two hands.
    \textsuperscript{45}\thinspace{}If your foot causes you to stum\-ble, cut it off. It's bet\-ter for you to en\-ter life lame than to be thrown in\-to hell with two feet.
    \textsuperscript{47}\thinspace{}If your eye causes you to stum\-ble, tear it out. It's bet\-ter for you to en\-ter God's king\-dom with one eye than to be thrown in\-to hell with two eyes.”
\end{para}

\begin{para}
    \textsuperscript{9:49}\thinspace{}“For eve\-ry\-one will be salt\-ed with fire.
    \textsuperscript{50}\thinspace{}Salt is good, but if the salt becomes un\-salt\-y, how will you make it salt\-y a\-gain? Have salt in your\-selves, and be at peace with one an\-oth\-er.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{10:1}\thinspace{}He left there and went in\-to the re\-gion of Judea and be\-yond the Jor\-dan. Crowds gathered a\-round him a\-gain. A\-gain he taught them, as was his cus\-tom.
\end{para}

\begin{para}
    \textsuperscript{10:2}\thinspace{}Pharisees came to him. Testing him, they asked, “Is it law\-ful for a man to di\-vorce a wife?”
\end{para}

\begin{para}
    \textsuperscript{10:3}\thinspace{}He answered them, “What did Mo\-ses com\-mand you?”
\end{para}

\begin{para}
    \textsuperscript{10:4}\thinspace{}They said, “Mo\-ses al\-lowed a man to ‘write a cer\-tif\-i\-cate of di\-vorce, and send her a\-way.’”
\end{para}

\begin{para}
    \textsuperscript{10:5}\thinspace{}But Jesus said to them, “He wrote you this com\-mand\-ment be\-cause of your hard hearts.
    \textsuperscript{6}\thinspace{}But from the be\-gin\-ning of cre\-a\-tion, ‘God made them male and fe\-male.’
    \textsuperscript{7}\thinspace{}‘There\-fore a man will leave his fa\-ther and moth\-er,
    \textsuperscript{8}\thinspace{}and the two will be\-come one flesh.’ So they are no long\-er two, but one flesh.
    \textsuperscript{9}\thinspace{}There\-fore what God has joined to\-geth\-er, let no one sep\-a\-rate.”
\end{para}

\begin{para}
    \textsuperscript{10:10}\thinspace{}In the house the disciples asked him a\-bout this a\-gain.
    \textsuperscript{11}\thinspace{}He said to them, “Who\-ev\-er divorces his wife, and marries an\-oth\-er, commits a\-dul\-ter\-y a\-gainst her.
    \textsuperscript{12}\thinspace{}If she divorces her hus\-band, and marries an\-oth\-er, she commits a\-dul\-ter\-y.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{10:13}\thinspace{}Peo\-ple were bring\-ing chil\-dren to him, so that he might tou\-ch them, but the disciples re\-buked them.
    \textsuperscript{14}\thinspace{}When Jesus saw it, he was in\-dig\-nant, and said to them, “Let the chil\-dren come to me. Don't hind\-er them, for God's king\-dom belongs to such as these.
    \textsuperscript{15}\thinspace{}Tru\-ly I tell you, who\-ev\-er does not re\-ceive God's king\-dom like a child, will nev\-er en\-ter it.”
    \textsuperscript{16}\thinspace{}He took them in his arms, laid his hands on them, and bless\-ed them.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{10:17}\thinspace{}As he was set\-ting out on his jour\-ney, a man ran up to him, knelt in front of him, and asked him, “Good teach\-er, what must I do to in\-her\-it e\-ter\-nal life?”
\end{para}

\begin{para}
    \textsuperscript{10:18}\thinspace{}Jesus said to him, “Why do you call me good? No one is good ex\-cept God a\-lone.
    \textsuperscript{19}\thinspace{}You know the commandments: ‘Do not mur\-der, Do not com\-mit a\-dul\-ter\-y, Do not steal, Do not bear false wit\-ness, Do not de\-fraud, Hon\-or your fa\-ther and moth\-er.’”
\end{para}

\begin{para}
    \textsuperscript{10:20}\thinspace{}He said to him, “I've kept all these from my youth.”
\end{para}

\begin{para}
    \textsuperscript{10:21}\thinspace{}Looking at him, Jesus loved him, and said to him, “You lack one thing: go, sell all you have, give to the poor, and you will have treas\-ure in heav\-en. Then come, fol\-low me.”
\end{para}

\begin{para}
    \textsuperscript{10:22}\thinspace{}But he was saddened by these words, and he went a\-way griev\-ing, for he had man\-y possessions.
\end{para}

\begin{para}
    \textsuperscript{10:23}\thinspace{}Jesus looked a\-round, and said to his disciples, “How hard it will be for those who have wealth to en\-ter God's king\-dom.”
    \textsuperscript{24}\thinspace{}The disciples were a\-mazed at his words. But Jesus said to them a\-gain, “Chil\-dren, how hard it is to en\-ter God's king\-dom.
    \textsuperscript{25}\thinspace{}It's eas\-i\-er for a cam\-el to go through the eye of a nee\-dle than for a rich per\-son to en\-ter God's king\-dom.”
\end{para}

\begin{para}
    \textsuperscript{10:26}\thinspace{}They were e\-ven more astonished, and asked him, “Then who can be saved?”
\end{para}

\begin{para}
    \textsuperscript{10:27}\thinspace{}Looking at them, Jesus said, “With man it's im\-pos\-si\-ble, but not with God. Be\-cause all things are pos\-si\-ble with God.”
\end{para}

\begin{para}
    \textsuperscript{10:28}\thinspace{}Pe\-ter be\-gan to tell him, “We've left eve\-ry\-thing, and followed you.”
\end{para}

\begin{para}
    \textsuperscript{10:29}\thinspace{}Jesus said, “Any\-one who has left house or brothers or sisters or moth\-er or fa\-ther or chil\-dren or fields, for my sa\-ke, and for the gos\-pel,
    \textsuperscript{30}\thinspace{}will re\-ceive a hun\-dred\-fold now in this time, hous\-es and brothers and sisters and mothers and chil\-dren and fields, with persecutions, and e\-ter\-nal life in the age to come.
    \textsuperscript{31}\thinspace{}But man\-y who are first will be last, and the last first.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{10:32}\thinspace{}They were on the road, go\-ing up to Je\-ru\-sa\-lem, and Jesus was walk\-ing a\-head of them. He took the twelve a\-gain, and be\-gan to tell them what was go\-ing to hap\-pen to him,
    \textsuperscript{33}\thinspace{}“We're go\-ing up to Je\-ru\-sa\-lem, and the son of man will be delivered to the chief priests and the scribes. They'll con\-demn him to death, and de\-liv\-er him to the Gentiles.
    \textsuperscript{34}\thinspace{}They'll mock him, spit on him, scourge him, and kill him. He'll rise af\-ter three days.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{10:42}\thinspace{}Jesus called them o\-ver, and said to them, “You know that those who are con\-sid\-ered rulers of the Gentiles lord it o\-ver them, and their great ones ex\-er\-cise au\-thor\-i\-ty o\-ver them.
    \textsuperscript{43}\thinspace{}But it is not this way a\-mong you. In\-stead, who\-ev\-er wants to be\-come great a\-mong you must be your serv\-ant,
    \textsuperscript{44}\thinspace{}and who\-ev\-er wants to be first a\-mong you must be slave of all.
    \textsuperscript{45}\thinspace{}For e\-ven the son of man did\-n't come to be served, but to serve, and to give his life as a ran\-som for man\-y.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{10:46}\thinspace{}As he was leav\-ing Jericho with his disciples and a large crowd, a blind beg\-gar was sit\-ting by the road.
    \textsuperscript{47}\thinspace{}When he heard that it was Jesus of Na\-za\-reth, he be\-gan to cry out, “Jesus, son of David, have mer\-cy on me.”
\end{para}

\begin{para}
    \textsuperscript{10:48}\thinspace{}Man\-y sternly told him to be si\-lent, but he cried out e\-ven louder, “Son of David, have mer\-cy on me.”
\end{para}

\begin{para}
    \textsuperscript{10:49}\thinspace{}Jesus stopped and said, “Call him.” They called the blind man, say\-ing to him, “Take cour\-age. Get up. He's call\-ing for you.”
    \textsuperscript{50}\thinspace{}Throwing a\-side his cloak, he jumped up, and came to Jesus.
\end{para}

\begin{para}
    \textsuperscript{10:51}\thinspace{}Jesus asked him, “What do you want me to do for you?” The blind man said to him, “Rab\-bi, I want to re\-gain my sight.”
\end{para}

\begin{para}
    \textsuperscript{10:52}\thinspace{}Jesus said to him, “Go, your faith has made you well.” Im\-me\-di\-ate\-ly he regained his sight, and be\-gan fol\-low\-ing him on the road.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{11:1}\thinspace{}As they approached Je\-ru\-sa\-lem, near the Mount of Olives, he sent two of his disciples,
    \textsuperscript{2}\thinspace{}and told them, “Go in\-to the vil\-lage a\-head of you. As soon as you en\-ter it, you'll find a colt tied, on which no one has ev\-er sat. Un\-tie it, and bring it.
    \textsuperscript{3}\thinspace{}If any\-one asks you, ‘Why are you do\-ing this?’ say, ‘The Lord needs it and will send it back im\-me\-di\-ate\-ly.’”
\end{para}

\begin{para}
    \textsuperscript{11:4}\thinspace{}They went, found a colt out\-side in the street tied by a door, and un\-tied it.
    \textsuperscript{5}\thinspace{}Some of the bystanders asked them, “Why are you un\-ty\-ing the colt?”
    \textsuperscript{6}\thinspace{}They told them what Jesus had said, and they gave them per\-mis\-sion.
    \textsuperscript{7}\thinspace{}They brought the colt to Jesus, and put their coats on it. He sat on it,
    \textsuperscript{11}\thinspace{}and went in\-to Je\-ru\-sa\-lem, and in\-to the tem\-ple. Af\-ter looking a\-round at eve\-ry\-thing, he left for Bethany with the twelve, since it was al\-read\-y late.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{11:15}\thinspace{}They came to Je\-ru\-sa\-lem. He entered the tem\-ple, and be\-gan to drive out those buy\-ing and sell\-ing in the tem\-ple, and overturned the tables of the mon\-ey changers and the seats of those sell\-ing doves.
    \textsuperscript{16}\thinspace{}He would not al\-low any\-one to car\-ry mer\-chan\-dise through the tem\-ple.
    \textsuperscript{17}\thinspace{}He taught, and said to them, “Is\-n't it writ\-ten, ‘My house will be called a house of pray\-er for all the nations’? But you have made it ‘a den of robbers.’”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{11:27}\thinspace{}As he was walk\-ing in the tem\-ple, the chief priests, the scribes, and the elders came to him.
    \textsuperscript{28}\thinspace{}They asked him, “By what au\-thor\-i\-ty are you do\-ing these things?”
\end{para}

\begin{para}
    \textsuperscript{11:29}\thinspace{}Jesus said to them, “I will ask you one ques\-tion. An\-swer me, and I will tell you by what au\-thor\-i\-ty I do these things.
    \textsuperscript{30}\thinspace{}Was John's bap\-tism from heav\-en or from men?”
\end{para}

\begin{para}
    \textsuperscript{11:31}\thinspace{}They discussed it with one an\-oth\-er, say\-ing, “If we say, ‘From heav\-en,’ he will say, ‘Then why did\-n't you be\-lieve him?’
    \textsuperscript{32}\thinspace{}But should we say, ‘From men’?” They feared the peo\-ple, for eve\-ry\-one thought that John re\-al\-ly was a proph\-et.
\end{para}

\begin{para}
    \textsuperscript{11:33}\thinspace{}They answered Jesus, “We don't know.” Jesus said to them, “Nei\-ther will I tell you by what au\-thor\-i\-ty I do these things.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{12:1}\thinspace{}He be\-gan to speak to them in parables. “A man planted a vine\-yard, put a fence a\-round it, dug a pit for a wine\-press, and built a tow\-er. Then he leased it to tenants, and went on a jour\-ney.
    \textsuperscript{2}\thinspace{}At har\-vest time, he sent a serv\-ant to the tenants to col\-lect some of the fruit of the vine\-yard from them.
    \textsuperscript{3}\thinspace{}They took him, beat him, and sent him a\-way empty-handed.
    \textsuperscript{4}\thinspace{}A\-gain he sent them an\-oth\-er serv\-ant. They hit him on the head, and treated him shamefully.
    \textsuperscript{5}\thinspace{}He sent an\-oth\-er, and they killed that one. He sent man\-y others; they beat some, and killed others.
    \textsuperscript{6}\thinspace{}He had one more to send, a be\-lov\-ed son. Fi\-nal\-ly he sent him to them, say\-ing, ‘They will re\-spect my son.’”
\end{para}

\begin{para}
    \textsuperscript{12:7}\thinspace{}“But those tenants said to one an\-oth\-er, ‘This is the heir. Come, let's kill him, and the in\-her\-it\-ance will be ours.’
    \textsuperscript{8}\thinspace{}They took him, killed him, and threw him out of the vine\-yard.”
\end{para}

\begin{para}
    \textsuperscript{12:9}\thinspace{}“What will the vineyard's own\-er do? He will come and de\-stroy the tenants, and give the vine\-yard to others.
    \textsuperscript{10}\thinspace{}Have\-n't you read this scrip\-ture: ‘The stone that the builders rejected has be\-come the cor\-ner\-stone.’”
\end{para}

\begin{para}
    \textsuperscript{12:12}\thinspace{}They were seek\-ing to ar\-rest him, be\-cause they un\-der\-stood that he spoke the par\-a\-ble a\-gainst them, but they feared the crowd. So they left him, and went a\-way.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{12:13}\thinspace{}They sent some of the Pharisees and Herodians to him, so that they might trap him in his words.
    \textsuperscript{14}\thinspace{}When they came, they said to him, “Teach\-er, we know that you're truth\-ful. You don't care what any\-one thinks, be\-cause you're im\-par\-tial. You truthfully teach the way of God. Is it law\-ful to pay taxes to Caesar, or not?”
    \textsuperscript{15}\thinspace{}Know\-ing their hy\-poc\-ri\-sy, he said to them, “Why are you testing me? Bring me a de\-nar\-i\-us to look at.”
\end{para}

\begin{para}
    \textsuperscript{12:16}\thinspace{}They brought one. He asked them, “Whose im\-age and in\-scrip\-tion is this?” They answered, “Caesar's.”
    \textsuperscript{17}\thinspace{}Jesus said to them, “Give to Caesar the things that are Caesar's, and to God the things that are God's.” They mar\-veled at him.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{12:18}\thinspace{}Some Sadducees, who say there is no res\-ur\-rec\-tion, came to Jesus and asked him,
    \textsuperscript{19}\thinspace{}“Teach\-er, Mo\-ses wrote for us that if a man's broth\-er dies, and leaves be\-hind a wife but no child, that his broth\-er must mar\-ry the wid\-ow, and raise up chil\-dren for his broth\-er.
    \textsuperscript{20}\thinspace{}There were sev\-en brothers. The first one took a wife, and died with\-out chil\-dren.
    \textsuperscript{21}\thinspace{}The sec\-ond one mar\-ried her, and died with\-out chil\-dren.
    \textsuperscript{22}\thinspace{}None of the sev\-en left chil\-dren. Last of all, the wom\-an died too.
    \textsuperscript{23}\thinspace{}In the res\-ur\-rec\-tion, when they rise, whose wife will she be? For all sev\-en had mar\-ried her.”
\end{para}

\begin{para}
    \textsuperscript{12:24}\thinspace{}Jesus said to them, “Is\-n't this the rea\-son you are mis\-tak\-en: you don't know the scriptures or the pow\-er of God?
    \textsuperscript{25}\thinspace{}For when the dead rise, they nei\-ther mar\-ry nor are giv\-en in mar\-riage, but are like angels in heav\-en.”
\end{para}

\begin{para}
    \textsuperscript{12:26}\thinspace{}“As for the dead be\-ing raised, have\-n't you read in the book of Mo\-ses, in the pas\-sage a\-bout the burn\-ing bush, how God said to him, ‘I am the God of Abraham, and the God of Isaac, and the God of Jacob’?
    \textsuperscript{27}\thinspace{}He is\-n't God of the dead, but of the liv\-ing. You're great\-ly mis\-tak\-en.”
\end{para}

\begin{para}
    \textsuperscript{12:28}\thinspace{}One of the scribes came, and heard them ar\-gu\-ing. See\-ing that he answered them well, he asked him, “Which com\-mand\-ment is the most im\-por\-tant?”
\end{para}

\begin{para}
    \textsuperscript{12:29}\thinspace{}Jesus answered, “The most im\-por\-tant is,
    \textsuperscript{30}\thinspace{}‘Love the Lord your God with all your heart, soul, mind, and strength.’
    \textsuperscript{31}\thinspace{}The sec\-ond is this, ‘Love your neigh\-bor as your\-self.’ There is no com\-mand\-ment greater than these.”
\end{para}

\begin{para}
    \textsuperscript{12:32}\thinspace{}The scr\-ibe said to him, “You stat\-ed cor\-rect\-ly that
    \textsuperscript{33}\thinspace{}to love him with all the heart, un\-der\-stand\-ing, and strength, and to love one's neigh\-bor as him\-self, is more im\-por\-tant than all burnt offerings and sacrifices.”
\end{para}

\begin{para}
    \textsuperscript{12:34}\thinspace{}When Jesus saw that he answered wisely, he said to him, “You aren't far from God's king\-dom.” Af\-ter that, no one dared to ask him any more questions.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{12:38}\thinspace{}In his teach\-ing he said, “Be\-ware of the scribes, who like walk\-ing a\-round in long robes, be\-ing greeted in the marketplaces,
    \textsuperscript{39}\thinspace{}the best seats in the synagogues, and places of hon\-or at banquets.
    \textsuperscript{40}\thinspace{}They de\-vour widows' hous\-es, and of\-fer long prayers for show. They will re\-ceive greater con\-dem\-na\-tion.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{13:1}\thinspace{}As he came out of the tem\-ple, one of his disciples said to him, “Teach\-er, look at these won\-der\-ful stones and buildings!”
\end{para}

\begin{para}
    \textsuperscript{13:2}\thinspace{}Jesus said to him, “Do you see these great buildings? Not one stone will be left up\-on an\-oth\-er; all will be thrown down.”
\end{para}

\begin{para}
    \textsuperscript{13:3}\thinspace{}As he sat on the Mount of Olives op\-po\-site the tem\-ple, Pe\-ter, James, John and Andrew asked him pri\-vate\-ly,
    \textsuperscript{4}\thinspace{}“When will these things hap\-pen? What will be the sign that they are a\-bout to be ful\-filled?”
\end{para}

\begin{para}
    \textsuperscript{13:5}\thinspace{}Jesus said to them, “See that no one leads you a\-stray.
    \textsuperscript{6}\thinspace{}Man\-y will come in my name, say\-ing, ‘I am he,’ and will mis\-lead man\-y.
    \textsuperscript{12}\thinspace{}Broth\-er will be\-tray broth\-er to death, and a fa\-ther his child. Chil\-dren will rise a\-gainst parents, and have them put to death.
    \textsuperscript{13}\thinspace{}Eve\-ry\-one will hate you be\-cause of my name, but he who endures to the end will be saved.”
\end{para}

\begin{para}
    \textsuperscript{13:31}\thinspace{}“Heav\-en and earth will pass a\-way, but my words won't pass a\-way.
    \textsuperscript{32}\thinspace{}No one knows a\-bout that day or that hour, not e\-ven the angels in heav\-en, nor the son, but on\-ly the fa\-ther.
    \textsuperscript{33}\thinspace{}Be on guard. Be a\-lert. For you don't know when the time will come.”
\end{para}

\begin{para}
    \textsuperscript{13:34}\thinspace{}“It's like a man on a jour\-ney: when he leaves home, he puts his servants in charge, each with a task, and commands the door\-keep\-er to be a\-lert.
    \textsuperscript{35}\thinspace{}There\-fore be a\-lert. You don't know when the mas\-ter of the house will come. It could be in the eve\-ning, at mid\-night, when the roost\-er crows, or in the morn\-ing.
    \textsuperscript{36}\thinspace{}Oth\-er\-wise, he may come sud\-den\-ly and find you a\-sleep.
    \textsuperscript{37}\thinspace{}What I say to you, I say to eve\-ry\-one: Be a\-lert.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{14:1}\thinspace{}The Passover and the Feast of Un\-leav\-ened Bread were now two days a\-way. The chief priests and the scribes sought a way to ar\-rest him by stealth, and kill him.
    \textsuperscript{2}\thinspace{}But they said, “Not dur\-ing the feast, oth\-er\-wise the peo\-ple might ri\-ot.”
\end{para}

\begin{para}
    \textsuperscript{14:3}\thinspace{}While he was in Bethany in the house of Simon the lep\-er, and re\-clin\-ing at the ta\-ble, a wom\-an came with an al\-a\-bas\-ter jar of ver\-y ex\-pen\-sive per\-fume. She broke the jar, and poured it o\-ver his head.
    \textsuperscript{4}\thinspace{}Some expressed in\-dig\-na\-tion to one an\-oth\-er, say\-ing, “Why has this per\-fume been wast\-ed?
    \textsuperscript{5}\thinspace{}This per\-fume might have been sold for more than three hun\-dred de\-nar\-i\-i, and the mon\-ey giv\-en to the poor.” And they scolded her.
\end{para}

\begin{para}
    \textsuperscript{14:6}\thinspace{}Jesus said, “Leave her a\-lone. Why are you bothering her? She's done a beau\-ti\-ful thing for me.
    \textsuperscript{7}\thinspace{}You al\-ways have the poor with you. When\-ev\-er you want, you can do good for them, but you won't al\-ways have me.
    \textsuperscript{8}\thinspace{}She has done what she could. She has anointed my bod\-y be\-fore\-hand for the bur\-i\-al.
    \textsuperscript{9}\thinspace{}Tru\-ly I tell you, wher\-ev\-er the gos\-pel is preached in the whole world, what she has done will be told in mem\-o\-ry of her.”
\end{para}

\begin{para}
    \textsuperscript{14:10}\thinspace{}Then Ju\-das Iscariot, one of the twelve, went to the chief priests to be\-tray him to them.
    \textsuperscript{11}\thinspace{}They were glad when they heard this, and prom\-ised to give him mon\-ey. He be\-gan looking for an op\-por\-tu\-ni\-ty to be\-tray him.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{14:12}\thinspace{}On the first day of Un\-leav\-ened Bread, when the Passover lamb was be\-ing sacrificed, his disciples asked him, “Where do you want us to go to pre\-pare for you to eat the Passover?”
\end{para}

\begin{para}
    \textsuperscript{14:13}\thinspace{}He sent two of his disciples, and told them, “Go in\-to the cit\-y. A man car\-ry\-ing a pitch\-er of wa\-ter will meet you. Fol\-low him.
    \textsuperscript{14}\thinspace{}Wher\-ev\-er he enters, say to the own\-er of the house, ‘The teach\-er says, Where is my guest room, where I may eat the passover with my disciples?’
    \textsuperscript{15}\thinspace{}He will show you a large up\-per room furnished and read\-y. Pre\-pare for us there.”
    \textsuperscript{16}\thinspace{}The disciples went out, entered the cit\-y, and found it just as he had told them. They pre\-pared the Passover.
\end{para}

\begin{para}
    \textsuperscript{14:17}\thinspace{}When it was eve\-ning, he ar\-rived with the twelve.
    \textsuperscript{18}\thinspace{}While they were re\-clin\-ing at ta\-ble and eat\-ing, Jesus said, “Tru\-ly I tell you, one of you will be\-tray me, one who is eat\-ing with me.”
\end{para}

\begin{para}
    \textsuperscript{14:19}\thinspace{}They be\-gan to be grieved, and to say to him one by one, “Sure\-ly not I?”
\end{para}

\begin{para}
    \textsuperscript{14:20}\thinspace{}He said to them, “It's one of the twelve, one who is dip\-ping bread in\-to the bowl with me.
    \textsuperscript{21}\thinspace{}For the son of man goes as it is writ\-ten of him, but woe to that man by whom the son of man is betrayed! It would have been bet\-ter for that man if he had not been born.”
\end{para}

\begin{para}
    \textsuperscript{14:22}\thinspace{}As they were eat\-ing, he took bread, bless\-ed it, broke it, gave it to them, and said, “Take it. This is my bod\-y.”
    \textsuperscript{23}\thinspace{}He took a cup, and af\-ter giv\-ing thanks, he gave it to them, and they all drank from it.
    \textsuperscript{24}\thinspace{}He said to them, “This is my blood of the cov\-e\-nant, which is poured out for man\-y.
    \textsuperscript{25}\thinspace{}Tru\-ly I tell you, I won't drink of the fruit of the vine a\-gain un\-til that day when I drink it new in God's king\-dom.”
\end{para}

\begin{para}
    \textsuperscript{14:26}\thinspace{}Af\-ter sing\-ing a hymn, they went out to the Mount of Olives.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{14:27}\thinspace{}Jesus said to them, “You will all fall a\-way, be\-cause it is writ\-ten, ‘I will strike down the shep\-herd, and the sheep will be scattered.’
    \textsuperscript{28}\thinspace{}But af\-ter I'm raised up, I'll go a\-head of you to Gal\-i\-lee.”
\end{para}

\begin{para}
    \textsuperscript{14:29}\thinspace{}Pe\-ter said to him, “E\-ven if eve\-ry\-one falls a\-way, I won't.”
\end{para}

\begin{para}
    \textsuperscript{14:30}\thinspace{}Jesus said to him, “Tru\-ly I tell you, that this ver\-y night, be\-fore a roost\-er crows twice, you will de\-ny me three times.”
\end{para}

\begin{para}
    \textsuperscript{14:31}\thinspace{}But he kept insisting, “If I have to die with you, I won't de\-ny you.” They all said the same thing.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{14:32}\thinspace{}They went to a place called Gethsemane. He told his disciples, “Sit here while I pray.”
    \textsuperscript{33}\thinspace{}He took Pe\-ter, James and John with him, and be\-gan to be ver\-y dis\-tressed and troubled.
    \textsuperscript{34}\thinspace{}He told them, “My soul is deep\-ly grieved to the point of death. Stay here, and keep watch.”
\end{para}

\begin{para}
    \textsuperscript{14:35}\thinspace{}He went a lit\-tle far\-ther, fell to the ground, and prayed that the hour might pass from him.
    \textsuperscript{36}\thinspace{}He said, “Fa\-ther, eve\-ry\-thing is pos\-si\-ble for you. Re\-move this cup from me. Yet not what I will, but what you will.”
\end{para}

\begin{para}
    \textsuperscript{14:37}\thinspace{}He came, found them sleep\-ing, and said to Pe\-ter, “Simon, are you a\-sleep? Could\-n't you keep watch for one hour?
    \textsuperscript{38}\thinspace{}Keep watch and pray, that you won't come in\-to temp\-ta\-tion. The spir\-it is will\-ing, but the flesh is weak.”
\end{para}

\begin{para}
    \textsuperscript{14:39}\thinspace{}A\-gain he went a\-way, and prayed, say\-ing the same words.
    \textsuperscript{40}\thinspace{}A\-gain he came, and found them sleep\-ing, be\-cause their eyes were ver\-y heav\-y. They did\-n't know what to say to him.
\end{para}

\begin{para}
    \textsuperscript{14:41}\thinspace{}He came the third time, and said to them, “Are you still sleep\-ing and rest\-ing? It's e\-nough. The time has come. See, the son of man is be\-ing betrayed in\-to the hands of sinners.
    \textsuperscript{42}\thinspace{}Get up. Let's go. See, my be\-tray\-er is near.”
\end{para}

\begin{para}
    \textsuperscript{14:43}\thinspace{}While he was still speak\-ing, Ju\-das, one of the twelve, ar\-rived. With him was a crowd, armed with swords and clubs, who was sent from the chief priests, the scribes and the elders.
    \textsuperscript{44}\thinspace{}The be\-tray\-er had giv\-en them a sign, say\-ing, “The one I kiss is the one. Ar\-rest him, and lead him a\-way un\-der guard.”
    \textsuperscript{45}\thinspace{}Go\-ing di\-rect\-ly to Jesus, he said, “Rab\-bi!” and kissed him.
\end{para}

\begin{para}
    \textsuperscript{14:46}\thinspace{}They laid hands on him, and arrested him.
    \textsuperscript{47}\thinspace{}One of the bystanders drew his sword, and struck the high priest's serv\-ant, cut\-ting off his ear.
\end{para}

\begin{para}
    \textsuperscript{14:48}\thinspace{}Jesus said to them, “Have you come out with swords and clubs to ar\-rest me, as you would ar\-rest a rob\-ber?
    \textsuperscript{49}\thinspace{}I was with you eve\-ry day, teach\-ing in the tem\-ple, and you did\-n't ar\-rest me.”
\end{para}

\begin{para}
    \textsuperscript{14:50}\thinspace{}They all left him, and fled.
    \textsuperscript{51}\thinspace{}A young man was fol\-low\-ing him, wear\-ing on\-ly a lin\-en cloth a\-round his na\-ked bod\-y. They seized him,
    \textsuperscript{52}\thinspace{}but he left the lin\-en cloth, and fled na\-ked.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{14:53}\thinspace{}They led Jesus to the high priest. All the chief priests, the elders and the scribes as\-sem\-bled.
    \textsuperscript{54}\thinspace{}Pe\-ter followed him at a dis\-tance, right in\-to the high priest's court\-yard. He was sit\-ting with the guards, and warming him\-self by the fire.
\end{para}

\begin{para}
    \textsuperscript{14:55}\thinspace{}The chief priests and the whole coun\-cil sought tes\-ti\-mo\-ny a\-gainst Jesus to put him to death, but found none.
    \textsuperscript{56}\thinspace{}Man\-y gave false tes\-ti\-mo\-ny a\-gainst him, but their tes\-ti\-mo\-ny was in\-con\-sist\-ent.
    \textsuperscript{57}\thinspace{}Some stood up, and gave false tes\-ti\-mo\-ny a\-gainst him, say\-ing,
    \textsuperscript{58}\thinspace{}“We heard him say, ‘I'll de\-stroy this tem\-ple made with hands, and in three days I'll build an\-oth\-er made with\-out hands.’”
    \textsuperscript{59}\thinspace{}But e\-ven this tes\-ti\-mo\-ny was in\-con\-sist\-ent.
\end{para}

\begin{para}
    \textsuperscript{14:60}\thinspace{}The high priest stood up in front of them, and asked Jesus, “Don't you an\-swer? What is it that these men tes\-ti\-fy a\-gainst you?”
    \textsuperscript{61}\thinspace{}But he kept si\-lent, and did not an\-swer. A\-gain the high priest asked him, “Are you the Christ, the son of the Bless\-ed One?”
\end{para}

\begin{para}
    \textsuperscript{14:62}\thinspace{}Jesus said, “I am. You'll see the son of man sit\-ting at the right hand of pow\-er, and com\-ing with the clouds of heav\-en.”
\end{para}

\begin{para}
    \textsuperscript{14:63}\thinspace{}The high priest tore his clothes, and said, “Why do we need any more witnesses?
    \textsuperscript{64}\thinspace{}You've heard the blas\-phe\-my. What's your de\-ci\-sion?” They all condemned him as de\-serv\-ing death.
\end{para}

\begin{para}
    \textsuperscript{14:65}\thinspace{}Some be\-gan to spit on him, and to cov\-er his face and strike him, say\-ing, “Proph\-e\-sy.” The guards re\-ceived him with slaps in the face.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{14:66}\thinspace{}While Pe\-ter was in the court\-yard be\-low, one of the high priest's serv\-ant girls came.
    \textsuperscript{67}\thinspace{}See\-ing Pe\-ter, she looked at him, and said, “You al\-so were with Jesus the Nazarene.”
\end{para}

\begin{para}
    \textsuperscript{14:68}\thinspace{}But he de\-nied it, say\-ing, “I don't know or un\-der\-stand what you're talking a\-bout.” He went out to the en\-try\-way, and a roost\-er crowed.
\end{para}

\begin{para}
    \textsuperscript{14:69}\thinspace{}The serv\-ant girl saw him, and be\-gan a\-gain to tell the bystanders, “This man is one of them.”
\end{para}

\begin{para}
    \textsuperscript{14:70}\thinspace{}But a\-gain he de\-nied it. Af\-ter a lit\-tle while the bystanders a\-gain said to Pe\-ter, “You're cer\-tain\-ly one of them, since you're a Galilean too.”
    \textsuperscript{71}\thinspace{}But he be\-gan to curse and swear, “I don't know this man you're talking a\-bout.”
\end{para}

\begin{para}
    \textsuperscript{14:72}\thinspace{}Im\-me\-di\-ate\-ly a roost\-er crowed a sec\-ond time. Pe\-ter remembered when Jesus said to him, “Be\-fore a roost\-er crows twice, you'll de\-ny me three times”. He be\-gan to weep.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{15:1}\thinspace{}Ear\-ly in the morn\-ing, the chief priests, elders, scribes, and the whole coun\-cil held a con\-sul\-ta\-tion. They bound Jesus, led him a\-way, and delivered him to Pilate.
\end{para}

\begin{para}
    \textsuperscript{15:2}\thinspace{}Pilate asked him, “Are you the king of the Jews?” He answered, “You have said so.”
\end{para}

\begin{para}
    \textsuperscript{15:3}\thinspace{}The chief priests ac\-cused him of man\-y things.
    \textsuperscript{4}\thinspace{}Pilate a\-gain asked him, “Don't you an\-swer? Look how man\-y charges they bring a\-gainst you.”
    \textsuperscript{5}\thinspace{}But Jesus answered no fur\-ther, so Pilate was a\-mazed.
\end{para}

\begin{para}
    \textsuperscript{15:6}\thinspace{}Now at the feast he used to re\-lease for them any one pris\-on\-er whom they requested.
    \textsuperscript{7}\thinspace{}The man named Barabbas was in pris\-on with rebels who had com\-mit\-ted mur\-der dur\-ing the in\-sur\-rec\-tion.
    \textsuperscript{8}\thinspace{}The crowd be\-gan to ask Pilate to do for them as he u\-su\-al\-ly did.
\end{para}

\begin{para}
    \textsuperscript{15:9}\thinspace{}Pilate answered them, say\-ing, “Do you want me to re\-lease the king of the Jews for you?”
    \textsuperscript{10}\thinspace{}For he knew that the chief priests had hand\-ed him o\-ver be\-cause of en\-vy.
    \textsuperscript{11}\thinspace{}But the chief priests stirred up the crowd to have him re\-lease Barabbas in\-stead.
    \textsuperscript{12}\thinspace{}Pilate a\-gain asked them, “Then what should I do with the one you call the king of the Jews?”
\end{para}

\begin{para}
    \textsuperscript{15:13}\thinspace{}They shouted back, “Cru\-ci\-fy him!”
\end{para}

\begin{para}
    \textsuperscript{15:14}\thinspace{}Pilate asked them, “Why? What evil has he done?” But they shouted all the more, “Cru\-ci\-fy him!”
\end{para}

\begin{para}
    \textsuperscript{15:15}\thinspace{}Wishing to sat\-is\-fy the crowd, Pilate re\-leased Barabbas for them. Af\-ter hav\-ing Jesus flogged, he hand\-ed him o\-ver to be cru\-ci\-fied.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{15:21}\thinspace{}Simon of Cyrene was pass\-ing by, com\-ing from the coun\-try. They forced him to car\-ry Jesus's cross.
    \textsuperscript{22}\thinspace{}They brought Jesus to the place called Gol\-go\-tha, which means place of a skull.
    \textsuperscript{23}\thinspace{}They offered him wine mixed with myrrh, but he did not take it.
\end{para}

\begin{para}
    \textsuperscript{15:24}\thinspace{}They cru\-ci\-fied him, and di\-vid\-ed his clothes a\-mong them\-selves, cast\-ing lots to de\-cide what each should take.
    \textsuperscript{25}\thinspace{}It was nine in the morn\-ing when they cru\-ci\-fied him.
    \textsuperscript{26}\thinspace{}The in\-scrip\-tion of the charge a\-gainst him read, “The King of the Jews.”
    \textsuperscript{27}\thinspace{}They cru\-ci\-fied two robbers with him, one on his right and one on his left.
\end{para}

\begin{para}
    \textsuperscript{15:29}\thinspace{}Those who passed by de\-rid\-ed him, shak\-ing their heads, and say\-ing, “Ha! You who would de\-stroy the tem\-ple and re\-build it in three days,
    \textsuperscript{30}\thinspace{}save your\-self, and come down from the cross.”
\end{para}

\begin{para}
    \textsuperscript{15:31}\thinspace{}In the same way, the chief priests and the scribes mocked him to one an\-oth\-er say\-ing, “He saved others. He can't save him\-self.
    \textsuperscript{32}\thinspace{}Let this Christ, the King of Israel, come down now from the cross, so that we may see and be\-lieve.” Those who were cru\-ci\-fied with him al\-so insulted him.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{15:33}\thinspace{}At noon, dark\-ness fell o\-ver the whole land un\-til three in the af\-ter\-noon.
    \textsuperscript{34}\thinspace{}At three in the af\-ter\-noon Jesus cried with a loud voice, “My God, my God, why have you a\-ban\-doned me?”
\end{para}

\begin{para}
    \textsuperscript{15:37}\thinspace{}Jesus let out a loud cry, and breathed his last.
    \textsuperscript{38}\thinspace{}The cur\-tain of the tem\-ple was torn in two, from the top to the bot\-tom.
\end{para}

\begin{para}
    \textsuperscript{15:39}\thinspace{}When the cen\-tu\-ri\-on who stood fac\-ing him saw the way he breathed his last, he said, “Tru\-ly this man was the son of God.”
\end{para}

\begin{para}
    \textsuperscript{15:40}\thinspace{}There were al\-so wom\-en watching from a dis\-tance. A\-mong them were Mar\-y Mag\-da\-lene, Salome, and Mar\-y the moth\-er of James the young\-er and of Joses.
    \textsuperscript{41}\thinspace{}When he was in Gal\-i\-lee, they followed him and took care of him. Man\-y oth\-er wom\-en had come up with him to Je\-ru\-sa\-lem.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{15:42}\thinspace{}When eve\-ning had come, be\-cause it was the day be\-fore the Sabbath,
    \textsuperscript{43}\thinspace{}Jo\-seph of Arimathea came, a prom\-i\-nent mem\-ber of the coun\-cil, who was him\-self waiting for God's king\-dom. He bold\-ly went to Pilate, and asked for Jesus's bod\-y.
    \textsuperscript{44}\thinspace{}Pilate was sur\-prised that he was al\-read\-y dead.
    \textsuperscript{45}\thinspace{}He grant\-ed the bod\-y to Jo\-seph.
\end{para}

\begin{para}
    \textsuperscript{15:46}\thinspace{}Jo\-seph bought a lin\-en cloth, took him down, wrapped him in the lin\-en cloth, laid him in a tomb cut out of a rock, and rolled a stone a\-gainst the en\-trance of the tomb.
    \textsuperscript{47}\thinspace{}Mar\-y Mag\-da\-lene and Mar\-y the moth\-er of Joses were watching to see where he was laid.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{16:1}\thinspace{}When the Sabbath was past, Mar\-y Mag\-da\-lene, Salome, and Mar\-y the moth\-er of James bought spices, so that they could go and a\-noint him.
    \textsuperscript{2}\thinspace{}Ver\-y ear\-ly on the first day of the week, when the sun had ris\-en, they went to the tomb.
    \textsuperscript{3}\thinspace{}They were say\-ing to one an\-oth\-er, “Who will roll a\-way the stone from the en\-trance of the tomb for us?”
    \textsuperscript{4}\thinspace{}Looking up, they saw that the stone had al\-read\-y been rolled a\-way, al\-though it was ex\-treme\-ly large.
\end{para}

\begin{para}
    \textsuperscript{16:5}\thinspace{}Entering the tomb, they saw a young man sit\-ting at the right, wear\-ing a white robe, and they were alarmed.
    \textsuperscript{6}\thinspace{}He told them, “Don't be alarmed. You're looking for Jesus the Nazarene, who was cru\-ci\-fied. He has ris\-en. He's not here. See, here is the place where they laid him.
    \textsuperscript{7}\thinspace{}But go, tell his disciples and Pe\-ter, ‘He's go\-ing a\-head of you to Gal\-i\-lee. You'll see him there, just as he told you.’”
\end{para}

\begin{para}
    \textsuperscript{16:8}\thinspace{}They went out, and fled from the tomb, for trem\-bling and as\-ton\-ish\-ment had seized them. They said noth\-ing to any\-one, be\-cause they were a\-fraid.
\end{para}

\chapter*{Mormon}
\markboth{Mormon}{Mormon}
\addcontentsline{toc}{section}{Mormon}\begin{changemargin}{.6in}{.6in}
  \begin{center}
    \itshape
    An a\-bridg\-ment of the Book of Mormon, with an in\-tro\-duc\-tion a\-bridged from Jo\-seph Smith---His\-to\-ry.
  \end{center}
\end{changemargin}
\vspace{0pt plus 8\baselineskip{}}\section*{Introduction}
\markright{Introduction}
\begin{para}
    \textsuperscript{1:1}\thinspace{}I, Jo\-seph Smith, write this his\-to\-ry to put all inquirers af\-ter truth in pos\-ses\-sion of the facts.
\end{para}

\begin{para}
    \textsuperscript{1:5}\thinspace{}There was in the place where I lived an un\-u\-su\-al ex\-cite\-ment on the sub\-ject of re\-li\-gion.
    \textsuperscript{6}\thinspace{}The cler\-gy pro\-mot\-ed this scene of re\-li\-gious feel\-ing in or\-der to have eve\-ry\-bod\-y con\-vert\-ed, let them join what sect they pleased. Yet when the converts be\-gan to file off, some to one par\-ty and some to an\-oth\-er, it was seen that the seemingly good feelings of both the priests and the converts were more pre\-tend\-ed than re\-al, lost in a strife of words: priest contending a\-gainst priest, and con\-vert a\-gainst con\-vert.
\end{para}

\begin{para}
    \textsuperscript{1:7}\thinspace{}I was at this time four\-teen years old.
    \textsuperscript{8}\thinspace{}My mind was called up to se\-ri\-ous re\-flec\-tion and great un\-eas\-i\-ness. My feelings were deep and of\-ten poign\-ant, but still I kept my\-self a\-loof from all the parties, though I attended their meetings as of\-ten as oc\-ca\-sion would per\-mit. So great were the con\-fu\-sion and strife a\-mong the dif\-fer\-ent denominations, that it was im\-pos\-si\-ble for a per\-son as young as I was to come to any cer\-tain con\-clu\-sion who was right and who was wrong.
\end{para}

\begin{para}
    \textsuperscript{1:10}\thinspace{}In the midst of this war of words and tu\-mult of opinions, I of\-ten said to my\-self, “What is to be done? If any of these parties is right, which is it, and how will I know it?”
\end{para}

\begin{para}
    \textsuperscript{1:11}\thinspace{}One day I was read\-ing James 1:5, which reads, “If any of you lacks wis\-dom, let him ask God, who gives gen\-er\-ous\-ly to all with\-out re\-proach, and it will be giv\-en to him.”
    \textsuperscript{12}\thinspace{}This pas\-sage entered with great force in\-to eve\-ry feel\-ing of my heart. I reflected on it a\-gain and a\-gain, know\-ing that if any per\-son needed wis\-dom from God, I did; for the dif\-fer\-ent sects un\-der\-stood the same passages of scrip\-ture so dif\-fer\-ent\-ly as to de\-stroy all con\-fi\-dence in set\-tling the ques\-tion by an ap\-peal to the Bible.
\end{para}

\begin{para}
    \textsuperscript{1:13}\thinspace{}At length I came to the de\-ter\-mi\-na\-tion to “ask God.”
    \textsuperscript{14}\thinspace{}So, in ac\-cord\-ance with my de\-ter\-mi\-na\-tion, I re\-tired to the woods to make the at\-tempt. It was the first time in my life that I had made the at\-tempt to pray vo\-cal\-ly.
\end{para}

\begin{para}
    \textsuperscript{1:15}\thinspace{}Hav\-ing looked a\-round me, and find\-ing my\-self a\-lone, I kneeled down and be\-gan to of\-fer up the desires of my heart to God. I had scarce\-ly done so when
    \textsuperscript{16}\thinspace{}I saw a pil\-lar of light o\-ver my head.
    \textsuperscript{17}\thinspace{}When the light rested up\-on me, I saw two personages. One of them said to me, “Jo\-seph, this is my be\-lov\-ed son. Hear him.”
\end{para}

\begin{para}
    \textsuperscript{1:18}\thinspace{}I asked the personages which of all the sects was right, and which I should join.
    \textsuperscript{19}\thinspace{}I was answered that I must join none of them.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{1:21}\thinspace{}A few days af\-ter I had this vi\-sion, I was with one of the preachers who was ver\-y ac\-tive in the re\-li\-gious ex\-cite\-ment. I gave him an ac\-count of the vi\-sion which I had had. He treated my com\-mu\-ni\-ca\-tion with con\-tempt, say\-ing that there were no such things as visions or revelations in these days.
\end{para}

\begin{para}
    \textsuperscript{1:22}\thinspace{}I soon found that my sto\-ry had ex\-cit\-ed a great deal of prej\-u\-dice a\-gainst me a\-mong the professors of re\-li\-gion. They ex\-cit\-ed the pub\-lic mind a\-gainst me and cre\-at\-ed a bit\-ter per\-se\-cu\-tion. This was com\-mon a\-mong all the sects,
    \textsuperscript{23}\thinspace{}and it of\-ten caused me great sor\-row.
\end{para}

\begin{para}
    \textsuperscript{1:24}\thinspace{}How\-ev\-er, it was nev\-er\-the\-less a fact that I had seen a vi\-sion, and all the per\-se\-cu\-tion un\-der heav\-en could not make it oth\-er\-wise.
    \textsuperscript{25}\thinspace{}I said in my heart, “Why per\-se\-cute me for tell\-ing the truth? For I have seen a vi\-sion. I know it, and I know that God knows it, and I can not de\-ny it.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{1:27}\thinspace{}I con\-tin\-ued to pur\-sue my com\-mon vocations in life, all the time be\-ing
    \textsuperscript{28}\thinspace{}per\-se\-cut\-ed by those who ought to have been my friends and to have treated me kind\-ly. If they sup\-posed me to be de\-lud\-ed, they ought to have endeavored to re\-claim me in a prop\-er and af\-fec\-tion\-ate man\-ner. I was left to mingling with all kinds of so\-ci\-e\-ty, and fre\-quent\-ly fell in\-to man\-y fool\-ish errors. In mak\-ing this con\-fes\-sion, no one need sup\-pose me guilt\-y of any great sins, but I was guilt\-y of lev\-i\-ty and some\-times as\-so\-ci\-at\-ed with jo\-vi\-al com\-pa\-ny.
\end{para}

\begin{para}
    \textsuperscript{1:29}\thinspace{}In con\-se\-quence of these things, I of\-ten felt condemned for my weaknesses and imperfections. One eve\-ning, I be\-took my\-self to pray\-er and sup\-pli\-ca\-tion to God for for\-give\-ness of my sins and follies.
    \textsuperscript{30}\thinspace{}While I was call\-ing up\-on God, I discovered a light appearing in my room, and im\-me\-di\-ate\-ly a per\-son\-age appeared.
\end{para}

\begin{para}
    \textsuperscript{1:33}\thinspace{}He called me by name, said that he was a mes\-sen\-ger from God, and that his name was Moroni.
    \textsuperscript{34}\thinspace{}He said that there was a book, writ\-ten up\-on gold plates, giv\-ing an ac\-count of the form\-er inhabitants of this con\-ti\-nent.
    \textsuperscript{42}\thinspace{}The vi\-sion was opened to my mind so that I could see the place where the plates were deposited.
    \textsuperscript{46}\thinspace{}He added a cau\-tion, tell\-ing me that Satan would try to tempt me to get the plates for the purposes of get\-ting rich. This he for\-bade me, say\-ing that I must on\-ly get the plates to glo\-ri\-fy God.
    \textsuperscript{53}\thinspace{}The mes\-sen\-ger in\-formed me that the time for bring\-ing forth the plates would ar\-rive in four years.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{1:54}\thinspace{}I went at the end of each year to the place where the plates were deposited. Each time I found the same mes\-sen\-ger there, and re\-ceived in\-struc\-tion from him.
\end{para}

\begin{para}
    \textsuperscript{1:59}\thinspace{}At length, the time ar\-rived for obtaining the plates. Hav\-ing gone as u\-su\-al at the end of an\-oth\-er year to the place where they were deposited, the same mes\-sen\-ger delivered them to me.
    \textsuperscript{67}\thinspace{}I com\-menced to trans\-late the Book of Mormon.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{1 Nephi}
\markright{1 Nephi}
\begin{para}
    \textsuperscript{1:1}\thinspace{}I, Nephi, am mak\-ing a rec\-ord of my proceedings dur\-ing my life\-time.
\end{para}

\begin{para}
    \textsuperscript{1:4}\thinspace{}In the first year of the reign of Zed\-e\-ki\-ah, king of Judah,
    \textsuperscript{18}\thinspace{}God showed my fa\-ther, Lehi, man\-y things a\-bout the de\-struc\-tion of Je\-ru\-sa\-lem. He went a\-mong the peo\-ple, and de\-clared to them what he had seen.
    \textsuperscript{20}\thinspace{}When the peo\-ple heard these things, they were an\-gry with him, and sought his life.
\end{para}

\begin{para}
    \textsuperscript{2:2}\thinspace{}God commanded my fa\-ther in a dream to take his fam\-i\-ly, and de\-part in\-to the wil\-der\-ness.
    \textsuperscript{4}\thinspace{}He left his house, land, gold, sil\-ver, and pre\-cious things, and took noth\-ing with him, ex\-cept his fam\-i\-ly, provisions, and tents, and de\-part\-ed in\-to the wil\-der\-ness.
\end{para}

\begin{para}
    \textsuperscript{2:11}\thinspace{}La\-man and Lem\-uel, my old\-er brothers, grumbled a\-gainst their fa\-ther be\-cause he had led them out of Je\-ru\-sa\-lem, leav\-ing their land, gold, sil\-ver, and pre\-cious things. They said he had done this be\-cause of the fool\-ish imaginations of his heart.
    \textsuperscript{13}\thinspace{}They al\-so did not be\-lieve that Je\-ru\-sa\-lem could be destroyed.
\end{para}

\begin{para}
    \textsuperscript{2:16}\thinspace{}I be\-lieved all that my fa\-ther had spo\-ken. There\-fore, I did not reb\-el a\-gainst him like my brothers.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{3:2}\thinspace{}My fa\-ther said to me,
    \textsuperscript{3}\thinspace{}“Laban has the rec\-ord of the Jews, and it's en\-graved on brass plates.
    \textsuperscript{4}\thinspace{}There\-fore, God has commanded me that you and your brothers should re\-turn to Je\-ru\-sa\-lem, go to Laban's house, seek the records, and bring them here in\-to the wil\-der\-ness.”
\end{para}

\begin{para}
    \textsuperscript{3:7}\thinspace{}I said to my fa\-ther, “I'll go and do what God has commanded, for I know that God gives no commandments un\-less he prepares a way to ac\-com\-plish them.”
\end{para}

\begin{para}
    \textsuperscript{3:19}\thinspace{}I said to my brothers, “It's wis\-dom in God that we should ob\-tain these records, that we may pre\-serve for our chil\-dren
    \textsuperscript{20}\thinspace{}the words spo\-ken by all the ho\-ly prophets.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{3:23}\thinspace{}We went up to the house of Laban.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{4:38}\thinspace{}My brothers and I took the brass plates, de\-part\-ed in\-to the wil\-der\-ness, and jour\-neyed to our father's tent.
\end{para}

\begin{para}
    \textsuperscript{5:1}\thinspace{}Af\-ter we returned to our fa\-ther, he was filled with joy, and al\-so my moth\-er was ver\-y glad.
    \textsuperscript{21}\thinspace{}We had obtained the records which God had commanded us, and found that they had great val\-ue to us, be\-cause we could pre\-serve the commandments of God for our chil\-dren.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{7:6}\thinspace{}As we jour\-neyed in the wil\-der\-ness, La\-man and Lem\-uel
    \textsuperscript{16}\thinspace{}were an\-gry with me, and tried to take a\-way my life,
    \textsuperscript{19}\thinspace{}but some of our com\-pa\-ny plead with them, and they stopped try\-ing to take a\-way my life.
    \textsuperscript{21}\thinspace{}I for\-gave them all that they had done, and we trav\-eled a\-gain on our jour\-ney.
\end{para}

\begin{para}
    \textsuperscript{17:3}\thinspace{}If the chil\-dren of men keep God's commandments, he provides means for them to ac\-com\-plish what he has commanded.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{17:5}\thinspace{}We came to the land which we called Boun\-ti\-ful, be\-cause it had much fruit and wild hon\-ey. God had pre\-pared all these things that we might not per\-ish.
    \textsuperscript{6}\thinspace{}We pitched our tents by the sea\-shore, and re\-joiced great\-ly.
\end{para}

\begin{para}
    \textsuperscript{17:7}\thinspace{}God said to me, Nephi,
    \textsuperscript{8}\thinspace{}“You shall con\-struct a ship af\-ter the man\-ner that I will show you, that I may car\-ry your peo\-ple a\-cross these waters.”
    \textsuperscript{15}\thinspace{}There\-fore, I la\-bored to keep the commandments of God.
\end{para}

\begin{para}
    \textsuperscript{17:17}\thinspace{}When my brothers saw that I was a\-bout to build a ship, they be\-gan to grum\-ble a\-gainst me, say\-ing, “Our broth\-er is a fool, for he thinks that he can build a ship, and cross these great waters.”
\end{para}

\begin{para}
    \textsuperscript{17:19}\thinspace{}I was deep\-ly grieved be\-cause of the hard\-ness of their hearts. When they saw this they were glad, say\-ing, “We knew that you could\-n't con\-struct a ship, and that you lacked judg\-ment.
    \textsuperscript{20}\thinspace{}You're like our fa\-ther, led a\-way by the fool\-ish imaginations of his heart. He has led us out of the land of Je\-ru\-sa\-lem, and we've wandered in the wil\-der\-ness all these years.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{18:3}\thinspace{}I, Nephi, of\-ten went up on the moun\-tain, and I of\-ten prayed to God.
\end{para}

\begin{para}
    \textsuperscript{18:4}\thinspace{}Af\-ter I had fin\-ished the ship, my brothers saw that it was good, and they humbled them\-selves a\-gain be\-fore God.
    \textsuperscript{6}\thinspace{}We pre\-pared man\-y provisions, went in\-to the ship with our wives and chil\-dren,
    \textsuperscript{8}\thinspace{}and put forth in\-to the sea.
\end{para}

\begin{para}
    \textsuperscript{18:23}\thinspace{}Af\-ter we had sailed for man\-y days we ar\-rived at the prom\-ised land.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{19:1}\thinspace{}I made plates of o\-re on which to en\-grave the rec\-ord of my peo\-ple,
    \textsuperscript{5}\thinspace{}to keep the more sa\-cred things for instructing my peo\-ple.
\end{para}

\begin{para}
    \textsuperscript{19:7}\thinspace{}The things which some men es\-teem to be of great worth, both to the bod\-y and soul, others set at naught.
    \textsuperscript{9}\thinspace{}The world, be\-cause of their in\-iq\-ui\-ty, will judge Jesus to be a thing of naught.
    \textsuperscript{18}\thinspace{}I have writ\-ten these things that per\-haps I might per\-suade my peo\-ple to re\-mem\-ber God.
\end{para}

\begin{para}
    \textsuperscript{22:23}\thinspace{}For the time soon co\-mes that all churches which are built up to make a prof\-it, and to get pow\-er, and to be\-come pop\-u\-lar in the eyes of the world, must be brought low in the dust.
\end{para}

\begin{para}
    \textsuperscript{22:31}\thinspace{}If you are o\-be\-di\-ent to the commandments, and en\-dure to the end, you will be saved at the last day.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{2 Nephi}
\markright{2 Nephi}
\begin{para}
    \textsuperscript{4:13}\thinspace{}Not man\-y days af\-ter Lehi's death, La\-man and Lem\-uel were an\-gry with me,
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{5:2}\thinspace{}so they sought to take a\-way my life.
\end{para}

\begin{para}
    \textsuperscript{5:5}\thinspace{}God warned me to de\-part from them, and flee in\-to the wil\-der\-ness a\-long with all those who would go with me.
    \textsuperscript{6}\thinspace{}There\-fore, I took my fam\-i\-ly, and Sam, my old\-er broth\-er, and Jacob and Jo\-seph, my young\-er brothers, and al\-so my sisters, and their families. All those who went with me were those who be\-lieved in the warnings and revelations of God.
\end{para}

\begin{para}
    \textsuperscript{5:7}\thinspace{}We jour\-neyed in the wil\-der\-ness for man\-y days, and pitched our tents.
    \textsuperscript{10}\thinspace{}We were care\-ful to keep the commandments of God in all things.
    \textsuperscript{11}\thinspace{}We prospered great\-ly, for we sowed seed, reaped in a\-bun\-dance, and raised flocks, and herds, and animals of eve\-ry kind.
\end{para}

\begin{para}
    \textsuperscript{5:14}\thinspace{}I made man\-y swords so that our enemies would not come up\-on us and de\-stroy us.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{6:1}\thinspace{}The words of Jacob, the broth\-er of Nephi:
\end{para}

\begin{para}
    \textsuperscript{9:27}\thinspace{}Woe to him who has all the commandments of God, and transgresses them, and wastes the days of his pro\-ba\-tion.
\end{para}

\begin{para}
    \textsuperscript{9:28}\thinspace{}Oh the van\-i\-ty, and fool\-ish\-ness of men! When they are learn\-ed they think they are wise, and they do not lis\-ten to God's coun\-sel, for they set it a\-side, sup\-pos\-ing they know of them\-selves, there\-fore, their wis\-dom is fool\-ish\-ness and it does not prof\-it them.
    \textsuperscript{29}\thinspace{}But to be learn\-ed is good if they lis\-ten to God's counsels.
\end{para}

\begin{para}
    \textsuperscript{9:30}\thinspace{}Woe to the rich, who des\-pise the poor, and per\-se\-cute the meek, and whose hearts are on their treasures; there\-fore, their treas\-ure is their God.
\end{para}

\begin{para}
    \textsuperscript{9:34}\thinspace{}Woe to the li\-ar,
    \textsuperscript{35}\thinspace{}to the mur\-der\-er who kills de\-lib\-er\-ate\-ly,
    \textsuperscript{36}\thinspace{}and to those who com\-mit for\-ni\-ca\-tion.
\end{para}

\begin{para}
    \textsuperscript{9:39}\thinspace{}Re\-mem\-ber, the mind set on the flesh is death, and the mind set on the spir\-it is e\-ter\-nal life.
    \textsuperscript{45}\thinspace{}Turn a\-way from your sins.
    \textsuperscript{46}\thinspace{}Pre\-pare your souls for the day of judg\-ment.
\end{para}

\begin{para}
    \textsuperscript{9:51}\thinspace{}Do not spend mon\-ey on what has no val\-ue, nor your la\-bor on what can\-not sat\-is\-fy.
\end{para}

\begin{para}
    \textsuperscript{10:23}\thinspace{}Cheer up your hearts, and re\-mem\-ber that you are free to act for your\-selves: to choose the way of ev\-er\-last\-ing death or the way of e\-ter\-nal life.
    \textsuperscript{24}\thinspace{}There\-fore, rec\-on\-cile your\-selves to the will of God, and not to the will of the flesh.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{11:1}\thinspace{}Jacob spoke man\-y more things to my peo\-ple, but I, Nephi, have on\-ly writ\-ten these.
\end{para}

\begin{para}
    \textsuperscript{25:23}\thinspace{}We la\-bor dil\-i\-gent\-ly to write, to per\-suade our chil\-dren, and al\-so our brothers, to be\-lieve in Christ, and to be rec\-on\-ciled to God.
    \textsuperscript{26}\thinspace{}We talk of Christ, we re\-joice in Christ, we preach of Christ, we proph\-e\-sy of Christ.
\end{para}

\begin{para}
    \textsuperscript{26:21}\thinspace{}There are man\-y churches built up which cause en\-vy, strife, and mal\-ice.
    \textsuperscript{22}\thinspace{}The dev\-il leads them by the neck with a flax\-en cord un\-til he binds them with his strong cords for\-ev\-er.
\end{para}

\begin{para}
    \textsuperscript{26:23}\thinspace{}God does not work in dark\-ness.
    \textsuperscript{25}\thinspace{}Does he say to any\-one, “De\-part from me?” No.
    \textsuperscript{26}\thinspace{}Has he commanded any\-one to de\-part from the hous\-es of wor\-ship? No.
\end{para}

\begin{para}
    \textsuperscript{26:29}\thinspace{}He commands that there shall be no priestcrafts, which are when men preach and set them\-selves up as a light to the world, to make a prof\-it, and get praise of the world.
    \textsuperscript{30}\thinspace{}God has for\-bid\-den this. There\-fore, he has commanded that all men should have love, and ex\-cept they have love they are noth\-ing.
\end{para}

\begin{para}
    \textsuperscript{26:32}\thinspace{}God has commanded men not to mur\-der, lie, steal, en\-vy, take his name in vain, have mal\-ice, con\-tend with one an\-oth\-er, or com\-mit for\-ni\-ca\-tion.
    \textsuperscript{33}\thinspace{}None of these iniquities come from God, for he does good a\-mong the chil\-dren of men.
\end{para}

\begin{para}
    \textsuperscript{28:3}\thinspace{}One church will say to the oth\-er, “Be\-hold, I, I am God's.” And the oth\-er churches will say, “I, I am God's.” Thus will eve\-ry one say that has built up churches, and not to God.
    \textsuperscript{4}\thinspace{}They will con\-tend with one an\-oth\-er, and their priests will con\-tend with one an\-oth\-er.
\end{para}

\begin{para}
    \textsuperscript{28:8}\thinspace{}Man\-y will say, “Eat, drink, and be mer\-ry, nev\-er\-the\-less fear God. He will jus\-ti\-fy in com\-mit\-ting a lit\-tle sin. Lie a lit\-tle, take ad\-van\-tage of one be\-cause of his words, dig a pit for your neigh\-bor. There is no harm in this. Do all these things, for to\-mor\-row we die. If we are guilt\-y, God will beat us with a few lashes, and at last we will be saved.”
\end{para}

\begin{para}
    \textsuperscript{28:12}\thinspace{}Be\-cause of pride their churches have be\-come corrupted.
    \textsuperscript{13}\thinspace{}They rob the poor be\-cause of their fi\-ne sanc\-tu\-ar\-ies, and their fi\-ne cloth\-ing. They per\-se\-cute the meek and the poor in heart, be\-cause in their pride they have be\-come ar\-ro\-gant.
\end{para}

\begin{para}
    \textsuperscript{28:16}\thinspace{}Woe to those who turn a\-side the just for noth\-ing, and re\-vile a\-gainst what is good, and say that it has no worth!
    \textsuperscript{29}\thinspace{}Woe to him who says, “We have re\-ceived the word of God, and we need no more of the word of God, for we have e\-nough!”
\end{para}

\begin{para}
    \textsuperscript{28:30}\thinspace{}God says, “To him who receives I will give more. From them who say, ‘We have e\-nough,’ will be tak\-en a\-way what\-ev\-er they have.”
\end{para}

\begin{para}
    \textsuperscript{29:3}\thinspace{}“Some will say, ‘We have a Bible, and there can\-not be any more Bible.’
    \textsuperscript{10}\thinspace{}Be\-cause you have a Bible you don't need to sup\-pose that it contains all my words.
    \textsuperscript{11}\thinspace{}For I com\-mand all men to write the words which I speak to them.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{30:17}\thinspace{}There is noth\-ing se\-cret that will not be revealed.
\end{para}

\begin{para}
    \textsuperscript{32:8}\thinspace{}Know that you must pray, for the evil spir\-it teaches a man that he must not pray.
\end{para}

\begin{para}
    \textsuperscript{33:10}\thinspace{}Lis\-ten to these words and be\-lieve in Christ, for they teach all men to do good.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Jacob}
\markright{Jacob}
\begin{para}
    \textsuperscript{1:2}\thinspace{}Nephi gave me, Jacob, a com\-mand\-ment to write a few of the most pre\-cious things.
    \textsuperscript{15}\thinspace{}The peo\-ple be\-gan to grow hard in their hearts, and to in\-dulge some\-what in wick\-ed\-ness,
    \textsuperscript{17}\thinspace{}there\-fore I gave them these words:
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{2:12}\thinspace{}Man\-y of you have be\-gun to search for gold, sil\-ver, and pre\-cious ores,
    \textsuperscript{13}\thinspace{}and prov\-i\-dence has smiled on you most pleas\-ing\-ly, that you have obtained man\-y rich\-es. Be\-cause some of you have obtained more a\-bun\-dant\-ly than others you are lifted up in the pride of your hearts be\-cause of the cost\-li\-ness of your ap\-par\-el, and per\-se\-cute others be\-cause you sup\-pose that you are bet\-ter than they. 
    \textsuperscript{14}\thinspace{}Do you sup\-pose that God justifies you in this thing? No, but he condemns you, and if you per\-sist his judgments will speed\-i\-ly come to you.
\end{para}

\begin{para}
    \textsuperscript{2:17}\thinspace{}Think of others like your\-selves, be fa\-mil\-iar with all, and free with your possessions, that they may be rich like you.
    \textsuperscript{18}\thinspace{}But be\-fore you seek for rich\-es, seek for the king\-dom of God.
    \textsuperscript{19}\thinspace{}Af\-ter you have obtained a hope in Christ you will ob\-tain rich\-es, if you seek them, and you will seek them with the in\-tent to do good: to clothe the na\-ked, to feed the hun\-gry, to lib\-er\-ate the cap\-tive, and to ad\-min\-is\-ter re\-lief to the sick and the afflicted.
    \textsuperscript{21}\thinspace{}Don't you sup\-pose that one be\-ing is as pre\-cious in God's sight as the oth\-er?
\end{para}

\begin{para}
    \textsuperscript{2:35}\thinspace{}You have bro\-ken your ten\-der wives' hearts, and lost your children's con\-fi\-dence, be\-cause of your bad examples be\-fore them. The sob\-bing of their hearts ascends up to God a\-gainst you.
\end{para}

\begin{para}
    \textsuperscript{4:8}\thinspace{}Don't des\-pise God's revelations.
    \textsuperscript{10}\thinspace{}Don't seek to coun\-sel God, but to take coun\-sel from him. For you know that he counsels in wis\-dom, jus\-tice, and great mer\-cy.
\end{para}

\begin{para}
    \textsuperscript{6:12}\thinspace{}Be wise; what more can I say?
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Zeniff}
\markright{Zeniff}
\begin{para}
    \textsuperscript{0}\thinspace{}Tak\-en from chapters 11-23 of the Book of Mo\-si\-ah.
\end{para}

\begin{para}
    \textsuperscript{11:1}\thinspace{}Zeniff con\-ferred the king\-dom on No\-ah, one of his sons.
    \textsuperscript{2}\thinspace{}No\-ah did not keep God's commandments, but followed the desires of his own heart.
    \textsuperscript{3}\thinspace{}He im\-posed a tax of one fifth of all the peo\-ple pos\-sessed
    \textsuperscript{4}\thinspace{}to sup\-port him\-self, and his priests.
\end{para}

\begin{para}
    \textsuperscript{11:20}\thinspace{}There was a man named A\-bin\-a\-di who went a\-mong the peo\-ple say\-ing, “God says, ‘Woe to this peo\-ple, for I have seen their abominations, their wick\-ed\-ness, and their for\-ni\-ca\-tion.
    \textsuperscript{21}\thinspace{}Un\-less they re\-pent they will be brought in\-to bond\-age.’”
\end{para}

\begin{para}
    \textsuperscript{11:27}\thinspace{}When king No\-ah heard of Abinadi's words, he was an\-gry. He said, “Who is A\-bin\-a\-di that he should judge me and my peo\-ple, or who is God to bring such great af\-flic\-tion on my peo\-ple?
    \textsuperscript{28}\thinspace{}Bring A\-bin\-a\-di here that I may kill him.”
\end{para}

\begin{para}
    \textsuperscript{12:9}\thinspace{}They took him and car\-ried him bound be\-fore the king.
    \textsuperscript{29}\thinspace{}A\-bin\-a\-di said to the priests, “Why do you set your hearts on rich\-es? Why do you spend your strength with prostitutes?”
\end{para}

\begin{para}
    \textsuperscript{13:4}\thinspace{}“You're an\-gry with me be\-cause I've told you the truth.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{15:26}\thinspace{}But fear and trem\-ble be\-fore God, for he redeems none who have wil\-ful\-ly rebelled a\-gainst him, who have known his commandments, and re\-fused to keep them.”
\end{para}

\begin{para}
    \textsuperscript{17:1}\thinspace{}The king commanded the priests to take A\-bin\-a\-di and cause him to be put to death.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{17:2}\thinspace{}There was a young man a\-mong the priests whose name was Al\-ma, and he be\-lieved Abinadi's words, there\-fore he be\-gan to plead with the king to let A\-bin\-a\-di de\-part in peace.
    \textsuperscript{3}\thinspace{}But the king was more an\-gry and sent his servants af\-ter Al\-ma to kill him.
    \textsuperscript{4}\thinspace{}But he fled from them and hid. Be\-ing concealed for man\-y days, he wrote all the words which A\-bin\-a\-di had spo\-ken.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{17:7}\thinspace{}The king said, “A\-bin\-a\-di,
    \textsuperscript{8}\thinspace{}you will be put to death un\-less you take back all the evil you've spo\-ken a\-bout me and my peo\-ple.”
\end{para}

\begin{para}
    \textsuperscript{17:9}\thinspace{}A\-bin\-a\-di said to him, “I will not take back what I've spo\-ken to you, for it is true.
    \textsuperscript{10}\thinspace{}It will stand as a tes\-ti\-mo\-ny a\-gainst you at the last day.”
\end{para}

\begin{para}
    \textsuperscript{17:11}\thinspace{}King No\-ah was a\-bout to re\-lease him, for he feared that God's judgments would come up\-on him.
    \textsuperscript{12}\thinspace{}But the priests said, “He has re\-viled the king.” There\-fore the king was in\-cit\-ed to an\-ger a\-gainst him, and delivered him to be killed.
\end{para}

\begin{para}
    \textsuperscript{17:20}\thinspace{}A\-bin\-a\-di suffered death by fire be\-cause he would not de\-ny the commandments of God, hav\-ing sealed the truth of his words with his death.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{18:1}\thinspace{}Al\-ma, who had fled from king Noah's servants, be\-gan to teach Abinadi's words.
    \textsuperscript{3}\thinspace{}Man\-y be\-lieved his words.
\end{para}

\begin{para}
    \textsuperscript{18:31}\thinspace{}These things were done in the borders of the land, that the king might not know.
    \textsuperscript{32}\thinspace{}But they were made known to the king.
    \textsuperscript{33}\thinspace{}He said that Al\-ma was in\-cit\-ing the peo\-ple to re\-bel\-lion a\-gainst him, there\-fore he sent his ar\-my to de\-stroy them.
\end{para}

\begin{para}
    \textsuperscript{23:1}\thinspace{}Now Al\-ma, hav\-ing been warned by God, told his peo\-ple, there\-fore they gathered to\-geth\-er their flocks, and their grain, and
    \textsuperscript{3}\thinspace{}they fled eight days' jour\-ney in\-to the wil\-der\-ness.
\end{para}

\begin{para}
    \textsuperscript{23:4}\thinspace{}They came to a beau\-ti\-ful and pleas\-ant land,
    \textsuperscript{5}\thinspace{}and be\-gan to till the ground, and build buildings. They were in\-dus\-tri\-ous, and worked ver\-y hard.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{23:6}\thinspace{}The peo\-ple wanted Al\-ma to be their king, for his peo\-ple loved him.
    \textsuperscript{7}\thinspace{}But he said to them, “It's not ex\-pe\-di\-ent for us to have a king.
    \textsuperscript{9}\thinspace{}Re\-mem\-ber the in\-iq\-ui\-ty of king No\-ah and his priests.
    \textsuperscript{12}\thinspace{}King No\-ah oppressed you, and you were in bond\-age to him and his priests.
    \textsuperscript{13}\thinspace{}Be\-cause you have been delivered out of the hands of king No\-ah and his peo\-ple, I ask you not to trust any man to be a king o\-ver you.”
\end{para}

\begin{para}
    \textsuperscript{23:15}\thinspace{}Thus did Al\-ma teach his peo\-ple that eve\-ry man should love his neigh\-bor as him\-self, and that there should be no con\-ten\-tion a\-mong them.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Mosiah}
\markright{Mosiah}
\begin{para}
    \textsuperscript{1:2}\thinspace{}King Ben\-ja\-min had three sons, and he caused them to be taught so that they might be\-come men of un\-der\-stand\-ing.
    \textsuperscript{3}\thinspace{}He al\-so taught them a\-bout the records which were en\-graved on the brass plates, say\-ing: “My sons,
    \textsuperscript{4}\thinspace{}it was\-n't pos\-si\-ble for our fa\-ther, Lehi, to re\-mem\-ber all these things; but he could read these engravings, and teach them to his chil\-dren, that they could teach them to their chil\-dren, e\-ven down to the pres\-ent time.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{1:9}\thinspace{}King Ben\-ja\-min grew old, and saw that he must ver\-y soon go the way of all the earth.
    \textsuperscript{10}\thinspace{}There\-fore, he
    \textsuperscript{18}\thinspace{}proclaimed to all the peo\-ple to gath\-er them\-selves to\-geth\-er to hear his words.
\end{para}

\begin{para}
    \textsuperscript{2:5}\thinspace{}When they ar\-rived, they pitched their tents on eve\-ry side, eve\-ry fam\-i\-ly be\-ing sep\-a\-rate from one an\-oth\-er.
\end{para}

\begin{para}
    \textsuperscript{2:8}\thinspace{}He caused his words to be writ\-ten and sent to those who could not hear him, that they might al\-so re\-ceive his words.
    \textsuperscript{9}\thinspace{}These are the words which he spoke:
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{2:10}\thinspace{}I have not commanded you to come here to fear me, or to think that I am more than a mor\-tal man.
    \textsuperscript{11}\thinspace{}But I am like you, sub\-ject to all kinds of infirmities in bod\-y and mind; yet the peo\-ple have cho\-sen me, and God has al\-lowed me to be a king; to serve you with all the might, mind and strength which he has grant\-ed me.
\end{para}

\begin{para}
    \textsuperscript{2:12}\thinspace{}I have been al\-lowed to spend my days in your serv\-ice, and have not sought gold, or sil\-ver, or any kind of rich\-es from you.
    \textsuperscript{13}\thinspace{}Nei\-ther have I al\-lowed you to be con\-fined in dungeons, or to make slaves of one an\-oth\-er, or to mur\-der, or steal, or com\-mit a\-dul\-ter\-y.
    \textsuperscript{14}\thinspace{}And e\-ven I, my\-self, have la\-bored with my own hands that I might serve you, and that you should not be bur\-dened with taxes.
\end{para}

\begin{para}
    \textsuperscript{2:15}\thinspace{}Yet, I have not done this to boast.
    \textsuperscript{17}\thinspace{}I tell you so you may learn wis\-dom; that you may learn that when you serve your fel\-low beings you are on\-ly serv\-ing God.
\end{para}

\begin{para}
    \textsuperscript{2:18}\thinspace{}If I, whom you call your king, la\-bor to serve you, then should\-n't you la\-bor to serve one an\-oth\-er?
    \textsuperscript{19}\thinspace{}If I, whom you call your king, mer\-it any thanks from you, O how you should thank your heav\-en\-ly king!
\end{para}

\begin{para}
    \textsuperscript{2:41}\thinspace{}Con\-sid\-er the bless\-ed and hap\-py state of those who keep God's commandments. They are bless\-ed in all things, both tem\-po\-ral and spir\-it\-u\-al, and if they re\-main faith\-ful to the end they are re\-ceived in\-to heav\-en to dwell with God in never-ending hap\-pi\-ness.
\end{para}

\begin{para}
    \textsuperscript{3:19}\thinspace{}The nat\-u\-ral man is an en\-e\-my to God, and will be for\-ev\-er, un\-less he puts off the nat\-u\-ral man and becomes like a child: sub\-mis\-sive, meek, hum\-ble, pa\-tient, full of love, will\-ing to sub\-mit to all things, e\-ven as a child submits to his fa\-ther.
\end{para}

\begin{para}
    \textsuperscript{4:9}\thinspace{}Be\-lieve in God. Be\-lieve that he is, and that he cre\-at\-ed all things. Be\-lieve that he has all wis\-dom. Be\-lieve that man does not com\-pre\-hend all that God can com\-pre\-hend.
    \textsuperscript{10}\thinspace{}Be\-lieve that you must re\-pent of your sins and for\-sake them, and hum\-ble your\-selves be\-fore God. Ask in sin\-cer\-i\-ty of heart that he would for\-give you. And now, if you be\-lieve all these things see that you do them.
\end{para}

\begin{para}
    \textsuperscript{4:11}\thinspace{}Hum\-ble your\-selves, call\-ing on God dai\-ly.
    \textsuperscript{12}\thinspace{}If you do this you will al\-ways re\-joice,
    \textsuperscript{13}\thinspace{}and you will not have a mind to in\-jure one an\-oth\-er, but to live at peace, and to re\-pay eve\-ry man what he is due.
    \textsuperscript{14}\thinspace{}You will not al\-low your chil\-dren to go hun\-gry, or na\-ked, or to trans\-gress God's laws and fight and quar\-rel with one an\-oth\-er.
    \textsuperscript{15}\thinspace{}But you will teach them to walk in the ways of so\-ber truth, to love one an\-oth\-er, and to serve one an\-oth\-er.
    \textsuperscript{16}\thinspace{}You will help those who need your help. You will give of your possessions to him who stands in need. You won't al\-low the beg\-gar to pe\-ti\-tion you in vain, and turn him a\-way to per\-ish.
\end{para}

\begin{para}
    \textsuperscript{4:17}\thinspace{}Per\-haps you will say, “The man has brought his mis\-er\-y on him\-self, there\-fore I won't give him of my food, or of my possessions, for his punishments are just.”
    \textsuperscript{18}\thinspace{}Who\-ev\-er does this has great cause to re\-pent. Un\-less he repents he perishes for\-ev\-er, and has no part in the king\-dom of God.
    \textsuperscript{19}\thinspace{}For aren't we all beggars? Don't we all de\-pend on God for all our possessions, for food and cloth\-ing, for gold and sil\-ver, and for all our rich\-es of eve\-ry kind?
\end{para}

\begin{para}
    \textsuperscript{4:22}\thinspace{}If you judge the man who petitions you for your possessions, that he per\-ish not, and con\-demn him, how much more just will be your con\-dem\-na\-tion for with\-hold\-ing your possessions, which don't be\-long to you but to God.
    \textsuperscript{23}\thinspace{}Woe to that man, for his possessions will per\-ish with him. I speak to those who are rich in the things of this world.
\end{para}

\begin{para}
    \textsuperscript{4:24}\thinspace{}I say to the poor, you who on\-ly have e\-nough to re\-main from day to day; I mean all you who de\-ny the beg\-gar, be\-cause you have not. Say in your hearts: “I don't give be\-cause I don't have, but if I had I would give.”
    \textsuperscript{25}\thinspace{}If you say this in your hearts you re\-main guilt\-less. Oth\-er\-wise you are condemned, for you cov\-et what you have not re\-ceived.
\end{para}

\begin{para}
    \textsuperscript{4:26}\thinspace{}To re\-tain for\-give\-ness of your sins: Share your possessions with the poor, eve\-ry man ac\-cord\-ing to what he has, such as feed\-ing the hun\-gry, cloth\-ing the na\-ked, visiting the sick and help\-ing them, both spiritually and tem\-po\-ral\-ly, ac\-cord\-ing to their needs.
    \textsuperscript{27}\thinspace{}And see that all this is done in wis\-dom and or\-der; for it is not nec\-es\-sar\-y for a man to run faster than he has strength. It is ex\-pe\-di\-ent for him to be dil\-i\-gent, that he might win the prize. There\-fore, all things must be done in or\-der.
\end{para}

\begin{para}
    \textsuperscript{4:28}\thinspace{}Re\-mem\-ber that who\-ev\-er borrows from his neigh\-bor should re\-turn the thing, as he has a\-greed, or else you will sin; and per\-haps you will cause your neigh\-bor to sin al\-so.
\end{para}

\begin{para}
    \textsuperscript{4:30}\thinspace{}If you do not watch your\-selves, and your thoughts, and your words, and your deeds, and ob\-serve God's commandments, e\-ven un\-til the end of your lives, you must per\-ish.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{5:13}\thinspace{}How does a man know the mas\-ter whom he has not served, and who is a strang\-er to him, and is far from the thoughts and intents of his heart?
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{6:4}\thinspace{}Mo\-si\-ah be\-gan to reign in his father's place.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{25:1}\thinspace{}He gathered all the peo\-ple to\-geth\-er, and
    \textsuperscript{5}\thinspace{}read the records of Zeniff to them.
    \textsuperscript{6}\thinspace{}He al\-so read the ac\-count of Al\-ma and his peo\-ple.
    \textsuperscript{15}\thinspace{}Al\-ma al\-so spoke to them, preaching re\-pent\-ance and faith on God.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{28:10}\thinspace{}King Mo\-si\-ah had no one to con\-fer the king\-dom on, for none of his sons would ac\-cept the king\-dom.
    \textsuperscript{11}\thinspace{}There\-fore he took the records
    \textsuperscript{20}\thinspace{}and con\-ferred them on Al\-ma, who was the son of Al\-ma, and commanded him to keep and pre\-serve them, and al\-so to keep a rec\-ord of the peo\-ple, handing them down from one gen\-er\-a\-tion to the next.
\end{para}

\begin{para}
    \textsuperscript{29:4}\thinspace{}King Mo\-si\-ah wrote to the peo\-ple say\-ing,
    \textsuperscript{5}\thinspace{}“You want to have a king.
    \textsuperscript{6}\thinspace{}He to whom the king\-dom right\-ly belongs has de\-clined.
    \textsuperscript{7}\thinspace{}If I ap\-point an\-oth\-er in his place, I fear there would a\-rise contentions which would cause wars, which would de\-stroy the souls of man\-y peo\-ple.”
\end{para}

\begin{para}
    \textsuperscript{29:10}\thinspace{}“Let us be wise and an\-tic\-i\-pate these things, and do what will make for peace.
    \textsuperscript{11}\thinspace{}There\-fore I will be your king for the rest of my life. Nev\-er\-the\-less, let us ap\-point judges, to judge this peo\-ple ac\-cord\-ing to our law, and we will new\-ly ar\-range the af\-fairs of this peo\-ple.”
\end{para}

\begin{para}
    \textsuperscript{29:16}\thinspace{}“Be\-cause all men are not just, it is not ex\-pe\-di\-ent to have a king or kings to rule o\-ver you.
    \textsuperscript{17}\thinspace{}For be\-hold, how much in\-iq\-ui\-ty does one wick\-ed king cause to be com\-mit\-ted, and what great de\-struc\-tion!”
\end{para}

\begin{para}
    \textsuperscript{29:21}\thinspace{}“You can on\-ly de\-throne an in\-iq\-ui\-tous king through much con\-ten\-tion, and blood\-shed.
    \textsuperscript{22}\thinspace{}For he has his friends in in\-iq\-ui\-ty, and he keeps his guards a\-round him, and he tears up the laws of those who have reigned in right\-eous\-ness be\-fore him, and he tramples God's commandments un\-der his feet,
    \textsuperscript{23}\thinspace{}and he enacts laws ac\-cord\-ing to his own wick\-ed\-ness, and who\-ev\-er does not o\-bey his laws he causes to be destroyed.”
\end{para}

\begin{para}
    \textsuperscript{29:25}\thinspace{}“There\-fore, choose judges by the voice of the peo\-ple, that you may be judged ac\-cord\-ing to the laws which have been giv\-en by our fathers.
    \textsuperscript{26}\thinspace{}Ob\-serve this and make it your law: do your busi\-ness by the voice of the peo\-ple.”
\end{para}

\begin{para}
    \textsuperscript{29:37}\thinspace{}The peo\-ple were con\-vinced of the truth of his words,
    \textsuperscript{38}\thinspace{}there\-fore they relinquished their de\-sire for a king.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{29:40}\thinspace{}The peo\-ple loved Mo\-si\-ah, for they did not con\-sid\-er him a ty\-rant who was seek\-ing for that mon\-ey which corrupts the soul, for he had not exacted rich\-es from them, nei\-ther had he de\-light\-ed in shed\-ding blood, but he had established peace in the land, and had grant\-ed to his peo\-ple that they should be delivered from all kinds of bond\-age, there\-fore they esteemed him be\-yond meas\-ure.
\end{para}

\begin{para}
    \textsuperscript{29:41}\thinspace{}They ap\-point\-ed judges to judge them ac\-cord\-ing to the law.
    \textsuperscript{42}\thinspace{}Al\-ma was ap\-point\-ed to be the first chief judge.
\end{para}

\begin{para}
    \textsuperscript{29:47}\thinspace{}Thus ended the reign of the kings o\-ver the peo\-ple.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Alma}
\markright{Alma}
\begin{para}
    \textsuperscript{1:2}\thinspace{}In the first year of Alma's reign in the judg\-ment seat, there was a man named Nehor
    \textsuperscript{3}\thinspace{}who was preaching that eve\-ry priest and teach\-er should be\-come pop\-u\-lar, and should not la\-bor with his hands, but be supported by the peo\-ple.
\end{para}

\begin{para}
    \textsuperscript{1:7}\thinspace{}As he was preaching, Nehor be\-gan to con\-tend sharp\-ly with a man
    \textsuperscript{8}\thinspace{}named Gid\-e\-on.
    \textsuperscript{9}\thinspace{}Nehor drew his sword and killed Gid\-e\-on.
\end{para}

\begin{para}
    \textsuperscript{1:10}\thinspace{}Nehor was brought be\-fore Al\-ma to be judged.
    \textsuperscript{12}\thinspace{}Al\-ma said to him,
    \textsuperscript{14}\thinspace{}“You are condemned to die, ac\-cord\-ing to the law.”
\end{para}

\begin{para}
    \textsuperscript{1:15}\thinspace{}He suffered a shame\-ful death,
    \textsuperscript{16}\thinspace{}yet this did not stop priest\-craft from spreading through the land, for there were man\-y who loved the vain things of the world.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{1:21}\thinspace{}There was a strict law a\-mong the peo\-ple of the church that no one should per\-se\-cute those who did not be\-long to the church.
    \textsuperscript{27}\thinspace{}They imparted of their possessions to the poor, and the needy, and the sick, and the afflicted. They did not wear ex\-pen\-sive cloth\-ing, yet they were neat and come\-ly.
\end{para}

\begin{para}
    \textsuperscript{1:29}\thinspace{}Be\-cause of the stead\-i\-ness of the church they be\-came ver\-y rich, hav\-ing an a\-bun\-dance of eve\-ry\-thing they needed.
    \textsuperscript{30}\thinspace{}In their pros\-per\-ous circumstances, they did not send a\-way any who were na\-ked, or hun\-gry, or sick. They did not set their hearts on rich\-es, there\-fore they were gen\-er\-ous to all.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{4:6}\thinspace{}The peo\-ple of the church be\-gan to grow proud be\-cause of the great rich\-es they had obtained by their dil\-i\-gence. They were lifted up in their pride, for they be\-gan to wear ver\-y ex\-pen\-sive cloth\-ing.
    \textsuperscript{7}\thinspace{}Now this caused Al\-ma much af\-flic\-tion.
\end{para}

\begin{para}
    \textsuperscript{4:9}\thinspace{}There be\-gan to be great contentions a\-mong the peo\-ple of the church. There was en\-vy, strife, mal\-ice, per\-se\-cu\-tion, and pride.
\end{para}

\begin{para}
    \textsuperscript{4:15}\thinspace{}Al\-ma, see\-ing the afflictions, persecutions, and in\-e\-qual\-i\-ty, be\-gan to be ver\-y sorrowful.
    \textsuperscript{16}\thinspace{}He selected a wise man, and
    \textsuperscript{18}\thinspace{}delivered the judg\-ment seat to him.
    \textsuperscript{19}\thinspace{}He did this that he him\-self might go out a\-mong his peo\-ple to re\-mind them of their du\-ty, and to tear down all the pride, craft\-i\-ness, and con\-ten\-tion a\-mong them.
    \textsuperscript{20}\thinspace{}And thus, Al\-ma delivered up the judg\-ment seat, and con\-fined him\-self whol\-ly to the priest\-hood of God.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{5:2}\thinspace{}This is what Al\-ma said to the peo\-ple of the church:
\end{para}

\begin{para}
    \textsuperscript{5:14}\thinspace{}I ask you, have you ex\-pe\-ri\-enced a might\-y change in your hearts?
    \textsuperscript{15}\thinspace{}Do you look for\-ward and view this mor\-tal bod\-y raised in im\-mor\-tal\-i\-ty to stand be\-fore God to be judged ac\-cord\-ing to the deeds which have been done in the mor\-tal bod\-y?
\end{para}

\begin{para}
    \textsuperscript{5:17}\thinspace{}Do you im\-ag\-ine that you can lie to God in that day, and say, “God, our works have been right\-eous,” and that he will save you?
\end{para}

\begin{para}
    \textsuperscript{5:26}\thinspace{}If you have ex\-pe\-ri\-enced a change of heart, can you feel so now?
    \textsuperscript{27}\thinspace{}Have you walked, keep\-ing your\-selves blame\-less be\-fore God? Could you say, if you were called to die at this time, that you have been sufficiently hum\-ble?
    \textsuperscript{28}\thinspace{}Are you strip\-ped of pride? If not, you are not pre\-pared to meet God.
\end{para}

\begin{para}
    \textsuperscript{5:30}\thinspace{}Is there one a\-mong you who mocks his broth\-er, or heaps persecutions on him?
    \textsuperscript{31}\thinspace{}Woe to such a one, for he is not pre\-pared, and must re\-pent!
\end{para}

\begin{para}
    \textsuperscript{5:57}\thinspace{}All of you who de\-sire to fol\-low the voice of the good shep\-herd, come out from the wick\-ed, be sep\-a\-rate, and don't tou\-ch their un\-clean things.
\end{para}

\begin{para}
    \textsuperscript{7:2}\thinspace{}If I had\-n't giv\-en the judg\-ment seat to an\-oth\-er, to reign in my place, I could not have come at this time.
\end{para}

\begin{para}
    \textsuperscript{7:15}\thinspace{}Lay a\-side eve\-ry sin which so eas\-i\-ly entangles you, which binds you down to de\-struc\-tion. Show God that you are will\-ing to re\-pent of your sins.
    \textsuperscript{23}\thinspace{}Be hum\-ble, sub\-mis\-sive and gen\-tle; o\-pen to rea\-son; full of pa\-tience and long-suffering; ex\-er\-cis\-ing self-control in all things; dil\-i\-gent in keep\-ing the commandments of God at all times; asking for what\-ev\-er you need, both spir\-it\-u\-al and tem\-po\-ral; al\-ways thanking God for what\-ev\-er you re\-ceive.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{8:6}\thinspace{}Al\-ma came to a cit\-y called Am\-mon\-i\-hah.
    \textsuperscript{19}\thinspace{}As he entered the cit\-y he said to a man, “Will you give a hum\-ble serv\-ant of God some\-thing to eat?”
\end{para}

\begin{para}
    \textsuperscript{8:21}\thinspace{}The man was named Am\-u\-lek, and he brought bread and meat and placed them be\-fore Al\-ma.
    \textsuperscript{27}\thinspace{}Al\-ma stayed with Am\-u\-lek for man\-y days be\-fore he be\-gan to preach to the peo\-ple.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{8:30}\thinspace{}Al\-ma and Am\-u\-lek went out a\-mong the peo\-ple, to de\-clare God's words.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{12:9}\thinspace{}Al\-ma be\-gan to ex\-plain things to them say\-ing, “God on\-ly imparts to the chil\-dren of men ac\-cord\-ing to the at\-ten\-tion and dil\-i\-gence which they give him.
    \textsuperscript{10}\thinspace{}There\-fore, he who hardens his heart receives the less\-er por\-tion of the word, and he who will not hard\-en his heart receives the greater por\-tion of the word.”
\end{para}

\begin{para}
    \textsuperscript{12:12}\thinspace{}“All men must die, and be brought be\-fore God to be judged ac\-cord\-ing to our works.
    \textsuperscript{14}\thinspace{}For our words, our works, and our thoughts will con\-demn us.”
\end{para}

\begin{para}
    \textsuperscript{12:24}\thinspace{}“We see that death co\-mes to man\-kind. Nev\-er\-the\-less there was a space grant\-ed to man in which he might re\-pent; there\-fore this life be\-came a probationary state; a time to pre\-pare to meet God.
    \textsuperscript{37}\thinspace{}Be\-cause we know these things, let us re\-pent, and not hard\-en our hearts.”
\end{para}

\begin{para}
    \textsuperscript{13:27}\thinspace{}“I wish, from the in\-ner\-most part of my heart, that you would lis\-ten to my words, and cast off your sins, and not pro\-cras\-ti\-nate the day of your re\-pent\-ance,
    \textsuperscript{28}\thinspace{}but that you would hum\-ble your\-selves be\-fore God, and watch and pray con\-tin\-u\-al\-ly, that you may not be tempted a\-bove what you can bear, be\-com\-ing hum\-ble, meek, sub\-mis\-sive, pa\-tient, full of love and all long-suffering;
    \textsuperscript{29}\thinspace{}hav\-ing faith in God; hav\-ing hope that you will re\-ceive e\-ter\-nal life; al\-ways hav\-ing the love of God in your hearts.”
\end{para}

\begin{para}
    \textsuperscript{13:31}\thinspace{}Al\-ma spoke man\-y more words to the peo\-ple which are not writ\-ten in this book.
\end{para}

\begin{para}
    \textsuperscript{14:1}\thinspace{}Af\-ter he had spo\-ken to the peo\-ple, man\-y of them be\-lieved his words, and be\-gan to re\-pent, and to search the scriptures.
\end{para}

\begin{para}
    \textsuperscript{15:16}\thinspace{}Am\-u\-lek had left all his gold, sil\-ver, and pre\-cious things, be\-ing rejected by those who were once his friends and al\-so by his fa\-ther and his relatives.
    \textsuperscript{18}\thinspace{}Al\-ma, see\-ing all these things, took Am\-u\-lek to his own house, and ministered to him in his tribulations.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{29:2}\thinspace{}I, Al\-ma, would de\-clare re\-pent\-ance to eve\-ry\-one, that they should re\-pent and come to God, that there might be less sor\-row on the earth.
    \textsuperscript{3}\thinspace{}But I sin in my wish, for I should be con\-tent with what God has al\-lot\-ted me.
    \textsuperscript{8}\thinspace{}For God grants to all nations, of their own na\-tion and lan\-guage, peo\-ple to teach what he sees fit that they should have.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{30:6}\thinspace{}A man be\-gan to preach a\-gainst the com\-ing of Christ.
    \textsuperscript{7}\thinspace{}Now there was no law a\-gainst a man's be\-lief.
    \textsuperscript{9}\thinspace{}If a man de\-sired to serve God, it was his priv\-i\-lege, but if he did not be\-lieve in him there was no law to pun\-ish him.
\end{para}

\begin{para}
    \textsuperscript{30:12}\thinspace{}This man, named Kor\-i\-hor, preached say\-ing,
    \textsuperscript{14}\thinspace{}“These things which you call prophecies are fool\-ish traditions of your fathers.
    \textsuperscript{15}\thinspace{}You can't know of things you don't see.
    \textsuperscript{16}\thinspace{}You say that you see a re\-mis\-sion of your sins, but it is the ef\-fect of a fren\-zied mind. This de\-range\-ment of your minds co\-mes be\-cause of the traditions of your fathers, which lead you a\-way to be\-lieve in things which are not so.”
    \textsuperscript{17}\thinspace{}And he told them that what\-ev\-er a man did was no crime.
\end{para}

\begin{para}
    \textsuperscript{30:31}\thinspace{}He re\-viled a\-gainst Al\-ma and the priests, ac\-cus\-ing them of lead\-ing a\-way the peo\-ple af\-ter the sil\-ly traditions of their fathers, to glut on the labors of the peo\-ple.
\end{para}

\begin{para}
    \textsuperscript{30:32}\thinspace{}Al\-ma said to him, “You know that we don't glut our\-selves on the labors of this peo\-ple. I've la\-bored with my own hands for my sup\-port.
    \textsuperscript{33}\thinspace{}Not\-with\-stand\-ing the man\-y labors I've performed in the church, I've nev\-er re\-ceived as much as one pen\-ny for my la\-bor; nei\-ther have any of my brothers.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{31:1}\thinspace{}Al\-ma re\-ceived news that one Zoram was lead\-ing the peo\-ple to bow down to idols.
    \textsuperscript{5}\thinspace{}As preaching the word had a great ten\-den\-cy to lead the peo\-ple to do what was just --- it had had a more pow\-er\-ful ef\-fect on the minds of the peo\-ple than the sword, or any\-thing else which had happened to them --- there\-fore Al\-ma thought it was ex\-pe\-di\-ent to try the pow\-er of the word of God.
    \textsuperscript{7}\thinspace{}There\-fore he went a\-mong the Zoramites to preach the word to them.
\end{para}

\begin{para}
    \textsuperscript{32:2}\thinspace{}Af\-ter much la\-bor, they be\-gan to have suc\-cess a\-mong the poor, for they were cast out of the synagogues be\-cause of their coarse cloth\-ing.
\end{para}

\begin{para}
    \textsuperscript{32:5}\thinspace{}One of them said, “They cast us out of our synagogues and we have no place to wor\-ship God. What shall we do?”
\end{para}

\begin{para}
    \textsuperscript{32:7}\thinspace{}Al\-ma said to them,
    \textsuperscript{10}\thinspace{}“Do you sup\-pose that you can on\-ly wor\-ship God in your synagogues?
    \textsuperscript{11}\thinspace{}Do you sup\-pose that you must on\-ly wor\-ship God once per week?”
\end{para}

\begin{para}
    \textsuperscript{32:22}\thinspace{}“God is mer\-ci\-ful to all who be\-lieve his word.
    \textsuperscript{23}\thinspace{}He imparts his word to men, and wom\-en. This is not all, lit\-tle chil\-dren of\-ten have words giv\-en to them which con\-found the wise and the learn\-ed.”
\end{para}

\begin{para}
    \textsuperscript{32:28}\thinspace{}“We will com\-pare the word to a seed. If you make room to plant a seed in your heart, you will be\-gin to say with\-in your\-selves, ‘It's a good seed, for it begins to en\-light\-en my un\-der\-stand\-ing.’
    \textsuperscript{33}\thinspace{}Be\-cause you have tried the ex\-per\-i\-ment, and planted the seed,
    \textsuperscript{34}\thinspace{}you know that the word has sprouted up, that your un\-der\-stand\-ing begins to be enlightened, and your mind begins to ex\-pand.”
    \textsuperscript{37}\thinspace{}As the tree begins to grow, you will say: “Let's nour\-ish it care\-ful\-ly, that it may bear fruit.
    \textsuperscript{38}\thinspace{}But if you ne\-glect the tree, and don't think a\-bout its nour\-ish\-ment, it with\-ers a\-way.
    \textsuperscript{39}\thinspace{}Now, this is\-n't be\-cause the seed was not good, or be\-cause its fruit would not be de\-sir\-a\-ble, but it's be\-cause you won't nour\-ish the tree.”
\end{para}

\begin{para}
    \textsuperscript{33:2}\thinspace{}“You've said that you could\-n't wor\-ship God be\-cause you're cast out of your synagogues, but you're great\-ly mis\-tak\-en.
    \textsuperscript{3}\thinspace{}Do you re\-mem\-ber what Zenos, the proph\-et of old, said a\-bout pray\-er?
    \textsuperscript{4}\thinspace{}He said, ‘You are mer\-ci\-ful, O God, for you have heard my pray\-er, e\-ven when I was in the wil\-der\-ness,
    \textsuperscript{5}\thinspace{}and when I cried to you in my field.
    \textsuperscript{6}\thinspace{}And a\-gain, when I turned to my house, you heard my pray\-er.
    \textsuperscript{8}\thinspace{}You are mer\-ci\-ful to your chil\-dren when they cry to you, to be heard by you and not by men, and you will hear them.’”
\end{para}

\begin{para}
    \textsuperscript{34:1}\thinspace{}Af\-ter Al\-ma had spo\-ken, Am\-u\-lek taught them, say\-ing,
    \textsuperscript{20}\thinspace{}“Cry to God when you're in your fields, o\-ver all your flocks.
    \textsuperscript{21}\thinspace{}Cry to him in your hous\-es, o\-ver all your house\-hold, both morn\-ing, mid\-day, and eve\-ning.
    \textsuperscript{24}\thinspace{}Cry to him o\-ver the crops of your fields, that you may pros\-per in them.
    \textsuperscript{25}\thinspace{}Cry o\-ver the flocks of your fields, that they may in\-crease.
    \textsuperscript{26}\thinspace{}You must pour out your souls in your se\-cret places, and in your wil\-der\-ness.
    \textsuperscript{28}\thinspace{}Don't sup\-pose that this is all; if you turn a\-way the needy, and the na\-ked, and don't vis\-it the sick and afflicted, your pray\-er is emp\-ty, and brings you noth\-ing, and you are hypocrites.”
\end{para}

\begin{para}
    \textsuperscript{34:32}\thinspace{}“This life is the time for men to la\-bor.
    \textsuperscript{33}\thinspace{}Don't pro\-cras\-ti\-nate the day of your re\-pent\-ance.
    \textsuperscript{40}\thinspace{}Have pa\-tience, and bear all kinds of af\-flic\-tion. Don't re\-vile a\-gainst those who cast you out be\-cause of your great pov\-er\-ty, lest you be\-come sinners like them.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{35:16}\thinspace{}Al\-ma gathered his sons to\-geth\-er:
\end{para}

\begin{para}
    \textsuperscript{37:1}\thinspace{}My son Helaman, take the records which have been entrusted to me,
    \textsuperscript{2}\thinspace{}and keep a rec\-ord of this peo\-ple.
    \textsuperscript{6}\thinspace{}You may sup\-pose that this is fool\-ish\-ness, but by small and sim\-ple things are great things brought to pass.
    \textsuperscript{8}\thinspace{}These things have en\-larged the mem\-o\-ry of this peo\-ple, and con\-vinced man\-y of the er\-ror of their ways.
\end{para}

\begin{para}
    \textsuperscript{37:35}\thinspace{}In your youth, learn wis\-dom and to keep the commandments of God.
    \textsuperscript{47}\thinspace{}Be so\-ber. My son, fare\-well.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{38:1}\thinspace{}My son Shiblon,
    \textsuperscript{10}\thinspace{}be dil\-i\-gent and tem\-per\-ate in all things.
    \textsuperscript{11}\thinspace{}Don't be lifted up in pride. Don't boast in your own wis\-dom, or in your much strength.
    \textsuperscript{12}\thinspace{}Use bold\-ness, but not overbearance. Bri\-dle all your passions, that you may be filled with love. Re\-frain from i\-dle\-ness.
\end{para}

\begin{para}
    \textsuperscript{38:14}\thinspace{}Don't say, “God, I thank you that we are bet\-ter than our brothers.” Rath\-er say, “God, for\-give my unworthiness, and re\-mem\-ber my brothers in mer\-cy.”
\end{para}

\begin{para}
    \textsuperscript{38:15}\thinspace{}Be so\-ber. My son, fare\-well.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{39:1}\thinspace{}My son Corianton,
    \textsuperscript{2}\thinspace{}this is what I have a\-gainst you: you went on boasting in your strength and your wis\-dom,
    \textsuperscript{3}\thinspace{}and you went af\-ter a pros\-ti\-tute.
    \textsuperscript{5}\thinspace{}Don't you know that these things are an a\-bom\-i\-na\-tion to God?
\end{para}

\begin{para}
    \textsuperscript{39:9}\thinspace{}Re\-pent and for\-sake your sins, and don't go af\-ter the lusts of your eyes any\-more.
    \textsuperscript{11}\thinspace{}Don't al\-low your\-self to be led a\-way by any vain or fool\-ish thing.
    \textsuperscript{14}\thinspace{}Don't seek af\-ter rich\-es or the vain things of this world, for you can\-not car\-ry them with you.
\end{para}

\begin{para}
    \textsuperscript{41:14}\thinspace{}Be mer\-ci\-ful to your brothers. Deal just\-ly, judge right\-eous\-ly, and do good con\-tin\-u\-al\-ly.
    \textsuperscript{15}\thinspace{}For what you send out will re\-turn to you a\-gain.
\end{para}

\begin{para}
    \textsuperscript{42:27}\thinspace{}Who\-ev\-er will come may par\-take of the waters of life free\-ly, and who\-ev\-er will not come is not com\-pelled to come.
\end{para}

\begin{para}
    \textsuperscript{42:31}\thinspace{}My son, de\-clare the word with truth and so\-ber\-ness.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Helaman}
\markright{Helaman}
\begin{para}
    \textsuperscript{3:37}\thinspace{}In the fifty-third year of the reign of the judges, Nephi, the son of Helaman, the son of Helaman, be\-gan to reign in the judg\-ment seat.
\end{para}

\begin{para}
    \textsuperscript{4:18}\thinspace{}In the sixty-second year of the reign of the judges,
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{5:1}\thinspace{}Nephi delivered the judg\-ment seat to Cezoram,
    \textsuperscript{4}\thinspace{}for Nephi was wea\-ry be\-cause of the in\-iq\-ui\-ty of the peo\-ple. He gave up the judg\-ment seat to preach the word of God for the rest of his life, and his broth\-er Lehi al\-so.
    \textsuperscript{5}\thinspace{}For they remembered what their fa\-ther Helaman had said to them:
\end{para}

\begin{para}
    \textsuperscript{5:6}\thinspace{}“My sons, de\-clare these words to the peo\-ple:
    \textsuperscript{12}\thinspace{}Re\-mem\-ber that it is up\-on the rock of Christ that you must build your founda\-tion; that when the dev\-il sends forth his might\-y winds, when all his hail and his might\-y storm beats up\-on you, it will have no pow\-er to dra\-g you down to the gulf of mis\-er\-y and end\-less woe, be\-cause of the rock on which you are built.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{6:15}\thinspace{}In the sixty-sixth year of the reign of the judges, Cezoram was murdered. His son, who had been ap\-point\-ed by the peo\-ple in his place, was al\-so murdered.
\end{para}

\begin{para}
    \textsuperscript{6:16}\thinspace{}The peo\-ple grew ver\-y wick\-ed.
    \textsuperscript{17}\thinspace{}For God had bless\-ed them so long with the rich\-es of the world that they be\-gan to set their hearts on their rich\-es. They be\-gan to seek for prof\-it that they might be lifted up one a\-bove an\-oth\-er. There\-fore they be\-gan to mur\-der, rob, and plun\-der, to get mon\-ey.
\end{para}

\begin{para}
    \textsuperscript{6:18}\thinspace{}Those murderers and plunderers were a band formed by Gadianton.
    \textsuperscript{22}\thinspace{}They had signs, and se\-cret words to rec\-og\-nize a broth\-er who had entered in\-to the cov\-e\-nant, that what\-ev\-er wick\-ed\-ness he did he should not be in\-jured by those who belonged to his band.
    \textsuperscript{23}\thinspace{}Thus they might mur\-der, plun\-der, steal, and com\-mit for\-ni\-ca\-tion and all kinds of wick\-ed\-ness, con\-tra\-ry to the laws of their coun\-try and the laws of God.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{7:4}\thinspace{}Nephi saw the peo\-ple in a state of aw\-ful wick\-ed\-ness, and the Gadianton robbers fill\-ing the judg\-ment seats, hav\-ing usurped the au\-thor\-i\-ty of the land;
    \textsuperscript{5}\thinspace{}let\-ting the guilt\-y go un\-pun\-ished be\-cause of their mon\-ey; be\-ing held in of\-fice to get mon\-ey and glo\-ry of the world, that they might more eas\-i\-ly com\-mit a\-dul\-ter\-y, steal, and kill.
\end{para}

\begin{para}
    \textsuperscript{7:6}\thinspace{}This great in\-iq\-ui\-ty had come up\-on the peo\-ple in on\-ly a few years.
\end{para}

\begin{para}
    \textsuperscript{7:13}\thinspace{}Nephi said to the peo\-ple,
    \textsuperscript{20}\thinspace{}“How could you have for\-got\-ten God?
    \textsuperscript{21}\thinspace{}Be\-hold, it is to get mon\-ey, and to be praised of men.
    \textsuperscript{26}\thinspace{}Woe will come to you be\-cause of that pride which you have al\-lowed to en\-ter your hearts, which has lifted you up be\-cause of your ver\-y great rich\-es!”
\end{para}

\begin{para}
    \textsuperscript{8:25}\thinspace{}“You have rejected the truth, and rebelled a\-gainst God. In\-stead of stor\-ing up for your\-selves treasures in heav\-en, where noth\-ing destroys, you are ac\-cu\-mu\-lat\-ing for your\-selves an\-ger for the day of judg\-ment.”
\end{para}

\begin{para}
    \textsuperscript{10:18}\thinspace{}They would not lis\-ten to his words.
\end{para}

\begin{para}
    \textsuperscript{12:1}\thinspace{}Thus we can see the unsteadiness of the hearts of the chil\-dren of men.
    \textsuperscript{2}\thinspace{}At the ver\-y time when God prospers his peo\-ple, do\-ing all things for their wel\-fare and hap\-pi\-ness, then is the time that they hard\-en their hearts, and for\-get God, be\-cause of their ease, and ver\-y great pros\-per\-i\-ty.
\end{para}

\begin{para}
    \textsuperscript{12:4}\thinspace{}How fool\-ish, vain, and quick to do in\-iq\-ui\-ty, and how slow to do good, are the chil\-dren of men. How quick to set their hearts on the vain things of the world!
    \textsuperscript{5}\thinspace{}How quick to be lifted up in pride; how quick to boast; and how slow to re\-mem\-ber God, and to lis\-ten to his guid\-ance. How slow to walk in wisdom's paths!
\end{para}

\begin{para}
    \textsuperscript{12:6}\thinspace{}They ne\-glect God's guid\-ance, and will not have him as their guide.
    \textsuperscript{23}\thinspace{}Bless\-ed are they who will re\-pent and lis\-ten to the voice of God.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{13:2}\thinspace{}There was one Samuel who came in\-to the land, and preached re\-pent\-ance to the peo\-ple.
    \textsuperscript{4}\thinspace{}They would not let him en\-ter the cit\-y, so he got on the cit\-y wall, and
    \textsuperscript{5}\thinspace{}said to them:
\end{para}

\begin{para}
    \textsuperscript{13:8}\thinspace{}God says, “Be\-cause of the hard\-ness of the hearts of the peo\-ple, un\-less they re\-pent I will take a\-way my word from them.
    \textsuperscript{10}\thinspace{}Your enemies, of the fourth gen\-er\-a\-tion, will live to see your ut\-ter de\-struc\-tion.”
    \textsuperscript{21}\thinspace{}Your rich\-es are curs\-ed be\-cause you have set your hearts on them, and have not listened to the words of him who gave them to you.
    \textsuperscript{22}\thinspace{}You don't re\-mem\-ber God in the things with which he has bless\-ed you, but you al\-ways re\-mem\-ber your rich\-es, not to thank God for them. Your hearts swell with great pride, un\-to boasting, en\-vy, strife, mal\-ice, per\-se\-cu\-tion, mur\-der, and all man\-ner of in\-iq\-ui\-ty.
\end{para}

\begin{para}
    \textsuperscript{13:24}\thinspace{}Woe to you, be\-cause you drive out the prophets, and mock them, and throw stones at them, and kill them, just as they did long a\-go.
    \textsuperscript{25}\thinspace{}And now you say, “If we had lived in the time of our fathers, we would not have killed the prophets. We would not have stoned them, and driv\-en them out.”
\end{para}

\begin{para}
    \textsuperscript{13:26}\thinspace{}You are worse than they. If a proph\-et co\-mes to you and declares your sins and iniquities, you are an\-gry with him, and look for ways to de\-stroy him. You will say that he is a false proph\-et, be\-cause he testifies that your deeds are evil.
\end{para}

\begin{para}
    \textsuperscript{13:27}\thinspace{}But if a man co\-mes to you and says, “Do this and there is no in\-iq\-ui\-ty. Walk af\-ter the pride of your eyes, and do what\-ev\-er your heart desires,” you will say that he is a proph\-et.
    \textsuperscript{28}\thinspace{}You will dress him in ex\-pen\-sive cloth\-ing be\-cause he speaks flattering words, and says that all is well.
\end{para}

\begin{para}
    \textsuperscript{14:30}\thinspace{}Re\-mem\-ber that you are free. You are permitted to act for your\-selves. God
    \textsuperscript{31}\thinspace{}has al\-lowed you to know good from evil, and to choose life or death. You can do good and have good re\-stored to you, or you can do evil and have evil re\-stored to you.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{16:1}\thinspace{}Man\-y heard the words of Samuel, and be\-lieved him, and confessed their sins.
    \textsuperscript{2}\thinspace{}But those who did not be\-lieve Samuel were an\-gry with him. They threw stones and shot ar\-rows at him as he stood on the wall, but they could not hit him.
\end{para}

\begin{para}
    \textsuperscript{16:7}\thinspace{}He jumped down from the wall, fled out of their land,
    \textsuperscript{8}\thinspace{}and was nev\-er heard from a\-gain.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{3 Nephi}
\markright{3 Nephi}
\begin{para}
    \textsuperscript{2:10}\thinspace{}The peo\-ple still remained in wick\-ed\-ness.
    \textsuperscript{11}\thinspace{}The Gadianton robbers had be\-come so nu\-mer\-ous, and killed so man\-y of the peo\-ple, that all the peo\-ple
    \textsuperscript{12}\thinspace{}were com\-pelled, for the safe\-ty of their lives and their wom\-en and their chil\-dren, to take up arms a\-gainst those Gadianton robbers to main\-tain their lib\-er\-ty.
\end{para}

\begin{para}
    \textsuperscript{6:4}\thinspace{}The peo\-ple be\-gan a\-gain to pros\-per, and there was great or\-der in the land.
    \textsuperscript{5}\thinspace{}There was noth\-ing to hind\-er the peo\-ple from prospering con\-tin\-u\-al\-ly, ex\-cept if they fell in\-to trans\-gres\-sion.
\end{para}

\begin{para}
    \textsuperscript{6:10}\thinspace{}There be\-gan to be some disputes a\-mong the peo\-ple, and some were lifted up in pride and boasting be\-cause of their ver\-y great rich\-es, e\-ven to great persecutions.
    \textsuperscript{12}\thinspace{}The peo\-ple be\-gan to be dis\-tin\-guished by rank, ac\-cord\-ing to their rich\-es and their chances for learn\-ing. Some were ig\-no\-rant be\-cause of their pov\-er\-ty, and others re\-ceived great learn\-ing be\-cause of their rich\-es.
    \textsuperscript{14}\thinspace{}And thus there a\-rose a great in\-e\-qual\-i\-ty in the land, so the church be\-gan to be bro\-ken up.
\end{para}

\begin{para}
    \textsuperscript{6:15}\thinspace{}Now the cause of the in\-iq\-ui\-ty was this: the peo\-ple were puffing up with pride, seek\-ing for pow\-er, au\-thor\-i\-ty, rich\-es, and the vain things of the world.
    \textsuperscript{18}\thinspace{}They did not sin ig\-no\-rant\-ly, be\-cause God's will had been taught to them, there\-fore they wil\-ful\-ly rebelled a\-gainst God.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{7:14}\thinspace{}The peo\-ple di\-vid\-ed in\-to tribes, eve\-ry man ac\-cord\-ing to his fam\-i\-ly. They had come to an a\-gree\-ment that they would not go to war with one an\-oth\-er, or tres\-pass a\-gainst one an\-oth\-er, so in some de\-gree they had peace in the land. Nev\-er\-the\-less, their hearts were turned from God, and they stoned the prophets and cast them out.
\end{para}

\begin{para}
    \textsuperscript{7:17}\thinspace{}Nephi preached man\-y things to the peo\-ple.
    \textsuperscript{18}\thinspace{}They were an\-gry with him.
    \textsuperscript{21}\thinspace{}On\-ly a few were con\-vert\-ed to God.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{11:1}\thinspace{}A great mul\-ti\-tude were gathered to\-geth\-er
    \textsuperscript{2}\thinspace{}talking a\-bout Jesus Christ.
    \textsuperscript{3}\thinspace{}While they were talking with one an\-oth\-er, they heard a voice as if it came from heav\-en.
    \textsuperscript{8}\thinspace{}They looked up, and saw a man descending out of heav\-en. He came down and stood a\-mong them.
    \textsuperscript{9}\thinspace{}He reached out his hand and spoke to the peo\-ple, say\-ing,
    \textsuperscript{10}\thinspace{}“I am Jesus Christ, whom the prophets tes\-ti\-fied would come in\-to the world.”
\end{para}

\begin{para}
    \textsuperscript{11:12}\thinspace{}When Jesus said these words the whole mul\-ti\-tude fell to the ground, for they remembered that it had been prophesied that Christ would show him\-self to them.
\end{para}

\begin{para}
    \textsuperscript{11:18}\thinspace{}Jesus commanded Nephi to come for\-ward,
    \textsuperscript{22}\thinspace{}and called others, and said to them,
    \textsuperscript{28}\thinspace{}“There shall be no disputes a\-mong you, as there have been un\-til now.
    \textsuperscript{29}\thinspace{}For he that has the spir\-it of con\-ten\-tion is not of me, but is of the dev\-il, who is the fa\-ther of con\-ten\-tion, and he stirs up the hearts of men to con\-tend with an\-ger, one with an\-oth\-er.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{12:1}\thinspace{}When Jesus had spo\-ken these words to Nephi, and those who had been called, he said to the mul\-ti\-tude:
\end{para}

\begin{para}
    \textsuperscript{12:3}\thinspace{}Bless\-ed are the poor in spir\-it who come to me, for the king\-dom of heav\-en is theirs.
    \textsuperscript{4}\thinspace{}Bless\-ed are all those who mourn, for they will be comforted.
    \textsuperscript{5}\thinspace{}Bless\-ed are the meek, for they will in\-her\-it the earth.
    \textsuperscript{6}\thinspace{}Bless\-ed are all those who hun\-ger and thirst for right\-eous\-ness, for they will be filled with the ho\-ly spir\-it.
    \textsuperscript{7}\thinspace{}Bless\-ed are the mer\-ci\-ful, for they will re\-ceive mer\-cy.
    \textsuperscript{8}\thinspace{}Bless\-ed are all the pure in heart, for they will see God.
    \textsuperscript{9}\thinspace{}Bless\-ed are all the peacemakers, for they will be called sons of God.
    \textsuperscript{10}\thinspace{}Bless\-ed are all those who are per\-se\-cut\-ed be\-cause of my name, for the king\-dom of heav\-en is theirs.
    \textsuperscript{11}\thinspace{}Bless\-ed are you when peo\-ple in\-sult you and per\-se\-cute you, and false\-ly say all kinds of evil a\-gainst you be\-cause of me,
    \textsuperscript{12}\thinspace{}for you will have great joy and be glad, for your re\-ward in heav\-en will be great; for in the same way they per\-se\-cut\-ed the prophets who were be\-fore you.
\end{para}

\begin{para}
    \textsuperscript{12:15}\thinspace{}Do peo\-ple light a lamp and put it un\-der a bas\-ket? No, but on a stand, and it gives light to eve\-ry\-one in the house.
    \textsuperscript{16}\thinspace{}There\-fore let your light shine be\-fore this peo\-ple, so that they may see your good works and glo\-ri\-fy God who is in heav\-en.
\end{para}

\begin{para}
    \textsuperscript{12:19}\thinspace{}You have the commandments.
    \textsuperscript{20}\thinspace{}Un\-less you keep my commandments, you will nev\-er en\-ter in\-to the king\-dom of heav\-en.
\end{para}

\begin{para}
    \textsuperscript{12:21}\thinspace{}It is writ\-ten that you shall not mur\-der, and who\-ev\-er murders will be sub\-ject to God's judg\-ment.
    \textsuperscript{22}\thinspace{}But I tell you that eve\-ry\-one who is an\-gry with his broth\-er will be sub\-ject to his judg\-ment. And who\-ev\-er says, “You fool,” will be sub\-ject to hell fire.
    \textsuperscript{23}\thinspace{}There\-fore, if you want to come to me, and re\-mem\-ber that your broth\-er has some\-thing a\-gainst you,
    \textsuperscript{24}\thinspace{}go to your broth\-er, and first be rec\-on\-ciled to him, and then come to me with full pur\-pose of heart, and I will re\-ceive you.
\end{para}

\begin{para}
    \textsuperscript{12:27}\thinspace{}It is writ\-ten that you shall not com\-mit a\-dul\-ter\-y.
    \textsuperscript{28}\thinspace{}But I tell you that eve\-ry\-one who looks at a wom\-an with lust for her has al\-read\-y com\-mit\-ted a\-dul\-ter\-y in his heart.
\end{para}

\begin{para}
    \textsuperscript{12:29}\thinspace{}Al\-low none of these things to en\-ter your heart,
    \textsuperscript{30}\thinspace{}for it is bet\-ter to de\-ny your\-selves these things than to be thrown in\-to hell.
\end{para}

\begin{para}
    \textsuperscript{12:33}\thinspace{}It is writ\-ten that you shall not break your oath, but shall ful\-fill your oaths to God.
    \textsuperscript{34}\thinspace{}But I tell you, make no oath at all,
    \textsuperscript{37}\thinspace{}but let your state\-ment be “Yes” or “No”. Any\-thing more than this is evil.
\end{para}

\begin{para}
    \textsuperscript{12:38}\thinspace{}It is writ\-ten, an eye for an eye, and a tooth for a tooth.
    \textsuperscript{39}\thinspace{}But I tell you, don't re\-sist an evil per\-son. In\-stead, if some\-one slaps you on the right cheek, turn to him the oth\-er al\-so.
    \textsuperscript{40}\thinspace{}If some\-one wants to sue you and take your shirt, let him have your coat al\-so.
    \textsuperscript{41}\thinspace{}If some\-one forces you to go one mile, go with him two.
    \textsuperscript{42}\thinspace{}Give to him who asks you, and don't turn a\-way from him who wants to bor\-row from you.
\end{para}

\begin{para}
    \textsuperscript{12:43}\thinspace{}It is writ\-ten, “Love your neigh\-bor and hate your en\-e\-my.”
    \textsuperscript{44}\thinspace{}But I tell you, love your enemies, bless those who curse you, do good to those who hate you, and pray for those who mis\-treat you and per\-se\-cute you
    \textsuperscript{45}\thinspace{}that you may be sons of God. For he makes his sun rise on the evil and the good.
\end{para}

\begin{para}
    \textsuperscript{13:1}\thinspace{}You should give to the poor, but be\-ware of prac\-tic\-ing your right\-eous\-ness be\-fore men to be seen by them.
    \textsuperscript{2}\thinspace{}So when you give to the poor, don't sound a trum\-pet be\-fore you, as the hypocrites do, to be praised by men. They have their re\-ward.
    \textsuperscript{3}\thinspace{}But when you give to the poor, don't let your left hand know what your right hand is do\-ing,
    \textsuperscript{4}\thinspace{}so that your giv\-ing may be in se\-cret.
\end{para}

\begin{para}
    \textsuperscript{13:5}\thinspace{}When you pray, don't do as the hypocrites, for they love to pray to be seen by men. They have their re\-ward.
    \textsuperscript{6}\thinspace{}But when you pray, go in\-to your room, shut your door, and pray to God who is in se\-cret.
    \textsuperscript{7}\thinspace{}When you pray, don't use mean\-ing\-less rep\-e\-ti\-tion like the pagans, for they think that they will be heard for their man\-y words.
\end{para}

\begin{para}
    \textsuperscript{13:16}\thinspace{}When you fast, don't look gloom\-y like the hypocrites, for they dis\-fig\-ure their faces to show men that they are fasting. They have their re\-ward.
    \textsuperscript{17}\thinspace{}But when you fast, a\-noint your head, and wash your face,
    \textsuperscript{18}\thinspace{}so that your fasting is not ob\-vi\-ous to men, but to God, who is in se\-cret.
\end{para}

\begin{para}
    \textsuperscript{13:19}\thinspace{}Don't store up for your\-selves treasures on earth, where moth and rust de\-stroy, and thieves break in and steal.
    \textsuperscript{20}\thinspace{}But store up for your\-selves treasures in heav\-en, where nei\-ther moth nor rust destroys, and where thieves do not break in and steal.
    \textsuperscript{21}\thinspace{}For where your treas\-ure is, there your heart will be al\-so.
\end{para}

\begin{para}
    \textsuperscript{13:24}\thinspace{}No one can serve two masters; for ei\-ther he will hate the one and love the oth\-er, or he will be de\-vot\-ed to the one and des\-pise the oth\-er. You can\-not serve God and mon\-ey.
\end{para}

\begin{para}
    \textsuperscript{14:1}\thinspace{}Don't judge, so that you won't be judged.
    \textsuperscript{2}\thinspace{}For in the way you judge, you will be judged; and with the meas\-ure you use, it will be meas\-ured to you.
\end{para}

\begin{para}
    \textsuperscript{14:3}\thinspace{}Why do you look at the speck in your brother's eye, but don't no\-tice the log in your own eye?
    \textsuperscript{12}\thinspace{}There\-fore, what\-ev\-er you want others to do to you, do al\-so to them.
\end{para}

\begin{para}
    \textsuperscript{14:15}\thinspace{}Be\-ware of false prophets who come to you in sheep's cloth\-ing, but in\-ward\-ly are rav\-en\-ous wolves.
    \textsuperscript{16}\thinspace{}You will rec\-og\-nize them by their fruits. Are grapes gathered from thornbushes, or figs from thistles?
    \textsuperscript{17}\thinspace{}Like\-wise, eve\-ry good tree bears good fruit, but a bad tree bears bad fruit.
    \textsuperscript{18}\thinspace{}A good tree can\-not bear bad fruit, nor can a bad tree bear good fruit.
    \textsuperscript{20}\thinspace{}So you'll rec\-og\-nize them by their fruits.
\end{para}

\begin{para}
    \textsuperscript{14:24}\thinspace{}Eve\-ry\-one who hears these words of mine and acts on them will be like a wise man who built his house on the rock.
    \textsuperscript{25}\thinspace{}The rain fell, the floods came, and the winds blew and beat a\-gainst that house. Yet it did not fall, be\-cause its founda\-tion was on the rock.
\end{para}

\begin{para}
    \textsuperscript{14:26}\thinspace{}Eve\-ry\-one who hears these words of mine and does\-n't act on them, will be like a fool\-ish man who built his house on the sand.
    \textsuperscript{27}\thinspace{}The rain fell, the floods came, and the winds blew and beat a\-gainst that house, and it fell.
\end{para}

\begin{para}
    \textsuperscript{15:9}\thinspace{}Turn to me, and en\-dure to the end, and you will live. For to him who endures to the end, I will give e\-ter\-nal life.
    \textsuperscript{10}\thinspace{}I've giv\-en you the commandments, there\-fore keep my commandments.
\end{para}

\begin{para}
    \textsuperscript{17:2}\thinspace{}I can see that you're weak, that you can\-not un\-der\-stand all that I'm speak\-ing to you.
    \textsuperscript{3}\thinspace{}There\-fore, go to your homes, and pon\-der on the things which I've said, and ask God that you may un\-der\-stand.
\end{para}

\begin{para}
    \textsuperscript{18:18}\thinspace{}You must watch and pray al\-ways to a\-void temp\-ta\-tion.
    \textsuperscript{21}\thinspace{}Pray to God in your families that your wives and your chil\-dren may be bless\-ed.
    \textsuperscript{22}\thinspace{}Meet to\-geth\-er of\-ten, and for\-bid no man from com\-ing to you when you meet to\-geth\-er,
    \textsuperscript{23}\thinspace{}but pray for them, and do not cast them out.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{18:36}\thinspace{}When Jesus fin\-ished these sayings,
    \textsuperscript{39}\thinspace{}he de\-part\-ed from them.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Mormon}
\markright{Mormon}
\begin{para}
    \textsuperscript{8:1}\thinspace{}I, Moroni, fin\-ish the rec\-ord of my fa\-ther, Mormon. I have on\-ly a few things to write, which things my fa\-ther commanded me.
\end{para}

\begin{para}
    \textsuperscript{8:3}\thinspace{}Our enemies killed my fa\-ther, and I a\-lone re\-main to write the sad tale of the de\-struc\-tion of my peo\-ple. I don't know wheth\-er they will kill me.
    \textsuperscript{5}\thinspace{}All my relatives have been killed in bat\-tle, and I have no friends, and no\-where to go.
\end{para}

\begin{para}
    \textsuperscript{8:12}\thinspace{}Who\-ev\-er receives this rec\-ord, and does\-n't con\-demn it be\-cause of its imperfections, will know of greater things than these;
    \textsuperscript{19}\thinspace{}for one who judges rash\-ly will be judged rash\-ly a\-gain; for his wages will be ac\-cord\-ing to his works.
    \textsuperscript{20}\thinspace{}The scrip\-ture says, “‘Man shall not strike, nei\-ther shall he judge, for judg\-ment is mine, and venge\-ance is mine al\-so, and I will re\-pay,’ says God.”
\end{para}

\begin{para}
    \textsuperscript{8:22}\thinspace{}For the e\-ter\-nal purposes of God will roll on un\-til all his promises are ful\-filled.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{8:31}\thinspace{}Man\-y will say, “Do this, or do that, and it does\-n't mat\-ter, for God will up\-hold them at the last day.” But woe to them, for they are in the gall of bit\-ter\-ness and in the bonds of in\-iq\-ui\-ty.
    \textsuperscript{32}\thinspace{}Churches will be built up that will say, “Come to me, and for your mon\-ey you will be for\-giv\-en of your sins.”
    \textsuperscript{33}\thinspace{}You wick\-ed and per\-verse and stub\-born peo\-ple.
\end{para}

\begin{para}
    \textsuperscript{8:36}\thinspace{}I know that you walk in the pride of your hearts, to wear\-ing ver\-y fi\-ne cloth\-ing, to en\-vy\-ing, strife, mal\-ice, per\-se\-cu\-tion, and all man\-ner of in\-iq\-ui\-ty.
    \textsuperscript{37}\thinspace{}For you love mon\-ey, and your possessions, and your fi\-ne cloth\-ing, and adorning your churches, more than you love the poor and the needy, the sick and the afflicted.
\end{para}

\begin{para}
    \textsuperscript{8:39}\thinspace{}Why do you a\-dorn your\-selves with what has no life and yet al\-low the hun\-gry, the needy, the na\-ked, the sick, and the afflicted to pass by you, and not no\-tice them?
\end{para}

\begin{para}
    \textsuperscript{9:27}\thinspace{}Don't doubt, but be\-lieve, and come to God with all your heart, and work out your own sal\-va\-tion with fear and trem\-bling be\-fore him.
    \textsuperscript{28}\thinspace{}Be wise dur\-ing your pro\-ba\-tion. Strip your\-selves of all im\-pu\-ri\-ty. Don't ask so that you may spend it on your pleasures, but ask with an un\-wa\-ver\-ing firm\-ness that you will yield to no temp\-ta\-tion, but serve the true and liv\-ing God.
\end{para}

\begin{para}
    \textsuperscript{9:31}\thinspace{}Don't con\-demn me or my fa\-ther be\-cause of our imperfections, nei\-ther those who have writ\-ten be\-fore us. Rath\-er, give thanks to God that he has revealed our imperfections to you, that you may learn to be wis\-er than we have been.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Moroni}
\markright{Moroni}
\begin{para}
    \textsuperscript{7:1}\thinspace{}I, Moroni, write some of the words of my fa\-ther Mormon.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{7:2}\thinspace{}I, Mormon, speak to you.
    \textsuperscript{5}\thinspace{}I re\-mem\-ber God's word, which says you will rec\-og\-nize them by their works, for if their works are good then they are al\-so good.
    \textsuperscript{6}\thinspace{}For God has said, an evil man can\-not do what is good. For if he offers a gift, or prays to God, un\-less he does it with re\-al in\-tent he gains noth\-ing,
    \textsuperscript{7}\thinspace{}for it is not counted to him as right\-eous\-ness.
    \textsuperscript{8}\thinspace{}For if an evil man gives a gift, he does it re\-luc\-tant\-ly, there\-fore it is counted to him the same as if he had retained the gift, there\-fore he is counted evil be\-fore God.
\end{para}

\begin{para}
    \textsuperscript{7:9}\thinspace{}Like\-wise, it is counted evil to a man if he prays with\-out re\-al in\-tent of heart. He gains noth\-ing, for God receives none such.
    \textsuperscript{10}\thinspace{}There\-fore, an evil man can\-not do what is good.
    \textsuperscript{11}\thinspace{}A bit\-ter spring can\-not pour out good wa\-ter; nei\-ther can a good spring pour out bit\-ter wa\-ter.
\end{para}

\begin{para}
    \textsuperscript{7:13}\thinspace{}That which is of God invites and entices to do good con\-tin\-u\-al\-ly. There\-fore, eve\-ry\-thing which invites and entices to do good, to love God, and to serve him, is in\-spired of God.
    \textsuperscript{14}\thinspace{}Be on guard that you don't judge what is evil to be of God, or what is good and of God to be of the dev\-il.
\end{para}

\begin{para}
    \textsuperscript{7:18}\thinspace{}See that you don't judge un\-just\-ly, for with the same judg\-ment that you judge you will al\-so be judged.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{7:44}\thinspace{}If a man does\-n't have love he is noth\-ing.
    \textsuperscript{45}\thinspace{}Love is pa\-tient, is kind, does not en\-vy, is not ar\-ro\-gant, is not self-seeking, is not ir\-ri\-ta\-ble, keeps no rec\-ord of wrongs, does not re\-joice in unrighteousness but rejoices in the truth, bears all things, believes all things, hopes all things, and endures all things.
\end{para}

\begin{para}
    \textsuperscript{7:46}\thinspace{}There\-fore, love nev\-er ends. Hold fast to love, which is the great\-est of all, for all things must end,
    \textsuperscript{47}\thinspace{}but love endures for\-ev\-er, and who\-ev\-er possesses it at the last day, it will go well with him.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{10:1}\thinspace{}Now I, Moroni,
    \textsuperscript{3}\thinspace{}ex\-hort you that when you read these things, you should re\-mem\-ber how mer\-ci\-ful God has been to the chil\-dren of men, and pon\-der it in your hearts.
\end{para}

\begin{para}
    \textsuperscript{10:8}\thinspace{}Don't de\-ny the gifts of God, for they are man\-y, and they come from the same God. There are dif\-fer\-ent ways that these gifts are administered, but it is the same God who works all things in all men, and they are giv\-en to prof\-it men.
    \textsuperscript{9}\thinspace{}To one is giv\-en that he may teach a mes\-sage of wis\-dom,
    \textsuperscript{10}\thinspace{}to an\-oth\-er teach\-ing a mes\-sage of knowl\-edge,
    \textsuperscript{11}\thinspace{}to an\-oth\-er ver\-y great faith, to an\-oth\-er the gifts of heal\-ing,
    \textsuperscript{12}\thinspace{}to an\-oth\-er work\-ing might\-y miracles,
    \textsuperscript{13}\thinspace{}to an\-oth\-er proph\-e\-sy a\-bout all things,
    \textsuperscript{14}\thinspace{}to an\-oth\-er see\-ing angels and ministering spirits,
    \textsuperscript{15}\thinspace{}to an\-oth\-er all kinds of tongues,
    \textsuperscript{16}\thinspace{}and to an\-oth\-er the in\-ter\-pre\-ta\-tion of languages.
    \textsuperscript{17}\thinspace{}All these gifts come from God to eve\-ry man in\-di\-vid\-u\-al\-ly, as he wills.
\end{para}

\begin{para}
    \textsuperscript{10:21}\thinspace{}Un\-less you have love, faith, and hope, you can\-not be saved.
    \textsuperscript{22}\thinspace{}If you don't have hope you must be in des\-pair, and des\-pair co\-mes be\-cause of in\-iq\-ui\-ty.
\end{para}

\begin{para}
    \textsuperscript{10:34}\thinspace{}I say to all, fare\-well. I soon go to rest in God's par\-a\-dise.
\end{para}

\chapter*{Muhammad}
\markboth{Muhammad}{Muhammad}
\addcontentsline{toc}{section}{Muhammad}\begin{changemargin}{.6in}{.6in}
  \begin{center}
    \itshape
    An a\-bridg\-ment of the Qur'an.
  \end{center}
\end{changemargin}
\begin{para}
    \textsuperscript{1:1}\thinspace{}In the name of God, Most Gra\-cious, Most Mer\-ci\-ful.
    \textsuperscript{2}\thinspace{}All praise is owed to God, sus\-tain\-er of the worlds,
    \textsuperscript{4}\thinspace{}Mas\-ter of the Day of Judg\-ment.
    \textsuperscript{5}\thinspace{}You a\-lone we wor\-ship, and you a\-lone we ask for help.
    \textsuperscript{6}\thinspace{}Guide us to the straight path;
    \textsuperscript{7}\thinspace{}the path of those you have bless\-ed, not of those who earned your an\-ger, nor of those who stray.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{2:2}\thinspace{}This book is a guid\-ance for those con\-scious of God,
    \textsuperscript{3}\thinspace{}who be\-lieve in the un\-seen, are stead\-fast in pray\-er, give out of what We have pro\-vid\-ed them,
    \textsuperscript{4}\thinspace{}be\-lieve in the rev\-e\-la\-tion sent to you, Muhammad, and what was sent be\-fore your time, and who have firm faith in the here\-af\-ter.
    \textsuperscript{5}\thinspace{}They are fol\-low\-ing guid\-ance from their Lord, and they will pros\-per.
    \textsuperscript{6}\thinspace{}As for those who dis\-be\-lieve, it is the same to them wheth\-er you warn them or not: they won't be\-lieve.
\end{para}

\begin{para}
    \textsuperscript{2:42}\thinspace{}Don't mix truth with false\-hood, nor know\-ing\-ly con\-ceal the truth.
    \textsuperscript{43}\thinspace{}Be stead\-fast in pray\-er, spend in char\-i\-ty, and bow down with those who bow down in wor\-ship.
\end{para}

\begin{para}
    \textsuperscript{2:44}\thinspace{}Do you en\-join right\-eous\-ness on the peo\-ple and for\-get to do it your\-selves, e\-ven though you re\-cite the Scrip\-ture? Have you no sense?
    \textsuperscript{45}\thinspace{}Seek help with pa\-tience and pray\-er. It is dif\-fi\-cult in\-deed, ex\-cept for the hum\-ble,
    \textsuperscript{46}\thinspace{}who are cer\-tain that they will meet their Lord and re\-turn to Him.
\end{para}

\begin{para}
    \textsuperscript{2:62}\thinspace{}All who be\-lieve in God and the Last Day, and do right\-eous\-ness, will have their re\-ward with their Lord. They will have no fear, nor will they grieve.
\end{para}

\begin{para}
    \textsuperscript{2:79}\thinspace{}Woe to those who write a book with their own hands then say, “This is from God,” in or\-der to sell it for a small price!
\end{para}

\begin{para}
    \textsuperscript{2:111}\thinspace{}They say, “None will en\-ter Par\-a\-dise un\-less he is a Jew or a Christian.”
    \textsuperscript{112}\thinspace{}In fact, who\-ev\-er submits him\-self whol\-ly to God and does good will have his re\-ward with his Lord.
\end{para}

\begin{para}
    \textsuperscript{2:135}\thinspace{}They say, “Be\-come Jews or Christians, then you will be right\-ly guid\-ed.”
    \textsuperscript{136}\thinspace{}Say, “We be\-lieve in God, and the rev\-e\-la\-tion giv\-en to us, and to Abraham, Ishmael, Isaac, Jacob, and their descendants, and the rev\-e\-la\-tion giv\-en to Mo\-ses, Jesus, and the prophets from their Lord. We make no dis\-tinc\-tion be\-tween any of them, and we sub\-mit to Him.”
\end{para}

\begin{para}
    \textsuperscript{2:153}\thinspace{}Believers, seek help through pa\-tience and pray\-er. For God is with the stead\-fast.
\end{para}

\begin{para}
    \textsuperscript{2:172}\thinspace{}Believers, eat the good things We have pro\-vid\-ed for you, and be grate\-ful to God, if it is Him you wor\-ship.
\end{para}

\begin{para}
    \textsuperscript{2:177}\thinspace{}Right\-eous\-ness is not turn\-ing your faces to\-wards east or west, but right\-eous\-ness is one who believes in God, the Last Day, the angels, the Book, and the prophets; who gives of his wealth, e\-ven though he loves it, to relatives, orphans, the needy, the trav\-el\-er, beggars, and for free\-ing slaves; who are stead\-fast in pray\-er, and spend in char\-i\-ty; who keep their promises; and are pa\-tient in pov\-er\-ty, ad\-ver\-si\-ty, and dur\-ing bat\-tle. These ones are true, and con\-scious of God.
\end{para}

\begin{para}
    \textsuperscript{2:180}\thinspace{}When death approaches any of you and he leaves be\-hind wealth, it is pre\-scribed that he make a prop\-er be\-quest to parents and near relatives. This is a du\-ty for those con\-scious of God.
    \textsuperscript{181}\thinspace{}If any\-one alters the be\-quest af\-ter hear\-ing it, the sin is on them: God hears and knows all.
    \textsuperscript{182}\thinspace{}But if any\-one fears that the tes\-ta\-tor made a mis\-take or did wrong, and makes peace be\-tween the heirs, he incurs no sin: God is for\-giv\-ing and mer\-ci\-ful.
\end{para}

\begin{para}
    \textsuperscript{2:183}\thinspace{}Believers, fasting is pre\-scribed for you, as it was for those be\-fore you, that you may be\-come con\-scious of God.
    \textsuperscript{184}\thinspace{}Fast for a spe\-cif\-ic num\-ber of days, but if you are sick or trav\-el\-ing, make up the days lat\-er. Those who can on\-ly fast with dif\-fi\-cul\-ty may feed a poor per\-son in\-stead. Who\-ev\-er volunteers to give more, it is bet\-ter for him.
    \textsuperscript{185}\thinspace{}In the month of Ramadan the Qur'an was revealed as a guid\-ance for man\-kind, as a clear sign of that guid\-ance, and as judg\-ment be\-tween right and wrong. So who\-ev\-er is pres\-ent dur\-ing that month should fast. God wants ease for you, not hard\-ship; and per\-haps you will be grate\-ful.
\end{para}

\begin{para}
    \textsuperscript{2:188}\thinspace{}Don't con\-sume one another's prop\-er\-ty un\-just\-ly, nor use it to bribe judges, intending to con\-sume some of the prop\-er\-ty of others sin\-ful\-ly and know\-ing\-ly.
\end{para}

\begin{para}
    \textsuperscript{2:215}\thinspace{}What\-ev\-er you spend in char\-i\-ty should be for parents, close relatives, orphans, the poor, and for travelers. What\-ev\-er good you do, God knows it well.
\end{para}

\begin{para}
    \textsuperscript{2:216}\thinspace{}You may dis\-like some\-thing that's good for you, and love some\-thing that's bad for you. God knows, and you do not.
\end{para}

\begin{para}
    \textsuperscript{2:219}\thinspace{}They ask you a\-bout wine and gambling. Say, “In them is great sin, yet some ben\-e\-fit, but the sin is greater than the ben\-e\-fit.” They ask you how much they should give. Say, “The ex\-cess.”
    \textsuperscript{220}\thinspace{}They ask you a\-bout orphans. Say, “Im\-prov\-ing their con\-di\-tion is best. God knows who spoils things and who improves them. If God had wished, He could have put you in\-to difficul\-ties.”
\end{para}

\begin{para}
    \textsuperscript{2:221}\thinspace{}Don't mar\-ry i\-dol\-a\-trous wom\-en un\-til they be\-lieve. A slave who believes is bet\-ter than an i\-dol\-a\-tress, e\-ven if she pleases you. These in\-vite you to the fire.
\end{para}

\begin{para}
    \textsuperscript{2:238}\thinspace{}Strict\-ly guard your hab\-it of prayers. Stand be\-fore God in de\-vout o\-be\-di\-ence.
    \textsuperscript{239}\thinspace{}If you are in dan\-ger, pray walk\-ing or rid\-ing, but when you are se\-cure, re\-mem\-ber God, for He taught you what you did not know.
\end{para}

\begin{para}
    \textsuperscript{2:256}\thinspace{}There shall be no com\-pul\-sion in re\-li\-gion. The right path has be\-come clear from the wrong. Who\-ev\-er rejects false gods and believes in God has grasped the most trust\-wor\-thy hand\-hold that nev\-er breaks.
\end{para}

\begin{para}
    \textsuperscript{2:261}\thinspace{}Those who spend their wealth in God's cause are like a grain of corn which grows sev\-en ears, and each ear has a hun\-dred grains.
    \textsuperscript{262}\thinspace{}Those who spend their wealth in God's cause will have their re\-ward with their Lord: they will have no fear, nor will they grieve.
\end{para}

\begin{para}
    \textsuperscript{2:263}\thinspace{}Kind words and for\-give\-ness are bet\-ter than char\-i\-ty followed by in\-ju\-ry.
    \textsuperscript{264}\thinspace{}Believers, don't can\-cel your char\-i\-ty with reminders of your gen\-er\-os\-i\-ty or with in\-ju\-ry, like those who spend their wealth to be seen of men, but do not be\-lieve in God or the Last Day. They are like a smooth rock with dust on it. Heav\-y rain falls, leav\-ing it bare. They are un\-a\-ble to keep any\-thing they have earned. God does not guide disbelievers.
    \textsuperscript{265}\thinspace{}Those who spend their wealth seek\-ing to please God are like a gar\-den on high ground. Heav\-y rain falls, yield\-ing a dou\-ble har\-vest. If heav\-y rain does\-n't fall, light rain will.
\end{para}

\begin{para}
    \textsuperscript{2:267}\thinspace{}Believers, give from the good things which you have earned. Do not give from what is bad, when you your\-selves would on\-ly ac\-cept it if your eyes were closed.
\end{para}

\begin{para}
    \textsuperscript{2:268}\thinspace{}Satan threatens you with pov\-er\-ty and orders you to com\-mit im\-mo\-ral\-i\-ty. God promises you His for\-give\-ness and boun\-ty.
\end{para}

\begin{para}
    \textsuperscript{2:270}\thinspace{}What\-ev\-er you give, or vow to give, God knows it. But the wrongdoers will have no helpers.
    \textsuperscript{271}\thinspace{}If you dis\-close your char\-i\-ty, it is good, but if you con\-ceal it and give it to the poor, it is bet\-ter for you. It will a\-tone for some of your bad deeds.
\end{para}

\begin{para}
    \textsuperscript{2:272}\thinspace{}What\-ev\-er char\-i\-ty you give benefits your own soul when you do it on\-ly to seek the face of God. What\-ev\-er you give will be repaid to you in full, and you will not be wronged.
    \textsuperscript{273}\thinspace{}Give to the poor who are whol\-ly occupied in God's cause, un\-a\-ble to trav\-el in the land for work. The un\-know\-ing might think they are rich be\-cause of their re\-straint, but you will rec\-og\-nize them be\-cause they don't beg per\-sist\-ent\-ly. What\-ev\-er good you give, be as\-sured that God knows it.
\end{para}

\begin{para}
    \textsuperscript{2:274}\thinspace{}Those who give of their possessions by night and by day, pri\-vate\-ly and pub\-lic\-ly, will have their re\-ward with their Lord: they will have no fear, nor will they grieve.
    \textsuperscript{277}\thinspace{}Those who be\-lieve, do good works, are stead\-fast in pray\-er, and spend in char\-i\-ty, will have their re\-ward with their Lord: they will have no fear, nor will they grieve.
\end{para}

\begin{para}
    \textsuperscript{2:282}\thinspace{}When you con\-tract a debt for a spec\-i\-fied term, write it down. Let the debt\-or write, and not di\-min\-ish what he owes. Bring two witnesses. They must not ref\-use when called to tes\-ti\-fy. Do not dis\-dain to write down your con\-tract, wheth\-er it be small or large, a\-long with the time it falls due: this is more just with God, more re\-li\-a\-ble as ev\-i\-dence, and more like\-ly to pre\-vent doubt be\-tween you. If it be an im\-me\-di\-ate trade, there is no blame if you don't write it down. Have witnesses pres\-ent when you trade with one an\-oth\-er.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{3:3}\thinspace{}He sent down to you the Book with the truth, confirming what went be\-fore it. He sent down the Torah and the Gos\-pel
    \textsuperscript{4}\thinspace{}as a guid\-ance for the peo\-ple.
\end{para}

\begin{para}
    \textsuperscript{3:14}\thinspace{}Al\-lur\-ing to men is the love of their desires: wom\-en, chil\-dren, heap\-ing treasures of gold and sil\-ver, fi\-ne hors\-es, cat\-tle, and farm\-land. These are pleasures of life in this world, but with God is the best place to re\-turn.
\end{para}

\begin{para}
    \textsuperscript{3:92}\thinspace{}You will on\-ly at\-tain true pi\-e\-ty if you give out of what you love. What\-ev\-er you give, tru\-ly, God knows it.
\end{para}

\begin{para}
    \textsuperscript{3:137}\thinspace{}Situations sim\-i\-lar to yours have happened be\-fore: trav\-el through the earth, and see how it ended for the disbelievers.
\end{para}

\begin{para}
    \textsuperscript{3:156}\thinspace{}Believers, don't say of your brothers who trav\-eled or went out to fight, “If they had stayed with us, they would not have died.” This will be\-come a source of re\-gret in your hearts.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{4:2}\thinspace{}Give orphans their prop\-er\-ty. Don't re\-place their good things with your bad things, nor con\-sume their prop\-er\-ty to\-geth\-er with your own. Sure\-ly, this is a great sin.
\end{para}

\begin{para}
    \textsuperscript{4:3}\thinspace{}Mar\-ry wom\-en of your choice, two or three or four. If you fear that you will not treat them fair\-ly, then on\-ly one. That will help you a\-void bi\-as.
    \textsuperscript{4}\thinspace{}Give the wom\-en their dow\-er gra\-cious\-ly. If they will\-ing\-ly re\-mit any of it to you, en\-joy it with a clear con\-science.
\end{para}

\begin{para}
    \textsuperscript{4:5}\thinspace{}Do not en\-trust your prop\-er\-ty to the weak-minded. God made it a means of sup\-port for you, but feed and clothe them with it, and speak to them with kind\-ness.
\end{para}

\begin{para}
    \textsuperscript{4:6}\thinspace{}Test orphans un\-til they reach mar\-riage\-a\-ble age. If they have sound judg\-ment, re\-lease their prop\-er\-ty to them. Don't con\-sume it wastefully or hast\-i\-ly be\-fore they grow up. If the guard\-i\-an is wealth\-y, he should take no com\-pen\-sa\-tion. If he is poor, he should use on\-ly what is rea\-son\-a\-ble. When you re\-lease their prop\-er\-ty to them, bring witnesses.
\end{para}

\begin{para}
    \textsuperscript{4:7}\thinspace{}Men and wom\-en shall each have a share of what parents and close relatives leave, wheth\-er the prop\-er\-ty is small or large.
    \textsuperscript{8}\thinspace{}If oth\-er relatives, or orphans, or poor are pres\-ent dur\-ing the di\-vi\-sion, then give them some\-thing from the es\-tate, and speak to them with kind\-ness.
    \textsuperscript{9}\thinspace{}Let executors have the same fear in their minds as if they had left help\-less chil\-dren be\-hind. Let them fear God, and speak just\-ly.
\end{para}

\begin{para}
    \textsuperscript{4:18}\thinspace{}Re\-pent\-ance has no ef\-fect for those who con\-tin\-ue to do evil un\-til death confronts them and they say, “Now I re\-pent.”
\end{para}

\begin{para}
    \textsuperscript{4:19}\thinspace{}Believers, you are for\-bid\-den to in\-her\-it wom\-en a\-gainst their will. Don't treat your wives harsh\-ly. Live with them in kind\-ness. If you dis\-like them, it may be that you dis\-like some\-thing in which God has placed much good.
\end{para}

\begin{para}
    \textsuperscript{4:22}\thinspace{}Don't mar\-ry wom\-en whom your fathers mar\-ried,
    \textsuperscript{23}\thinspace{}wom\-en who are close relatives, your wet nurses, your milk sisters, your wives' mothers, your stepdaughters, those whom your sons mar\-ried, two sisters at the same time,
    \textsuperscript{24}\thinspace{}and wom\-en al\-read\-y mar\-ried. All oth\-er wom\-en are law\-ful to you, pro\-vid\-ed that you seek them in mar\-riage, with dow\-er from your prop\-er\-ty, de\-sir\-ing mar\-riage not for\-ni\-ca\-tion.
\end{para}

\begin{para}
    \textsuperscript{4:29}\thinspace{}Believers, don't con\-sume one another's wealth un\-just\-ly, but let there be trade by mu\-tu\-al con\-sent. Don't kill your\-selves or one an\-oth\-er, for God is mer\-ci\-ful to you.
\end{para}

\begin{para}
    \textsuperscript{4:31}\thinspace{}If you a\-void the great sins which you are for\-bid\-den, We will re\-move your mi\-nor sins, and ad\-mit you through the en\-trance of hon\-or.
    \textsuperscript{32}\thinspace{}Don't cov\-et what God has bestowed more free\-ly on some than on others. Men and wom\-en re\-ceive a share of what they earn.
\end{para}

\begin{para}
    \textsuperscript{4:36}\thinspace{}Wor\-ship God, and join noth\-ing with Him. Do good to parents, relatives, orphans, the poor, neighbors, the com\-pan\-ion by your side, and travelers. God does not like the ar\-ro\-gant, the boast\-ful,
    \textsuperscript{37}\thinspace{}the sting\-y; those who com\-mand others to be sting\-y, hid\-ing the boun\-ty God has giv\-en them;
    \textsuperscript{38}\thinspace{}those who spend their wealth to be seen by men, and don't be\-lieve in God nor the Last Day.
    \textsuperscript{39}\thinspace{}What harm would it do them to be\-lieve in God and the Last Day, and to spend out of what God has pro\-vid\-ed them?
    \textsuperscript{40}\thinspace{}God wrongs no one in the least de\-gree. If there is a good deed, He dou\-bles it, and gives a great re\-ward.
\end{para}

\begin{para}
    \textsuperscript{4:58}\thinspace{}God commands you to re\-turn what you've been entrusted to its right\-ful own\-er. When you judge be\-tween peo\-ple, do so with jus\-tice.
    \textsuperscript{59}\thinspace{}Believers, o\-bey God, the Mes\-sen\-ger, and those in au\-thor\-i\-ty a\-mong you.
\end{para}

\begin{para}
    \textsuperscript{4:85}\thinspace{}Who\-ev\-er intercedes for a good cause will share in its re\-ward. Who\-ev\-er intercedes for an evil cause will share in its bur\-den.
\end{para}

\begin{para}
    \textsuperscript{4:86}\thinspace{}When offered a greet\-ing, re\-spond with a bet\-ter one, or at least its e\-qual. God takes ac\-count of eve\-ry\-thing.
\end{para}

\begin{para}
    \textsuperscript{4:95}\thinspace{}God prefers the believers who strive with their wealth and lives o\-ver those who sit --- oth\-er than the dis\-a\-bled. Al\-though He has prom\-ised good to all of them, those who strive re\-ceive a greater re\-ward than those who sit.
\end{para}

\begin{para}
    \textsuperscript{4:100}\thinspace{}Who\-ev\-er emigrates for God's cause will find on the earth a\-bun\-dance, and man\-y places of ref\-uge. If he then dies as a ref\-u\-gee, his re\-ward from God is cer\-tain.
\end{para}

\begin{para}
    \textsuperscript{4:101}\thinspace{}When you trav\-el, it's no sin to short\-en your prayers, if you fear that disbelievers may at\-tack you.
    \textsuperscript{103}\thinspace{}When you have com\-plet\-ed the pray\-er, re\-mem\-ber God: stand\-ing, sit\-ting, and ly\-ing down. Once you are se\-cure, re\-es\-tab\-lish reg\-u\-lar pray\-er, for pray\-er is man\-da\-to\-ry for believers at spe\-cif\-ic times.
\end{para}

\begin{para}
    \textsuperscript{4:110}\thinspace{}One who does evil or wrongs him\-self, but then asks God for for\-give\-ness, will find God is for\-giv\-ing and mer\-ci\-ful.
    \textsuperscript{112}\thinspace{}But if one sins then blames an in\-no\-cent per\-son, he burdens him\-self with a slan\-der and a fla\-grant sin.
\end{para}

\begin{para}
    \textsuperscript{4:123}\thinspace{}Any\-one who does wrong will be repaid ac\-cord\-ing\-ly and will find no pro\-tec\-tor or help\-er be\-sides God.
    \textsuperscript{124}\thinspace{}Any be\-liev\-er, male or fe\-male, who does good deeds will en\-ter heav\-en, and will not re\-ceive the least in\-jus\-tice.
\end{para}

\begin{para}
    \textsuperscript{4:129}\thinspace{}You will nev\-er be a\-ble to treat your wives with per\-fect fair\-ness, how\-ev\-er much you may want to, but don't fa\-vor one com\-plete\-ly so as to leave an\-oth\-er hang\-ing. If you make a\-mends and are con\-scious of God, He is for\-giv\-ing and mer\-ci\-ful.
\end{para}

\begin{para}
    \textsuperscript{4:135}\thinspace{}Believers, stead\-fast\-ly up\-hold jus\-tice, as witnesses for God, e\-ven if it is a\-gainst your\-selves, your parents, or your relatives. Wheth\-er rich or poor, God can best pro\-tect them both. Don't fol\-low your own de\-sire, lest you swerve from jus\-tice. If you dis\-tort or ne\-glect jus\-tice, God knows what you do.
\end{para}

\begin{para}
    \textsuperscript{4:140}\thinspace{}When you hear peo\-ple de\-ny and rid\-i\-cule God's messages, do not sit with them un\-til they change topics. Oth\-er\-wise you will be\-come like them: God will gath\-er the hypocrites and disbelievers in hell.
\end{para}

\begin{para}
    \textsuperscript{4:142}\thinspace{}When hypocrites stand to pray, they stand la\-zi\-ly, to be seen of men, remembering God on\-ly a lit\-tle,
    \textsuperscript{143}\thinspace{}wavering be\-tween this and that, be\-long\-ing to nei\-ther these nor those.
\end{para}

\begin{para}
    \textsuperscript{4:148}\thinspace{}God does not like evil to be mentioned in pub\-lic, ex\-cept by the one who was wronged. God hears and knows all.
    \textsuperscript{149}\thinspace{}Wheth\-er you do good openly or pri\-vate\-ly, or par\-don an evil, God is for\-giv\-ing and pow\-er\-ful.
\end{para}

\begin{para}
    \textsuperscript{4:163}\thinspace{}Muhammad, we've sent you rev\-e\-la\-tion as We did to No\-ah and the prophets af\-ter him: Abraham, Ishmael, Isaac, Jacob, Jesus, Job, Jonah, Aaron, and Solomon; to David We gave the Psalms;
    \textsuperscript{164}\thinspace{}to oth\-er messengers We've mentioned be\-fore; to others We have\-n't mentioned; and to Mo\-ses God spoke di\-rect\-ly.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{5:2}\thinspace{}Some peo\-ple shut you out of the Sa\-cred Mosque. Don't let your ha\-tred for them lead you to trans\-gress. Help one an\-oth\-er in right\-eous\-ness and pi\-e\-ty, but don't help one an\-oth\-er in sin and ag\-gres\-sion. Fear God, for God is se\-vere in pun\-ish\-ment.
\end{para}

\begin{para}
    \textsuperscript{5:7}\thinspace{}Re\-mem\-ber God's bless\-ing on you. Fear God, for He knows all the secrets of your heart.
    \textsuperscript{8}\thinspace{}Believers, be stead\-fast in de\-vo\-tion to God, as im\-par\-tial witnesses. Don't let ha\-tred of others lead you a\-way from jus\-tice. Be just: that is nearer to pi\-e\-ty.
    \textsuperscript{9}\thinspace{}God has prom\-ised for\-give\-ness and a great re\-ward to those who be\-lieve and do good works.
\end{para}

\begin{para}
    \textsuperscript{5:35}\thinspace{}Believers, fear God, seek ways to come clos\-er to Him, and strive in His cause that you may pros\-per.
\end{para}

\begin{para}
    \textsuperscript{5:38}\thinspace{}Cut off the hands of thieves as rec\-om\-pense for what they have done: a de\-ter\-rent from God.
    \textsuperscript{39}\thinspace{}But who\-ev\-er repents af\-ter his crime, and makes a\-mends, God will ac\-cept his re\-pent\-ance, for God is for\-giv\-ing and mer\-ci\-ful.
\end{para}

\begin{para}
    \textsuperscript{5:44}\thinspace{}We revealed the Torah containing guid\-ance and light.
    \textsuperscript{45}\thinspace{}In it we ordained a life for a life, an eye for an eye, and sim\-i\-lar ret\-ri\-bu\-tion for wounds. But if any\-one forgoes this out of char\-i\-ty, it is an a\-tone\-ment for him\-self. Those who do not judge by what God has revealed are the wrongdoers.
\end{para}

\begin{para}
    \textsuperscript{5:48}\thinspace{}Muhammad, We revealed to you this Book with the truth, confirming the scriptures that came be\-fore it. To each of you, We have pre\-scribed a law and a clear path. Had God wanted it, He would have made you all a sin\-gle com\-mu\-ni\-ty, but He wanted to test you in what He has giv\-en you, so race to do good. You will all re\-turn to God, then He will clar\-i\-fy your differences.
\end{para}

\begin{para}
    \textsuperscript{5:57}\thinspace{}Believers, don't take al\-lies who view your re\-li\-gion with rid\-i\-cule and a\-muse\-ment.
\end{para}

\begin{para}
    \textsuperscript{5:89}\thinspace{}God will not pun\-ish you for your thought\-less oaths, on\-ly for your de\-lib\-er\-ate oaths. The a\-tone\-ment for break\-ing an oath is to feed ten poor peo\-ple as you would feed your own fam\-i\-ly, or to clothe them, or to free a slave. If you can't af\-ford it, fast for three days. That is the a\-tone\-ment for bro\-ken oaths, but keep your oaths.
\end{para}

\begin{para}
    \textsuperscript{5:90}\thinspace{}Believers: intoxicants, gambling, i\-dol\-a\-try, and div\-i\-na\-tion are abominations from Satan. A\-void them so that you may pros\-per.
    \textsuperscript{91}\thinspace{}Satan on\-ly wants to use intoxicants and gambling to cause en\-mi\-ty and ha\-tred a\-mong you, and to hind\-er you from remembering God and pray\-er. So will you quit?
\end{para}

\begin{para}
    \textsuperscript{5:106}\thinspace{}Believers, when mak\-ing bequests as death approaches you, let there be two hon\-est witnesses.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{6:7}\thinspace{}Muhammad, e\-ven if We had sent down to you a mes\-sage writ\-ten on pa\-per, and the disbelievers had touched it with their hands, they would have said, “This is noth\-ing but ob\-vi\-ous mag\-ic!”
    \textsuperscript{8}\thinspace{}They say, “Why was no an\-gel sent down to him?” Had We sent down an an\-gel, their judg\-ment would have come at once, with\-out res\-pite.
    \textsuperscript{9}\thinspace{}E\-ven if We had sent an an\-gel, We would have sent him as a man, adding to their con\-fu\-sion.
\end{para}

\begin{para}
    \textsuperscript{6:48}\thinspace{}We send the messengers on\-ly to give good news and to warn. Who\-ev\-er believes and reforms will have no fear, nor will they grieve.
    \textsuperscript{49}\thinspace{}But suf\-fer\-ing afflicts those who re\-ject our messages, be\-cause of their de\-fi\-ant dis\-o\-be\-di\-ence.
\end{para}

\begin{para}
    \textsuperscript{6:52}\thinspace{}Don't send a\-way those who call on their Lord morn\-ing and eve\-ning, seek\-ing His face. You are not ac\-count\-a\-ble for them, nor they for you. If you turn them a\-way, you be\-come one of the un\-just.
    \textsuperscript{54}\thinspace{}Your Lord has pre\-scribed for Him\-self mer\-cy: if any of you does wrong out of ig\-no\-rance, but then repents and corrects him\-self, He is for\-giv\-ing and mer\-ci\-ful.
\end{para}

\begin{para}
    \textsuperscript{6:70}\thinspace{}Leave a\-lone those who take play and a\-muse\-ment as their re\-li\-gion, and are de\-ceived by the life of this world. Re\-mind them, lest a soul be destroyed for what it has earned, that it will find no pro\-tec\-tor or in\-ter\-ces\-sor ex\-cept God.
\end{para}

\begin{para}
    \textsuperscript{6:108}\thinspace{}Don't re\-vile those whom idolaters call up\-on, lest they re\-vile God in their spite and ig\-no\-rance. To eve\-ry com\-mu\-ni\-ty We have made its own deeds seem pleas\-ing. In the end they will re\-turn to their Lord, and He will in\-form them of all they did.
\end{para}

\begin{para}
    \textsuperscript{6:120}\thinspace{}A\-void sin, both o\-pen and se\-cret, for those who sin will be repaid for what they've done.
    \textsuperscript{121}\thinspace{}Eat noth\-ing o\-ver which God's name has not been pro\-nounced: that is dis\-o\-be\-di\-ence in\-deed.
\end{para}

\begin{para}
    \textsuperscript{6:141}\thinspace{}Don't be ex\-ces\-sive, for God dislikes the waste\-ful.
    \textsuperscript{142}\thinspace{}Eat what God has pro\-vid\-ed for you, wheth\-er it's live\-stock for burdens or for meat.
\end{para}

\begin{para}
    \textsuperscript{6:151}\thinspace{}Don't as\-cribe any equals with God. Be good to your parents. Don't kill your chil\-dren for fear of pov\-er\-ty --- We will pro\-vide for you and for them. Don't ap\-proach shame\-ful deeds, wheth\-er o\-pen or se\-cret. Don't take the life that God has made sa\-cred, ex\-cept by le\-gal right.
    \textsuperscript{152}\thinspace{}Don't ap\-proach the orphan's prop\-er\-ty be\-fore he co\-mes of age, ex\-cept to im\-prove it. Give full meas\-ure and weight with jus\-tice; we bur\-den no one with more than he can bear. When you speak, be just, e\-ven if it concerns a rel\-a\-tive.
    \textsuperscript{153}\thinspace{}This is My straight path, so fol\-low it. Don't fol\-low oth\-er paths. They will lead you a\-way from His. He has commanded you to do this, so that you may be right\-eous.
\end{para}

\begin{para}
    \textsuperscript{6:164}\thinspace{}Eve\-ry soul is re\-spon\-si\-ble for its own actions, and no one bears another's bur\-den. You will all re\-turn to your Lord in the end, and He will clar\-i\-fy that on which you used to dif\-fer.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{7:3}\thinspace{}Believers, fol\-low the rev\-e\-la\-tion to you from your Lord, and fol\-low no masters oth\-er than Him.
\end{para}

\begin{para}
    \textsuperscript{7:204}\thinspace{}When the Qur'an is re\-cit\-ed, pay at\-ten\-tion and lis\-ten si\-lent\-ly that you may re\-ceive mer\-cy.
    \textsuperscript{205}\thinspace{}In the morn\-ings and eve\-nings, re\-mem\-ber your Lord in\-ward\-ly, with hu\-mil\-i\-ty, rev\-er\-ence, and a qui\-et voice. Don't be a\-mong the heed\-less.
    \textsuperscript{206}\thinspace{}E\-ven those who are near your Lord aren't too proud to wor\-ship Him: they glo\-ri\-fy Him, and pros\-trate be\-fore Him.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{8:2}\thinspace{}Believers are those whose hearts trem\-ble with awe when God is mentioned, whose faith strengthens when His messages are re\-cit\-ed to them, who trust in their Lord,
    \textsuperscript{3}\thinspace{}who are stead\-fast in pray\-er, and who give from what We have pro\-vid\-ed them.
    \textsuperscript{4}\thinspace{}They are the true believers. They have for\-give\-ness with their Lord, and gen\-er\-ous sus\-te\-nance.
\end{para}

\begin{para}
    \textsuperscript{8:20}\thinspace{}Believers, o\-bey God and His Mes\-sen\-ger. Don't turn a\-way from him when you are listening.
    \textsuperscript{21}\thinspace{}Don't be like those who say, “We have heard,” but don't ac\-tu\-al\-ly lis\-ten.
    \textsuperscript{22}\thinspace{}In\-deed, the worst creatures in God's sight are the deaf and dumb who don't use rea\-son.
\end{para}

\begin{para}
    \textsuperscript{8:47}\thinspace{}Don't be like those who came out of their homes con\-ceit\-ed, show\-ing off, and hindering others from God's path.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{9:24}\thinspace{}If your fathers, your sons, your brothers, your wives, your tribes, the wealth you have gained, the com\-merce you fear will de\-cline, and the dwellings you love are dearer to you than God, His Mes\-sen\-ger, and strug\-gle in His cause, then wait un\-til God executes His com\-mand. God does not guide the re\-bel\-lious.
\end{para}

\begin{para}
    \textsuperscript{9:34}\thinspace{}Believers, man\-y rabbis and monks false\-ly de\-vour men's wealth and di\-vert men from God's path. Tell those who hoard gold and sil\-ver, with\-out giv\-ing for God's cause, that they will re\-ceive a pain\-ful pun\-ish\-ment.
\end{para}

\begin{para}
    \textsuperscript{9:54}\thinspace{}On\-ly this prevents a gift from be\-ing ac\-cept\-ed: they don't be\-lieve God and His Mes\-sen\-ger, they come to pray\-er la\-zi\-ly, and they give un\-will\-ing\-ly.
\end{para}

\begin{para}
    \textsuperscript{9:60}\thinspace{}Alms are for the poor, the needy, those who ad\-min\-is\-ter the funds, those whose hearts are to be won o\-ver, to free slaves, to help debtors, for God's cause, and needy travelers.
\end{para}

\begin{para}
    \textsuperscript{9:71}\thinspace{}The believers pro\-tect one an\-oth\-er. They en\-join what is right, for\-bid what is wrong, are stead\-fast in pray\-er, spend in char\-i\-ty, and o\-bey God and His Mes\-sen\-ger. God will have mer\-cy on them.
\end{para}

\begin{para}
    \textsuperscript{9:91}\thinspace{}There is no blame on the weak, the ill, or those with\-out resources to spend, when they are sin\-cere to God and His Mes\-sen\-ger. There is no rea\-son to re\-proach those who do good.
    \textsuperscript{92}\thinspace{}There is al\-so no blame on those who came to you for mounts, and when you said, “I can't find a mount for you,” they turned back, their eyes overflowing with tears of grief that they had noth\-ing to spend.
    \textsuperscript{93}\thinspace{}On\-ly those who re\-quest ex\-emp\-tion, de\-spite be\-ing rich, are o\-pen to blame. They are con\-tent to be with those who stay be\-hind.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{10:11}\thinspace{}If God were to has\-ten the ill that men have earned, as they wish to has\-ten the good, their end would al\-read\-y be here.
    \textsuperscript{12}\thinspace{}And when trou\-ble touches man, he cries out to Us wheth\-er ly\-ing on his side, sit\-ting, or stand\-ing, but when We re\-move his trou\-ble, he goes on his way as if he had nev\-er cried out to Us.
\end{para}

\begin{para}
    \textsuperscript{10:41}\thinspace{}If they don't be\-lieve you, say, “My deeds for me, and yours for you. You are not re\-spon\-si\-ble for what I do, nor I for what you do.”
\end{para}

\begin{para}
    \textsuperscript{10:47}\thinspace{}Eve\-ry com\-mu\-ni\-ty is sent a mes\-sen\-ger. When their mes\-sen\-ger co\-mes, they will be judged with jus\-tice, and will not be wronged.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{11:9}\thinspace{}If We let man taste mer\-cy, and then with\-draw it, he despairs and is un\-grate\-ful.
    \textsuperscript{10}\thinspace{}But if We let him taste mer\-cy af\-ter ad\-ver\-si\-ty has touched him, he is sure to say, “Mis\-for\-tune has left me.” He becomes ex\-ult\-ant and boast\-ful.
    \textsuperscript{11}\thinspace{}Those who are stead\-fast and do right\-eous deeds don't act this way. They will have for\-give\-ness and a great re\-ward.
\end{para}

\begin{para}
    \textsuperscript{11:112}\thinspace{}Stay on the right path as you have been commanded, to\-geth\-er with those who have turned to God. Don't trans\-gress, for He sees all that you do.
    \textsuperscript{113}\thinspace{}Don't in\-cline to\-ward nor re\-ly on those who do wrong.
    \textsuperscript{114}\thinspace{}Be stead\-fast in pray\-er at both ends of the day and dur\-ing parts of the night. Good deeds drive a\-way evil deeds.
    \textsuperscript{115}\thinspace{}Be stead\-fast: God won't lose the re\-ward of those who do good.
\end{para}

\begin{para}
    \textsuperscript{11:117}\thinspace{}Your Lord would nev\-er de\-stroy communities un\-just\-ly if its peo\-ple act right\-eous\-ly.
    \textsuperscript{118}\thinspace{}If your Lord had wanted, He could have made man\-kind one com\-mu\-ni\-ty, but they con\-tin\-ue to dif\-fer.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{13:4}\thinspace{}There are neighboring plots in the land: vineyards, cornfields, and date palms. They are all watered with the same wa\-ter, yet We make some bet\-ter to eat than others. These are signs for peo\-ple who rea\-son.
\end{para}

\begin{para}
    \textsuperscript{13:11}\thinspace{}God does not change the con\-di\-tion of a peo\-ple un\-less they change what is with\-in them\-selves.
\end{para}

\begin{para}
    \textsuperscript{13:22}\thinspace{}Those who are pa\-tient, seek\-ing their Lord's face; who are stead\-fast in pray\-er; who spend se\-cret\-ly and openly from what We have pro\-vid\-ed them; and who re\-pel evil with good: those will have the re\-ward of an e\-ter\-nal home.
\end{para}

\begin{para}
    \textsuperscript{13:26}\thinspace{}They re\-joice in the life of this world, but it is a fleet\-ing pleas\-ure com\-pared to the here\-af\-ter.
    \textsuperscript{27}\thinspace{}The disbelievers say, “Why has\-n't his Lord sent down a sign to him?” Say, “God lets stray whom He will, and guides to Him\-self those who turn to Him,
    \textsuperscript{28}\thinspace{}those who be\-lieve, and those whose hearts find rest in remembering God.
    \textsuperscript{29}\thinspace{}Joy and a beau\-ti\-ful home\-com\-ing a\-wait those who be\-lieve and do right\-eous deeds.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{14:3}\thinspace{}Those who pre\-fer this world o\-ver the here\-af\-ter, and turn others from God's way, mak\-ing it seem crook\-ed: they are far a\-stray.
\end{para}

\begin{para}
    \textsuperscript{14:4}\thinspace{}We have on\-ly sent messengers who speak the lan\-guage of their peo\-ple, in or\-der to make things clear for them.
\end{para}

\begin{para}
    \textsuperscript{14:10}\thinspace{}The messengers said to them, “Can there be any doubt a\-bout God, the cre\-a\-tor of the heavens and earth? He invites you in or\-der to for\-give your sins, and let you live un\-til the ap\-point\-ed time.” They said, “You're on\-ly men like us. You want to turn us a\-way from what our fathers wor\-shipped. In that case, bring us clear proof.”
    \textsuperscript{11}\thinspace{}Their messengers answered, “True, we are on\-ly men like you, but God fa\-vors His servants as He pleases. We can't bring you proof un\-less God permits it, so let all believers put their trust in God.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{15:2}\thinspace{}Per\-haps the disbelievers will wish that they had sub\-mit\-ted to God.
    \textsuperscript{3}\thinspace{}Let them eat and en\-joy. Let false hope dis\-tract them. They will come to know.
\end{para}

\begin{para}
    \textsuperscript{15:19}\thinspace{}We have spread out the earth, placed firm mountains on it, and let eve\-ry\-thing grow on it in due bal\-ance.
    \textsuperscript{20}\thinspace{}We have pro\-vid\-ed sus\-te\-nance in it for you and for oth\-er creatures.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{16:5}\thinspace{}He has cre\-at\-ed live\-stock for you. You de\-rive warmth, food, and oth\-er benefits from them.
    \textsuperscript{6}\thinspace{}You find beau\-ty in them when you bring them home in the eve\-ning, and as you lead them to pas\-ture in the morn\-ing.
    \textsuperscript{7}\thinspace{}They car\-ry your loads to lands that you could on\-ly reach on your own with great hard\-ship. Tru\-ly your Lord is kind and mer\-ci\-ful.
    \textsuperscript{8}\thinspace{}He has cre\-at\-ed hors\-es, mules, and donkeys for you to ride, and as a\-dorn\-ment. He al\-so creates things you know noth\-ing a\-bout.
\end{para}

\begin{para}
    \textsuperscript{16:10}\thinspace{}He sends down wa\-ter from the sky. It provides a drink for you and for the veg\-e\-ta\-tion which feeds your live\-stock.
    \textsuperscript{11}\thinspace{}With it He grows for you crops, olives, date palms, grapes, and eve\-ry kind of fruit. These are signs for those who re\-flect.
\end{para}

\begin{para}
    \textsuperscript{16:14}\thinspace{}He has subjected the sea for you, that you may eat fresh meat from it, and ex\-tract ornaments from it to wear. You see the ships plowing through its waves seek\-ing the boun\-ty of God that you may be grate\-ful.
    \textsuperscript{15}\thinspace{}He has placed rivers and roads on the earth, so you may find your way;
    \textsuperscript{16}\thinspace{}and landmarks and stars that men may guide them\-selves.
    \textsuperscript{17}\thinspace{}Is He who creates like one who does\-n't cre\-ate?
\end{para}

\begin{para}
    \textsuperscript{16:36}\thinspace{}We sent a mes\-sen\-ger to eve\-ry com\-mu\-ni\-ty say\-ing, “Serve God and a\-void false gods.” A\-mong them were some whom God guid\-ed and some who strayed. So trav\-el through the earth and ob\-serve how it ended for the deniers.
\end{para}

\begin{para}
    \textsuperscript{16:61}\thinspace{}If God were to pun\-ish men for their wrong\-do\-ing, He would leave no liv\-ing crea\-ture on the earth, but He gives them res\-pite for an ap\-point\-ed time. When their time co\-mes, they are un\-a\-ble to de\-lay it or ad\-vance it by a sin\-gle hour.
\end{para}

\begin{para}
    \textsuperscript{16:90}\thinspace{}God commands jus\-tice, do\-ing good, and gen\-er\-os\-i\-ty to\-wards relatives. He forbids all shame\-ful deeds, bad con\-duct, and op\-pres\-sion. He instructs you so that you may take heed.
    \textsuperscript{91}\thinspace{}Ful\-fill the cov\-e\-nant of God once you have made it, and don't break your oaths.
    \textsuperscript{92}\thinspace{}Don't be like a wom\-an who unravels the yarn which she has firm\-ly spun.
\end{para}

\begin{para}
    \textsuperscript{16:93}\thinspace{}If God wanted, He could have made you all one peo\-ple, but He lets some stray and guides others. You will be ac\-count\-a\-ble for your actions.
    \textsuperscript{94}\thinspace{}Don't use your oaths to de\-ceive one an\-oth\-er, lest a foot slip that was firm\-ly planted. Then you will taste the evil of hindering some\-one from God's path, and will suf\-fer great pun\-ish\-ment.
\end{para}

\begin{para}
    \textsuperscript{16:96}\thinspace{}What you have will van\-ish, but what God has will en\-dure. We will cer\-tain\-ly re\-ward the stead\-fast ac\-cord\-ing to the best of their actions.
    \textsuperscript{97}\thinspace{}We will give a new life, one that is good and pure, to who\-ev\-er works right\-eous\-ness and has faith, wheth\-er male or fe\-male.
\end{para}

\begin{para}
    \textsuperscript{16:99}\thinspace{}Satan has no pow\-er o\-ver those who be\-lieve and trust in their Lord.
    \textsuperscript{100}\thinspace{}He on\-ly has pow\-er o\-ver those who al\-ly with him and who join part\-ners with God.
\end{para}

\begin{para}
    \textsuperscript{16:110}\thinspace{}Your Lord is for\-giv\-ing and mer\-ci\-ful to those who em\-i\-grate be\-cause of per\-se\-cu\-tion, then fight for the faith and re\-main stead\-fast.
    \textsuperscript{111}\thinspace{}On the Day when eve\-ry soul co\-mes to plead for him\-self, eve\-ry soul will be ful\-ly paid for his actions and none will be wronged.
\end{para}

\begin{para}
    \textsuperscript{16:114}\thinspace{}Eat the good and law\-ful food which God has pro\-vid\-ed for you, and be grate\-ful for His fa\-vors.
\end{para}

\begin{para}
    \textsuperscript{16:119}\thinspace{}Your Lord is for\-giv\-ing and mer\-ci\-ful to those who do wrong out of ig\-no\-rance, but af\-ter\-wards re\-pent and make a\-mends.
\end{para}

\begin{para}
    \textsuperscript{16:125}\thinspace{}In\-vite eve\-ry\-one to your Lord's path with wis\-dom and beau\-ti\-ful teach\-ing. Ar\-gue with them in the most po\-lite way, for your Lord knows who has strayed from His path and who is right\-ly guid\-ed.
    \textsuperscript{126}\thinspace{}If you must re\-spond to an at\-tack, re\-spond no worse than the at\-tack, but it is bet\-ter to show pa\-tience.
    \textsuperscript{127}\thinspace{}Be pa\-tient, for your pa\-tience is from God. Don't grieve o\-ver them or be dis\-tressed by their schemes.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{17:15}\thinspace{}Who\-ev\-er receives guid\-ance does so for his own ben\-e\-fit. Who\-ev\-er strays does so to his own det\-ri\-ment. No one will bear another's bur\-den, but We on\-ly pun\-ish af\-ter We have sent a mes\-sen\-ger.
\end{para}

\begin{para}
    \textsuperscript{17:20}\thinspace{}We be\-stow some of our boun\-ty on all. Your Lord's boun\-ty is not closed to any\-one.
    \textsuperscript{21}\thinspace{}No\-tice that We have bestowed more on some than on others, but the here\-af\-ter holds greater rank and boun\-ty.
\end{para}

\begin{para}
    \textsuperscript{17:22}\thinspace{}Don't make an\-oth\-er de\-i\-ty e\-qual with God or you will be\-come dis\-graced and for\-sak\-en.
    \textsuperscript{23}\thinspace{}Your Lord has de\-creed that you wor\-ship on\-ly Him, and that you be kind to your parents. Don't speak a word of dis\-re\-spect to them, nor scold them, but ad\-dress them re\-spect\-ful\-ly.
    \textsuperscript{24}\thinspace{}Say, “My Lord, have mer\-cy on them just as they reared me when I was small.”
\end{para}

\begin{para}
    \textsuperscript{17:25}\thinspace{}Your Lord knows what is in your hearts. If you are right\-eous, He forgives those who turn to Him a\-gain and a\-gain.
    \textsuperscript{26}\thinspace{}Give what is due to relatives, the needy, and travelers; but don't spend your wealth wastefully.
    \textsuperscript{27}\thinspace{}In\-deed, the waste\-ful are brothers of Satan, and Satan is al\-ways un\-grate\-ful to his Lord.
    \textsuperscript{28}\thinspace{}E\-ven if you must turn a\-way from the needy in pur\-suit of mer\-cy from your Lord, then at least speak a kind word to them.
\end{para}

\begin{para}
    \textsuperscript{17:31}\thinspace{}Don't kill your chil\-dren for fear of pov\-er\-ty. We will pro\-vide for them and for you. In\-deed, kill\-ing them is a great sin.
    \textsuperscript{32}\thinspace{}Don't go near a\-dul\-ter\-y, for it is shame\-ful and an evil path.
    \textsuperscript{33}\thinspace{}Don't take life, which God has made sa\-cred, ex\-cept in pur\-suit of jus\-tice.
    \textsuperscript{34}\thinspace{}Don't go near an orphan's prop\-er\-ty, ex\-cept to im\-prove it, un\-til he reaches ma\-tur\-i\-ty. Ful\-fill eve\-ry com\-mit\-ment, for you will be questioned a\-bout your commitments.
    \textsuperscript{35}\thinspace{}Give full meas\-ure when you meas\-ure, and weigh with hon\-est scales. That is best and ad\-van\-ta\-geous in the end.
    \textsuperscript{36}\thinspace{}Don't con\-cern your\-self with things you know noth\-ing a\-bout: you will be questioned a\-bout hear\-ing, see\-ing, and feel\-ing.
    \textsuperscript{37}\thinspace{}Don't walk on the earth with ar\-ro\-gance: you can't tear the earth a\-part, nor match the mountains in height.
    \textsuperscript{38}\thinspace{}Your Lord detests all these actions.
    \textsuperscript{39}\thinspace{}This is some of the wis\-dom your Lord has revealed to you.
\end{para}

\begin{para}
    \textsuperscript{17:53}\thinspace{}Say what is best. Satan sows dis\-cord a\-mong My servants. He is a clear en\-e\-my to man.
\end{para}

\begin{para}
    \textsuperscript{17:70}\thinspace{}We have honored Adam's chil\-dren, car\-ried them on land and sea, giv\-en them good sus\-te\-nance, and fa\-vored them a\-bove much of our cre\-a\-tion.
    \textsuperscript{71}\thinspace{}One day We will call to\-geth\-er all men a\-long with their leaders. Who\-ev\-er is giv\-en his rec\-ord in his right hand will read it with pleas\-ure, and will not be wronged in the least.
    \textsuperscript{72}\thinspace{}But who\-ev\-er is blind in this world will be blind in the here\-af\-ter and e\-ven fur\-ther a\-stray.
\end{para}

\begin{para}
    \textsuperscript{17:78}\thinspace{}Es\-tab\-lish reg\-u\-lar pray\-er from mid\-day un\-til the dark\-ness of night. Re\-cite the Qur'an at dawn, for morn\-ing rec\-i\-ta\-tion is al\-ways witnessed.
\end{para}

\begin{para}
    \textsuperscript{17:100}\thinspace{}If you pos\-sessed the treas\-ur\-ies of my Lord's mer\-cy, you would with\-hold them for fear of spend\-ing too much: man is ev\-er sting\-y.
\end{para}

\begin{para}
    \textsuperscript{17:110}\thinspace{}Call on God, or on the Most Mer\-ci\-ful. Which\-ev\-er name you call is the same, for the best names be\-long to Him. Don't pray too loud\-ly or too qui\-et\-ly, but seek a mid\-dle way.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{18:23}\thinspace{}Nev\-er say, “I'll do that to\-mor\-row,”
    \textsuperscript{24}\thinspace{}with\-out adding, “God will\-ing.” When you for\-get, re\-mem\-ber your Lord and say, “May my Lord guide me clos\-er to the right path.”
\end{para}

\begin{para}
    \textsuperscript{18:32}\thinspace{}There were two men. We gave one of them two vineyards surrounded by date palms with cornfields in be\-tween.
    \textsuperscript{33}\thinspace{}Both vineyards pro\-duced fruit, did\-n't fail, and We caused a riv\-er to flow through them,
    \textsuperscript{34}\thinspace{}so this man had a\-bun\-dant fruit. He said to his friend, “I have more wealth and men than you.”
    \textsuperscript{35}\thinspace{}Hav\-ing thus sinned a\-gainst him\-self, he went in\-to his gar\-den. He said, “I don't think this will ev\-er per\-ish,
    \textsuperscript{36}\thinspace{}nor that the Hour of Judg\-ment will ev\-er come. E\-ven if I were brought back to my Lord, I would sure\-ly find some\-thing e\-ven bet\-ter than this.”
\end{para}

\begin{para}
    \textsuperscript{18:37}\thinspace{}His friend said to him, “Do you dis\-be\-lieve in Him who cre\-at\-ed you?
    \textsuperscript{38}\thinspace{}But as for me, He is God, My Lord, and I as\-so\-ci\-ate no part\-ner with Him.
    \textsuperscript{39}\thinspace{}When you entered your gar\-den, why did\-n't you say, ‘This is God's will. There is no pow\-er but with God.’ Al\-though I have less wealth and off\-spring than you,
    \textsuperscript{40}\thinspace{}my Lord may give me some\-thing bet\-ter than your gar\-den, or a ca\-lam\-i\-ty may turn your gar\-den in\-to slip\-per\-y sand,
    \textsuperscript{41}\thinspace{}or its wa\-ter might sink un\-der\-ground so that you can't reach it.”
\end{para}

\begin{para}
    \textsuperscript{18:42}\thinspace{}And thus it happened: his fruit was ruined. He remained, wring\-ing his hands o\-ver what he had spent on it, while it col\-lapsed up\-on its trellises, say\-ing, “I wish that I had nev\-er as\-cribed part\-ners to my Lord.”
    \textsuperscript{43}\thinspace{}He had none to help him oth\-er than God, nor could he help him\-self.
\end{para}

\begin{para}
    \textsuperscript{18:45}\thinspace{}This world's life is like rain sent down from the sky. The earth's veg\-e\-ta\-tion absorbs it but soon becomes dry stub\-ble, scattered by the wind.
    \textsuperscript{46}\thinspace{}Wealth and chil\-dren a\-dorn this world's life, but en\-dur\-ing good deeds are bet\-ter to your Lord, and a bet\-ter source of hope.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{20:2}\thinspace{}We did\-n't send down the Qur'an to dis\-tress you,
    \textsuperscript{3}\thinspace{}but on\-ly as a reminder for those who fear God.
\end{para}

\begin{para}
    \textsuperscript{20:15}\thinspace{}Al\-though I keep the hour hid\-den, the time is com\-ing for each soul to be rewarded for its ef\-fort.
    \textsuperscript{16}\thinspace{}Don't let those who de\-ny it and fol\-low their own desires dis\-tract you from it, lest you per\-ish.
\end{para}

\begin{para}
    \textsuperscript{20:81}\thinspace{}Eat from the good things We have pro\-vid\-ed you, but don't trans\-gress, lest My an\-ger de\-scend on you.
    \textsuperscript{82}\thinspace{}Yet I re\-peat\-ed\-ly for\-give those who re\-pent, be\-lieve, do right, and con\-tin\-ue re\-ceiv\-ing guid\-ance.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{22:39}\thinspace{}Those who have been wrong\-ful\-ly attacked may fight back. In\-deed, God has the pow\-er to help
    \textsuperscript{40}\thinspace{}those who have been un\-just\-ly driv\-en from their homes on\-ly for say\-ing, “Our Lord is God”. If God did\-n't re\-pel one peo\-ple with an\-oth\-er, man\-y monasteries, churches, synagogues, and mosques would have been destroyed; places in which the name of God is fre\-quent\-ly in\-voked.
\end{para}

\begin{para}
    \textsuperscript{22:67}\thinspace{}We have ap\-point\-ed a dif\-fer\-ent man\-ner of wor\-ship for each com\-mu\-ni\-ty to ob\-serve, so don't let others ar\-gue with you a\-bout the mat\-ter.
    \textsuperscript{68}\thinspace{}If they do ar\-gue with you, say, “God knows best what you are do\-ing.”
    \textsuperscript{69}\thinspace{}On Res\-ur\-rec\-tion Day, God will judge the differences be\-tween you.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{23:1}\thinspace{}The believers are in\-deed suc\-cess\-ful:
    \textsuperscript{2}\thinspace{}those who pray hum\-bly,
    \textsuperscript{3}\thinspace{}a\-void i\-dle talk,
    \textsuperscript{4}\thinspace{}give alms,
    \textsuperscript{5}\thinspace{}guard their chas\-ti\-ty
    \textsuperscript{6}\thinspace{}(ex\-cept with their spouses),
    \textsuperscript{8}\thinspace{}who are at\-ten\-tive to their trusts and promises,
    \textsuperscript{9}\thinspace{}and guard their prayers.
    \textsuperscript{11}\thinspace{}These will in\-her\-it Par\-a\-dise and live there for\-ev\-er.
\end{para}

\begin{para}
    \textsuperscript{23:44}\thinspace{}We sent our messengers one af\-ter an\-oth\-er. Eve\-ry time a community's mes\-sen\-ger ar\-rived, they de\-nied him. So We let them fol\-low each oth\-er to de\-struc\-tion and they be\-came cau\-tion\-ar\-y ta\-les.
\end{para}

\begin{para}
    \textsuperscript{23:68}\thinspace{}Have they nev\-er con\-tem\-plat\-ed the word of God? Has some\-thing new come to them that did\-n't come to their forefathers?
    \textsuperscript{69}\thinspace{}Do they de\-ny their Mes\-sen\-ger be\-cause they don't rec\-og\-nize him?
    \textsuperscript{70}\thinspace{}Do they say, “He is pos\-sessed”? No, he brought them the truth and most of them hate it.
    \textsuperscript{71}\thinspace{}If truth had followed their desires, the heavens, the earth, and eve\-ry\-thing that lives in them would be ruined. We sent them a reminder, but they turn a\-way from it.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{24:27}\thinspace{}Believers, be\-fore entering an\-oth\-er person's house, ask per\-mis\-sion and greet its occupants.
    \textsuperscript{28}\thinspace{}E\-ven if no one is in the house, wait un\-til you re\-ceive per\-mis\-sion. If you are told, “Go a\-way,” do so. God is a\-ware of all that you do.
\end{para}

\begin{para}
    \textsuperscript{24:30}\thinspace{}Tell believing men to low\-er their gaze and guard their pri\-vate parts.
    \textsuperscript{31}\thinspace{}Tell believing wom\-en to low\-er their gaze and guard their pri\-vate parts, to cov\-er their breasts, and not to stamp their feet to draw at\-ten\-tion to what is hid\-den.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{25:7}\thinspace{}They say, “What sort of a mes\-sen\-ger is this? He eats food and walks through the mar\-ket\-place. Why was\-n't an an\-gel sent down to help him warn?
    \textsuperscript{8}\thinspace{}Why has\-n't he been giv\-en treas\-ure, or a gar\-den to feed him? The man you fol\-low is simp\-ly bewitched.”
    \textsuperscript{9}\thinspace{}They've gone a\-stray and can't find the right path.
\end{para}

\begin{para}
    \textsuperscript{25:20}\thinspace{}Muhammad, all the messengers We sent be\-fore you were al\-so men who ate food and walked through the mar\-ket\-place. We've made some peo\-ple a test for others. Will you have pa\-tience? Your Lord sees eve\-ry\-thing.
\end{para}

\begin{para}
    \textsuperscript{25:27}\thinspace{}On that Day the wrong\-do\-er will bite his hand and say, “If on\-ly I had tak\-en the same path as the Mes\-sen\-ger!
    \textsuperscript{28}\thinspace{}Woe is me! If on\-ly I had not tak\-en so-and-so as a friend!
    \textsuperscript{29}\thinspace{}He lead me a\-way from the Reminder af\-ter it came to me. Satan al\-ways betrays man\-kind.”
\end{para}

\begin{para}
    \textsuperscript{25:32}\thinspace{}Disbelievers say, “Why is\-n't the Qur'an revealed to him all at once?” This is to strength\-en your heart. We have revealed it grad\-u\-al\-ly, in stages.
\end{para}

\begin{para}
    \textsuperscript{25:47}\thinspace{}He makes the night as cloth\-ing for you, sleep as rest, and the day like a res\-ur\-rec\-tion.
    \textsuperscript{48}\thinspace{}He sends the winds as messengers of good news be\-fore His mer\-cy. We send down pure wa\-ter from the sky
    \textsuperscript{49}\thinspace{}to re\-vive a dead land and as a drink for our creations: animals and men in great numbers.
    \textsuperscript{50}\thinspace{}We have re\-peat\-ed this man\-y times a\-mong them as a reminder, but most re\-main un\-grate\-ful.
\end{para}

\begin{para}
    \textsuperscript{25:56}\thinspace{}Muhammad, We have on\-ly sent you to de\-liv\-er good news and a warn\-ing.
    \textsuperscript{57}\thinspace{}Say, “The on\-ly pay\-ment I ask for it is that who\-ev\-er wishes takes a path to his Lord.”
\end{para}

\begin{para}
    \textsuperscript{25:63}\thinspace{}The servants of the Most Gra\-cious are those who walk hum\-bly on the earth, and when fools ad\-dress them, they re\-ply, “Peace;”
    \textsuperscript{67}\thinspace{}those who, when they spend, are nei\-ther ex\-trav\-a\-gant nor sting\-y, but mod\-er\-ate be\-tween those extremes;
    \textsuperscript{68}\thinspace{}those who nev\-er in\-voke a de\-i\-ty be\-side God, nor take a life which God has made sa\-cred, nor com\-mit for\-ni\-ca\-tion.
\end{para}

\begin{para}
    \textsuperscript{25:70}\thinspace{}God will change evil in\-to good for him who repents, believes, and does good deeds. God is for\-giv\-ing and mer\-ci\-ful.
    \textsuperscript{71}\thinspace{}Who\-ev\-er repents and does good has turned to\-wards God with true re\-pent\-ance.
\end{para}

\begin{para}
    \textsuperscript{25:72}\thinspace{}Servants of God are those who give no false wit\-ness and pass by fri\-vol\-i\-ty with dig\-ni\-ty;
    \textsuperscript{73}\thinspace{}those who don't ig\-nore their Lord's messages when reminded of them;
    \textsuperscript{74}\thinspace{}and those who pray, “Our Lord, give us joy in our wives and chil\-dren, and let us be examples to the right\-eous.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{28:50}\thinspace{}Who is fur\-ther a\-stray than one who follows his own desires with\-out guid\-ance from God? In\-deed, God does\-n't guide wrongdoers.
    \textsuperscript{51}\thinspace{}We have conveyed the Word to them, so they may re\-mem\-ber.
\end{para}

\begin{para}
    \textsuperscript{28:54}\thinspace{}Those who be\-lieve the Scriptures will be rewarded dou\-ble, be\-cause they are stead\-fast, re\-pel evil with good, give out of what We have giv\-en them,
    \textsuperscript{55}\thinspace{}and turn a\-way from friv\-o\-lous speech say\-ing, “We have our actions and you have yours. Peace be with you. We don't seek as\-so\-ci\-a\-tion with the ig\-no\-rant.”
    \textsuperscript{56}\thinspace{}You can't guide eve\-ry\-one you love.
\end{para}

\begin{para}
    \textsuperscript{28:76}\thinspace{}Qarun had great rich\-es. His peo\-ple said to him, “Don't gloat, for God does\-n't like gloaters.
    \textsuperscript{77}\thinspace{}In\-stead, use your wealth to seek the home of the here\-af\-ter with\-out for\-get\-ting your share in this world. Do good, as God has been good to you. Don't spread cor\-rup\-tion in the land, for God does\-n't like those who spread cor\-rup\-tion.”
    \textsuperscript{79}\thinspace{}He went out a\-mong the peo\-ple in his pomp. Those who de\-sired a world\-ly life said, “If on\-ly we had what Qarun has. He is tru\-ly for\-tu\-nate.”
    \textsuperscript{80}\thinspace{}But those with knowl\-edge said, “Woe to you. God's re\-ward is bet\-ter for those who be\-lieve and do good. On\-ly the stead\-fast will at\-tain it.”
\end{para}

\begin{para}
    \textsuperscript{28:84}\thinspace{}Who\-ev\-er does good will re\-ceive some\-thing bet\-ter, but who\-ev\-er does evil will on\-ly be punished for what he has done.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{29:7}\thinspace{}For those who be\-lieve and do good deeds, We will blot out all their evil and re\-ward them ac\-cord\-ing to their best deeds.
\end{para}

\begin{para}
    \textsuperscript{29:8}\thinspace{}We have commanded man to treat his parents with kind\-ness, but if they try to make you wor\-ship gods be\-side Me, dis\-o\-bey them. You will all re\-turn to me, and I'll tell you the truth a\-bout eve\-ry\-thing you've done.
\end{para}

\begin{para}
    \textsuperscript{29:10}\thinspace{}Some men say, “We be\-lieve in God,” but when they suf\-fer for His cause, they con\-sid\-er per\-se\-cu\-tion from men as if it were God's pun\-ish\-ment. If help co\-mes to you from your Lord, they say, “We've al\-ways been with you.” God knows best what is in everyone's heart.
\end{para}

\begin{para}
    \textsuperscript{29:45}\thinspace{}Pray\-er restrains im\-mor\-al and un\-just be\-hav\-ior.
    \textsuperscript{46}\thinspace{}Don't ar\-gue with followers of ear\-li\-er rev\-e\-la\-tion, but say, “We be\-lieve in what was revealed to us and in what was revealed to you. Our God and your God are the same. We all bow to Him.”
\end{para}

\begin{para}
    \textsuperscript{29:57}\thinspace{}Eve\-ry\-one will taste death, then re\-turn to Us.
    \textsuperscript{58}\thinspace{}To them who be\-lieve and do good, We will give mansions to live in for\-ev\-er in a par\-a\-dise with flow\-ing rivers. The re\-ward is ex\-cel\-lent for those who la\-bor,
    \textsuperscript{59}\thinspace{}are stead\-fast, and trust in their Lord.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{30:21}\thinspace{}A\-mong His signs is this: He cre\-at\-ed spouses for you that you may live in tran\-quil\-i\-ty with them. He put love and mer\-cy be\-tween you. In\-deed, there are signs in this for those who re\-flect.
    \textsuperscript{22}\thinspace{}A\-mong His signs is the cre\-a\-tion of the heavens and the earth, and the di\-ver\-si\-ty of your languages and your col\-ors. In\-deed, in that are signs for those who know.
    \textsuperscript{23}\thinspace{}A\-mong His signs is your sleep, by night and day, and your quest for some of His boun\-ty.
\end{para}

\begin{para}
    \textsuperscript{30:33}\thinspace{}When men en\-coun\-ter trou\-ble, they cry to their Lord and turn to Him for help, but as soon as He gives them a taste of mer\-cy, some of them wor\-ship gods be\-side Him,
    \textsuperscript{34}\thinspace{}show\-ing in\-grat\-i\-tude for what We gave them.
\end{para}

\begin{para}
    \textsuperscript{30:36}\thinspace{}When We give men a taste of mer\-cy, they re\-joice, but when evil afflicts them be\-cause of their own actions, they des\-pair.
\end{para}

\begin{para}
    \textsuperscript{30:38}\thinspace{}Give what is due to relatives, the needy, and the trav\-el\-er. That is best for those who seek God's coun\-te\-nance, and they will pros\-per.
    \textsuperscript{39}\thinspace{}What\-ev\-er you lend in u\-su\-ry, to in\-crease through the prop\-er\-ty of others, has no in\-crease with God, but what\-ev\-er you give in char\-i\-ty, seek\-ing God's coun\-te\-nance, will in\-crease re\-peat\-ed\-ly.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{31:4}\thinspace{}Those who are stead\-fast in pray\-er, spend in char\-i\-ty, and are cer\-tain of the here\-af\-ter,
    \textsuperscript{5}\thinspace{}are guid\-ed by their Lord and will pros\-per.
\end{para}

\begin{para}
    \textsuperscript{31:14}\thinspace{}We have commanded man to be good to his parents. His moth\-er car\-ried him in weak\-ness, and it took two years to wean him. Be grate\-ful to Me and your parents.
    \textsuperscript{15}\thinspace{}But if they try to make you as\-so\-ci\-ate others with Me, dis\-o\-bey them. Yet ac\-com\-pa\-ny them in this life with kind\-ness, and fol\-low the path of those who turn to me. You will all re\-turn to Me in the end, and I will tell you the truth a\-bout all you did.
\end{para}

\begin{para}
    \textsuperscript{31:17}\thinspace{}Be stead\-fast in pray\-er, en\-join what is right, for\-bid what is wrong, and pa\-tient\-ly bear what\-ev\-er happens to you: as\-pire to these things.
    \textsuperscript{18}\thinspace{}Don't turn a\-way from men in pride, nor walk ar\-ro\-gant\-ly through the earth, for God dislikes the ar\-ro\-gant and boast\-ful.
    \textsuperscript{19}\thinspace{}Be mod\-er\-ate in your pa\-ce and low\-er your voice, for the ugliest sound is the braying of as\-ses.
\end{para}

\begin{para}
    \textsuperscript{31:32}\thinspace{}When waves cov\-er them like a can\-o\-py, they call to God, of\-fer\-ing Him sin\-cere de\-vo\-tion. But once He has safe\-ly delivered them to land, some of them wav\-er.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{33:35}\thinspace{}God has pre\-pared for\-give\-ness and a great re\-ward for those who are de\-vot\-ed to God, for the believers, the de\-vout, the hon\-est, the stead\-fast, the hum\-ble, the char\-i\-ta\-ble, those who fast, the chaste, and those who fre\-quent\-ly re\-mem\-ber God.
\end{para}

\begin{para}
    \textsuperscript{33:54}\thinspace{}God has full knowl\-edge of eve\-ry\-thing, wheth\-er you re\-veal or con\-ceal it.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{34:34}\thinspace{}When\-ev\-er we send a warn\-er to a com\-mu\-ni\-ty, the wealth\-y say, “We don't be\-lieve in the mes\-sage you bring.
    \textsuperscript{35}\thinspace{}We have more wealth and chil\-dren than you, and will not be punished.”
\end{para}

\begin{para}
    \textsuperscript{34:37}\thinspace{}Nei\-ther wealth nor chil\-dren will bring you nearer to Us: on\-ly faith and right\-eous\-ness.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{35:45}\thinspace{}If God im\-me\-di\-ate\-ly punished men for their wrong\-do\-ing, He would\-n't leave a sin\-gle crea\-ture on the earth. How\-ev\-er, He defers for a stat\-ed time.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{39:7}\thinspace{}If you are un\-grate\-ful, God has no need of you. He dislikes in\-grat\-i\-tude in His servants. If you are grate\-ful, He is pleased with you. No one will bear another's bur\-den. You will re\-turn to your Lord in the end, and He will in\-form you a\-bout what you did. He knows what is in men's hearts.
\end{para}

\begin{para}
    \textsuperscript{39:10}\thinspace{}Those who do good in this world re\-ceive a good re\-ward. The pa\-tient will re\-ceive a re\-ward with\-out meas\-ure.
\end{para}

\begin{para}
    \textsuperscript{39:49}\thinspace{}When man suffers, he cries out to Us, but when We fa\-vor him, he says, “I've re\-ceived this be\-cause of my own wis\-dom.” This is on\-ly a test, but most don't know it.
\end{para}

\begin{para}
    \textsuperscript{39:69}\thinspace{}The earth will shine with the light of its Lord, the rec\-ord of deeds will be opened, the prophets and witnesses will be brought for\-ward, and fair judg\-ment pro\-nounced be\-tween them. They will not be wronged.
    \textsuperscript{70}\thinspace{}Eve\-ry man will be paid in full for his actions.
    \textsuperscript{71}\thinspace{}Disbelievers will be led to Hell in crowds. When they ar\-rive, its gates will o\-pen, and its keepers will say, “Did\-n't messengers come from a\-mong you, re\-cit\-ing your Lord's messages to you, and warn\-ing you a\-bout today's meet\-ing?”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{40:55}\thinspace{}Be pa\-tient, for God's prom\-ise is true. Ask for\-give\-ness for your sins, and praise your Lord eve\-ning and morn\-ing.
    \textsuperscript{56}\thinspace{}Those who dis\-pute God's messages with\-out au\-thor\-i\-ty have hearts full of pride, which they will nev\-er sat\-is\-fy. Seek ref\-uge in God.
\end{para}

\begin{para}
    \textsuperscript{40:61}\thinspace{}God made the night for you, so that you may rest. He made the day to help you see. God is boun\-ti\-ful to\-wards men, but most men give no thanks.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{41:33}\thinspace{}Who speaks bet\-ter than one who calls men to God, does what is right, and says, “I sub\-mit to God”?
    \textsuperscript{34}\thinspace{}Re\-pel evil with some\-thing bet\-ter, then your en\-e\-my will be\-come like a close friend.
    \textsuperscript{36}\thinspace{}If whispers from Satan stir you up, seek ref\-uge in God.
\end{para}

\begin{para}
    \textsuperscript{41:49}\thinspace{}Man nev\-er tires of asking for good things, but if evil touches him, he gives up all hope and is lost in des\-pair.
    \textsuperscript{50}\thinspace{}When we let him taste mer\-cy af\-ter ad\-ver\-si\-ty, he is sure to say, “This is due to my mer\-it. I don't think that the Hour will come, but if I am brought back to my Lord, I will re\-ceive the best re\-ward.” We will in\-form the disbelievers of eve\-ry\-thing they did, and make them taste a se\-vere tor\-ment.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{42:13}\thinspace{}He has established the same re\-li\-gion for you that He gave to No\-ah, Abraham, Mo\-ses, and Jesus: up\-hold the faith, and make no factions there\-in.
    \textsuperscript{14}\thinspace{}They be\-came di\-vid\-ed through jeal\-ous an\-i\-mos\-i\-ty a\-mong them\-selves.
    \textsuperscript{15}\thinspace{}There\-fore say, “God is our Lord and your Lord. Our deeds for us, and your deeds for you. Let there be no arguments be\-tween us. God will bring us to\-geth\-er, and we will re\-turn to Him.”
\end{para}

\begin{para}
    \textsuperscript{42:27}\thinspace{}If God were to grant a\-bun\-dant pro\-vi\-sion to His servants, they would mis\-be\-have; but he sends down in due meas\-ure what He pleases. He knows His servants and watches o\-ver them.
\end{para}

\begin{para}
    \textsuperscript{42:36}\thinspace{}What you've been giv\-en is for tem\-po\-rar\-y en\-joy\-ment in this life. What is with God is bet\-ter and more last\-ing for those who be\-lieve and trust in their Lord;
    \textsuperscript{37}\thinspace{}a\-void great sins and immoralities; for\-give, e\-ven when an\-gry;
    \textsuperscript{38}\thinspace{}re\-spond to their Lord; are stead\-fast in pray\-er; con\-duct their af\-fairs by mu\-tu\-al con\-sul\-ta\-tion; give from what We have pro\-vid\-ed them;
    \textsuperscript{39}\thinspace{}and de\-fend them\-selves when tyr\-an\-ny afflicts them.
\end{para}

\begin{para}
    \textsuperscript{42:40}\thinspace{}The rec\-om\-pense for harm is an e\-qual harm, but God rewards him who forgives and reconciles.
    \textsuperscript{41}\thinspace{}There is no rea\-son to blame some\-one who defends him\-self af\-ter be\-ing wronged.
    \textsuperscript{43}\thinspace{}It requires great de\-ter\-mi\-na\-tion to be pa\-tient and for\-giv\-ing.
\end{para}

\begin{para}
    \textsuperscript{42:48}\thinspace{}Man rejoices when We let him taste Our mer\-cy, but he is un\-grate\-ful when his own deeds bring mis\-for\-tune.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{43:6}\thinspace{}We sent man\-y prophets a\-mong ear\-li\-er peo\-ple,
    \textsuperscript{7}\thinspace{}but they mocked each one.
\end{para}

\begin{para}
    \textsuperscript{43:22}\thinspace{}They said, “Our fathers followed a cer\-tain re\-li\-gion, and we guide our\-selves by their footsteps.”
    \textsuperscript{23}\thinspace{}In the same way, when\-ev\-er We sent a mes\-sen\-ger to warn a com\-mu\-ni\-ty, the wealth\-y ones said, “Our fathers followed a cer\-tain re\-li\-gion, and we will cer\-tain\-ly fol\-low them.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{45:14}\thinspace{}Tell the believers to for\-give those who don't ex\-pect the Days of God. He a\-lone will re\-pay peo\-ple for what they have earned.
    \textsuperscript{15}\thinspace{}Who\-ev\-er does good benefits him\-self. Who\-ev\-er does evil harms him\-self. You will all re\-turn to your Lord.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{47:2}\thinspace{}For those who have faith, do good, and be\-lieve in the rev\-e\-la\-tion giv\-en to Muhammad, God will re\-move their misdeeds and im\-prove their con\-di\-tion.
\end{para}

\begin{para}
    \textsuperscript{47:36}\thinspace{}If you be\-lieve and fear God, He will give you your rewards. He does\-n't ask you to give up all your possessions.
    \textsuperscript{37}\thinspace{}If He pressed you for them, you would be\-grudge Him, and it would bring out your ill feelings.
\end{para}

\begin{para}
    \textsuperscript{47:38}\thinspace{}You're in\-vit\-ed to give to\-wards God's cause, but some a\-mong you are grudg\-ing. Who\-ev\-er withholds, on\-ly withholds from him\-self. God is free from all wants. You are the needy ones. If you turn a\-way, He'll re\-place you with some\-one else.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{49:6}\thinspace{}Believers, if a trou\-ble\-mak\-er brings you news, ver\-i\-fy it, lest you un\-wit\-ting\-ly harm some\-one, and lat\-er re\-gret what you've done.
\end{para}

\begin{para}
    \textsuperscript{49:11}\thinspace{}Believers, one group should not rid\-i\-cule an\-oth\-er, who may af\-ter all be bet\-ter than them. Don't in\-sult one an\-oth\-er. Don't use of\-fen\-sive nicknames for one an\-oth\-er.
    \textsuperscript{12}\thinspace{}A\-void mak\-ing too man\-y assumptions, for some assumptions are sins. Don't spy on or back\-bite one an\-oth\-er.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{53:29}\thinspace{}A\-void those who turn a\-way from Our mes\-sage and on\-ly de\-sire a world\-ly life.
    \textsuperscript{30}\thinspace{}That is the ex\-tent of their knowl\-edge.
    \textsuperscript{31}\thinspace{}Eve\-ry\-thing in the heavens and on earth belongs to God. He repays those who do evil, ac\-cord\-ing to their deeds, and those who do good, with what is best.
    \textsuperscript{32}\thinspace{}Your Lord is am\-ple in for\-give\-ness to those who a\-void great sins and shame\-ful deeds, though they may com\-mit small sins.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{57:7}\thinspace{}Be\-lieve in God and His mes\-sen\-ger, and give out of what you in\-her\-it from Him. Those who be\-lieve and give will re\-ceive a great re\-ward. 
    \textsuperscript{20}\thinspace{}This life is like rain: the growth it causes delights the tillers, but soon it with\-ers. You see it yel\-low, then dry and crum\-ble a\-way. World\-ly life is on\-ly a pleas\-ant il\-lu\-sion.
    \textsuperscript{21}\thinspace{}Race for your Lord's for\-give\-ness and for a gar\-den as wide as heav\-en and earth, pre\-pared for those who be\-lieve in God and His messengers.
\end{para}

\begin{para}
    \textsuperscript{57:23}\thinspace{}You don't need to des\-pair o\-ver what eludes you or gloat o\-ver what you re\-ceive. For God dislikes those who are con\-ceit\-ed, boast\-ful,
    \textsuperscript{24}\thinspace{}sting\-y, and en\-join stin\-gi\-ness on others.
\end{para}

\begin{para}
    \textsuperscript{57:26}\thinspace{}We sent No\-ah and Abraham, and gave prophethood and scrip\-ture to their descendants. Some of them were right\-ly guid\-ed, but man\-y were transgressors.
    \textsuperscript{27}\thinspace{}We sent oth\-er messengers in their footsteps. We sent Jesus the son of Mar\-y, gave him the Gos\-pel, and put com\-pas\-sion and mer\-cy in\-to the hearts of his followers. But We did not pre\-scribe mo\-nas\-ti\-cism for them; they invented that them\-selves. We gave the believers a\-mong them their re\-ward, but man\-y of them were transgressors.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{61:2}\thinspace{}Believers, why do you say one thing and do an\-oth\-er?
    \textsuperscript{3}\thinspace{}It is most hate\-ful in God's sight.
\end{para}

\begin{para}
    \textsuperscript{61:7}\thinspace{}Who does greater wrong than one who invents a lie a\-gainst God while he is be\-ing in\-vit\-ed to Islam?
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{62:9}\thinspace{}Believers, when the call to pray\-er is proclaimed on Friday, hur\-ry to God's rememberance, and leave your com\-merce. That is best for you if on\-ly you knew.
    \textsuperscript{10}\thinspace{}When the pray\-er is fin\-ished, dis\-perse through\-out the land, seek God's boun\-ty, and re\-mem\-ber Him of\-ten, so that you may pros\-per.
\end{para}

\begin{para}
    \textsuperscript{62:11}\thinspace{}Muhammad, when they see com\-merce or a\-muse\-ment, they rush to\-wards it, and leave you stand\-ing. Say, “What God has is bet\-ter than any a\-muse\-ment or com\-merce. God is the best pro\-vid\-er.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{68:7}\thinspace{}Your Lord knows best who strays from His path and who receives guid\-ance,
    \textsuperscript{8}\thinspace{}so don't o\-bey the deniers.
    \textsuperscript{10}\thinspace{}Don't o\-bey a worth\-less oath-maker,
    \textsuperscript{11}\thinspace{}a slanderer, a ma\-li\-cious gos\-sip,
    \textsuperscript{12}\thinspace{}one who hinders good, a transgressor, sin\-ful,
    \textsuperscript{13}\thinspace{}cru\-el, a pre\-tend\-er.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{70:19}\thinspace{}Man was cre\-at\-ed anx\-ious.
    \textsuperscript{20}\thinspace{}He's fret\-ful when mis\-for\-tune touches him,
    \textsuperscript{21}\thinspace{}and sting\-y when good for\-tune touches him.
    \textsuperscript{22}\thinspace{}Ex\-cept for those de\-vot\-ed to pray\-er
    \textsuperscript{23}\thinspace{}who re\-main con\-stant in their prayers,
    \textsuperscript{24}\thinspace{}and who give a share of their wealth
    \textsuperscript{25}\thinspace{}to beggars and the de\-prived;
    \textsuperscript{26}\thinspace{}those who be\-lieve in the Day of Judg\-ment;
    \textsuperscript{27}\thinspace{}those who fear their Lord's pun\-ish\-ment,
    \textsuperscript{28}\thinspace{}for none is safe from it;
    \textsuperscript{29}\thinspace{}those who guard their chas\-ti\-ty,
    \textsuperscript{30}\thinspace{}ex\-cept with their wives;
    \textsuperscript{32}\thinspace{}those who are at\-ten\-tive to their trusts and pledges;
    \textsuperscript{33}\thinspace{}those who bear wit\-ness hon\-est\-ly;
    \textsuperscript{34}\thinspace{}those who guard their prayers.
    \textsuperscript{35}\thinspace{}They will be honored in the Gardens.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{73:7}\thinspace{}You are bus\-y for much of the day,
    \textsuperscript{8}\thinspace{}but re\-mem\-ber your Lord and de\-vote your\-self to Him whole\-heart\-ed\-ly.
    \textsuperscript{11}\thinspace{}Let Me deal with the deniers who live in lux\-u\-ry. Bear with them for a lit\-tle while.
\end{para}

\begin{para}
    \textsuperscript{73:20}\thinspace{}Re\-cite as much of the Qur'an as may be eas\-y for you. He knows that some of you are sick, others trav\-el through\-out the land seek\-ing God's boun\-ty, and others fight in God's cause. There\-fore, re\-cite as much of the Qur'an as may be eas\-y for you, es\-tab\-lish reg\-u\-lar pray\-er, and spend in char\-i\-ty.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{74:3}\thinspace{}Glo\-ri\-fy your Lord,
    \textsuperscript{4}\thinspace{}pu\-ri\-fy your\-self,
    \textsuperscript{5}\thinspace{}and shun all filth.
    \textsuperscript{6}\thinspace{}Don't be\-stow fa\-vors to re\-ceive in\-crease,
    \textsuperscript{7}\thinspace{}but be pa\-tient for your Lord.
\end{para}

\begin{para}
    \textsuperscript{74:11}\thinspace{}Let Me a\-lone deal with him whom I a\-lone cre\-at\-ed,
    \textsuperscript{12}\thinspace{}to whom I gave vast wealth
    \textsuperscript{13}\thinspace{}and chil\-dren by his side,
    \textsuperscript{14}\thinspace{}for whom I made life com\-fort\-a\-ble,
    \textsuperscript{15}\thinspace{}yet he still wants Me to give him more.
\end{para}

\begin{para}
    \textsuperscript{74:40}\thinspace{}The companions of the right will be in gardens asking one an\-oth\-er
    \textsuperscript{41}\thinspace{}a\-bout the sinners,
    \textsuperscript{42}\thinspace{}“What led you in\-to hell-fire?”
    \textsuperscript{43}\thinspace{}They will an\-swer, “We nei\-ther prayed
    \textsuperscript{44}\thinspace{}nor fed the poor,
    \textsuperscript{45}\thinspace{}but we in\-dulged in vain con\-ver\-sa\-tion with others
    \textsuperscript{46}\thinspace{}and de\-nied the Day of Judg\-ment
    \textsuperscript{47}\thinspace{}un\-til cer\-tain\-ty came up\-on us.”
\end{para}

\begin{para}
    \textsuperscript{74:49}\thinspace{}Why do they turn a\-way from ad\-mo\-ni\-tion
    \textsuperscript{50}\thinspace{}like fright\-ened donkeys
    \textsuperscript{51}\thinspace{}flee\-ing from a li\-on?
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{76:7}\thinspace{}The right\-eous ful\-fill their vows; they fear a day of wide\-spread evil;
    \textsuperscript{8}\thinspace{}they feed the poor, the or\-phan, and the cap\-tive,
    \textsuperscript{9}\thinspace{}say\-ing, “We feed you for the sa\-ke of God a\-lone. We seek nei\-ther re\-ward nor grat\-i\-tude from you.”
\end{para}

\begin{para}
    \textsuperscript{76:24}\thinspace{}Pa\-tient\-ly a\-wait your Lord's judg\-ment, and don't o\-bey the sin\-ner or dis\-be\-liev\-er.
    \textsuperscript{25}\thinspace{}Re\-mem\-ber your Lord's name morn\-ing and eve\-ning.
    \textsuperscript{26}\thinspace{}Bow to Him dur\-ing the night, and glo\-ri\-fy Him through\-out the long night.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{80:1}\thinspace{}Muhammad frowned and turned a\-way
    \textsuperscript{2}\thinspace{}be\-cause the blind man came to him.
    \textsuperscript{3}\thinspace{}For all you know, he might be\-come pure,
    \textsuperscript{4}\thinspace{}or be reminded of some\-thing use\-ful.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{81:1}\thinspace{}When the sun is wrapped in dark\-ness,
    \textsuperscript{2}\thinspace{}when the stars fall,
    \textsuperscript{3}\thinspace{}when the mountains van\-ish,
    \textsuperscript{8}\thinspace{}when the ba\-by girl bur\-ied a\-live is asked
    \textsuperscript{9}\thinspace{}for what sin she was killed,
    \textsuperscript{10}\thinspace{}when the pages are laid o\-pen,
    \textsuperscript{14}\thinspace{}then eve\-ry per\-son will know what he has pre\-pared for him\-self.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{83:2}\thinspace{}Woe to those who de\-mand a full meas\-ure for them\-selves
    \textsuperscript{3}\thinspace{}but give to others less than what's due.
    \textsuperscript{4}\thinspace{}Do they not think that they will be resurrected,
    \textsuperscript{6}\thinspace{}on a Day when all men will stand be\-fore the Lord of the worlds?
\end{para}

\begin{para}
    \textsuperscript{83:29}\thinspace{}Sinners used to laugh at believers,
    \textsuperscript{30}\thinspace{}wink at each oth\-er when they passed by,
    \textsuperscript{31}\thinspace{}joke a\-bout them when they returned to their own peo\-ple,
    \textsuperscript{32}\thinspace{}and say, “These peo\-ple are tru\-ly lost,”
    \textsuperscript{33}\thinspace{}but they were not sent as guardians o\-ver them.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{87:8}\thinspace{}We will make the path to\-wards ease eas\-y for you.
    \textsuperscript{9}\thinspace{}So re\-mind others, if a reminder will help them.
    \textsuperscript{10}\thinspace{}Those who fear God will re\-ceive the reminder,
    \textsuperscript{11}\thinspace{}but the wick\-ed will a\-void it.
\end{para}

\begin{para}
    \textsuperscript{87:14}\thinspace{}Those will pros\-per who pu\-ri\-fy them\-selves,
    \textsuperscript{15}\thinspace{}re\-mem\-ber the name of their Lord, and pray.
    \textsuperscript{16}\thinspace{}Yet you pre\-fer the life of this world
    \textsuperscript{17}\thinspace{}e\-ven though the Here\-af\-ter is bet\-ter and more en\-dur\-ing.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{89:15}\thinspace{}As for man, when his Lord tries him by giv\-ing him hon\-or and gifts, he says “My Lord has honored me.”
    \textsuperscript{16}\thinspace{}But when He tries him by restricting his pro\-vi\-sion, he says, “My Lord has hu\-mil\-i\-at\-ed me.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{92:4}\thinspace{}You all strive for di\-verse goals.
    \textsuperscript{5}\thinspace{}For him who gives, fears God,
    \textsuperscript{6}\thinspace{}and believes in the best,
    \textsuperscript{7}\thinspace{}We will smooth his path to\-wards ease.
    \textsuperscript{8}\thinspace{}But for him who is sting\-y, considers him\-self self-sufficient,
    \textsuperscript{9}\thinspace{}and denies the best,
    \textsuperscript{10}\thinspace{}We will smooth his path to\-wards dif\-fi\-cul\-ty,
    \textsuperscript{11}\thinspace{}and his wealth will not help him when he falls.
\end{para}

\begin{para}
    \textsuperscript{92:17}\thinspace{}The right\-eous will re\-main far from hell.
    \textsuperscript{18}\thinspace{}Those who spend their wealth on self-purification,
    \textsuperscript{19}\thinspace{}not to re\-turn a fa\-vor
    \textsuperscript{20}\thinspace{}but on\-ly seek\-ing their Lord's face,
    \textsuperscript{21}\thinspace{}will be sat\-is\-fied.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{93:6}\thinspace{}Muhammad, did He not find you as an or\-phan and shel\-ter you,
    \textsuperscript{7}\thinspace{}find you lost and guide you,
    \textsuperscript{8}\thinspace{}find you in need and make you rich?
    \textsuperscript{9}\thinspace{}There\-fore, don't op\-press the or\-phan,
    \textsuperscript{10}\thinspace{}nor re\-pel the beg\-gar,
    \textsuperscript{11}\thinspace{}but talk a\-bout your Lord's blessings.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{96:3}\thinspace{}Your Lord is the most Boun\-ti\-ful,
    \textsuperscript{4}\thinspace{}who taught by the pen,
    \textsuperscript{5}\thinspace{}who taught man what he did\-n't know.
    \textsuperscript{6}\thinspace{}But man transgresses
    \textsuperscript{7}\thinspace{}by con\-sid\-er\-ing him\-self self-sufficient.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{104:1}\thinspace{}Woe to eve\-ry slanderer and back\-bit\-er
    \textsuperscript{2}\thinspace{}who amasses wealth and con\-tin\-u\-ous\-ly counts it,
    \textsuperscript{3}\thinspace{}think\-ing that his wealth will make him im\-mor\-tal.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{107:1}\thinspace{}Have you con\-sid\-ered the man who denies the Judg\-ment?
    \textsuperscript{2}\thinspace{}He is the one who pushes the or\-phan a\-way,
    \textsuperscript{3}\thinspace{}and does\-n't en\-cour\-age the feed\-ing of the poor.
    \textsuperscript{4}\thinspace{}So woe to those who pray
    \textsuperscript{5}\thinspace{}but are heed\-less of their pray\-er;
    \textsuperscript{6}\thinspace{}who on\-ly want to be seen,
    \textsuperscript{7}\thinspace{}and with\-hold small kindnesses.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{109:1}\thinspace{}Muhammad, say, “Disbelievers,
    \textsuperscript{2}\thinspace{}I don't wor\-ship what you wor\-ship,
    \textsuperscript{3}\thinspace{}nor do you wor\-ship what I wor\-ship.
    \textsuperscript{6}\thinspace{}You have your re\-li\-gion and I have mine.”
\end{para}


\end{document}

% Copyright 2021 Michael Hendricks
%
% Permission is granted to use, copy, modify, and distribute
% this work for any purpose with or without fee.
%
% This work is provided as-is without any express or implied
% warranty. In no event is the author liable for any damages
% arising from this work.
