%
% see tex.stackexchange.com/a/600/151276
\def\changemargin#1#2{\list{}{\rightmargin#2\leftmargin#1}\item[]}
\let\endchangemargin=\endlist

\documentclass[openany,12pt,english]{book}
\usepackage[utf8]{inputenc}
\usepackage{babel}
\usepackage{microtype}
\usepackage{csquotes}

\usepackage[
  paperwidth=8.5in,
  paperheight=11in,
  top=1in,
  outer=1in,
  bottom=1.25in,
  inner=1.5in
]{geometry}

\title{Muhammad}
\author{}
\date{}

\newenvironment{para}{\par\pretolerance=100\tolerance=200\setlength{\emergencystretch}{0.6em}\relax}{\par}

\begin{document}

\maketitle

% Copyright page
{
\thispagestyle{empty}
\vspace*{\fill}
\begin{changemargin}{.8in}{.8in}
\begin{center}
\textcopyright{}~2021 Michael Hendricks

\vspace{0.5\baselineskip}

Permission is granted to use, copy, modify, and distribute
this work for any purpose with or without fee.

\vspace{0.5\baselineskip}

This work is provided as-is without any express or implied
warranty. In no event is the author liable for any damages
arising from this work.
\end{center}
\end{changemargin}
}
\pagenumbering{arabic}
\clearpage{}

% Table of Contents
\clearpage{}

\markboth{}{}
\twocolumn[\begin{changemargin}{.6in}{.6in}
  
  \begin{center}
    \itshape
    An a\-bridg\-ment of the Qur'an.
    Verse references such
    as~\textsuperscript{2:2}\thinspace{}indicate
    the original source of the text, but
    don't imply a full quotation.
  \end{center}
  
\end{changemargin}]{}


\begin{para}
    \textsuperscript{1:1}\thinspace{}In the name of God, Most Gra\-cious, Most Mer\-ci\-ful.
    \textsuperscript{2}\thinspace{}All praise is owed to God, sus\-tain\-er of the worlds,
    \textsuperscript{4}\thinspace{}Mas\-ter of the Day of Judg\-ment.
    \textsuperscript{5}\thinspace{}You a\-lone we wor\-ship, and you a\-lone we ask for help.
    \textsuperscript{6}\thinspace{}Guide us to the straight path;
    \textsuperscript{7}\thinspace{}the path of those you have bless\-ed, not of those who earned your an\-ger, nor of those who stray.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{2:2}\thinspace{}This book is a guid\-ance for those con\-scious of God,
    \textsuperscript{3}\thinspace{}who be\-lieve in the un\-seen, are stead\-fast in pray\-er, give out of what We have pro\-vid\-ed them,
    \textsuperscript{4}\thinspace{}be\-lieve in the rev\-e\-la\-tion sent to you, Muhammad, and what was sent be\-fore your time, and who have firm faith in the here\-af\-ter.
    \textsuperscript{5}\thinspace{}They are fol\-low\-ing guid\-ance from their Lord, and they will pros\-per.
    \textsuperscript{6}\thinspace{}As for those who dis\-be\-lieve, it is the same to them wheth\-er you warn them or not: they won't be\-lieve.
\end{para}

\begin{para}
    \textsuperscript{2:42}\thinspace{}Don't mix truth with false\-hood, nor know\-ing\-ly con\-ceal the truth.
    \textsuperscript{43}\thinspace{}Be stead\-fast in pray\-er, spend in char\-i\-ty, and bow down with those who bow down in wor\-ship.
\end{para}

\begin{para}
    \textsuperscript{2:44}\thinspace{}Do you en\-join right\-eous\-ness on the peo\-ple and for\-get to do it your\-selves, e\-ven though you re\-cite the Scrip\-ture? Have you no sense?
    \textsuperscript{45}\thinspace{}Seek help with pa\-tience and pray\-er. It is dif\-fi\-cult in\-deed, ex\-cept for the hum\-ble,
    \textsuperscript{46}\thinspace{}who are cer\-tain that they will meet their Lord and re\-turn to Him.
\end{para}

\begin{para}
    \textsuperscript{2:62}\thinspace{}All who be\-lieve in God and the Last Day, and do right\-eous\-ness, will have their re\-ward with their Lord. They will have no fear, nor will they grieve.
\end{para}

\begin{para}
    \textsuperscript{2:79}\thinspace{}Woe to those who write a book with their own hands then say, “This is from God,” in or\-der to sell it for a small price!
\end{para}

\begin{para}
    \textsuperscript{2:111}\thinspace{}They say, “None will en\-ter Par\-a\-dise un\-less he is a Jew or a Christian.”
    \textsuperscript{112}\thinspace{}In fact, who\-ev\-er submits him\-self whol\-ly to God and does good will have his re\-ward with his Lord.
\end{para}

\begin{para}
    \textsuperscript{2:135}\thinspace{}They say, “Be\-come Jews or Christians, then you will be right\-ly guid\-ed.”
    \textsuperscript{136}\thinspace{}Say, “We be\-lieve in God, and the rev\-e\-la\-tion giv\-en to us, and to Abraham, Ishmael, Isaac, Jacob, and their descendants, and the rev\-e\-la\-tion giv\-en to Mo\-ses, Jesus, and the prophets from their Lord. We make no dis\-tinc\-tion be\-tween any of them, and we sub\-mit to Him.”
\end{para}

\begin{para}
    \textsuperscript{2:153}\thinspace{}Believers, seek help through pa\-tience and pray\-er. For God is with the stead\-fast.
\end{para}

\begin{para}
    \textsuperscript{2:172}\thinspace{}Believers, eat the good things We have pro\-vid\-ed for you, and be grate\-ful to God, if it is Him you wor\-ship.
\end{para}

\begin{para}
    \textsuperscript{2:177}\thinspace{}Right\-eous\-ness is not turn\-ing your faces to\-wards east or west, but right\-eous\-ness is one who believes in God, the Last Day, the angels, the Book, and the prophets; who gives of his wealth, e\-ven though he loves it, to relatives, orphans, the needy, the trav\-el\-er, beggars, and for free\-ing slaves; who are stead\-fast in pray\-er, and spend in char\-i\-ty; who keep their promises; and are pa\-tient in pov\-er\-ty, ad\-ver\-si\-ty, and dur\-ing bat\-tle. These ones are true, and con\-scious of God.
\end{para}

\begin{para}
    \textsuperscript{2:180}\thinspace{}When death approaches any of you and he leaves be\-hind wealth, it is pre\-scribed that he make a prop\-er be\-quest to parents and near relatives. This is a du\-ty for those con\-scious of God.
    \textsuperscript{181}\thinspace{}If any\-one alters the be\-quest af\-ter hear\-ing it, the sin is on them: God hears and knows all.
    \textsuperscript{182}\thinspace{}But if any\-one fears that the tes\-ta\-tor made a mis\-take or did wrong, and makes peace be\-tween the heirs, he incurs no sin: God is for\-giv\-ing and mer\-ci\-ful.
\end{para}

\begin{para}
    \textsuperscript{2:183}\thinspace{}Believers, fasting is pre\-scribed for you, as it was for those be\-fore you, that you may be\-come con\-scious of God.
    \textsuperscript{184}\thinspace{}Fast for a spe\-cif\-ic num\-ber of days, but if you are sick or trav\-el\-ing, make up the days lat\-er. Those who can on\-ly fast with dif\-fi\-cul\-ty may feed a poor per\-son in\-stead. Who\-ev\-er volunteers to give more, it is bet\-ter for him.
    \textsuperscript{185}\thinspace{}In the month of Ramadan the Qur'an was revealed as a guid\-ance for man\-kind, as a clear sign of that guid\-ance, and as judg\-ment be\-tween right and wrong. So who\-ev\-er is pres\-ent dur\-ing that month should fast. God wants ease for you, not hard\-ship; and per\-haps you will be grate\-ful.
\end{para}

\begin{para}
    \textsuperscript{2:188}\thinspace{}Don't con\-sume one another's prop\-er\-ty un\-just\-ly, nor use it to bribe judges, intending to con\-sume some of the prop\-er\-ty of others sin\-ful\-ly and know\-ing\-ly.
\end{para}

\begin{para}
    \textsuperscript{2:215}\thinspace{}What\-ev\-er you spend in char\-i\-ty should be for parents, close relatives, orphans, the poor, and for travelers. What\-ev\-er good you do, God knows it well.
\end{para}

\begin{para}
    \textsuperscript{2:216}\thinspace{}You may dis\-like some\-thing that's good for you, and love some\-thing that's bad for you. God knows, and you do not.
\end{para}

\begin{para}
    \textsuperscript{2:219}\thinspace{}They ask you a\-bout wine and gambling. Say, “In them is great sin, yet some ben\-e\-fit, but the sin is greater than the ben\-e\-fit.” They ask you how much they should give. Say, “The ex\-cess.”
    \textsuperscript{220}\thinspace{}They ask you a\-bout orphans. Say, “Im\-prov\-ing their con\-di\-tion is best. God knows who spoils things and who improves them. If God had wished, He could have put you in\-to difficul\-ties.”
\end{para}

\begin{para}
    \textsuperscript{2:221}\thinspace{}Don't mar\-ry i\-dol\-a\-trous wom\-en un\-til they be\-lieve. A slave who believes is bet\-ter than an i\-dol\-a\-tress, e\-ven if she pleases you. These in\-vite you to the fire.
\end{para}

\begin{para}
    \textsuperscript{2:238}\thinspace{}Strict\-ly guard your hab\-it of prayers. Stand be\-fore God in de\-vout o\-be\-di\-ence.
    \textsuperscript{239}\thinspace{}If you are in dan\-ger, pray walk\-ing or rid\-ing, but when you are se\-cure, re\-mem\-ber God, for He taught you what you did not know.
\end{para}

\begin{para}
    \textsuperscript{2:256}\thinspace{}There shall be no com\-pul\-sion in re\-li\-gion. The right path has be\-come clear from the wrong. Who\-ev\-er rejects false gods and believes in God has grasped the most trust\-wor\-thy hand\-hold that nev\-er breaks.
\end{para}

\begin{para}
    \textsuperscript{2:261}\thinspace{}Those who spend their wealth in God's cause are like a grain of corn which grows sev\-en ears, and each ear has a hun\-dred grains.
    \textsuperscript{262}\thinspace{}Those who spend their wealth in God's cause will have their re\-ward with their Lord: they will have no fear, nor will they grieve.
\end{para}

\begin{para}
    \textsuperscript{2:263}\thinspace{}Kind words and for\-give\-ness are bet\-ter than char\-i\-ty followed by in\-ju\-ry.
    \textsuperscript{264}\thinspace{}Believers, don't can\-cel your char\-i\-ty with reminders of your gen\-er\-os\-i\-ty or with in\-ju\-ry, like those who spend their wealth to be seen of men, but do not be\-lieve in God or the Last Day. They are like a smooth rock with dust on it. Heav\-y rain falls, leav\-ing it bare. They are un\-a\-ble to keep any\-thing they have earned. God does not guide disbelievers.
    \textsuperscript{265}\thinspace{}Those who spend their wealth seek\-ing to please God are like a gar\-den on high ground. Heav\-y rain falls, yield\-ing a dou\-ble har\-vest. If heav\-y rain does\-n't fall, light rain will.
\end{para}

\begin{para}
    \textsuperscript{2:267}\thinspace{}Believers, give from the good things which you have earned. Do not give from what is bad, when you your\-selves would on\-ly ac\-cept it if your eyes were closed.
\end{para}

\begin{para}
    \textsuperscript{2:268}\thinspace{}Satan threatens you with pov\-er\-ty and orders you to com\-mit im\-mo\-ral\-i\-ty. God promises you His for\-give\-ness and boun\-ty.
\end{para}

\begin{para}
    \textsuperscript{2:270}\thinspace{}What\-ev\-er you give, or vow to give, God knows it. But the wrongdoers will have no helpers.
    \textsuperscript{271}\thinspace{}If you dis\-close your char\-i\-ty, it is good, but if you con\-ceal it and give it to the poor, it is bet\-ter for you. It will a\-tone for some of your bad deeds.
\end{para}

\begin{para}
    \textsuperscript{2:272}\thinspace{}What\-ev\-er char\-i\-ty you give benefits your own soul when you do it on\-ly to seek the face of God. What\-ev\-er you give will be repaid to you in full, and you will not be wronged.
    \textsuperscript{273}\thinspace{}Give to the poor who are whol\-ly occupied in God's cause, un\-a\-ble to trav\-el in the land for work. The un\-know\-ing might think they are rich be\-cause of their re\-straint, but you will rec\-og\-nize them be\-cause they don't beg per\-sist\-ent\-ly. What\-ev\-er good you give, be as\-sured that God knows it.
\end{para}

\begin{para}
    \textsuperscript{2:274}\thinspace{}Those who give of their possessions by night and by day, pri\-vate\-ly and pub\-lic\-ly, will have their re\-ward with their Lord: they will have no fear, nor will they grieve.
    \textsuperscript{277}\thinspace{}Those who be\-lieve, do good works, are stead\-fast in pray\-er, and spend in char\-i\-ty, will have their re\-ward with their Lord: they will have no fear, nor will they grieve.
\end{para}

\begin{para}
    \textsuperscript{2:282}\thinspace{}When you con\-tract a debt for a spec\-i\-fied term, write it down. Let the debt\-or write, and not di\-min\-ish what he owes. Bring two witnesses. They must not ref\-use when called to tes\-ti\-fy. Do not dis\-dain to write down your con\-tract, wheth\-er it be small or large, a\-long with the time it falls due: this is more just with God, more re\-li\-a\-ble as ev\-i\-dence, and more like\-ly to pre\-vent doubt be\-tween you. If it be an im\-me\-di\-ate trade, there is no blame if you don't write it down. Have witnesses pres\-ent when you trade with one an\-oth\-er.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{3:3}\thinspace{}He sent down to you the Book with the truth, confirming what went be\-fore it. He sent down the Torah and the Gos\-pel
    \textsuperscript{4}\thinspace{}as a guid\-ance for the peo\-ple.
\end{para}

\begin{para}
    \textsuperscript{3:14}\thinspace{}Al\-lur\-ing to men is the love of their desires: wom\-en, chil\-dren, heap\-ing treasures of gold and sil\-ver, fi\-ne hors\-es, cat\-tle, and farm\-land. These are pleasures of life in this world, but with God is the best place to re\-turn.
\end{para}

\begin{para}
    \textsuperscript{3:92}\thinspace{}You will on\-ly at\-tain true pi\-e\-ty if you give out of what you love. What\-ev\-er you give, tru\-ly, God knows it.
\end{para}

\begin{para}
    \textsuperscript{3:137}\thinspace{}Situations sim\-i\-lar to yours have happened be\-fore: trav\-el through the earth, and see how it ended for the disbelievers.
\end{para}

\begin{para}
    \textsuperscript{3:156}\thinspace{}Believers, don't say of your brothers who trav\-eled or went out to fight, “If they had stayed with us, they would not have died.” This will be\-come a source of re\-gret in your hearts.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{4:2}\thinspace{}Give orphans their prop\-er\-ty. Don't re\-place their good things with your bad things, nor con\-sume their prop\-er\-ty to\-geth\-er with your own. Sure\-ly, this is a great sin.
\end{para}

\begin{para}
    \textsuperscript{4:3}\thinspace{}Mar\-ry wom\-en of your choice, two or three or four. If you fear that you will not treat them fair\-ly, then on\-ly one. That will help you a\-void bi\-as.
    \textsuperscript{4}\thinspace{}Give the wom\-en their dow\-er gra\-cious\-ly. If they will\-ing\-ly re\-mit any of it to you, en\-joy it with a clear con\-science.
\end{para}

\begin{para}
    \textsuperscript{4:5}\thinspace{}Do not en\-trust your prop\-er\-ty to the weak-minded. God made it a means of sup\-port for you, but feed and clothe them with it, and speak to them with kind\-ness.
\end{para}

\begin{para}
    \textsuperscript{4:6}\thinspace{}Test orphans un\-til they reach mar\-riage\-a\-ble age. If they have sound judg\-ment, re\-lease their prop\-er\-ty to them. Don't con\-sume it wastefully or hast\-i\-ly be\-fore they grow up. If the guard\-i\-an is wealth\-y, he should take no com\-pen\-sa\-tion. If he is poor, he should use on\-ly what is rea\-son\-a\-ble. When you re\-lease their prop\-er\-ty to them, bring witnesses.
\end{para}

\begin{para}
    \textsuperscript{4:7}\thinspace{}Men and wom\-en shall each have a share of what parents and close relatives leave, wheth\-er the prop\-er\-ty is small or large.
    \textsuperscript{8}\thinspace{}If oth\-er relatives, or orphans, or poor are pres\-ent dur\-ing the di\-vi\-sion, then give them some\-thing from the es\-tate, and speak to them with kind\-ness.
    \textsuperscript{9}\thinspace{}Let executors have the same fear in their minds as if they had left help\-less chil\-dren be\-hind. Let them fear God, and speak just\-ly.
\end{para}

\begin{para}
    \textsuperscript{4:18}\thinspace{}Re\-pent\-ance has no ef\-fect for those who con\-tin\-ue to do evil un\-til death confronts them and they say, “Now I re\-pent.”
\end{para}

\begin{para}
    \textsuperscript{4:19}\thinspace{}Believers, you are for\-bid\-den to in\-her\-it wom\-en a\-gainst their will. Don't treat your wives harsh\-ly. Live with them in kind\-ness. If you dis\-like them, it may be that you dis\-like some\-thing in which God has placed much good.
\end{para}

\begin{para}
    \textsuperscript{4:22}\thinspace{}Don't mar\-ry wom\-en whom your fathers mar\-ried,
    \textsuperscript{23}\thinspace{}wom\-en who are close relatives, your wet nurses, your milk sisters, your wives' mothers, your stepdaughters, those whom your sons mar\-ried, two sisters at the same time,
    \textsuperscript{24}\thinspace{}and wom\-en al\-read\-y mar\-ried. All oth\-er wom\-en are law\-ful to you, pro\-vid\-ed that you seek them in mar\-riage, with dow\-er from your prop\-er\-ty, de\-sir\-ing mar\-riage not for\-ni\-ca\-tion.
\end{para}

\begin{para}
    \textsuperscript{4:29}\thinspace{}Believers, don't con\-sume one another's wealth un\-just\-ly, but let there be trade by mu\-tu\-al con\-sent. Don't kill your\-selves or one an\-oth\-er, for God is mer\-ci\-ful to you.
\end{para}

\begin{para}
    \textsuperscript{4:31}\thinspace{}If you a\-void the great sins which you are for\-bid\-den, We will re\-move your mi\-nor sins, and ad\-mit you through the en\-trance of hon\-or.
    \textsuperscript{32}\thinspace{}Don't cov\-et what God has bestowed more free\-ly on some than on others. Men and wom\-en re\-ceive a share of what they earn.
\end{para}

\begin{para}
    \textsuperscript{4:36}\thinspace{}Wor\-ship God, and join noth\-ing with Him. Do good to parents, relatives, orphans, the poor, neighbors, the com\-pan\-ion by your side, and travelers. God does not like the ar\-ro\-gant, the boast\-ful,
    \textsuperscript{37}\thinspace{}the sting\-y; those who com\-mand others to be sting\-y, hid\-ing the boun\-ty God has giv\-en them;
    \textsuperscript{38}\thinspace{}those who spend their wealth to be seen by men, and don't be\-lieve in God nor the Last Day.
    \textsuperscript{39}\thinspace{}What harm would it do them to be\-lieve in God and the Last Day, and to spend out of what God has pro\-vid\-ed them?
    \textsuperscript{40}\thinspace{}God wrongs no one in the least de\-gree. If there is a good deed, He dou\-bles it, and gives a great re\-ward.
\end{para}

\begin{para}
    \textsuperscript{4:58}\thinspace{}God commands you to re\-turn what you've been entrusted to its right\-ful own\-er. When you judge be\-tween peo\-ple, do so with jus\-tice.
    \textsuperscript{59}\thinspace{}Believers, o\-bey God, the Mes\-sen\-ger, and those in au\-thor\-i\-ty a\-mong you.
\end{para}

\begin{para}
    \textsuperscript{4:85}\thinspace{}Who\-ev\-er intercedes for a good cause will share in its re\-ward. Who\-ev\-er intercedes for an evil cause will share in its bur\-den.
\end{para}

\begin{para}
    \textsuperscript{4:86}\thinspace{}When offered a greet\-ing, re\-spond with a bet\-ter one, or at least its e\-qual. God takes ac\-count of eve\-ry\-thing.
\end{para}

\begin{para}
    \textsuperscript{4:95}\thinspace{}God prefers the believers who strive with their wealth and lives o\-ver those who sit --- oth\-er than the dis\-a\-bled. Al\-though He has prom\-ised good to all of them, those who strive re\-ceive a greater re\-ward than those who sit.
\end{para}

\begin{para}
    \textsuperscript{4:100}\thinspace{}Who\-ev\-er emigrates for God's cause will find on the earth a\-bun\-dance, and man\-y places of ref\-uge. If he then dies as a ref\-u\-gee, his re\-ward from God is cer\-tain.
\end{para}

\begin{para}
    \textsuperscript{4:101}\thinspace{}When you trav\-el, it's no sin to short\-en your prayers, if you fear that disbelievers may at\-tack you.
    \textsuperscript{103}\thinspace{}When you have com\-plet\-ed the pray\-er, re\-mem\-ber God: stand\-ing, sit\-ting, and ly\-ing down. Once you are se\-cure, re\-es\-tab\-lish reg\-u\-lar pray\-er, for pray\-er is man\-da\-to\-ry for believers at spe\-cif\-ic times.
\end{para}

\begin{para}
    \textsuperscript{4:110}\thinspace{}One who does evil or wrongs him\-self, but then asks God for for\-give\-ness, will find God is for\-giv\-ing and mer\-ci\-ful.
    \textsuperscript{112}\thinspace{}But if one sins then blames an in\-no\-cent per\-son, he burdens him\-self with a slan\-der and a fla\-grant sin.
\end{para}

\begin{para}
    \textsuperscript{4:123}\thinspace{}Any\-one who does wrong will be repaid ac\-cord\-ing\-ly and will find no pro\-tec\-tor or help\-er be\-sides God.
    \textsuperscript{124}\thinspace{}Any be\-liev\-er, male or fe\-male, who does good deeds will en\-ter heav\-en, and will not re\-ceive the least in\-jus\-tice.
\end{para}

\begin{para}
    \textsuperscript{4:129}\thinspace{}You will nev\-er be a\-ble to treat your wives with per\-fect fair\-ness, how\-ev\-er much you may want to, but don't fa\-vor one com\-plete\-ly so as to leave an\-oth\-er hang\-ing. If you make a\-mends and are con\-scious of God, He is for\-giv\-ing and mer\-ci\-ful.
\end{para}

\begin{para}
    \textsuperscript{4:135}\thinspace{}Believers, stead\-fast\-ly up\-hold jus\-tice, as witnesses for God, e\-ven if it is a\-gainst your\-selves, your parents, or your relatives. Wheth\-er rich or poor, God can best pro\-tect them both. Don't fol\-low your own de\-sire, lest you swerve from jus\-tice. If you dis\-tort or ne\-glect jus\-tice, God knows what you do.
\end{para}

\begin{para}
    \textsuperscript{4:140}\thinspace{}When you hear peo\-ple de\-ny and rid\-i\-cule God's messages, do not sit with them un\-til they change topics. Oth\-er\-wise you will be\-come like them: God will gath\-er the hypocrites and disbelievers in hell.
\end{para}

\begin{para}
    \textsuperscript{4:142}\thinspace{}When hypocrites stand to pray, they stand la\-zi\-ly, to be seen of men, remembering God on\-ly a lit\-tle,
    \textsuperscript{143}\thinspace{}wavering be\-tween this and that, be\-long\-ing to nei\-ther these nor those.
\end{para}

\begin{para}
    \textsuperscript{4:148}\thinspace{}God does not like evil to be mentioned in pub\-lic, ex\-cept by the one who was wronged. God hears and knows all.
    \textsuperscript{149}\thinspace{}Wheth\-er you do good openly or pri\-vate\-ly, or par\-don an evil, God is for\-giv\-ing and pow\-er\-ful.
\end{para}

\begin{para}
    \textsuperscript{4:163}\thinspace{}Muhammad, we've sent you rev\-e\-la\-tion as We did to No\-ah and the prophets af\-ter him: Abraham, Ishmael, Isaac, Jacob, Jesus, Job, Jonah, Aaron, and Solomon; to David We gave the Psalms;
    \textsuperscript{164}\thinspace{}to oth\-er messengers We've mentioned be\-fore; to others We have\-n't mentioned; and to Mo\-ses God spoke di\-rect\-ly.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{5:2}\thinspace{}Some peo\-ple shut you out of the Sa\-cred Mosque. Don't let your ha\-tred for them lead you to trans\-gress. Help one an\-oth\-er in right\-eous\-ness and pi\-e\-ty, but don't help one an\-oth\-er in sin and ag\-gres\-sion. Fear God, for God is se\-vere in pun\-ish\-ment.
\end{para}

\begin{para}
    \textsuperscript{5:7}\thinspace{}Re\-mem\-ber God's bless\-ing on you. Fear God, for He knows all the secrets of your heart.
    \textsuperscript{8}\thinspace{}Believers, be stead\-fast in de\-vo\-tion to God, as im\-par\-tial witnesses. Don't let ha\-tred of others lead you a\-way from jus\-tice. Be just: that is nearer to pi\-e\-ty.
    \textsuperscript{9}\thinspace{}God has prom\-ised for\-give\-ness and a great re\-ward to those who be\-lieve and do good works.
\end{para}

\begin{para}
    \textsuperscript{5:35}\thinspace{}Believers, fear God, seek ways to come clos\-er to Him, and strive in His cause that you may pros\-per.
\end{para}

\begin{para}
    \textsuperscript{5:38}\thinspace{}Cut off the hands of thieves as rec\-om\-pense for what they have done: a de\-ter\-rent from God.
    \textsuperscript{39}\thinspace{}But who\-ev\-er repents af\-ter his crime, and makes a\-mends, God will ac\-cept his re\-pent\-ance, for God is for\-giv\-ing and mer\-ci\-ful.
\end{para}

\begin{para}
    \textsuperscript{5:44}\thinspace{}We revealed the Torah containing guid\-ance and light.
    \textsuperscript{45}\thinspace{}In it we ordained a life for a life, an eye for an eye, and sim\-i\-lar ret\-ri\-bu\-tion for wounds. But if any\-one forgoes this out of char\-i\-ty, it is an a\-tone\-ment for him\-self. Those who do not judge by what God has revealed are the wrongdoers.
\end{para}

\begin{para}
    \textsuperscript{5:48}\thinspace{}Muhammad, We revealed to you this Book with the truth, confirming the scriptures that came be\-fore it. To each of you, We have pre\-scribed a law and a clear path. Had God wanted it, He would have made you all a sin\-gle com\-mu\-ni\-ty, but He wanted to test you in what He has giv\-en you, so race to do good. You will all re\-turn to God, then He will clar\-i\-fy your differences.
\end{para}

\begin{para}
    \textsuperscript{5:57}\thinspace{}Believers, don't take al\-lies who view your re\-li\-gion with rid\-i\-cule and a\-muse\-ment.
\end{para}

\begin{para}
    \textsuperscript{5:89}\thinspace{}God will not pun\-ish you for your thought\-less oaths, on\-ly for your de\-lib\-er\-ate oaths. The a\-tone\-ment for break\-ing an oath is to feed ten poor peo\-ple as you would feed your own fam\-i\-ly, or to clothe them, or to free a slave. If you can't af\-ford it, fast for three days. That is the a\-tone\-ment for bro\-ken oaths, but keep your oaths.
\end{para}

\begin{para}
    \textsuperscript{5:90}\thinspace{}Believers: intoxicants, gambling, i\-dol\-a\-try, and div\-i\-na\-tion are abominations from Satan. A\-void them so that you may pros\-per.
    \textsuperscript{91}\thinspace{}Satan on\-ly wants to use intoxicants and gambling to cause en\-mi\-ty and ha\-tred a\-mong you, and to hind\-er you from remembering God and pray\-er. So will you quit?
\end{para}

\begin{para}
    \textsuperscript{5:106}\thinspace{}Believers, when mak\-ing bequests as death approaches you, let there be two hon\-est witnesses.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{6:7}\thinspace{}Muhammad, e\-ven if We had sent down to you a mes\-sage writ\-ten on pa\-per, and the disbelievers had touched it with their hands, they would have said, “This is noth\-ing but ob\-vi\-ous mag\-ic!”
    \textsuperscript{8}\thinspace{}They say, “Why was no an\-gel sent down to him?” Had We sent down an an\-gel, their judg\-ment would have come at once, with\-out res\-pite.
    \textsuperscript{9}\thinspace{}E\-ven if We had sent an an\-gel, We would have sent him as a man, adding to their con\-fu\-sion.
\end{para}

\begin{para}
    \textsuperscript{6:48}\thinspace{}We send the messengers on\-ly to give good news and to warn. Who\-ev\-er believes and reforms will have no fear, nor will they grieve.
    \textsuperscript{49}\thinspace{}But suf\-fer\-ing afflicts those who re\-ject our messages, be\-cause of their de\-fi\-ant dis\-o\-be\-di\-ence.
\end{para}

\begin{para}
    \textsuperscript{6:52}\thinspace{}Don't send a\-way those who call on their Lord morn\-ing and eve\-ning, seek\-ing His face. You are not ac\-count\-a\-ble for them, nor they for you. If you turn them a\-way, you be\-come one of the un\-just.
    \textsuperscript{54}\thinspace{}Your Lord has pre\-scribed for Him\-self mer\-cy: if any of you does wrong out of ig\-no\-rance, but then repents and corrects him\-self, He is for\-giv\-ing and mer\-ci\-ful.
\end{para}

\begin{para}
    \textsuperscript{6:70}\thinspace{}Leave a\-lone those who take play and a\-muse\-ment as their re\-li\-gion, and are de\-ceived by the life of this world. Re\-mind them, lest a soul be destroyed for what it has earned, that it will find no pro\-tec\-tor or in\-ter\-ces\-sor ex\-cept God.
\end{para}

\begin{para}
    \textsuperscript{6:108}\thinspace{}Don't re\-vile those whom idolaters call up\-on, lest they re\-vile God in their spite and ig\-no\-rance. To eve\-ry com\-mu\-ni\-ty We have made its own deeds seem pleas\-ing. In the end they will re\-turn to their Lord, and He will in\-form them of all they did.
\end{para}

\begin{para}
    \textsuperscript{6:120}\thinspace{}A\-void sin, both o\-pen and se\-cret, for those who sin will be repaid for what they've done.
    \textsuperscript{121}\thinspace{}Eat noth\-ing o\-ver which God's name has not been pro\-nounced: that is dis\-o\-be\-di\-ence in\-deed.
\end{para}

\begin{para}
    \textsuperscript{6:141}\thinspace{}Don't be ex\-ces\-sive, for God dislikes the waste\-ful.
    \textsuperscript{142}\thinspace{}Eat what God has pro\-vid\-ed for you, wheth\-er it's live\-stock for burdens or for meat.
\end{para}

\begin{para}
    \textsuperscript{6:151}\thinspace{}Don't as\-cribe any equals with God. Be good to your parents. Don't kill your chil\-dren for fear of pov\-er\-ty --- We will pro\-vide for you and for them. Don't ap\-proach shame\-ful deeds, wheth\-er o\-pen or se\-cret. Don't take the life that God has made sa\-cred, ex\-cept by le\-gal right.
    \textsuperscript{152}\thinspace{}Don't ap\-proach the orphan's prop\-er\-ty be\-fore he co\-mes of age, ex\-cept to im\-prove it. Give full meas\-ure and weight with jus\-tice; we bur\-den no one with more than he can bear. When you speak, be just, e\-ven if it concerns a rel\-a\-tive.
    \textsuperscript{153}\thinspace{}This is My straight path, so fol\-low it. Don't fol\-low oth\-er paths. They will lead you a\-way from His. He has commanded you to do this, so that you may be right\-eous.
\end{para}

\begin{para}
    \textsuperscript{6:164}\thinspace{}Eve\-ry soul is re\-spon\-si\-ble for its own actions, and no one bears another's bur\-den. You will all re\-turn to your Lord in the end, and He will clar\-i\-fy that on which you used to dif\-fer.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{7:3}\thinspace{}Believers, fol\-low the rev\-e\-la\-tion to you from your Lord, and fol\-low no masters oth\-er than Him.
\end{para}

\begin{para}
    \textsuperscript{7:204}\thinspace{}When the Qur'an is re\-cit\-ed, pay at\-ten\-tion and lis\-ten si\-lent\-ly that you may re\-ceive mer\-cy.
    \textsuperscript{205}\thinspace{}In the morn\-ings and eve\-nings, re\-mem\-ber your Lord in\-ward\-ly, with hu\-mil\-i\-ty, rev\-er\-ence, and a qui\-et voice. Don't be a\-mong the heed\-less.
    \textsuperscript{206}\thinspace{}E\-ven those who are near your Lord aren't too proud to wor\-ship Him: they glo\-ri\-fy Him, and pros\-trate be\-fore Him.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{8:2}\thinspace{}Believers are those whose hearts trem\-ble with awe when God is mentioned, whose faith strengthens when His messages are re\-cit\-ed to them, who trust in their Lord,
    \textsuperscript{3}\thinspace{}who are stead\-fast in pray\-er, and who give from what We have pro\-vid\-ed them.
    \textsuperscript{4}\thinspace{}They are the true believers. They have for\-give\-ness with their Lord, and gen\-er\-ous sus\-te\-nance.
\end{para}

\begin{para}
    \textsuperscript{8:20}\thinspace{}Believers, o\-bey God and His Mes\-sen\-ger. Don't turn a\-way from him when you are listening.
    \textsuperscript{21}\thinspace{}Don't be like those who say, “We have heard,” but don't ac\-tu\-al\-ly lis\-ten.
    \textsuperscript{22}\thinspace{}In\-deed, the worst creatures in God's sight are the deaf and dumb who don't use rea\-son.
\end{para}

\begin{para}
    \textsuperscript{8:47}\thinspace{}Don't be like those who came out of their homes con\-ceit\-ed, show\-ing off, and hindering others from God's path.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{9:24}\thinspace{}If your fathers, your sons, your brothers, your wives, your tribes, the wealth you have gained, the com\-merce you fear will de\-cline, and the dwellings you love are dearer to you than God, His Mes\-sen\-ger, and strug\-gle in His cause, then wait un\-til God executes His com\-mand. God does not guide the re\-bel\-lious.
\end{para}

\begin{para}
    \textsuperscript{9:34}\thinspace{}Believers, man\-y rabbis and monks false\-ly de\-vour men's wealth and di\-vert men from God's path. Tell those who hoard gold and sil\-ver, with\-out giv\-ing for God's cause, that they will re\-ceive a pain\-ful pun\-ish\-ment.
\end{para}

\begin{para}
    \textsuperscript{9:54}\thinspace{}On\-ly this prevents a gift from be\-ing ac\-cept\-ed: they don't be\-lieve God and His Mes\-sen\-ger, they come to pray\-er la\-zi\-ly, and they give un\-will\-ing\-ly.
\end{para}

\begin{para}
    \textsuperscript{9:60}\thinspace{}Alms are for the poor, the needy, those who ad\-min\-is\-ter the funds, those whose hearts are to be won o\-ver, to free slaves, to help debtors, for God's cause, and needy travelers.
\end{para}

\begin{para}
    \textsuperscript{9:71}\thinspace{}The believers pro\-tect one an\-oth\-er. They en\-join what is right, for\-bid what is wrong, are stead\-fast in pray\-er, spend in char\-i\-ty, and o\-bey God and His Mes\-sen\-ger. God will have mer\-cy on them.
\end{para}

\begin{para}
    \textsuperscript{9:91}\thinspace{}There is no blame on the weak, the ill, or those with\-out resources to spend, when they are sin\-cere to God and His Mes\-sen\-ger. There is no rea\-son to re\-proach those who do good.
    \textsuperscript{92}\thinspace{}There is al\-so no blame on those who came to you for mounts, and when you said, “I can't find a mount for you,” they turned back, their eyes overflowing with tears of grief that they had noth\-ing to spend.
    \textsuperscript{93}\thinspace{}On\-ly those who re\-quest ex\-emp\-tion, de\-spite be\-ing rich, are o\-pen to blame. They are con\-tent to be with those who stay be\-hind.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{10:11}\thinspace{}If God were to has\-ten the ill that men have earned, as they wish to has\-ten the good, their end would al\-read\-y be here.
    \textsuperscript{12}\thinspace{}And when trou\-ble touches man, he cries out to Us wheth\-er ly\-ing on his side, sit\-ting, or stand\-ing, but when We re\-move his trou\-ble, he goes on his way as if he had nev\-er cried out to Us.
\end{para}

\begin{para}
    \textsuperscript{10:41}\thinspace{}If they don't be\-lieve you, say, “My deeds for me, and yours for you. You are not re\-spon\-si\-ble for what I do, nor I for what you do.”
\end{para}

\begin{para}
    \textsuperscript{10:47}\thinspace{}Eve\-ry com\-mu\-ni\-ty is sent a mes\-sen\-ger. When their mes\-sen\-ger co\-mes, they will be judged with jus\-tice, and will not be wronged.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{11:9}\thinspace{}If We let man taste mer\-cy, and then with\-draw it, he despairs and is un\-grate\-ful.
    \textsuperscript{10}\thinspace{}But if We let him taste mer\-cy af\-ter ad\-ver\-si\-ty has touched him, he is sure to say, “Mis\-for\-tune has left me.” He becomes ex\-ult\-ant and boast\-ful.
    \textsuperscript{11}\thinspace{}Those who are stead\-fast and do right\-eous deeds don't act this way. They will have for\-give\-ness and a great re\-ward.
\end{para}

\begin{para}
    \textsuperscript{11:112}\thinspace{}Stay on the right path as you have been commanded, to\-geth\-er with those who have turned to God. Don't trans\-gress, for He sees all that you do.
    \textsuperscript{113}\thinspace{}Don't in\-cline to\-ward nor re\-ly on those who do wrong.
    \textsuperscript{114}\thinspace{}Be stead\-fast in pray\-er at both ends of the day and dur\-ing parts of the night. Good deeds drive a\-way evil deeds.
    \textsuperscript{115}\thinspace{}Be stead\-fast: God won't lose the re\-ward of those who do good.
\end{para}

\begin{para}
    \textsuperscript{11:117}\thinspace{}Your Lord would nev\-er de\-stroy communities un\-just\-ly if its peo\-ple act right\-eous\-ly.
    \textsuperscript{118}\thinspace{}If your Lord had wanted, He could have made man\-kind one com\-mu\-ni\-ty, but they con\-tin\-ue to dif\-fer.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{13:4}\thinspace{}There are neighboring plots in the land: vineyards, cornfields, and date palms. They are all watered with the same wa\-ter, yet We make some bet\-ter to eat than others. These are signs for peo\-ple who rea\-son.
\end{para}

\begin{para}
    \textsuperscript{13:11}\thinspace{}God does not change the con\-di\-tion of a peo\-ple un\-less they change what is with\-in them\-selves.
\end{para}

\begin{para}
    \textsuperscript{13:22}\thinspace{}Those who are pa\-tient, seek\-ing their Lord's face; who are stead\-fast in pray\-er; who spend se\-cret\-ly and openly from what We have pro\-vid\-ed them; and who re\-pel evil with good: those will have the re\-ward of an e\-ter\-nal home.
\end{para}

\begin{para}
    \textsuperscript{13:26}\thinspace{}They re\-joice in the life of this world, but it is a fleet\-ing pleas\-ure com\-pared to the here\-af\-ter.
    \textsuperscript{27}\thinspace{}The disbelievers say, “Why has\-n't his Lord sent down a sign to him?” Say, “God lets stray whom He will, and guides to Him\-self those who turn to Him,
    \textsuperscript{28}\thinspace{}those who be\-lieve, and those whose hearts find rest in remembering God.
    \textsuperscript{29}\thinspace{}Joy and a beau\-ti\-ful home\-com\-ing a\-wait those who be\-lieve and do right\-eous deeds.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{14:3}\thinspace{}Those who pre\-fer this world o\-ver the here\-af\-ter, and turn others from God's way, mak\-ing it seem crook\-ed: they are far a\-stray.
\end{para}

\begin{para}
    \textsuperscript{14:4}\thinspace{}We have on\-ly sent messengers who speak the lan\-guage of their peo\-ple, in or\-der to make things clear for them.
\end{para}

\begin{para}
    \textsuperscript{14:10}\thinspace{}The messengers said to them, “Can there be any doubt a\-bout God, the cre\-a\-tor of the heavens and earth? He invites you in or\-der to for\-give your sins, and let you live un\-til the ap\-point\-ed time.” They said, “You're on\-ly men like us. You want to turn us a\-way from what our fathers wor\-shipped. In that case, bring us clear proof.”
    \textsuperscript{11}\thinspace{}Their messengers answered, “True, we are on\-ly men like you, but God fa\-vors His servants as He pleases. We can't bring you proof un\-less God permits it, so let all believers put their trust in God.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{15:2}\thinspace{}Per\-haps the disbelievers will wish that they had sub\-mit\-ted to God.
    \textsuperscript{3}\thinspace{}Let them eat and en\-joy. Let false hope dis\-tract them. They will come to know.
\end{para}

\begin{para}
    \textsuperscript{15:19}\thinspace{}We have spread out the earth, placed firm mountains on it, and let eve\-ry\-thing grow on it in due bal\-ance.
    \textsuperscript{20}\thinspace{}We have pro\-vid\-ed sus\-te\-nance in it for you and for oth\-er creatures.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{16:5}\thinspace{}He has cre\-at\-ed live\-stock for you. You de\-rive warmth, food, and oth\-er benefits from them.
    \textsuperscript{6}\thinspace{}You find beau\-ty in them when you bring them home in the eve\-ning, and as you lead them to pas\-ture in the morn\-ing.
    \textsuperscript{7}\thinspace{}They car\-ry your loads to lands that you could on\-ly reach on your own with great hard\-ship. Tru\-ly your Lord is kind and mer\-ci\-ful.
    \textsuperscript{8}\thinspace{}He has cre\-at\-ed hors\-es, mules, and donkeys for you to ride, and as a\-dorn\-ment. He al\-so creates things you know noth\-ing a\-bout.
\end{para}

\begin{para}
    \textsuperscript{16:10}\thinspace{}He sends down wa\-ter from the sky. It provides a drink for you and for the veg\-e\-ta\-tion which feeds your live\-stock.
    \textsuperscript{11}\thinspace{}With it He grows for you crops, olives, date palms, grapes, and eve\-ry kind of fruit. These are signs for those who re\-flect.
\end{para}

\begin{para}
    \textsuperscript{16:14}\thinspace{}He has subjected the sea for you, that you may eat fresh meat from it, and ex\-tract ornaments from it to wear. You see the ships plowing through its waves seek\-ing the boun\-ty of God that you may be grate\-ful.
    \textsuperscript{15}\thinspace{}He has placed rivers and roads on the earth, so you may find your way;
    \textsuperscript{16}\thinspace{}and landmarks and stars that men may guide them\-selves.
    \textsuperscript{17}\thinspace{}Is He who creates like one who does\-n't cre\-ate?
\end{para}

\begin{para}
    \textsuperscript{16:36}\thinspace{}We sent a mes\-sen\-ger to eve\-ry com\-mu\-ni\-ty say\-ing, “Serve God and a\-void false gods.” A\-mong them were some whom God guid\-ed and some who strayed. So trav\-el through the earth and ob\-serve how it ended for the deniers.
\end{para}

\begin{para}
    \textsuperscript{16:61}\thinspace{}If God were to pun\-ish men for their wrong\-do\-ing, He would leave no liv\-ing crea\-ture on the earth, but He gives them res\-pite for an ap\-point\-ed time. When their time co\-mes, they are un\-a\-ble to de\-lay it or ad\-vance it by a sin\-gle hour.
\end{para}

\begin{para}
    \textsuperscript{16:90}\thinspace{}God commands jus\-tice, do\-ing good, and gen\-er\-os\-i\-ty to\-wards relatives. He forbids all shame\-ful deeds, bad con\-duct, and op\-pres\-sion. He instructs you so that you may take heed.
    \textsuperscript{91}\thinspace{}Ful\-fill the cov\-e\-nant of God once you have made it, and don't break your oaths.
    \textsuperscript{92}\thinspace{}Don't be like a wom\-an who unravels the yarn which she has firm\-ly spun.
\end{para}

\begin{para}
    \textsuperscript{16:93}\thinspace{}If God wanted, He could have made you all one peo\-ple, but He lets some stray and guides others. You will be ac\-count\-a\-ble for your actions.
    \textsuperscript{94}\thinspace{}Don't use your oaths to de\-ceive one an\-oth\-er, lest a foot slip that was firm\-ly planted. Then you will taste the evil of hindering some\-one from God's path, and will suf\-fer great pun\-ish\-ment.
\end{para}

\begin{para}
    \textsuperscript{16:96}\thinspace{}What you have will van\-ish, but what God has will en\-dure. We will cer\-tain\-ly re\-ward the stead\-fast ac\-cord\-ing to the best of their actions.
    \textsuperscript{97}\thinspace{}We will give a new life, one that is good and pure, to who\-ev\-er works right\-eous\-ness and has faith, wheth\-er male or fe\-male.
\end{para}

\begin{para}
    \textsuperscript{16:99}\thinspace{}Satan has no pow\-er o\-ver those who be\-lieve and trust in their Lord.
    \textsuperscript{100}\thinspace{}He on\-ly has pow\-er o\-ver those who al\-ly with him and who join part\-ners with God.
\end{para}

\begin{para}
    \textsuperscript{16:110}\thinspace{}Your Lord is for\-giv\-ing and mer\-ci\-ful to those who em\-i\-grate be\-cause of per\-se\-cu\-tion, then fight for the faith and re\-main stead\-fast.
    \textsuperscript{111}\thinspace{}On the Day when eve\-ry soul co\-mes to plead for him\-self, eve\-ry soul will be ful\-ly paid for his actions and none will be wronged.
\end{para}

\begin{para}
    \textsuperscript{16:114}\thinspace{}Eat the good and law\-ful food which God has pro\-vid\-ed for you, and be grate\-ful for His fa\-vors.
\end{para}

\begin{para}
    \textsuperscript{16:119}\thinspace{}Your Lord is for\-giv\-ing and mer\-ci\-ful to those who do wrong out of ig\-no\-rance, but af\-ter\-wards re\-pent and make a\-mends.
\end{para}

\begin{para}
    \textsuperscript{16:125}\thinspace{}In\-vite eve\-ry\-one to your Lord's path with wis\-dom and beau\-ti\-ful teach\-ing. Ar\-gue with them in the most po\-lite way, for your Lord knows who has strayed from His path and who is right\-ly guid\-ed.
    \textsuperscript{126}\thinspace{}If you must re\-spond to an at\-tack, re\-spond no worse than the at\-tack, but it is bet\-ter to show pa\-tience.
    \textsuperscript{127}\thinspace{}Be pa\-tient, for your pa\-tience is from God. Don't grieve o\-ver them or be dis\-tressed by their schemes.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{17:15}\thinspace{}Who\-ev\-er receives guid\-ance does so for his own ben\-e\-fit. Who\-ev\-er strays does so to his own det\-ri\-ment. No one will bear another's bur\-den, but We on\-ly pun\-ish af\-ter We have sent a mes\-sen\-ger.
\end{para}

\begin{para}
    \textsuperscript{17:20}\thinspace{}We be\-stow some of our boun\-ty on all. Your Lord's boun\-ty is not closed to any\-one.
    \textsuperscript{21}\thinspace{}No\-tice that We have bestowed more on some than on others, but the here\-af\-ter holds greater rank and boun\-ty.
\end{para}

\begin{para}
    \textsuperscript{17:22}\thinspace{}Don't make an\-oth\-er de\-i\-ty e\-qual with God or you will be\-come dis\-graced and for\-sak\-en.
    \textsuperscript{23}\thinspace{}Your Lord has de\-creed that you wor\-ship on\-ly Him, and that you be kind to your parents. Don't speak a word of dis\-re\-spect to them, nor scold them, but ad\-dress them re\-spect\-ful\-ly.
    \textsuperscript{24}\thinspace{}Say, “My Lord, have mer\-cy on them just as they reared me when I was small.”
\end{para}

\begin{para}
    \textsuperscript{17:25}\thinspace{}Your Lord knows what is in your hearts. If you are right\-eous, He forgives those who turn to Him a\-gain and a\-gain.
    \textsuperscript{26}\thinspace{}Give what is due to relatives, the needy, and travelers; but don't spend your wealth wastefully.
    \textsuperscript{27}\thinspace{}In\-deed, the waste\-ful are brothers of Satan, and Satan is al\-ways un\-grate\-ful to his Lord.
    \textsuperscript{28}\thinspace{}E\-ven if you must turn a\-way from the needy in pur\-suit of mer\-cy from your Lord, then at least speak a kind word to them.
\end{para}

\begin{para}
    \textsuperscript{17:31}\thinspace{}Don't kill your chil\-dren for fear of pov\-er\-ty. We will pro\-vide for them and for you. In\-deed, kill\-ing them is a great sin.
    \textsuperscript{32}\thinspace{}Don't go near a\-dul\-ter\-y, for it is shame\-ful and an evil path.
    \textsuperscript{33}\thinspace{}Don't take life, which God has made sa\-cred, ex\-cept in pur\-suit of jus\-tice.
    \textsuperscript{34}\thinspace{}Don't go near an orphan's prop\-er\-ty, ex\-cept to im\-prove it, un\-til he reaches ma\-tur\-i\-ty. Ful\-fill eve\-ry com\-mit\-ment, for you will be questioned a\-bout your commitments.
    \textsuperscript{35}\thinspace{}Give full meas\-ure when you meas\-ure, and weigh with hon\-est scales. That is best and ad\-van\-ta\-geous in the end.
    \textsuperscript{36}\thinspace{}Don't con\-cern your\-self with things you know noth\-ing a\-bout: you will be questioned a\-bout hear\-ing, see\-ing, and feel\-ing.
    \textsuperscript{37}\thinspace{}Don't walk on the earth with ar\-ro\-gance: you can't tear the earth a\-part, nor match the mountains in height.
    \textsuperscript{38}\thinspace{}Your Lord detests all these actions.
    \textsuperscript{39}\thinspace{}This is some of the wis\-dom your Lord has revealed to you.
\end{para}

\begin{para}
    \textsuperscript{17:53}\thinspace{}Say what is best. Satan sows dis\-cord a\-mong My servants. He is a clear en\-e\-my to man.
\end{para}

\begin{para}
    \textsuperscript{17:70}\thinspace{}We have honored Adam's chil\-dren, car\-ried them on land and sea, giv\-en them good sus\-te\-nance, and fa\-vored them a\-bove much of our cre\-a\-tion.
    \textsuperscript{71}\thinspace{}One day We will call to\-geth\-er all men a\-long with their leaders. Who\-ev\-er is giv\-en his rec\-ord in his right hand will read it with pleas\-ure, and will not be wronged in the least.
    \textsuperscript{72}\thinspace{}But who\-ev\-er is blind in this world will be blind in the here\-af\-ter and e\-ven fur\-ther a\-stray.
\end{para}

\begin{para}
    \textsuperscript{17:78}\thinspace{}Es\-tab\-lish reg\-u\-lar pray\-er from mid\-day un\-til the dark\-ness of night. Re\-cite the Qur'an at dawn, for morn\-ing rec\-i\-ta\-tion is al\-ways witnessed.
\end{para}

\begin{para}
    \textsuperscript{17:100}\thinspace{}If you pos\-sessed the treas\-ur\-ies of my Lord's mer\-cy, you would with\-hold them for fear of spend\-ing too much: man is ev\-er sting\-y.
\end{para}

\begin{para}
    \textsuperscript{17:110}\thinspace{}Call on God, or on the Most Mer\-ci\-ful. Which\-ev\-er name you call is the same, for the best names be\-long to Him. Don't pray too loud\-ly or too qui\-et\-ly, but seek a mid\-dle way.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{18:23}\thinspace{}Nev\-er say, “I'll do that to\-mor\-row,”
    \textsuperscript{24}\thinspace{}with\-out adding, “God will\-ing.” When you for\-get, re\-mem\-ber your Lord and say, “May my Lord guide me clos\-er to the right path.”
\end{para}

\begin{para}
    \textsuperscript{18:32}\thinspace{}There were two men. We gave one of them two vineyards surrounded by date palms with cornfields in be\-tween.
    \textsuperscript{33}\thinspace{}Both vineyards pro\-duced fruit, did\-n't fail, and We caused a riv\-er to flow through them,
    \textsuperscript{34}\thinspace{}so this man had a\-bun\-dant fruit. He said to his friend, “I have more wealth and men than you.”
    \textsuperscript{35}\thinspace{}Hav\-ing thus sinned a\-gainst him\-self, he went in\-to his gar\-den. He said, “I don't think this will ev\-er per\-ish,
    \textsuperscript{36}\thinspace{}nor that the Hour of Judg\-ment will ev\-er come. E\-ven if I were brought back to my Lord, I would sure\-ly find some\-thing e\-ven bet\-ter than this.”
\end{para}

\begin{para}
    \textsuperscript{18:37}\thinspace{}His friend said to him, “Do you dis\-be\-lieve in Him who cre\-at\-ed you?
    \textsuperscript{38}\thinspace{}But as for me, He is God, My Lord, and I as\-so\-ci\-ate no part\-ner with Him.
    \textsuperscript{39}\thinspace{}When you entered your gar\-den, why did\-n't you say, ‘This is God's will. There is no pow\-er but with God.’ Al\-though I have less wealth and off\-spring than you,
    \textsuperscript{40}\thinspace{}my Lord may give me some\-thing bet\-ter than your gar\-den, or a ca\-lam\-i\-ty may turn your gar\-den in\-to slip\-per\-y sand,
    \textsuperscript{41}\thinspace{}or its wa\-ter might sink un\-der\-ground so that you can't reach it.”
\end{para}

\begin{para}
    \textsuperscript{18:42}\thinspace{}And thus it happened: his fruit was ruined. He remained, wring\-ing his hands o\-ver what he had spent on it, while it col\-lapsed up\-on its trellises, say\-ing, “I wish that I had nev\-er as\-cribed part\-ners to my Lord.”
    \textsuperscript{43}\thinspace{}He had none to help him oth\-er than God, nor could he help him\-self.
\end{para}

\begin{para}
    \textsuperscript{18:45}\thinspace{}This world's life is like rain sent down from the sky. The earth's veg\-e\-ta\-tion absorbs it but soon becomes dry stub\-ble, scattered by the wind.
    \textsuperscript{46}\thinspace{}Wealth and chil\-dren a\-dorn this world's life, but en\-dur\-ing good deeds are bet\-ter to your Lord, and a bet\-ter source of hope.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{20:2}\thinspace{}We did\-n't send down the Qur'an to dis\-tress you,
    \textsuperscript{3}\thinspace{}but on\-ly as a reminder for those who fear God.
\end{para}

\begin{para}
    \textsuperscript{20:15}\thinspace{}Al\-though I keep the hour hid\-den, the time is com\-ing for each soul to be rewarded for its ef\-fort.
    \textsuperscript{16}\thinspace{}Don't let those who de\-ny it and fol\-low their own desires dis\-tract you from it, lest you per\-ish.
\end{para}

\begin{para}
    \textsuperscript{20:81}\thinspace{}Eat from the good things We have pro\-vid\-ed you, but don't trans\-gress, lest My an\-ger de\-scend on you.
    \textsuperscript{82}\thinspace{}Yet I re\-peat\-ed\-ly for\-give those who re\-pent, be\-lieve, do right, and con\-tin\-ue re\-ceiv\-ing guid\-ance.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{22:39}\thinspace{}Those who have been wrong\-ful\-ly attacked may fight back. In\-deed, God has the pow\-er to help
    \textsuperscript{40}\thinspace{}those who have been un\-just\-ly driv\-en from their homes on\-ly for say\-ing, “Our Lord is God”. If God did\-n't re\-pel one peo\-ple with an\-oth\-er, man\-y monasteries, churches, synagogues, and mosques would have been destroyed; places in which the name of God is fre\-quent\-ly in\-voked.
\end{para}

\begin{para}
    \textsuperscript{22:67}\thinspace{}We have ap\-point\-ed a dif\-fer\-ent man\-ner of wor\-ship for each com\-mu\-ni\-ty to ob\-serve, so don't let others ar\-gue with you a\-bout the mat\-ter.
    \textsuperscript{68}\thinspace{}If they do ar\-gue with you, say, “God knows best what you are do\-ing.”
    \textsuperscript{69}\thinspace{}On Res\-ur\-rec\-tion Day, God will judge the differences be\-tween you.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{23:1}\thinspace{}The believers are in\-deed suc\-cess\-ful:
    \textsuperscript{2}\thinspace{}those who pray hum\-bly,
    \textsuperscript{3}\thinspace{}a\-void i\-dle talk,
    \textsuperscript{4}\thinspace{}give alms,
    \textsuperscript{5}\thinspace{}guard their chas\-ti\-ty
    \textsuperscript{6}\thinspace{}(ex\-cept with their spouses),
    \textsuperscript{8}\thinspace{}who are at\-ten\-tive to their trusts and promises,
    \textsuperscript{9}\thinspace{}and guard their prayers.
    \textsuperscript{11}\thinspace{}These will in\-her\-it Par\-a\-dise and live there for\-ev\-er.
\end{para}

\begin{para}
    \textsuperscript{23:44}\thinspace{}We sent our messengers one af\-ter an\-oth\-er. Eve\-ry time a community's mes\-sen\-ger ar\-rived, they de\-nied him. So We let them fol\-low each oth\-er to de\-struc\-tion and they be\-came cau\-tion\-ar\-y ta\-les.
\end{para}

\begin{para}
    \textsuperscript{23:68}\thinspace{}Have they nev\-er con\-tem\-plat\-ed the word of God? Has some\-thing new come to them that did\-n't come to their forefathers?
    \textsuperscript{69}\thinspace{}Do they de\-ny their Mes\-sen\-ger be\-cause they don't rec\-og\-nize him?
    \textsuperscript{70}\thinspace{}Do they say, “He is pos\-sessed”? No, he brought them the truth and most of them hate it.
    \textsuperscript{71}\thinspace{}If truth had followed their desires, the heavens, the earth, and eve\-ry\-thing that lives in them would be ruined. We sent them a reminder, but they turn a\-way from it.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{24:27}\thinspace{}Believers, be\-fore entering an\-oth\-er person's house, ask per\-mis\-sion and greet its occupants.
    \textsuperscript{28}\thinspace{}E\-ven if no one is in the house, wait un\-til you re\-ceive per\-mis\-sion. If you are told, “Go a\-way,” do so. God is a\-ware of all that you do.
\end{para}

\begin{para}
    \textsuperscript{24:30}\thinspace{}Tell believing men to low\-er their gaze and guard their pri\-vate parts.
    \textsuperscript{31}\thinspace{}Tell believing wom\-en to low\-er their gaze and guard their pri\-vate parts, to cov\-er their breasts, and not to stamp their feet to draw at\-ten\-tion to what is hid\-den.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{25:7}\thinspace{}They say, “What sort of a mes\-sen\-ger is this? He eats food and walks through the mar\-ket\-place. Why was\-n't an an\-gel sent down to help him warn?
    \textsuperscript{8}\thinspace{}Why has\-n't he been giv\-en treas\-ure, or a gar\-den to feed him? The man you fol\-low is simp\-ly bewitched.”
    \textsuperscript{9}\thinspace{}They've gone a\-stray and can't find the right path.
\end{para}

\begin{para}
    \textsuperscript{25:20}\thinspace{}Muhammad, all the messengers We sent be\-fore you were al\-so men who ate food and walked through the mar\-ket\-place. We've made some peo\-ple a test for others. Will you have pa\-tience? Your Lord sees eve\-ry\-thing.
\end{para}

\begin{para}
    \textsuperscript{25:27}\thinspace{}On that Day the wrong\-do\-er will bite his hand and say, “If on\-ly I had tak\-en the same path as the Mes\-sen\-ger!
    \textsuperscript{28}\thinspace{}Woe is me! If on\-ly I had not tak\-en so-and-so as a friend!
    \textsuperscript{29}\thinspace{}He lead me a\-way from the Reminder af\-ter it came to me. Satan al\-ways betrays man\-kind.”
\end{para}

\begin{para}
    \textsuperscript{25:32}\thinspace{}Disbelievers say, “Why is\-n't the Qur'an revealed to him all at once?” This is to strength\-en your heart. We have revealed it grad\-u\-al\-ly, in stages.
\end{para}

\begin{para}
    \textsuperscript{25:47}\thinspace{}He makes the night as cloth\-ing for you, sleep as rest, and the day like a res\-ur\-rec\-tion.
    \textsuperscript{48}\thinspace{}He sends the winds as messengers of good news be\-fore His mer\-cy. We send down pure wa\-ter from the sky
    \textsuperscript{49}\thinspace{}to re\-vive a dead land and as a drink for our creations: animals and men in great numbers.
    \textsuperscript{50}\thinspace{}We have re\-peat\-ed this man\-y times a\-mong them as a reminder, but most re\-main un\-grate\-ful.
\end{para}

\begin{para}
    \textsuperscript{25:56}\thinspace{}Muhammad, We have on\-ly sent you to de\-liv\-er good news and a warn\-ing.
    \textsuperscript{57}\thinspace{}Say, “The on\-ly pay\-ment I ask for it is that who\-ev\-er wishes takes a path to his Lord.”
\end{para}

\begin{para}
    \textsuperscript{25:63}\thinspace{}The servants of the Most Gra\-cious are those who walk hum\-bly on the earth, and when fools ad\-dress them, they re\-ply, “Peace;”
    \textsuperscript{67}\thinspace{}those who, when they spend, are nei\-ther ex\-trav\-a\-gant nor sting\-y, but mod\-er\-ate be\-tween those extremes;
    \textsuperscript{68}\thinspace{}those who nev\-er in\-voke a de\-i\-ty be\-side God, nor take a life which God has made sa\-cred, nor com\-mit for\-ni\-ca\-tion.
\end{para}

\begin{para}
    \textsuperscript{25:70}\thinspace{}God will change evil in\-to good for him who repents, believes, and does good deeds. God is for\-giv\-ing and mer\-ci\-ful.
    \textsuperscript{71}\thinspace{}Who\-ev\-er repents and does good has turned to\-wards God with true re\-pent\-ance.
\end{para}

\begin{para}
    \textsuperscript{25:72}\thinspace{}Servants of God are those who give no false wit\-ness and pass by fri\-vol\-i\-ty with dig\-ni\-ty;
    \textsuperscript{73}\thinspace{}those who don't ig\-nore their Lord's messages when reminded of them;
    \textsuperscript{74}\thinspace{}and those who pray, “Our Lord, give us joy in our wives and chil\-dren, and let us be examples to the right\-eous.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{28:50}\thinspace{}Who is fur\-ther a\-stray than one who follows his own desires with\-out guid\-ance from God? In\-deed, God does\-n't guide wrongdoers.
    \textsuperscript{51}\thinspace{}We have conveyed the Word to them, so they may re\-mem\-ber.
\end{para}

\begin{para}
    \textsuperscript{28:54}\thinspace{}Those who be\-lieve the Scriptures will be rewarded dou\-ble, be\-cause they are stead\-fast, re\-pel evil with good, give out of what We have giv\-en them,
    \textsuperscript{55}\thinspace{}and turn a\-way from friv\-o\-lous speech say\-ing, “We have our actions and you have yours. Peace be with you. We don't seek as\-so\-ci\-a\-tion with the ig\-no\-rant.”
    \textsuperscript{56}\thinspace{}You can't guide eve\-ry\-one you love.
\end{para}

\begin{para}
    \textsuperscript{28:76}\thinspace{}Qarun had great rich\-es. His peo\-ple said to him, “Don't gloat, for God does\-n't like gloaters.
    \textsuperscript{77}\thinspace{}In\-stead, use your wealth to seek the home of the here\-af\-ter with\-out for\-get\-ting your share in this world. Do good, as God has been good to you. Don't spread cor\-rup\-tion in the land, for God does\-n't like those who spread cor\-rup\-tion.”
    \textsuperscript{79}\thinspace{}He went out a\-mong the peo\-ple in his pomp. Those who de\-sired a world\-ly life said, “If on\-ly we had what Qarun has. He is tru\-ly for\-tu\-nate.”
    \textsuperscript{80}\thinspace{}But those with knowl\-edge said, “Woe to you. God's re\-ward is bet\-ter for those who be\-lieve and do good. On\-ly the stead\-fast will at\-tain it.”
\end{para}

\begin{para}
    \textsuperscript{28:84}\thinspace{}Who\-ev\-er does good will re\-ceive some\-thing bet\-ter, but who\-ev\-er does evil will on\-ly be punished for what he has done.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{29:7}\thinspace{}For those who be\-lieve and do good deeds, We will blot out all their evil and re\-ward them ac\-cord\-ing to their best deeds.
\end{para}

\begin{para}
    \textsuperscript{29:8}\thinspace{}We have commanded man to treat his parents with kind\-ness, but if they try to make you wor\-ship gods be\-side Me, dis\-o\-bey them. You will all re\-turn to me, and I'll tell you the truth a\-bout eve\-ry\-thing you've done.
\end{para}

\begin{para}
    \textsuperscript{29:10}\thinspace{}Some men say, “We be\-lieve in God,” but when they suf\-fer for His cause, they con\-sid\-er per\-se\-cu\-tion from men as if it were God's pun\-ish\-ment. If help co\-mes to you from your Lord, they say, “We've al\-ways been with you.” God knows best what is in everyone's heart.
\end{para}

\begin{para}
    \textsuperscript{29:45}\thinspace{}Pray\-er restrains im\-mor\-al and un\-just be\-hav\-ior.
    \textsuperscript{46}\thinspace{}Don't ar\-gue with followers of ear\-li\-er rev\-e\-la\-tion, but say, “We be\-lieve in what was revealed to us and in what was revealed to you. Our God and your God are the same. We all bow to Him.”
\end{para}

\begin{para}
    \textsuperscript{29:57}\thinspace{}Eve\-ry\-one will taste death, then re\-turn to Us.
    \textsuperscript{58}\thinspace{}To them who be\-lieve and do good, We will give mansions to live in for\-ev\-er in a par\-a\-dise with flow\-ing rivers. The re\-ward is ex\-cel\-lent for those who la\-bor,
    \textsuperscript{59}\thinspace{}are stead\-fast, and trust in their Lord.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{30:21}\thinspace{}A\-mong His signs is this: He cre\-at\-ed spouses for you that you may live in tran\-quil\-i\-ty with them. He put love and mer\-cy be\-tween you. In\-deed, there are signs in this for those who re\-flect.
    \textsuperscript{22}\thinspace{}A\-mong His signs is the cre\-a\-tion of the heavens and the earth, and the di\-ver\-si\-ty of your languages and your col\-ors. In\-deed, in that are signs for those who know.
    \textsuperscript{23}\thinspace{}A\-mong His signs is your sleep, by night and day, and your quest for some of His boun\-ty.
\end{para}

\begin{para}
    \textsuperscript{30:33}\thinspace{}When men en\-coun\-ter trou\-ble, they cry to their Lord and turn to Him for help, but as soon as He gives them a taste of mer\-cy, some of them wor\-ship gods be\-side Him,
    \textsuperscript{34}\thinspace{}show\-ing in\-grat\-i\-tude for what We gave them.
\end{para}

\begin{para}
    \textsuperscript{30:36}\thinspace{}When We give men a taste of mer\-cy, they re\-joice, but when evil afflicts them be\-cause of their own actions, they des\-pair.
\end{para}

\begin{para}
    \textsuperscript{30:38}\thinspace{}Give what is due to relatives, the needy, and the trav\-el\-er. That is best for those who seek God's coun\-te\-nance, and they will pros\-per.
    \textsuperscript{39}\thinspace{}What\-ev\-er you lend in u\-su\-ry, to in\-crease through the prop\-er\-ty of others, has no in\-crease with God, but what\-ev\-er you give in char\-i\-ty, seek\-ing God's coun\-te\-nance, will in\-crease re\-peat\-ed\-ly.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{31:4}\thinspace{}Those who are stead\-fast in pray\-er, spend in char\-i\-ty, and are cer\-tain of the here\-af\-ter,
    \textsuperscript{5}\thinspace{}are guid\-ed by their Lord and will pros\-per.
\end{para}

\begin{para}
    \textsuperscript{31:14}\thinspace{}We have commanded man to be good to his parents. His moth\-er car\-ried him in weak\-ness, and it took two years to wean him. Be grate\-ful to Me and your parents.
    \textsuperscript{15}\thinspace{}But if they try to make you as\-so\-ci\-ate others with Me, dis\-o\-bey them. Yet ac\-com\-pa\-ny them in this life with kind\-ness, and fol\-low the path of those who turn to me. You will all re\-turn to Me in the end, and I will tell you the truth a\-bout all you did.
\end{para}

\begin{para}
    \textsuperscript{31:17}\thinspace{}Be stead\-fast in pray\-er, en\-join what is right, for\-bid what is wrong, and pa\-tient\-ly bear what\-ev\-er happens to you: as\-pire to these things.
    \textsuperscript{18}\thinspace{}Don't turn a\-way from men in pride, nor walk ar\-ro\-gant\-ly through the earth, for God dislikes the ar\-ro\-gant and boast\-ful.
    \textsuperscript{19}\thinspace{}Be mod\-er\-ate in your pa\-ce and low\-er your voice, for the ugliest sound is the braying of as\-ses.
\end{para}

\begin{para}
    \textsuperscript{31:32}\thinspace{}When waves cov\-er them like a can\-o\-py, they call to God, of\-fer\-ing Him sin\-cere de\-vo\-tion. But once He has safe\-ly delivered them to land, some of them wav\-er.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{33:35}\thinspace{}God has pre\-pared for\-give\-ness and a great re\-ward for those who are de\-vot\-ed to God, for the believers, the de\-vout, the hon\-est, the stead\-fast, the hum\-ble, the char\-i\-ta\-ble, those who fast, the chaste, and those who fre\-quent\-ly re\-mem\-ber God.
\end{para}

\begin{para}
    \textsuperscript{33:54}\thinspace{}God has full knowl\-edge of eve\-ry\-thing, wheth\-er you re\-veal or con\-ceal it.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{34:34}\thinspace{}When\-ev\-er we send a warn\-er to a com\-mu\-ni\-ty, the wealth\-y say, “We don't be\-lieve in the mes\-sage you bring.
    \textsuperscript{35}\thinspace{}We have more wealth and chil\-dren than you, and will not be punished.”
\end{para}

\begin{para}
    \textsuperscript{34:37}\thinspace{}Nei\-ther wealth nor chil\-dren will bring you nearer to Us: on\-ly faith and right\-eous\-ness.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{35:45}\thinspace{}If God im\-me\-di\-ate\-ly punished men for their wrong\-do\-ing, He would\-n't leave a sin\-gle crea\-ture on the earth. How\-ev\-er, He defers for a stat\-ed time.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{39:7}\thinspace{}If you are un\-grate\-ful, God has no need of you. He dislikes in\-grat\-i\-tude in His servants. If you are grate\-ful, He is pleased with you. No one will bear another's bur\-den. You will re\-turn to your Lord in the end, and He will in\-form you a\-bout what you did. He knows what is in men's hearts.
\end{para}

\begin{para}
    \textsuperscript{39:10}\thinspace{}Those who do good in this world re\-ceive a good re\-ward. The pa\-tient will re\-ceive a re\-ward with\-out meas\-ure.
\end{para}

\begin{para}
    \textsuperscript{39:49}\thinspace{}When man suffers, he cries out to Us, but when We fa\-vor him, he says, “I've re\-ceived this be\-cause of my own wis\-dom.” This is on\-ly a test, but most don't know it.
\end{para}

\begin{para}
    \textsuperscript{39:69}\thinspace{}The earth will shine with the light of its Lord, the rec\-ord of deeds will be opened, the prophets and witnesses will be brought for\-ward, and fair judg\-ment pro\-nounced be\-tween them. They will not be wronged.
    \textsuperscript{70}\thinspace{}Eve\-ry man will be paid in full for his actions.
    \textsuperscript{71}\thinspace{}Disbelievers will be led to Hell in crowds. When they ar\-rive, its gates will o\-pen, and its keepers will say, “Did\-n't messengers come from a\-mong you, re\-cit\-ing your Lord's messages to you, and warn\-ing you a\-bout today's meet\-ing?”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{40:55}\thinspace{}Be pa\-tient, for God's prom\-ise is true. Ask for\-give\-ness for your sins, and praise your Lord eve\-ning and morn\-ing.
    \textsuperscript{56}\thinspace{}Those who dis\-pute God's messages with\-out au\-thor\-i\-ty have hearts full of pride, which they will nev\-er sat\-is\-fy. Seek ref\-uge in God.
\end{para}

\begin{para}
    \textsuperscript{40:61}\thinspace{}God made the night for you, so that you may rest. He made the day to help you see. God is boun\-ti\-ful to\-wards men, but most men give no thanks.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{41:33}\thinspace{}Who speaks bet\-ter than one who calls men to God, does what is right, and says, “I sub\-mit to God”?
    \textsuperscript{34}\thinspace{}Re\-pel evil with some\-thing bet\-ter, then your en\-e\-my will be\-come like a close friend.
    \textsuperscript{36}\thinspace{}If whispers from Satan stir you up, seek ref\-uge in God.
\end{para}

\begin{para}
    \textsuperscript{41:49}\thinspace{}Man nev\-er tires of asking for good things, but if evil touches him, he gives up all hope and is lost in des\-pair.
    \textsuperscript{50}\thinspace{}When we let him taste mer\-cy af\-ter ad\-ver\-si\-ty, he is sure to say, “This is due to my mer\-it. I don't think that the Hour will come, but if I am brought back to my Lord, I will re\-ceive the best re\-ward.” We will in\-form the disbelievers of eve\-ry\-thing they did, and make them taste a se\-vere tor\-ment.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{42:13}\thinspace{}He has established the same re\-li\-gion for you that He gave to No\-ah, Abraham, Mo\-ses, and Jesus: up\-hold the faith, and make no factions there\-in.
    \textsuperscript{14}\thinspace{}They be\-came di\-vid\-ed through jeal\-ous an\-i\-mos\-i\-ty a\-mong them\-selves.
    \textsuperscript{15}\thinspace{}There\-fore say, “God is our Lord and your Lord. Our deeds for us, and your deeds for you. Let there be no arguments be\-tween us. God will bring us to\-geth\-er, and we will re\-turn to Him.”
\end{para}

\begin{para}
    \textsuperscript{42:27}\thinspace{}If God were to grant a\-bun\-dant pro\-vi\-sion to His servants, they would mis\-be\-have; but he sends down in due meas\-ure what He pleases. He knows His servants and watches o\-ver them.
\end{para}

\begin{para}
    \textsuperscript{42:36}\thinspace{}What you've been giv\-en is for tem\-po\-rar\-y en\-joy\-ment in this life. What is with God is bet\-ter and more last\-ing for those who be\-lieve and trust in their Lord;
    \textsuperscript{37}\thinspace{}a\-void great sins and immoralities; for\-give, e\-ven when an\-gry;
    \textsuperscript{38}\thinspace{}re\-spond to their Lord; are stead\-fast in pray\-er; con\-duct their af\-fairs by mu\-tu\-al con\-sul\-ta\-tion; give from what We have pro\-vid\-ed them;
    \textsuperscript{39}\thinspace{}and de\-fend them\-selves when tyr\-an\-ny afflicts them.
\end{para}

\begin{para}
    \textsuperscript{42:40}\thinspace{}The rec\-om\-pense for harm is an e\-qual harm, but God rewards him who forgives and reconciles.
    \textsuperscript{41}\thinspace{}There is no rea\-son to blame some\-one who defends him\-self af\-ter be\-ing wronged.
    \textsuperscript{43}\thinspace{}It requires great de\-ter\-mi\-na\-tion to be pa\-tient and for\-giv\-ing.
\end{para}

\begin{para}
    \textsuperscript{42:48}\thinspace{}Man rejoices when We let him taste Our mer\-cy, but he is un\-grate\-ful when his own deeds bring mis\-for\-tune.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{43:6}\thinspace{}We sent man\-y prophets a\-mong ear\-li\-er peo\-ple,
    \textsuperscript{7}\thinspace{}but they mocked each one.
\end{para}

\begin{para}
    \textsuperscript{43:22}\thinspace{}They said, “Our fathers followed a cer\-tain re\-li\-gion, and we guide our\-selves by their footsteps.”
    \textsuperscript{23}\thinspace{}In the same way, when\-ev\-er We sent a mes\-sen\-ger to warn a com\-mu\-ni\-ty, the wealth\-y ones said, “Our fathers followed a cer\-tain re\-li\-gion, and we will cer\-tain\-ly fol\-low them.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{45:14}\thinspace{}Tell the believers to for\-give those who don't ex\-pect the Days of God. He a\-lone will re\-pay peo\-ple for what they have earned.
    \textsuperscript{15}\thinspace{}Who\-ev\-er does good benefits him\-self. Who\-ev\-er does evil harms him\-self. You will all re\-turn to your Lord.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{47:2}\thinspace{}For those who have faith, do good, and be\-lieve in the rev\-e\-la\-tion giv\-en to Muhammad, God will re\-move their misdeeds and im\-prove their con\-di\-tion.
\end{para}

\begin{para}
    \textsuperscript{47:36}\thinspace{}If you be\-lieve and fear God, He will give you your rewards. He does\-n't ask you to give up all your possessions.
    \textsuperscript{37}\thinspace{}If He pressed you for them, you would be\-grudge Him, and it would bring out your ill feelings.
\end{para}

\begin{para}
    \textsuperscript{47:38}\thinspace{}You're in\-vit\-ed to give to\-wards God's cause, but some a\-mong you are grudg\-ing. Who\-ev\-er withholds, on\-ly withholds from him\-self. God is free from all wants. You are the needy ones. If you turn a\-way, He'll re\-place you with some\-one else.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{49:6}\thinspace{}Believers, if a trou\-ble\-mak\-er brings you news, ver\-i\-fy it, lest you un\-wit\-ting\-ly harm some\-one, and lat\-er re\-gret what you've done.
\end{para}

\begin{para}
    \textsuperscript{49:11}\thinspace{}Believers, one group should not rid\-i\-cule an\-oth\-er, who may af\-ter all be bet\-ter than them. Don't in\-sult one an\-oth\-er. Don't use of\-fen\-sive nicknames for one an\-oth\-er.
    \textsuperscript{12}\thinspace{}A\-void mak\-ing too man\-y assumptions, for some assumptions are sins. Don't spy on or back\-bite one an\-oth\-er.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{53:29}\thinspace{}A\-void those who turn a\-way from Our mes\-sage and on\-ly de\-sire a world\-ly life.
    \textsuperscript{30}\thinspace{}That is the ex\-tent of their knowl\-edge.
    \textsuperscript{31}\thinspace{}Eve\-ry\-thing in the heavens and on earth belongs to God. He repays those who do evil, ac\-cord\-ing to their deeds, and those who do good, with what is best.
    \textsuperscript{32}\thinspace{}Your Lord is am\-ple in for\-give\-ness to those who a\-void great sins and shame\-ful deeds, though they may com\-mit small sins.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{57:7}\thinspace{}Be\-lieve in God and His mes\-sen\-ger, and give out of what you in\-her\-it from Him. Those who be\-lieve and give will re\-ceive a great re\-ward. 
    \textsuperscript{20}\thinspace{}This life is like rain: the growth it causes delights the tillers, but soon it with\-ers. You see it yel\-low, then dry and crum\-ble a\-way. World\-ly life is on\-ly a pleas\-ant il\-lu\-sion.
    \textsuperscript{21}\thinspace{}Race for your Lord's for\-give\-ness and for a gar\-den as wide as heav\-en and earth, pre\-pared for those who be\-lieve in God and His messengers.
\end{para}

\begin{para}
    \textsuperscript{57:23}\thinspace{}You don't need to des\-pair o\-ver what eludes you or gloat o\-ver what you re\-ceive. For God dislikes those who are con\-ceit\-ed, boast\-ful,
    \textsuperscript{24}\thinspace{}sting\-y, and en\-join stin\-gi\-ness on others.
\end{para}

\begin{para}
    \textsuperscript{57:26}\thinspace{}We sent No\-ah and Abraham, and gave prophethood and scrip\-ture to their descendants. Some of them were right\-ly guid\-ed, but man\-y were transgressors.
    \textsuperscript{27}\thinspace{}We sent oth\-er messengers in their footsteps. We sent Jesus the son of Mar\-y, gave him the Gos\-pel, and put com\-pas\-sion and mer\-cy in\-to the hearts of his followers. But We did not pre\-scribe mo\-nas\-ti\-cism for them; they invented that them\-selves. We gave the believers a\-mong them their re\-ward, but man\-y of them were transgressors.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{61:2}\thinspace{}Believers, why do you say one thing and do an\-oth\-er?
    \textsuperscript{3}\thinspace{}It is most hate\-ful in God's sight.
\end{para}

\begin{para}
    \textsuperscript{61:7}\thinspace{}Who does greater wrong than one who invents a lie a\-gainst God while he is be\-ing in\-vit\-ed to Islam?
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{62:9}\thinspace{}Believers, when the call to pray\-er is proclaimed on Friday, hur\-ry to God's rememberance, and leave your com\-merce. That is best for you if on\-ly you knew.
    \textsuperscript{10}\thinspace{}When the pray\-er is fin\-ished, dis\-perse through\-out the land, seek God's boun\-ty, and re\-mem\-ber Him of\-ten, so that you may pros\-per.
\end{para}

\begin{para}
    \textsuperscript{62:11}\thinspace{}Muhammad, when they see com\-merce or a\-muse\-ment, they rush to\-wards it, and leave you stand\-ing. Say, “What God has is bet\-ter than any a\-muse\-ment or com\-merce. God is the best pro\-vid\-er.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{68:7}\thinspace{}Your Lord knows best who strays from His path and who receives guid\-ance,
    \textsuperscript{8}\thinspace{}so don't o\-bey the deniers.
    \textsuperscript{10}\thinspace{}Don't o\-bey a worth\-less oath-maker,
    \textsuperscript{11}\thinspace{}a slanderer, a ma\-li\-cious gos\-sip,
    \textsuperscript{12}\thinspace{}one who hinders good, a transgressor, sin\-ful,
    \textsuperscript{13}\thinspace{}cru\-el, a pre\-tend\-er.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{70:19}\thinspace{}Man was cre\-at\-ed anx\-ious.
    \textsuperscript{20}\thinspace{}He's fret\-ful when mis\-for\-tune touches him,
    \textsuperscript{21}\thinspace{}and sting\-y when good for\-tune touches him.
    \textsuperscript{22}\thinspace{}Ex\-cept for those de\-vot\-ed to pray\-er
    \textsuperscript{23}\thinspace{}who re\-main con\-stant in their prayers,
    \textsuperscript{24}\thinspace{}and who give a share of their wealth
    \textsuperscript{25}\thinspace{}to beggars and the de\-prived;
    \textsuperscript{26}\thinspace{}those who be\-lieve in the Day of Judg\-ment;
    \textsuperscript{27}\thinspace{}those who fear their Lord's pun\-ish\-ment,
    \textsuperscript{28}\thinspace{}for none is safe from it;
    \textsuperscript{29}\thinspace{}those who guard their chas\-ti\-ty,
    \textsuperscript{30}\thinspace{}ex\-cept with their wives;
    \textsuperscript{32}\thinspace{}those who are at\-ten\-tive to their trusts and pledges;
    \textsuperscript{33}\thinspace{}those who bear wit\-ness hon\-est\-ly;
    \textsuperscript{34}\thinspace{}those who guard their prayers.
    \textsuperscript{35}\thinspace{}They will be honored in the Gardens.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{73:7}\thinspace{}You are bus\-y for much of the day,
    \textsuperscript{8}\thinspace{}but re\-mem\-ber your Lord and de\-vote your\-self to Him whole\-heart\-ed\-ly.
    \textsuperscript{11}\thinspace{}Let Me deal with the deniers who live in lux\-u\-ry. Bear with them for a lit\-tle while.
\end{para}

\begin{para}
    \textsuperscript{73:20}\thinspace{}Re\-cite as much of the Qur'an as may be eas\-y for you. He knows that some of you are sick, others trav\-el through\-out the land seek\-ing God's boun\-ty, and others fight in God's cause. There\-fore, re\-cite as much of the Qur'an as may be eas\-y for you, es\-tab\-lish reg\-u\-lar pray\-er, and spend in char\-i\-ty.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{74:3}\thinspace{}Glo\-ri\-fy your Lord,
    \textsuperscript{4}\thinspace{}pu\-ri\-fy your\-self,
    \textsuperscript{5}\thinspace{}and shun all filth.
    \textsuperscript{6}\thinspace{}Don't be\-stow fa\-vors to re\-ceive in\-crease,
    \textsuperscript{7}\thinspace{}but be pa\-tient for your Lord.
\end{para}

\begin{para}
    \textsuperscript{74:11}\thinspace{}Let Me a\-lone deal with him whom I a\-lone cre\-at\-ed,
    \textsuperscript{12}\thinspace{}to whom I gave vast wealth
    \textsuperscript{13}\thinspace{}and chil\-dren by his side,
    \textsuperscript{14}\thinspace{}for whom I made life com\-fort\-a\-ble,
    \textsuperscript{15}\thinspace{}yet he still wants Me to give him more.
\end{para}

\begin{para}
    \textsuperscript{74:40}\thinspace{}The companions of the right will be in gardens asking one an\-oth\-er
    \textsuperscript{41}\thinspace{}a\-bout the sinners,
    \textsuperscript{42}\thinspace{}“What led you in\-to hell-fire?”
    \textsuperscript{43}\thinspace{}They will an\-swer, “We nei\-ther prayed
    \textsuperscript{44}\thinspace{}nor fed the poor,
    \textsuperscript{45}\thinspace{}but we in\-dulged in vain con\-ver\-sa\-tion with others
    \textsuperscript{46}\thinspace{}and de\-nied the Day of Judg\-ment
    \textsuperscript{47}\thinspace{}un\-til cer\-tain\-ty came up\-on us.”
\end{para}

\begin{para}
    \textsuperscript{74:49}\thinspace{}Why do they turn a\-way from ad\-mo\-ni\-tion
    \textsuperscript{50}\thinspace{}like fright\-ened donkeys
    \textsuperscript{51}\thinspace{}flee\-ing from a li\-on?
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{76:7}\thinspace{}The right\-eous ful\-fill their vows; they fear a day of wide\-spread evil;
    \textsuperscript{8}\thinspace{}they feed the poor, the or\-phan, and the cap\-tive,
    \textsuperscript{9}\thinspace{}say\-ing, “We feed you for the sa\-ke of God a\-lone. We seek nei\-ther re\-ward nor grat\-i\-tude from you.”
\end{para}

\begin{para}
    \textsuperscript{76:24}\thinspace{}Pa\-tient\-ly a\-wait your Lord's judg\-ment, and don't o\-bey the sin\-ner or dis\-be\-liev\-er.
    \textsuperscript{25}\thinspace{}Re\-mem\-ber your Lord's name morn\-ing and eve\-ning.
    \textsuperscript{26}\thinspace{}Bow to Him dur\-ing the night, and glo\-ri\-fy Him through\-out the long night.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{80:1}\thinspace{}Muhammad frowned and turned a\-way
    \textsuperscript{2}\thinspace{}be\-cause the blind man came to him.
    \textsuperscript{3}\thinspace{}For all you know, he might be\-come pure,
    \textsuperscript{4}\thinspace{}or be reminded of some\-thing use\-ful.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{81:1}\thinspace{}When the sun is wrapped in dark\-ness,
    \textsuperscript{2}\thinspace{}when the stars fall,
    \textsuperscript{3}\thinspace{}when the mountains van\-ish,
    \textsuperscript{8}\thinspace{}when the ba\-by girl bur\-ied a\-live is asked
    \textsuperscript{9}\thinspace{}for what sin she was killed,
    \textsuperscript{10}\thinspace{}when the pages are laid o\-pen,
    \textsuperscript{14}\thinspace{}then eve\-ry per\-son will know what he has pre\-pared for him\-self.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{83:2}\thinspace{}Woe to those who de\-mand a full meas\-ure for them\-selves
    \textsuperscript{3}\thinspace{}but give to others less than what's due.
    \textsuperscript{4}\thinspace{}Do they not think that they will be resurrected,
    \textsuperscript{6}\thinspace{}on a Day when all men will stand be\-fore the Lord of the worlds?
\end{para}

\begin{para}
    \textsuperscript{83:29}\thinspace{}Sinners used to laugh at believers,
    \textsuperscript{30}\thinspace{}wink at each oth\-er when they passed by,
    \textsuperscript{31}\thinspace{}joke a\-bout them when they returned to their own peo\-ple,
    \textsuperscript{32}\thinspace{}and say, “These peo\-ple are tru\-ly lost,”
    \textsuperscript{33}\thinspace{}but they were not sent as guardians o\-ver them.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{87:8}\thinspace{}We will make the path to\-wards ease eas\-y for you.
    \textsuperscript{9}\thinspace{}So re\-mind others, if a reminder will help them.
    \textsuperscript{10}\thinspace{}Those who fear God will re\-ceive the reminder,
    \textsuperscript{11}\thinspace{}but the wick\-ed will a\-void it.
\end{para}

\begin{para}
    \textsuperscript{87:14}\thinspace{}Those will pros\-per who pu\-ri\-fy them\-selves,
    \textsuperscript{15}\thinspace{}re\-mem\-ber the name of their Lord, and pray.
    \textsuperscript{16}\thinspace{}Yet you pre\-fer the life of this world
    \textsuperscript{17}\thinspace{}e\-ven though the Here\-af\-ter is bet\-ter and more en\-dur\-ing.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{89:15}\thinspace{}As for man, when his Lord tries him by giv\-ing him hon\-or and gifts, he says “My Lord has honored me.”
    \textsuperscript{16}\thinspace{}But when He tries him by restricting his pro\-vi\-sion, he says, “My Lord has hu\-mil\-i\-at\-ed me.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{92:4}\thinspace{}You all strive for di\-verse goals.
    \textsuperscript{5}\thinspace{}For him who gives, fears God,
    \textsuperscript{6}\thinspace{}and believes in the best,
    \textsuperscript{7}\thinspace{}We will smooth his path to\-wards ease.
    \textsuperscript{8}\thinspace{}But for him who is sting\-y, considers him\-self self-sufficient,
    \textsuperscript{9}\thinspace{}and denies the best,
    \textsuperscript{10}\thinspace{}We will smooth his path to\-wards dif\-fi\-cul\-ty,
    \textsuperscript{11}\thinspace{}and his wealth will not help him when he falls.
\end{para}

\begin{para}
    \textsuperscript{92:17}\thinspace{}The right\-eous will re\-main far from hell.
    \textsuperscript{18}\thinspace{}Those who spend their wealth on self-purification,
    \textsuperscript{19}\thinspace{}not to re\-turn a fa\-vor
    \textsuperscript{20}\thinspace{}but on\-ly seek\-ing their Lord's face,
    \textsuperscript{21}\thinspace{}will be sat\-is\-fied.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{93:6}\thinspace{}Muhammad, did He not find you as an or\-phan and shel\-ter you,
    \textsuperscript{7}\thinspace{}find you lost and guide you,
    \textsuperscript{8}\thinspace{}find you in need and make you rich?
    \textsuperscript{9}\thinspace{}There\-fore, don't op\-press the or\-phan,
    \textsuperscript{10}\thinspace{}nor re\-pel the beg\-gar,
    \textsuperscript{11}\thinspace{}but talk a\-bout your Lord's blessings.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{96:3}\thinspace{}Your Lord is the most Boun\-ti\-ful,
    \textsuperscript{4}\thinspace{}who taught by the pen,
    \textsuperscript{5}\thinspace{}who taught man what he did\-n't know.
    \textsuperscript{6}\thinspace{}But man transgresses
    \textsuperscript{7}\thinspace{}by con\-sid\-er\-ing him\-self self-sufficient.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{104:1}\thinspace{}Woe to eve\-ry slanderer and back\-bit\-er
    \textsuperscript{2}\thinspace{}who amasses wealth and con\-tin\-u\-ous\-ly counts it,
    \textsuperscript{3}\thinspace{}think\-ing that his wealth will make him im\-mor\-tal.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{107:1}\thinspace{}Have you con\-sid\-ered the man who denies the Judg\-ment?
    \textsuperscript{2}\thinspace{}He is the one who pushes the or\-phan a\-way,
    \textsuperscript{3}\thinspace{}and does\-n't en\-cour\-age the feed\-ing of the poor.
    \textsuperscript{4}\thinspace{}So woe to those who pray
    \textsuperscript{5}\thinspace{}but are heed\-less of their pray\-er;
    \textsuperscript{6}\thinspace{}who on\-ly want to be seen,
    \textsuperscript{7}\thinspace{}and with\-hold small kindnesses.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{109:1}\thinspace{}Muhammad, say, “Disbelievers,
    \textsuperscript{2}\thinspace{}I don't wor\-ship what you wor\-ship,
    \textsuperscript{3}\thinspace{}nor do you wor\-ship what I wor\-ship.
    \textsuperscript{6}\thinspace{}You have your re\-li\-gion and I have mine.”
\end{para}

\begin{para}
\end{para}


\end{document}

% Copyright 2021 Michael Hendricks
%
% Permission is granted to use, copy, modify, and distribute
% this work for any purpose with or without fee.
%
% This work is provided as-is without any express or implied
% warranty. In no event is the author liable for any damages
% arising from this work.
