%
% see tex.stackexchange.com/a/600/151276
\def\changemargin#1#2{\list{}{\rightmargin#2\leftmargin#1}\item[]}
\let\endchangemargin=\endlist

\documentclass[openany,12pt,english]{book}
\usepackage[utf8]{inputenc}
\usepackage{babel}
\usepackage{microtype}
\usepackage{csquotes}

\usepackage[
  paperwidth=8.5in,
  paperheight=11in,
  top=1in,
  outer=1in,
  bottom=1.25in,
  inner=1.5in
]{geometry}

\title{Jesus}
\author{}
\date{}

\newenvironment{para}{\par\pretolerance=100\tolerance=200\setlength{\emergencystretch}{0.6em}\relax}{\par}

\begin{document}

\maketitle

% Copyright page
{
\thispagestyle{empty}
\vspace*{\fill}
\begin{changemargin}{.8in}{.8in}
\begin{center}
\textcopyright{}~2021 Michael Hendricks

\vspace{0.5\baselineskip}

Permission is granted to use, copy, modify, and distribute
this work for any purpose with or without fee.

\vspace{0.5\baselineskip}

This work is provided as-is without any express or implied
warranty. In no event is the author liable for any damages
arising from this work.
\end{center}
\end{changemargin}
}
\pagenumbering{arabic}
\clearpage{}

% Table of Contents
\clearpage{}

\markboth{}{}
\twocolumn[\begin{changemargin}{.6in}{.6in}
  \begin{center}
    \itshape
    An a\-bridg\-ment of the Gos\-pel of Mark fo\-cus\-ing on the Tri\-ple Tra\-di\-tion.
    Verse references such
    as~\textsuperscript{2:6}\thinspace{}indicate
    the original source of the text, but
    don't imply a full quotation.
  \end{center}
\end{changemargin}]{}


\begin{para}
    \textsuperscript{1:4}\thinspace{}John the Baptist appeared in the wil\-der\-ness, preaching a bap\-tism of re\-pent\-ance for the for\-give\-ness of sins.
    \textsuperscript{5}\thinspace{}All the peo\-ple were go\-ing out to him and were be\-ing bap\-tized by him in the riv\-er Jor\-dan, con\-fess\-ing their sins.
    \textsuperscript{6}\thinspace{}John was clothed with camel's hair, wore a leath\-er belt a\-round his waist, and ate locusts and wild hon\-ey.
\end{para}

\begin{para}
    \textsuperscript{1:7}\thinspace{}He preached, say\-ing, “Af\-ter me co\-mes he who is might\-i\-er than I. I'm not wor\-thy to stoop down and un\-tie the strap of his sandals.
    \textsuperscript{8}\thinspace{}I bap\-tized you with wa\-ter, but he'll bap\-tize you with the ho\-ly spir\-it.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{1:9}\thinspace{}In those days Jesus was bap\-tized by John in the Jor\-dan.
    \textsuperscript{10}\thinspace{}Just as he was com\-ing up out of the wa\-ter, he saw the heavens be\-ing torn o\-pen and the spir\-it descending on him like a dove.
    \textsuperscript{11}\thinspace{}And a voice came out of the heavens, “You are my be\-lov\-ed son; with you I am well pleased.”
\end{para}

\begin{para}
    \textsuperscript{1:12}\thinspace{}Im\-me\-di\-ate\-ly the spir\-it drove him in\-to the wil\-der\-ness.
    \textsuperscript{13}\thinspace{}He was in the wil\-der\-ness for\-ty days, be\-ing tempted by the temp\-ter. He was with the wild animals, and the angels ministered to him.
\end{para}

\begin{para}
    \textsuperscript{1:14}\thinspace{}Af\-ter John was arrested, Jesus came in\-to Gal\-i\-lee, preaching the gos\-pel of God,
    \textsuperscript{15}\thinspace{}and say\-ing, “The time is ful\-filled, and God's king\-dom is at hand. Re\-pent and be\-lieve in the gos\-pel.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{1:16}\thinspace{}Walk\-ing a\-long by the Sea of Gal\-i\-lee, he saw Simon and Andrew the broth\-er of Simon cast\-ing a net in\-to the sea, for they were fishermen.
    \textsuperscript{17}\thinspace{}Jesus said to them, “Come, fol\-low me, and I'll make you be\-come fishers of men.”
    \textsuperscript{18}\thinspace{}Im\-me\-di\-ate\-ly they left the nets, and followed him.
\end{para}

\begin{para}
    \textsuperscript{1:19}\thinspace{}Go\-ing on a lit\-tle far\-ther, he saw James the son of Zebedee, and John his broth\-er, who were in the boat mend\-ing their nets.
    \textsuperscript{20}\thinspace{}Im\-me\-di\-ate\-ly he called them, and they left their fa\-ther Zebedee in the boat with the hired servants, and followed him.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{1:29}\thinspace{}They entered the house of Simon and Andrew, with James and John.
\end{para}

\begin{para}
    \textsuperscript{1:30}\thinspace{}Now Simon's mother-in-law lay sick with a fe\-ver, and im\-me\-di\-ate\-ly they told him a\-bout her.
    \textsuperscript{31}\thinspace{}He came, took her by the hand, and helped her up. The fe\-ver left her, and she be\-gan to wait on them.
\end{para}

\begin{para}
    \textsuperscript{1:32}\thinspace{}That eve\-ning, af\-ter the sun had set, they brought to him all who were sick and pos\-sessed by demons.
    \textsuperscript{33}\thinspace{}The whole cit\-y gathered to\-geth\-er at the door.
    \textsuperscript{34}\thinspace{}He healed man\-y who were sick with var\-i\-ous diseases, and cast out man\-y demons. He would not al\-low the demons to speak, be\-cause they knew him.
    \textsuperscript{39}\thinspace{}He went in\-to their synagogues through\-out all Gal\-i\-lee, preaching and cast\-ing out demons.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{1:40}\thinspace{}A lep\-er came to him, and kneel\-ing said to him, “If you're will\-ing, you can make me clean.”
\end{para}

\begin{para}
    \textsuperscript{1:41}\thinspace{}Moved with com\-pas\-sion, he touched him, and said to him, “I am will\-ing. Be clean.”
    \textsuperscript{42}\thinspace{}Im\-me\-di\-ate\-ly the lep\-ro\-sy left him, and he was cleansed.
    \textsuperscript{43}\thinspace{}Jesus sent him a\-way at once and sternly warned him,
    \textsuperscript{44}\thinspace{}“Say noth\-ing to any\-one, but go show your\-self to the priest and of\-fer for your cleans\-ing what Mo\-ses commanded, as a tes\-ti\-mo\-ny to them.”
\end{para}

\begin{para}
    \textsuperscript{1:45}\thinspace{}But he went out and be\-gan to talk free\-ly a\-bout it, spreading the news, so that Jesus could no long\-er en\-ter a cit\-y pub\-lic\-ly, but stayed out in de\-sert\-ed places, and peo\-ple came to him from eve\-ry\-where.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{2:1}\thinspace{}When he returned to Capernaum some days lat\-er, peo\-ple heard that he was at home.
    \textsuperscript{2}\thinspace{}Man\-y were gathered to\-geth\-er, so that there was no more room for them, not e\-ven near the door. He preached the word to them.
    \textsuperscript{3}\thinspace{}They came, bring\-ing to him a par\-a\-lyt\-ic, car\-ried by four of them.
    \textsuperscript{4}\thinspace{}When they could not come near him be\-cause of the crowd, they re\-moved the roof a\-bove him, made an o\-pen\-ing, and let down the mat on which the par\-a\-lyt\-ic was ly\-ing.
    \textsuperscript{5}\thinspace{}When Jesus saw their faith, he said to the par\-a\-lyt\-ic, “Son, your sins are for\-giv\-en.”
\end{para}

\begin{para}
    \textsuperscript{2:6}\thinspace{}But some of the scribes were sit\-ting there, think\-ing to them\-selves,
    \textsuperscript{7}\thinspace{}“Why does he talk like that? He's blas\-phem\-ing. Who can for\-give sins but God a\-lone?”
\end{para}

\begin{para}
    \textsuperscript{2:8}\thinspace{}Im\-me\-di\-ate\-ly Jesus perceived in his spir\-it what they were think\-ing, and said to them, “Why do you ques\-tion these things in your hearts?
    \textsuperscript{9}\thinspace{}Which is eas\-i\-er, to say to the par\-a\-lyt\-ic, ‘Your sins are for\-giv\-en,’ or to say, ‘Rise, pick up your mat, and walk’?
    \textsuperscript{10}\thinspace{}But that you may know that the son of man has au\-thor\-i\-ty on earth to for\-give sins,” he said to the par\-a\-lyt\-ic,
    \textsuperscript{11}\thinspace{}“I tell you, Rise, pick up your mat, and go home.”
\end{para}

\begin{para}
    \textsuperscript{2:12}\thinspace{}He got up, im\-me\-di\-ate\-ly pick\-ed up the mat, and went out in front of them all, so that they were all a\-mazed, and glo\-ri\-fied God, say\-ing, “We've nev\-er seen any\-thing like this.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{2:13}\thinspace{}He went out a\-gain be\-side the sea, and all the crowd came to him, and he taught them.
    \textsuperscript{14}\thinspace{}As he passed by, he saw Levi the son of Alphaeus sit\-ting in the tax booth, and said to him, “Fol\-low me.” He got up and followed him.
\end{para}

\begin{para}
    \textsuperscript{2:15}\thinspace{}As he re\-clined at the ta\-ble in his house, man\-y tax collectors and sinners re\-clined with Jesus and his disciples, for there were man\-y who followed him.
    \textsuperscript{16}\thinspace{}The scribes of the Pharisees asked his disciples, “Why does he eat with tax collectors and sinners?”
\end{para}

\begin{para}
    \textsuperscript{2:17}\thinspace{}When Jesus heard it, he said to them, “Those who are health\-y don't need a doc\-tor, but those who are sick. I did\-n't come to call the right\-eous, but sinners.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{2:18}\thinspace{}Some peo\-ple came and asked him, “Why do John's disciples and the disciples of the Pharisees fast, but your disciples don't fast?”
\end{para}

\begin{para}
    \textsuperscript{2:19}\thinspace{}Jesus said to them, “Can wed\-ding guests fast while the groom is with them? As long as they have the groom with them, they can't fast.
    \textsuperscript{20}\thinspace{}The days will come, when the groom is tak\-en a\-way from them, and then they will fast in that day.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{2:23}\thinspace{}One Sabbath he was go\-ing through the grainfields, and his disciples be\-gan to pluck some heads of grain as they went.
    \textsuperscript{24}\thinspace{}The Pharisees said to him, “Look. Why are they do\-ing what's un\-law\-ful on the Sabbath?”
\end{para}

\begin{para}
    \textsuperscript{2:25}\thinspace{}He said to them, “Have\-n't you read what David did when he and his companions were in need and hun\-gry?
    \textsuperscript{26}\thinspace{}He entered the house of God and ate the con\-se\-crat\-ed bread, which is on\-ly law\-ful for the priests to eat, and he al\-so gave some to his companions.”
\end{para}

\begin{para}
    \textsuperscript{2:27}\thinspace{}He said to them, “The Sabbath was made for man, not man for the Sabbath.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{3:1}\thinspace{}He entered the syn\-a\-gogue a\-gain, and a man with a withered hand was there.
    \textsuperscript{2}\thinspace{}They watched Jesus to see if he would heal him on the Sabbath, so that they might ac\-cuse him.
\end{para}

\begin{para}
    \textsuperscript{3:3}\thinspace{}He said to the man with the withered hand, “Come for\-ward.”
    \textsuperscript{4}\thinspace{}He asked them, “Is it law\-ful on the Sabbath to do good or to do harm, to save life or to kill?” But they were si\-lent.
\end{para}

\begin{para}
    \textsuperscript{3:5}\thinspace{}He looked a\-round at them with an\-ger, grieved at the hard\-ness of their hearts, and said to the man, “Stretch out your hand.” He stretched it out, and his hand was re\-stored.
    \textsuperscript{6}\thinspace{}The Pharisees went out and im\-me\-di\-ate\-ly held coun\-sel with the Herodians a\-gainst him, how to de\-stroy him.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{3:7}\thinspace{}Jesus with\-drew with his disciples to the sea, and a large crowd followed him.
    \textsuperscript{8}\thinspace{}Hear\-ing the great things he did, a large crowd came to him.
\end{para}

\begin{para}
    \textsuperscript{3:9}\thinspace{}Be\-cause of the crowd, he told his disciples to have a boat wait for him, so they would\-n't crush him.
    \textsuperscript{10}\thinspace{}Since he had healed so man\-y, eve\-ry\-one who had diseases pressed a\-round him to tou\-ch him.
    \textsuperscript{11}\thinspace{}When\-ev\-er the un\-clean spirits saw him, they fell down in front of him, and cried out, “You are the son of God.”
    \textsuperscript{12}\thinspace{}And he sternly ordered them not to make him known.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{3:13}\thinspace{}He went up on the moun\-tain and summoned those he wanted, and they came to him.
    \textsuperscript{14}\thinspace{}He ap\-point\-ed twelve, whom he al\-so named apostles, to be with him, and to send them out to preach,
    \textsuperscript{15}\thinspace{}and to have au\-thor\-i\-ty to cast out demons.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{3:22}\thinspace{}The scribes said, “He casts out demons by the prince of demons.”
\end{para}

\begin{para}
    \textsuperscript{3:23}\thinspace{}He called them o\-ver to him, and said to them in parables, “How can the temp\-ter cast out the temp\-ter?
    \textsuperscript{24}\thinspace{}If a king\-dom is di\-vid\-ed a\-gainst it\-self, that king\-dom can\-not stand.
    \textsuperscript{25}\thinspace{}If a house is di\-vid\-ed a\-gainst it\-self, that house will not be a\-ble to stand.
    \textsuperscript{26}\thinspace{}If the temp\-ter has ris\-en up a\-gainst him\-self, and is di\-vid\-ed, he can\-not stand, but is fin\-ished.
    \textsuperscript{27}\thinspace{}No one can en\-ter a strong man's house, and plun\-der his goods, un\-less he first ties up the strong man, and then he can plun\-der his house.”
\end{para}

\begin{para}
    \textsuperscript{3:28}\thinspace{}“Tru\-ly I say to you, the sons of men will be for\-giv\-en all sins and what\-ev\-er blasphemies they speak,
    \textsuperscript{29}\thinspace{}but who\-ev\-er blasphemes a\-gainst the ho\-ly spir\-it will nev\-er be for\-giv\-en, but is guilt\-y of an e\-ter\-nal sin,”
    \textsuperscript{30}\thinspace{}be\-cause they said, “He has an un\-clean spir\-it.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{3:32}\thinspace{}A crowd was sit\-ting a\-round him, and said to him, “Your moth\-er and brothers are out\-side looking for you.”
\end{para}

\begin{para}
    \textsuperscript{3:33}\thinspace{}He answered them, “Who are my moth\-er and my brothers?”
    \textsuperscript{34}\thinspace{}Looking at those seated a\-round him, he said, “Here are my moth\-er and my brothers.
    \textsuperscript{35}\thinspace{}Who\-ev\-er does the will of God, he is my broth\-er and sis\-ter and moth\-er.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{4:1}\thinspace{}He be\-gan to teach by the sea a\-gain. A ver\-y great crowd gathered a\-round him, so he got in\-to a boat on the sea and sat down. The whole crowd was by the sea on the land.
    \textsuperscript{2}\thinspace{}He taught them man\-y things in parables, and in his teach\-ing said to them,
    \textsuperscript{3}\thinspace{}“A sower went out to sow.
    \textsuperscript{4}\thinspace{}As he sowed, some seed fell a\-long the path, and the birds came and devoured it.
    \textsuperscript{5}\thinspace{}Oth\-er seed fell on rock\-y ground, where it had lit\-tle soil. It sprang up quick\-ly, be\-cause it had no depth of soil.
    \textsuperscript{6}\thinspace{}When the sun rose, it was scorched. Be\-cause it had no root, it withered a\-way.
    \textsuperscript{7}\thinspace{}Oth\-er seed fell a\-mong thorns. The thorns grew up, and choked it, and it pro\-duced no grain.
    \textsuperscript{8}\thinspace{}Oth\-er seed fell on good soil. It grew up, in\-creased, and pro\-duced grain, bring\-ing forth thir\-ty, six\-ty, and a hun\-dred\-fold.
    \textsuperscript{9}\thinspace{}He who has ears to hear, let him hear.”
\end{para}

\begin{para}
    \textsuperscript{4:10}\thinspace{}When he was a\-lone, those a\-round him with the twelve asked him a\-bout the parables.
    \textsuperscript{14}\thinspace{}He said to them, “The sower sows the word.
    \textsuperscript{15}\thinspace{}These are the ones a\-long the path, where the word is sown. When they hear, im\-me\-di\-ate\-ly the temp\-ter co\-mes, and takes a\-way the word sown in them.
    \textsuperscript{16}\thinspace{}These are the ones sown on rock\-y ground. When they hear the word, they im\-me\-di\-ate\-ly re\-ceive it with joy,
    \textsuperscript{17}\thinspace{}but they have no root in them\-selves, and en\-dure brief\-ly. When trou\-ble or per\-se\-cu\-tion co\-mes be\-cause of the word, they im\-me\-di\-ate\-ly fall a\-way.
    \textsuperscript{18}\thinspace{}These are the ones sown a\-mong thorns. When they hear the word,
    \textsuperscript{19}\thinspace{}the worries of the world, the de\-ceit\-ful\-ness of rich\-es, and the de\-sire for oth\-er things en\-ter in and choke the word, and it becomes un\-fruit\-ful.
    \textsuperscript{20}\thinspace{}These are the ones sown on good soil. They hear the word, ac\-cept it, and bear fruit: thir\-ty, six\-ty, and a hun\-dred\-fold.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{4:21}\thinspace{}He said to them, “Is a lamp brought in to be put un\-der a bas\-ket, or un\-der a bed, and not on a stand?
    \textsuperscript{22}\thinspace{}Noth\-ing is hid\-den that will not be revealed, and noth\-ing is se\-cret that will not come to light.”
\end{para}

\begin{para}
    \textsuperscript{4:24}\thinspace{}He said to them, “Pay at\-ten\-tion to what you hear. With the meas\-ure you use, it will be meas\-ured to you, and e\-ven more will be added to you.
    \textsuperscript{25}\thinspace{}For who\-ev\-er has will be giv\-en more; and who\-ev\-er does\-n't have, e\-ven what he has will be tak\-en a\-way.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{4:30}\thinspace{}He said, “With what can we com\-pare God's king\-dom? What par\-a\-ble can we use to de\-scribe it?
    \textsuperscript{31}\thinspace{}It's like a mus\-tard seed. When sown up\-on the soil, it's smaller than all oth\-er seeds on the ground,
    \textsuperscript{32}\thinspace{}yet when it's sown, it grows up and becomes larg\-er than all the gar\-den plants, and puts out large branches, so the birds of the air can nest in its shade.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{4:36}\thinspace{}Leav\-ing the crowd, they took him with them since he was in the boat; and oth\-er boats were with him.
    \textsuperscript{37}\thinspace{}A great wind\-storm a\-rose, and the waves were break\-ing o\-ver the boat, so that the boat was al\-read\-y fill\-ing up.
\end{para}

\begin{para}
    \textsuperscript{4:38}\thinspace{}He was in the stern, sleep\-ing on the cush\-ion. They woke him up, and said to him, “Teach\-er, don't you care that we're go\-ing to die?”
\end{para}

\begin{para}
    \textsuperscript{4:39}\thinspace{}He woke up, re\-buked the wind, and said to the sea, “Peace. Be still.” The wind stopped, and there was a great calm.
    \textsuperscript{40}\thinspace{}He said to them, “Why are you a\-fraid? Do you still have no faith?”
\end{para}

\begin{para}
    \textsuperscript{4:41}\thinspace{}They were ter\-ri\-fied, and asked each oth\-er, “Who is this, that e\-ven the wind and the sea o\-bey him?”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{5:1}\thinspace{}They came to the oth\-er side of the sea.
    \textsuperscript{2}\thinspace{}As soon as Jesus got out of the boat, a man with an un\-clean spir\-it came out of the tombs and met him.
    \textsuperscript{3}\thinspace{}He lived in the tombs. No one could bind him any\-more, not e\-ven with a chain,
    \textsuperscript{4}\thinspace{}be\-cause he had of\-ten been bound with shackles and chains, but had torn the chains a\-part, and broke the shackles in pieces. No one was strong e\-nough to sub\-due him.
    \textsuperscript{5}\thinspace{}Night and day, in the tombs and in the mountains, he was al\-ways cry\-ing out, and cut\-ting him\-self with stones.
\end{para}

\begin{para}
    \textsuperscript{5:6}\thinspace{}When he saw Jesus from a dis\-tance, he ran and knelt down in front of him.
    \textsuperscript{7}\thinspace{}He cried out with a loud voice, “What do you want with me, Jesus, you son of the most high God? I beg you by God, not to tor\-ment me.”
    \textsuperscript{8}\thinspace{}For he had said to him, “Come out of the man, you un\-clean spir\-it.”
\end{para}

\begin{para}
    \textsuperscript{5:11}\thinspace{}There was a large herd of pigs on the hill\-side, feed\-ing.
    \textsuperscript{13}\thinspace{}So he gave the spirits per\-mis\-sion to en\-ter the pigs. The un\-clean spirits came out, and entered the pigs. The herd of a\-bout two thou\-sand rushed down the steep bank in\-to the sea and drowned.
\end{para}

\begin{para}
    \textsuperscript{5:14}\thinspace{}The herdsmen fled and told it in the cit\-y and the coun\-try\-side. Peo\-ple came to see what had happened.
    \textsuperscript{15}\thinspace{}They came to Jesus, and saw him who had been pos\-sessed with the demons, sit\-ting, clothed, and in his right mind; and they were a\-fraid.
    \textsuperscript{16}\thinspace{}Those who had seen it de\-scribed to them what had happened to the demon-possessed man and the pigs.
    \textsuperscript{17}\thinspace{}They be\-gan to beg Jesus to leave their re\-gion.
\end{para}

\begin{para}
    \textsuperscript{5:18}\thinspace{}As he was entering the boat, the man who had been pos\-sessed with demons begged to go with him.
    \textsuperscript{19}\thinspace{}He did not let him, but said to him, “Go home to your own peo\-ple, and tell them how much the Lord has done for you, and how he had mer\-cy on you.”
    \textsuperscript{20}\thinspace{}He went a\-way and be\-gan to pro\-claim how much Jesus had done for him. Eve\-ry\-one was a\-mazed.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{5:21}\thinspace{}When Jesus had crossed a\-gain by boat to the oth\-er side, a large crowd gathered a\-round him, and he was by the sea.
    \textsuperscript{22}\thinspace{}One of the syn\-a\-gogue leaders, named Jairus, came. See\-ing Jesus, he fell at his feet,
    \textsuperscript{23}\thinspace{}begging him ear\-nest\-ly, “My lit\-tle daugh\-ter is dy\-ing. Come and lay your hands on her, so that she will get well, and live.”
\end{para}

\begin{para}
    \textsuperscript{5:24}\thinspace{}He went with him. A large crowd followed him, and pressed a\-round him.
    \textsuperscript{25}\thinspace{}A wom\-an, bleed\-ing for twelve years,
    \textsuperscript{26}\thinspace{}had suffered much un\-der man\-y doctors, and had spent all she had, and was no bet\-ter, but rath\-er grew worse.
    \textsuperscript{27}\thinspace{}Hav\-ing heard a\-bout Jesus, she came up be\-hind him in the crowd, and touched his cloak.
    \textsuperscript{28}\thinspace{}For she said, “If I can just tou\-ch his clothes, I'll get well.”
    \textsuperscript{29}\thinspace{}Im\-me\-di\-ate\-ly, the flow of blood dried up. She felt in her bod\-y that she was healed of her af\-flic\-tion.
\end{para}

\begin{para}
    \textsuperscript{5:30}\thinspace{}Jesus re\-al\-ized at once that pow\-er had gone out from him, and turned a\-round in the crowd, and said, “Who touched my clothes?”
\end{para}

\begin{para}
    \textsuperscript{5:31}\thinspace{}His disciples said to him, “You see the crowd press\-ing a\-round you, and yet you ask, ‘Who touched me?’”
\end{para}

\begin{para}
    \textsuperscript{5:32}\thinspace{}He looked a\-round to see who had done it.
    \textsuperscript{33}\thinspace{}The wom\-an, fearing and trem\-bling, know\-ing what had happened to her, came and fell down in front of him, and told him the whole truth.
    \textsuperscript{34}\thinspace{}He said to her, “Daugh\-ter, your faith has made you well. Go in peace, and be healed of your af\-flic\-tion.”
\end{para}

\begin{para}
    \textsuperscript{5:35}\thinspace{}While he was still speak\-ing, peo\-ple came from Jairus's house and said, “Your daugh\-ter is dead. Why both\-er the teach\-er any\-more?”
\end{para}

\begin{para}
    \textsuperscript{5:36}\thinspace{}Jesus o\-ver\-heard what they said, and told Jairus, “Don't be a\-fraid, just be\-lieve.”
\end{para}

\begin{para}
    \textsuperscript{5:37}\thinspace{}He al\-lowed no one to fol\-low him ex\-cept Pe\-ter, James, and John.
    \textsuperscript{38}\thinspace{}When they came to Jairus's house, he saw a com\-mo\-tion, and peo\-ple weep\-ing and wailing loud\-ly.
    \textsuperscript{39}\thinspace{}He went in\-side, and asked them, “Why are you mak\-ing a com\-mo\-tion, and weep\-ing? The child is\-n't dead but a\-sleep.”
\end{para}

\begin{para}
    \textsuperscript{5:40}\thinspace{}They laughed at him, but he sent them all out\-side. He took the child's fa\-ther and moth\-er and those who were with him, and went in to where the child was.
    \textsuperscript{41}\thinspace{}Tak\-ing her by the hand, he said to her, “Lit\-tle girl, I say to you, get up.”
    \textsuperscript{42}\thinspace{}Im\-me\-di\-ate\-ly the girl got up and be\-gan walk\-ing. They were com\-plete\-ly a\-mazed.
    \textsuperscript{43}\thinspace{}He gave them strict orders that no one should know a\-bout this, and told them to give her some\-thing to eat.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{6:1}\thinspace{}He left there, and went to his home\-town, and his disciples followed him.
    \textsuperscript{2}\thinspace{}When the Sabbath came, he be\-gan to teach in the syn\-a\-gogue. Man\-y who heard him were astonished, say\-ing, “Where did this man get these things? What is this wis\-dom giv\-en to him? How are such miracles done by his hands?
    \textsuperscript{3}\thinspace{}Is\-n't this the car\-pen\-ter, the son of Mar\-y, and broth\-er of James, and Joses, and Ju\-das, and Simon? Aren't his sisters here with us?” And they took of\-fense at him.
\end{para}

\begin{para}
    \textsuperscript{6:4}\thinspace{}Jesus said to them, “A proph\-et is with\-out hon\-or on\-ly in his home\-town, a\-mong his relatives, and in his own house\-hold.”
    \textsuperscript{5}\thinspace{}He could do no mir\-a\-cle there, ex\-cept to lay his hands on a few sick peo\-ple, and heal them.
    \textsuperscript{6}\thinspace{}He was a\-mazed at their un\-be\-lief.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{6:6}\thinspace{}He went a\-round the villages teach\-ing.
    \textsuperscript{7}\thinspace{}He summoned the twelve, and be\-gan to send them out in pairs. He gave them au\-thor\-i\-ty o\-ver un\-clean spirits.
    \textsuperscript{8}\thinspace{}He instructed them to take noth\-ing for their jour\-ney ex\-cept a staff: no bread, no bag, no mon\-ey in their belts;
    \textsuperscript{9}\thinspace{}but to wear sandals and not a sec\-ond shirt.
\end{para}

\begin{para}
    \textsuperscript{6:12}\thinspace{}They went out, and preached that peo\-ple should re\-pent.
    \textsuperscript{13}\thinspace{}They cast out man\-y demons, anointed man\-y sick peo\-ple with oil, and healed them.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{6:14}\thinspace{}King Her\-od heard a\-bout this be\-cause Jesus's name had be\-come well known.
    \textsuperscript{16}\thinspace{}When Her\-od heard it, he said, “John, whom I beheaded, has ris\-en.”
    \textsuperscript{17}\thinspace{}For Her\-od had mar\-ried Herodias, his brother's wife. He had ordered John to be arrested and bound in pris\-on for her sa\-ke.
    \textsuperscript{18}\thinspace{}For John had been tell\-ing Her\-od, “It's not law\-ful for you to have your brother's wife.”
    \textsuperscript{19}\thinspace{}So Herodias had a grudge a\-gainst him, and wanted to kill him. But she could not,
    \textsuperscript{20}\thinspace{}be\-cause Her\-od feared John, and kept him safe, know\-ing that he was a right\-eous and ho\-ly man. When he heard him, he was great\-ly per\-plexed, yet he enjoyed listening to him.
\end{para}

\begin{para}
    \textsuperscript{6:21}\thinspace{}An op\-por\-tu\-ni\-ty came on Herod's birth\-day. He hosted a ban\-quet for his nobles, the mil\-i\-tar\-y commanders, and the lead\-ing men of Gal\-i\-lee.
    \textsuperscript{22}\thinspace{}When Herodias's daugh\-ter came in and danced, she pleased Her\-od and his guests. The king said to the girl, “Ask me for what\-ev\-er you want, and I'll give it to you.”
\end{para}

\begin{para}
    \textsuperscript{6:25}\thinspace{}She said, “I want you to give me John the Baptist's head on a plat\-ter right now.”
\end{para}

\begin{para}
    \textsuperscript{6:26}\thinspace{}The king was great\-ly dis\-tressed, but be\-cause of his vow and his guests he did not want to ref\-use her.
    \textsuperscript{27}\thinspace{}Im\-me\-di\-ate\-ly the king sent an ex\-e\-cu\-tion\-er who went and beheaded him in the pris\-on,
    \textsuperscript{28}\thinspace{}brought his head on a plat\-ter, and gave it to the girl, and the girl gave it to her moth\-er.
\end{para}

\begin{para}
    \textsuperscript{6:29}\thinspace{}When his disciples heard a\-bout it, they came and took his bod\-y, and laid it in a tomb.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{6:32}\thinspace{}They went a\-way in the boat by them\-selves to a de\-sert\-ed place.
    \textsuperscript{33}\thinspace{}Man\-y saw them go\-ing and rec\-og\-nized them, and ran on foot from all the towns and ar\-rived a\-head of them.
    \textsuperscript{34}\thinspace{}When he went a\-shore, he saw a large crowd, and had com\-pas\-sion on them, be\-cause they were like sheep with\-out a shep\-herd. He be\-gan to teach them man\-y things.
\end{para}

\begin{para}
    \textsuperscript{6:35}\thinspace{}When it grew late, his disciples came to him, and said, “This is a des\-o\-late place, and it's get\-ting late.
    \textsuperscript{36}\thinspace{}Send them a\-way to go in\-to the sur\-round\-ing coun\-try\-side and villages to buy them\-selves some\-thing to eat.”
\end{para}

\begin{para}
    \textsuperscript{6:37}\thinspace{}He answered, “You give them some\-thing to eat.” They asked him, “Should we go and buy two hun\-dred de\-nar\-i\-i worth of bread, and give it to them to eat?”
\end{para}

\begin{para}
    \textsuperscript{6:38}\thinspace{}He said, “How man\-y loaves do you have? Go and see.” Once they had found out, they said, “Five, and two fish.”
\end{para}

\begin{para}
    \textsuperscript{6:39}\thinspace{}He commanded them all to sit down in groups on the green grass.
    \textsuperscript{40}\thinspace{}They sat down in groups of hundreds and fifties.
\end{para}

\begin{para}
    \textsuperscript{6:41}\thinspace{}He took the five loaves and the two fish. Looking up to heav\-en, he bless\-ed and broke the loaves. He gave them to his disciples to set be\-fore the peo\-ple, and di\-vid\-ed the two fish a\-mong them all.
    \textsuperscript{42}\thinspace{}Eve\-ry\-one ate and was sat\-is\-fied.
    \textsuperscript{43}\thinspace{}They pick\-ed up twelve baskets full of bro\-ken pieces and of fish.
    \textsuperscript{44}\thinspace{}There were five thou\-sand men who ate the loaves.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{7:1}\thinspace{}The Pharisees, and some of the scribes who had come from Je\-ru\-sa\-lem, gathered a\-round him,
    \textsuperscript{2}\thinspace{}and saw some of his disciples eat\-ing their bread with un\-washed hands.
    \textsuperscript{3}\thinspace{}The Pharisees on\-ly eat if they have care\-ful\-ly washed their hands, keep\-ing the tra\-di\-tion of the elders.
    \textsuperscript{4}\thinspace{}When they come from the mar\-ket\-place, they on\-ly eat af\-ter they have washed. They keep man\-y oth\-er traditions such as the wash\-ing of cups, pitchers, cop\-per pots, and din\-ing couches.
\end{para}

\begin{para}
    \textsuperscript{7:5}\thinspace{}The Pharisees and the scribes asked him, “Why don't your disciples keep the tra\-di\-tion of the elders? In\-stead, they eat with un\-washed hands.”
\end{para}

\begin{para}
    \textsuperscript{7:6}\thinspace{}He replied, “Isaiah prophesied cor\-rect\-ly a\-bout you hypocrites, as it is writ\-ten: ‘This peo\-ple hon\-ors me with their lips, but their heart is far from me.
    \textsuperscript{7}\thinspace{}They wor\-ship me in vain, teach\-ing as doctrines the commandments of men.’
    \textsuperscript{8}\thinspace{}Neglecting the com\-mand\-ment of God, you hold on to the tra\-di\-tion of men.
    \textsuperscript{9}\thinspace{}You have a fi\-ne way of rejecting the com\-mand\-ment of God in or\-der to keep your tra\-di\-tion.
    \textsuperscript{13}\thinspace{}The tra\-di\-tion that you have hand\-ed down nullifies the word of God. You do man\-y things like that.”
\end{para}

\begin{para}
    \textsuperscript{7:14}\thinspace{}He summoned the crowd a\-gain, and said to them, “Eve\-ry\-one lis\-ten to me and un\-der\-stand:
    \textsuperscript{15}\thinspace{}Noth\-ing that enters a per\-son from out\-side can de\-file him, but the things that come out of a per\-son are what de\-file him.”
\end{para}

\begin{para}
    \textsuperscript{7:17}\thinspace{}When he entered the house a\-way from the crowd, his disciples asked him a\-bout the par\-a\-ble.
    \textsuperscript{18}\thinspace{}He said to them, “Are you al\-so with\-out un\-der\-stand\-ing? Do you not see that what\-ev\-er goes in\-to a per\-son from out\-side can\-not de\-file him,
    \textsuperscript{19}\thinspace{}be\-cause it does\-n't go in\-to his heart, but in\-to his stom\-ach and is e\-lim\-i\-nat\-ed?” Thus he de\-clared all foods clean.
\end{para}

\begin{para}
    \textsuperscript{7:20}\thinspace{}He said, “What co\-mes out of a per\-son is what defiles him.
    \textsuperscript{21}\thinspace{}For from with\-in, out of a person's heart, come evil thoughts, sex\-u\-al im\-mo\-ral\-i\-ty, theft, mur\-der, a\-dul\-ter\-y,
    \textsuperscript{22}\thinspace{}coveting, wick\-ed\-ness, de\-ceit, sen\-su\-al\-i\-ty, en\-vy, slan\-der, pride, fool\-ish\-ness.
    \textsuperscript{23}\thinspace{}All these evil things pro\-ceed from with\-in, and de\-file a per\-son.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{8:11}\thinspace{}The Pharisees came and be\-gan to ar\-gue with him, de\-mand\-ing of him a sign from heav\-en to test him.
    \textsuperscript{12}\thinspace{}He sighed deep\-ly in his spir\-it, and said, “Why does this gen\-er\-a\-tion de\-mand a sign? No sign will be giv\-en to this gen\-er\-a\-tion.”
    \textsuperscript{13}\thinspace{}He left them, got back in\-to the boat, and went to the oth\-er side.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{8:27}\thinspace{}Jesus and his disciples went to the villages of Caesarea Phi\-lip\-pi. On the way he asked his disciples, “Who do peo\-ple say that I am?”
\end{para}

\begin{para}
    \textsuperscript{8:28}\thinspace{}They told him, “John the Baptist; others say Elijah; but others, one of the prophets.”
\end{para}

\begin{para}
    \textsuperscript{8:29}\thinspace{}He asked them, “But who do you say that I am?” Pe\-ter answered, “You are the Christ.”
\end{para}

\begin{para}
    \textsuperscript{8:30}\thinspace{}He warned them to tell no one a\-bout him.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{8:31}\thinspace{}He be\-gan to teach them that the son of man must suf\-fer man\-y things, and be rejected by the elders, chief priests, and scribes, and be killed, and af\-ter three days rise a\-gain.
    \textsuperscript{32}\thinspace{}He spoke openly a\-bout this. Pe\-ter took him a\-side, and be\-gan to re\-buke him.
\end{para}

\begin{para}
    \textsuperscript{8:33}\thinspace{}Turn\-ing and see\-ing his disciples, he re\-buked Pe\-ter, and said, “Get be\-hind me, Temp\-ter. For you're not think\-ing a\-bout the things of God, but the things of man.”
\end{para}

\begin{para}
    \textsuperscript{8:34}\thinspace{}He summoned the crowd with his disciples, and said to them, “If any\-one wants to come af\-ter me, let him de\-ny him\-self, take up his cross, and fol\-low me.
    \textsuperscript{35}\thinspace{}For who\-ev\-er wants to save his life will lose it, but who\-ev\-er loses his life for my sa\-ke and the gospel's will save it.
    \textsuperscript{36}\thinspace{}What does it prof\-it a man to gain the whole world, and for\-feit his soul?
    \textsuperscript{37}\thinspace{}What can a man give in ex\-change for his soul?
    \textsuperscript{38}\thinspace{}Who\-ev\-er is a\-shamed of me and my words in this a\-dul\-ter\-ous and sin\-ful gen\-er\-a\-tion, the son of man will al\-so be a\-shamed of him when he co\-mes in his father's glo\-ry with the ho\-ly angels.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{9:2}\thinspace{}Af\-ter six days Jesus took Pe\-ter, James, and John with him, and led them up a high moun\-tain by them\-selves. He was transfigured in front of them,
    \textsuperscript{3}\thinspace{}and his clothes be\-came ra\-di\-ant, in\-tense\-ly white, as no one on earth could whit\-en them.
    \textsuperscript{4}\thinspace{}Elijah and Mo\-ses appeared to them, and they were talking with Jesus.
\end{para}

\begin{para}
    \textsuperscript{9:7}\thinspace{}A cloud overshadowed them, and a voice from the cloud said, “This is my be\-lov\-ed son. Lis\-ten to him.”
    \textsuperscript{8}\thinspace{}Sud\-den\-ly, looking a\-round, they no long\-er saw any\-one with them ex\-cept Jesus.
\end{para}

\begin{para}
    \textsuperscript{9:9}\thinspace{}As they were com\-ing down the moun\-tain, he ordered them to tell no one what they had seen, un\-til the son of man had ris\-en from the dead.
    \textsuperscript{10}\thinspace{}They kept it to them\-selves, discussing what it meant to rise from the dead.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{9:14}\thinspace{}When they came to the disciples, they saw a large crowd a\-round them, and scribes ar\-gu\-ing with them.
    \textsuperscript{15}\thinspace{}As soon as the crowd saw him, they were a\-mazed, and ran to greet him.
    \textsuperscript{16}\thinspace{}He asked them, “What are you ar\-gu\-ing with them a\-bout?”
\end{para}

\begin{para}
    \textsuperscript{9:17}\thinspace{}Some\-one in the crowd answered him, “Teach\-er, I brought my son to you. He has a spir\-it that makes him mute.
    \textsuperscript{18}\thinspace{}When\-ev\-er it seizes him, it throws him down. He foams at the mouth, grinds his teeth, and becomes rig\-id. I asked your disciples to cast it out, but they could\-n't.”
\end{para}

\begin{para}
    \textsuperscript{9:19}\thinspace{}He answered them, “You un\-be\-liev\-ing gen\-er\-a\-tion, how long will I be with you? How long will I put up with you? Bring him to me.”
\end{para}

\begin{para}
    \textsuperscript{9:20}\thinspace{}They brought the boy to him. When the spir\-it saw him, it im\-me\-di\-ate\-ly threw the boy in\-to a con\-vul\-sion. He fell to the ground, and rolled a\-round, foaming at the mouth.
\end{para}

\begin{para}
    \textsuperscript{9:21}\thinspace{}Jesus asked his fa\-ther, “How long has this been hap\-pen\-ing to him?” And he said, “From child\-hood.
    \textsuperscript{22}\thinspace{}It has of\-ten thrown him in\-to fire or wa\-ter to de\-stroy him. If you can do any\-thing, have com\-pas\-sion on us, and help us.”
\end{para}

\begin{para}
    \textsuperscript{9:23}\thinspace{}Jesus said to him, “‘If you can?’ Eve\-ry\-thing is pos\-si\-ble for one who believes.”
\end{para}

\begin{para}
    \textsuperscript{9:24}\thinspace{}Im\-me\-di\-ate\-ly the boy's fa\-ther cried out, “I do be\-lieve. Help my un\-be\-lief.”
\end{para}

\begin{para}
    \textsuperscript{9:25}\thinspace{}When Jesus saw that a crowd was rap\-id\-ly gath\-er\-ing, he re\-buked the un\-clean spir\-it, say\-ing to it, “You mute and deaf spir\-it, I com\-mand you: come out of him and nev\-er en\-ter him a\-gain.”
\end{para}

\begin{para}
    \textsuperscript{9:26}\thinspace{}Af\-ter cry\-ing out, and throwing him in\-to ter\-ri\-ble convulsions, it came out. The boy was like a corpse, so that most of them said, “He's dead.”
    \textsuperscript{27}\thinspace{}But Jesus took him by the hand, and lifted him up, and he got up.
\end{para}

\begin{para}
    \textsuperscript{9:28}\thinspace{}Af\-ter he entered the house, his disciples asked him pri\-vate\-ly, “Why could\-n't we cast it out?”
\end{para}

\begin{para}
    \textsuperscript{9:29}\thinspace{}He said to them, “This kind on\-ly co\-mes out by pray\-er.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{9:30}\thinspace{}They left that place, and passed through Gal\-i\-lee. He did not want any\-one to know,
    \textsuperscript{31}\thinspace{}be\-cause he was teach\-ing his disciples, say\-ing to them, “The son of man will be delivered in\-to the hands of men, and they will kill him. Af\-ter he's killed, he'll rise three days lat\-er.”
    \textsuperscript{32}\thinspace{}But they did not un\-der\-stand this state\-ment, and were a\-fraid to ask him.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{9:33}\thinspace{}They came to Capernaum. He asked them, “What were you discussing on the way?”
    \textsuperscript{34}\thinspace{}But they were si\-lent, be\-cause they had been ar\-gu\-ing a\-bout which of them was the great\-est.
    \textsuperscript{35}\thinspace{}He sat down, called the twelve, and said to them, “If any\-one wants to be first, he must be last of all, and serv\-ant of all.”
\end{para}

\begin{para}
    \textsuperscript{9:36}\thinspace{}He took a child, and had him stand a\-mong them. Tak\-ing him in his arms, he said to them,
    \textsuperscript{37}\thinspace{}“Who\-ev\-er welcomes one child like this in my name welcomes me. Who\-ev\-er welcomes me, does\-n't wel\-come me, but him that sent me.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{9:38}\thinspace{}John said to him, “Teach\-er, we saw some\-one cast\-ing out demons in your name, and we tried to stop him, be\-cause he did\-n't fol\-low us.”
\end{para}

\begin{para}
    \textsuperscript{9:39}\thinspace{}But Jesus said, “Don't stop him. No one can do a mir\-a\-cle in my name, then soon af\-ter\-ward speak evil of me.
    \textsuperscript{40}\thinspace{}Who\-ev\-er is not a\-gainst us is for us.
    \textsuperscript{41}\thinspace{}Tru\-ly I tell you, who\-ev\-er gives you a cup of wa\-ter to drink, be\-cause you are Christ's, will nev\-er lose his re\-ward.”
\end{para}

\begin{para}
    \textsuperscript{9:42}\thinspace{}“Who\-ev\-er causes one of these lit\-tle ones who be\-lieve in me to stum\-ble, it would be bet\-ter for him if a heav\-y mill\-stone were hung a\-round his neck, and he were thrown in\-to the sea.
    \textsuperscript{43}\thinspace{}If your hand causes you to stum\-ble, cut it off. It's bet\-ter for you to en\-ter life crippled than to go to hell with two hands.
    \textsuperscript{45}\thinspace{}If your foot causes you to stum\-ble, cut it off. It's bet\-ter for you to en\-ter life lame than to be thrown in\-to hell with two feet.
    \textsuperscript{47}\thinspace{}If your eye causes you to stum\-ble, tear it out. It's bet\-ter for you to en\-ter God's king\-dom with one eye than to be thrown in\-to hell with two eyes.”
\end{para}

\begin{para}
    \textsuperscript{9:49}\thinspace{}“For eve\-ry\-one will be salt\-ed with fire.
    \textsuperscript{50}\thinspace{}Salt is good, but if the salt becomes un\-salt\-y, how will you make it salt\-y a\-gain? Have salt in your\-selves, and be at peace with one an\-oth\-er.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{10:1}\thinspace{}He left there and went in\-to the re\-gion of Judea and be\-yond the Jor\-dan. Crowds gathered a\-round him a\-gain. A\-gain he taught them, as was his cus\-tom.
\end{para}

\begin{para}
    \textsuperscript{10:2}\thinspace{}Pharisees came to him. Testing him, they asked, “Is it law\-ful for a man to di\-vorce a wife?”
\end{para}

\begin{para}
    \textsuperscript{10:3}\thinspace{}He answered them, “What did Mo\-ses com\-mand you?”
\end{para}

\begin{para}
    \textsuperscript{10:4}\thinspace{}They said, “Mo\-ses al\-lowed a man to ‘write a cer\-tif\-i\-cate of di\-vorce, and send her a\-way.’”
\end{para}

\begin{para}
    \textsuperscript{10:5}\thinspace{}But Jesus said to them, “He wrote you this com\-mand\-ment be\-cause of your hard hearts.
    \textsuperscript{6}\thinspace{}But from the be\-gin\-ning of cre\-a\-tion, ‘God made them male and fe\-male.’
    \textsuperscript{7}\thinspace{}‘There\-fore a man will leave his fa\-ther and moth\-er,
    \textsuperscript{8}\thinspace{}and the two will be\-come one flesh.’ So they are no long\-er two, but one flesh.
    \textsuperscript{9}\thinspace{}There\-fore what God has joined to\-geth\-er, let no one sep\-a\-rate.”
\end{para}

\begin{para}
    \textsuperscript{10:10}\thinspace{}In the house the disciples asked him a\-bout this a\-gain.
    \textsuperscript{11}\thinspace{}He said to them, “Who\-ev\-er divorces his wife, and marries an\-oth\-er, commits a\-dul\-ter\-y a\-gainst her.
    \textsuperscript{12}\thinspace{}If she divorces her hus\-band, and marries an\-oth\-er, she commits a\-dul\-ter\-y.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{10:13}\thinspace{}Peo\-ple were bring\-ing chil\-dren to him, so that he might tou\-ch them, but the disciples re\-buked them.
    \textsuperscript{14}\thinspace{}When Jesus saw it, he was in\-dig\-nant, and said to them, “Let the chil\-dren come to me. Don't hind\-er them, for God's king\-dom belongs to such as these.
    \textsuperscript{15}\thinspace{}Tru\-ly I tell you, who\-ev\-er does not re\-ceive God's king\-dom like a child, will nev\-er en\-ter it.”
    \textsuperscript{16}\thinspace{}He took them in his arms, laid his hands on them, and bless\-ed them.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{10:17}\thinspace{}As he was set\-ting out on his jour\-ney, a man ran up to him, knelt in front of him, and asked him, “Good teach\-er, what must I do to in\-her\-it e\-ter\-nal life?”
\end{para}

\begin{para}
    \textsuperscript{10:18}\thinspace{}Jesus said to him, “Why do you call me good? No one is good ex\-cept God a\-lone.
    \textsuperscript{19}\thinspace{}You know the commandments: ‘Do not mur\-der, Do not com\-mit a\-dul\-ter\-y, Do not steal, Do not bear false wit\-ness, Do not de\-fraud, Hon\-or your fa\-ther and moth\-er.’”
\end{para}

\begin{para}
    \textsuperscript{10:20}\thinspace{}He said to him, “I've kept all these from my youth.”
\end{para}

\begin{para}
    \textsuperscript{10:21}\thinspace{}Looking at him, Jesus loved him, and said to him, “You lack one thing: go, sell all you have, give to the poor, and you will have treas\-ure in heav\-en. Then come, fol\-low me.”
\end{para}

\begin{para}
    \textsuperscript{10:22}\thinspace{}But he was saddened by these words, and he went a\-way griev\-ing, for he had man\-y possessions.
\end{para}

\begin{para}
    \textsuperscript{10:23}\thinspace{}Jesus looked a\-round, and said to his disciples, “How hard it will be for those who have wealth to en\-ter God's king\-dom.”
    \textsuperscript{24}\thinspace{}The disciples were a\-mazed at his words. But Jesus said to them a\-gain, “Chil\-dren, how hard it is to en\-ter God's king\-dom.
    \textsuperscript{25}\thinspace{}It's eas\-i\-er for a cam\-el to go through the eye of a nee\-dle than for a rich per\-son to en\-ter God's king\-dom.”
\end{para}

\begin{para}
    \textsuperscript{10:26}\thinspace{}They were e\-ven more astonished, and asked him, “Then who can be saved?”
\end{para}

\begin{para}
    \textsuperscript{10:27}\thinspace{}Looking at them, Jesus said, “With man it's im\-pos\-si\-ble, but not with God. Be\-cause all things are pos\-si\-ble with God.”
\end{para}

\begin{para}
    \textsuperscript{10:28}\thinspace{}Pe\-ter be\-gan to tell him, “We've left eve\-ry\-thing, and followed you.”
\end{para}

\begin{para}
    \textsuperscript{10:29}\thinspace{}Jesus said, “Any\-one who has left house or brothers or sisters or moth\-er or fa\-ther or chil\-dren or fields, for my sa\-ke, and for the gos\-pel,
    \textsuperscript{30}\thinspace{}will re\-ceive a hun\-dred\-fold now in this time, hous\-es and brothers and sisters and mothers and chil\-dren and fields, with persecutions, and e\-ter\-nal life in the age to come.
    \textsuperscript{31}\thinspace{}But man\-y who are first will be last, and the last first.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{10:32}\thinspace{}They were on the road, go\-ing up to Je\-ru\-sa\-lem, and Jesus was walk\-ing a\-head of them. He took the twelve a\-gain, and be\-gan to tell them what was go\-ing to hap\-pen to him,
    \textsuperscript{33}\thinspace{}“We're go\-ing up to Je\-ru\-sa\-lem, and the son of man will be delivered to the chief priests and the scribes. They'll con\-demn him to death, and de\-liv\-er him to the Gentiles.
    \textsuperscript{34}\thinspace{}They'll mock him, spit on him, scourge him, and kill him. He'll rise af\-ter three days.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{10:42}\thinspace{}Jesus called them o\-ver, and said to them, “You know that those who are con\-sid\-ered rulers of the Gentiles lord it o\-ver them, and their great ones ex\-er\-cise au\-thor\-i\-ty o\-ver them.
    \textsuperscript{43}\thinspace{}But it is not this way a\-mong you. In\-stead, who\-ev\-er wants to be\-come great a\-mong you must be your serv\-ant,
    \textsuperscript{44}\thinspace{}and who\-ev\-er wants to be first a\-mong you must be slave of all.
    \textsuperscript{45}\thinspace{}For e\-ven the son of man did\-n't come to be served, but to serve, and to give his life as a ran\-som for man\-y.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{10:46}\thinspace{}As he was leav\-ing Jericho with his disciples and a large crowd, a blind beg\-gar was sit\-ting by the road.
    \textsuperscript{47}\thinspace{}When he heard that it was Jesus of Na\-za\-reth, he be\-gan to cry out, “Jesus, son of David, have mer\-cy on me.”
\end{para}

\begin{para}
    \textsuperscript{10:48}\thinspace{}Man\-y sternly told him to be si\-lent, but he cried out e\-ven louder, “Son of David, have mer\-cy on me.”
\end{para}

\begin{para}
    \textsuperscript{10:49}\thinspace{}Jesus stopped and said, “Call him.” They called the blind man, say\-ing to him, “Take cour\-age. Get up. He's call\-ing for you.”
    \textsuperscript{50}\thinspace{}Throwing a\-side his cloak, he jumped up, and came to Jesus.
\end{para}

\begin{para}
    \textsuperscript{10:51}\thinspace{}Jesus asked him, “What do you want me to do for you?” The blind man said to him, “Rab\-bi, I want to re\-gain my sight.”
\end{para}

\begin{para}
    \textsuperscript{10:52}\thinspace{}Jesus said to him, “Go, your faith has made you well.” Im\-me\-di\-ate\-ly he regained his sight, and be\-gan fol\-low\-ing him on the road.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{11:1}\thinspace{}As they approached Je\-ru\-sa\-lem, near the Mount of Olives, he sent two of his disciples,
    \textsuperscript{2}\thinspace{}and told them, “Go in\-to the vil\-lage a\-head of you. As soon as you en\-ter it, you'll find a colt tied, on which no one has ev\-er sat. Un\-tie it, and bring it.
    \textsuperscript{3}\thinspace{}If any\-one asks you, ‘Why are you do\-ing this?’ say, ‘The Lord needs it and will send it back im\-me\-di\-ate\-ly.’”
\end{para}

\begin{para}
    \textsuperscript{11:4}\thinspace{}They went, found a colt out\-side in the street tied by a door, and un\-tied it.
    \textsuperscript{5}\thinspace{}Some of the bystanders asked them, “Why are you un\-ty\-ing the colt?”
    \textsuperscript{6}\thinspace{}They told them what Jesus had said, and they gave them per\-mis\-sion.
    \textsuperscript{7}\thinspace{}They brought the colt to Jesus, and put their coats on it. He sat on it,
    \textsuperscript{11}\thinspace{}and went in\-to Je\-ru\-sa\-lem, and in\-to the tem\-ple. Af\-ter looking a\-round at eve\-ry\-thing, he left for Bethany with the twelve, since it was al\-read\-y late.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{11:15}\thinspace{}They came to Je\-ru\-sa\-lem. He entered the tem\-ple, and be\-gan to drive out those buy\-ing and sell\-ing in the tem\-ple, and overturned the tables of the mon\-ey changers and the seats of those sell\-ing doves.
    \textsuperscript{16}\thinspace{}He would not al\-low any\-one to car\-ry mer\-chan\-dise through the tem\-ple.
    \textsuperscript{17}\thinspace{}He taught, and said to them, “Is\-n't it writ\-ten, ‘My house will be called a house of pray\-er for all the nations’? But you have made it ‘a den of robbers.’”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{11:27}\thinspace{}As he was walk\-ing in the tem\-ple, the chief priests, the scribes, and the elders came to him.
    \textsuperscript{28}\thinspace{}They asked him, “By what au\-thor\-i\-ty are you do\-ing these things?”
\end{para}

\begin{para}
    \textsuperscript{11:29}\thinspace{}Jesus said to them, “I will ask you one ques\-tion. An\-swer me, and I will tell you by what au\-thor\-i\-ty I do these things.
    \textsuperscript{30}\thinspace{}Was John's bap\-tism from heav\-en or from men?”
\end{para}

\begin{para}
    \textsuperscript{11:31}\thinspace{}They discussed it with one an\-oth\-er, say\-ing, “If we say, ‘From heav\-en,’ he will say, ‘Then why did\-n't you be\-lieve him?’
    \textsuperscript{32}\thinspace{}But should we say, ‘From men’?” They feared the peo\-ple, for eve\-ry\-one thought that John re\-al\-ly was a proph\-et.
\end{para}

\begin{para}
    \textsuperscript{11:33}\thinspace{}They answered Jesus, “We don't know.” Jesus said to them, “Nei\-ther will I tell you by what au\-thor\-i\-ty I do these things.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{12:1}\thinspace{}He be\-gan to speak to them in parables. “A man planted a vine\-yard, put a fence a\-round it, dug a pit for a wine\-press, and built a tow\-er. Then he leased it to tenants, and went on a jour\-ney.
    \textsuperscript{2}\thinspace{}At har\-vest time, he sent a serv\-ant to the tenants to col\-lect some of the fruit of the vine\-yard from them.
    \textsuperscript{3}\thinspace{}They took him, beat him, and sent him a\-way empty-handed.
    \textsuperscript{4}\thinspace{}A\-gain he sent them an\-oth\-er serv\-ant. They hit him on the head, and treated him shamefully.
    \textsuperscript{5}\thinspace{}He sent an\-oth\-er, and they killed that one. He sent man\-y others; they beat some, and killed others.
    \textsuperscript{6}\thinspace{}He had one more to send, a be\-lov\-ed son. Fi\-nal\-ly he sent him to them, say\-ing, ‘They will re\-spect my son.’”
\end{para}

\begin{para}
    \textsuperscript{12:7}\thinspace{}“But those tenants said to one an\-oth\-er, ‘This is the heir. Come, let's kill him, and the in\-her\-it\-ance will be ours.’
    \textsuperscript{8}\thinspace{}They took him, killed him, and threw him out of the vine\-yard.”
\end{para}

\begin{para}
    \textsuperscript{12:9}\thinspace{}“What will the vineyard's own\-er do? He will come and de\-stroy the tenants, and give the vine\-yard to others.
    \textsuperscript{10}\thinspace{}Have\-n't you read this scrip\-ture: ‘The stone that the builders rejected has be\-come the cor\-ner\-stone.’”
\end{para}

\begin{para}
    \textsuperscript{12:12}\thinspace{}They were seek\-ing to ar\-rest him, be\-cause they un\-der\-stood that he spoke the par\-a\-ble a\-gainst them, but they feared the crowd. So they left him, and went a\-way.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{12:13}\thinspace{}They sent some of the Pharisees and Herodians to him, so that they might trap him in his words.
    \textsuperscript{14}\thinspace{}When they came, they said to him, “Teach\-er, we know that you're truth\-ful. You don't care what any\-one thinks, be\-cause you're im\-par\-tial. You truthfully teach the way of God. Is it law\-ful to pay taxes to Caesar, or not?”
    \textsuperscript{15}\thinspace{}Know\-ing their hy\-poc\-ri\-sy, he said to them, “Why are you testing me? Bring me a de\-nar\-i\-us to look at.”
\end{para}

\begin{para}
    \textsuperscript{12:16}\thinspace{}They brought one. He asked them, “Whose im\-age and in\-scrip\-tion is this?” They answered, “Caesar's.”
    \textsuperscript{17}\thinspace{}Jesus said to them, “Give to Caesar the things that are Caesar's, and to God the things that are God's.” They mar\-veled at him.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{12:18}\thinspace{}Some Sadducees, who say there is no res\-ur\-rec\-tion, came to Jesus and asked him,
    \textsuperscript{19}\thinspace{}“Teach\-er, Mo\-ses wrote for us that if a man's broth\-er dies, and leaves be\-hind a wife but no child, that his broth\-er must mar\-ry the wid\-ow, and raise up chil\-dren for his broth\-er.
    \textsuperscript{20}\thinspace{}There were sev\-en brothers. The first one took a wife, and died with\-out chil\-dren.
    \textsuperscript{21}\thinspace{}The sec\-ond one mar\-ried her, and died with\-out chil\-dren.
    \textsuperscript{22}\thinspace{}None of the sev\-en left chil\-dren. Last of all, the wom\-an died too.
    \textsuperscript{23}\thinspace{}In the res\-ur\-rec\-tion, when they rise, whose wife will she be? For all sev\-en had mar\-ried her.”
\end{para}

\begin{para}
    \textsuperscript{12:24}\thinspace{}Jesus said to them, “Is\-n't this the rea\-son you are mis\-tak\-en: you don't know the scriptures or the pow\-er of God?
    \textsuperscript{25}\thinspace{}For when the dead rise, they nei\-ther mar\-ry nor are giv\-en in mar\-riage, but are like angels in heav\-en.”
\end{para}

\begin{para}
    \textsuperscript{12:26}\thinspace{}“As for the dead be\-ing raised, have\-n't you read in the book of Mo\-ses, in the pas\-sage a\-bout the burn\-ing bush, how God said to him, ‘I am the God of Abraham, and the God of Isaac, and the God of Jacob’?
    \textsuperscript{27}\thinspace{}He is\-n't God of the dead, but of the liv\-ing. You're great\-ly mis\-tak\-en.”
\end{para}

\begin{para}
    \textsuperscript{12:28}\thinspace{}One of the scribes came, and heard them ar\-gu\-ing. See\-ing that he answered them well, he asked him, “Which com\-mand\-ment is the most im\-por\-tant?”
\end{para}

\begin{para}
    \textsuperscript{12:29}\thinspace{}Jesus answered, “The most im\-por\-tant is,
    \textsuperscript{30}\thinspace{}‘Love the Lord your God with all your heart, soul, mind, and strength.’
    \textsuperscript{31}\thinspace{}The sec\-ond is this, ‘Love your neigh\-bor as your\-self.’ There is no com\-mand\-ment greater than these.”
\end{para}

\begin{para}
    \textsuperscript{12:32}\thinspace{}The scr\-ibe said to him, “You stat\-ed cor\-rect\-ly that
    \textsuperscript{33}\thinspace{}to love him with all the heart, un\-der\-stand\-ing, and strength, and to love one's neigh\-bor as him\-self, is more im\-por\-tant than all burnt offerings and sacrifices.”
\end{para}

\begin{para}
    \textsuperscript{12:34}\thinspace{}When Jesus saw that he answered wisely, he said to him, “You aren't far from God's king\-dom.” Af\-ter that, no one dared to ask him any more questions.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{12:38}\thinspace{}In his teach\-ing he said, “Be\-ware of the scribes, who like walk\-ing a\-round in long robes, be\-ing greeted in the marketplaces,
    \textsuperscript{39}\thinspace{}the best seats in the synagogues, and places of hon\-or at banquets.
    \textsuperscript{40}\thinspace{}They de\-vour widows' hous\-es, and of\-fer long prayers for show. They will re\-ceive greater con\-dem\-na\-tion.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{13:1}\thinspace{}As he came out of the tem\-ple, one of his disciples said to him, “Teach\-er, look at these won\-der\-ful stones and buildings!”
\end{para}

\begin{para}
    \textsuperscript{13:2}\thinspace{}Jesus said to him, “Do you see these great buildings? Not one stone will be left up\-on an\-oth\-er; all will be thrown down.”
\end{para}

\begin{para}
    \textsuperscript{13:3}\thinspace{}As he sat on the Mount of Olives op\-po\-site the tem\-ple, Pe\-ter, James, John and Andrew asked him pri\-vate\-ly,
    \textsuperscript{4}\thinspace{}“When will these things hap\-pen? What will be the sign that they are a\-bout to be ful\-filled?”
\end{para}

\begin{para}
    \textsuperscript{13:5}\thinspace{}Jesus said to them, “See that no one leads you a\-stray.
    \textsuperscript{6}\thinspace{}Man\-y will come in my name, say\-ing, ‘I am he,’ and will mis\-lead man\-y.
    \textsuperscript{12}\thinspace{}Broth\-er will be\-tray broth\-er to death, and a fa\-ther his child. Chil\-dren will rise a\-gainst parents, and have them put to death.
    \textsuperscript{13}\thinspace{}Eve\-ry\-one will hate you be\-cause of my name, but he who endures to the end will be saved.”
\end{para}

\begin{para}
    \textsuperscript{13:31}\thinspace{}“Heav\-en and earth will pass a\-way, but my words won't pass a\-way.
    \textsuperscript{32}\thinspace{}No one knows a\-bout that day or that hour, not e\-ven the angels in heav\-en, nor the son, but on\-ly the fa\-ther.
    \textsuperscript{33}\thinspace{}Be on guard. Be a\-lert. For you don't know when the time will come.”
\end{para}

\begin{para}
    \textsuperscript{13:34}\thinspace{}“It's like a man on a jour\-ney: when he leaves home, he puts his servants in charge, each with a task, and commands the door\-keep\-er to be a\-lert.
    \textsuperscript{35}\thinspace{}There\-fore be a\-lert. You don't know when the mas\-ter of the house will come. It could be in the eve\-ning, at mid\-night, when the roost\-er crows, or in the morn\-ing.
    \textsuperscript{36}\thinspace{}Oth\-er\-wise, he may come sud\-den\-ly and find you a\-sleep.
    \textsuperscript{37}\thinspace{}What I say to you, I say to eve\-ry\-one: Be a\-lert.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{14:1}\thinspace{}The Passover and the Feast of Un\-leav\-ened Bread were now two days a\-way. The chief priests and the scribes sought a way to ar\-rest him by stealth, and kill him.
    \textsuperscript{2}\thinspace{}But they said, “Not dur\-ing the feast, oth\-er\-wise the peo\-ple might ri\-ot.”
\end{para}

\begin{para}
    \textsuperscript{14:3}\thinspace{}While he was in Bethany in the house of Simon the lep\-er, and re\-clin\-ing at the ta\-ble, a wom\-an came with an al\-a\-bas\-ter jar of ver\-y ex\-pen\-sive per\-fume. She broke the jar, and poured it o\-ver his head.
    \textsuperscript{4}\thinspace{}Some expressed in\-dig\-na\-tion to one an\-oth\-er, say\-ing, “Why has this per\-fume been wast\-ed?
    \textsuperscript{5}\thinspace{}This per\-fume might have been sold for more than three hun\-dred de\-nar\-i\-i, and the mon\-ey giv\-en to the poor.” And they scolded her.
\end{para}

\begin{para}
    \textsuperscript{14:6}\thinspace{}Jesus said, “Leave her a\-lone. Why are you bothering her? She's done a beau\-ti\-ful thing for me.
    \textsuperscript{7}\thinspace{}You al\-ways have the poor with you. When\-ev\-er you want, you can do good for them, but you won't al\-ways have me.
    \textsuperscript{8}\thinspace{}She has done what she could. She has anointed my bod\-y be\-fore\-hand for the bur\-i\-al.
    \textsuperscript{9}\thinspace{}Tru\-ly I tell you, wher\-ev\-er the gos\-pel is preached in the whole world, what she has done will be told in mem\-o\-ry of her.”
\end{para}

\begin{para}
    \textsuperscript{14:10}\thinspace{}Then Ju\-das Iscariot, one of the twelve, went to the chief priests to be\-tray him to them.
    \textsuperscript{11}\thinspace{}They were glad when they heard this, and prom\-ised to give him mon\-ey. He be\-gan looking for an op\-por\-tu\-ni\-ty to be\-tray him.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{14:12}\thinspace{}On the first day of Un\-leav\-ened Bread, when the Passover lamb was be\-ing sacrificed, his disciples asked him, “Where do you want us to go to pre\-pare for you to eat the Passover?”
\end{para}

\begin{para}
    \textsuperscript{14:13}\thinspace{}He sent two of his disciples, and told them, “Go in\-to the cit\-y. A man car\-ry\-ing a pitch\-er of wa\-ter will meet you. Fol\-low him.
    \textsuperscript{14}\thinspace{}Wher\-ev\-er he enters, say to the own\-er of the house, ‘The teach\-er says, Where is my guest room, where I may eat the passover with my disciples?’
    \textsuperscript{15}\thinspace{}He will show you a large up\-per room furnished and read\-y. Pre\-pare for us there.”
    \textsuperscript{16}\thinspace{}The disciples went out, entered the cit\-y, and found it just as he had told them. They pre\-pared the Passover.
\end{para}

\begin{para}
    \textsuperscript{14:17}\thinspace{}When it was eve\-ning, he ar\-rived with the twelve.
    \textsuperscript{18}\thinspace{}While they were re\-clin\-ing at ta\-ble and eat\-ing, Jesus said, “Tru\-ly I tell you, one of you will be\-tray me, one who is eat\-ing with me.”
\end{para}

\begin{para}
    \textsuperscript{14:19}\thinspace{}They be\-gan to be grieved, and to say to him one by one, “Sure\-ly not I?”
\end{para}

\begin{para}
    \textsuperscript{14:20}\thinspace{}He said to them, “It's one of the twelve, one who is dip\-ping bread in\-to the bowl with me.
    \textsuperscript{21}\thinspace{}For the son of man goes as it is writ\-ten of him, but woe to that man by whom the son of man is betrayed! It would have been bet\-ter for that man if he had not been born.”
\end{para}

\begin{para}
    \textsuperscript{14:22}\thinspace{}As they were eat\-ing, he took bread, bless\-ed it, broke it, gave it to them, and said, “Take it. This is my bod\-y.”
    \textsuperscript{23}\thinspace{}He took a cup, and af\-ter giv\-ing thanks, he gave it to them, and they all drank from it.
    \textsuperscript{24}\thinspace{}He said to them, “This is my blood of the cov\-e\-nant, which is poured out for man\-y.
    \textsuperscript{25}\thinspace{}Tru\-ly I tell you, I won't drink of the fruit of the vine a\-gain un\-til that day when I drink it new in God's king\-dom.”
\end{para}

\begin{para}
    \textsuperscript{14:26}\thinspace{}Af\-ter sing\-ing a hymn, they went out to the Mount of Olives.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{14:27}\thinspace{}Jesus said to them, “You will all fall a\-way, be\-cause it is writ\-ten, ‘I will strike down the shep\-herd, and the sheep will be scattered.’
    \textsuperscript{28}\thinspace{}But af\-ter I'm raised up, I'll go a\-head of you to Gal\-i\-lee.”
\end{para}

\begin{para}
    \textsuperscript{14:29}\thinspace{}Pe\-ter said to him, “E\-ven if eve\-ry\-one falls a\-way, I won't.”
\end{para}

\begin{para}
    \textsuperscript{14:30}\thinspace{}Jesus said to him, “Tru\-ly I tell you, that this ver\-y night, be\-fore a roost\-er crows twice, you will de\-ny me three times.”
\end{para}

\begin{para}
    \textsuperscript{14:31}\thinspace{}But he kept insisting, “If I have to die with you, I won't de\-ny you.” They all said the same thing.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{14:32}\thinspace{}They went to a place called Gethsemane. He told his disciples, “Sit here while I pray.”
    \textsuperscript{33}\thinspace{}He took Pe\-ter, James and John with him, and be\-gan to be ver\-y dis\-tressed and troubled.
    \textsuperscript{34}\thinspace{}He told them, “My soul is deep\-ly grieved to the point of death. Stay here, and keep watch.”
\end{para}

\begin{para}
    \textsuperscript{14:35}\thinspace{}He went a lit\-tle far\-ther, fell to the ground, and prayed that the hour might pass from him.
    \textsuperscript{36}\thinspace{}He said, “Fa\-ther, eve\-ry\-thing is pos\-si\-ble for you. Re\-move this cup from me. Yet not what I will, but what you will.”
\end{para}

\begin{para}
    \textsuperscript{14:37}\thinspace{}He came, found them sleep\-ing, and said to Pe\-ter, “Simon, are you a\-sleep? Could\-n't you keep watch for one hour?
    \textsuperscript{38}\thinspace{}Keep watch and pray, that you won't come in\-to temp\-ta\-tion. The spir\-it is will\-ing, but the flesh is weak.”
\end{para}

\begin{para}
    \textsuperscript{14:39}\thinspace{}A\-gain he went a\-way, and prayed, say\-ing the same words.
    \textsuperscript{40}\thinspace{}A\-gain he came, and found them sleep\-ing, be\-cause their eyes were ver\-y heav\-y. They did\-n't know what to say to him.
\end{para}

\begin{para}
    \textsuperscript{14:41}\thinspace{}He came the third time, and said to them, “Are you still sleep\-ing and rest\-ing? It's e\-nough. The time has come. See, the son of man is be\-ing betrayed in\-to the hands of sinners.
    \textsuperscript{42}\thinspace{}Get up. Let's go. See, my be\-tray\-er is near.”
\end{para}

\begin{para}
    \textsuperscript{14:43}\thinspace{}While he was still speak\-ing, Ju\-das, one of the twelve, ar\-rived. With him was a crowd, armed with swords and clubs, who was sent from the chief priests, the scribes and the elders.
    \textsuperscript{44}\thinspace{}The be\-tray\-er had giv\-en them a sign, say\-ing, “The one I kiss is the one. Ar\-rest him, and lead him a\-way un\-der guard.”
    \textsuperscript{45}\thinspace{}Go\-ing di\-rect\-ly to Jesus, he said, “Rab\-bi!” and kissed him.
\end{para}

\begin{para}
    \textsuperscript{14:46}\thinspace{}They laid hands on him, and arrested him.
    \textsuperscript{47}\thinspace{}One of the bystanders drew his sword, and struck the high priest's serv\-ant, cut\-ting off his ear.
\end{para}

\begin{para}
    \textsuperscript{14:48}\thinspace{}Jesus said to them, “Have you come out with swords and clubs to ar\-rest me, as you would ar\-rest a rob\-ber?
    \textsuperscript{49}\thinspace{}I was with you eve\-ry day, teach\-ing in the tem\-ple, and you did\-n't ar\-rest me.”
\end{para}

\begin{para}
    \textsuperscript{14:50}\thinspace{}They all left him, and fled.
    \textsuperscript{51}\thinspace{}A young man was fol\-low\-ing him, wear\-ing on\-ly a lin\-en cloth a\-round his na\-ked bod\-y. They seized him,
    \textsuperscript{52}\thinspace{}but he left the lin\-en cloth, and fled na\-ked.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{14:53}\thinspace{}They led Jesus to the high priest. All the chief priests, the elders and the scribes as\-sem\-bled.
    \textsuperscript{54}\thinspace{}Pe\-ter followed him at a dis\-tance, right in\-to the high priest's court\-yard. He was sit\-ting with the guards, and warming him\-self by the fire.
\end{para}

\begin{para}
    \textsuperscript{14:55}\thinspace{}The chief priests and the whole coun\-cil sought tes\-ti\-mo\-ny a\-gainst Jesus to put him to death, but found none.
    \textsuperscript{56}\thinspace{}Man\-y gave false tes\-ti\-mo\-ny a\-gainst him, but their tes\-ti\-mo\-ny was in\-con\-sist\-ent.
    \textsuperscript{57}\thinspace{}Some stood up, and gave false tes\-ti\-mo\-ny a\-gainst him, say\-ing,
    \textsuperscript{58}\thinspace{}“We heard him say, ‘I'll de\-stroy this tem\-ple made with hands, and in three days I'll build an\-oth\-er made with\-out hands.’”
    \textsuperscript{59}\thinspace{}But e\-ven this tes\-ti\-mo\-ny was in\-con\-sist\-ent.
\end{para}

\begin{para}
    \textsuperscript{14:60}\thinspace{}The high priest stood up in front of them, and asked Jesus, “Don't you an\-swer? What is it that these men tes\-ti\-fy a\-gainst you?”
    \textsuperscript{61}\thinspace{}But he kept si\-lent, and did not an\-swer. A\-gain the high priest asked him, “Are you the Christ, the son of the Bless\-ed One?”
\end{para}

\begin{para}
    \textsuperscript{14:62}\thinspace{}Jesus said, “I am. You'll see the son of man sit\-ting at the right hand of pow\-er, and com\-ing with the clouds of heav\-en.”
\end{para}

\begin{para}
    \textsuperscript{14:63}\thinspace{}The high priest tore his clothes, and said, “Why do we need any more witnesses?
    \textsuperscript{64}\thinspace{}You've heard the blas\-phe\-my. What's your de\-ci\-sion?” They all condemned him as de\-serv\-ing death.
\end{para}

\begin{para}
    \textsuperscript{14:65}\thinspace{}Some be\-gan to spit on him, and to cov\-er his face and strike him, say\-ing, “Proph\-e\-sy.” The guards re\-ceived him with slaps in the face.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{14:66}\thinspace{}While Pe\-ter was in the court\-yard be\-low, one of the high priest's serv\-ant girls came.
    \textsuperscript{67}\thinspace{}See\-ing Pe\-ter, she looked at him, and said, “You al\-so were with Jesus the Nazarene.”
\end{para}

\begin{para}
    \textsuperscript{14:68}\thinspace{}But he de\-nied it, say\-ing, “I don't know or un\-der\-stand what you're talking a\-bout.” He went out to the en\-try\-way, and a roost\-er crowed.
\end{para}

\begin{para}
    \textsuperscript{14:69}\thinspace{}The serv\-ant girl saw him, and be\-gan a\-gain to tell the bystanders, “This man is one of them.”
\end{para}

\begin{para}
    \textsuperscript{14:70}\thinspace{}But a\-gain he de\-nied it. Af\-ter a lit\-tle while the bystanders a\-gain said to Pe\-ter, “You're cer\-tain\-ly one of them, since you're a Galilean too.”
    \textsuperscript{71}\thinspace{}But he be\-gan to curse and swear, “I don't know this man you're talking a\-bout.”
\end{para}

\begin{para}
    \textsuperscript{14:72}\thinspace{}Im\-me\-di\-ate\-ly a roost\-er crowed a sec\-ond time. Pe\-ter remembered when Jesus said to him, “Be\-fore a roost\-er crows twice, you'll de\-ny me three times”. He be\-gan to weep.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{15:1}\thinspace{}Ear\-ly in the morn\-ing, the chief priests, elders, scribes, and the whole coun\-cil held a con\-sul\-ta\-tion. They bound Jesus, led him a\-way, and delivered him to Pilate.
\end{para}

\begin{para}
    \textsuperscript{15:2}\thinspace{}Pilate asked him, “Are you the king of the Jews?” He answered, “You have said so.”
\end{para}

\begin{para}
    \textsuperscript{15:3}\thinspace{}The chief priests ac\-cused him of man\-y things.
    \textsuperscript{4}\thinspace{}Pilate a\-gain asked him, “Don't you an\-swer? Look how man\-y charges they bring a\-gainst you.”
    \textsuperscript{5}\thinspace{}But Jesus answered no fur\-ther, so Pilate was a\-mazed.
\end{para}

\begin{para}
    \textsuperscript{15:6}\thinspace{}Now at the feast he used to re\-lease for them any one pris\-on\-er whom they requested.
    \textsuperscript{7}\thinspace{}The man named Barabbas was in pris\-on with rebels who had com\-mit\-ted mur\-der dur\-ing the in\-sur\-rec\-tion.
    \textsuperscript{8}\thinspace{}The crowd be\-gan to ask Pilate to do for them as he u\-su\-al\-ly did.
\end{para}

\begin{para}
    \textsuperscript{15:9}\thinspace{}Pilate answered them, say\-ing, “Do you want me to re\-lease the king of the Jews for you?”
    \textsuperscript{10}\thinspace{}For he knew that the chief priests had hand\-ed him o\-ver be\-cause of en\-vy.
    \textsuperscript{11}\thinspace{}But the chief priests stirred up the crowd to have him re\-lease Barabbas in\-stead.
    \textsuperscript{12}\thinspace{}Pilate a\-gain asked them, “Then what should I do with the one you call the king of the Jews?”
\end{para}

\begin{para}
    \textsuperscript{15:13}\thinspace{}They shouted back, “Cru\-ci\-fy him!”
\end{para}

\begin{para}
    \textsuperscript{15:14}\thinspace{}Pilate asked them, “Why? What evil has he done?” But they shouted all the more, “Cru\-ci\-fy him!”
\end{para}

\begin{para}
    \textsuperscript{15:15}\thinspace{}Wishing to sat\-is\-fy the crowd, Pilate re\-leased Barabbas for them. Af\-ter hav\-ing Jesus flogged, he hand\-ed him o\-ver to be cru\-ci\-fied.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{15:21}\thinspace{}Simon of Cyrene was pass\-ing by, com\-ing from the coun\-try. They forced him to car\-ry Jesus's cross.
    \textsuperscript{22}\thinspace{}They brought Jesus to the place called Gol\-go\-tha, which means place of a skull.
    \textsuperscript{23}\thinspace{}They offered him wine mixed with myrrh, but he did not take it.
\end{para}

\begin{para}
    \textsuperscript{15:24}\thinspace{}They cru\-ci\-fied him, and di\-vid\-ed his clothes a\-mong them\-selves, cast\-ing lots to de\-cide what each should take.
    \textsuperscript{25}\thinspace{}It was nine in the morn\-ing when they cru\-ci\-fied him.
    \textsuperscript{26}\thinspace{}The in\-scrip\-tion of the charge a\-gainst him read, “The King of the Jews.”
    \textsuperscript{27}\thinspace{}They cru\-ci\-fied two robbers with him, one on his right and one on his left.
\end{para}

\begin{para}
    \textsuperscript{15:29}\thinspace{}Those who passed by de\-rid\-ed him, shak\-ing their heads, and say\-ing, “Ha! You who would de\-stroy the tem\-ple and re\-build it in three days,
    \textsuperscript{30}\thinspace{}save your\-self, and come down from the cross.”
\end{para}

\begin{para}
    \textsuperscript{15:31}\thinspace{}In the same way, the chief priests and the scribes mocked him to one an\-oth\-er say\-ing, “He saved others. He can't save him\-self.
    \textsuperscript{32}\thinspace{}Let this Christ, the King of Israel, come down now from the cross, so that we may see and be\-lieve.” Those who were cru\-ci\-fied with him al\-so insulted him.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{15:33}\thinspace{}At noon, dark\-ness fell o\-ver the whole land un\-til three in the af\-ter\-noon.
    \textsuperscript{34}\thinspace{}At three in the af\-ter\-noon Jesus cried with a loud voice, “My God, my God, why have you a\-ban\-doned me?”
\end{para}

\begin{para}
    \textsuperscript{15:37}\thinspace{}Jesus let out a loud cry, and breathed his last.
    \textsuperscript{38}\thinspace{}The cur\-tain of the tem\-ple was torn in two, from the top to the bot\-tom.
\end{para}

\begin{para}
    \textsuperscript{15:39}\thinspace{}When the cen\-tu\-ri\-on who stood fac\-ing him saw the way he breathed his last, he said, “Tru\-ly this man was the son of God.”
\end{para}

\begin{para}
    \textsuperscript{15:40}\thinspace{}There were al\-so wom\-en watching from a dis\-tance. A\-mong them were Mar\-y Mag\-da\-lene, Salome, and Mar\-y the moth\-er of James the young\-er and of Joses.
    \textsuperscript{41}\thinspace{}When he was in Gal\-i\-lee, they followed him and took care of him. Man\-y oth\-er wom\-en had come up with him to Je\-ru\-sa\-lem.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{15:42}\thinspace{}When eve\-ning had come, be\-cause it was the day be\-fore the Sabbath,
    \textsuperscript{43}\thinspace{}Jo\-seph of Arimathea came, a prom\-i\-nent mem\-ber of the coun\-cil, who was him\-self waiting for God's king\-dom. He bold\-ly went to Pilate, and asked for Jesus's bod\-y.
    \textsuperscript{44}\thinspace{}Pilate was sur\-prised that he was al\-read\-y dead.
    \textsuperscript{45}\thinspace{}He grant\-ed the bod\-y to Jo\-seph.
\end{para}

\begin{para}
    \textsuperscript{15:46}\thinspace{}Jo\-seph bought a lin\-en cloth, took him down, wrapped him in the lin\-en cloth, laid him in a tomb cut out of a rock, and rolled a stone a\-gainst the en\-trance of the tomb.
    \textsuperscript{47}\thinspace{}Mar\-y Mag\-da\-lene and Mar\-y the moth\-er of Joses were watching to see where he was laid.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{16:1}\thinspace{}When the Sabbath was past, Mar\-y Mag\-da\-lene, Salome, and Mar\-y the moth\-er of James bought spices, so that they could go and a\-noint him.
    \textsuperscript{2}\thinspace{}Ver\-y ear\-ly on the first day of the week, when the sun had ris\-en, they went to the tomb.
    \textsuperscript{3}\thinspace{}They were say\-ing to one an\-oth\-er, “Who will roll a\-way the stone from the en\-trance of the tomb for us?”
    \textsuperscript{4}\thinspace{}Looking up, they saw that the stone had al\-read\-y been rolled a\-way, al\-though it was ex\-treme\-ly large.
\end{para}

\begin{para}
    \textsuperscript{16:5}\thinspace{}Entering the tomb, they saw a young man sit\-ting at the right, wear\-ing a white robe, and they were alarmed.
    \textsuperscript{6}\thinspace{}He told them, “Don't be alarmed. You're looking for Jesus the Nazarene, who was cru\-ci\-fied. He has ris\-en. He's not here. See, here is the place where they laid him.
    \textsuperscript{7}\thinspace{}But go, tell his disciples and Pe\-ter, ‘He's go\-ing a\-head of you to Gal\-i\-lee. You'll see him there, just as he told you.’”
\end{para}

\begin{para}
    \textsuperscript{16:8}\thinspace{}They went out, and fled from the tomb, for trem\-bling and as\-ton\-ish\-ment had seized them. They said noth\-ing to any\-one, be\-cause they were a\-fraid.
\end{para}

\begin{para}
\end{para}


\end{document}

% Copyright 2021 Michael Hendricks
%
% Permission is granted to use, copy, modify, and distribute
% this work for any purpose with or without fee.
%
% This work is provided as-is without any express or implied
% warranty. In no event is the author liable for any damages
% arising from this work.
