%
% see tex.stackexchange.com/a/600/151276
\def\changemargin#1#2{\list{}{\rightmargin#2\leftmargin#1}\item[]}
\let\endchangemargin=\endlist

\documentclass[openany,12pt,english]{book}
\usepackage[utf8]{inputenc}
\usepackage{babel}
\usepackage{microtype}
\usepackage{csquotes}

\usepackage[
  paperwidth=8.5in,
  paperheight=11in,
  top=1in,
  outer=1in,
  bottom=1.25in,
  inner=1.5in
]{geometry}

\title{Mormon}
\author{}
\date{}

\newenvironment{para}{\par\pretolerance=100\tolerance=200\setlength{\emergencystretch}{0.6em}\relax}{\par}

\begin{document}

\maketitle

% Copyright page
{
\thispagestyle{empty}
\vspace*{\fill}
\begin{changemargin}{.8in}{.8in}
\begin{center}
\textcopyright{}~2021 Michael Hendricks

\vspace{0.5\baselineskip}

Permission is granted to use, copy, modify, and distribute
this work for any purpose with or without fee.

\vspace{0.5\baselineskip}

This work is provided as-is without any express or implied
warranty. In no event is the author liable for any damages
arising from this work.
\end{center}
\end{changemargin}
}
\pagenumbering{arabic}
\clearpage{}

% Table of Contents
\tableofcontents
\clearpage{}

\markboth{}{}
\twocolumn[\begin{changemargin}{.6in}{.6in}
  
  \begin{center}
    \itshape
    An a\-bridg\-ment of the Book of Mormon, with an in\-tro\-duc\-tion a\-bridged from Jo\-seph Smith---His\-to\-ry.
    Verse references such
    as~\textsuperscript{1:5}\thinspace{}indicate
    the original source of the text, but
    don't imply a full quotation.
  \end{center}
  
\end{changemargin}]{}


\vspace{0pt plus 8\baselineskip{}}\section*{Introduction}\addcontentsline{toc}{section}{Introduction}
\begin{para}
    \textsuperscript{1:1}\thinspace{}I, Jo\-seph Smith, write this his\-to\-ry to put all inquirers af\-ter truth in pos\-ses\-sion of the facts.
\end{para}

\begin{para}
    \textsuperscript{1:5}\thinspace{}There was in the place where I lived an un\-u\-su\-al ex\-cite\-ment on the sub\-ject of re\-li\-gion.
    \textsuperscript{6}\thinspace{}The cler\-gy pro\-mot\-ed this scene of re\-li\-gious feel\-ing in or\-der to have eve\-ry\-bod\-y con\-vert\-ed, let them join what sect they pleased. Yet when the converts be\-gan to file off, some to one par\-ty and some to an\-oth\-er, it was seen that the seemingly good feelings of both the priests and the converts were more pre\-tend\-ed than re\-al, lost in a strife of words: priest contending a\-gainst priest, and con\-vert a\-gainst con\-vert.
\end{para}

\begin{para}
    \textsuperscript{1:7}\thinspace{}I was at this time four\-teen years old.
    \textsuperscript{8}\thinspace{}My mind was called up to se\-ri\-ous re\-flec\-tion and great un\-eas\-i\-ness. My feelings were deep and of\-ten poign\-ant, but still I kept my\-self a\-loof from all the parties, though I attended their meetings as of\-ten as oc\-ca\-sion would per\-mit. So great were the con\-fu\-sion and strife a\-mong the dif\-fer\-ent denominations, that it was im\-pos\-si\-ble for a per\-son as young as I was to come to any cer\-tain con\-clu\-sion who was right and who was wrong.
\end{para}

\begin{para}
    \textsuperscript{1:10}\thinspace{}In the midst of this war of words and tu\-mult of opinions, I of\-ten said to my\-self, “What is to be done? If any of these parties is right, which is it, and how will I know it?”
\end{para}

\begin{para}
    \textsuperscript{1:11}\thinspace{}One day I was read\-ing James 1:5, which reads, “If any of you lacks wis\-dom, let him ask God, who gives gen\-er\-ous\-ly to all with\-out re\-proach, and it will be giv\-en to him.”
    \textsuperscript{12}\thinspace{}This pas\-sage entered with great force in\-to eve\-ry feel\-ing of my heart. I reflected on it a\-gain and a\-gain, know\-ing that if any per\-son needed wis\-dom from God, I did; for the dif\-fer\-ent sects un\-der\-stood the same passages of scrip\-ture so dif\-fer\-ent\-ly as to de\-stroy all con\-fi\-dence in set\-tling the ques\-tion by an ap\-peal to the Bible.
\end{para}

\begin{para}
    \textsuperscript{1:13}\thinspace{}At length I came to the de\-ter\-mi\-na\-tion to “ask God.”
    \textsuperscript{14}\thinspace{}So, in ac\-cord\-ance with my de\-ter\-mi\-na\-tion, I re\-tired to the woods to make the at\-tempt. It was the first time in my life that I had made the at\-tempt to pray vo\-cal\-ly.
\end{para}

\begin{para}
    \textsuperscript{1:15}\thinspace{}Hav\-ing looked a\-round me, and find\-ing my\-self a\-lone, I kneeled down and be\-gan to of\-fer up the desires of my heart to God. I had scarce\-ly done so when
    \textsuperscript{16}\thinspace{}I saw a pil\-lar of light o\-ver my head.
    \textsuperscript{17}\thinspace{}When the light rested up\-on me, I saw two personages. One of them said to me, “Jo\-seph, this is my be\-lov\-ed son. Hear him.”
\end{para}

\begin{para}
    \textsuperscript{1:18}\thinspace{}I asked the personages which of all the sects was right, and which I should join.
    \textsuperscript{19}\thinspace{}I was answered that I must join none of them.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{1:21}\thinspace{}A few days af\-ter I had this vi\-sion, I was with one of the preachers who was ver\-y ac\-tive in the re\-li\-gious ex\-cite\-ment. I gave him an ac\-count of the vi\-sion which I had had. He treated my com\-mu\-ni\-ca\-tion with con\-tempt, say\-ing that there were no such things as visions or revelations in these days.
\end{para}

\begin{para}
    \textsuperscript{1:22}\thinspace{}I soon found that my sto\-ry had ex\-cit\-ed a great deal of prej\-u\-dice a\-gainst me a\-mong the professors of re\-li\-gion. They ex\-cit\-ed the pub\-lic mind a\-gainst me and cre\-at\-ed a bit\-ter per\-se\-cu\-tion. This was com\-mon a\-mong all the sects,
    \textsuperscript{23}\thinspace{}and it of\-ten caused me great sor\-row.
\end{para}

\begin{para}
    \textsuperscript{1:24}\thinspace{}How\-ev\-er, it was nev\-er\-the\-less a fact that I had seen a vi\-sion, and all the per\-se\-cu\-tion un\-der heav\-en could not make it oth\-er\-wise.
    \textsuperscript{25}\thinspace{}I said in my heart, “Why per\-se\-cute me for tell\-ing the truth? For I have seen a vi\-sion. I know it, and I know that God knows it, and I can not de\-ny it.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{1:27}\thinspace{}I con\-tin\-ued to pur\-sue my com\-mon vocations in life, all the time be\-ing
    \textsuperscript{28}\thinspace{}per\-se\-cut\-ed by those who ought to have been my friends and to have treated me kind\-ly. If they sup\-posed me to be de\-lud\-ed, they ought to have endeavored to re\-claim me in a prop\-er and af\-fec\-tion\-ate man\-ner. I was left to mingling with all kinds of so\-ci\-e\-ty, and fre\-quent\-ly fell in\-to man\-y fool\-ish errors. In mak\-ing this con\-fes\-sion, no one need sup\-pose me guilt\-y of any great sins, but I was guilt\-y of lev\-i\-ty and some\-times as\-so\-ci\-at\-ed with jo\-vi\-al com\-pa\-ny.
\end{para}

\begin{para}
    \textsuperscript{1:29}\thinspace{}In con\-se\-quence of these things, I of\-ten felt condemned for my weaknesses and imperfections. One eve\-ning, I be\-took my\-self to pray\-er and sup\-pli\-ca\-tion to God for for\-give\-ness of my sins and follies.
    \textsuperscript{30}\thinspace{}While I was call\-ing up\-on God, I discovered a light appearing in my room, and im\-me\-di\-ate\-ly a per\-son\-age appeared.
\end{para}

\begin{para}
    \textsuperscript{1:33}\thinspace{}He called me by name, said that he was a mes\-sen\-ger from God, and that his name was Moroni.
    \textsuperscript{34}\thinspace{}He said that there was a book, writ\-ten up\-on gold plates, giv\-ing an ac\-count of the form\-er inhabitants of this con\-ti\-nent.
    \textsuperscript{42}\thinspace{}The vi\-sion was opened to my mind so that I could see the place where the plates were deposited.
    \textsuperscript{46}\thinspace{}He added a cau\-tion, tell\-ing me that Satan would try to tempt me to get the plates for the purposes of get\-ting rich. This he for\-bade me, say\-ing that I must on\-ly get the plates to glo\-ri\-fy God.
    \textsuperscript{53}\thinspace{}The mes\-sen\-ger in\-formed me that the time for bring\-ing forth the plates would ar\-rive in four years.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{1:54}\thinspace{}I went at the end of each year to the place where the plates were deposited. Each time I found the same mes\-sen\-ger there, and re\-ceived in\-struc\-tion from him.
\end{para}

\begin{para}
    \textsuperscript{1:59}\thinspace{}At length, the time ar\-rived for obtaining the plates. Hav\-ing gone as u\-su\-al at the end of an\-oth\-er year to the place where they were deposited, the same mes\-sen\-ger delivered them to me.
    \textsuperscript{67}\thinspace{}I com\-menced to trans\-late the Book of Mormon.
\end{para}

\begin{para}
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{1 Nephi}\addcontentsline{toc}{section}{1 Nephi}
\begin{para}
    \textsuperscript{1:1}\thinspace{}I, Nephi, am mak\-ing a rec\-ord of my proceedings dur\-ing my life\-time.
\end{para}

\begin{para}
    \textsuperscript{1:4}\thinspace{}In the first year of the reign of Zed\-e\-ki\-ah, king of Judah,
    \textsuperscript{18}\thinspace{}God showed my fa\-ther, Lehi, man\-y things a\-bout the de\-struc\-tion of Je\-ru\-sa\-lem. He went a\-mong the peo\-ple, and de\-clared to them what he had seen.
    \textsuperscript{20}\thinspace{}When the peo\-ple heard these things, they were an\-gry with him, and sought his life.
\end{para}

\begin{para}
    \textsuperscript{2:2}\thinspace{}God commanded my fa\-ther in a dream to take his fam\-i\-ly, and de\-part in\-to the wil\-der\-ness.
    \textsuperscript{4}\thinspace{}He left his house, land, gold, sil\-ver, and pre\-cious things, and took noth\-ing with him, ex\-cept his fam\-i\-ly, provisions, and tents, and de\-part\-ed in\-to the wil\-der\-ness.
\end{para}

\begin{para}
    \textsuperscript{2:11}\thinspace{}La\-man and Lem\-uel, my old\-er brothers, grumbled a\-gainst their fa\-ther be\-cause he had led them out of Je\-ru\-sa\-lem, leav\-ing their land, gold, sil\-ver, and pre\-cious things. They said he had done this be\-cause of the fool\-ish imaginations of his heart.
    \textsuperscript{13}\thinspace{}They al\-so did not be\-lieve that Je\-ru\-sa\-lem could be destroyed.
\end{para}

\begin{para}
    \textsuperscript{2:16}\thinspace{}I be\-lieved all that my fa\-ther had spo\-ken. There\-fore, I did not reb\-el a\-gainst him like my brothers.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{3:2}\thinspace{}My fa\-ther said to me,
    \textsuperscript{3}\thinspace{}“Laban has the rec\-ord of the Jews, and it's en\-graved on brass plates.
    \textsuperscript{4}\thinspace{}There\-fore, God has commanded me that you and your brothers should re\-turn to Je\-ru\-sa\-lem, go to Laban's house, seek the records, and bring them here in\-to the wil\-der\-ness.”
\end{para}

\begin{para}
    \textsuperscript{3:7}\thinspace{}I said to my fa\-ther, “I'll go and do what God has commanded, for I know that God gives no commandments un\-less he prepares a way to ac\-com\-plish them.”
\end{para}

\begin{para}
    \textsuperscript{3:19}\thinspace{}I said to my brothers, “It's wis\-dom in God that we should ob\-tain these records, that we may pre\-serve for our chil\-dren
    \textsuperscript{20}\thinspace{}the words spo\-ken by all the ho\-ly prophets.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{3:23}\thinspace{}We went up to the house of Laban.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{4:38}\thinspace{}My brothers and I took the brass plates, de\-part\-ed in\-to the wil\-der\-ness, and jour\-neyed to our father's tent.
\end{para}

\begin{para}
    \textsuperscript{5:1}\thinspace{}Af\-ter we returned to our fa\-ther, he was filled with joy, and al\-so my moth\-er was ver\-y glad.
    \textsuperscript{21}\thinspace{}We had obtained the records which God had commanded us, and found that they had great val\-ue to us, be\-cause we could pre\-serve the commandments of God for our chil\-dren.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{7:6}\thinspace{}As we jour\-neyed in the wil\-der\-ness, La\-man and Lem\-uel
    \textsuperscript{16}\thinspace{}were an\-gry with me, and tried to take a\-way my life,
    \textsuperscript{19}\thinspace{}but some of our com\-pa\-ny plead with them, and they stopped try\-ing to take a\-way my life.
    \textsuperscript{21}\thinspace{}I for\-gave them all that they had done, and we trav\-eled a\-gain on our jour\-ney.
\end{para}

\begin{para}
    \textsuperscript{17:3}\thinspace{}If the chil\-dren of men keep God's commandments, he provides means for them to ac\-com\-plish what he has commanded.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{17:5}\thinspace{}We came to the land which we called Boun\-ti\-ful, be\-cause it had much fruit and wild hon\-ey. God had pre\-pared all these things that we might not per\-ish.
    \textsuperscript{6}\thinspace{}We pitched our tents by the sea\-shore, and re\-joiced great\-ly.
\end{para}

\begin{para}
    \textsuperscript{17:7}\thinspace{}God said to me, Nephi,
    \textsuperscript{8}\thinspace{}“You shall con\-struct a ship af\-ter the man\-ner that I will show you, that I may car\-ry your peo\-ple a\-cross these waters.”
    \textsuperscript{15}\thinspace{}There\-fore, I la\-bored to keep the commandments of God.
\end{para}

\begin{para}
    \textsuperscript{17:17}\thinspace{}When my brothers saw that I was a\-bout to build a ship, they be\-gan to grum\-ble a\-gainst me, say\-ing, “Our broth\-er is a fool, for he thinks that he can build a ship, and cross these great waters.”
\end{para}

\begin{para}
    \textsuperscript{17:19}\thinspace{}I was deep\-ly grieved be\-cause of the hard\-ness of their hearts. When they saw this they were glad, say\-ing, “We knew that you could\-n't con\-struct a ship, and that you lacked judg\-ment.
    \textsuperscript{20}\thinspace{}You're like our fa\-ther, led a\-way by the fool\-ish imaginations of his heart. He has led us out of the land of Je\-ru\-sa\-lem, and we've wandered in the wil\-der\-ness all these years.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{18:3}\thinspace{}I, Nephi, of\-ten went up on the moun\-tain, and I of\-ten prayed to God.
\end{para}

\begin{para}
    \textsuperscript{18:4}\thinspace{}Af\-ter I had fin\-ished the ship, my brothers saw that it was good, and they humbled them\-selves a\-gain be\-fore God.
    \textsuperscript{6}\thinspace{}We pre\-pared man\-y provisions, went in\-to the ship with our wives and chil\-dren,
    \textsuperscript{8}\thinspace{}and put forth in\-to the sea.
\end{para}

\begin{para}
    \textsuperscript{18:23}\thinspace{}Af\-ter we had sailed for man\-y days we ar\-rived at the prom\-ised land.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{19:1}\thinspace{}I made plates of o\-re on which to en\-grave the rec\-ord of my peo\-ple,
    \textsuperscript{5}\thinspace{}to keep the more sa\-cred things for instructing my peo\-ple.
\end{para}

\begin{para}
    \textsuperscript{19:7}\thinspace{}The things which some men es\-teem to be of great worth, both to the bod\-y and soul, others set at naught.
    \textsuperscript{9}\thinspace{}The world, be\-cause of their in\-iq\-ui\-ty, will judge Jesus to be a thing of naught.
    \textsuperscript{18}\thinspace{}I have writ\-ten these things that per\-haps I might per\-suade my peo\-ple to re\-mem\-ber God.
\end{para}

\begin{para}
    \textsuperscript{22:23}\thinspace{}For the time soon co\-mes that all churches which are built up to make a prof\-it, and to get pow\-er, and to be\-come pop\-u\-lar in the eyes of the world, must be brought low in the dust.
\end{para}

\begin{para}
    \textsuperscript{22:31}\thinspace{}If you are o\-be\-di\-ent to the commandments, and en\-dure to the end, you will be saved at the last day.
\end{para}

\begin{para}
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{2 Nephi}\addcontentsline{toc}{section}{2 Nephi}
\begin{para}
    \textsuperscript{4:13}\thinspace{}Not man\-y days af\-ter Lehi's death, La\-man and Lem\-uel were an\-gry with me,
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{5:2}\thinspace{}so they sought to take a\-way my life.
\end{para}

\begin{para}
    \textsuperscript{5:5}\thinspace{}God warned me to de\-part from them, and flee in\-to the wil\-der\-ness a\-long with all those who would go with me.
    \textsuperscript{6}\thinspace{}There\-fore, I took my fam\-i\-ly, and Sam, my old\-er broth\-er, and Jacob and Jo\-seph, my young\-er brothers, and al\-so my sisters, and their families. All those who went with me were those who be\-lieved in the warnings and revelations of God.
\end{para}

\begin{para}
    \textsuperscript{5:7}\thinspace{}We jour\-neyed in the wil\-der\-ness for man\-y days, and pitched our tents.
    \textsuperscript{10}\thinspace{}We were care\-ful to keep the commandments of God in all things.
    \textsuperscript{11}\thinspace{}We prospered great\-ly, for we sowed seed, reaped in a\-bun\-dance, and raised flocks, and herds, and animals of eve\-ry kind.
\end{para}

\begin{para}
    \textsuperscript{5:14}\thinspace{}I made man\-y swords so that our enemies would not come up\-on us and de\-stroy us.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{6:1}\thinspace{}The words of Jacob, the broth\-er of Nephi:
\end{para}

\begin{para}
    \textsuperscript{9:27}\thinspace{}Woe to him who has all the commandments of God, and transgresses them, and wastes the days of his pro\-ba\-tion.
\end{para}

\begin{para}
    \textsuperscript{9:28}\thinspace{}Oh the van\-i\-ty, and fool\-ish\-ness of men! When they are learn\-ed they think they are wise, and they do not lis\-ten to God's coun\-sel, for they set it a\-side, sup\-pos\-ing they know of them\-selves, there\-fore, their wis\-dom is fool\-ish\-ness and it does not prof\-it them.
    \textsuperscript{29}\thinspace{}But to be learn\-ed is good if they lis\-ten to God's counsels.
\end{para}

\begin{para}
    \textsuperscript{9:30}\thinspace{}Woe to the rich, who des\-pise the poor, and per\-se\-cute the meek, and whose hearts are on their treasures; there\-fore, their treas\-ure is their God.
\end{para}

\begin{para}
    \textsuperscript{9:34}\thinspace{}Woe to the li\-ar,
    \textsuperscript{35}\thinspace{}to the mur\-der\-er who kills de\-lib\-er\-ate\-ly,
    \textsuperscript{36}\thinspace{}and to those who com\-mit for\-ni\-ca\-tion.
\end{para}

\begin{para}
    \textsuperscript{9:39}\thinspace{}Re\-mem\-ber, the mind set on the flesh is death, and the mind set on the spir\-it is e\-ter\-nal life.
    \textsuperscript{45}\thinspace{}Turn a\-way from your sins.
    \textsuperscript{46}\thinspace{}Pre\-pare your souls for the day of judg\-ment.
\end{para}

\begin{para}
    \textsuperscript{9:51}\thinspace{}Do not spend mon\-ey on what has no val\-ue, nor your la\-bor on what can\-not sat\-is\-fy.
\end{para}

\begin{para}
    \textsuperscript{10:23}\thinspace{}Cheer up your hearts, and re\-mem\-ber that you are free to act for your\-selves: to choose the way of ev\-er\-last\-ing death or the way of e\-ter\-nal life.
    \textsuperscript{24}\thinspace{}There\-fore, rec\-on\-cile your\-selves to the will of God, and not to the will of the flesh.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{11:1}\thinspace{}Jacob spoke man\-y more things to my peo\-ple, but I, Nephi, have on\-ly writ\-ten these.
\end{para}

\begin{para}
    \textsuperscript{25:23}\thinspace{}We la\-bor dil\-i\-gent\-ly to write, to per\-suade our chil\-dren, and al\-so our brothers, to be\-lieve in Christ, and to be rec\-on\-ciled to God.
    \textsuperscript{26}\thinspace{}We talk of Christ, we re\-joice in Christ, we preach of Christ, we proph\-e\-sy of Christ.
\end{para}

\begin{para}
    \textsuperscript{26:21}\thinspace{}There are man\-y churches built up which cause en\-vy, strife, and mal\-ice.
    \textsuperscript{22}\thinspace{}The dev\-il leads them by the neck with a flax\-en cord un\-til he binds them with his strong cords for\-ev\-er.
\end{para}

\begin{para}
    \textsuperscript{26:23}\thinspace{}God does not work in dark\-ness.
    \textsuperscript{25}\thinspace{}Does he say to any\-one, “De\-part from me?” No.
    \textsuperscript{26}\thinspace{}Has he commanded any\-one to de\-part from the hous\-es of wor\-ship? No.
\end{para}

\begin{para}
    \textsuperscript{26:29}\thinspace{}He commands that there shall be no priestcrafts, which are when men preach and set them\-selves up as a light to the world, to make a prof\-it, and get praise of the world.
    \textsuperscript{30}\thinspace{}God has for\-bid\-den this. There\-fore, he has commanded that all men should have love, and ex\-cept they have love they are noth\-ing.
\end{para}

\begin{para}
    \textsuperscript{26:32}\thinspace{}God has commanded men not to mur\-der, lie, steal, en\-vy, take his name in vain, have mal\-ice, con\-tend with one an\-oth\-er, or com\-mit for\-ni\-ca\-tion.
    \textsuperscript{33}\thinspace{}None of these iniquities come from God, for he does good a\-mong the chil\-dren of men.
\end{para}

\begin{para}
    \textsuperscript{28:3}\thinspace{}One church will say to the oth\-er, “Be\-hold, I, I am God's.” And the oth\-er churches will say, “I, I am God's.” Thus will eve\-ry one say that has built up churches, and not to God.
    \textsuperscript{4}\thinspace{}They will con\-tend with one an\-oth\-er, and their priests will con\-tend with one an\-oth\-er.
\end{para}

\begin{para}
    \textsuperscript{28:8}\thinspace{}Man\-y will say, “Eat, drink, and be mer\-ry, nev\-er\-the\-less fear God. He will jus\-ti\-fy in com\-mit\-ting a lit\-tle sin. Lie a lit\-tle, take ad\-van\-tage of one be\-cause of his words, dig a pit for your neigh\-bor. There is no harm in this. Do all these things, for to\-mor\-row we die. If we are guilt\-y, God will beat us with a few lashes, and at last we will be saved.”
\end{para}

\begin{para}
    \textsuperscript{28:12}\thinspace{}Be\-cause of pride their churches have be\-come corrupted.
    \textsuperscript{13}\thinspace{}They rob the poor be\-cause of their fi\-ne sanc\-tu\-ar\-ies, and their fi\-ne cloth\-ing. They per\-se\-cute the meek and the poor in heart, be\-cause in their pride they have be\-come ar\-ro\-gant.
\end{para}

\begin{para}
    \textsuperscript{28:16}\thinspace{}Woe to those who turn a\-side the just for noth\-ing, and re\-vile a\-gainst what is good, and say that it has no worth!
    \textsuperscript{29}\thinspace{}Woe to him who says, “We have re\-ceived the word of God, and we need no more of the word of God, for we have e\-nough!”
\end{para}

\begin{para}
    \textsuperscript{28:30}\thinspace{}God says, “To him who receives I will give more. From them who say, ‘We have e\-nough,’ will be tak\-en a\-way what\-ev\-er they have.”
\end{para}

\begin{para}
    \textsuperscript{29:3}\thinspace{}“Some will say, ‘We have a Bible, and there can\-not be any more Bible.’
    \textsuperscript{10}\thinspace{}Be\-cause you have a Bible you don't need to sup\-pose that it contains all my words.
    \textsuperscript{11}\thinspace{}For I com\-mand all men to write the words which I speak to them.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{30:17}\thinspace{}There is noth\-ing se\-cret that will not be revealed.
\end{para}

\begin{para}
    \textsuperscript{32:8}\thinspace{}Know that you must pray, for the evil spir\-it teaches a man that he must not pray.
\end{para}

\begin{para}
    \textsuperscript{33:10}\thinspace{}Lis\-ten to these words and be\-lieve in Christ, for they teach all men to do good.
\end{para}

\begin{para}
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Jacob}\addcontentsline{toc}{section}{Jacob}
\begin{para}
    \textsuperscript{1:2}\thinspace{}Nephi gave me, Jacob, a com\-mand\-ment to write a few of the most pre\-cious things.
    \textsuperscript{15}\thinspace{}The peo\-ple be\-gan to grow hard in their hearts, and to in\-dulge some\-what in wick\-ed\-ness,
    \textsuperscript{17}\thinspace{}there\-fore I gave them these words:
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{2:12}\thinspace{}Man\-y of you have be\-gun to search for gold, sil\-ver, and pre\-cious ores,
    \textsuperscript{13}\thinspace{}and prov\-i\-dence has smiled on you most pleas\-ing\-ly, that you have obtained man\-y rich\-es. Be\-cause some of you have obtained more a\-bun\-dant\-ly than others you are lifted up in the pride of your hearts be\-cause of the cost\-li\-ness of your ap\-par\-el, and per\-se\-cute others be\-cause you sup\-pose that you are bet\-ter than they. 
    \textsuperscript{14}\thinspace{}Do you sup\-pose that God justifies you in this thing? No, but he condemns you, and if you per\-sist his judgments will speed\-i\-ly come to you.
\end{para}

\begin{para}
    \textsuperscript{2:17}\thinspace{}Think of others like your\-selves, be fa\-mil\-iar with all, and free with your possessions, that they may be rich like you.
    \textsuperscript{18}\thinspace{}But be\-fore you seek for rich\-es, seek for the king\-dom of God.
    \textsuperscript{19}\thinspace{}Af\-ter you have obtained a hope in Christ you will ob\-tain rich\-es, if you seek them, and you will seek them with the in\-tent to do good: to clothe the na\-ked, to feed the hun\-gry, to lib\-er\-ate the cap\-tive, and to ad\-min\-is\-ter re\-lief to the sick and the afflicted.
    \textsuperscript{21}\thinspace{}Don't you sup\-pose that one be\-ing is as pre\-cious in God's sight as the oth\-er?
\end{para}

\begin{para}
    \textsuperscript{2:35}\thinspace{}You have bro\-ken your ten\-der wives' hearts, and lost your children's con\-fi\-dence, be\-cause of your bad examples be\-fore them. The sob\-bing of their hearts ascends up to God a\-gainst you.
\end{para}

\begin{para}
    \textsuperscript{4:8}\thinspace{}Don't des\-pise God's revelations.
    \textsuperscript{10}\thinspace{}Don't seek to coun\-sel God, but to take coun\-sel from him. For you know that he counsels in wis\-dom, jus\-tice, and great mer\-cy.
\end{para}

\begin{para}
    \textsuperscript{6:12}\thinspace{}Be wise; what more can I say?
\end{para}

\begin{para}
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Zeniff}\addcontentsline{toc}{section}{Zeniff}
\begin{para}
    \textsuperscript{0}\thinspace{}Tak\-en from chapters 11-23 of the Book of Mo\-si\-ah.
\end{para}

\begin{para}
    \textsuperscript{11:1}\thinspace{}Zeniff con\-ferred the king\-dom on No\-ah, one of his sons.
    \textsuperscript{2}\thinspace{}No\-ah did not keep God's commandments, but followed the desires of his own heart.
    \textsuperscript{3}\thinspace{}He im\-posed a tax of one fifth of all the peo\-ple pos\-sessed
    \textsuperscript{4}\thinspace{}to sup\-port him\-self, and his priests.
\end{para}

\begin{para}
    \textsuperscript{11:20}\thinspace{}There was a man named A\-bin\-a\-di who went a\-mong the peo\-ple say\-ing, “God says, ‘Woe to this peo\-ple, for I have seen their abominations, their wick\-ed\-ness, and their for\-ni\-ca\-tion.
    \textsuperscript{21}\thinspace{}Un\-less they re\-pent they will be brought in\-to bond\-age.’”
\end{para}

\begin{para}
    \textsuperscript{11:27}\thinspace{}When king No\-ah heard of Abinadi's words, he was an\-gry. He said, “Who is A\-bin\-a\-di that he should judge me and my peo\-ple, or who is God to bring such great af\-flic\-tion on my peo\-ple?
    \textsuperscript{28}\thinspace{}Bring A\-bin\-a\-di here that I may kill him.”
\end{para}

\begin{para}
    \textsuperscript{12:9}\thinspace{}They took him and car\-ried him bound be\-fore the king.
    \textsuperscript{29}\thinspace{}A\-bin\-a\-di said to the priests, “Why do you set your hearts on rich\-es? Why do you spend your strength with prostitutes?”
\end{para}

\begin{para}
    \textsuperscript{13:4}\thinspace{}“You're an\-gry with me be\-cause I've told you the truth.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{15:26}\thinspace{}But fear and trem\-ble be\-fore God, for he redeems none who have wil\-ful\-ly rebelled a\-gainst him, who have known his commandments, and re\-fused to keep them.”
\end{para}

\begin{para}
    \textsuperscript{17:1}\thinspace{}The king commanded the priests to take A\-bin\-a\-di and cause him to be put to death.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{17:2}\thinspace{}There was a young man a\-mong the priests whose name was Al\-ma, and he be\-lieved Abinadi's words, there\-fore he be\-gan to plead with the king to let A\-bin\-a\-di de\-part in peace.
    \textsuperscript{3}\thinspace{}But the king was more an\-gry and sent his servants af\-ter Al\-ma to kill him.
    \textsuperscript{4}\thinspace{}But he fled from them and hid. Be\-ing concealed for man\-y days, he wrote all the words which A\-bin\-a\-di had spo\-ken.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{17:7}\thinspace{}The king said, “A\-bin\-a\-di,
    \textsuperscript{8}\thinspace{}you will be put to death un\-less you take back all the evil you've spo\-ken a\-bout me and my peo\-ple.”
\end{para}

\begin{para}
    \textsuperscript{17:9}\thinspace{}A\-bin\-a\-di said to him, “I will not take back what I've spo\-ken to you, for it is true.
    \textsuperscript{10}\thinspace{}It will stand as a tes\-ti\-mo\-ny a\-gainst you at the last day.”
\end{para}

\begin{para}
    \textsuperscript{17:11}\thinspace{}King No\-ah was a\-bout to re\-lease him, for he feared that God's judgments would come up\-on him.
    \textsuperscript{12}\thinspace{}But the priests said, “He has re\-viled the king.” There\-fore the king was in\-cit\-ed to an\-ger a\-gainst him, and delivered him to be killed.
\end{para}

\begin{para}
    \textsuperscript{17:20}\thinspace{}A\-bin\-a\-di suffered death by fire be\-cause he would not de\-ny the commandments of God, hav\-ing sealed the truth of his words with his death.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{18:1}\thinspace{}Al\-ma, who had fled from king Noah's servants, be\-gan to teach Abinadi's words.
    \textsuperscript{3}\thinspace{}Man\-y be\-lieved his words.
\end{para}

\begin{para}
    \textsuperscript{18:31}\thinspace{}These things were done in the borders of the land, that the king might not know.
    \textsuperscript{32}\thinspace{}But they were made known to the king.
    \textsuperscript{33}\thinspace{}He said that Al\-ma was in\-cit\-ing the peo\-ple to re\-bel\-lion a\-gainst him, there\-fore he sent his ar\-my to de\-stroy them.
\end{para}

\begin{para}
    \textsuperscript{23:1}\thinspace{}Now Al\-ma, hav\-ing been warned by God, told his peo\-ple, there\-fore they gathered to\-geth\-er their flocks, and their grain, and
    \textsuperscript{3}\thinspace{}they fled eight days' jour\-ney in\-to the wil\-der\-ness.
\end{para}

\begin{para}
    \textsuperscript{23:4}\thinspace{}They came to a beau\-ti\-ful and pleas\-ant land,
    \textsuperscript{5}\thinspace{}and be\-gan to till the ground, and build buildings. They were in\-dus\-tri\-ous, and worked ver\-y hard.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{23:6}\thinspace{}The peo\-ple wanted Al\-ma to be their king, for his peo\-ple loved him.
    \textsuperscript{7}\thinspace{}But he said to them, “It's not ex\-pe\-di\-ent for us to have a king.
    \textsuperscript{9}\thinspace{}Re\-mem\-ber the in\-iq\-ui\-ty of king No\-ah and his priests.
    \textsuperscript{12}\thinspace{}King No\-ah oppressed you, and you were in bond\-age to him and his priests.
    \textsuperscript{13}\thinspace{}Be\-cause you have been delivered out of the hands of king No\-ah and his peo\-ple, I ask you not to trust any man to be a king o\-ver you.”
\end{para}

\begin{para}
    \textsuperscript{23:15}\thinspace{}Thus did Al\-ma teach his peo\-ple that eve\-ry man should love his neigh\-bor as him\-self, and that there should be no con\-ten\-tion a\-mong them.
\end{para}

\begin{para}
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Mosiah}\addcontentsline{toc}{section}{Mosiah}
\begin{para}
    \textsuperscript{1:2}\thinspace{}King Ben\-ja\-min had three sons, and he caused them to be taught so that they might be\-come men of un\-der\-stand\-ing.
    \textsuperscript{3}\thinspace{}He al\-so taught them a\-bout the records which were en\-graved on the brass plates, say\-ing: “My sons,
    \textsuperscript{4}\thinspace{}it was\-n't pos\-si\-ble for our fa\-ther, Lehi, to re\-mem\-ber all these things; but he could read these engravings, and teach them to his chil\-dren, that they could teach them to their chil\-dren, e\-ven down to the pres\-ent time.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{1:9}\thinspace{}King Ben\-ja\-min grew old, and saw that he must ver\-y soon go the way of all the earth.
    \textsuperscript{10}\thinspace{}There\-fore, he
    \textsuperscript{18}\thinspace{}proclaimed to all the peo\-ple to gath\-er them\-selves to\-geth\-er to hear his words.
\end{para}

\begin{para}
    \textsuperscript{2:5}\thinspace{}When they ar\-rived, they pitched their tents on eve\-ry side, eve\-ry fam\-i\-ly be\-ing sep\-a\-rate from one an\-oth\-er.
\end{para}

\begin{para}
    \textsuperscript{2:8}\thinspace{}He caused his words to be writ\-ten and sent to those who could not hear him, that they might al\-so re\-ceive his words.
    \textsuperscript{9}\thinspace{}These are the words which he spoke:
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{2:10}\thinspace{}I have not commanded you to come here to fear me, or to think that I am more than a mor\-tal man.
    \textsuperscript{11}\thinspace{}But I am like you, sub\-ject to all kinds of infirmities in bod\-y and mind; yet the peo\-ple have cho\-sen me, and God has al\-lowed me to be a king; to serve you with all the might, mind and strength which he has grant\-ed me.
\end{para}

\begin{para}
    \textsuperscript{2:12}\thinspace{}I have been al\-lowed to spend my days in your serv\-ice, and have not sought gold, or sil\-ver, or any kind of rich\-es from you.
    \textsuperscript{13}\thinspace{}Nei\-ther have I al\-lowed you to be con\-fined in dungeons, or to make slaves of one an\-oth\-er, or to mur\-der, or steal, or com\-mit a\-dul\-ter\-y.
    \textsuperscript{14}\thinspace{}And e\-ven I, my\-self, have la\-bored with my own hands that I might serve you, and that you should not be bur\-dened with taxes.
\end{para}

\begin{para}
    \textsuperscript{2:15}\thinspace{}Yet, I have not done this to boast.
    \textsuperscript{17}\thinspace{}I tell you so you may learn wis\-dom; that you may learn that when you serve your fel\-low beings you are on\-ly serv\-ing God.
\end{para}

\begin{para}
    \textsuperscript{2:18}\thinspace{}If I, whom you call your king, la\-bor to serve you, then should\-n't you la\-bor to serve one an\-oth\-er?
    \textsuperscript{19}\thinspace{}If I, whom you call your king, mer\-it any thanks from you, O how you should thank your heav\-en\-ly king!
\end{para}

\begin{para}
    \textsuperscript{2:41}\thinspace{}Con\-sid\-er the bless\-ed and hap\-py state of those who keep God's commandments. They are bless\-ed in all things, both tem\-po\-ral and spir\-it\-u\-al, and if they re\-main faith\-ful to the end they are re\-ceived in\-to heav\-en to dwell with God in never-ending hap\-pi\-ness.
\end{para}

\begin{para}
    \textsuperscript{3:19}\thinspace{}The nat\-u\-ral man is an en\-e\-my to God, and will be for\-ev\-er, un\-less he puts off the nat\-u\-ral man and becomes like a child: sub\-mis\-sive, meek, hum\-ble, pa\-tient, full of love, will\-ing to sub\-mit to all things, e\-ven as a child submits to his fa\-ther.
\end{para}

\begin{para}
    \textsuperscript{4:9}\thinspace{}Be\-lieve in God. Be\-lieve that he is, and that he cre\-at\-ed all things. Be\-lieve that he has all wis\-dom. Be\-lieve that man does not com\-pre\-hend all that God can com\-pre\-hend.
    \textsuperscript{10}\thinspace{}Be\-lieve that you must re\-pent of your sins and for\-sake them, and hum\-ble your\-selves be\-fore God. Ask in sin\-cer\-i\-ty of heart that he would for\-give you. And now, if you be\-lieve all these things see that you do them.
\end{para}

\begin{para}
    \textsuperscript{4:11}\thinspace{}Hum\-ble your\-selves, call\-ing on God dai\-ly.
    \textsuperscript{12}\thinspace{}If you do this you will al\-ways re\-joice,
    \textsuperscript{13}\thinspace{}and you will not have a mind to in\-jure one an\-oth\-er, but to live at peace, and to re\-pay eve\-ry man what he is due.
    \textsuperscript{14}\thinspace{}You will not al\-low your chil\-dren to go hun\-gry, or na\-ked, or to trans\-gress God's laws and fight and quar\-rel with one an\-oth\-er.
    \textsuperscript{15}\thinspace{}But you will teach them to walk in the ways of so\-ber truth, to love one an\-oth\-er, and to serve one an\-oth\-er.
    \textsuperscript{16}\thinspace{}You will help those who need your help. You will give of your possessions to him who stands in need. You won't al\-low the beg\-gar to pe\-ti\-tion you in vain, and turn him a\-way to per\-ish.
\end{para}

\begin{para}
    \textsuperscript{4:17}\thinspace{}Per\-haps you will say, “The man has brought his mis\-er\-y on him\-self, there\-fore I won't give him of my food, or of my possessions, for his punishments are just.”
    \textsuperscript{18}\thinspace{}Who\-ev\-er does this has great cause to re\-pent. Un\-less he repents he perishes for\-ev\-er, and has no part in the king\-dom of God.
    \textsuperscript{19}\thinspace{}For aren't we all beggars? Don't we all de\-pend on God for all our possessions, for food and cloth\-ing, for gold and sil\-ver, and for all our rich\-es of eve\-ry kind?
\end{para}

\begin{para}
    \textsuperscript{4:22}\thinspace{}If you judge the man who petitions you for your possessions, that he per\-ish not, and con\-demn him, how much more just will be your con\-dem\-na\-tion for with\-hold\-ing your possessions, which don't be\-long to you but to God.
    \textsuperscript{23}\thinspace{}Woe to that man, for his possessions will per\-ish with him. I speak to those who are rich in the things of this world.
\end{para}

\begin{para}
    \textsuperscript{4:24}\thinspace{}I say to the poor, you who on\-ly have e\-nough to re\-main from day to day; I mean all you who de\-ny the beg\-gar, be\-cause you have not. Say in your hearts: “I don't give be\-cause I don't have, but if I had I would give.”
    \textsuperscript{25}\thinspace{}If you say this in your hearts you re\-main guilt\-less. Oth\-er\-wise you are condemned, for you cov\-et what you have not re\-ceived.
\end{para}

\begin{para}
    \textsuperscript{4:26}\thinspace{}To re\-tain for\-give\-ness of your sins: Share your possessions with the poor, eve\-ry man ac\-cord\-ing to what he has, such as feed\-ing the hun\-gry, cloth\-ing the na\-ked, visiting the sick and help\-ing them, both spiritually and tem\-po\-ral\-ly, ac\-cord\-ing to their needs.
    \textsuperscript{27}\thinspace{}And see that all this is done in wis\-dom and or\-der; for it is not nec\-es\-sar\-y for a man to run faster than he has strength. It is ex\-pe\-di\-ent for him to be dil\-i\-gent, that he might win the prize. There\-fore, all things must be done in or\-der.
\end{para}

\begin{para}
    \textsuperscript{4:28}\thinspace{}Re\-mem\-ber that who\-ev\-er borrows from his neigh\-bor should re\-turn the thing, as he has a\-greed, or else you will sin; and per\-haps you will cause your neigh\-bor to sin al\-so.
\end{para}

\begin{para}
    \textsuperscript{4:30}\thinspace{}If you do not watch your\-selves, and your thoughts, and your words, and your deeds, and ob\-serve God's commandments, e\-ven un\-til the end of your lives, you must per\-ish.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{5:13}\thinspace{}How does a man know the mas\-ter whom he has not served, and who is a strang\-er to him, and is far from the thoughts and intents of his heart?
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{6:4}\thinspace{}Mo\-si\-ah be\-gan to reign in his father's place.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{25:1}\thinspace{}He gathered all the peo\-ple to\-geth\-er, and
    \textsuperscript{5}\thinspace{}read the records of Zeniff to them.
    \textsuperscript{6}\thinspace{}He al\-so read the ac\-count of Al\-ma and his peo\-ple.
    \textsuperscript{15}\thinspace{}Al\-ma al\-so spoke to them, preaching re\-pent\-ance and faith on God.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{28:10}\thinspace{}King Mo\-si\-ah had no one to con\-fer the king\-dom on, for none of his sons would ac\-cept the king\-dom.
    \textsuperscript{11}\thinspace{}There\-fore he took the records
    \textsuperscript{20}\thinspace{}and con\-ferred them on Al\-ma, who was the son of Al\-ma, and commanded him to keep and pre\-serve them, and al\-so to keep a rec\-ord of the peo\-ple, handing them down from one gen\-er\-a\-tion to the next.
\end{para}

\begin{para}
    \textsuperscript{29:4}\thinspace{}King Mo\-si\-ah wrote to the peo\-ple say\-ing,
    \textsuperscript{5}\thinspace{}“You want to have a king.
    \textsuperscript{6}\thinspace{}He to whom the king\-dom right\-ly belongs has de\-clined.
    \textsuperscript{7}\thinspace{}If I ap\-point an\-oth\-er in his place, I fear there would a\-rise contentions which would cause wars, which would de\-stroy the souls of man\-y peo\-ple.”
\end{para}

\begin{para}
    \textsuperscript{29:10}\thinspace{}“Let us be wise and an\-tic\-i\-pate these things, and do what will make for peace.
    \textsuperscript{11}\thinspace{}There\-fore I will be your king for the rest of my life. Nev\-er\-the\-less, let us ap\-point judges, to judge this peo\-ple ac\-cord\-ing to our law, and we will new\-ly ar\-range the af\-fairs of this peo\-ple.”
\end{para}

\begin{para}
    \textsuperscript{29:16}\thinspace{}“Be\-cause all men are not just, it is not ex\-pe\-di\-ent to have a king or kings to rule o\-ver you.
    \textsuperscript{17}\thinspace{}For be\-hold, how much in\-iq\-ui\-ty does one wick\-ed king cause to be com\-mit\-ted, and what great de\-struc\-tion!”
\end{para}

\begin{para}
    \textsuperscript{29:21}\thinspace{}“You can on\-ly de\-throne an in\-iq\-ui\-tous king through much con\-ten\-tion, and blood\-shed.
    \textsuperscript{22}\thinspace{}For he has his friends in in\-iq\-ui\-ty, and he keeps his guards a\-round him, and he tears up the laws of those who have reigned in right\-eous\-ness be\-fore him, and he tramples God's commandments un\-der his feet,
    \textsuperscript{23}\thinspace{}and he enacts laws ac\-cord\-ing to his own wick\-ed\-ness, and who\-ev\-er does not o\-bey his laws he causes to be destroyed.”
\end{para}

\begin{para}
    \textsuperscript{29:25}\thinspace{}“There\-fore, choose judges by the voice of the peo\-ple, that you may be judged ac\-cord\-ing to the laws which have been giv\-en by our fathers.
    \textsuperscript{26}\thinspace{}Ob\-serve this and make it your law: do your busi\-ness by the voice of the peo\-ple.”
\end{para}

\begin{para}
    \textsuperscript{29:37}\thinspace{}The peo\-ple were con\-vinced of the truth of his words,
    \textsuperscript{38}\thinspace{}there\-fore they relinquished their de\-sire for a king.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{29:40}\thinspace{}The peo\-ple loved Mo\-si\-ah, for they did not con\-sid\-er him a ty\-rant who was seek\-ing for that mon\-ey which corrupts the soul, for he had not exacted rich\-es from them, nei\-ther had he de\-light\-ed in shed\-ding blood, but he had established peace in the land, and had grant\-ed to his peo\-ple that they should be delivered from all kinds of bond\-age, there\-fore they esteemed him be\-yond meas\-ure.
\end{para}

\begin{para}
    \textsuperscript{29:41}\thinspace{}They ap\-point\-ed judges to judge them ac\-cord\-ing to the law.
    \textsuperscript{42}\thinspace{}Al\-ma was ap\-point\-ed to be the first chief judge.
\end{para}

\begin{para}
    \textsuperscript{29:47}\thinspace{}Thus ended the reign of the kings o\-ver the peo\-ple.
\end{para}

\begin{para}
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Alma}\addcontentsline{toc}{section}{Alma}
\begin{para}
    \textsuperscript{1:2}\thinspace{}In the first year of Alma's reign in the judg\-ment seat, there was a man named Nehor
    \textsuperscript{3}\thinspace{}who was preaching that eve\-ry priest and teach\-er should be\-come pop\-u\-lar, and should not la\-bor with his hands, but be supported by the peo\-ple.
\end{para}

\begin{para}
    \textsuperscript{1:7}\thinspace{}As he was preaching, Nehor be\-gan to con\-tend sharp\-ly with a man
    \textsuperscript{8}\thinspace{}named Gid\-e\-on.
    \textsuperscript{9}\thinspace{}Nehor drew his sword and killed Gid\-e\-on.
\end{para}

\begin{para}
    \textsuperscript{1:10}\thinspace{}Nehor was brought be\-fore Al\-ma to be judged.
    \textsuperscript{12}\thinspace{}Al\-ma said to him,
    \textsuperscript{14}\thinspace{}“You are condemned to die, ac\-cord\-ing to the law.”
\end{para}

\begin{para}
    \textsuperscript{1:15}\thinspace{}He suffered a shame\-ful death,
    \textsuperscript{16}\thinspace{}yet this did not stop priest\-craft from spreading through the land, for there were man\-y who loved the vain things of the world.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{1:21}\thinspace{}There was a strict law a\-mong the peo\-ple of the church that no one should per\-se\-cute those who did not be\-long to the church.
    \textsuperscript{27}\thinspace{}They imparted of their possessions to the poor, and the needy, and the sick, and the afflicted. They did not wear ex\-pen\-sive cloth\-ing, yet they were neat and come\-ly.
\end{para}

\begin{para}
    \textsuperscript{1:29}\thinspace{}Be\-cause of the stead\-i\-ness of the church they be\-came ver\-y rich, hav\-ing an a\-bun\-dance of eve\-ry\-thing they needed.
    \textsuperscript{30}\thinspace{}In their pros\-per\-ous circumstances, they did not send a\-way any who were na\-ked, or hun\-gry, or sick. They did not set their hearts on rich\-es, there\-fore they were gen\-er\-ous to all.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{4:6}\thinspace{}The peo\-ple of the church be\-gan to grow proud be\-cause of the great rich\-es they had obtained by their dil\-i\-gence. They were lifted up in their pride, for they be\-gan to wear ver\-y ex\-pen\-sive cloth\-ing.
    \textsuperscript{7}\thinspace{}Now this caused Al\-ma much af\-flic\-tion.
\end{para}

\begin{para}
    \textsuperscript{4:9}\thinspace{}There be\-gan to be great contentions a\-mong the peo\-ple of the church. There was en\-vy, strife, mal\-ice, per\-se\-cu\-tion, and pride.
\end{para}

\begin{para}
    \textsuperscript{4:15}\thinspace{}Al\-ma, see\-ing the afflictions, persecutions, and in\-e\-qual\-i\-ty, be\-gan to be ver\-y sorrowful.
    \textsuperscript{16}\thinspace{}He selected a wise man, and
    \textsuperscript{18}\thinspace{}delivered the judg\-ment seat to him.
    \textsuperscript{19}\thinspace{}He did this that he him\-self might go out a\-mong his peo\-ple to re\-mind them of their du\-ty, and to tear down all the pride, craft\-i\-ness, and con\-ten\-tion a\-mong them.
    \textsuperscript{20}\thinspace{}And thus, Al\-ma delivered up the judg\-ment seat, and con\-fined him\-self whol\-ly to the priest\-hood of God.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{5:2}\thinspace{}This is what Al\-ma said to the peo\-ple of the church:
\end{para}

\begin{para}
    \textsuperscript{5:14}\thinspace{}I ask you, have you ex\-pe\-ri\-enced a might\-y change in your hearts?
    \textsuperscript{15}\thinspace{}Do you look for\-ward and view this mor\-tal bod\-y raised in im\-mor\-tal\-i\-ty to stand be\-fore God to be judged ac\-cord\-ing to the deeds which have been done in the mor\-tal bod\-y?
\end{para}

\begin{para}
    \textsuperscript{5:17}\thinspace{}Do you im\-ag\-ine that you can lie to God in that day, and say, “God, our works have been right\-eous,” and that he will save you?
\end{para}

\begin{para}
    \textsuperscript{5:26}\thinspace{}If you have ex\-pe\-ri\-enced a change of heart, can you feel so now?
    \textsuperscript{27}\thinspace{}Have you walked, keep\-ing your\-selves blame\-less be\-fore God? Could you say, if you were called to die at this time, that you have been sufficiently hum\-ble?
    \textsuperscript{28}\thinspace{}Are you strip\-ped of pride? If not, you are not pre\-pared to meet God.
\end{para}

\begin{para}
    \textsuperscript{5:30}\thinspace{}Is there one a\-mong you who mocks his broth\-er, or heaps persecutions on him?
    \textsuperscript{31}\thinspace{}Woe to such a one, for he is not pre\-pared, and must re\-pent!
\end{para}

\begin{para}
    \textsuperscript{5:57}\thinspace{}All of you who de\-sire to fol\-low the voice of the good shep\-herd, come out from the wick\-ed, be sep\-a\-rate, and don't tou\-ch their un\-clean things.
\end{para}

\begin{para}
    \textsuperscript{7:2}\thinspace{}If I had\-n't giv\-en the judg\-ment seat to an\-oth\-er, to reign in my place, I could not have come at this time.
\end{para}

\begin{para}
    \textsuperscript{7:15}\thinspace{}Lay a\-side eve\-ry sin which so eas\-i\-ly entangles you, which binds you down to de\-struc\-tion. Show God that you are will\-ing to re\-pent of your sins.
    \textsuperscript{23}\thinspace{}Be hum\-ble, sub\-mis\-sive and gen\-tle; o\-pen to rea\-son; full of pa\-tience and long-suffering; ex\-er\-cis\-ing self-control in all things; dil\-i\-gent in keep\-ing the commandments of God at all times; asking for what\-ev\-er you need, both spir\-it\-u\-al and tem\-po\-ral; al\-ways thanking God for what\-ev\-er you re\-ceive.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{8:6}\thinspace{}Al\-ma came to a cit\-y called Am\-mon\-i\-hah.
    \textsuperscript{19}\thinspace{}As he entered the cit\-y he said to a man, “Will you give a hum\-ble serv\-ant of God some\-thing to eat?”
\end{para}

\begin{para}
    \textsuperscript{8:21}\thinspace{}The man was named Am\-u\-lek, and he brought bread and meat and placed them be\-fore Al\-ma.
    \textsuperscript{27}\thinspace{}Al\-ma stayed with Am\-u\-lek for man\-y days be\-fore he be\-gan to preach to the peo\-ple.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{8:30}\thinspace{}Al\-ma and Am\-u\-lek went out a\-mong the peo\-ple, to de\-clare God's words.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{12:9}\thinspace{}Al\-ma be\-gan to ex\-plain things to them say\-ing, “God on\-ly imparts to the chil\-dren of men ac\-cord\-ing to the at\-ten\-tion and dil\-i\-gence which they give him.
    \textsuperscript{10}\thinspace{}There\-fore, he who hardens his heart receives the less\-er por\-tion of the word, and he who will not hard\-en his heart receives the greater por\-tion of the word.”
\end{para}

\begin{para}
    \textsuperscript{12:12}\thinspace{}“All men must die, and be brought be\-fore God to be judged ac\-cord\-ing to our works.
    \textsuperscript{14}\thinspace{}For our words, our works, and our thoughts will con\-demn us.”
\end{para}

\begin{para}
    \textsuperscript{12:24}\thinspace{}“We see that death co\-mes to man\-kind. Nev\-er\-the\-less there was a space grant\-ed to man in which he might re\-pent; there\-fore this life be\-came a probationary state; a time to pre\-pare to meet God.
    \textsuperscript{37}\thinspace{}Be\-cause we know these things, let us re\-pent, and not hard\-en our hearts.”
\end{para}

\begin{para}
    \textsuperscript{13:27}\thinspace{}“I wish, from the in\-ner\-most part of my heart, that you would lis\-ten to my words, and cast off your sins, and not pro\-cras\-ti\-nate the day of your re\-pent\-ance,
    \textsuperscript{28}\thinspace{}but that you would hum\-ble your\-selves be\-fore God, and watch and pray con\-tin\-u\-al\-ly, that you may not be tempted a\-bove what you can bear, be\-com\-ing hum\-ble, meek, sub\-mis\-sive, pa\-tient, full of love and all long-suffering;
    \textsuperscript{29}\thinspace{}hav\-ing faith in God; hav\-ing hope that you will re\-ceive e\-ter\-nal life; al\-ways hav\-ing the love of God in your hearts.”
\end{para}

\begin{para}
    \textsuperscript{13:31}\thinspace{}Al\-ma spoke man\-y more words to the peo\-ple which are not writ\-ten in this book.
\end{para}

\begin{para}
    \textsuperscript{14:1}\thinspace{}Af\-ter he had spo\-ken to the peo\-ple, man\-y of them be\-lieved his words, and be\-gan to re\-pent, and to search the scriptures.
\end{para}

\begin{para}
    \textsuperscript{15:16}\thinspace{}Am\-u\-lek had left all his gold, sil\-ver, and pre\-cious things, be\-ing rejected by those who were once his friends and al\-so by his fa\-ther and his relatives.
    \textsuperscript{18}\thinspace{}Al\-ma, see\-ing all these things, took Am\-u\-lek to his own house, and ministered to him in his tribulations.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{29:2}\thinspace{}I, Al\-ma, would de\-clare re\-pent\-ance to eve\-ry\-one, that they should re\-pent and come to God, that there might be less sor\-row on the earth.
    \textsuperscript{3}\thinspace{}But I sin in my wish, for I should be con\-tent with what God has al\-lot\-ted me.
    \textsuperscript{8}\thinspace{}For God grants to all nations, of their own na\-tion and lan\-guage, peo\-ple to teach what he sees fit that they should have.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{30:6}\thinspace{}A man be\-gan to preach a\-gainst the com\-ing of Christ.
    \textsuperscript{7}\thinspace{}Now there was no law a\-gainst a man's be\-lief.
    \textsuperscript{9}\thinspace{}If a man de\-sired to serve God, it was his priv\-i\-lege, but if he did not be\-lieve in him there was no law to pun\-ish him.
\end{para}

\begin{para}
    \textsuperscript{30:12}\thinspace{}This man, named Kor\-i\-hor, preached say\-ing,
    \textsuperscript{14}\thinspace{}“These things which you call prophecies are fool\-ish traditions of your fathers.
    \textsuperscript{15}\thinspace{}You can't know of things you don't see.
    \textsuperscript{16}\thinspace{}You say that you see a re\-mis\-sion of your sins, but it is the ef\-fect of a fren\-zied mind. This de\-range\-ment of your minds co\-mes be\-cause of the traditions of your fathers, which lead you a\-way to be\-lieve in things which are not so.”
    \textsuperscript{17}\thinspace{}And he told them that what\-ev\-er a man did was no crime.
\end{para}

\begin{para}
    \textsuperscript{30:31}\thinspace{}He re\-viled a\-gainst Al\-ma and the priests, ac\-cus\-ing them of lead\-ing a\-way the peo\-ple af\-ter the sil\-ly traditions of their fathers, to glut on the labors of the peo\-ple.
\end{para}

\begin{para}
    \textsuperscript{30:32}\thinspace{}Al\-ma said to him, “You know that we don't glut our\-selves on the labors of this peo\-ple. I've la\-bored with my own hands for my sup\-port.
    \textsuperscript{33}\thinspace{}Not\-with\-stand\-ing the man\-y labors I've performed in the church, I've nev\-er re\-ceived as much as one pen\-ny for my la\-bor; nei\-ther have any of my brothers.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{31:1}\thinspace{}Al\-ma re\-ceived news that one Zoram was lead\-ing the peo\-ple to bow down to idols.
    \textsuperscript{5}\thinspace{}As preaching the word had a great ten\-den\-cy to lead the peo\-ple to do what was just --- it had had a more pow\-er\-ful ef\-fect on the minds of the peo\-ple than the sword, or any\-thing else which had happened to them --- there\-fore Al\-ma thought it was ex\-pe\-di\-ent to try the pow\-er of the word of God.
    \textsuperscript{7}\thinspace{}There\-fore he went a\-mong the Zoramites to preach the word to them.
\end{para}

\begin{para}
    \textsuperscript{32:2}\thinspace{}Af\-ter much la\-bor, they be\-gan to have suc\-cess a\-mong the poor, for they were cast out of the synagogues be\-cause of their coarse cloth\-ing.
\end{para}

\begin{para}
    \textsuperscript{32:5}\thinspace{}One of them said, “They cast us out of our synagogues and we have no place to wor\-ship God. What shall we do?”
\end{para}

\begin{para}
    \textsuperscript{32:7}\thinspace{}Al\-ma said to them,
    \textsuperscript{10}\thinspace{}“Do you sup\-pose that you can on\-ly wor\-ship God in your synagogues?
    \textsuperscript{11}\thinspace{}Do you sup\-pose that you must on\-ly wor\-ship God once per week?”
\end{para}

\begin{para}
    \textsuperscript{32:22}\thinspace{}“God is mer\-ci\-ful to all who be\-lieve his word.
    \textsuperscript{23}\thinspace{}He imparts his word to men, and wom\-en. This is not all, lit\-tle chil\-dren of\-ten have words giv\-en to them which con\-found the wise and the learn\-ed.”
\end{para}

\begin{para}
    \textsuperscript{32:28}\thinspace{}“We will com\-pare the word to a seed. If you make room to plant a seed in your heart, you will be\-gin to say with\-in your\-selves, ‘It's a good seed, for it begins to en\-light\-en my un\-der\-stand\-ing.’
    \textsuperscript{33}\thinspace{}Be\-cause you have tried the ex\-per\-i\-ment, and planted the seed,
    \textsuperscript{34}\thinspace{}you know that the word has sprouted up, that your un\-der\-stand\-ing begins to be enlightened, and your mind begins to ex\-pand.”
    \textsuperscript{37}\thinspace{}As the tree begins to grow, you will say: “Let's nour\-ish it care\-ful\-ly, that it may bear fruit.
    \textsuperscript{38}\thinspace{}But if you ne\-glect the tree, and don't think a\-bout its nour\-ish\-ment, it with\-ers a\-way.
    \textsuperscript{39}\thinspace{}Now, this is\-n't be\-cause the seed was not good, or be\-cause its fruit would not be de\-sir\-a\-ble, but it's be\-cause you won't nour\-ish the tree.”
\end{para}

\begin{para}
    \textsuperscript{33:2}\thinspace{}“You've said that you could\-n't wor\-ship God be\-cause you're cast out of your synagogues, but you're great\-ly mis\-tak\-en.
    \textsuperscript{3}\thinspace{}Do you re\-mem\-ber what Zenos, the proph\-et of old, said a\-bout pray\-er?
    \textsuperscript{4}\thinspace{}He said, ‘You are mer\-ci\-ful, O God, for you have heard my pray\-er, e\-ven when I was in the wil\-der\-ness,
    \textsuperscript{5}\thinspace{}and when I cried to you in my field.
    \textsuperscript{6}\thinspace{}And a\-gain, when I turned to my house, you heard my pray\-er.
    \textsuperscript{8}\thinspace{}You are mer\-ci\-ful to your chil\-dren when they cry to you, to be heard by you and not by men, and you will hear them.’”
\end{para}

\begin{para}
    \textsuperscript{34:1}\thinspace{}Af\-ter Al\-ma had spo\-ken, Am\-u\-lek taught them, say\-ing,
    \textsuperscript{20}\thinspace{}“Cry to God when you're in your fields, o\-ver all your flocks.
    \textsuperscript{21}\thinspace{}Cry to him in your hous\-es, o\-ver all your house\-hold, both morn\-ing, mid\-day, and eve\-ning.
    \textsuperscript{24}\thinspace{}Cry to him o\-ver the crops of your fields, that you may pros\-per in them.
    \textsuperscript{25}\thinspace{}Cry o\-ver the flocks of your fields, that they may in\-crease.
    \textsuperscript{26}\thinspace{}You must pour out your souls in your se\-cret places, and in your wil\-der\-ness.
    \textsuperscript{28}\thinspace{}Don't sup\-pose that this is all; if you turn a\-way the needy, and the na\-ked, and don't vis\-it the sick and afflicted, your pray\-er is emp\-ty, and brings you noth\-ing, and you are hypocrites.”
\end{para}

\begin{para}
    \textsuperscript{34:32}\thinspace{}“This life is the time for men to la\-bor.
    \textsuperscript{33}\thinspace{}Don't pro\-cras\-ti\-nate the day of your re\-pent\-ance.
    \textsuperscript{40}\thinspace{}Have pa\-tience, and bear all kinds of af\-flic\-tion. Don't re\-vile a\-gainst those who cast you out be\-cause of your great pov\-er\-ty, lest you be\-come sinners like them.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{35:16}\thinspace{}Al\-ma gathered his sons to\-geth\-er:
\end{para}

\begin{para}
    \textsuperscript{37:1}\thinspace{}My son Helaman, take the records which have been entrusted to me,
    \textsuperscript{2}\thinspace{}and keep a rec\-ord of this peo\-ple.
    \textsuperscript{6}\thinspace{}You may sup\-pose that this is fool\-ish\-ness, but by small and sim\-ple things are great things brought to pass.
    \textsuperscript{8}\thinspace{}These things have en\-larged the mem\-o\-ry of this peo\-ple, and con\-vinced man\-y of the er\-ror of their ways.
\end{para}

\begin{para}
    \textsuperscript{37:35}\thinspace{}In your youth, learn wis\-dom and to keep the commandments of God.
    \textsuperscript{47}\thinspace{}Be so\-ber. My son, fare\-well.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{38:1}\thinspace{}My son Shiblon,
    \textsuperscript{10}\thinspace{}be dil\-i\-gent and tem\-per\-ate in all things.
    \textsuperscript{11}\thinspace{}Don't be lifted up in pride. Don't boast in your own wis\-dom, or in your much strength.
    \textsuperscript{12}\thinspace{}Use bold\-ness, but not overbearance. Bri\-dle all your passions, that you may be filled with love. Re\-frain from i\-dle\-ness.
\end{para}

\begin{para}
    \textsuperscript{38:14}\thinspace{}Don't say, “God, I thank you that we are bet\-ter than our brothers.” Rath\-er say, “God, for\-give my unworthiness, and re\-mem\-ber my brothers in mer\-cy.”
\end{para}

\begin{para}
    \textsuperscript{38:15}\thinspace{}Be so\-ber. My son, fare\-well.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{39:1}\thinspace{}My son Corianton,
    \textsuperscript{2}\thinspace{}this is what I have a\-gainst you: you went on boasting in your strength and your wis\-dom,
    \textsuperscript{3}\thinspace{}and you went af\-ter a pros\-ti\-tute.
    \textsuperscript{5}\thinspace{}Don't you know that these things are an a\-bom\-i\-na\-tion to God?
\end{para}

\begin{para}
    \textsuperscript{39:9}\thinspace{}Re\-pent and for\-sake your sins, and don't go af\-ter the lusts of your eyes any\-more.
    \textsuperscript{11}\thinspace{}Don't al\-low your\-self to be led a\-way by any vain or fool\-ish thing.
    \textsuperscript{14}\thinspace{}Don't seek af\-ter rich\-es or the vain things of this world, for you can\-not car\-ry them with you.
\end{para}

\begin{para}
    \textsuperscript{41:14}\thinspace{}Be mer\-ci\-ful to your brothers. Deal just\-ly, judge right\-eous\-ly, and do good con\-tin\-u\-al\-ly.
    \textsuperscript{15}\thinspace{}For what you send out will re\-turn to you a\-gain.
\end{para}

\begin{para}
    \textsuperscript{42:27}\thinspace{}Who\-ev\-er will come may par\-take of the waters of life free\-ly, and who\-ev\-er will not come is not com\-pelled to come.
\end{para}

\begin{para}
    \textsuperscript{42:31}\thinspace{}My son, de\-clare the word with truth and so\-ber\-ness.
\end{para}

\begin{para}
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Helaman}\addcontentsline{toc}{section}{Helaman}
\begin{para}
    \textsuperscript{3:37}\thinspace{}In the fifty-third year of the reign of the judges, Nephi, the son of Helaman, the son of Helaman, be\-gan to reign in the judg\-ment seat.
\end{para}

\begin{para}
    \textsuperscript{4:18}\thinspace{}In the sixty-second year of the reign of the judges,
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{5:1}\thinspace{}Nephi delivered the judg\-ment seat to Cezoram,
    \textsuperscript{4}\thinspace{}for Nephi was wea\-ry be\-cause of the in\-iq\-ui\-ty of the peo\-ple. He gave up the judg\-ment seat to preach the word of God for the rest of his life, and his broth\-er Lehi al\-so.
    \textsuperscript{5}\thinspace{}For they remembered what their fa\-ther Helaman had said to them:
\end{para}

\begin{para}
    \textsuperscript{5:6}\thinspace{}“My sons, de\-clare these words to the peo\-ple:
    \textsuperscript{12}\thinspace{}Re\-mem\-ber that it is up\-on the rock of Christ that you must build your founda\-tion; that when the dev\-il sends forth his might\-y winds, when all his hail and his might\-y storm beats up\-on you, it will have no pow\-er to dra\-g you down to the gulf of mis\-er\-y and end\-less woe, be\-cause of the rock on which you are built.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{6:15}\thinspace{}In the sixty-sixth year of the reign of the judges, Cezoram was murdered. His son, who had been ap\-point\-ed by the peo\-ple in his place, was al\-so murdered.
\end{para}

\begin{para}
    \textsuperscript{6:16}\thinspace{}The peo\-ple grew ver\-y wick\-ed.
    \textsuperscript{17}\thinspace{}For God had bless\-ed them so long with the rich\-es of the world that they be\-gan to set their hearts on their rich\-es. They be\-gan to seek for prof\-it that they might be lifted up one a\-bove an\-oth\-er. There\-fore they be\-gan to mur\-der, rob, and plun\-der, to get mon\-ey.
\end{para}

\begin{para}
    \textsuperscript{6:18}\thinspace{}Those murderers and plunderers were a band formed by Gadianton.
    \textsuperscript{22}\thinspace{}They had signs, and se\-cret words to rec\-og\-nize a broth\-er who had entered in\-to the cov\-e\-nant, that what\-ev\-er wick\-ed\-ness he did he should not be in\-jured by those who belonged to his band.
    \textsuperscript{23}\thinspace{}Thus they might mur\-der, plun\-der, steal, and com\-mit for\-ni\-ca\-tion and all kinds of wick\-ed\-ness, con\-tra\-ry to the laws of their coun\-try and the laws of God.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{7:4}\thinspace{}Nephi saw the peo\-ple in a state of aw\-ful wick\-ed\-ness, and the Gadianton robbers fill\-ing the judg\-ment seats, hav\-ing usurped the au\-thor\-i\-ty of the land;
    \textsuperscript{5}\thinspace{}let\-ting the guilt\-y go un\-pun\-ished be\-cause of their mon\-ey; be\-ing held in of\-fice to get mon\-ey and glo\-ry of the world, that they might more eas\-i\-ly com\-mit a\-dul\-ter\-y, steal, and kill.
\end{para}

\begin{para}
    \textsuperscript{7:6}\thinspace{}This great in\-iq\-ui\-ty had come up\-on the peo\-ple in on\-ly a few years.
\end{para}

\begin{para}
    \textsuperscript{7:13}\thinspace{}Nephi said to the peo\-ple,
    \textsuperscript{20}\thinspace{}“How could you have for\-got\-ten God?
    \textsuperscript{21}\thinspace{}Be\-hold, it is to get mon\-ey, and to be praised of men.
    \textsuperscript{26}\thinspace{}Woe will come to you be\-cause of that pride which you have al\-lowed to en\-ter your hearts, which has lifted you up be\-cause of your ver\-y great rich\-es!”
\end{para}

\begin{para}
    \textsuperscript{8:25}\thinspace{}“You have rejected the truth, and rebelled a\-gainst God. In\-stead of stor\-ing up for your\-selves treasures in heav\-en, where noth\-ing destroys, you are ac\-cu\-mu\-lat\-ing for your\-selves an\-ger for the day of judg\-ment.”
\end{para}

\begin{para}
    \textsuperscript{10:18}\thinspace{}They would not lis\-ten to his words.
\end{para}

\begin{para}
    \textsuperscript{12:1}\thinspace{}Thus we can see the unsteadiness of the hearts of the chil\-dren of men.
    \textsuperscript{2}\thinspace{}At the ver\-y time when God prospers his peo\-ple, do\-ing all things for their wel\-fare and hap\-pi\-ness, then is the time that they hard\-en their hearts, and for\-get God, be\-cause of their ease, and ver\-y great pros\-per\-i\-ty.
\end{para}

\begin{para}
    \textsuperscript{12:4}\thinspace{}How fool\-ish, vain, and quick to do in\-iq\-ui\-ty, and how slow to do good, are the chil\-dren of men. How quick to set their hearts on the vain things of the world!
    \textsuperscript{5}\thinspace{}How quick to be lifted up in pride; how quick to boast; and how slow to re\-mem\-ber God, and to lis\-ten to his guid\-ance. How slow to walk in wisdom's paths!
\end{para}

\begin{para}
    \textsuperscript{12:6}\thinspace{}They ne\-glect God's guid\-ance, and will not have him as their guide.
    \textsuperscript{23}\thinspace{}Bless\-ed are they who will re\-pent and lis\-ten to the voice of God.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{13:2}\thinspace{}There was one Samuel who came in\-to the land, and preached re\-pent\-ance to the peo\-ple.
    \textsuperscript{4}\thinspace{}They would not let him en\-ter the cit\-y, so he got on the cit\-y wall, and
    \textsuperscript{5}\thinspace{}said to them:
\end{para}

\begin{para}
    \textsuperscript{13:8}\thinspace{}God says, “Be\-cause of the hard\-ness of the hearts of the peo\-ple, un\-less they re\-pent I will take a\-way my word from them.
    \textsuperscript{10}\thinspace{}Your enemies, of the fourth gen\-er\-a\-tion, will live to see your ut\-ter de\-struc\-tion.”
    \textsuperscript{21}\thinspace{}Your rich\-es are curs\-ed be\-cause you have set your hearts on them, and have not listened to the words of him who gave them to you.
    \textsuperscript{22}\thinspace{}You don't re\-mem\-ber God in the things with which he has bless\-ed you, but you al\-ways re\-mem\-ber your rich\-es, not to thank God for them. Your hearts swell with great pride, un\-to boasting, en\-vy, strife, mal\-ice, per\-se\-cu\-tion, mur\-der, and all man\-ner of in\-iq\-ui\-ty.
\end{para}

\begin{para}
    \textsuperscript{13:24}\thinspace{}Woe to you, be\-cause you drive out the prophets, and mock them, and throw stones at them, and kill them, just as they did long a\-go.
    \textsuperscript{25}\thinspace{}And now you say, “If we had lived in the time of our fathers, we would not have killed the prophets. We would not have stoned them, and driv\-en them out.”
\end{para}

\begin{para}
    \textsuperscript{13:26}\thinspace{}You are worse than they. If a proph\-et co\-mes to you and declares your sins and iniquities, you are an\-gry with him, and look for ways to de\-stroy him. You will say that he is a false proph\-et, be\-cause he testifies that your deeds are evil.
\end{para}

\begin{para}
    \textsuperscript{13:27}\thinspace{}But if a man co\-mes to you and says, “Do this and there is no in\-iq\-ui\-ty. Walk af\-ter the pride of your eyes, and do what\-ev\-er your heart desires,” you will say that he is a proph\-et.
    \textsuperscript{28}\thinspace{}You will dress him in ex\-pen\-sive cloth\-ing be\-cause he speaks flattering words, and says that all is well.
\end{para}

\begin{para}
    \textsuperscript{14:30}\thinspace{}Re\-mem\-ber that you are free. You are permitted to act for your\-selves. God
    \textsuperscript{31}\thinspace{}has al\-lowed you to know good from evil, and to choose life or death. You can do good and have good re\-stored to you, or you can do evil and have evil re\-stored to you.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{16:1}\thinspace{}Man\-y heard the words of Samuel, and be\-lieved him, and confessed their sins.
    \textsuperscript{2}\thinspace{}But those who did not be\-lieve Samuel were an\-gry with him. They threw stones and shot ar\-rows at him as he stood on the wall, but they could not hit him.
\end{para}

\begin{para}
    \textsuperscript{16:7}\thinspace{}He jumped down from the wall, fled out of their land,
    \textsuperscript{8}\thinspace{}and was nev\-er heard from a\-gain.
\end{para}

\begin{para}
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{3 Nephi}\addcontentsline{toc}{section}{3 Nephi}
\begin{para}
    \textsuperscript{2:10}\thinspace{}The peo\-ple still remained in wick\-ed\-ness.
    \textsuperscript{11}\thinspace{}The Gadianton robbers had be\-come so nu\-mer\-ous, and killed so man\-y of the peo\-ple, that all the peo\-ple
    \textsuperscript{12}\thinspace{}were com\-pelled, for the safe\-ty of their lives and their wom\-en and their chil\-dren, to take up arms a\-gainst those Gadianton robbers to main\-tain their lib\-er\-ty.
\end{para}

\begin{para}
    \textsuperscript{6:4}\thinspace{}The peo\-ple be\-gan a\-gain to pros\-per, and there was great or\-der in the land.
    \textsuperscript{5}\thinspace{}There was noth\-ing to hind\-er the peo\-ple from prospering con\-tin\-u\-al\-ly, ex\-cept if they fell in\-to trans\-gres\-sion.
\end{para}

\begin{para}
    \textsuperscript{6:10}\thinspace{}There be\-gan to be some disputes a\-mong the peo\-ple, and some were lifted up in pride and boasting be\-cause of their ver\-y great rich\-es, e\-ven to great persecutions.
    \textsuperscript{12}\thinspace{}The peo\-ple be\-gan to be dis\-tin\-guished by rank, ac\-cord\-ing to their rich\-es and their chances for learn\-ing. Some were ig\-no\-rant be\-cause of their pov\-er\-ty, and others re\-ceived great learn\-ing be\-cause of their rich\-es.
    \textsuperscript{14}\thinspace{}And thus there a\-rose a great in\-e\-qual\-i\-ty in the land, so the church be\-gan to be bro\-ken up.
\end{para}

\begin{para}
    \textsuperscript{6:15}\thinspace{}Now the cause of the in\-iq\-ui\-ty was this: the peo\-ple were puffing up with pride, seek\-ing for pow\-er, au\-thor\-i\-ty, rich\-es, and the vain things of the world.
    \textsuperscript{18}\thinspace{}They did not sin ig\-no\-rant\-ly, be\-cause God's will had been taught to them, there\-fore they wil\-ful\-ly rebelled a\-gainst God.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{7:14}\thinspace{}The peo\-ple di\-vid\-ed in\-to tribes, eve\-ry man ac\-cord\-ing to his fam\-i\-ly. They had come to an a\-gree\-ment that they would not go to war with one an\-oth\-er, or tres\-pass a\-gainst one an\-oth\-er, so in some de\-gree they had peace in the land. Nev\-er\-the\-less, their hearts were turned from God, and they stoned the prophets and cast them out.
\end{para}

\begin{para}
    \textsuperscript{7:17}\thinspace{}Nephi preached man\-y things to the peo\-ple.
    \textsuperscript{18}\thinspace{}They were an\-gry with him.
    \textsuperscript{21}\thinspace{}On\-ly a few were con\-vert\-ed to God.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{11:1}\thinspace{}A great mul\-ti\-tude were gathered to\-geth\-er
    \textsuperscript{2}\thinspace{}talking a\-bout Jesus Christ.
    \textsuperscript{3}\thinspace{}While they were talking with one an\-oth\-er, they heard a voice as if it came from heav\-en.
    \textsuperscript{8}\thinspace{}They looked up, and saw a man descending out of heav\-en. He came down and stood a\-mong them.
    \textsuperscript{9}\thinspace{}He reached out his hand and spoke to the peo\-ple, say\-ing,
    \textsuperscript{10}\thinspace{}“I am Jesus Christ, whom the prophets tes\-ti\-fied would come in\-to the world.”
\end{para}

\begin{para}
    \textsuperscript{11:12}\thinspace{}When Jesus said these words the whole mul\-ti\-tude fell to the ground, for they remembered that it had been prophesied that Christ would show him\-self to them.
\end{para}

\begin{para}
    \textsuperscript{11:18}\thinspace{}Jesus commanded Nephi to come for\-ward,
    \textsuperscript{22}\thinspace{}and called others, and said to them,
    \textsuperscript{28}\thinspace{}“There shall be no disputes a\-mong you, as there have been un\-til now.
    \textsuperscript{29}\thinspace{}For he that has the spir\-it of con\-ten\-tion is not of me, but is of the dev\-il, who is the fa\-ther of con\-ten\-tion, and he stirs up the hearts of men to con\-tend with an\-ger, one with an\-oth\-er.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{12:1}\thinspace{}When Jesus had spo\-ken these words to Nephi, and those who had been called, he said to the mul\-ti\-tude:
\end{para}

\begin{para}
    \textsuperscript{12:3}\thinspace{}Bless\-ed are the poor in spir\-it who come to me, for the king\-dom of heav\-en is theirs.
    \textsuperscript{4}\thinspace{}Bless\-ed are all those who mourn, for they will be comforted.
    \textsuperscript{5}\thinspace{}Bless\-ed are the meek, for they will in\-her\-it the earth.
    \textsuperscript{6}\thinspace{}Bless\-ed are all those who hun\-ger and thirst for right\-eous\-ness, for they will be filled with the ho\-ly spir\-it.
    \textsuperscript{7}\thinspace{}Bless\-ed are the mer\-ci\-ful, for they will re\-ceive mer\-cy.
    \textsuperscript{8}\thinspace{}Bless\-ed are all the pure in heart, for they will see God.
    \textsuperscript{9}\thinspace{}Bless\-ed are all the peacemakers, for they will be called sons of God.
    \textsuperscript{10}\thinspace{}Bless\-ed are all those who are per\-se\-cut\-ed be\-cause of my name, for the king\-dom of heav\-en is theirs.
    \textsuperscript{11}\thinspace{}Bless\-ed are you when peo\-ple in\-sult you and per\-se\-cute you, and false\-ly say all kinds of evil a\-gainst you be\-cause of me,
    \textsuperscript{12}\thinspace{}for you will have great joy and be glad, for your re\-ward in heav\-en will be great; for in the same way they per\-se\-cut\-ed the prophets who were be\-fore you.
\end{para}

\begin{para}
    \textsuperscript{12:15}\thinspace{}Do peo\-ple light a lamp and put it un\-der a bas\-ket? No, but on a stand, and it gives light to eve\-ry\-one in the house.
    \textsuperscript{16}\thinspace{}There\-fore let your light shine be\-fore this peo\-ple, so that they may see your good works and glo\-ri\-fy God who is in heav\-en.
\end{para}

\begin{para}
    \textsuperscript{12:19}\thinspace{}You have the commandments.
    \textsuperscript{20}\thinspace{}Un\-less you keep my commandments, you will nev\-er en\-ter in\-to the king\-dom of heav\-en.
\end{para}

\begin{para}
    \textsuperscript{12:21}\thinspace{}It is writ\-ten that you shall not mur\-der, and who\-ev\-er murders will be sub\-ject to God's judg\-ment.
    \textsuperscript{22}\thinspace{}But I tell you that eve\-ry\-one who is an\-gry with his broth\-er will be sub\-ject to his judg\-ment. And who\-ev\-er says, “You fool,” will be sub\-ject to hell fire.
    \textsuperscript{23}\thinspace{}There\-fore, if you want to come to me, and re\-mem\-ber that your broth\-er has some\-thing a\-gainst you,
    \textsuperscript{24}\thinspace{}go to your broth\-er, and first be rec\-on\-ciled to him, and then come to me with full pur\-pose of heart, and I will re\-ceive you.
\end{para}

\begin{para}
    \textsuperscript{12:27}\thinspace{}It is writ\-ten that you shall not com\-mit a\-dul\-ter\-y.
    \textsuperscript{28}\thinspace{}But I tell you that eve\-ry\-one who looks at a wom\-an with lust for her has al\-read\-y com\-mit\-ted a\-dul\-ter\-y in his heart.
\end{para}

\begin{para}
    \textsuperscript{12:29}\thinspace{}Al\-low none of these things to en\-ter your heart,
    \textsuperscript{30}\thinspace{}for it is bet\-ter to de\-ny your\-selves these things than to be thrown in\-to hell.
\end{para}

\begin{para}
    \textsuperscript{12:33}\thinspace{}It is writ\-ten that you shall not break your oath, but shall ful\-fill your oaths to God.
    \textsuperscript{34}\thinspace{}But I tell you, make no oath at all,
    \textsuperscript{37}\thinspace{}but let your state\-ment be “Yes” or “No”. Any\-thing more than this is evil.
\end{para}

\begin{para}
    \textsuperscript{12:38}\thinspace{}It is writ\-ten, an eye for an eye, and a tooth for a tooth.
    \textsuperscript{39}\thinspace{}But I tell you, don't re\-sist an evil per\-son. In\-stead, if some\-one slaps you on the right cheek, turn to him the oth\-er al\-so.
    \textsuperscript{40}\thinspace{}If some\-one wants to sue you and take your shirt, let him have your coat al\-so.
    \textsuperscript{41}\thinspace{}If some\-one forces you to go one mile, go with him two.
    \textsuperscript{42}\thinspace{}Give to him who asks you, and don't turn a\-way from him who wants to bor\-row from you.
\end{para}

\begin{para}
    \textsuperscript{12:43}\thinspace{}It is writ\-ten, “Love your neigh\-bor and hate your en\-e\-my.”
    \textsuperscript{44}\thinspace{}But I tell you, love your enemies, bless those who curse you, do good to those who hate you, and pray for those who mis\-treat you and per\-se\-cute you
    \textsuperscript{45}\thinspace{}that you may be sons of God. For he makes his sun rise on the evil and the good.
\end{para}

\begin{para}
    \textsuperscript{13:1}\thinspace{}You should give to the poor, but be\-ware of prac\-tic\-ing your right\-eous\-ness be\-fore men to be seen by them.
    \textsuperscript{2}\thinspace{}So when you give to the poor, don't sound a trum\-pet be\-fore you, as the hypocrites do, to be praised by men. They have their re\-ward.
    \textsuperscript{3}\thinspace{}But when you give to the poor, don't let your left hand know what your right hand is do\-ing,
    \textsuperscript{4}\thinspace{}so that your giv\-ing may be in se\-cret.
\end{para}

\begin{para}
    \textsuperscript{13:5}\thinspace{}When you pray, don't do as the hypocrites, for they love to pray to be seen by men. They have their re\-ward.
    \textsuperscript{6}\thinspace{}But when you pray, go in\-to your room, shut your door, and pray to God who is in se\-cret.
    \textsuperscript{7}\thinspace{}When you pray, don't use mean\-ing\-less rep\-e\-ti\-tion like the pagans, for they think that they will be heard for their man\-y words.
\end{para}

\begin{para}
    \textsuperscript{13:16}\thinspace{}When you fast, don't look gloom\-y like the hypocrites, for they dis\-fig\-ure their faces to show men that they are fasting. They have their re\-ward.
    \textsuperscript{17}\thinspace{}But when you fast, a\-noint your head, and wash your face,
    \textsuperscript{18}\thinspace{}so that your fasting is not ob\-vi\-ous to men, but to God, who is in se\-cret.
\end{para}

\begin{para}
    \textsuperscript{13:19}\thinspace{}Don't store up for your\-selves treasures on earth, where moth and rust de\-stroy, and thieves break in and steal.
    \textsuperscript{20}\thinspace{}But store up for your\-selves treasures in heav\-en, where nei\-ther moth nor rust destroys, and where thieves do not break in and steal.
    \textsuperscript{21}\thinspace{}For where your treas\-ure is, there your heart will be al\-so.
\end{para}

\begin{para}
    \textsuperscript{13:24}\thinspace{}No one can serve two masters; for ei\-ther he will hate the one and love the oth\-er, or he will be de\-vot\-ed to the one and des\-pise the oth\-er. You can\-not serve God and mon\-ey.
\end{para}

\begin{para}
    \textsuperscript{14:1}\thinspace{}Don't judge, so that you won't be judged.
    \textsuperscript{2}\thinspace{}For in the way you judge, you will be judged; and with the meas\-ure you use, it will be meas\-ured to you.
\end{para}

\begin{para}
    \textsuperscript{14:3}\thinspace{}Why do you look at the speck in your brother's eye, but don't no\-tice the log in your own eye?
    \textsuperscript{12}\thinspace{}There\-fore, what\-ev\-er you want others to do to you, do al\-so to them.
\end{para}

\begin{para}
    \textsuperscript{14:15}\thinspace{}Be\-ware of false prophets who come to you in sheep's cloth\-ing, but in\-ward\-ly are rav\-en\-ous wolves.
    \textsuperscript{16}\thinspace{}You will rec\-og\-nize them by their fruits. Are grapes gathered from thornbushes, or figs from thistles?
    \textsuperscript{17}\thinspace{}Like\-wise, eve\-ry good tree bears good fruit, but a bad tree bears bad fruit.
    \textsuperscript{18}\thinspace{}A good tree can\-not bear bad fruit, nor can a bad tree bear good fruit.
    \textsuperscript{20}\thinspace{}So you'll rec\-og\-nize them by their fruits.
\end{para}

\begin{para}
    \textsuperscript{14:24}\thinspace{}Eve\-ry\-one who hears these words of mine and acts on them will be like a wise man who built his house on the rock.
    \textsuperscript{25}\thinspace{}The rain fell, the floods came, and the winds blew and beat a\-gainst that house. Yet it did not fall, be\-cause its founda\-tion was on the rock.
\end{para}

\begin{para}
    \textsuperscript{14:26}\thinspace{}Eve\-ry\-one who hears these words of mine and does\-n't act on them, will be like a fool\-ish man who built his house on the sand.
    \textsuperscript{27}\thinspace{}The rain fell, the floods came, and the winds blew and beat a\-gainst that house, and it fell.
\end{para}

\begin{para}
    \textsuperscript{15:9}\thinspace{}Turn to me, and en\-dure to the end, and you will live. For to him who endures to the end, I will give e\-ter\-nal life.
    \textsuperscript{10}\thinspace{}I've giv\-en you the commandments, there\-fore keep my commandments.
\end{para}

\begin{para}
    \textsuperscript{17:2}\thinspace{}I can see that you're weak, that you can\-not un\-der\-stand all that I'm speak\-ing to you.
    \textsuperscript{3}\thinspace{}There\-fore, go to your homes, and pon\-der on the things which I've said, and ask God that you may un\-der\-stand.
\end{para}

\begin{para}
    \textsuperscript{18:18}\thinspace{}You must watch and pray al\-ways to a\-void temp\-ta\-tion.
    \textsuperscript{21}\thinspace{}Pray to God in your families that your wives and your chil\-dren may be bless\-ed.
    \textsuperscript{22}\thinspace{}Meet to\-geth\-er of\-ten, and for\-bid no man from com\-ing to you when you meet to\-geth\-er,
    \textsuperscript{23}\thinspace{}but pray for them, and do not cast them out.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{18:36}\thinspace{}When Jesus fin\-ished these sayings,
    \textsuperscript{39}\thinspace{}he de\-part\-ed from them.
\end{para}

\begin{para}
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Mormon}\addcontentsline{toc}{section}{Mormon}
\begin{para}
    \textsuperscript{8:1}\thinspace{}I, Moroni, fin\-ish the rec\-ord of my fa\-ther, Mormon. I have on\-ly a few things to write, which things my fa\-ther commanded me.
\end{para}

\begin{para}
    \textsuperscript{8:3}\thinspace{}Our enemies killed my fa\-ther, and I a\-lone re\-main to write the sad tale of the de\-struc\-tion of my peo\-ple. I don't know wheth\-er they will kill me.
    \textsuperscript{5}\thinspace{}All my relatives have been killed in bat\-tle, and I have no friends, and no\-where to go.
\end{para}

\begin{para}
    \textsuperscript{8:12}\thinspace{}Who\-ev\-er receives this rec\-ord, and does\-n't con\-demn it be\-cause of its imperfections, will know of greater things than these;
    \textsuperscript{19}\thinspace{}for one who judges rash\-ly will be judged rash\-ly a\-gain; for his wages will be ac\-cord\-ing to his works.
    \textsuperscript{20}\thinspace{}The scrip\-ture says, “‘Man shall not strike, nei\-ther shall he judge, for judg\-ment is mine, and venge\-ance is mine al\-so, and I will re\-pay,’ says God.”
\end{para}

\begin{para}
    \textsuperscript{8:22}\thinspace{}For the e\-ter\-nal purposes of God will roll on un\-til all his promises are ful\-filled.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{8:31}\thinspace{}Man\-y will say, “Do this, or do that, and it does\-n't mat\-ter, for God will up\-hold them at the last day.” But woe to them, for they are in the gall of bit\-ter\-ness and in the bonds of in\-iq\-ui\-ty.
    \textsuperscript{32}\thinspace{}Churches will be built up that will say, “Come to me, and for your mon\-ey you will be for\-giv\-en of your sins.”
    \textsuperscript{33}\thinspace{}You wick\-ed and per\-verse and stub\-born peo\-ple.
\end{para}

\begin{para}
    \textsuperscript{8:36}\thinspace{}I know that you walk in the pride of your hearts, to wear\-ing ver\-y fi\-ne cloth\-ing, to en\-vy\-ing, strife, mal\-ice, per\-se\-cu\-tion, and all man\-ner of in\-iq\-ui\-ty.
    \textsuperscript{37}\thinspace{}For you love mon\-ey, and your possessions, and your fi\-ne cloth\-ing, and adorning your churches, more than you love the poor and the needy, the sick and the afflicted.
\end{para}

\begin{para}
    \textsuperscript{8:39}\thinspace{}Why do you a\-dorn your\-selves with what has no life and yet al\-low the hun\-gry, the needy, the na\-ked, the sick, and the afflicted to pass by you, and not no\-tice them?
\end{para}

\begin{para}
    \textsuperscript{9:27}\thinspace{}Don't doubt, but be\-lieve, and come to God with all your heart, and work out your own sal\-va\-tion with fear and trem\-bling be\-fore him.
    \textsuperscript{28}\thinspace{}Be wise dur\-ing your pro\-ba\-tion. Strip your\-selves of all im\-pu\-ri\-ty. Don't ask so that you may spend it on your pleasures, but ask with an un\-wa\-ver\-ing firm\-ness that you will yield to no temp\-ta\-tion, but serve the true and liv\-ing God.
\end{para}

\begin{para}
    \textsuperscript{9:31}\thinspace{}Don't con\-demn me or my fa\-ther be\-cause of our imperfections, nei\-ther those who have writ\-ten be\-fore us. Rath\-er, give thanks to God that he has revealed our imperfections to you, that you may learn to be wis\-er than we have been.
\end{para}

\begin{para}
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Moroni}\addcontentsline{toc}{section}{Moroni}
\begin{para}
    \textsuperscript{7:1}\thinspace{}I, Moroni, write some of the words of my fa\-ther Mormon.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{7:2}\thinspace{}I, Mormon, speak to you.
    \textsuperscript{5}\thinspace{}I re\-mem\-ber God's word, which says you will rec\-og\-nize them by their works, for if their works are good then they are al\-so good.
    \textsuperscript{6}\thinspace{}For God has said, an evil man can\-not do what is good. For if he offers a gift, or prays to God, un\-less he does it with re\-al in\-tent he gains noth\-ing,
    \textsuperscript{7}\thinspace{}for it is not counted to him as right\-eous\-ness.
    \textsuperscript{8}\thinspace{}For if an evil man gives a gift, he does it re\-luc\-tant\-ly, there\-fore it is counted to him the same as if he had retained the gift, there\-fore he is counted evil be\-fore God.
\end{para}

\begin{para}
    \textsuperscript{7:9}\thinspace{}Like\-wise, it is counted evil to a man if he prays with\-out re\-al in\-tent of heart. He gains noth\-ing, for God receives none such.
    \textsuperscript{10}\thinspace{}There\-fore, an evil man can\-not do what is good.
    \textsuperscript{11}\thinspace{}A bit\-ter spring can\-not pour out good wa\-ter; nei\-ther can a good spring pour out bit\-ter wa\-ter.
\end{para}

\begin{para}
    \textsuperscript{7:13}\thinspace{}That which is of God invites and entices to do good con\-tin\-u\-al\-ly. There\-fore, eve\-ry\-thing which invites and entices to do good, to love God, and to serve him, is in\-spired of God.
    \textsuperscript{14}\thinspace{}Be on guard that you don't judge what is evil to be of God, or what is good and of God to be of the dev\-il.
\end{para}

\begin{para}
    \textsuperscript{7:18}\thinspace{}See that you don't judge un\-just\-ly, for with the same judg\-ment that you judge you will al\-so be judged.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{7:44}\thinspace{}If a man does\-n't have love he is noth\-ing.
    \textsuperscript{45}\thinspace{}Love is pa\-tient, is kind, does not en\-vy, is not ar\-ro\-gant, is not self-seeking, is not ir\-ri\-ta\-ble, keeps no rec\-ord of wrongs, does not re\-joice in unrighteousness but rejoices in the truth, bears all things, believes all things, hopes all things, and endures all things.
\end{para}

\begin{para}
    \textsuperscript{7:46}\thinspace{}There\-fore, love nev\-er ends. Hold fast to love, which is the great\-est of all, for all things must end,
    \textsuperscript{47}\thinspace{}but love endures for\-ev\-er, and who\-ev\-er possesses it at the last day, it will go well with him.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{10:1}\thinspace{}Now I, Moroni,
    \textsuperscript{3}\thinspace{}ex\-hort you that when you read these things, you should re\-mem\-ber how mer\-ci\-ful God has been to the chil\-dren of men, and pon\-der it in your hearts.
\end{para}

\begin{para}
    \textsuperscript{10:8}\thinspace{}Don't de\-ny the gifts of God, for they are man\-y, and they come from the same God. There are dif\-fer\-ent ways that these gifts are administered, but it is the same God who works all things in all men, and they are giv\-en to prof\-it men.
    \textsuperscript{9}\thinspace{}To one is giv\-en that he may teach a mes\-sage of wis\-dom,
    \textsuperscript{10}\thinspace{}to an\-oth\-er teach\-ing a mes\-sage of knowl\-edge,
    \textsuperscript{11}\thinspace{}to an\-oth\-er ver\-y great faith, to an\-oth\-er the gifts of heal\-ing,
    \textsuperscript{12}\thinspace{}to an\-oth\-er work\-ing might\-y miracles,
    \textsuperscript{13}\thinspace{}to an\-oth\-er proph\-e\-sy a\-bout all things,
    \textsuperscript{14}\thinspace{}to an\-oth\-er see\-ing angels and ministering spirits,
    \textsuperscript{15}\thinspace{}to an\-oth\-er all kinds of tongues,
    \textsuperscript{16}\thinspace{}and to an\-oth\-er the in\-ter\-pre\-ta\-tion of languages.
    \textsuperscript{17}\thinspace{}All these gifts come from God to eve\-ry man in\-di\-vid\-u\-al\-ly, as he wills.
\end{para}

\begin{para}
    \textsuperscript{10:21}\thinspace{}Un\-less you have love, faith, and hope, you can\-not be saved.
    \textsuperscript{22}\thinspace{}If you don't have hope you must be in des\-pair, and des\-pair co\-mes be\-cause of in\-iq\-ui\-ty.
\end{para}

\begin{para}
    \textsuperscript{10:34}\thinspace{}I say to all, fare\-well. I soon go to rest in God's par\-a\-dise.
\end{para}

\begin{para}
\end{para}


\end{document}

% Copyright 2021 Michael Hendricks
%
% Permission is granted to use, copy, modify, and distribute
% this work for any purpose with or without fee.
%
% This work is provided as-is without any express or implied
% warranty. In no event is the author liable for any damages
% arising from this work.
