%
% see tex.stackexchange.com/a/600/151276
\def\changemargin#1#2{\list{}{\rightmargin#2\leftmargin#1}\item[]}
\let\endchangemargin=\endlist

\documentclass[openany,12pt,english]{book}
\usepackage[utf8]{inputenc}
\usepackage{babel}
\usepackage{microtype}
\usepackage{csquotes}

\usepackage[
  paperwidth=8.5in,
  paperheight=11in,
  top=1in,
  outer=1in,
  bottom=1.25in,
  inner=1.5in
]{geometry}

\title{Buddha}
\author{}
\date{}

\newenvironment{para}{\par\pretolerance=100\tolerance=200\setlength{\emergencystretch}{0.6em}\relax}{\par}

\begin{document}

\maketitle

% Copyright page
{
\thispagestyle{empty}
\vspace*{\fill}
\begin{changemargin}{.8in}{.8in}
\begin{center}
\textcopyright{}~2021 Michael Hendricks

\vspace{0.5\baselineskip}

Permission is granted to use, copy, modify, and distribute
this work for any purpose with or without fee.

\vspace{0.5\baselineskip}

This work is provided as-is without any express or implied
warranty. In no event is the author liable for any damages
arising from this work.
\end{center}
\end{changemargin}
}
\pagenumbering{arabic}
\clearpage{}

% Table of Contents
\tableofcontents
\clearpage{}

\markboth{}{}
\twocolumn[\begin{changemargin}{.6in}{.6in}
  \begin{center}
    \itshape
    A col\-lec\-tion of sayings of the Bud\-dha tak\-en from the Dham\-ma\-pa\-da.
    Verse references such
    as~\textsuperscript{3}\thinspace{}indicate
    the original source of the text, but
    don't imply a full quotation.
  \end{center}
\end{changemargin}]{}


\vspace{0pt plus 8\baselineskip{}}\section*{Pairs}\addcontentsline{toc}{section}{Pairs}
\begin{para}
    \textsuperscript{1}\thinspace{}Our life is shaped by our mind: it is ruled by our mind, it is made by our mind. If one speaks or acts with an evil mind, suf\-fer\-ing follows him, as the wheel follows the foot of the ox that draws the cart.
    \textsuperscript{2}\thinspace{}If one speaks or acts with a pure mind, hap\-pi\-ness follows him, like a shad\-ow that nev\-er leaves him.
\end{para}

\begin{para}
    \textsuperscript{3}\thinspace{}“He insulted me, hit me, defeated me, robbed me.” In those who har\-bor such thoughts, ha\-tred will nev\-er cease.
    \textsuperscript{4}\thinspace{}In those who do not har\-bor such thoughts, ha\-tred will cease.
    \textsuperscript{5}\thinspace{}For ha\-tred nev\-er ceases by ha\-tred. Ha\-tred ceases by love.
    \textsuperscript{6}\thinspace{}Some do not re\-al\-ize that we all must die, but those who do re\-al\-ize it cease their quarrels at once.
\end{para}

\begin{para}
    \textsuperscript{7}\thinspace{}Just as the wind throws down a weak tree, so will the temp\-ter over\-throw him who fo\-cus\-es on\-ly on pleasures, his sen\-ses un\-con\-trolled, im\-mod\-er\-ate in his food, i\-dle, and weak.
    \textsuperscript{8}\thinspace{}Just as the wind can\-not throw down a rock\-y moun\-tain, so the temp\-ter will not over\-throw him who does not fo\-cus on pleasures, his sen\-ses well con\-trolled, mod\-er\-ate in his food, faith\-ful and strong.
\end{para}

\begin{para}
    \textsuperscript{9}\thinspace{}He who puts on the monk's yel\-low robe with\-out hav\-ing cleansed him\-self from sin, who disregards self-control and truth, is un\-wor\-thy of the robe.
    \textsuperscript{10}\thinspace{}But he who has cleansed him\-self from sin, is well grounded in vir\-tues, and regards self-control and truth, is in\-deed wor\-thy of the robe.
\end{para}

\begin{para}
    \textsuperscript{11}\thinspace{}They who mis\-take the triv\-i\-al for the es\-sen\-tial, and the es\-sen\-tial for the triv\-i\-al, nev\-er ar\-rive at the es\-sen\-tial, but fol\-low vain desires.
    \textsuperscript{12}\thinspace{}They who know the es\-sen\-tial as es\-sen\-tial, and the triv\-i\-al as triv\-i\-al, ar\-rive at the es\-sen\-tial, and fol\-low true desires.
\end{para}

\begin{para}
    \textsuperscript{13}\thinspace{}Just as rain seeps through a poor\-ly tha\-tched house, pas\-sion seeps through an undeveloped mind.
    \textsuperscript{14}\thinspace{}Just as rain can\-not seep through a well tha\-tched house, pas\-sion can\-not seep through a well de\-vel\-oped mind.
\end{para}

\begin{para}
    \textsuperscript{15}\thinspace{}The evil\-do\-er grieves in both this world and the next. He grieves and suffers when he sees the evil of his own work.
    \textsuperscript{16}\thinspace{}The vir\-tu\-ous man delights in both this world and the next. He delights and rejoices when he sees the pu\-ri\-ty of his own work.
\end{para}

\begin{para}
    \textsuperscript{17}\thinspace{}The evil\-do\-er suffers both in this world and the next. He suffers when he thinks of the evil he has done. He suffers more when go\-ing to the next world.
    \textsuperscript{18}\thinspace{}The vir\-tu\-ous man is hap\-py both in this world and the next. He is hap\-py when he thinks of the good he has done. He is hap\-pi\-er still when go\-ing to the next world.
\end{para}

\begin{para}
    \textsuperscript{19}\thinspace{}The thought\-less man, if he recites man\-y scriptures, but does\-n't do what they say, has no share in the blessings of the ho\-ly life, but is like a cow\-herd counting the cows of others.
    \textsuperscript{20}\thinspace{}He who recites on\-ly a few scriptures, but does what they say, hav\-ing for\-sak\-en lust and ha\-tred and de\-lu\-sion, possesses true knowl\-edge and se\-ren\-i\-ty of mind, car\-ing for noth\-ing in this world or that to come, he shares in the blessings of the ho\-ly life.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Vigilance}\addcontentsline{toc}{section}{Vigilance}
\begin{para}
    \textsuperscript{21}\thinspace{}Vig\-i\-lance is the path to sal\-va\-tion. Care\-less\-ness is the path to death. Those who are vig\-i\-lant don't die. Those who are care\-less are as if dead al\-read\-y.
    \textsuperscript{22}\thinspace{}The wise, hav\-ing un\-der\-stood this clear\-ly, de\-light in vig\-i\-lance, and re\-joice in the knowl\-edge of the no\-ble ones.
    \textsuperscript{23}\thinspace{}These wise peo\-ple, meditative, stead\-fast\-ly per\-se\-ver\-ing, at\-tain to sal\-va\-tion, the highest hap\-pi\-ness.
\end{para}

\begin{para}
    \textsuperscript{24}\thinspace{}If a vig\-i\-lant per\-son is en\-er\-get\-ic, if he is mind\-ful, if his deeds are pure, if he acts with con\-sid\-er\-a\-tion, if he has self-control, and is right\-eous, then his glo\-ry will in\-crease.
    \textsuperscript{25}\thinspace{}By ef\-fort, by vig\-i\-lance, by dis\-ci\-pline and self-control, the wise man may make for him\-self an is\-land which no flood can over\-whelm.
\end{para}

\begin{para}
    \textsuperscript{26}\thinspace{}The fool\-ish and ig\-no\-rant in\-dulge in care\-less\-ness. The wise man keeps vig\-i\-lance as his great\-est treas\-ure.
    \textsuperscript{27}\thinspace{}Do not give way to care\-less\-ness or to the en\-joy\-ment of sen\-su\-al pleasures. He who is vig\-i\-lant and meditative, obtains am\-ple joy.
\end{para}

\begin{para}
    \textsuperscript{28}\thinspace{}When the wise man drives a\-way care\-less\-ness by vig\-i\-lance, he climbs the tow\-er of wis\-dom and looks down up\-on the fools. With\-out suf\-fer\-ing, he looks at the suf\-fer\-ing crowd, as one that stands on a moun\-tain looks at them that stand on the ground be\-low.
\end{para}

\begin{para}
    \textsuperscript{29}\thinspace{}Vig\-i\-lant a\-mong the care\-less, a\-wake a\-mong the sleep\-ing, the wise man ad\-vanc\-es like a fast horse, leav\-ing the weak be\-hind.
    \textsuperscript{30}\thinspace{}Vig\-i\-lance is al\-ways praised; care\-less\-ness is al\-ways de\-spised.
\end{para}

\begin{para}
    \textsuperscript{31}\thinspace{}The monk who delights in vig\-i\-lance, who looks with fear at care\-less\-ness: moves a\-bout like fire, burn\-ing all fetters, small and large;
    \textsuperscript{32}\thinspace{}can\-not fall a\-way. He is close to sal\-va\-tion.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Mind}\addcontentsline{toc}{section}{Mind}
\begin{para}
    \textsuperscript{33}\thinspace{}Just as a fletch\-er straightens his ar\-row, a wise man straightens his trem\-bling and un\-stead\-y mind, which is dif\-fi\-cult to guard, dif\-fi\-cult to re\-strain.
\end{para}

\begin{para}
    \textsuperscript{34}\thinspace{}Just as a fish quivers when tak\-en from its home in the wa\-ter and thrown on land, our mind trem\-bles all over to es\-cape the in\-flu\-ence of the temp\-ter.
\end{para}

\begin{para}
    \textsuperscript{35}\thinspace{}It is good to tame the mind, which is dif\-fi\-cult to tame and flight\-y, rush\-ing wher\-ev\-er it wants. A tamed mind brings hap\-pi\-ness.
    \textsuperscript{36}\thinspace{}Let the wise man guard his mind, for it is sub\-tle and e\-lu\-sive, and it rushes wher\-ev\-er it wants. A guard\-ed mind brings hap\-pi\-ness.
\end{para}

\begin{para}
    \textsuperscript{37}\thinspace{}The mind wanders far, moves a\-bout a\-lone, is with\-out a bod\-y, and hides in the ca\-ve of the heart. Those who bri\-dle it are free from the bonds of the temp\-ter.
\end{para}

\begin{para}
    \textsuperscript{38}\thinspace{}If a man's mind is un\-stead\-y, if he does\-n't know the true law, and if his faith is troubled, then his knowl\-edge will nev\-er be per\-fect.
    \textsuperscript{39}\thinspace{}If a man's mind is free from pas\-sion, if his mind is free from ill will, if he is un\-af\-fect\-ed by good and evil, then there is no dan\-ger for him.
\end{para}

\begin{para}
    \textsuperscript{40}\thinspace{}Know\-ing that this bod\-y is frag\-ile like a clay pot, and securing this mind like a for\-tress, at\-tack the temp\-ter with the sword of knowl\-edge. Then guard what is won with\-out feel\-ing at\-tach\-ment to it.
    \textsuperscript{41}\thinspace{}Soon, this bod\-y will lie on the ground, discarded, with\-out con\-scious\-ness, like a use\-less log.
\end{para}

\begin{para}
    \textsuperscript{42}\thinspace{}What\-ev\-er an en\-e\-my might do to an en\-e\-my, or a foe to a foe, a wrong\-ly di\-rect\-ed mind can do to you e\-ven worse.
    \textsuperscript{43}\thinspace{}What\-ev\-er a moth\-er, a fa\-ther, or any oth\-er rel\-a\-tive might do for you, a well di\-rect\-ed mind can do for you e\-ven more.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Flowers}\addcontentsline{toc}{section}{Flowers}
\begin{para}
    \textsuperscript{44}\thinspace{}Who will over\-come this earth, and the world of the dead, and the world of the gods? Who will choose the well taught path of vir\-tue, as an ex\-pert flo\-rist chooses the right flow\-er?
    \textsuperscript{45}\thinspace{}A dis\-ci\-ple will.
\end{para}

\begin{para}
    \textsuperscript{46}\thinspace{}Know\-ing that this bod\-y is like froth, and learn\-ing that it is in\-sub\-stan\-tial like a mi\-rage, you will re\-move the flower-tipped ar\-rows of the temp\-ter, and nev\-er see the king of death.
\end{para}

\begin{para}
    \textsuperscript{47}\thinspace{}As a flood sweeps a\-way a sleep\-ing vil\-lage, so death sweeps a\-way a man who is im\-mersed in gath\-er\-ing flowers, his mind dis\-tract\-ed.
    \textsuperscript{48}\thinspace{}Death subdues a man who is im\-mersed in gath\-er\-ing flowers, his mind dis\-tract\-ed, be\-fore he is sa\-ti\-at\-ed in his pleasures.
\end{para}

\begin{para}
    \textsuperscript{49}\thinspace{}As a bee collects nec\-tar and departs with\-out dam\-ag\-ing the flow\-er or its col\-or or its scent, so should a sage col\-lect alms in a vil\-lage.
\end{para}

\begin{para}
    \textsuperscript{50}\thinspace{}Do not no\-tice the faults of others, their sins of com\-mis\-sion or o\-mis\-sion, but no\-tice your own mis\-con\-duct and neg\-li\-gence.
\end{para}

\begin{para}
    \textsuperscript{51}\thinspace{}Like a beau\-ti\-ful flow\-er, full of col\-or but with\-out scent, are fair but fruit\-less words of him who does not prac\-tice them.
    \textsuperscript{52}\thinspace{}But, like a beau\-ti\-ful flow\-er, full of col\-or and full of scent, are fair and fruit\-ful words of him who practices them.
\end{para}

\begin{para}
    \textsuperscript{53}\thinspace{}Just as man\-y wreaths can be made from a heap of flowers, so can man\-y good things be done by a mor\-tal.
\end{para}

\begin{para}
    \textsuperscript{54}\thinspace{}The scent of flowers does not trav\-el against the wind, not e\-ven that of san\-dal\-wood, tagara or jas\-mine; but the rep\-u\-ta\-tion of the vir\-tu\-ous does trav\-el against the wind. The rep\-u\-ta\-tion of the vir\-tu\-ous pervades eve\-ry place.
    \textsuperscript{55}\thinspace{}A\-mong the scent of san\-dal\-wood, tagara, lo\-tus, and jas\-mine, the scent of vir\-tue is un\-sur\-passed.
\end{para}

\begin{para}
    \textsuperscript{56}\thinspace{}The scent that co\-mes from tagara and san\-dal\-wood is faint. The scent of the vir\-tu\-ous is best, it rises up to the gods.
\end{para}

\begin{para}
    \textsuperscript{58}\thinspace{}On a gar\-bage heap by the roadside, a lo\-tus grows full of sweet per\-fume and de\-light.
    \textsuperscript{59}\thinspace{}Like\-wise, on a gar\-bage heap of blind mortals, a true dis\-ci\-ple of the Bud\-dha shines forth in wis\-dom.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{The Fool}\addcontentsline{toc}{section}{The Fool}
\begin{para}
    \textsuperscript{60}\thinspace{}The night is long to the sleep\-less; a mile is long to the tired; life is long to the fool\-ish who do not know the true law.
\end{para}

\begin{para}
    \textsuperscript{61}\thinspace{}If a trav\-el\-er does not meet a com\-pan\-ion who is his bet\-ter, or his e\-qual, let him firm\-ly keep to his sol\-i\-tar\-y jour\-ney. There is no com\-pan\-ion\-ship with a fool.
    \textsuperscript{62}\thinspace{}He worries, “I have sons, and I have wealth,” but he him\-self is not his own, how much less sons and wealth?
\end{para}

\begin{para}
    \textsuperscript{63}\thinspace{}A fool who knows his own fool\-ish\-ness is wise to that ex\-tent. But a fool who thinks he is wise is a fool in\-deed.
\end{para}

\begin{para}
    \textsuperscript{64}\thinspace{}E\-ven if a fool associates with a wise man all his life, he will per\-ceive the truth no more than a spoon perceives the fla\-vor of the soup.
    \textsuperscript{65}\thinspace{}If an in\-tel\-li\-gent man associates with a wise man for on\-ly one min\-ute, he will soon per\-ceive the truth as the tongue perceives the taste of soup.
\end{para}

\begin{para}
    \textsuperscript{66}\thinspace{}Fools of lit\-tle un\-der\-stand\-ing are their own enemies, for they do evil deeds that bear bit\-ter fruit.
    \textsuperscript{67}\thinspace{}That deed is poor\-ly done if a man must re\-pent af\-ter\-ward, and its re\-ward is cry\-ing and a tear\-ful face.
    \textsuperscript{68}\thinspace{}That deed is well done if a man does not re\-pent af\-ter\-ward, and its re\-ward is de\-light and hap\-pi\-ness.
\end{para}

\begin{para}
    \textsuperscript{69}\thinspace{}As long as an evil deed does not bear fruit, the fool thinks it's like hon\-ey. But when it ripens, then the fool suffers grief.
\end{para}

\begin{para}
    \textsuperscript{70}\thinspace{}E\-ven if a fool, month af\-ter month, eats his food like an as\-cet\-ic with the tip of a blade of grass, he is not worth e\-ven one six\-teenth of those who have un\-der\-stood the truth.
\end{para}

\begin{para}
    \textsuperscript{71}\thinspace{}An evil deed, like fresh\-ly drawn milk, does not sour im\-me\-di\-ate\-ly; it follows the fool, smoldering, like fire cov\-ered by ash\-es.
\end{para}

\begin{para}
    \textsuperscript{73}\thinspace{}The fool wishes for a false rep\-u\-ta\-tion, for prec\-e\-dence a\-mong monks, for au\-thor\-i\-ty in the monasteries, for hon\-or a\-mong oth\-er peo\-ple.
    \textsuperscript{74}\thinspace{}“May both laymen and monks think that I did it. Let them o\-bey me in eve\-ry\-thing, great and small.” Such is the mind of the fool, and thus his de\-sire and pride in\-crease.
\end{para}

\begin{para}
    \textsuperscript{75}\thinspace{}One road leads to wealth, an\-oth\-er road leads to sal\-va\-tion. If the monk, a dis\-ci\-ple of Bud\-dha, has learn\-ed this, he will not yearn for hon\-or, he will strive for sep\-a\-ra\-tion from the world.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{The Wise}\addcontentsline{toc}{section}{The Wise}
\begin{para}
    \textsuperscript{76}\thinspace{}If you find a wise man who points out faults, and who reproves, fol\-low that wise man as one would fol\-low a guide to hid\-den treas\-ure. It will be bet\-ter, not worse, for those who fol\-low him.
    \textsuperscript{77}\thinspace{}Let him ad\-mon\-ish, let him teach, let him re\-strain you from what is im\-prop\-er. The good will love him. The bad will hate him.
\end{para}

\begin{para}
    \textsuperscript{78}\thinspace{}Don't have evildoers as companions, don't have vile peo\-ple as friends. Have vir\-tu\-ous peo\-ple as companions, have the best as friends.
\end{para}

\begin{para}
    \textsuperscript{79}\thinspace{}He who drinks in the law lives hap\-pi\-ly with a se\-rene mind. The wise man rejoices al\-ways in the law, as preached by the no\-ble ones.
\end{para}

\begin{para}
    \textsuperscript{80}\thinspace{}Irrigators guide the wa\-ter. Fletchers straight\-en the ar\-row. Carpenters shape the wood. The wise con\-trol them\-selves.
\end{para}

\begin{para}
    \textsuperscript{81}\thinspace{}Just as a sol\-id rock is not shak\-en by the wind, the wise don't fal\-ter a\-midst blame or praise.
    \textsuperscript{82}\thinspace{}Af\-ter they have listened to the law, the wise be\-come se\-rene, like a deep, smooth, and still lake.
\end{para}

\begin{para}
    \textsuperscript{83}\thinspace{}The good keep walk\-ing what\-ev\-er happens. The good don't prat\-tle, long\-ing for pleas\-ure. Wheth\-er touched by hap\-pi\-ness or sor\-row, the wise nev\-er ap\-pear e\-lat\-ed or de\-pressed.
\end{para}

\begin{para}
    \textsuperscript{84}\thinspace{}A man is good, wise, and vir\-tu\-ous if he does no wrong for his own sa\-ke or for the sa\-ke of others; if he wishes nei\-ther for wealth, a son, or a king\-dom; and if he does\-n't wish for suc\-cess by un\-fair means.
\end{para}

\begin{para}
    \textsuperscript{85}\thinspace{}Few a\-mong men ar\-rive at the oth\-er shore. The rest run up and down on this shore.
    \textsuperscript{86}\thinspace{}But those who fol\-low the law, when it has been well taught to them, will ar\-rive at the oth\-er shore hav\-ing crossed the do\-min\-ion of death, how\-ev\-er dif\-fi\-cult to over\-come.
\end{para}

\begin{para}
    \textsuperscript{87}\thinspace{}Let a wise man leave the dark way, and fol\-low the bright path. Af\-ter go\-ing from his home to home\-less\-ness, let him yearn for de\-light in de\-tach\-ment, so dif\-fi\-cult to en\-joy.
    \textsuperscript{88}\thinspace{}Leav\-ing all pleasures be\-hind, and call\-ing noth\-ing his own, let the wise man cleanse him\-self of that which defiles the mind.
    \textsuperscript{89}\thinspace{}Those whose minds are well grounded in the factors of en\-light\-en\-ment, who with\-out cling\-ing to any\-thing, re\-joice in free\-dom from at\-tach\-ment, whose appetites have been conquered, and who are full of light, are free e\-ven in this world.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{The Saint}\addcontentsline{toc}{section}{The Saint}
\begin{para}
    \textsuperscript{91}\thinspace{}The mind\-ful strive dil\-i\-gent\-ly. They're not at\-tached to any home. Like swans who leave a lake, they leave their house and home.
\end{para}

\begin{para}
    \textsuperscript{92}\thinspace{}Like birds in the air, the path of the saint is hard to trace. They have no rich\-es, are wise re\-gard\-ing food, and their ob\-jec\-tive is emp\-ti\-ness and free\-dom.
\end{para}

\begin{para}
    \textsuperscript{94}\thinspace{}E\-ven the gods a\-dore him whose sen\-ses have been sub\-dued, like hors\-es well trained by a char\-i\-ot\-eer, who's free from pride, and free from appetites.
    \textsuperscript{95}\thinspace{}Like the earth, he's not pro\-voked. He's firm like a pil\-lar. He's like a pure lake with\-out mud. No new births are in store for him.
\end{para}

\begin{para}
    \textsuperscript{96}\thinspace{}His mind is calm, and calm are his words and deeds. By true knowl\-edge he has obtained free\-dom, and thus be\-come tran\-quil.
\end{para}

\begin{para}
    \textsuperscript{98}\thinspace{}In a vil\-lage or in a for\-est, in a val\-ley or on a hill, wher\-ev\-er saints dwell, that place is de\-light\-ful.
    \textsuperscript{99}\thinspace{}De\-light\-ful are the forests in which the crowds find no pleas\-ure. Those with\-out pas\-sion will find de\-light there, for they don't seek pleas\-ure.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Thousands}\addcontentsline{toc}{section}{Thousands}
\begin{para}
    \textsuperscript{100}\thinspace{}Bet\-ter than a speech of a thou\-sand vain words is one thought\-ful word, hear\-ing which brings peace.
    \textsuperscript{101}\thinspace{}Bet\-ter than a po\-em of a thou\-sand vain verses is one thought\-ful verse, hear\-ing which brings peace.
\end{para}

\begin{para}
    \textsuperscript{103}\thinspace{}Greater than he who conquers a thou\-sand times a thou\-sand men in bat\-tle, is he who conquerors just one---him\-self.
    \textsuperscript{104}\thinspace{}It is far bet\-ter to con\-quer one\-self than to con\-quer others.
    \textsuperscript{105}\thinspace{}For a man who has conquered him\-self and al\-ways lives with self-control, not e\-ven a god, an an\-gel, or a de\-mon with a god can change that vic\-to\-ry in\-to de\-feat.
\end{para}

\begin{para}
    \textsuperscript{106}\thinspace{}A man could sac\-ri\-fice by the thousands for month af\-ter month for a hun\-dred years, or he could pay hom\-age for one mo\-ment to a man liv\-ing in wis\-dom. Bet\-ter is that hom\-age than sacrifices for a cen\-tu\-ry.
    \textsuperscript{107}\thinspace{}A man could tend the sac\-ri\-fi\-cial fire in the for\-est for a hun\-dred years, or he could pay hom\-age for one mo\-ment to a man liv\-ing in wis\-dom. Bet\-ter is that hom\-age than sacrifices for a cen\-tu\-ry.
    \textsuperscript{108}\thinspace{}In this world, a man may make sacrifices and oblations for a whole year in or\-der to gain mer\-it, yet all of it is not worth a quar\-ter of the mer\-it gained by re\-ver\-ing the right\-eous.
\end{para}

\begin{para}
    \textsuperscript{109}\thinspace{}For him who al\-ways respects and hon\-ors the wise, four things will in\-crease: life, beau\-ty, hap\-pi\-ness, pow\-er.
\end{para}

\begin{para}
    \textsuperscript{110}\thinspace{}Bet\-ter than a hun\-dred years lived im\-mor\-al and un\-re\-strained, is one day lived vir\-tu\-ous and reflecting.
    \textsuperscript{111}\thinspace{}Bet\-ter than a hun\-dred years lived ig\-no\-rant and un\-re\-strained, is one day lived wise and reflecting.
    \textsuperscript{112}\thinspace{}Bet\-ter than a hun\-dred years i\-dle and weak, is one day lived en\-er\-get\-ic and firm.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Evil}\addcontentsline{toc}{section}{Evil}
\begin{para}
    \textsuperscript{116}\thinspace{}Has\-ten to do good. Keep your mind from evil. If a man is slow to do good, his mind delights in evil.
\end{para}

\begin{para}
    \textsuperscript{117}\thinspace{}If a man does what is evil, let him not do it again and again. Let him not de\-light in it, for ac\-cu\-mu\-lat\-ing evil brings pain.
    \textsuperscript{118}\thinspace{}If a man does what is good, let him do it again and again. Let him de\-light in it, for ac\-cu\-mu\-lat\-ing good brings joy.
\end{para}

\begin{para}
    \textsuperscript{119}\thinspace{}E\-ven an evil\-do\-er may be hap\-py be\-fore his evil deed ripens, but when it ripens, he suffers.
    \textsuperscript{120}\thinspace{}E\-ven a good man may suf\-fer be\-fore his good deed ripens, but when it ripens, he is hap\-py.
\end{para}

\begin{para}
    \textsuperscript{121}\thinspace{}Let no man think light\-ly of evil, say\-ing in his heart, “It won't a\-mount to much.” As a wa\-ter pot is filled by the fall\-ing of wa\-ter drops, so the fool becomes full of evil, e\-ven if he gathers it lit\-tle by lit\-tle.
    \textsuperscript{122}\thinspace{}Let no man think light\-ly of good, say\-ing in his heart, “It won't a\-mount to much.” As a wa\-ter pot is filled by the fall\-ing of wa\-ter drops, so the wise man becomes full of good, e\-ven if he gathers it lit\-tle by lit\-tle.
\end{para}

\begin{para}
    \textsuperscript{123}\thinspace{}Let a man a\-void evil deeds, as a wealth\-y mer\-chant with few companions avoids a dan\-ger\-ous road; as a man who loves life avoids poi\-son.
\end{para}

\begin{para}
    \textsuperscript{125}\thinspace{}If a man offends a harm\-less, pure, and in\-no\-cent per\-son, the evil falls back up\-on that fool, like dust thrown against the wind.
\end{para}

\begin{para}
    \textsuperscript{127}\thinspace{}Not in the sky, not in the mid\-dle of the sea, not in a cleft of the mountains, is there a spot to be found where a man can es\-cape from the con\-se\-quenc\-es of his evil deed.
    \textsuperscript{128}\thinspace{}Not in the sky, not in the mid\-dle of the sea, not in a cleft of the mountains, is there a spot to be found where a man will not be over\-come by death.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Punishment}\addcontentsline{toc}{section}{Punishment}
\begin{para}
    \textsuperscript{129}\thinspace{}Eve\-ry\-one trem\-bles at pun\-ish\-ment, eve\-ry\-one fears death. Put\-ting your\-self in his place, don't kill or cause others to kill.
    \textsuperscript{130}\thinspace{}Eve\-ry\-one trem\-bles at pun\-ish\-ment, eve\-ry\-one loves his own life. Put\-ting your\-self in his place, don't kill or cause others to kill.
\end{para}

\begin{para}
    \textsuperscript{131}\thinspace{}He who, seek\-ing hap\-pi\-ness, punishes beings who al\-so de\-sire hap\-pi\-ness, won't find hap\-pi\-ness af\-ter death.
    \textsuperscript{132}\thinspace{}He who, seek\-ing hap\-pi\-ness, does\-n't pun\-ish beings who al\-so de\-sire hap\-pi\-ness, will find hap\-pi\-ness af\-ter death.
\end{para}

\begin{para}
    \textsuperscript{133}\thinspace{}Speak harsh\-ly to no one, or those who are spo\-ken to will an\-swer you in the same way. Con\-ten\-tious speech hurts, and re\-tal\-i\-a\-tion will come to you.
    \textsuperscript{134}\thinspace{}If your mind is still, like a bro\-ken gong, you have attained sal\-va\-tion, for there's no con\-ten\-tion in you.
\end{para}

\begin{para}
    \textsuperscript{135}\thinspace{}As a cow\-herd drives cows to pas\-ture with a staff, so age\-ing and death drive the life of men.
    \textsuperscript{136}\thinspace{}A fool does\-n't re\-al\-ize when he commits evil, but he suffers for those deeds as if burned by fire.
\end{para}

\begin{para}
    \textsuperscript{137}\thinspace{}He who inflicts pain on the in\-no\-cent and un\-armed will soon suf\-fer in one of these ten ways:
    \textsuperscript{138}\thinspace{}se\-vere pain, im\-pov\-er\-ish\-ment, bod\-i\-ly in\-ju\-ry, se\-ri\-ous ill\-ness, men\-tal ill\-ness,
    \textsuperscript{139}\thinspace{}trou\-ble from the gov\-ern\-ment, a fear\-ful ac\-cu\-sa\-tion, loss of relatives, loss of wealth,
    \textsuperscript{140}\thinspace{}or fire will de\-stroy his hous\-es. When his bod\-y is destroyed, the fool will go to hell.
\end{para}

\begin{para}
    \textsuperscript{141}\thinspace{}Nei\-ther na\-ked\-ness, nor mat\-ted hair, nor dirt, nor fasting, nor ly\-ing on the earth, nor rub\-bing with dust, nor sit\-ting mo\-tion\-less, can pu\-ri\-fy a mor\-tal who has not over\-come doubt.
    \textsuperscript{142}\thinspace{}He who exercises tran\-quil\-i\-ty, is qui\-et, sub\-dued, restrained, chaste, and has set a\-side vi\-o\-lence to\-wards all others, he in\-deed is a ho\-ly man, an as\-cet\-ic, a monk, e\-ven if dressed in fi\-ne ap\-par\-el.
\end{para}

\begin{para}
    \textsuperscript{143}\thinspace{}Who in this world is so restrained by con\-science that he needs no re\-proof, as a fi\-ne horse needs no whip?
    \textsuperscript{144}\thinspace{}Like a fi\-ne horse when touched by the whip, be ac\-tive and live\-ly. By faith, by vir\-tue, by ef\-fort, by med\-i\-ta\-tion, by dis\-cern\-ing the truth, by per\-fect knowl\-edge and con\-duct, by mind\-ful\-ness, you will leave this pain be\-hind.
\end{para}

\begin{para}
    \textsuperscript{145}\thinspace{}Irrigators guide the wa\-ter. Fletchers straight\-en the ar\-row. Carpenters shape the wood. Good peo\-ple fash\-ion them\-selves.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Old Age}\addcontentsline{toc}{section}{Old Age}
\begin{para}
    \textsuperscript{146}\thinspace{}Why is there laugh\-ter and joy when this world is al\-ways burn\-ing? Surrounded by dark\-ness, don't you look for a light?
\end{para}

\begin{para}
    \textsuperscript{147}\thinspace{}Look at this bod\-y---a paint\-ed im\-age, cov\-ered with festering wounds, sick\-ly, full of man\-y thoughts---nei\-ther last\-ing nor sta\-ble.
    \textsuperscript{148}\thinspace{}This bod\-y is worn out, full of sick\-ness, and frail. This heap of cor\-rup\-tion breaks to pieces, for life in\-deed ends with death.
    \textsuperscript{149}\thinspace{}These white bones are like gourds thrown a\-way in the au\-tumn. What pleas\-ure is there in see\-ing them?
\end{para}

\begin{para}
    \textsuperscript{150}\thinspace{}A cit\-y made of bones is cov\-ered with flesh and blood. Liv\-ing in it are old age and death, pride and de\-ceit.
    \textsuperscript{151}\thinspace{}The beau\-ti\-ful chariots of kings get worn out and so does the bod\-y. But the vir\-tue of good peo\-ple does\-n't age: thus do the good say a\-mong them\-selves.
\end{para}

\begin{para}
    \textsuperscript{152}\thinspace{}A man who has learn\-ed lit\-tle grows old like an ox. His bod\-y grows, but his wis\-dom does\-n't.
\end{para}

\begin{para}
    \textsuperscript{153}\thinspace{}I roamed through man\-y births, with\-out re\-ward and with\-out rest, looking for the build\-er of this house. Pain\-ful is birth again and again.
    \textsuperscript{154}\thinspace{}But now, build\-er of this house, I have seen you. You won't build this house again. All your rafters are bro\-ken, your ridge pole is dismantled. My mind has attained the ex\-tinc\-tion of all desires.
\end{para}

\begin{para}
    \textsuperscript{155}\thinspace{}Those who, in their youth, have\-n't lived the ho\-ly life nor gained wealth per\-ish like old herons in a lake with\-out fish.
    \textsuperscript{156}\thinspace{}They lie sighing over the past like ar\-rows misfired from a bow.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Self}\addcontentsline{toc}{section}{Self}
\begin{para}
    \textsuperscript{157}\thinspace{}If a man holds him\-self dear, let him watch him\-self care\-ful\-ly. Let a wise man keep vig\-il dur\-ing one of the three watches of the night.
\end{para}

\begin{para}
    \textsuperscript{158}\thinspace{}Let each man first di\-rect him\-self to what is prop\-er, then let him teach others. Thus a wise man will not be reproached.
\end{para}

\begin{para}
    \textsuperscript{159}\thinspace{}If a man makes him\-self as he teaches others to be, then, be\-ing him\-self well tamed, he may tame others. One's self is in\-deed dif\-fi\-cult to tame.
\end{para}

\begin{para}
    \textsuperscript{160}\thinspace{}A man is tru\-ly his own pro\-tec\-tor. Who else could be the pro\-tec\-tor? With self well con\-trolled, one finds a pro\-tec\-tor such as few can find.
\end{para}

\begin{para}
    \textsuperscript{161}\thinspace{}The evil a fool\-ish man does by him\-self, self-born, self-created, grinds him down, as a dia\-mond breaks a hard gem.
    \textsuperscript{162}\thinspace{}When over\-spread with great wick\-ed\-ness, like a tree cov\-ered by vines, a man brings him\-self down to where his en\-e\-my wishes.
    \textsuperscript{163}\thinspace{}Evil deeds, which harm our\-selves, are eas\-y to do. Good and ben\-e\-fi\-cial deeds are ver\-y dif\-fi\-cult to do.
\end{para}

\begin{para}
    \textsuperscript{165}\thinspace{}Evil is done by one\-self, by one\-self one is de\-filed. Evil is left un\-done by one\-self, by one\-self is one pu\-ri\-fied. Pu\-ri\-ty and im\-pu\-ri\-ty are one's own do\-ing. No one can pu\-ri\-fy an\-oth\-er.
\end{para}

\begin{para}
    \textsuperscript{166}\thinspace{}Let no man sac\-ri\-fice his own wel\-fare for that of an\-oth\-er, how\-ev\-er great. Let a man, af\-ter dis\-cern\-ing his own wel\-fare, be al\-ways at\-ten\-tive to it.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{The World}\addcontentsline{toc}{section}{The World}
\begin{para}
    \textsuperscript{167}\thinspace{}Don't fol\-low vul\-gar ways. Don't live with a\-ban\-don. Don't fol\-low false doc\-trine. Don't be a friend of the world.
\end{para}

\begin{para}
    \textsuperscript{168}\thinspace{}Wake up! Don't be i\-dle. Live a vir\-tu\-ous life. The vir\-tu\-ous sleeps with ease in this world and the next.
    \textsuperscript{169}\thinspace{}Live a vir\-tu\-ous life. Don't live a life of sin. The vir\-tu\-ous sleeps with ease in this world and the next.
\end{para}

\begin{para}
    \textsuperscript{170}\thinspace{}The king of death does\-n't see him who looks up\-on the world as a bub\-ble, as a mi\-rage.
    \textsuperscript{171}\thinspace{}Come look at this world, dec\-o\-rat\-ed like a roy\-al char\-i\-ot. Fools are im\-mersed in it, but the wise don't cling to it.
\end{para}

\begin{para}
    \textsuperscript{172}\thinspace{}He who was reck\-less and then becomes mind\-ful, brightens up this world like the moon when freed from clouds.
    \textsuperscript{173}\thinspace{}He who replaces evil deeds with good deeds, brightens up this world, like the moon when freed from clouds.
\end{para}

\begin{para}
    \textsuperscript{174}\thinspace{}In this dark world, on\-ly a few can see. On\-ly a few, like birds es\-cap\-ing from the net, go to heav\-en.
    \textsuperscript{175}\thinspace{}Swans fly on the path of the sun, through the air by their won\-der\-ful pow\-er. The wise flee from the world af\-ter defeating the temp\-ter and his armies.
\end{para}

\begin{para}
    \textsuperscript{176}\thinspace{}A man who transgresses the one law, speaks lies, and scoffs at the world be\-yond is ca\-pa\-ble of any evil.
\end{para}

\begin{para}
    \textsuperscript{177}\thinspace{}Misers don't go to the world of the gods; nei\-ther do fools praise gen\-er\-os\-i\-ty. A wise man rejoices in gen\-er\-os\-i\-ty, and through it becomes bless\-ed in the world be\-yond.
\end{para}

\begin{para}
    \textsuperscript{178}\thinspace{}The first step to\-wards ho\-li\-ness is bet\-ter than sov\-er\-eign\-ty over the earth, bet\-ter than go\-ing to heav\-en, and bet\-ter than be\-ing lord over all the worlds.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{The Awakened}\addcontentsline{toc}{section}{The Awakened}
\begin{para}
    \textsuperscript{181}\thinspace{}E\-ven the gods en\-vy those who are awakened, who are de\-vot\-ed to med\-i\-ta\-tion, who are wise, and who de\-light in the calm of re\-nun\-ci\-a\-tion.
\end{para}

\begin{para}
    \textsuperscript{182}\thinspace{}It's dif\-fi\-cult to ob\-tain a hu\-man birth, dif\-fi\-cult is the life of mortals. It's dif\-fi\-cult to ob\-tain the chance to hear the true law, dif\-fi\-cult is the birth of the awakened.
\end{para}

\begin{para}
    \textsuperscript{183}\thinspace{}To a\-void all evil, to cul\-ti\-vate good, and to pu\-ri\-fy one's mind, that is the teach\-ing of the awakened.
    \textsuperscript{184}\thinspace{}Pa\-tient en\-dur\-ance is the highest aus\-ter\-i\-ty. “Sal\-va\-tion is su\-preme”, say the awakened. He is no monk who injures others. He is no as\-cet\-ic who insults others.
    \textsuperscript{185}\thinspace{}Not to re\-vile, not to in\-jure, to live restrained un\-der the law, to be mod\-er\-ate in food, to live in sol\-i\-tude, and to dwell on the highest thoughts: this is the teach\-ing of the awakened.
\end{para}

\begin{para}
    \textsuperscript{186}\thinspace{}There's no way to sat\-is\-fy lusts, e\-ven with a show\-er of gold coins. He is wise who knows that lusts give lit\-tle sat\-is\-fac\-tion and much pain.
    \textsuperscript{187}\thinspace{}E\-ven in heav\-en\-ly pleasures the wise finds no de\-light. The dis\-ci\-ple who is ful\-ly awakened delights on\-ly in the de\-struc\-tion of crav\-ing.
\end{para}

\begin{para}
    \textsuperscript{188}\thinspace{}When threatened with dan\-ger, men go to man\-y a ref\-uge: to mountains and forests, to groves and shrines.
    \textsuperscript{189}\thinspace{}But that is not a se\-cure ref\-uge, that is not the best ref\-uge. A man is not delivered from all pain af\-ter go\-ing to that ref\-uge.
\end{para}

\begin{para}
    \textsuperscript{190}\thinspace{}He who takes ref\-uge with the law, the awakened, and their or\-der, sees with clear un\-der\-stand\-ing the four ho\-ly truths:
    \textsuperscript{191}\thinspace{}pain, the cause of pain, the de\-struc\-tion of pain, and the ho\-ly eight\-fold path that leads to the quieting of pain;
    \textsuperscript{192}\thinspace{}That is a se\-cure ref\-uge, that is the best ref\-uge. Af\-ter go\-ing to that ref\-uge, a man is delivered from all pain.
\end{para}

\begin{para}
    \textsuperscript{193}\thinspace{}An awakened man is hard to find. He's not born eve\-ry\-where. Wher\-ev\-er such a wise man is born, that clan prospers.
\end{para}

\begin{para}
    \textsuperscript{194}\thinspace{}Bless\-ed is the arising of awakened ones, bless\-ed is the teach\-ing of the true law, bless\-ed is har\-mo\-ny in the or\-der, bless\-ed is the de\-vo\-tion of those who are at peace.
\end{para}

\begin{para}
    \textsuperscript{195}\thinspace{}He who reveres those who de\-serve rev\-er\-ence---the awakened ones or their disciples, who have over\-come all obstacles, sor\-row, and lam\-en\-ta\-tion,
    \textsuperscript{196}\thinspace{}who found de\-liv\-er\-ance and know no fear---his mer\-it can nev\-er be meas\-ured.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Happiness}\addcontentsline{toc}{section}{Happiness}
\begin{para}
    \textsuperscript{197}\thinspace{}We live hap\-pi\-ly, not hat\-ing those who hate us. When a\-mong men who hate us, we dwell with\-out ha\-tred.
    \textsuperscript{198}\thinspace{}We live hap\-pi\-ly, with\-out ill\-ness a\-mong the ill. When a\-mong men who are ill, we dwell with\-out ill\-ness.
    \textsuperscript{199}\thinspace{}We live hap\-pi\-ly, with\-out greed a\-mong the greed\-y. When a\-mong men who are greed\-y, we dwell with\-out greed.
    \textsuperscript{200}\thinspace{}We live hap\-pi\-ly, we who own noth\-ing.
\end{para}

\begin{para}
    \textsuperscript{201}\thinspace{}Vic\-to\-ry breeds ha\-tred, for the conquered live in pain. The con\-tent\-ed is hap\-py, hav\-ing giv\-en up both vic\-to\-ry and de\-feat.
\end{para}

\begin{para}
    \textsuperscript{202}\thinspace{}There is no fire like lust, no loss like ha\-tred, no pain like this ex\-ist\-ence, and no hap\-pi\-ness higher than peace.
\end{para}

\begin{para}
    \textsuperscript{203}\thinspace{}Hun\-ger is the worst dis\-ease, this ex\-ist\-ence the great\-est pain. If one tru\-ly knows this, he realizes sal\-va\-tion, the highest hap\-pi\-ness.
    \textsuperscript{204}\thinspace{}Health is the great\-est gift, con\-tent\-ment the great\-est wealth. Trust is the best kins\-man, sal\-va\-tion the highest hap\-pi\-ness.
\end{para}

\begin{para}
    \textsuperscript{205}\thinspace{}He who has savored the taste of sol\-i\-tude and tran\-quil\-i\-ty is free from fear and evil, refreshed with the nour\-ish\-ment of the law.
\end{para}

\begin{para}
    \textsuperscript{206}\thinspace{}The sight of no\-ble ones is good. To live with them is al\-ways hap\-pi\-ness. If a man does\-n't see fools, he will be tru\-ly hap\-py.
\end{para}

\begin{para}
    \textsuperscript{207}\thinspace{}He who walks with fools suffers for a long time. Com\-pa\-ny with fools, as with an en\-e\-my, is al\-ways pain\-ful. Com\-pa\-ny with the wise is a pleas\-ure, like re\-un\-ion with fam\-i\-ly.
    \textsuperscript{208}\thinspace{}There\-fore, fol\-low the wise, the in\-tel\-li\-gent, the learn\-ed, the stead\-fast, the du\-ti\-ful, the no\-ble. Fol\-low a good and dis\-cern\-ing man, as the moon follows the path of the stars.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Pleasure}\addcontentsline{toc}{section}{Pleasure}
\begin{para}
    \textsuperscript{209}\thinspace{}He who gives him\-self to van\-i\-ty, in\-stead of to med\-i\-ta\-tion, who forgets the goal and grasps at pleas\-ure, will come to en\-vy him who has exerted him\-self.
\end{para}

\begin{para}
    \textsuperscript{210}\thinspace{}Don't seek for what's pleas\-ant or un\-pleas\-ant, for it's pain not to see the pleas\-ant, and it's pain to see what's un\-pleas\-ant.
\end{para}

\begin{para}
    \textsuperscript{211}\thinspace{}There\-fore, let no man hold any\-thing dear, for loss of the dear is pain\-ful. Those who hold noth\-ing dear and hate noth\-ing have no fetters.
\end{para}

\begin{para}
    \textsuperscript{212}\thinspace{}From pleas\-ure co\-mes grief, from pleas\-ure co\-mes fear. He who is free from pleas\-ure knows nei\-ther grief nor fear.
    \textsuperscript{213}\thinspace{}From af\-fec\-tion co\-mes grief, from af\-fec\-tion co\-mes fear. He who is free from af\-fec\-tion knows nei\-ther grief nor fear.
    \textsuperscript{214}\thinspace{}From at\-tach\-ment co\-mes grief, from at\-tach\-ment co\-mes fear. He who is free from at\-tach\-ment knows nei\-ther grief nor fear.
    \textsuperscript{215}\thinspace{}From lust co\-mes grief, from lust co\-mes fear. He who is free from lust knows nei\-ther grief nor fear.
    \textsuperscript{216}\thinspace{}From crav\-ing co\-mes grief, from crav\-ing co\-mes fear. He who is free from crav\-ing knows nei\-ther grief nor fear.
\end{para}

\begin{para}
    \textsuperscript{217}\thinspace{}He who has vir\-tue and in\-sight, who is just, who speaks the truth, and does his own busi\-ness, him the world holds dear.
\end{para}

\begin{para}
    \textsuperscript{219}\thinspace{}Fam\-i\-ly, friends, and well-wishers wel\-come a man long ab\-sent who returns safe\-ly from a\-far.
    \textsuperscript{220}\thinspace{}As fam\-i\-ly receives a friend on his re\-turn, so do his good works wel\-come him who has done good, and has gone from this world to the next.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Anger}\addcontentsline{toc}{section}{Anger}
\begin{para}
    \textsuperscript{221}\thinspace{}Let a man leave an\-ger, for\-sake pride, and over\-come all bond\-age. No suf\-fer\-ing befalls the man who is de\-tached from name and form, and who calls noth\-ing his own.
\end{para}

\begin{para}
    \textsuperscript{223}\thinspace{}Over\-come an\-ger with love, evil with good, greed with gen\-er\-os\-i\-ty, and lies with truth.
    \textsuperscript{224}\thinspace{}Speak the truth, don't yield to an\-ger, and give, if asked, e\-ven if it's on\-ly a lit\-tle. By these three steps you will go near the gods.
\end{para}

\begin{para}
    \textsuperscript{225}\thinspace{}The sages who harm no one, and al\-ways con\-trol their bod\-y, go to the un\-change\-a\-ble state, where they suf\-fer no more.
    \textsuperscript{226}\thinspace{}Those who are al\-ways vig\-i\-lant, stud\-y\-ing day and night, in\-tent on sal\-va\-tion, their passions fade a\-way.
\end{para}

\begin{para}
    \textsuperscript{227}\thinspace{}This is an old say\-ing, not just of to\-day: “They blame him who is si\-lent, they blame him who speaks much, they blame him who speaks lit\-tle.” There is none on earth who escapes blame.
    \textsuperscript{228}\thinspace{}There nev\-er was, there nev\-er will be, nor is there now, a man who is al\-ways blamed or al\-ways praised.
\end{para}

\begin{para}
    \textsuperscript{229}\thinspace{}If the knowl\-edge\-a\-ble praise him, hav\-ing ob\-served him day af\-ter day to be with\-out blem\-ish, wise, and rich in knowl\-edge and vir\-tue,
    \textsuperscript{230}\thinspace{}who can blame him? E\-ven the gods praise him. E\-ven the Cre\-a\-tor praises him.
\end{para}

\begin{para}
    \textsuperscript{231}\thinspace{}With your bod\-y, tongue, and mind: be\-ware of an\-ger, be restrained, leave its sins, and with it prac\-tise vir\-tue.
    \textsuperscript{234}\thinspace{}The wise who con\-trol their bod\-y, their tongue, and their mind, are well con\-trolled in\-deed.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Impurity}\addcontentsline{toc}{section}{Impurity}
\begin{para}
    \textsuperscript{235}\thinspace{}You are now like a withered leaf, the messengers of death stand near\-by. You stand at the door of de\-par\-ture, yet have no pro\-vi\-sion for your jour\-ney.
    \textsuperscript{237}\thinspace{}Your life has come to an end, you are near death, there is no rest\-ing place for you on the road, yet you have no pro\-vi\-sion for your jour\-ney.
    \textsuperscript{238}\thinspace{}Make an is\-land for your\-self, work hard, be wise. When your impurities are blown a\-way, and you are free from guilt, you will not en\-ter again in\-to birth and de\-cay.
\end{para}

\begin{para}
    \textsuperscript{239}\thinspace{}Let a wise man re\-move his own impurities, as a smith removes impurities from sil\-ver one by one, lit\-tle by lit\-tle, and mo\-ment by mo\-ment.
    \textsuperscript{240}\thinspace{}Just as rust destroys the i\-ron from which it springs, so do a transgressor's deeds lead him to an evil des\-ti\-na\-tion.
\end{para}

\begin{para}
    \textsuperscript{241}\thinspace{}The im\-pu\-ri\-ty of prayers is not re\-cit\-ing them. The im\-pu\-ri\-ty of hous\-es is not repairing them. The im\-pu\-ri\-ty of beau\-ty is sloth. The im\-pu\-ri\-ty of a watch\-man is mind\-less\-ness.
    \textsuperscript{242}\thinspace{}Sex\-u\-al mis\-con\-duct is the im\-pu\-ri\-ty of wom\-an, stin\-gi\-ness the im\-pu\-ri\-ty of a do\-nor. All evil ways are impurities, both in this world and in the next.
    \textsuperscript{243}\thinspace{}But there is an im\-pu\-ri\-ty worse than all others: ig\-no\-rance. Throw off that im\-pu\-ri\-ty to be\-come pure.
\end{para}

\begin{para}
    \textsuperscript{246}\thinspace{}He who destroys life, speaks lies, takes what is not giv\-en, goes to an\-oth\-er man's wife,
    \textsuperscript{247}\thinspace{}and gives him\-self to in\-tox\-i\-cat\-ing liq\-uor, he digs up his own root here in this world.
    \textsuperscript{248}\thinspace{}Know this, good man, that evil is hard to con\-trol. Do not let greed and vi\-ce bring you to prolonged grief.
\end{para}

\begin{para}
    \textsuperscript{249}\thinspace{}Peo\-ple give ac\-cord\-ing to their faith or con\-vic\-tion. If a man worries a\-bout the food and drink giv\-en to others, he will find no rest by day or night.
    \textsuperscript{250}\thinspace{}He in whom that feel\-ing is cut down, uprooted, and destroyed finds rest by day and night.
\end{para}

\begin{para}
    \textsuperscript{251}\thinspace{}There's no fire like pas\-sion, no grip like ha\-tred, no snare like de\-lu\-sion, no riv\-er like crav\-ing.
\end{para}

\begin{para}
    \textsuperscript{252}\thinspace{}It's eas\-y to see the faults of others, but dif\-fi\-cult to see one's own. A man winnows the faults of others like chaff, but hides his own like a cheat hides an un\-luck\-y throw of the dice.
    \textsuperscript{253}\thinspace{}If a man fo\-cus\-es on the errors of others, and is al\-ways in\-clined to find fault, his own faults will grow, and he is far from destroying them.
\end{para}

\begin{para}
    \textsuperscript{254}\thinspace{}The world delights in van\-i\-ty, but the awakened are free from it.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{The Just}\addcontentsline{toc}{section}{The Just}
\begin{para}
    \textsuperscript{256}\thinspace{}A man does\-n't be\-come just by pass\-ing ar\-bi\-trar\-y judge\-ment. No, he's called just who weighs both right and wrong,
    \textsuperscript{257}\thinspace{}who judges im\-par\-tial\-ly and with\-out hur\-ry, in line with law, guarding the law, guard\-ed by the law, and who is in\-tel\-li\-gent.
\end{para}

\begin{para}
    \textsuperscript{258}\thinspace{}A man is\-n't wise just be\-cause he talks a lot. He's called wise who is pa\-tient, and free from hos\-til\-i\-ty and fear.
\end{para}

\begin{para}
    \textsuperscript{259}\thinspace{}A man does\-n't up\-hold the law just be\-cause he talks a lot. He who has learn\-ed on\-ly a lit\-tle of the law, yet comprehends it and nev\-er neglects it, he upholds the law.
\end{para}

\begin{para}
    \textsuperscript{260}\thinspace{}A man is\-n't an el\-der just be\-cause his hair is grey. He may be ad\-vanced in years, but is called “one grown old in vain.”
    \textsuperscript{261}\thinspace{}He in whom there is truth, vir\-tue, love, re\-straint, self con\-trol, who is free from im\-pu\-ri\-ty, and is wise, he is called an el\-der.
\end{para}

\begin{para}
    \textsuperscript{262}\thinspace{}El\-o\-quence or phys\-i\-cal beau\-ty don't make an en\-vi\-ous, greed\-y, and dis\-hon\-est man re\-spect\-a\-ble.
    \textsuperscript{263}\thinspace{}He's called re\-spect\-a\-ble when all these are cut down, uprooted, and destroyed, when he's free from ha\-tred, and is wise.
\end{para}

\begin{para}
    \textsuperscript{264}\thinspace{}An un\-dis\-ci\-plined man who speaks lies does\-n't be\-come a monk just by shav\-ing his head. How can he be a monk who is held cap\-tive by de\-sire and greed?
    \textsuperscript{265}\thinspace{}He who quiets evil, both small and large, is called a monk, be\-cause he has over\-come all evil.
\end{para}

\begin{para}
    \textsuperscript{266}\thinspace{}A man is\-n't a monk just be\-cause he asks for alms. He who adopts the whole law is a monk.
    \textsuperscript{267}\thinspace{}He who is a\-bove both good and evil, who is chaste, and who goes through the world with un\-der\-stand\-ing, he in\-deed is called a monk.
\end{para}

\begin{para}
    \textsuperscript{268}\thinspace{}A fool\-ish and ig\-no\-rant man does\-n't be\-come a sage just by si\-lence. But the wise, as if hold\-ing the scales, chooses good and avoids evil.
    \textsuperscript{269}\thinspace{}For this rea\-son he is a sage. He who weighs both sides of this world is called a sage.
\end{para}

\begin{para}
    \textsuperscript{270}\thinspace{}A man does\-n't be\-come no\-ble by harming liv\-ing creatures. He's called no\-ble for harming no liv\-ing thing.
\end{para}

\begin{para}
    \textsuperscript{271}\thinspace{}Not by rules and practices, not by much learn\-ing, not by great con\-cen\-tra\-tion, not by life in se\-clu\-sion,
    \textsuperscript{272}\thinspace{}not by think\-ing, “I en\-joy the hap\-pi\-ness of re\-nun\-ci\-a\-tion which no world\-ling can know”, should the monk be con\-fi\-dent with\-out attaining the ex\-tinc\-tion of desires.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{The Path}\addcontentsline{toc}{section}{The Path}
\begin{para}
    \textsuperscript{273}\thinspace{}Of paths, the Eight\-fold is best. Of truths, the Four No\-ble are best. Of vir\-tues, de\-tach\-ment is best. Of men, he who has eyes to see is best.
\end{para}

\begin{para}
    \textsuperscript{274}\thinspace{}This is the on\-ly path, there is no oth\-er, to pu\-ri\-fy vi\-sion. Fol\-low this path and it will be\-wil\-der the temp\-ter.
    \textsuperscript{275}\thinspace{}Fol\-low\-ing this path will end pain. Once I un\-der\-stood how to re\-move thorns in the flesh, I preached this path.
\end{para}

\begin{para}
    \textsuperscript{276}\thinspace{}You your\-self must make an ef\-fort. Buddhas on\-ly show the way. The thought\-ful who fol\-low the path are freed from the tempter's bonds.
\end{para}

\begin{para}
    \textsuperscript{277}\thinspace{}“All cre\-at\-ed things are tem\-po\-rar\-y.” He who sees and knows this turns a\-way from pain. This is the path to pu\-ri\-ty.
\end{para}

\begin{para}
    \textsuperscript{278}\thinspace{}“All cre\-at\-ed things are pain\-ful.” He who sees and knows this turns a\-way from pain. This is the path to pu\-ri\-ty.
\end{para}

\begin{para}
    \textsuperscript{280}\thinspace{}He who does\-n't ex\-ert him\-self when it's time, who, though young and strong, is full of sloth, whose will and mind are weak, that la\-zy and i\-dle man will nev\-er find the path to knowl\-edge.
\end{para}

\begin{para}
    \textsuperscript{281}\thinspace{}Let a man watch his speech, be well restrained in mind, and nev\-er com\-mit wrong with his bod\-y. If a man purifies these three cours\-es of ac\-tion, he will a\-chieve the path taught by the wise.
\end{para}

\begin{para}
    \textsuperscript{282}\thinspace{}Zeal brings knowl\-edge. Lack of zeal loses knowl\-edge. Know\-ing these two paths of gain and loss, let a man con\-duct him\-self so that knowl\-edge may grow.
\end{para}

\begin{para}
    \textsuperscript{283}\thinspace{}Cut down the whole for\-est of de\-sire, not just one tree. Dan\-ger co\-mes from the for\-est. Once you have cut down this for\-est and its un\-der\-growth, you will be free.
\end{para}

\begin{para}
    \textsuperscript{284}\thinspace{}As long as the smallest crav\-ing of man for wom\-an is not destroyed, his mind is in bond\-age, as the suck\-ing calf to its moth\-er.
    \textsuperscript{285}\thinspace{}Cut out the love of self, like an au\-tumn lo\-tus with your hand. Cher\-ish the road of peace, as shown by the awakened.
\end{para}

\begin{para}
    \textsuperscript{286}\thinspace{}“I will live here dur\-ing the rains, and over here in win\-ter and sum\-mer.” Thus thinks the fool, un\-a\-ware of obstructions.
\end{para}

\begin{para}
    \textsuperscript{287}\thinspace{}Death carries off the man praised for his chil\-dren and flocks, his mind dis\-tract\-ed, as a flood carries off a sleep\-ing vil\-lage.
\end{para}

\begin{para}
    \textsuperscript{288}\thinspace{}For him whom death has seized, kinsmen are no pro\-tec\-tion. Nei\-ther sons, a fa\-ther, nor re\-la\-tions can give shel\-ter.
    \textsuperscript{289}\thinspace{}Know\-ing this, a wise and good man should quick\-ly clear the path lead\-ing to sal\-va\-tion.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Miscellaneous}\addcontentsline{toc}{section}{Miscellaneous}
\begin{para}
    \textsuperscript{290}\thinspace{}If by leav\-ing a small pleas\-ure one sees a great pleas\-ure, let a wise man leave the small pleas\-ure and look to the great.
\end{para}

\begin{para}
    \textsuperscript{291}\thinspace{}He who seeks pleas\-ure for him\-self by caus\-ing pain to others is entangled by the bonds of ha\-tred, and can\-not be free from ha\-tred.
\end{para}

\begin{para}
    \textsuperscript{292}\thinspace{}Desires al\-ways in\-crease for un\-ru\-ly, heed\-less peo\-ple who ne\-glect what should be done and do what should not be done.
    \textsuperscript{293}\thinspace{}But desires will come to an end for watch\-ful, wise peo\-ple who don't do what should not be done, who stead\-fast\-ly do what should be done, and whose whole watch\-ful\-ness is al\-ways di\-rect\-ed to their bod\-y.
\end{para}

\begin{para}
    \textsuperscript{296}\thinspace{}The disciples of Gotama are wide a\-wake and vig\-i\-lant, and their minds day and night are al\-ways set on the Awakened,
    \textsuperscript{297}\thinspace{}on the law,
    \textsuperscript{298}\thinspace{}on the church,
    \textsuperscript{299}\thinspace{}and on their bod\-y.
\end{para}

\begin{para}
    \textsuperscript{300}\thinspace{}The disciples of Gotama are wide a\-wake and vig\-i\-lant, and their minds day and night al\-ways de\-light in com\-pas\-sion,
    \textsuperscript{301}\thinspace{}and in med\-i\-ta\-tion.
\end{para}

\begin{para}
    \textsuperscript{302}\thinspace{}It's hard to be a monk. It's hard to en\-joy that life. House\-hold life is al\-so hard and pain\-ful. It's pain\-ful to dwell with unequals and the wan\-der\-ing monk is be\-set with pain. There\-fore let no man wan\-der aim\-less\-ly and he will not be be\-set with pain.
\end{para}

\begin{para}
    \textsuperscript{303}\thinspace{}A faith\-ful, vir\-tu\-ous, rep\-u\-ta\-ble, and wealth\-y man is respected wher\-ev\-er he goes.
    \textsuperscript{304}\thinspace{}Like the Himalayas, good peo\-ple shine from a\-far. Like ar\-rows shot by night, bad peo\-ple are un\-seen e\-ven when they are near.
\end{para}

\begin{para}
    \textsuperscript{305}\thinspace{}He who sits a\-lone, sleeps a\-lone, walks a\-lone, is dil\-i\-gent, and subdues him\-self a\-lone, will re\-joice in the sol\-i\-tude of the for\-est.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Hell}\addcontentsline{toc}{section}{Hell}
\begin{para}
    \textsuperscript{306}\thinspace{}He goes to hell, the one who asserts what did\-n't hap\-pen, and the one who did a thing but says, “I have\-n't done it”. They're both evildoers, and are e\-qual in the next world.
\end{para}

\begin{para}
    \textsuperscript{307}\thinspace{}Man\-y men who wear the saf\-fron robe are ill-mannered and un\-re\-strained. Such evildoers by their evil deeds go to hell.
    \textsuperscript{308}\thinspace{}It would be bet\-ter to swal\-low an i\-ron ball, red-hot and a\-flame, than for an im\-mor\-al, un\-re\-strained man to eat the alms of the peo\-ple.
\end{para}

\begin{para}
    \textsuperscript{309}\thinspace{}A wreckless man who lies down with his neighbor's wife gains four things: a bad rep\-u\-ta\-tion, poor sleep, pun\-ish\-ment, and hell.
    \textsuperscript{310}\thinspace{}There is bad rep\-u\-ta\-tion, an evil des\-ti\-na\-tion, the brief pleas\-ure of the fright\-ened man and wom\-an, and the king imposes a heav\-y pun\-ish\-ment. There\-fore, let no man lie down with his neighbor's wife.
\end{para}

\begin{para}
    \textsuperscript{311}\thinspace{}Just as a blade of grass, if wrong\-ly han\-dled, cuts the hand, so does as\-cet\-i\-cism, if wrong\-ly prac\-tised, lead to hell.
\end{para}

\begin{para}
    \textsuperscript{312}\thinspace{}An act performed care\-less\-ly, a bro\-ken vow, and ques\-tion\-a\-ble cel\-i\-ba\-cy, none of these bring much re\-ward.
    \textsuperscript{313}\thinspace{}If any\-thing is to be done, then do it vig\-or\-ous\-ly. A care\-less pil\-grim on\-ly scatters the dust of his passions more wide\-ly.
\end{para}

\begin{para}
    \textsuperscript{314}\thinspace{}It's bet\-ter not to do an evil deed, for an evil deed torments a man af\-ter\-wards. It's bet\-ter to do a good deed, for hav\-ing done it, one is not tormented.
    \textsuperscript{315}\thinspace{}Like a fron\-tier fort with defences in\-side and out, so let a man guard him\-self. Don't let a mo\-ment pass by, for they who al\-low the right mo\-ment to pass suf\-fer pain when consigned to hell.
\end{para}

\begin{para}
    \textsuperscript{316}\thinspace{}They who are a\-shamed of what is not shame\-ful, and are not a\-shamed of what is shame\-ful;
    \textsuperscript{317}\thinspace{}they who see dan\-ger where there is none, and see no dan\-ger where there is;
    \textsuperscript{318}\thinspace{}they who see wrong where there is none, and see no wrong where there is; such men, em\-brac\-ing false doctrines, go to an evil des\-ti\-na\-tion.
\end{para}

\begin{para}
    \textsuperscript{319}\thinspace{}They who see wrong as wrong, and see right as right, such men, em\-brac\-ing true doc\-trine go to a good des\-ti\-na\-tion.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{The Elephant}\addcontentsline{toc}{section}{The Elephant}
\begin{para}
    \textsuperscript{320}\thinspace{}As an el\-e\-phant in bat\-tle endures an ar\-row shot from a bow, so will I en\-dure def\-a\-ma\-tion. Man\-y peo\-ple lack vir\-tue.
\end{para}

\begin{para}
    \textsuperscript{321}\thinspace{}They on\-ly lead a tamed el\-e\-phant in\-to crowds. The king on\-ly rides a tamed el\-e\-phant. The tamed man who endures def\-a\-ma\-tion is the best a\-mong men.
\end{para}

\begin{para}
    \textsuperscript{322}\thinspace{}Trained mules, thor\-ough\-bred hors\-es, and elephants with large tusks are ex\-cel\-lent, but he who tames him\-self is bet\-ter still.
    \textsuperscript{323}\thinspace{}For with these animals no man reaches the un\-trod\-den coun\-try where a tamed man goes by tam\-ing him\-self.
\end{para}

\begin{para}
    \textsuperscript{324}\thinspace{}The el\-e\-phant called Dhanapalaka, deep in rut, is dif\-fi\-cult to hold. He does\-n't eat a mor\-sel in cap\-tiv\-i\-ty, but longs for the el\-e\-phant grove.
\end{para}

\begin{para}
    \textsuperscript{325}\thinspace{}When a man is in\-ac\-tive and glut\-ton\-ous, sleep\-ing and roll\-ing a\-round in bed like an o\-ver\-fed hog, that fool is born again and again.
\end{para}

\begin{para}
    \textsuperscript{326}\thinspace{}This mind used to wan\-der a\-bout as it liked, wher\-ev\-er it wished, how\-ev\-er it pleased. Now I will hold it in check, as a rid\-er with a hook controls an el\-e\-phant in rut.
\end{para}

\begin{para}
    \textsuperscript{327}\thinspace{}De\-light in mind\-ful\-ness. Watch over your mind. Lift your\-self out of the evil way, as an el\-e\-phant lifts it\-self out of the mud.
\end{para}

\begin{para}
    \textsuperscript{328}\thinspace{}If you find a wise, pru\-dent, and vir\-tu\-ous friend, then, over\-com\-ing all dangers, walk with him joy\-ful\-ly and mind\-ful\-ly.
    \textsuperscript{329}\thinspace{}If you find no friend who is wise, pru\-dent, and vir\-tu\-ous, then walk a\-lone like a king leav\-ing be\-hind his king\-dom, like an el\-e\-phant in the for\-est.
\end{para}

\begin{para}
    \textsuperscript{330}\thinspace{}It's bet\-ter to live a\-lone. There's no com\-pan\-ion\-ship with a fool. Live a\-lone, do no evil, and be con\-tent, like an el\-e\-phant in the for\-est.
\end{para}

\begin{para}
    \textsuperscript{331}\thinspace{}These are good: friends when the need arises, con\-tent\-ment with what you have, mer\-it in the hour of death, giv\-ing up all grief,
    \textsuperscript{332}\thinspace{}serv\-ing your moth\-er, serv\-ing your fa\-ther, serv\-ing monks, serv\-ing ho\-ly men,
    \textsuperscript{333}\thinspace{}vir\-tue last\-ing in\-to old age, firm\-ly root\-ed faith, attaining dis\-cern\-ment, avoiding evil.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Craving}\addcontentsline{toc}{section}{Craving}
\begin{para}
    \textsuperscript{334}\thinspace{}The crav\-ing of a heed\-less man grows like a creep\-ing vine. He runs from life to life, like a mon\-key seek\-ing fruit in the for\-est.
\end{para}

\begin{para}
    \textsuperscript{335}\thinspace{}He who is over\-come by this vile and cling\-ing crav\-ing, his suf\-fer\-ing increases like grass af\-ter rain.
    \textsuperscript{336}\thinspace{}He who overcomes this vile crav\-ing, so dif\-fi\-cult to con\-quer, suf\-fer\-ing falls off him like wa\-ter drops from a lo\-tus leaf.
\end{para}

\begin{para}
    \textsuperscript{337}\thinspace{}To all as\-sem\-bled here I say, “Good luck. Dig up the root of crav\-ing like one seek\-ing the sweet scented root of birana grass. Let not the temp\-ter crush you again and again as the stream crushes a reed.”
    \textsuperscript{338}\thinspace{}A tree will grow again, e\-ven if cut down, as long as its root is safe and firm. Thus, un\-less la\-tent crav\-ing is destroyed, suf\-fer\-ing will re\-turn again and again.
\end{para}

\begin{para}
    \textsuperscript{339}\thinspace{}The thir\-ty six streams of crav\-ing flow to\-wards pleas\-ure. The mis\-guid\-ed man, in whom they flow strongly, is car\-ried a\-way.
    \textsuperscript{340}\thinspace{}The streams flow eve\-ry\-where. The creep\-ing vines of crav\-ing grow eve\-ry\-where. If you see the vine spring up, cut its root with knowl\-edge.
\end{para}

\begin{para}
    \textsuperscript{342}\thinspace{}Driv\-en by crav\-ing, men run a\-bout like a snared rab\-bit. Held by fetters and bonds, they suf\-fer again and again for a long time.
    \textsuperscript{343}\thinspace{}So a monk should drive out crav\-ing, and strive to be with\-out pas\-sion.
\end{para}

\begin{para}
    \textsuperscript{345}\thinspace{}The wise don't say that a fet\-ter made of i\-ron, wood, or hemp is strong. Far stronger is a long\-ing for jewels and ornaments, chil\-dren and wives.
    \textsuperscript{346}\thinspace{}The wise say that a strong fet\-ter drags one down, though seem\-ing loose, it's dif\-fi\-cult to re\-move. Af\-ter hav\-ing cut it, the wise are free from cares, and leave the world, desires, and pleasures be\-hind.
\end{para}

\begin{para}
    \textsuperscript{348}\thinspace{}Give up the past, the fu\-ture, and the pres\-ent to cross to the oth\-er shore of ex\-ist\-ence. If your mind is com\-plete\-ly free, you won't en\-ter in\-to birth and de\-cay again.
\end{para}

\begin{para}
    \textsuperscript{349}\thinspace{}If a man is tossed a\-bout by doubts, full of pas\-sion, and fo\-cused on pleas\-ure, his crav\-ing grows more and more. He makes his fetters strong in\-deed.
    \textsuperscript{350}\thinspace{}If a man delights in quieting doubts, is al\-ways mind\-ful, and meditates on the impurities, he will end crav\-ing and cut the fet\-ter of the temp\-ter.
\end{para}

\begin{para}
    \textsuperscript{351}\thinspace{}He who has reached the goal is with\-out fear, is with\-out crav\-ing, is with\-out blem\-ish, and has re\-moved all the thorns of life: this will be his last bod\-y.
\end{para}

\begin{para}
    \textsuperscript{354}\thinspace{}The gift of the law exceeds all gifts. The taste of the law exceeds all tastes. De\-light in the law exceeds all delights. The ex\-tinc\-tion of crav\-ing overcomes all suf\-fer\-ing.
\end{para}

\begin{para}
    \textsuperscript{355}\thinspace{}Rich\-es de\-stroy the fool\-ish, but not those who seek the be\-yond. The fool, by crav\-ing for rich\-es, destroys him\-self and others.
\end{para}

\begin{para}
    \textsuperscript{356}\thinspace{}Weeds spoil a field. Pas\-sion, ha\-tred, de\-lu\-sion, or de\-sire spoil a man.
    \textsuperscript{359}\thinspace{}There\-fore, a gift giv\-en to those with\-out pas\-sion, ha\-tred, de\-lu\-sion, or de\-sire yields great fruit.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{The Monk}\addcontentsline{toc}{section}{The Monk}
\begin{para}
    \textsuperscript{360}\thinspace{}Good is re\-straint over the eye, over the ear, over the nose, and over the tongue.
    \textsuperscript{361}\thinspace{}Good is re\-straint in the bod\-y, in speech, in mind, in all things. A monk, restrained in all things, is freed from all suf\-fer\-ing.
\end{para}

\begin{para}
    \textsuperscript{362}\thinspace{}He who controls his hands, feet, and speech; who is well con\-trolled, delights in the in\-ward, is col\-lect\-ed, sol\-i\-tar\-y and con\-tent, him they call a monk.
\end{para}

\begin{para}
    \textsuperscript{363}\thinspace{}The monk who controls his tongue, speaks mod\-er\-ate\-ly and calm\-ly, and teaches the law and its mean\-ing, his speech is sweet.
\end{para}

\begin{para}
    \textsuperscript{364}\thinspace{}The monk who abides in the law, delights in the law, meditates on the law, and keeps the law in mind, he will nev\-er fall a\-way from the true law.
\end{para}

\begin{para}
    \textsuperscript{365}\thinspace{}Let him not de\-spise what he has re\-ceived, nor en\-vy others. A monk who envies others does\-n't ob\-tain con\-cen\-tra\-tion.
\end{para}

\begin{para}
    \textsuperscript{366}\thinspace{}If a monk lives pure\-ly, is dil\-i\-gent, and does\-n't de\-spise what he has re\-ceived, though he receives lit\-tle, then e\-ven the gods will praise him.
    \textsuperscript{368}\thinspace{}The monk who abides in kind\-ness, and is de\-vot\-ed to the teachings of Bud\-dha, will at\-tain peace, hap\-pi\-ness, and ces\-sa\-tion of nat\-u\-ral desires.
\end{para}

\begin{para}
    \textsuperscript{369}\thinspace{}Monk, bail out this boat. Once bailed, it will sail light\-ly. Hav\-ing cut out pas\-sion and ha\-tred, you will reach sal\-va\-tion.
\end{para}

\begin{para}
    \textsuperscript{371}\thinspace{}Med\-i\-tate monk. Be mind\-ful. Don't di\-rect your mind to sen\-su\-al pleas\-ure. Heed\-less, don't swal\-low the flam\-ing i\-ron ball lest you cry out when burn\-ing, “This is suf\-fer\-ing.”
\end{para}

\begin{para}
    \textsuperscript{372}\thinspace{}There is no med\-i\-ta\-tion with\-out in\-sight. There is no in\-sight with\-out med\-i\-ta\-tion. He who has in\-sight and med\-i\-ta\-tion is near sal\-va\-tion.
\end{para}

\begin{para}
    \textsuperscript{373}\thinspace{}A monk who has entered his emp\-ty house, whose mind is tran\-quil, and who sees the law clear\-ly, feels a de\-light more than hu\-man.
\end{para}

\begin{para}
    \textsuperscript{375}\thinspace{}This is the be\-gin\-ning for a wise monk: guarding the sen\-ses, con\-tent\-ment, re\-straint un\-der the law.
    \textsuperscript{376}\thinspace{}He should have friends who are no\-ble, en\-er\-get\-ic, and pure. He should be friend\-ly, and skilled in con\-duct. Then, hav\-ing much joy, he will put an end to suf\-fer\-ing.
\end{para}

\begin{para}
    \textsuperscript{377}\thinspace{}Shed pas\-sion and ha\-tred, as the jas\-mine plant sheds its withered flowers.
    \textsuperscript{378}\thinspace{}That monk is called qui\-et whose bod\-y and speech and mind are calm, who is col\-lect\-ed, and who has purged the baits of the world.
\end{para}

\begin{para}
    \textsuperscript{379}\thinspace{}Re\-prove your\-self and ex\-am\-ine your\-self by your\-self. By be\-ing vig\-i\-lant and at\-ten\-tive you will live hap\-pi\-ly.
    \textsuperscript{380}\thinspace{}You are your own mas\-ter and ref\-uge. There\-fore, watch over your\-self as a mer\-chant watches over a good horse.
\end{para}

\begin{para}
    \textsuperscript{381}\thinspace{}The monk who is full of joy, and who has faith in the teachings of Bud\-dha will reach the qui\-et place, ces\-sa\-tion of nat\-u\-ral desires, and hap\-pi\-ness.
    \textsuperscript{382}\thinspace{}E\-ven a young monk who applies him\-self to the teachings of Bud\-dha brightens this world, like the moon set free from clouds.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{The Holy Man}\addcontentsline{toc}{section}{The Holy Man}
\begin{para}
    \textsuperscript{386}\thinspace{}He who is mind\-ful, blame\-less, settled, du\-ti\-ful, with\-out passions, and who has attained the highest goal, him I call a ho\-ly man.
\end{para}

\begin{para}
    \textsuperscript{387}\thinspace{}The sun shines by day, the moon shines by night, the war\-ri\-or shines in ar\-mor, the ho\-ly man shines in med\-i\-ta\-tion.
\end{para}

\begin{para}
    \textsuperscript{388}\thinspace{}A man is called ho\-ly for discarding evil, con\-tem\-pla\-tive for walk\-ing qui\-et\-ly, a pil\-grim for leav\-ing his own impurities.
\end{para}

\begin{para}
    \textsuperscript{389}\thinspace{}No one should at\-tack a ho\-ly man, but no ho\-ly man, if attacked, should give in to his an\-ger. Shame on him who attacks a ho\-ly man, but more shame to him who gives in to his an\-ger.
\end{para}

\begin{para}
    \textsuperscript{390}\thinspace{}Noth\-ing is bet\-ter for a ho\-ly man than to hold his mind back from the pleasures of life. When all de\-sire to in\-jure has vanished, pain ceases.
\end{para}

\begin{para}
    \textsuperscript{391}\thinspace{}I call him a ho\-ly man who does\-n't of\-fend in bod\-y, word, or mind, and is restrained on these three points.
\end{para}

\begin{para}
    \textsuperscript{392}\thinspace{}Re\-vere that teach\-er who has taught you the law as taught by the Bud\-dha, just as a ho\-ly man\-y reveres the sac\-ri\-fi\-cial fire.
\end{para}

\begin{para}
    \textsuperscript{393}\thinspace{}A man does\-n't be\-come ho\-ly by his mat\-ted hair, by his fam\-i\-ly, or by birth. He in whom there is truth and right\-eous\-ness, he is pure, he is a ho\-ly man.
    \textsuperscript{394}\thinspace{}What use is mat\-ted hair, O fool? What use is your coat of goat skins? The tan\-gle is with\-in you, but you comb the out\-side.
\end{para}

\begin{para}
    \textsuperscript{397}\thinspace{}I call him a ho\-ly man who has cut eve\-ry fet\-ter, is fear\-less, in\-de\-pend\-ent and unshackled.
\end{para}

\begin{para}
    \textsuperscript{399}\thinspace{}I call him a ho\-ly man who, with\-out an\-ger, endures in\-sult, im\-pris\-on\-ment, and beat\-ing, who has pa\-tience like the strength of an ar\-my.
\end{para}

\begin{para}
    \textsuperscript{400}\thinspace{}I call him a ho\-ly man who is free from an\-ger, du\-ti\-ful, vir\-tu\-ous, with\-out crav\-ing, sub\-dued, and who has re\-ceived his last bod\-y.
\end{para}

\begin{para}
    \textsuperscript{401}\thinspace{}I call him a ho\-ly man who does\-n't cling to pleasures; like wa\-ter on a lo\-tus leaf or a mus\-tard seed on the tip of a nee\-dle.
    \textsuperscript{402}\thinspace{}I call him a ho\-ly man who, e\-ven here, discerns the end of his suf\-fer\-ing, has put down his bur\-den, and is unshackled.
\end{para}

\begin{para}
    \textsuperscript{403}\thinspace{}I call him a ho\-ly man who is wise, has deep un\-der\-stand\-ing, discerns the right and wrong path, and has reached the ul\-ti\-mate goal.
\end{para}

\begin{para}
    \textsuperscript{404}\thinspace{}I call him a ho\-ly man who keeps a\-loof from both laymen and as\-cet\-ic, who has no per\-ma\-nent home, and few desires.
\end{para}

\begin{para}
    \textsuperscript{405}\thinspace{}I call him a ho\-ly man who renounces vi\-o\-lence to\-wards all beings, wheth\-er weak or strong, and nei\-ther kills nor causes others to kill.
\end{para}

\begin{para}
    \textsuperscript{407}\thinspace{}I call him a ho\-ly man from whom pas\-sion and ha\-tred, pride and en\-vy have dropped like a mus\-tard seed from the tip of a nee\-dle.
\end{para}

\begin{para}
    \textsuperscript{408}\thinspace{}I call him a ho\-ly man who speaks true and in\-struc\-tive words, free from harsh\-ness, so that he offends no one.
    \textsuperscript{409}\thinspace{}I call him a ho\-ly man who on\-ly takes what is giv\-en to him, be it long or short, small or large, good or bad.
\end{para}

\begin{para}
    \textsuperscript{412}\thinspace{}I call him a ho\-ly man who, in this world, is a\-bove at\-tach\-ment to mer\-it and evil, is with\-out grief, stain\-less, and pure.
\end{para}

\begin{para}
    \textsuperscript{416}\thinspace{}I call him a ho\-ly man who, leav\-ing all crav\-ing, travels with\-out a home, and in whom all crav\-ing is ex\-tinct.
\end{para}

\begin{para}
    \textsuperscript{417}\thinspace{}I call him a ho\-ly man who, af\-ter leav\-ing all at\-tach\-ment to men, has ris\-en a\-bove all at\-tach\-ment to the gods, and is free from eve\-ry at\-tach\-ment.
\end{para}

\begin{para}
    \textsuperscript{421}\thinspace{}I call him a ho\-ly man who does\-n't cling to past, pres\-ent, or fu\-ture, who has noth\-ing, who clings to noth\-ing.
\end{para}


\end{document}

% Copyright 2021 Michael Hendricks
%
% Permission is granted to use, copy, modify, and distribute
% this work for any purpose with or without fee.
%
% This work is provided as-is without any express or implied
% warranty. In no event is the author liable for any damages
% arising from this work.
