%
% see tex.stackexchange.com/a/600/151276
\def\changemargin#1#2{\list{}{\rightmargin#2\leftmargin#1}\item[]}
\let\endchangemargin=\endlist

\documentclass[openany,12pt,english]{book}
\usepackage[utf8]{inputenc}
\usepackage{babel}
\usepackage{microtype}
\usepackage{csquotes}

\usepackage[
  paperwidth=8.5in,
  paperheight=11in,
  top=1in,
  outer=1in,
  bottom=1.25in,
  inner=1.5in
]{geometry}

\title{Introduction}
\author{}
\date{}

\newenvironment{para}{\par\pretolerance=100\tolerance=200\setlength{\emergencystretch}{0.6em}\relax}{\par}

\begin{document}

\maketitle

% Copyright page
{
\thispagestyle{empty}
\vspace*{\fill}
\begin{changemargin}{.8in}{.8in}
\begin{center}
\textcopyright{}~2021 Michael Hendricks

\vspace{0.5\baselineskip}

Permission is granted to use, copy, modify, and distribute
this work for any purpose with or without fee.

\vspace{0.5\baselineskip}

This work is provided as-is without any express or implied
warranty. In no event is the author liable for any damages
arising from this work.
\end{center}
\end{changemargin}
}
\pagenumbering{arabic}
\clearpage{}

% Table of Contents
\clearpage{}

\markboth{}{}
\twocolumn[\begin{changemargin}{.6in}{.6in}
\end{changemargin}]{}


\begin{para}
    \textsuperscript{1}\thinspace{}God loves eve\-ry\-one and wants each of us to find joy and sat\-is\-fac\-tion.
    \textsuperscript{2}\thinspace{}To help us to\-wards that goal, God gave us rea\-son, con\-science, and the a\-bil\-i\-ty to choose.
    \textsuperscript{3}\thinspace{}Our re\-spon\-si\-bil\-i\-ty is to fol\-low, sin\-cere\-ly, the dictates of our con\-science, guid\-ed by rea\-son, while re\-spect\-ing the e\-qual priv\-i\-lege of others to do the same.
\end{para}

\begin{para}
    \textsuperscript{4}\thinspace{}We're ac\-count\-a\-ble to God for our choices.
    \textsuperscript{5}\thinspace{}Each of us can have a per\-son\-al re\-la\-tion\-ship with God.
    \textsuperscript{6}\thinspace{}God communicates with each of us as needed,
    \textsuperscript{7}\thinspace{}and sends messengers to dif\-fer\-ent times, places, and cultures to guide us. Stud\-y\-ing the messages that are in\-tend\-ed for us can en\-light\-en our un\-der\-stand\-ing.
\end{para}

\begin{para}
    \textsuperscript{8}\thinspace{}God knows the vast\-ness of hu\-man di\-ver\-si\-ty. Just as a skilled op\-tom\-e\-trist prescribes eye\-glass\-es to com\-pen\-sate for a patient's shortcomings, God guides us to a pre\-scrip\-tion that's op\-ti\-mized for our spir\-it\-u\-al growth. Our pre\-scrip\-tion may dif\-fer from, or co\-in\-cide with, that of our peers.
    \textsuperscript{9}\thinspace{}Be\-cause of this di\-ver\-si\-ty, a com\-mu\-ni\-ty of believers can strength\-en and in\-spire one an\-oth\-er.
\end{para}

\begin{para}
    \textsuperscript{10}\thinspace{}Details a\-bout God's na\-ture and methods are large\-ly a dis\-trac\-tion from the prop\-er fo\-cus of our lives: a life\-long mi\-gra\-tion to\-wards joy through our per\-son\-al choices.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{11}\thinspace{}While striving to live the principles de\-scribed a\-bove, I felt guid\-ed to prayer\-ful\-ly un\-der\-take the actions be\-low. The results are for my own ben\-e\-fit and may not ben\-e\-fit others in the same way.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{12}\thinspace{}Cre\-ate a prioritized list of es\-sen\-tial commandments to help me fo\-cus on what matters most and to re\-solve conflicts that may a\-rise be\-tween those commandments. This be\-came the Di\-gest of Priorities.
\end{para}

\begin{para}
    \textsuperscript{13}\thinspace{}Write pairs of es\-sen\-tial vir\-tues whose opposing demands I must bal\-ance in my own life. This be\-came the Di\-gest of Vir\-tues.
\end{para}

\begin{para}
    \textsuperscript{14}\thinspace{}Try dif\-fer\-ent methods of pray\-er, listening care\-ful\-ly to un\-der\-stand which are most ef\-fec\-tive at strengthening my re\-la\-tion\-ship with God. Cod\-i\-fy those practices and ob\-serve them faith\-ful\-ly. This be\-came the Di\-gest of Pray\-er.
\end{para}

\begin{para}
    \textsuperscript{15}\thinspace{}Rec\-ord short stories that God has used to clar\-i\-fy my un\-der\-stand\-ing of cer\-tain principles. This be\-came the Di\-gest of Parables.
\end{para}

\begin{para}
    \textsuperscript{16}\thinspace{}Stud\-y the teachings of those who are wide\-ly con\-sid\-ered to be messengers of God. A\-bridge those messages by selecting the most prac\-ti\-cal and im\-por\-tant teachings for guid\-ing my life. This be\-came the Messages. I re\-fer to them as The Mes\-sage of Mo\-ses, The Mes\-sage of Jesus, etc.
\end{para}


\end{document}

% Copyright 2021 Michael Hendricks
%
% Permission is granted to use, copy, modify, and distribute
% this work for any purpose with or without fee.
%
% This work is provided as-is without any express or implied
% warranty. In no event is the author liable for any damages
% arising from this work.
