%
% see tex.stackexchange.com/a/600/151276
\def\changemargin#1#2{\list{}{\rightmargin#2\leftmargin#1}\item[]}
\let\endchangemargin=\endlist

\documentclass[openany,12pt,english]{book}
\usepackage[utf8]{inputenc}
\usepackage{babel}
\usepackage{microtype}
\usepackage{csquotes}

\usepackage[
  paperwidth=8.5in,
  paperheight=11in,
  top=1in,
  outer=1in,
  bottom=1.25in,
  inner=1.5in
]{geometry}

\title{Moses}
\author{}
\date{}

\newenvironment{para}{\par\pretolerance=100\tolerance=200\setlength{\emergencystretch}{0.6em}\relax}{\par}

\begin{document}

\maketitle

% Copyright page
{
\thispagestyle{empty}
\vspace*{\fill}
\begin{changemargin}{.8in}{.8in}
\begin{center}
\textcopyright{}~2021 Michael Hendricks

\vspace{0.5\baselineskip}

Permission is granted to use, copy, modify, and distribute
this work for any purpose with or without fee.

\vspace{0.5\baselineskip}

This work is provided as-is without any express or implied
warranty. In no event is the author liable for any damages
arising from this work.
\end{center}
\end{changemargin}
}
\pagenumbering{arabic}
\clearpage{}

% Table of Contents
\clearpage{}

\markboth{}{}
\twocolumn[\begin{changemargin}{.6in}{.6in}
  \begin{center}
    \itshape
    An a\-bridg\-ment of the Book of Deuteronomy.
    Verse references such
    as~\textsuperscript{5:6}\thinspace{}indicate
    the original source of the text, but
    don't imply a full quotation.
  \end{center}
\end{changemargin}]{}


\begin{para}
    \textsuperscript{5:1}\thinspace{}Mo\-ses summoned all Israel and said to them, Hear the statutes and ordinances that I speak to\-day. Learn them and ob\-serve them care\-ful\-ly.
    \textsuperscript{2}\thinspace{}\textsc{God} made a cov\-e\-nant with us.
    \textsuperscript{3}\thinspace{}\textsc{God} did not make this cov\-e\-nant with our fathers.
    \textsuperscript{4}\thinspace{}\textsc{God} spoke with you on the moun\-tain, say\-ing,
\end{para}

\begin{para}
    \textsuperscript{5:6}\thinspace{}I am \textsc{God}, who brought you out of the land of Egypt, out of the house of slav\-er\-y.
    \textsuperscript{7}\thinspace{}Don't have oth\-er gods be\-fore me.
    \textsuperscript{8}\thinspace{}Don't make an i\-dol for your\-self in the form of any\-thing in the heavens or on the earth or in the waters.
    \textsuperscript{9}\thinspace{}Don't wor\-ship them or serve them, for I am a jeal\-ous god, visiting the in\-iq\-ui\-ty of the fathers on the chil\-dren to the third and fourth gen\-er\-a\-tion,
    \textsuperscript{10}\thinspace{}but show\-ing love to thousands who love me and keep my commandments.
\end{para}

\begin{para}
    \textsuperscript{5:11}\thinspace{}Don't mis\-use the name of \textsc{God}.
\end{para}

\begin{para}
    \textsuperscript{5:12}\thinspace{}Ob\-serve the sabbath day by keep\-ing it ho\-ly.
    \textsuperscript{13}\thinspace{}La\-bor six days and do all your work,
    \textsuperscript{14}\thinspace{}but the sev\-enth day is a sabbath to \textsc{God}. On it you shall not do any work: nei\-ther you, your chil\-dren, your serv\-ant, your live\-stock, nor any for\-eign\-er liv\-ing a\-mong you, that they may rest as you do.
    \textsuperscript{15}\thinspace{}Re\-mem\-ber that you were slaves in Egypt, and that \textsc{God} brought you out. There\-fore, \textsc{God} commanded you to keep the sabbath day.
\end{para}

\begin{para}
    \textsuperscript{5:16}\thinspace{}Hon\-or your parents so that you may live long and that it may go well with you in the land that \textsc{God} is giv\-ing you.
\end{para}

\begin{para}
    \textsuperscript{5:17}\thinspace{}Don't mur\-der.
    \textsuperscript{18}\thinspace{}Don't com\-mit a\-dul\-ter\-y.
    \textsuperscript{19}\thinspace{}Don't steal.
    \textsuperscript{20}\thinspace{}Don't give false tes\-ti\-mo\-ny a\-gainst your neigh\-bor.
    \textsuperscript{21}\thinspace{}Don't cov\-et your neighbor's wife, house, field, servants, live\-stock, or any\-thing that belongs to your neigh\-bor.
\end{para}

\begin{para}
    \textsuperscript{5:32}\thinspace{}Be care\-ful to do as \textsc{God} has commanded you. Don't turn a\-side to the right or to the left.
    \textsuperscript{33}\thinspace{}Walk on the path which \textsc{God} has commanded you, that you may live, pros\-per, and pro\-long your days in the land that you will pos\-sess.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{6:5}\thinspace{}Love \textsc{God} with all your heart, soul, and strength.
    \textsuperscript{6}\thinspace{}These words that I com\-mand you to\-day shall be on your heart.
\end{para}

\begin{para}
    \textsuperscript{6:10}\thinspace{}When \textsc{God} brings you in\-to the land that he swore to your fathers Abraham, Isaac, and Jacob that he would give you --- a land with large and splen\-did cit\-ies,
    \textsuperscript{11}\thinspace{}hous\-es filled with eve\-ry good thing, cisterns, vineyards, and ol\-ive trees --- when you eat and are sat\-is\-fied,
    \textsuperscript{12}\thinspace{}then be care\-ful not to for\-get \textsc{God}.
\end{para}

\begin{para}
    \textsuperscript{6:14}\thinspace{}Don't fol\-low oth\-er gods, the gods of the peo\-ple a\-round you,
    \textsuperscript{15}\thinspace{}lest the an\-ger of \textsc{God} be kindled a\-gainst you, and de\-stroy you from the face of the earth.
\end{para}

\begin{para}
    \textsuperscript{6:20}\thinspace{}When your son asks you, “What is the mean\-ing of the testimonies, the statutes, and the rules that \textsc{God} has commanded you?”
    \textsuperscript{21}\thinspace{}tell him, “We were slaves to Pharaoh in Egypt, but \textsc{God} brought us out of Egypt.
    \textsuperscript{23}\thinspace{}He brought us out from there in or\-der to give us the land that he swore to our fathers.
    \textsuperscript{24}\thinspace{}So \textsc{God} commanded us to ob\-serve all these statutes and to fear \textsc{God} for our good and our sur\-viv\-al.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{7:1}\thinspace{}When \textsc{God} brings you in\-to the land, and clears a\-way man\-y nations be\-fore you,
    \textsuperscript{3}\thinspace{}don't in\-ter\-mar\-ry with them.
    \textsuperscript{4}\thinspace{}They will turn your chil\-dren a\-way from me, to wor\-ship oth\-er gods.
\end{para}

\begin{para}
    \textsuperscript{7:7}\thinspace{}\textsc{God} did\-n't choose you be\-cause you were more nu\-mer\-ous than oth\-er peo\-ple,
    \textsuperscript{8}\thinspace{}but be\-cause \textsc{God} loves you and is keep\-ing the oath that he swore to your fathers.
\end{para}

\begin{para}
    \textsuperscript{7:12}\thinspace{}If you lis\-ten to and care\-ful\-ly keep these ordinances, \textsc{God} will keep the cov\-e\-nant of love with you that he swore to your fathers.
    \textsuperscript{13}\thinspace{}He will love you, bless you, and mul\-ti\-ply you. He will bless the fruit of your womb, the crops of your land, and the in\-crease of your herd.
\end{para}

\begin{para}
    \textsuperscript{7:17}\thinspace{}If you say in your heart, “These nations are greater than I. How can I dis\-pos\-sess them?”
    \textsuperscript{18}\thinspace{}Don't be a\-fraid of them. Re\-mem\-ber what \textsc{God} did to Pharaoh and to all Egypt.
    \textsuperscript{21}\thinspace{}Don't dread them, for \textsc{God} is a\-mong you, a great and awe\-some god.
    \textsuperscript{22}\thinspace{}\textsc{God} will clear a\-way these nations be\-fore you lit\-tle by lit\-tle. You will not e\-lim\-i\-nate them all at once, lest the wild animals be\-come too nu\-mer\-ous for you.
    \textsuperscript{25}\thinspace{}Burn the carved images of their gods. Don't cov\-et the sil\-ver or gold that is on them, nor take it for your\-self, lest you be en\-snared by it.
    \textsuperscript{26}\thinspace{}Don't bring an a\-bom\-i\-na\-tion in\-to your house, lest you be set a\-part for de\-struc\-tion. In\-stead, utterly de\-test it and ab\-hor it.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{8:2}\thinspace{}Re\-mem\-ber how \textsc{God} led you these for\-ty years in the wil\-der\-ness, to hum\-ble you, and to test what was in your heart, wheth\-er you would keep his commandments.
    \textsuperscript{3}\thinspace{}He humbled you, let you go hun\-gry, then fed you with man\-na, which nei\-ther you nor your fathers had known, to teach you that man does not live by bread a\-lone, but by eve\-ry word that co\-mes from the mouth of \textsc{God}.
    \textsuperscript{5}\thinspace{}Know in your heart that \textsc{God} disciplines you just as a man disciplines his son.
    \textsuperscript{6}\thinspace{}There\-fore keep the commandments of \textsc{God} by walk\-ing in his ways and by fearing him.
    \textsuperscript{10}\thinspace{}When you've eat\-en and are sat\-is\-fied, then bless \textsc{God} for the good land he has giv\-en you.
\end{para}

\begin{para}
    \textsuperscript{8:11}\thinspace{}Be care\-ful not to for\-get \textsc{God}, by fail\-ing to keep his commandments, his ordinances, and his statutes.
    \textsuperscript{12}\thinspace{}Oth\-er\-wise, once you've eat\-en and are sat\-is\-fied, and have built good hous\-es to live in,
    \textsuperscript{13}\thinspace{}and when your herds mul\-ti\-ply, and your sil\-ver and gold mul\-ti\-ply, and eve\-ry\-thing you have multiplies,
    \textsuperscript{14}\thinspace{}then your heart will be\-come proud, and you will for\-get \textsc{God},
    \textsuperscript{17}\thinspace{}and you may say in your heart, “My pow\-er and the strength of my hand has made me this wealth.”
    \textsuperscript{18}\thinspace{}But re\-mem\-ber that \textsc{God} gives you pow\-er to make wealth in or\-der to con\-firm his cov\-e\-nant.
    \textsuperscript{19}\thinspace{}If you for\-get \textsc{God}, fol\-low oth\-er gods, serve them, and wor\-ship them, I tes\-ti\-fy a\-gainst you to\-day that you will sure\-ly per\-ish.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{9:4}\thinspace{}Once \textsc{God} has driv\-en the peo\-ple out be\-fore you, don't say in your heart, “\textsc{God} brought me in to pos\-sess this land be\-cause of my right\-eous\-ness.” In\-stead, \textsc{God} drives them out be\-cause of their wick\-ed\-ness.
    \textsuperscript{6}\thinspace{}Un\-der\-stand that \textsc{God} is\-n't giv\-ing you this good land be\-cause of your right\-eous\-ness, for you are a stub\-born peo\-ple.
    \textsuperscript{7}\thinspace{}Don't for\-get how you pro\-voked \textsc{God} in the wil\-der\-ness. From the day you de\-part\-ed Egypt un\-til you ar\-rived in this place, you have been re\-bel\-lious a\-gainst \textsc{God}.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{10:17}\thinspace{}\textsc{God} is not par\-tial and takes no bribe.
    \textsuperscript{18}\thinspace{}He executes jus\-tice for orphans and widows, and loves the for\-eign\-er by giv\-ing him food and cloth\-ing.
    \textsuperscript{19}\thinspace{}There\-fore, love the for\-eign\-er, for you were foreigners in Egypt.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{11:2}\thinspace{}Re\-mem\-ber that your chil\-dren did\-n't ex\-pe\-ri\-ence the dis\-ci\-pline of \textsc{God}; his works which he did in Egypt;
    \textsuperscript{4}\thinspace{}what he did to the ar\-my of Egypt; how he made the wa\-ter of the Red sea en\-gulf them as they pur\-sued you;
    \textsuperscript{5}\thinspace{}and what he did to you in the wil\-der\-ness, un\-til you reached this place;
    \textsuperscript{7}\thinspace{}but your eyes have seen eve\-ry great work \textsc{God} has done.
\end{para}

\begin{para}
    \textsuperscript{11:16}\thinspace{}Be care\-ful not to be en\-ticed to turn a\-side and serve oth\-er gods and wor\-ship them.
    \textsuperscript{18}\thinspace{}There\-fore im\-print my words on your heart and on your soul, bind them as a sign on your hand, and wear them on your foreheads.
    \textsuperscript{19}\thinspace{}Teach them to your chil\-dren, speak of them when you sit in your house, when you walk a\-long the road, when you lie down, and when you get up.
    \textsuperscript{20}\thinspace{}Write them on the doorposts of your house and on your gates,
    \textsuperscript{21}\thinspace{}that your days, and the days of your chil\-dren, may be mul\-ti\-plied as long as the heavens are a\-bove the earth.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{12:4}\thinspace{}Don't wor\-ship \textsc{God} in the same way as the nations you dis\-pos\-sess.
    \textsuperscript{5}\thinspace{}In\-stead, seek the place where \textsc{God} chooses to es\-tab\-lish his name. Go there
    \textsuperscript{6}\thinspace{}and bring your burnt offerings, your sacrifices, your tithes, your contributions, your vow offerings, your free\-will offerings, and the firstborn of your live\-stock.
    \textsuperscript{7}\thinspace{}There, in the pres\-ence of \textsc{God}, you and your families will eat and re\-joice in eve\-ry\-thing you un\-der\-take, be\-cause \textsc{God} has bless\-ed you.
\end{para}

\begin{para}
    \textsuperscript{12:20}\thinspace{}When \textsc{God} enlarges your ter\-ri\-to\-ry as he has prom\-ised you, and you say, “I want to eat meat,” you may eat meat when\-ev\-er you want.
\end{para}

\begin{para}
    \textsuperscript{12:29}\thinspace{}When \textsc{God} cuts off the nations be\-fore you, and you live in their land,
    \textsuperscript{30}\thinspace{}be care\-ful not to be en\-snared by their ways,
    \textsuperscript{31}\thinspace{}for they have done eve\-ry de\-test\-a\-ble act for their gods; e\-ven burn\-ing their chil\-dren as sacrifices.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{13:1}\thinspace{}If a proph\-et arises a\-mong you, and gives you a sign or a won\-der,
    \textsuperscript{2}\thinspace{}and it co\-mes true, but he says, “Let us fol\-low oth\-er gods and wor\-ship them,”
    \textsuperscript{3}\thinspace{}don't lis\-ten to that proph\-et. For \textsc{God} is testing you to find out wheth\-er you love \textsc{God} with all your heart and with all your soul.
\end{para}

\begin{para}
    \textsuperscript{13:6}\thinspace{}If your broth\-er, child, wife, or friend se\-cret\-ly entices you, say\-ing, “Let us wor\-ship oth\-er gods,”
    \textsuperscript{8}\thinspace{}don't yield to him and don't lis\-ten to him.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{14:22}\thinspace{}Set a\-side a tenth of the pro\-duce that is harvested from your fields each year.
    \textsuperscript{23}\thinspace{}Eat the tithe of your grain, of your wine, and of your oil, and the firstborn of your live\-stock be\-fore \textsc{God}, in the place where he chooses to es\-tab\-lish his name, so that you may learn to fear \textsc{God} al\-ways.
\end{para}

\begin{para}
    \textsuperscript{14:24}\thinspace{}If the dis\-tance is so great that you can't car\-ry it,
    \textsuperscript{25}\thinspace{}then turn it in\-to mon\-ey and take the mon\-ey to the place where \textsc{God} chooses to es\-tab\-lish his name.
    \textsuperscript{26}\thinspace{}Spend the mon\-ey for what\-ev\-er you want, for ox\-en, or sheep, or wine, or strong drink. Eat it be\-fore \textsc{God} and re\-joice, you and your house\-hold.
    \textsuperscript{27}\thinspace{}Don't ne\-glect the Levite who is in your town, for he has no por\-tion or in\-her\-it\-ance with you.
\end{para}

\begin{para}
    \textsuperscript{14:28}\thinspace{}At the end of eve\-ry third year, bring all the tithe of that year's pro\-duce and store it in your town.
    \textsuperscript{29}\thinspace{}The Levite, the for\-eign\-er, the fa\-ther\-less, and the wid\-ow in your town shall come and eat and be sat\-is\-fied, so that \textsc{God} may bless you in all your work.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{15:1}\thinspace{}At the end of eve\-ry sev\-enth year, can\-cel debts.
    \textsuperscript{2}\thinspace{}Eve\-ry cred\-i\-tor must can\-cel the loans he made to his neigh\-bor. He must not de\-mand pay\-ment from his neigh\-bor.
    \textsuperscript{3}\thinspace{}You may re\-quire pay\-ment from a for\-eign\-er, but you must for\-give what\-ev\-er your broth\-er owes you.
    \textsuperscript{6}\thinspace{}For \textsc{God} will bless you, as he has prom\-ised. You will lend to man\-y nations, but you will not bor\-row. You will rule man\-y nations, but none will rule you.
\end{para}

\begin{para}
    \textsuperscript{15:7}\thinspace{}If there is a poor man a\-mong you, in any of your towns, don't hard\-en your heart or shut your hand to\-wards him.
    \textsuperscript{8}\thinspace{}In\-stead, o\-pen your hand to him and lend him suf\-fi\-cient for his need, what\-ev\-er it may be.
\end{para}

\begin{para}
    \textsuperscript{15:9}\thinspace{}Be care\-ful not to have this wick\-ed thought: “The sev\-enth year, the year of re\-lease, is near,” and you look grudg\-ing\-ly on your poor broth\-er and give him noth\-ing. He will cry to \textsc{God} a\-gainst you, and you will be guilt\-y.
    \textsuperscript{10}\thinspace{}Give gen\-er\-ous\-ly to him with\-out a grudg\-ing heart. Be\-cause of this \textsc{God} will bless you in eve\-ry\-thing you do.
    \textsuperscript{11}\thinspace{}There will al\-ways be poor peo\-ple in the land. There\-fore free\-ly o\-pen your hand to your broth\-er, to the poor, and to the needy.
\end{para}

\begin{para}
    \textsuperscript{15:12}\thinspace{}If a fel\-low Hebrew is sold to you, and serves you for six years, free him in the sev\-enth year.
    \textsuperscript{13}\thinspace{}When you free him, don't send him a\-way empty-handed.
    \textsuperscript{14}\thinspace{}Fur\-nish him lib\-er\-al\-ly from your flock, from your threshing floor, and from your wine\-press. Give to him as \textsc{God} has bless\-ed you.
    \textsuperscript{15}\thinspace{}Re\-mem\-ber that you were a slave in Egypt, and \textsc{God} redeemed you. That is why I am giv\-ing you this com\-mand.
    \textsuperscript{18}\thinspace{}Don't con\-sid\-er it a hard\-ship when you set him free, for he served you six years at half the cost of a hired serv\-ant. \textsc{God} will bless you in all you do.
\end{para}

\begin{para}
    \textsuperscript{15:19}\thinspace{}Ded\-i\-cate to \textsc{God} eve\-ry firstborn male of your live\-stock. Do no work with the firstborn of your herd, nor shear the firstborn of your flock.
    \textsuperscript{20}\thinspace{}Each year you and your fam\-i\-ly are to eat it be\-fore \textsc{God} in the place \textsc{God} chooses.
    \textsuperscript{21}\thinspace{}If it has any de\-fect, is lame or blind, don't sac\-ri\-fice it to \textsc{God}.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{16:1}\thinspace{}Ob\-serve the month of Aviv and cel\-e\-brate the Passover to \textsc{God}, for in the month of Aviv \textsc{God} brought you out of Egypt by night.
    \textsuperscript{3}\thinspace{}For sev\-en days you are to eat un\-leav\-ened bread---be\-cause you left Egypt in haste---so that you may re\-mem\-ber the day you left Egypt for the rest of your life.
    \textsuperscript{4}\thinspace{}Let no yeast be found in your ter\-ri\-to\-ry for sev\-en days.
    \textsuperscript{6}\thinspace{}Sac\-ri\-fice the Passover at sun\-set, at the time you came out of Egypt.
    \textsuperscript{8}\thinspace{}On the sev\-enth day hold a sol\-emn as\-sem\-bly to \textsc{God}, and do no work.
\end{para}

\begin{para}
    \textsuperscript{16:9}\thinspace{}Sev\-en weeks af\-ter you start the grain har\-vest,
    \textsuperscript{10}\thinspace{}cel\-e\-brate the Feast of Weeks to \textsc{God} by giv\-ing a free\-will of\-fer\-ing in pro\-por\-tion to how \textsc{God} has bless\-ed you.
    \textsuperscript{11}\thinspace{}Re\-joice be\-fore \textsc{God}; you, your chil\-dren, your servants, and the Levites in your town; the for\-eign\-er, the or\-phan, and the wid\-ow, who are a\-mong you.
\end{para}

\begin{para}
    \textsuperscript{16:13}\thinspace{}Cel\-e\-brate the Feast of Tabernacles for sev\-en days af\-ter gath\-er\-ing the pro\-duce from your threshing floor and wine\-press.
    \textsuperscript{14}\thinspace{}Re\-joice dur\-ing your feast; you, your chil\-dren, your servants, and the Levites; the strang\-er, the or\-phan, and the wid\-ow who are a\-mong you.
    \textsuperscript{15}\thinspace{}Cel\-e\-brate the feast to \textsc{God} for sev\-en days, be\-cause \textsc{God} will bless you in all your pro\-duce and in all the work of your hands, so that you will be al\-to\-geth\-er joy\-ful.
\end{para}

\begin{para}
    \textsuperscript{16:16}\thinspace{}All males must ap\-pear three times a year be\-fore \textsc{God} where he chooses: at the Feast of Un\-leav\-ened Bread, the Feast of Weeks, and the Feast of Tabernacles. None shall ap\-pear be\-fore \textsc{God} empty-handed.
    \textsuperscript{17}\thinspace{}Each of you must bring a gift in pro\-por\-tion to how \textsc{God} has bless\-ed you.
\end{para}

\begin{para}
    \textsuperscript{16:18}\thinspace{}Ap\-point judges and officers for your tribes in all your towns. They shall judge the peo\-ple right\-eous\-ly.
    \textsuperscript{19}\thinspace{}Don't per\-vert jus\-tice or be par\-tial. Don't take a bribe, for it blinds the wise and twists the words of the right\-eous.
    \textsuperscript{20}\thinspace{}Pur\-sue jus\-tice, and on\-ly jus\-tice.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{17:1}\thinspace{}Don't sac\-ri\-fice to \textsc{God} an ox or a sheep with any blem\-ish or de\-fect, for that is de\-test\-a\-ble to \textsc{God}.
\end{para}

\begin{para}
    \textsuperscript{17:2}\thinspace{}If any\-one in any of your towns
    \textsuperscript{3}\thinspace{}has wor\-shiped oth\-er gods,
    \textsuperscript{4}\thinspace{}and if you hear a\-bout it, then in\-ves\-ti\-gate dil\-i\-gent\-ly. If it is true and cer\-tain,
    \textsuperscript{5}\thinspace{}bring him to your cit\-y gates and stone him to death.
    \textsuperscript{6}\thinspace{}He shall be put to death on the tes\-ti\-mo\-ny of two or three witnesses, but not on the tes\-ti\-mo\-ny of one wit\-ness.
    \textsuperscript{7}\thinspace{}The hands of the witnesses shall be first a\-gainst him to put him to death, and af\-ter\-ward the hands of all the peo\-ple.
\end{para}

\begin{para}
    \textsuperscript{17:8}\thinspace{}If a case is too dif\-fi\-cult for you to judge, wheth\-er hom\-i\-cide, law\-suit, or as\-sault, then go to the place \textsc{God} chooses,
    \textsuperscript{9}\thinspace{}and ask the priests or the judge who is in of\-fice, and they will give you the ver\-dict.
    \textsuperscript{10}\thinspace{}You must a\-bide by the ver\-dict they give you. Be care\-ful to do eve\-ry\-thing they tell you.
\end{para}

\begin{para}
    \textsuperscript{17:14}\thinspace{}When you en\-ter the land \textsc{God} is giv\-ing you, and you pos\-sess it, and live in it, and say, “I will set a king o\-ver me like all the nations a\-round me,”
    \textsuperscript{15}\thinspace{}ap\-point a king whom \textsc{God} chooses, one from a\-mong your brothers. Don't ap\-point a for\-eign\-er who is not your broth\-er.
    \textsuperscript{16}\thinspace{}He must not ac\-quire man\-y hors\-es for him\-self or send the peo\-ple to Egypt to ac\-quire man\-y hors\-es. For \textsc{God} has told you, “Nev\-er re\-turn that way a\-gain.”
    \textsuperscript{17}\thinspace{}He must not ac\-quire man\-y wives for him\-self, lest his heart go a\-stray. He must not ac\-quire ex\-ces\-sive sil\-ver and gold for him\-self.
\end{para}

\begin{para}
    \textsuperscript{17:18}\thinspace{}When he sits on the throne, he must write for him\-self a cop\-y of this law, in the pres\-ence of the Levites.
    \textsuperscript{19}\thinspace{}He must read from it all the days of his life, so he may learn to fear \textsc{God} by ob\-serv\-ing this law and these statutes,
    \textsuperscript{20}\thinspace{}lest his heart be lifted up a\-bove his countrymen or he turn a\-side from the com\-mand\-ment.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{18:9}\thinspace{}When you en\-ter the land which \textsc{God} is giv\-ing you, don't im\-i\-tate the de\-test\-a\-ble practices of those nations.
    \textsuperscript{10}\thinspace{}No one a\-mong you is to sac\-ri\-fice his child in the fire, prac\-tice div\-i\-na\-tion, tell fortunes, prac\-tice sor\-cer\-y,
    \textsuperscript{11}\thinspace{}cast spells, con\-sult a me\-di\-um or spir\-it\-ist, or in\-quire of the dead.
\end{para}

\begin{para}
    \textsuperscript{18:15}\thinspace{}\textsc{God} will raise up for you a proph\-et like me from a\-mong your countrymen. Lis\-ten to him.
    \textsuperscript{18}\thinspace{}I will put my words in his mouth, and he will tell them eve\-ry\-thing I com\-mand him.
    \textsuperscript{19}\thinspace{}I will hold ac\-count\-a\-ble those who do not lis\-ten to my words that he speaks in my name.
\end{para}

\begin{para}
    \textsuperscript{18:21}\thinspace{}You may ask your\-self, “How can we know when a mes\-sage is not from \textsc{God}?”
    \textsuperscript{22}\thinspace{}When a proph\-et speaks in \textsc{God}'s name and it does not come true, then \textsc{God} has not spo\-ken that mes\-sage. The proph\-et spoke pre\-sump\-tu\-ous\-ly. Don't be a\-fraid of him.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{19:2}\thinspace{}Set a\-part three cit\-ies for your\-selves in the mid\-dle of the land that \textsc{God} is giv\-ing you.
    \textsuperscript{3}\thinspace{}Di\-vide the land in\-to three parts, so that any man\-slay\-er can flee to these cit\-ies.
    \textsuperscript{4}\thinspace{}A man\-slay\-er is one who kills his neigh\-bor ac\-ci\-den\-tal\-ly with\-out pre\-vi\-ous hos\-til\-i\-ty.
    \textsuperscript{5}\thinspace{}He may flee to one of these cit\-ies and live,
    \textsuperscript{6}\thinspace{}lest the a\-veng\-er of blood pur\-sue the man\-slay\-er in an\-ger, and kill him, though he did not de\-serve death.
\end{para}

\begin{para}
    \textsuperscript{19:14}\thinspace{}Don't move your neighbor's bound\-a\-ry mark, established by your predecessors.
\end{para}

\begin{para}
    \textsuperscript{19:15}\thinspace{}One wit\-ness is in\-suf\-fi\-cient to es\-tab\-lish wrong\-do\-ing by any man. A charge must be established on the ev\-i\-dence of two or three witnesses.
    \textsuperscript{16}\thinspace{}If a ma\-li\-cious wit\-ness accuses some\-one of wrong\-do\-ing,
    \textsuperscript{17}\thinspace{}then both peo\-ple in the dis\-pute must ap\-pear be\-fore \textsc{God}, be\-fore the priests and the judges.
    \textsuperscript{18}\thinspace{}The judges must in\-ves\-ti\-gate thor\-ough\-ly, and if the wit\-ness has ac\-cused his broth\-er false\-ly,
    \textsuperscript{19}\thinspace{}then you must do to him as he in\-tend\-ed to do to his broth\-er.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{20:1}\thinspace{}When you go to war a\-gainst your enemies and see hors\-es, chariots, and an ar\-my larg\-er than yours, don't be a\-fraid of them, for \textsc{God}, who brought you out of Egypt, is with you.
    \textsuperscript{2}\thinspace{}When you ap\-proach the bat\-tle, the priest shall come for\-ward and speak to the ar\-my,
    \textsuperscript{3}\thinspace{}say\-ing, “Hear, O Israel, to\-day you ap\-proach bat\-tle a\-gainst your enemies. Don't be faint\-heart\-ed. Don't fear or pan\-ic or be ter\-ri\-fied of them,
    \textsuperscript{4}\thinspace{}for \textsc{God} goes with you to fight a\-gainst your enemies, to give you vic\-to\-ry.”
    \textsuperscript{5}\thinspace{}The officers shall speak to the ar\-my, say\-ing, “Has any man built a new house and not ded\-i\-cat\-ed it?
    \textsuperscript{6}\thinspace{}Has any man planted a vine\-yard and not yet enjoyed its fruit?
    \textsuperscript{7}\thinspace{}Has any man be\-come en\-gaged to a wom\-an and not yet mar\-ried her?
    \textsuperscript{8}\thinspace{}Is any man a\-fraid or faint\-heart\-ed? Let them re\-turn home.”
\end{para}

\begin{para}
    \textsuperscript{20:19}\thinspace{}When you be\-siege a cit\-y for a long time, don't de\-stroy its trees. Are the trees hu\-man that you should be\-siege them?
    \textsuperscript{20}\thinspace{}You may on\-ly cut down trees that you know to pro\-duce no food. Use them to con\-struct siegeworks a\-gainst the cit\-y un\-til it falls.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{21:15}\thinspace{}If a man has two wives, one loved and the oth\-er un\-loved, and both bear him sons, and the firstborn belongs to the un\-loved,
    \textsuperscript{16}\thinspace{}then, when his sons in\-her\-it what he has, he may not fa\-vor the son of the loved wife o\-ver the son of the un\-loved wife.
    \textsuperscript{17}\thinspace{}He must ac\-knowl\-edge the firstborn, the son of the un\-loved wife, by giv\-ing him a dou\-ble por\-tion of all he has. The right of the firstborn is his.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{22:1}\thinspace{}If you see your brother's an\-i\-mal straying, don't ig\-nore it. Bring it back to him.
    \textsuperscript{2}\thinspace{}If he lives far a\-way or you don't know who he is, bring it home and care for it un\-til your broth\-er looks for it, then re\-turn it to him.
    \textsuperscript{3}\thinspace{}Do the same with any\-thing else that your broth\-er loses and you find. Don't ig\-nore it.
    \textsuperscript{4}\thinspace{}Don't ig\-nore your brother's don\-key or ox when it collapses on the road. Help him lift it up a\-gain.
\end{para}

\begin{para}
    \textsuperscript{22:5}\thinspace{}A wom\-an shall not wear men's clothes, nei\-ther shall a man wear women's clothes, for \textsc{God} detests those who do it.
\end{para}

\begin{para}
    \textsuperscript{22:6}\thinspace{}If you come a\-cross a bird's nest with chicks or eggs, and the moth\-er is sit\-ting on them, don't take the moth\-er with the young.
    \textsuperscript{7}\thinspace{}You may take the young but let the moth\-er go so that you may pros\-per and have a long life.
\end{para}

\begin{para}
    \textsuperscript{22:8}\thinspace{}When you build a new house, con\-struct a rail\-ing a\-round your roof, lest you bring bloodguilt on your house if any\-one falls from it.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{23:9}\thinspace{}When you are encamped a\-gainst your enemies,
    \textsuperscript{12}\thinspace{}des\-ig\-nate a place out\-side the camp where you can go to re\-lieve your\-self.
    \textsuperscript{13}\thinspace{}Keep a dig\-ging tool in your e\-quip\-ment so that when you re\-lieve your\-self you can dig a hole and cov\-er your ex\-cre\-ment.
    \textsuperscript{14}\thinspace{}For \textsc{God} walks through your camp to pro\-tect you and to de\-feat your enemies. Your camp must be ho\-ly lest he see some\-thing in\-de\-cent a\-mong you and turn a\-way from you.
\end{para}

\begin{para}
    \textsuperscript{23:15}\thinspace{}Don't re\-turn a slave to his mas\-ter af\-ter he has es\-caped to you.
    \textsuperscript{16}\thinspace{}Let him live a\-mong you wher\-ev\-er he chooses. Don't mis\-treat him.
\end{para}

\begin{para}
    \textsuperscript{23:17}\thinspace{}No man or wom\-an of Israel shall be a shrine pros\-ti\-tute.
    \textsuperscript{18}\thinspace{}Don't bring the earn\-ings of a pros\-ti\-tute, fe\-male or male, in\-to the house of \textsc{God} to pay any vow, be\-cause both are de\-test\-a\-ble to \textsc{God}.
\end{para}

\begin{para}
    \textsuperscript{23:19}\thinspace{}Don't charge your broth\-er in\-ter\-est.
    \textsuperscript{20}\thinspace{}You may charge in\-ter\-est to a for\-eign\-er, but not to your broth\-er, so that \textsc{God} may bless you in eve\-ry\-thing you un\-der\-take.
\end{para}

\begin{para}
    \textsuperscript{23:21}\thinspace{}When you make a vow to \textsc{God}, don't de\-lay ful\-fil\-ling it, for \textsc{God} will sure\-ly re\-quire it of you, and you will be guilt\-y of sin.
    \textsuperscript{22}\thinspace{}But it's not a sin to re\-frain from mak\-ing a vow.
    \textsuperscript{23}\thinspace{}Be care\-ful to do what you say, be\-cause you made your prom\-ise free\-ly to \textsc{God} with your own mouth.
\end{para}

\begin{para}
    \textsuperscript{23:24}\thinspace{}If you en\-ter your neighbor's vine\-yard, you may eat your fill of grapes, but don't put any in your bas\-ket.
    \textsuperscript{25}\thinspace{}If you en\-ter your neighbor's stand\-ing grain, you may pluck grain with your hand, but don't har\-vest it with a sick\-le.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{24:1}\thinspace{}If a man marries a wom\-an, but she displeases him be\-cause he finds some in\-de\-cen\-cy in her, then let him write her a di\-vorce cer\-tif\-i\-cate, and send her out of his house.
    \textsuperscript{2}\thinspace{}When she leaves his house, she may mar\-ry an\-oth\-er man.
\end{para}

\begin{para}
    \textsuperscript{24:5}\thinspace{}When a man takes a new wife, he shall not go to war or be li\-a\-ble for any pub\-lic du\-ty for one year. He is free to stay home, so that he can bring joy to his new wife.
\end{para}

\begin{para}
    \textsuperscript{24:6}\thinspace{}Don't take a mill or an up\-per mill\-stone in pledge, for that would be tak\-ing a life in pledge.
\end{para}

\begin{para}
    \textsuperscript{24:10}\thinspace{}If you lend any\-thing to your neigh\-bor, don't en\-ter his house to col\-lect his pledge.
    \textsuperscript{11}\thinspace{}Re\-main out\-side and let your neigh\-bor bring the pledge out to you.
    \textsuperscript{12}\thinspace{}If the man is poor, don't keep his pledge o\-ver\-night.
    \textsuperscript{13}\thinspace{}Re\-turn it to him when the sun sets, so that he may sleep in his cloak and bless you.
\end{para}

\begin{para}
    \textsuperscript{24:14}\thinspace{}Don't op\-press a hired work\-er who is poor and needy, wheth\-er he is a broth\-er or a for\-eign\-er liv\-ing in your land.
    \textsuperscript{15}\thinspace{}Give him his wages each day be\-fore the sun sets, for he is poor and depends on them; lest he cry to \textsc{God} a\-gainst you, and you be\-come guilt\-y of sin.
\end{para}

\begin{para}
    \textsuperscript{24:16}\thinspace{}Fathers shall not be put to death for their chil\-dren, nor chil\-dren put to death for their fathers. Each one shall be put to death for his own sin.
\end{para}

\begin{para}
    \textsuperscript{24:17}\thinspace{}Don't de\-ny jus\-tice to a for\-eign\-er or an or\-phan, and don't take a widow's cloak as a pledge.
    \textsuperscript{18}\thinspace{}Re\-mem\-ber that you were a slave in Egypt, and that \textsc{God} redeemed you from there.
\end{para}

\begin{para}
    \textsuperscript{24:19}\thinspace{}When you reap your har\-vest and for\-get a sheaf in the field, don't go back to get it. Leave it for the for\-eign\-er, the or\-phan, and the wid\-ow, so that \textsc{God} may bless you in all the work of your hands.
    \textsuperscript{20}\thinspace{}When you beat the olives from your trees, don't go o\-ver the branches a\-gain.
    \textsuperscript{21}\thinspace{}When you gath\-er the grapes from your vine\-yard, don't go o\-ver the vines a\-gain. What re\-mains is for the for\-eign\-er, the or\-phan, and the wid\-ow.
    \textsuperscript{22}\thinspace{}Re\-mem\-ber that you were a slave in Egypt.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{25:4}\thinspace{}Don't muz\-zle the ox while it treads out the grain.
\end{para}

\begin{para}
    \textsuperscript{25:13}\thinspace{}Don't car\-ry differing weights in your bag, one heav\-y and one light.
    \textsuperscript{14}\thinspace{}Don't have differing meas\-ures in your house, one large and one small.
    \textsuperscript{15}\thinspace{}Use ac\-cu\-rate and hon\-est weights and meas\-ures.
    \textsuperscript{16}\thinspace{}For \textsc{God} detests those who are dis\-hon\-est in such things.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{26:2}\thinspace{}Put in\-to a bas\-ket some of the first fruit of the har\-vest.
    \textsuperscript{3}\thinspace{}Go to the priest.
    \textsuperscript{4}\thinspace{}The priest shall take the bas\-ket from your hand and set it down be\-fore the al\-tar of \textsc{God}.
    \textsuperscript{5}\thinspace{}Say be\-fore \textsc{God}, “My fa\-ther was a wan\-der\-ing Aramean. He went down to Egypt with a few peo\-ple, sojourned there, and be\-came a great, might\-y, and pop\-u\-lous na\-tion.
    \textsuperscript{6}\thinspace{}The Egyptians mistreated us, hu\-mil\-i\-at\-ed us, and im\-posed hard la\-bor on us.
    \textsuperscript{7}\thinspace{}Then we cried to \textsc{God}, the god of our fathers, and \textsc{God} heard our voice and saw our af\-flic\-tion, our toil, and our op\-pres\-sion,
    \textsuperscript{8}\thinspace{}\textsc{God} brought us out of Egypt with a might\-y hand, an out\-stretched arm, great deeds of ter\-ror, signs, and wonders.
    \textsuperscript{9}\thinspace{}He brought us to this place and gave us this land, a land flow\-ing with milk and hon\-ey.
    \textsuperscript{10}\thinspace{}I bring the first fruit of the ground, which you, O \textsc{God}, have giv\-en me.” Set it be\-fore \textsc{God} and wor\-ship be\-fore \textsc{God}.
    \textsuperscript{11}\thinspace{}You, the Levite, and the for\-eign\-er shall re\-joice in all the good things which \textsc{God} has giv\-en to you and your house\-hold.
\end{para}

\begin{para}
    \textsuperscript{26:12}\thinspace{}When you have fin\-ished pay\-ing all the tithes of your pro\-duce in the third year, give it to the Levite, the for\-eign\-er, the or\-phan, and the wid\-ow, so that they may eat in your towns and be sat\-is\-fied.
    \textsuperscript{13}\thinspace{}Say be\-fore \textsc{God}, “I have tak\-en the sa\-cred por\-tion out of my house. I have giv\-en it to the Levite, the for\-eign\-er, the or\-phan, and the wid\-ow, ac\-cord\-ing to all the commandments you gave me. I have not transgressed or for\-got\-ten your commandments.
    \textsuperscript{14}\thinspace{}I have not eat\-en any of the tithe while in mourn\-ing, or re\-moved any of it while un\-clean, or offered any of it to the dead. I have obeyed \textsc{God}. I have done all you commanded me.
    \textsuperscript{15}\thinspace{}Look down from heav\-en, your ho\-ly hab\-i\-ta\-tion, and bless your peo\-ple Israel and the land you have giv\-en us, as you swore to our fathers, a land flow\-ing with milk and hon\-ey.”
\end{para}

\begin{para}
\end{para}


\end{document}

% Copyright 2021 Michael Hendricks
%
% Permission is granted to use, copy, modify, and distribute
% this work for any purpose with or without fee.
%
% This work is provided as-is without any express or implied
% warranty. In no event is the author liable for any damages
% arising from this work.
