%
% see tex.stackexchange.com/a/600/151276
\def\changemargin#1#2{\list{}{\rightmargin#2\leftmargin#1}\item[]}
\let\endchangemargin=\endlist

\documentclass[openany,12pt,english]{book}
\usepackage[utf8]{inputenc}
\usepackage{babel}
\usepackage{microtype}
\usepackage{csquotes}

\usepackage[
  paperwidth=8.5in,
  paperheight=11in,
  top=1in,
  outer=1in,
  bottom=1.25in,
  inner=1.5in
]{geometry}

\title{Laozi}
\author{}
\date{}

\newenvironment{para}{\par\pretolerance=100\tolerance=200\setlength{\emergencystretch}{0.6em}\relax}{\par}

\begin{document}

\maketitle

% Copyright page
{
\thispagestyle{empty}
\vspace*{\fill}
\begin{changemargin}{.8in}{.8in}
\begin{center}
\textcopyright{}~2021 Michael Hendricks

\vspace{0.5\baselineskip}

Permission is granted to use, copy, modify, and distribute
this work for any purpose with or without fee.

\vspace{0.5\baselineskip}

This work is provided as-is without any express or implied
warranty. In no event is the author liable for any damages
arising from this work.
\end{center}
\end{changemargin}
}
\pagenumbering{arabic}
\clearpage{}

% Table of Contents
\clearpage{}

\markboth{}{}
\twocolumn[\begin{changemargin}{.6in}{.6in}
  \begin{center}
    \itshape
    An a\-bridg\-ment of the Daodejing.
    Verse references such
    as~\textsuperscript{2:3}\thinspace{}indicate
    the original source of the text, but
    don't imply a full quotation.
  \end{center}
\end{changemargin}]{}


\begin{para}
    \textsuperscript{1:1}\thinspace{}The Way that can be put in\-to words is not the con\-stant Way.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{2:1}\thinspace{}As soon as eve\-ry\-one in the world knows beau\-ty as beau\-ty, there is al\-read\-y ug\-li\-ness.
    \textsuperscript{2}\thinspace{}As soon as eve\-ry\-one knows good as good, there is bad.
\end{para}

\begin{para}
    \textsuperscript{2:3}\thinspace{}Be\-ing and non\-be\-ing pro\-duce each oth\-er;
    \textsuperscript{4}\thinspace{}dif\-fi\-cult and eas\-y com\-plete each oth\-er;
    \textsuperscript{5}\thinspace{}long and short shape each oth\-er;
    \textsuperscript{6}\thinspace{}high and low fill each oth\-er;
    \textsuperscript{7}\thinspace{}note and voice har\-mo\-nize with each oth\-er;
    \textsuperscript{8}\thinspace{}front and back fol\-low each oth\-er:
    \textsuperscript{9}\thinspace{}these are con\-stant.
\end{para}

\begin{para}
    \textsuperscript{2:10}\thinspace{}There\-fore the sage dwells in the work of nonforced ac\-tion,
    \textsuperscript{11}\thinspace{}and practices word\-less teach\-ing.
    \textsuperscript{12}\thinspace{}All things a\-rise, but he does\-n't start them;
    \textsuperscript{13}\thinspace{}he acts on their be\-half, but does\-n't make them de\-pend\-ent;
    \textsuperscript{14}\thinspace{}he succeeds, but does\-n't dwell on it.
    \textsuperscript{15}\thinspace{}It's on\-ly be\-cause he does\-n't dwell on it, that it does\-n't leave.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{3:1}\thinspace{}By not el\-e\-vat\-ing the wor\-thy, you keep the peo\-ple from con\-ten\-tion.
    \textsuperscript{2}\thinspace{}By not valuing rare treasures, you keep the peo\-ple from steal\-ing.
    \textsuperscript{3}\thinspace{}By not displaying what is de\-sir\-a\-ble, you keep the peo\-ple from con\-fu\-sion.
\end{para}

\begin{para}
    \textsuperscript{3:4}\thinspace{}There\-fore the sage governs by clear\-ing the people's minds, but fill\-ing their bellies; weakening their ambitions, but strengthening their bones.
\end{para}

\begin{para}
    \textsuperscript{3:5}\thinspace{}He con\-stant\-ly keeps the peo\-ple with\-out cun\-ning and with\-out desires
    \textsuperscript{6}\thinspace{}so that the intellectuals won't dare to med\-dle.
    \textsuperscript{7}\thinspace{}Act with\-out forc\-ing, then eve\-ry\-thing is governed prop\-er\-ly.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{7:5}\thinspace{}The sage puts him\-self in the back\-ground yet finds him\-self in the fore\-ground;
    \textsuperscript{6}\thinspace{}puts him\-self out of mind yet finds him\-self cared for.
    \textsuperscript{7}\thinspace{}Is\-n't it be\-cause he's self\-less that
    \textsuperscript{8}\thinspace{}he's a\-ble to a\-chieve his self-interest?
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{8:1}\thinspace{}The highest good is like wa\-ter.
    \textsuperscript{2}\thinspace{}Water's good\-ness is that it benefits all things,
    \textsuperscript{3}\thinspace{}yet it strives to dwell in the places that the crowd hates,
    \textsuperscript{4}\thinspace{}there\-fore it co\-mes close to the Way.
\end{para}

\begin{para}
    \textsuperscript{8:6}\thinspace{}In a dwell\-ing, its good\-ness is lo\-ca\-tion;
    \textsuperscript{7}\thinspace{}in a mind, depth;
    \textsuperscript{8}\thinspace{}in giv\-ing, be\-ing like Heav\-en;
    \textsuperscript{9}\thinspace{}in speak\-ing, sin\-cer\-i\-ty;
    \textsuperscript{10}\thinspace{}in gov\-ern\-ment, or\-der;
    \textsuperscript{11}\thinspace{}in han\-dling af\-fairs, a\-bil\-i\-ty;
    \textsuperscript{12}\thinspace{}in ac\-tion, time\-li\-ness.
\end{para}

\begin{para}
    \textsuperscript{8:13}\thinspace{}Be\-cause there is no con\-ten\-tion, there is no blame.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{9:1}\thinspace{}If you hold and o\-ver\-fill a cup, it's not as good as stop\-ping short.
    \textsuperscript{2}\thinspace{}If you ham\-mer and oversharpen a blade, you can't main\-tain it for long.
    \textsuperscript{3}\thinspace{}If gold and jade fill your rooms, no one is a\-ble to pro\-tect it.
    \textsuperscript{4}\thinspace{}If sta\-tus and wealth lead to ar\-ro\-gance, you bring dis\-as\-ter up\-on your\-self.
\end{para}

\begin{para}
    \textsuperscript{9:5}\thinspace{}When the work is done, with\-draw; this is the Way of Heav\-en.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{11:1}\thinspace{}Thir\-ty spokes come to\-geth\-er at one hub,
    \textsuperscript{2}\thinspace{}but it's the emp\-ti\-ness in\-side the hub that makes it use\-ful to the cart.
    \textsuperscript{3}\thinspace{}We form clay in\-to a pot,
    \textsuperscript{4}\thinspace{}but it's the emp\-ti\-ness in\-side that makes the clay pot use\-ful.
    \textsuperscript{5}\thinspace{}We chis\-el out doors and windows to make a room,
    \textsuperscript{6}\thinspace{}but it's the emp\-ti\-ness in\-side that makes the room use\-ful.
\end{para}

\begin{para}
    \textsuperscript{11:7}\thinspace{}There\-fore, what is pres\-ent brings ben\-e\-fit,
    \textsuperscript{8}\thinspace{}but what is ab\-sent makes it use\-ful.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{15:1}\thinspace{}In an\-tiq\-ui\-ty, those who were good at the Way
    \textsuperscript{2}\thinspace{}were sub\-tle, won\-der\-ful, mys\-te\-ri\-ous, pen\-e\-trat\-ing, and too deep to know.
    \textsuperscript{3}\thinspace{}Be\-cause they can't be known, we're forced to de\-scribe their ap\-pear\-ance:
    \textsuperscript{4}\thinspace{}Hes\-i\-tant, as if cross\-ing a riv\-er in win\-ter.
    \textsuperscript{5}\thinspace{}A\-lert, as if fearing the neighbors on all sides.
    \textsuperscript{6}\thinspace{}Sol\-emn, like a visiting guest.
    \textsuperscript{7}\thinspace{}Yield\-ing, like ice a\-bout to melt.
    \textsuperscript{8}\thinspace{}Gen\-u\-ine, like un\-carved wood.
    \textsuperscript{9}\thinspace{}O\-paque, like mud\-dy wa\-ter.
    \textsuperscript{10}\thinspace{}Vast, like a val\-ley.
\end{para}

\begin{para}
    \textsuperscript{15:11}\thinspace{}If mud\-dy wa\-ter is still, it grad\-u\-al\-ly becomes clear.
    \textsuperscript{12}\thinspace{}If some\-thing at rest is moved, it grad\-u\-al\-ly co\-mes to life.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{17:1}\thinspace{}With the ver\-y best rulers, those be\-low on\-ly know that the rulers ex\-ist,
    \textsuperscript{2}\thinspace{}the next best are the rulers they love and praise,
    \textsuperscript{3}\thinspace{}next are the rulers they fear,
    \textsuperscript{4}\thinspace{}and the worst are the rulers they des\-pise.
\end{para}

\begin{para}
    \textsuperscript{17:7}\thinspace{}When the task is ac\-com\-plished and the work is done, all the com\-mon peo\-ple say, “We did it our\-selves.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{18:1}\thinspace{}It's when the great Way is a\-ban\-doned that be\-nev\-o\-lence and right\-eous\-ness ap\-pear.
    \textsuperscript{2}\thinspace{}It's when learn\-ing and clev\-er\-ness e\-merge that the great hy\-poc\-ri\-sy appears.
    \textsuperscript{3}\thinspace{}It's when fam\-i\-ly relationships are not har\-mo\-ni\-ous that fil\-ial pi\-e\-ty and pa\-ren\-tal com\-pas\-sion ap\-pear.
    \textsuperscript{4}\thinspace{}It's when the coun\-try is con\-fused and cha\-ot\-ic that up\-right officials ap\-pear.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{19:5}\thinspace{}Dem\-on\-strate plain\-ness, em\-brace sim\-plic\-i\-ty, have lit\-tle self\-ish\-ness and few desires.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{22:1}\thinspace{}Bent o\-ver then pre\-served whole;
    \textsuperscript{2}\thinspace{}crook\-ed then straight;
    \textsuperscript{3}\thinspace{}hol\-low then filled;
    \textsuperscript{4}\thinspace{}worn out then renewed;
    \textsuperscript{5}\thinspace{}hav\-ing lit\-tle then sat\-is\-fied;
    \textsuperscript{6}\thinspace{}hav\-ing much then con\-fused.
    \textsuperscript{7}\thinspace{}There\-fore the sage holds to the one and becomes shep\-herd to the world.
\end{para}

\begin{para}
    \textsuperscript{22:8}\thinspace{}He does\-n't pro\-mote him\-self, so he's ac\-knowl\-edged.
    \textsuperscript{9}\thinspace{}He does\-n't dis\-play him\-self, so he shines bright\-ly.
    \textsuperscript{10}\thinspace{}He does\-n't brag a\-bout him\-self, so he has mer\-it.
    \textsuperscript{11}\thinspace{}He's not boast\-ful, so he can en\-dure.
    \textsuperscript{12}\thinspace{}It's on\-ly be\-cause he does\-n't con\-tend that no one can con\-tend with him.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{23:1}\thinspace{}It's nat\-u\-ral to speak rare\-ly.
    \textsuperscript{2}\thinspace{}A vi\-o\-lent wind does\-n't last all morn\-ing.
    \textsuperscript{3}\thinspace{}A tor\-ren\-tial rain does\-n't last all day.
    \textsuperscript{5}\thinspace{}If heav\-en and earth can't sus\-tain it for long, how much less can man?
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{24:1}\thinspace{}He who blows hard can't stand.
    \textsuperscript{2}\thinspace{}He who promotes him\-self is\-n't ac\-knowl\-edged.
    \textsuperscript{3}\thinspace{}He who displays him\-self does\-n't shine bright\-ly.
    \textsuperscript{5}\thinspace{}He who brags a\-bout him\-self has no mer\-it.
    \textsuperscript{6}\thinspace{}He who boasts a\-bout him\-self won't en\-dure.
    \textsuperscript{8}\thinspace{}These are so de\-spised that e\-ven those who want things don't a\-bide them.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{27:1}\thinspace{}A skill\-ful trav\-el\-er leaves no ruts or tracks.
    \textsuperscript{2}\thinspace{}A skill\-ful speak\-er makes no slips or gaffes.
    \textsuperscript{3}\thinspace{}A skill\-ful count\-er needs no tallies or papers.
\end{para}

\begin{para}
    \textsuperscript{27:6}\thinspace{}The sage is al\-ways skill\-ful at sav\-ing peo\-ple, and nev\-er abandons any\-one;
    \textsuperscript{7}\thinspace{}with things, he nev\-er abandons val\-u\-a\-ble prop\-er\-ty.
\end{para}

\begin{para}
    \textsuperscript{27:9}\thinspace{}Skill\-ful peo\-ple are the teachers of skill\-ful peo\-ple;
    \textsuperscript{10}\thinspace{}un\-skilled peo\-ple are the raw ma\-te\-ri\-als of skill\-ful peo\-ple.
    \textsuperscript{11}\thinspace{}He who does\-n't val\-ue his teach\-er and cher\-ish his raw ma\-te\-ri\-als is great\-ly con\-fused, no mat\-ter how clev\-er he is.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{29:1}\thinspace{}Those who want to take the world and rule it,
    \textsuperscript{2}\thinspace{}I see that they won't suc\-ceed.
    \textsuperscript{3}\thinspace{}The world is a sa\-cred ves\-sel,
    \textsuperscript{4}\thinspace{}and it can't be ruled.
    \textsuperscript{5}\thinspace{}Those who rule it ru\-in it.
    \textsuperscript{6}\thinspace{}Those who hold it lose it.
\end{para}

\begin{para}
    \textsuperscript{29:7}\thinspace{}It's the na\-ture of things that some lead and some fol\-low,
    \textsuperscript{8}\thinspace{}some blow hot and some blow cold,
    \textsuperscript{9}\thinspace{}some are strong and some are weak,
    \textsuperscript{10}\thinspace{}some cul\-ti\-vate and some de\-te\-ri\-o\-rate.
    \textsuperscript{11}\thinspace{}There\-fore the sage avoids extremes, ar\-ro\-gance, and ex\-trav\-a\-gance.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{30:1}\thinspace{}Those with the Way who as\-sist the rul\-er don't use the force of arms to con\-quer the world;
    \textsuperscript{2}\thinspace{}such methods would on\-ly back\-fire.
    \textsuperscript{3}\thinspace{}Wher\-ev\-er armies camp, brambles and thorns will grow.
\end{para}

\begin{para}
    \textsuperscript{30:5}\thinspace{}A skill\-ful lead\-er achieves results then stops,
    \textsuperscript{6}\thinspace{}and does\-n't use the oc\-ca\-sion to sieze pow\-er.
    \textsuperscript{7}\thinspace{}He achieves results with\-out ar\-ro\-gance.
    \textsuperscript{8}\thinspace{}He achieves results with\-out prais\-ing his deeds.
    \textsuperscript{9}\thinspace{}He achieves results with\-out boasting.
    \textsuperscript{10}\thinspace{}He achieves results on\-ly as a last re\-sort.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{31:1}\thinspace{}The mil\-i\-tar\-y is an om\-i\-nous in\-stru\-ment.
    \textsuperscript{2}\thinspace{}It's so de\-spised that e\-ven those who want things don't a\-bide it.
\end{para}

\begin{para}
    \textsuperscript{31:5}\thinspace{}The mil\-i\-tar\-y is an om\-i\-nous in\-stru\-ment,
    \textsuperscript{6}\thinspace{}so on\-ly use it as a last re\-sort;
    \textsuperscript{7}\thinspace{}with calm de\-tach\-ment is best.
    \textsuperscript{8}\thinspace{}Don't glo\-ri\-fy it;
    \textsuperscript{9}\thinspace{}to glo\-ri\-fy it would be to de\-light in kill\-ing peo\-ple.
    \textsuperscript{10}\thinspace{}He who delights in kill\-ing peo\-ple can't a\-chieve his ambitions in the world.
    \textsuperscript{16}\thinspace{}When man\-y peo\-ple have been killed,
    \textsuperscript{17}\thinspace{}weep for them with sor\-row.
    \textsuperscript{18}\thinspace{}When the war has been won, treat it like a fu\-ner\-al cer\-e\-mo\-ny.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{33:1}\thinspace{}He who knows others is in\-tel\-li\-gent;
    \textsuperscript{2}\thinspace{}he who knows him\-self is enlightened.
    \textsuperscript{3}\thinspace{}He who overcomes others is force\-ful;
    \textsuperscript{4}\thinspace{}he who overcomes him\-self is pow\-er\-ful.
    \textsuperscript{5}\thinspace{}He who knows con\-tent\-ment is rich.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{34:1}\thinspace{}The Way floats and drifts;
    \textsuperscript{2}\thinspace{}it can go left or right.
    \textsuperscript{4}\thinspace{}When the task is ac\-com\-plished and the work is done, it claims no cred\-it.
\end{para}

\begin{para}
    \textsuperscript{34:5}\thinspace{}All things turn to it
    \textsuperscript{6}\thinspace{}and yet it does\-n't act as their mas\-ter,
    \textsuperscript{8}\thinspace{}so it can be called a small thing.
\end{para}

\begin{para}
    \textsuperscript{34:9}\thinspace{}All things turn to it
    \textsuperscript{10}\thinspace{}and yet it does\-n't act as their mas\-ter,
    \textsuperscript{11}\thinspace{}so it can be called a great thing.
\end{para}

\begin{para}
    \textsuperscript{34:12}\thinspace{}That's why the sage is a\-ble to ac\-com\-plish great things:
    \textsuperscript{13}\thinspace{}he does\-n't try to do great things, there\-fore he accomplishes great things.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{35:3}\thinspace{}Com\-ing to you, the world suffers no harm, but finds safe\-ty and peace.
\end{para}

\begin{para}
    \textsuperscript{35:4}\thinspace{}Pass\-ing travelers will stop for mu\-sic and food.
    \textsuperscript{5}\thinspace{}But if the Way were de\-scribed,
    \textsuperscript{6}\thinspace{}it would seem in\-sip\-id and with\-out fla\-vor.
    \textsuperscript{7}\thinspace{}When you look at it, it's not suf\-fi\-cient to be seen;
    \textsuperscript{8}\thinspace{}when you lis\-ten to it, it's not suf\-fi\-cient to be heard;
    \textsuperscript{9}\thinspace{}yet when you use it, it can't be exhausted.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{36:1}\thinspace{}If you want to shrink some\-thing,
    \textsuperscript{2}\thinspace{}it must first be ex\-pand\-ed;
    \textsuperscript{3}\thinspace{}if you want to weak\-en some\-thing,
    \textsuperscript{4}\thinspace{}it must first be strengthened;
    \textsuperscript{5}\thinspace{}if you want to leave some\-thing,
    \textsuperscript{6}\thinspace{}it must first be joined;
    \textsuperscript{7}\thinspace{}if you want to take some\-thing a\-way,
    \textsuperscript{8}\thinspace{}it must first be giv\-en.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{38:1}\thinspace{}A man of su\-pe\-ri\-or vir\-tue does\-n't cling to vir\-tue,
    \textsuperscript{2}\thinspace{}there\-fore he has true vir\-tue.
    \textsuperscript{5}\thinspace{}A man of su\-pe\-ri\-or vir\-tue acts with\-out force,
    \textsuperscript{6}\thinspace{}and with\-out an a\-gen\-da.
\end{para}

\begin{para}
    \textsuperscript{38:16}\thinspace{}There\-fore, af\-ter los\-ing the Way there is vir\-tue;
    \textsuperscript{17}\thinspace{}af\-ter los\-ing vir\-tue there is be\-nev\-o\-lence;
    \textsuperscript{18}\thinspace{}af\-ter los\-ing be\-nev\-o\-lence there is right\-eous\-ness;
    \textsuperscript{19}\thinspace{}af\-ter los\-ing right\-eous\-ness there is pro\-pri\-e\-ty.
\end{para}

\begin{para}
    \textsuperscript{38:20}\thinspace{}As for pro\-pri\-e\-ty, it's a thin ve\-neer of loy\-al\-ty and sin\-cer\-i\-ty,
    \textsuperscript{21}\thinspace{}and the be\-gin\-ning of dis\-or\-der.
\end{para}

\begin{para}
    \textsuperscript{38:24}\thinspace{}There\-fore a tru\-ly great man dwells in the thick sub\-stance,
    \textsuperscript{25}\thinspace{}not in the thin ve\-neer;
    \textsuperscript{26}\thinspace{}in the fruit, not in the flow\-er.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{43:1}\thinspace{}The softest, most pli\-a\-ble thing in the world
    \textsuperscript{2}\thinspace{}runs rough\-shod o\-ver the hardest, most sol\-id thing in the world.
    \textsuperscript{3}\thinspace{}That which has no sub\-stance enters in\-to that which has no openings.
    \textsuperscript{4}\thinspace{}That's how I know the benefits of nonforced ac\-tion.
\end{para}

\begin{para}
    \textsuperscript{43:5}\thinspace{}Teachings that use no words,
    \textsuperscript{6}\thinspace{}and the benefits of nonforced ac\-tion:
    \textsuperscript{7}\thinspace{}few in the world can un\-der\-stand them.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{44:4}\thinspace{}Ex\-ces\-sive at\-tach\-ment co\-mes at great cost,
    \textsuperscript{5}\thinspace{}and great ac\-cu\-mu\-la\-tion leads to heav\-y loss.
    \textsuperscript{6}\thinspace{}There\-fore, know\-ing con\-tent\-ment avoids dis\-grace,
    \textsuperscript{7}\thinspace{}and know\-ing when to stop avoids dan\-ger;
    \textsuperscript{8}\thinspace{}thus you can en\-dure for a long time.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{46:1}\thinspace{}When the world has the Way,
    \textsuperscript{2}\thinspace{}fi\-ne hors\-es are rel\-e\-gat\-ed to fer\-ti\-liz\-ing the fields.
    \textsuperscript{3}\thinspace{}When the world lacks the Way,
    \textsuperscript{4}\thinspace{}warhorses are bred just out\-side the cit\-y walls.
\end{para}

\begin{para}
    \textsuperscript{46:5}\thinspace{}There is no greater crime than de\-sire.
    \textsuperscript{6}\thinspace{}There is no greater dis\-as\-ter than be\-ing dis\-con\-tent.
    \textsuperscript{7}\thinspace{}There is no greater fault than want\-ing more.
    \textsuperscript{8}\thinspace{}There\-fore the sat\-is\-fac\-tion that co\-mes from be\-ing con\-tent is last\-ing sat\-is\-fac\-tion.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{47:1}\thinspace{}You don't have to go out the door to know the world.
    \textsuperscript{2}\thinspace{}You don't have to peek out the win\-dow to know the Way of Heav\-en.
    \textsuperscript{3}\thinspace{}The far\-ther you go, the less you know.
\end{para}

\begin{para}
    \textsuperscript{47:4}\thinspace{}There\-fore the sage does\-n't trav\-el but knows;
    \textsuperscript{5}\thinspace{}does\-n't look, but discerns;
    \textsuperscript{6}\thinspace{}and does\-n't do, but succeeds.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{48:1}\thinspace{}He who pursues learn\-ing gains some\-thing each day.
    \textsuperscript{2}\thinspace{}He who hears the Way loses some\-thing each day.
    \textsuperscript{3}\thinspace{}He loses, and loses a\-gain, un\-til he acts with\-out force.
    \textsuperscript{4}\thinspace{}He acts with\-out force and yet noth\-ing is left un\-done.
\end{para}

\begin{para}
    \textsuperscript{48:5}\thinspace{}If you want to rule the world, con\-stant\-ly a\-void in\-ter\-fer\-ing with it.
    \textsuperscript{6}\thinspace{}If you in\-ter\-fere, you're un\-fit to rule the world.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{49:1}\thinspace{}The sage con\-stant\-ly has no mind of his own.
    \textsuperscript{2}\thinspace{}He takes the mind of the peo\-ple as his mind.
\end{para}

\begin{para}
    \textsuperscript{49:3}\thinspace{}He's good to good peo\-ple
    \textsuperscript{4}\thinspace{}and al\-so good to bad peo\-ple,
    \textsuperscript{5}\thinspace{}thus he gains in good\-ness.
    \textsuperscript{6}\thinspace{}He trusts trust\-wor\-thy peo\-ple
    \textsuperscript{7}\thinspace{}and al\-so trusts un\-trust\-wor\-thy peo\-ple,
    \textsuperscript{8}\thinspace{}thus he gains in trust.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{53:1}\thinspace{}If I had the least bit of knowl\-edge,
    \textsuperscript{2}\thinspace{}I would walk on the great Way
    \textsuperscript{3}\thinspace{}and my on\-ly fear would be straying from it.
    \textsuperscript{4}\thinspace{}The great Way is ver\-y much lev\-el,
    \textsuperscript{5}\thinspace{}but peo\-ple ver\-y much pre\-fer moun\-tain trails.
\end{para}

\begin{para}
    \textsuperscript{53:6}\thinspace{}The court is full of splen\-dor,
    \textsuperscript{7}\thinspace{}but the fields are full of weeds,
    \textsuperscript{8}\thinspace{}and the granaries are full of noth\-ing;
    \textsuperscript{9}\thinspace{}their clothes are embroidered and col\-or\-ful,
    \textsuperscript{10}\thinspace{}their belts have sharp swords,
    \textsuperscript{11}\thinspace{}they're stuffed with food,
    \textsuperscript{12}\thinspace{}and have mon\-ey and possessions in ex\-cess:
    \textsuperscript{13}\thinspace{}this is called rob\-ber\-y,
    \textsuperscript{14}\thinspace{}and rob\-ber\-y is cer\-tain\-ly not the Way.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{56:1}\thinspace{}Those who know it, don't talk a\-bout it;
    \textsuperscript{2}\thinspace{}those who talk a\-bout it, don't know it.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{57:1}\thinspace{}Gov\-ern the state with jus\-tice,
    \textsuperscript{2}\thinspace{}de\-ploy the mil\-i\-tar\-y with sur\-prise,
    \textsuperscript{3}\thinspace{}and take the world with\-out in\-ter\-fer\-ence.
    \textsuperscript{4}\thinspace{}How do I know this? From the fol\-low\-ing:
\end{para}

\begin{para}
    \textsuperscript{57:5}\thinspace{}The world has man\-y prohibitions and taboos,
    \textsuperscript{6}\thinspace{}and yet the peo\-ple are full of pov\-er\-ty.
    \textsuperscript{7}\thinspace{}The peo\-ple have man\-y ben\-e\-fi\-cial tools,
    \textsuperscript{8}\thinspace{}and yet the state and fam\-i\-ly are in\-creas\-ing\-ly faint.
    \textsuperscript{9}\thinspace{}Peo\-ple have man\-y clev\-er tricks,
    \textsuperscript{10}\thinspace{}and yet strange things in\-creas\-ing\-ly hap\-pen.
    \textsuperscript{11}\thinspace{}The laws are in\-creas\-ing\-ly pub\-li\-cized,
    \textsuperscript{12}\thinspace{}and yet there are man\-y thieves and robbers.
\end{para}

\begin{para}
    \textsuperscript{57:13}\thinspace{}There\-fore, in the words of the sage:
    \textsuperscript{14}\thinspace{}“I act with\-out force, and the peo\-ple trans\-form them\-selves.
    \textsuperscript{15}\thinspace{}I en\-joy qui\-et, and the peo\-ple cor\-rect them\-selves.
    \textsuperscript{16}\thinspace{}I don't intefere, and the peo\-ple pros\-per them\-selves.
    \textsuperscript{17}\thinspace{}I have no desires, and the peo\-ple sim\-pli\-fy them\-selves.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{58:5}\thinspace{}Good for\-tune perch\-es on mis\-for\-tune;
    \textsuperscript{6}\thinspace{}mis\-for\-tune lurks be\-neath good for\-tune.
    \textsuperscript{7}\thinspace{}Who knows where it will end?
\end{para}

\begin{para}
    \textsuperscript{58:8}\thinspace{}Noth\-ing is ob\-vi\-ous.
    \textsuperscript{9}\thinspace{}Ob\-vi\-ous a\-gain becomes sur\-pris\-ing;
    \textsuperscript{10}\thinspace{}good a\-gain becomes mon\-strous.
    \textsuperscript{11}\thinspace{}The peo\-ple have been con\-fused for a long time.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{59:1}\thinspace{}In governing the peo\-ple and serv\-ing Heav\-en,
    \textsuperscript{2}\thinspace{}there is noth\-ing like fru\-gal\-i\-ty.
    \textsuperscript{3}\thinspace{}On\-ly with fru\-gal\-i\-ty can you sub\-mit ear\-ly.
    \textsuperscript{4}\thinspace{}Sub\-mit\-ting ear\-ly is called re\-peat\-ed\-ly ac\-cu\-mu\-lat\-ing vir\-tue.
    \textsuperscript{5}\thinspace{}If you re\-peat\-ed\-ly ac\-cu\-mu\-late vir\-tue, then there is noth\-ing you can't o\-ver\-come.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{62:2}\thinspace{}The Way is the treas\-ure of a good man,
    \textsuperscript{3}\thinspace{}and the shel\-ter of a bad man.
\end{para}

\begin{para}
    \textsuperscript{62:4}\thinspace{}Beau\-ti\-ful words can buy hon\-or,
    \textsuperscript{5}\thinspace{}and good deeds can raise you a\-bove others.
\end{para}

\begin{para}
    \textsuperscript{62:6}\thinspace{}E\-ven if a man is bad, how can you a\-ban\-don him?
\end{para}

\begin{para}
    \textsuperscript{62:7}\thinspace{}At the emperor's cor\-o\-na\-tion,
    \textsuperscript{9}\thinspace{}of\-fer\-ing jade discs brought in by a team of hors\-es
    \textsuperscript{10}\thinspace{}is not as good as sit\-ting still and of\-fer\-ing this Way.
\end{para}

\begin{para}
    \textsuperscript{62:11}\thinspace{}Why was this Way so val\-u\-a\-ble to the ancients?
    \textsuperscript{12}\thinspace{}Is\-n't it said, “Those who seek, with this will ob\-tain,
    \textsuperscript{13}\thinspace{}and those who have guilt, with this will es\-cape”?
    \textsuperscript{14}\thinspace{}There\-fore it's the most val\-u\-a\-ble thing in the world.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{63:1}\thinspace{}Act with\-out forc\-ing,
    \textsuperscript{2}\thinspace{}serve with\-out in\-ter\-fer\-ing,
    \textsuperscript{3}\thinspace{}sa\-vor the fla\-vor\-less,
    \textsuperscript{4}\thinspace{}con\-sid\-er the small as great and the few as man\-y,
    \textsuperscript{5}\thinspace{}re\-spond to re\-sent\-ment with vir\-tue.
\end{para}

\begin{para}
    \textsuperscript{63:6}\thinspace{}Plan for some\-thing dif\-fi\-cult while it's still eas\-y.
    \textsuperscript{7}\thinspace{}Deal with some\-thing great while it's still small.
    \textsuperscript{8}\thinspace{}The dif\-fi\-cult things of the world start off eas\-y.
    \textsuperscript{9}\thinspace{}The great things of the world start off small.
    \textsuperscript{10}\thinspace{}There\-fore, the sage nev\-er does great things,
    \textsuperscript{11}\thinspace{}and thus he is a\-ble to ac\-com\-plish great things.
\end{para}

\begin{para}
    \textsuperscript{63:12}\thinspace{}He who makes promises light\-ly is sure to have lit\-tle trust,
    \textsuperscript{13}\thinspace{}and he who treats man\-y things as eas\-y is sure to have man\-y difficul\-ties.
    \textsuperscript{14}\thinspace{}There\-fore, the sage treats things as dif\-fi\-cult,
    \textsuperscript{15}\thinspace{}and thus, in the end, noth\-ing is dif\-fi\-cult.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{64:1}\thinspace{}It's eas\-y to hold what is at rest.
    \textsuperscript{2}\thinspace{}It's eas\-y to plan for what has\-n't be\-gun.
    \textsuperscript{3}\thinspace{}It's eas\-y to break what is brit\-tle.
    \textsuperscript{4}\thinspace{}It's eas\-y to scat\-ter what is ti\-ny.
    \textsuperscript{5}\thinspace{}Deal with it be\-fore it happens;
    \textsuperscript{6}\thinspace{}man\-age it be\-fore it's cha\-ot\-ic.
\end{para}

\begin{para}
    \textsuperscript{64:7}\thinspace{}A tree that's thick e\-nough to em\-brace rises from a ti\-ny shoot.
    \textsuperscript{8}\thinspace{}A ter\-race that's nine stories tall rises from a heap of earth.
    \textsuperscript{9}\thinspace{}A wall that's three hun\-dred yards high begins at the ground be\-neath your feet.
\end{para}

\begin{para}
    \textsuperscript{64:10}\thinspace{}Those who rule it ru\-in it.
    \textsuperscript{11}\thinspace{}Those who hold it lose it.
    \textsuperscript{12}\thinspace{}There\-fore, the sage rules noth\-ing and thus ruins noth\-ing.
    \textsuperscript{13}\thinspace{}He holds on to noth\-ing and thus loses noth\-ing.
\end{para}

\begin{para}
    \textsuperscript{64:14}\thinspace{}Peo\-ple u\-su\-al\-ly fail in their endeavors on the verge of suc\-cess.
    \textsuperscript{15}\thinspace{}There\-fore we say, “If you're as care\-ful at the end as you are at the be\-gin\-ning,
    \textsuperscript{16}\thinspace{}you'll have no fail\-ure.”
\end{para}

\begin{para}
    \textsuperscript{64:17}\thinspace{}The sage desires not to de\-sire,
    \textsuperscript{18}\thinspace{}and does\-n't val\-ue goods that are hard to ob\-tain;
    \textsuperscript{19}\thinspace{}he learns not to stud\-y,
    \textsuperscript{20}\thinspace{}and returns to what the mul\-ti\-tude has passed o\-ver.
    \textsuperscript{21}\thinspace{}He could help all things to be nat\-u\-ral,
    \textsuperscript{22}\thinspace{}but dares not do it.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{66:1}\thinspace{}The rivers and lakes are a\-ble to be the kings of the valleys
    \textsuperscript{2}\thinspace{}be\-cause of their skill at be\-ing low.
    \textsuperscript{4}\thinspace{}There\-fore, want\-ing to be a\-bove the peo\-ple, the sage must speak to them hum\-bly from be\-low;
    \textsuperscript{5}\thinspace{}want\-ing to be be\-fore the peo\-ple, he must put him\-self be\-hind them.
    \textsuperscript{6}\thinspace{}There\-fore, he dwells a\-bove them yet the peo\-ple don't feel that he's a bur\-den.
    \textsuperscript{7}\thinspace{}He dwells be\-fore them yet the peo\-ple don't feel that he's an ob\-struc\-tion.
    \textsuperscript{8}\thinspace{}Eve\-ry\-one in the world delights in pro\-mot\-ing him and nev\-er tires of him.
\end{para}

\begin{para}
    \textsuperscript{66:9}\thinspace{}It's be\-cause he does\-n't con\-tend
    \textsuperscript{10}\thinspace{}that no one in the world can con\-tend with him.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{67:4}\thinspace{}I con\-stant\-ly have three treasures which I hold and treas\-ure:
    \textsuperscript{5}\thinspace{}the first is com\-pas\-sion,
    \textsuperscript{6}\thinspace{}the sec\-ond is fru\-gal\-i\-ty,
    \textsuperscript{7}\thinspace{}the third is not dar\-ing to be pre\-em\-i\-nent in the world.
    \textsuperscript{8}\thinspace{}It's be\-cause of com\-pas\-sion that I can be cou\-ra\-geous;
    \textsuperscript{9}\thinspace{}it's be\-cause of fru\-gal\-i\-ty that I can be gen\-er\-ous;
    \textsuperscript{10}\thinspace{}it's be\-cause of not dar\-ing to be pre\-em\-i\-nent in the world that I can be head of all things.
\end{para}

\begin{para}
    \textsuperscript{67:11}\thinspace{}Now, abandoning com\-pas\-sion to be cou\-ra\-geous,
    \textsuperscript{12}\thinspace{}abandoning fru\-gal\-i\-ty to be gen\-er\-ous,
    \textsuperscript{13}\thinspace{}abandoning the rear to be in front:
    \textsuperscript{14}\thinspace{}this is courting death.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{68:1}\thinspace{}A skill\-ful war\-ri\-or does\-n't show his might.
    \textsuperscript{2}\thinspace{}A skill\-ful fight\-er does\-n't get an\-gry.
    \textsuperscript{3}\thinspace{}A skill\-ful vic\-tor does\-n't en\-gage his en\-e\-my.
    \textsuperscript{4}\thinspace{}A skill\-ful em\-ploy\-er places him\-self be\-low others.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{73:1}\thinspace{}Those who are brave in dar\-ing will die;
    \textsuperscript{2}\thinspace{}those who are brave in cau\-tion will live.
    \textsuperscript{3}\thinspace{}In these two cases, one is ben\-e\-fi\-cial and the oth\-er is harm\-ful.
    \textsuperscript{4}\thinspace{}Who understands why Heav\-en hates what it does?
\end{para}

\begin{para}
    \textsuperscript{73:6}\thinspace{}The Way of Heav\-en does\-n't fight yet wins skill\-ful\-ly,
    \textsuperscript{7}\thinspace{}does\-n't speak yet responds skill\-ful\-ly,
    \textsuperscript{8}\thinspace{}is\-n't summoned yet co\-mes on its own,
    \textsuperscript{9}\thinspace{}is at ease yet plans skill\-ful\-ly.
\end{para}

\begin{para}
    \textsuperscript{73:10}\thinspace{}The net of Heav\-en is vast; it has a coarse mesh yet noth\-ing slips through.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{75:1}\thinspace{}The peo\-ple starve be\-cause those a\-bove con\-sume too much in taxes.
    \textsuperscript{2}\thinspace{}The peo\-ple reb\-el be\-cause those a\-bove in\-ter\-fere.
    \textsuperscript{3}\thinspace{}The peo\-ple scoff at death be\-cause those a\-bove de\-mand so much from life.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{76:1}\thinspace{}While liv\-ing, a man is sup\-ple and soft,
    \textsuperscript{2}\thinspace{}but once dead, he is hard and rig\-id.
    \textsuperscript{3}\thinspace{}While liv\-ing, all things, grasses, and trees are sup\-ple and crisp,
    \textsuperscript{4}\thinspace{}but once dead, they are withered and dry.
\end{para}

\begin{para}
    \textsuperscript{76:5}\thinspace{}There\-fore it is said: those who are hard and rig\-id are the followers of death;
    \textsuperscript{6}\thinspace{}those who are sup\-ple and soft are the followers of life.
    \textsuperscript{7}\thinspace{}There\-fore, a rig\-id ar\-my won't win,
    \textsuperscript{8}\thinspace{}and a rig\-id tree will snap.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{77:1}\thinspace{}The Way of Heav\-en is like an arch\-er draw\-ing his bow.
    \textsuperscript{2}\thinspace{}If it's too high, he presses down; if it's too low, he lifts up;
    \textsuperscript{3}\thinspace{}if there is ex\-cess, he reduces it;
    \textsuperscript{4}\thinspace{}if there is not e\-nough, he adds to it.
    \textsuperscript{5}\thinspace{}The Way of Heav\-en is al\-so to re\-duce what has ex\-cess and add to what does\-n't have e\-nough.
\end{para}

\begin{para}
    \textsuperscript{77:6}\thinspace{}The Way of man is dif\-fer\-ent:
    \textsuperscript{7}\thinspace{}he reduces what does\-n't have e\-nough and offers it to that which has ex\-cess.
\end{para}

\begin{para}
    \textsuperscript{77:8}\thinspace{}Who can have ex\-cess and of\-fer it to Heav\-en?
    \textsuperscript{9}\thinspace{}On\-ly he who has the Way.
\end{para}

\begin{para}
    \textsuperscript{77:10}\thinspace{}There\-fore, the sage takes actions with\-out claiming the results;
    \textsuperscript{11}\thinspace{}he succeeds with\-out dwell\-ing on it;
    \textsuperscript{12}\thinspace{}sim\-i\-lar\-ly, he does\-n't de\-sire to make a dis\-play of his wor\-thi\-ness.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{79:1}\thinspace{}When har\-mo\-ny is made af\-ter great re\-sent\-ment,
    \textsuperscript{2}\thinspace{}there is al\-ways some re\-sent\-ment remaining.
    \textsuperscript{3}\thinspace{}How can this be con\-sid\-ered good?
    \textsuperscript{4}\thinspace{}There\-fore, the sage keeps his part of the con\-tract, but makes no demands on the oth\-er par\-ty.
\end{para}

\begin{para}
    \textsuperscript{79:7}\thinspace{}The Way of Heav\-en has no favorites,
    \textsuperscript{8}\thinspace{}it's con\-stant\-ly on the side of the good man.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{80:1}\thinspace{}Let there be a small state with few peo\-ple.
    \textsuperscript{2}\thinspace{}Let them have pow\-er\-ful tools and not use them.
    \textsuperscript{3}\thinspace{}Let the peo\-ple take death se\-ri\-ous\-ly and not mi\-grate far a\-way.
    \textsuperscript{4}\thinspace{}There are boats and vehicles but no rea\-son to ride them.
    \textsuperscript{5}\thinspace{}There are ar\-mor and weapons but no rea\-son to pa\-rade them.
    \textsuperscript{6}\thinspace{}Let the peo\-ple keep their records with knot\-ted string a\-gain.
    \textsuperscript{7}\thinspace{}Their food is sweet,
    \textsuperscript{8}\thinspace{}their clothes are beau\-ti\-ful,
    \textsuperscript{9}\thinspace{}their homes are se\-cure,
    \textsuperscript{10}\thinspace{}and their cus\-toms are en\-joy\-a\-ble.
\end{para}

\begin{para}
    \textsuperscript{80:11}\thinspace{}Neighboring states can see each oth\-er,
    \textsuperscript{12}\thinspace{}they can hear each other's dogs and roosters,
    \textsuperscript{13}\thinspace{}yet the peo\-ple grow old and die
    \textsuperscript{14}\thinspace{}with\-out any deal\-ings be\-tween each oth\-er.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{81:1}\thinspace{}Cred\-i\-ble words aren't el\-o\-quent;
    \textsuperscript{2}\thinspace{}el\-o\-quent words aren't cred\-i\-ble.
    \textsuperscript{3}\thinspace{}The skill\-ful aren't polymaths;
    \textsuperscript{4}\thinspace{}those who are polymaths aren't skill\-ful.
    \textsuperscript{5}\thinspace{}Those who know aren't learn\-ed;
    \textsuperscript{6}\thinspace{}the learn\-ed don't know.
\end{para}

\begin{para}
    \textsuperscript{81:7}\thinspace{}The sage accumulates noth\-ing.
    \textsuperscript{8}\thinspace{}The more he does for others, the greater his possessions.
    \textsuperscript{9}\thinspace{}The more he gives to others, the greater his a\-bun\-dance.
\end{para}

\begin{para}
    \textsuperscript{81:10}\thinspace{}There\-fore, the Way of Heav\-en is to ben\-e\-fit with\-out harming.
    \textsuperscript{11}\thinspace{}The Way of the sage is to act with\-out contending.
\end{para}

\begin{para}
\end{para}


\end{document}

% Copyright 2021 Michael Hendricks
%
% Permission is granted to use, copy, modify, and distribute
% this work for any purpose with or without fee.
%
% This work is provided as-is without any express or implied
% warranty. In no event is the author liable for any damages
% arising from this work.
