%
% see tex.stackexchange.com/a/600/151276
\def\changemargin#1#2{\list{}{\rightmargin#2\leftmargin#1}\item[]}
\let\endchangemargin=\endlist

\documentclass[openany,12pt,english]{book}
\usepackage[utf8]{inputenc}
\usepackage{babel}
\usepackage{microtype}
\usepackage{csquotes}

\usepackage[
  paperwidth=8.5in,
  paperheight=11in,
  top=1in,
  outer=1in,
  bottom=1.25in,
  inner=1.5in
]{geometry}

\title{Digests}
\author{}
\date{}

\newenvironment{para}{\par\pretolerance=100\tolerance=200\setlength{\emergencystretch}{0.6em}\relax}{\par}

\begin{document}

\maketitle

% Copyright page
{
\thispagestyle{empty}
\vspace*{\fill}
\begin{changemargin}{.8in}{.8in}
\begin{center}
\textcopyright{}~2021 Michael Hendricks

\vspace{0.5\baselineskip}

Permission is granted to use, copy, modify, and distribute
this work for any purpose with or without fee.

\vspace{0.5\baselineskip}

This work is provided as-is without any express or implied
warranty. In no event is the author liable for any damages
arising from this work.
\end{center}
\end{changemargin}
}
\pagenumbering{arabic}
\clearpage{}

% Table of Contents
\tableofcontents
\clearpage{}

\markboth{}{}
\twocolumn[\begin{changemargin}{.6in}{.6in}
\end{changemargin}]{}


\vspace{0pt plus 8\baselineskip{}}\section*{Priorities}\addcontentsline{toc}{section}{Priorities}
\begin{para}
    \textsuperscript{0}\thinspace{}You have rea\-son and con\-science to gov\-ern your choices. The fol\-low\-ing priorities are a guide to keep con\-science with\-in prop\-er bounds. They are list\-ed by de\-creas\-ing im\-por\-tance. Try to ob\-serve all the priorities. If two must con\-flict, pre\-fer the more im\-por\-tant one:
\end{para}

\begin{para}
    \textsuperscript{1}\thinspace{}Com\-pul\-sion may on\-ly be used to stop an im\-me\-di\-ate threat to life, prop\-er\-ty, or free will.
    \textsuperscript{2}\thinspace{}Choose what you will have at the ex\-pense of what you won't.
    \textsuperscript{3}\thinspace{}Give no im\-por\-tance to that which has none.
    \textsuperscript{4}\thinspace{}Tend to your own af\-fairs.
\end{para}

\begin{para}
    \textsuperscript{5}\thinspace{}Make promises spar\-ing\-ly.
    \textsuperscript{6}\thinspace{}Keep your word.
    \textsuperscript{7}\thinspace{}Make a\-mends.
    \textsuperscript{8}\thinspace{}Be grate\-ful.
\end{para}

\begin{para}
    \textsuperscript{9}\thinspace{}Take care of your\-self.
    \textsuperscript{10}\thinspace{}Be spar\-ing and cau\-tious with behaviors that might be ad\-dic\-tive.
\end{para}

\begin{para}
    \textsuperscript{11}\thinspace{}Re\-spect others.
    \textsuperscript{12}\thinspace{}Speak kind\-ly and qui\-et\-ly.
    \textsuperscript{13}\thinspace{}Main\-tain a few strong relationships.
    \textsuperscript{14}\thinspace{}Main\-tain sta\-ble, long-term relationships with your spouses.
    \textsuperscript{15}\thinspace{}Ob\-tain con\-sent from your spouses for all in\-ti\-mate re\-la\-tions.
    \textsuperscript{16}\thinspace{}Have chil\-dren with your spouses.
\end{para}

\begin{para}
    \textsuperscript{17}\thinspace{}Spend some time each day qui\-et and a\-lone.
    \textsuperscript{18}\thinspace{}Pray dai\-ly.
    \textsuperscript{19}\thinspace{}Stud\-y the Digests and Messages dai\-ly, bal\-anc\-ing lit\-er\-al and fig\-ur\-a\-tive in\-ter\-pre\-ta\-tion.
\end{para}

\begin{para}
    \textsuperscript{20}\thinspace{}A\-void con\-ten\-tion.
    \textsuperscript{21}\thinspace{}For\-give wrongs, re\-al and perceived.
    \textsuperscript{22}\thinspace{}Main\-tain po\-lit\-i\-cal neu\-tral\-i\-ty.
\end{para}

\begin{para}
    \textsuperscript{23}\thinspace{}Ac\-quire use\-ful knowl\-edge, then act on it.
    \textsuperscript{24}\thinspace{}En\-gage skill\-ful\-ly in an oc\-cu\-pa\-tion, trade, or craft.
    \textsuperscript{25}\thinspace{}Make an hon\-est prof\-it.
    \textsuperscript{26}\thinspace{}In\-cur debts rare\-ly and spar\-ing\-ly, and pay them back quick\-ly.
    \textsuperscript{27}\thinspace{}Ten per\-cent of your an\-nu\-al expenditures should be donations.
    \textsuperscript{28}\thinspace{}Keep a mod\-est re\-serve of use\-ful as\-sets.
\end{para}

\begin{para}
    \textsuperscript{29}\thinspace{}Smile.
    \textsuperscript{30}\thinspace{}Serve a\-non\-y\-mous\-ly, with\-out re\-mu\-ner\-a\-tion.
\end{para}

\begin{para}
    \textsuperscript{31}\thinspace{}Main\-tain your bod\-y in strength, speed, flex\-i\-bil\-i\-ty, and bal\-ance.
    \textsuperscript{32}\thinspace{}De\-vote one day in sev\-en to rejuvenation and spir\-it\-u\-al growth.
    \textsuperscript{33}\thinspace{}Im\-prove your sur\-round\-ings.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Virtues}\addcontentsline{toc}{section}{Virtues}
\begin{para}
    \textsuperscript{0}\thinspace{}Eve\-ry vir\-tue tak\-en to ex\-treme becomes vi\-ce. The vir\-tues be\-low are list\-ed in pairs. Bal\-ance the two vir\-tues in a pair. If you're un\-sure how to bal\-ance them, err on the side of the first. The pairs are list\-ed al\-pha\-bet\-i\-cal\-ly with no rel\-a\-tive im\-por\-tance in\-tend\-ed be\-tween them:
\end{para}

\begin{para}
    \textsuperscript{1}\thinspace{}Chal\-lenge, com\-fort.
    \textsuperscript{2}\thinspace{}Com\-pan\-ion\-ship, sol\-i\-tude.
    \textsuperscript{3}\thinspace{}Con\-tent\-ment, am\-bi\-tion.
    \textsuperscript{4}\thinspace{}De\-tach\-ment, at\-tach\-ment.
    \textsuperscript{5}\thinspace{}Flex\-i\-ble, te\-na\-cious.
    \textsuperscript{6}\thinspace{}Gen\-er\-ous, fru\-gal.
    \textsuperscript{7}\thinspace{}Grad\-u\-al, swift.
    \textsuperscript{8}\thinspace{}Ho\-lis\-tic, at\-om\-is\-tic.
    \textsuperscript{9}\thinspace{}Hum\-ble, con\-fi\-dent.
    \textsuperscript{10}\thinspace{}In\-de\-pend\-ence, con\-form\-i\-ty.
    \textsuperscript{11}\thinspace{}In\-tel\-lect, e\-mo\-tion.
    \textsuperscript{12}\thinspace{}Mer\-cy, jus\-tice.
    \textsuperscript{13}\thinspace{}O\-pen, guard\-ed.
    \textsuperscript{14}\thinspace{}Pre\-pare, im\-pro\-vise.
    \textsuperscript{15}\thinspace{}Si\-lence, com\-mu\-ni\-ca\-tion.
    \textsuperscript{16}\thinspace{}Sim\-ple, a\-bun\-dant.
    \textsuperscript{17}\thinspace{}So\-lem\-ni\-ty, hu\-mor.
    \textsuperscript{18}\thinspace{}Sta\-ble, dy\-nam\-ic.
    \textsuperscript{19}\thinspace{}Tact\-ful, hon\-est.
    \textsuperscript{20}\thinspace{}Tra\-di\-tion, in\-no\-va\-tion.
    \textsuperscript{21}\thinspace{}Tran\-quil\-i\-ty, ex\-cite\-ment.
    \textsuperscript{22}\thinspace{}Work, re\-pose.
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Prayer}\addcontentsline{toc}{section}{Prayer}
\begin{para}
    \textsuperscript{1:1}\thinspace{}Of\-fer in\-for\-mal pray\-er at least three times dur\-ing your wak\-ing hours.
\end{para}

\begin{para}
    \textsuperscript{1:2}\thinspace{}Po\-si\-tion your\-self how\-ev\-er is mean\-ing\-ful, com\-fort\-a\-ble, and ap\-pro\-pri\-ate to your circumstances.
    \textsuperscript{3}\thinspace{}Pray si\-lent\-ly or a\-loud.
    \textsuperscript{4}\thinspace{}Be\-gin by say\-ing, “God in heav\-en.”
    \textsuperscript{5}\thinspace{}In your own words, ex\-press grat\-i\-tude, ac\-knowl\-edge your faults, and re\-quest guid\-ance and as\-sis\-tance.
    \textsuperscript{6}\thinspace{}Pause to lis\-ten.
\end{para}

\begin{para}
    \textsuperscript{1:7}\thinspace{}Con\-clude by say\-ing, “A\-men.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{2:1}\thinspace{}Of\-fer for\-mal pray\-er once or twice dur\-ing your wak\-ing hours,
    \textsuperscript{2}\thinspace{}with at least one performed out\-side.
\end{para}

\begin{para}
    \textsuperscript{2:3}\thinspace{}When praying out\-side, con\-sid\-er turn\-ing your back to\-wards the ap\-prox\-i\-mate lo\-ca\-tion of the sun.
    \textsuperscript{4}\thinspace{}Pray a\-loud if pos\-si\-ble.
    \textsuperscript{5}\thinspace{}Make your best ef\-fort to per\-form each pos\-ture of pray\-er in turn:
\end{para}

\begin{para}
    \textsuperscript{2:6}\thinspace{}Grat\-i\-tude: Stand, face up\-ward, eyes closed, hands clasped, un\-til tran\-quil.
    \textsuperscript{7}\thinspace{}Re\-cite, “God in heav\-en, I thank you for life, air to breathe, qui\-et places, and the a\-bil\-i\-ty to choose;
    \textsuperscript{8}\thinspace{}for challenges, work, e\-mo\-tion, and the gift of rea\-son.”
    \textsuperscript{9}\thinspace{}In your own words, ex\-press heart\-felt grat\-i\-tude to God.
\end{para}

\begin{para}
    \textsuperscript{2:10}\thinspace{}Con\-fes\-sion: Bow, face down\-ward, eyes closed, hands on knees.
    \textsuperscript{11}\thinspace{}Re\-cite, “God in heav\-en, I strive to keep the Priorities, to bal\-ance the Vir\-tues, and to learn from my failures when I don't.
    \textsuperscript{12}\thinspace{}I want to be more faith\-ful.
    \textsuperscript{13}\thinspace{}I fall short in these ways.”
    \textsuperscript{14}\thinspace{}In your own words, con\-fess to God your failures, weaknesses, and ways you need to im\-prove.
\end{para}

\begin{para}
    \textsuperscript{2:15}\thinspace{}Pe\-ti\-tion: Squat, face for\-ward, eyes closed, arms out\-stretched, hands cupped to form a bowl.
    \textsuperscript{16}\thinspace{}Re\-cite, “God in heav\-en, you have giv\-en me a beau\-ti\-ful world to live in, messengers to teach me, and guid\-ance when I ask.
    \textsuperscript{17}\thinspace{}Hear my pray\-er.”
    \textsuperscript{18}\thinspace{}In your own words, re\-quest God's guid\-ance.
\end{para}

\begin{para}
    \textsuperscript{2:19}\thinspace{}Re\-cep\-tion: Stand, face down\-ward, eyes closed, hands clasped.
    \textsuperscript{20}\thinspace{}Re\-cite, “God in heav\-en, you let the sun rise on good and evil peo\-ple a\-like.
    \textsuperscript{21}\thinspace{}You send guid\-ance through a gen\-tle thought, a soft feel\-ing, or a qui\-et voice.
    \textsuperscript{22}\thinspace{}I'm listening.” Lis\-ten for God's guid\-ance.
\end{para}

\begin{para}
    \textsuperscript{2:23}\thinspace{}Ac\-tion: Bow, face down\-ward, eyes closed, hands on knees.
    \textsuperscript{24}\thinspace{}Re\-cite, “God in heav\-en, the great\-est gift brings no prof\-it when left un\-used.
    \textsuperscript{25}\thinspace{}I thank you for giv\-ing me guid\-ance.
    \textsuperscript{26}\thinspace{}I plan to act.”
    \textsuperscript{27}\thinspace{}In your own words, tell God how you will act on the guid\-ance you have re\-ceived.
\end{para}

\begin{para}
    \textsuperscript{2:28}\thinspace{}Con\-clude by say\-ing, “A\-men.”
\end{para}

\vspace{0pt plus 8\baselineskip{}}\section*{Parables}\addcontentsline{toc}{section}{Parables}
\begin{para}
    \textsuperscript{1:1}\thinspace{}A young child asks his moth\-er, “How is a ba\-by made?”
    \textsuperscript{2}\thinspace{}Know\-ing the child's ru\-di\-men\-ta\-ry un\-der\-stand\-ing, she responds,
    \textsuperscript{3}\thinspace{}“The fa\-ther places a seed in the mother's bel\-ly, and it grows in\-to a ba\-by.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{2:1}\thinspace{}A king had two vaults filled with treas\-ure, each protected by its own lock.
    \textsuperscript{2}\thinspace{}The king said that who\-ev\-er opened a vault could have the treas\-ure in\-side.
    \textsuperscript{3}\thinspace{}Two men sought the treas\-ure.
\end{para}

\begin{para}
    \textsuperscript{2:4}\thinspace{}The first man turned the dials of the lock.
    \textsuperscript{5}\thinspace{}He listened care\-ful\-ly as the mech\-a\-nism moved in\-side.
    \textsuperscript{6}\thinspace{}He no\-ticed patterns of sound and vi\-bra\-tion with\-in the dials.
    \textsuperscript{7}\thinspace{}Af\-ter man\-y days of per\-sis\-tence and stud\-y, he discovered the com\-bi\-na\-tion and unlocked the vault.
\end{para}

\begin{para}
    \textsuperscript{2:8}\thinspace{}The second man offered to be the king's serv\-ant.
    \textsuperscript{9}\thinspace{}He listened care\-ful\-ly to the king's wishes, and ex\-e\-cut\-ed them faith\-ful\-ly.
    \textsuperscript{10}\thinspace{}He no\-ticed patterns a\-mong the king's requests and an\-tic\-i\-pat\-ed his wishes.
    \textsuperscript{11}\thinspace{}Af\-ter man\-y days of per\-sis\-tence and ser\-vice, the second man asked the king for the com\-bi\-na\-tion and re\-ceived it.
\end{para}

\begin{para}
    \textsuperscript{2:13}\thinspace{}Both men re\-ceived treas\-ure.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{3:1}\thinspace{}There were two lakes.
    \textsuperscript{2}\thinspace{}The first lake said to him\-self, “I will re\-tain eve\-ry drop of wa\-ter that flows in\-to me so that I can be great.”
    \textsuperscript{3}\thinspace{}His waters be\-came salt\-y and pu\-trid.
    \textsuperscript{4}\thinspace{}His fish died and the plants a\-long his shore wilted.
    \textsuperscript{5}\thinspace{}Animals and travelers avoided him.
\end{para}

\begin{para}
    \textsuperscript{3:6}\thinspace{}The second lake said to him\-self, “I want to pros\-per all life on the moun\-tain be\-low me. I will re\-tain a mod\-est pool for my\-self then al\-low each drop to pass down the riv\-er to those who might use it.”
    \textsuperscript{7}\thinspace{}His waters be\-came clean and sweet.
    \textsuperscript{8}\thinspace{}His fish grew large and the plants a\-long his shore prospered.
    \textsuperscript{9}\thinspace{}Animals and travelers came from a\-far to en\-joy his splen\-dor.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{4:1}\thinspace{}Rain, fresh\-ly fall\-en from a cloud, is clean and good to drink.
    \textsuperscript{2}\thinspace{}As it flows down mountains and a\-cross plains, it becomes soiled and bit\-ter.
    \textsuperscript{3}\thinspace{}The sun must ex\-tract pure wa\-ter from the sea a\-gain, leav\-ing impurities be\-hind, to pro\-duce fresh rain once more.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{5:1}\thinspace{}A horse, a dol\-phin, and an ea\-gle each needs dif\-fer\-ent nour\-ish\-ment,
    \textsuperscript{2}\thinspace{}and yet each grows to full stat\-ure and maj\-es\-ty.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{6:1}\thinspace{}A child asks his moth\-er, “Please get me a cup from the top shelf.”
    \textsuperscript{2}\thinspace{}The moth\-er reaches the cup for him.
    \textsuperscript{3}\thinspace{}A grown man asks his moth\-er, “Please get me a cup from the top shelf.”
    \textsuperscript{4}\thinspace{}She responds, “You can reach it your\-self.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{7:1}\thinspace{}A man mixed one meas\-ure of sug\-ar and the juice from one lem\-on in\-to 2,000 meas\-ures of wa\-ter,
    \textsuperscript{2}\thinspace{}then served it to his guests as lem\-on\-ade.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{8:1}\thinspace{}A man brought eggs home from the mar\-ket.
    \textsuperscript{2}\thinspace{}Up\-on ar\-riv\-ing, he an\-nounced, “I purchased one doz\-en eggs.”
    \textsuperscript{3}\thinspace{}His wife said, “No you did\-n't. You bought twelve eggs.”
    \textsuperscript{4}\thinspace{}A son said, “You're both wrong. There are clear\-ly three rows of four eggs each.”
    \textsuperscript{5}\thinspace{}A daugh\-ter dis\-a\-greed, “Not at all. There are four columns of three eggs each.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{9:1}\thinspace{}A man de\-cid\-ed to climb a moun\-tain.
    \textsuperscript{2}\thinspace{}He fixed his gaze per\-ma\-nent\-ly on the sum\-mit and be\-gan hik\-ing.
    \textsuperscript{3}\thinspace{}He tripped o\-ver un\-seen rocks at his feet,
    \textsuperscript{4}\thinspace{}and was un\-a\-ble to reach the moun\-tain top.
\end{para}

\begin{para}
    \textsuperscript{9:5}\thinspace{}An\-oth\-er man de\-cid\-ed to climb the same moun\-tain.
    \textsuperscript{6}\thinspace{}He gazed per\-ma\-nent\-ly at his feet, to a\-void the errors of the first man, and be\-gan hik\-ing.
    \textsuperscript{7}\thinspace{}With\-out trip\-ping once, he wandered off in\-to a val\-ley,
    \textsuperscript{8}\thinspace{}and was un\-a\-ble to reach the moun\-tain top.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{10:1}\thinspace{}A man no\-ticed jig\-saw puz\-zle pieces rest\-ing on a ta\-ble.
    \textsuperscript{2}\thinspace{}He wanted to as\-sem\-ble the puz\-zle but pieces were mis\-sing, and he did\-n't know how the puz\-zle should look when fin\-ished.
\end{para}

\begin{para}
    \textsuperscript{10:3}\thinspace{}From time to time he lo\-cat\-ed mis\-sing pieces, but there were still too few to guess how the puz\-zle should look.
    \textsuperscript{4}\thinspace{}One day he found a sketch of the fin\-ished puz\-zle.
    \textsuperscript{5}\thinspace{}He as\-sem\-bled the pieces he had, on\-ly to re\-al\-ize that he was still mis\-sing some.
    \textsuperscript{6}\thinspace{}He con\-tin\-ued search\-ing for mis\-sing pieces, plac\-ing them where they belonged.
    \textsuperscript{7}\thinspace{}In time, he found all the pieces, and as\-sem\-bled the puz\-zle in\-to its beau\-ti\-ful, com\-plete form.
\end{para}


\end{document}

% Copyright 2021 Michael Hendricks
%
% Permission is granted to use, copy, modify, and distribute
% this work for any purpose with or without fee.
%
% This work is provided as-is without any express or implied
% warranty. In no event is the author liable for any damages
% arising from this work.
