%
% see tex.stackexchange.com/a/600/151276
\def\changemargin#1#2{\list{}{\rightmargin#2\leftmargin#1}\item[]}
\let\endchangemargin=\endlist

\documentclass[openany,12pt,english]{book}
\usepackage[utf8]{inputenc}
\usepackage{babel}
\usepackage{microtype}
\usepackage{csquotes}

\usepackage[
  paperwidth=8.5in,
  paperheight=11in,
  top=1in,
  outer=1in,
  bottom=1.25in,
  inner=1.5in
]{geometry}

\title{David}
\author{}
\date{}

\newenvironment{para}{\par\pretolerance=100\tolerance=200\setlength{\emergencystretch}{0.6em}\relax}{\par}

\begin{document}

\maketitle

% Copyright page
{
\thispagestyle{empty}
\vspace*{\fill}
\begin{changemargin}{.8in}{.8in}
\begin{center}
\textcopyright{}~2021 Michael Hendricks

\vspace{0.5\baselineskip}

Permission is granted to use, copy, modify, and distribute
this work for any purpose with or without fee.

\vspace{0.5\baselineskip}

This work is provided as-is without any express or implied
warranty. In no event is the author liable for any damages
arising from this work.
\end{center}
\end{changemargin}
}
\pagenumbering{arabic}
\clearpage{}

% Table of Contents
\clearpage{}

\markboth{}{}
\twocolumn[\begin{changemargin}{.6in}{.6in}
  \begin{center}
    \itshape
    A se\-lec\-tion of proverbs tak\-en from the Book of Psalms in the Hebrew Bible.
    Verse references such
    as~\textsuperscript{12:8}\thinspace{}indicate
    the original source of the text, but
    don't imply a full quotation.
  \end{center}
\end{changemargin}]{}


\begin{para}
    \textsuperscript{2:11}\thinspace{}Serve \textsc{God} with fear and re\-joice with trem\-bling.
\end{para}

\begin{para}
    \textsuperscript{4:4}\thinspace{}In your an\-ger, do not sin. Pon\-der in your heart on your bed and be still.
    \textsuperscript{5}\thinspace{}Of\-fer the sacrifices of right\-eous\-ness and put your trust in \textsc{God}.
\end{para}

\begin{para}
    \textsuperscript{12:8}\thinspace{}The wick\-ed walk on eve\-ry side when vile\-ness is ex\-alt\-ed a\-mong the chil\-dren of men.
\end{para}

\begin{para}
    \textsuperscript{15:1}\thinspace{}\textsc{God}, who may so\-journ in your tent? Who may dwell on your ho\-ly hill?
    \textsuperscript{2}\thinspace{}He who walks with in\-teg\-ri\-ty and does what is right and speaks truth in his heart;
    \textsuperscript{3}\thinspace{}who does not slan\-der with his tongue and does no evil to his neigh\-bor and does not take up a re\-proach a\-gainst his friend;
    \textsuperscript{4}\thinspace{}who despises a vile per\-son but hon\-ors those who fear \textsc{God}; who keeps a prom\-ise e\-ven when it hurts and does not change;
    \textsuperscript{5}\thinspace{}who does not loan out his mon\-ey with in\-ter\-est and does not take a bribe a\-gainst the in\-no\-cent. He who does these things shall nev\-er be moved.
\end{para}

\begin{para}
    \textsuperscript{18:25}\thinspace{}\textsc{God} shows mer\-cy to the mer\-ci\-ful.
\end{para}

\begin{para}
    \textsuperscript{18:30}\thinspace{}The word of \textsc{God} proves true. He is a shield to all who take ref\-uge in him.
\end{para}

\begin{para}
    \textsuperscript{19:1}\thinspace{}The heavens de\-clare the glo\-ry of God and the skies show his hand\-i\-work.
\end{para}

\begin{para}
    \textsuperscript{19:7}\thinspace{}The law of \textsc{God} is per\-fect, re\-viv\-ing the soul. The tes\-ti\-mo\-ny of \textsc{God} is trust\-wor\-thy, mak\-ing wise the sim\-ple.
    \textsuperscript{8}\thinspace{}The precepts of \textsc{God} are right, re\-joic\-ing the heart. The com\-mand\-ment of \textsc{God} is pure, enlightening the eyes.
    \textsuperscript{9}\thinspace{}The fear of \textsc{God} is clean, en\-dur\-ing for\-ev\-er. The laws of \textsc{God} are true and right\-eous al\-to\-geth\-er.
\end{para}

\begin{para}
    \textsuperscript{27:14}\thinspace{}Wait for \textsc{God}. Be strong and let your heart take cour\-age. Wait for \textsc{God}.
\end{para}

\begin{para}
    \textsuperscript{30:5}\thinspace{}Weep\-ing may last for a night, but joy co\-mes in the morn\-ing.
\end{para}

\begin{para}
    \textsuperscript{32:10}\thinspace{}Man\-y sorrows come to the wick\-ed, but lov\-ing-kind\-ness surrounds him who trusts in \textsc{God}.
\end{para}

\begin{para}
    \textsuperscript{33:6}\thinspace{}By the word of \textsc{God} the heavens were made and all their host by the breath of his mouth.
\end{para}

\begin{para}
    \textsuperscript{34:13}\thinspace{}Keep your tongue from evil and your lips from speak\-ing de\-ceit.
    \textsuperscript{14}\thinspace{}De\-part from evil and do good. Seek peace and pur\-sue it.
\end{para}

\begin{para}
    \textsuperscript{37:1}\thinspace{}Don't wor\-ry a\-bout evildoers, and don't en\-vy wrongdoers.
    \textsuperscript{7}\thinspace{}Be still be\-fore \textsc{God} and wait pa\-tient\-ly for him. Don't wor\-ry a\-bout him who prospers or a\-bout him who carries out wick\-ed schemes.
    \textsuperscript{8}\thinspace{}Cease from an\-ger and for\-sake wrath. Don't wor\-ry: it on\-ly leads to evil\-do\-ing.
\end{para}

\begin{para}
    \textsuperscript{37:16}\thinspace{}A lit\-tle that the right\-eous has is bet\-ter than the a\-bun\-dance of man\-y wick\-ed.
\end{para}

\begin{para}
    \textsuperscript{37:21}\thinspace{}The wick\-ed borrows and does not pay back, but the right\-eous is gen\-er\-ous and gives.
\end{para}

\begin{para}
    \textsuperscript{39:1}\thinspace{}I will guard my ways so that I may not sin with my tongue. I will muz\-zle my mouth while the wick\-ed are in my pres\-ence.
\end{para}

\begin{para}
    \textsuperscript{41:1}\thinspace{}Bless\-ed is he who considers the poor. \textsc{God} will de\-liv\-er him in the day of trou\-ble.
\end{para}

\begin{para}
    \textsuperscript{46:1}\thinspace{}God is our ref\-uge and strength, a ver\-y pres\-ent help in trou\-ble.
    \textsuperscript{10}\thinspace{}Be still and know that I am God.
\end{para}

\begin{para}
    \textsuperscript{49:16}\thinspace{}Do not be a\-fraid when one is made rich, when the splen\-dor of his house increases.
    \textsuperscript{17}\thinspace{}For when he dies he will car\-ry noth\-ing a\-way. His glo\-ry will not de\-scend af\-ter him.
\end{para}

\begin{para}
    \textsuperscript{50:14}\thinspace{}Of\-fer to God a sac\-ri\-fice of thanks\-giv\-ing, and pay your vows to the Most High,
    \textsuperscript{15}\thinspace{}and call on me in the day of trou\-ble. I will de\-liv\-er you and you will glo\-ri\-fy me.
\end{para}

\begin{para}
    \textsuperscript{55:22}\thinspace{}Cast your bur\-den on \textsc{God} and he will sus\-tain you. He will nev\-er al\-low the right\-eous to be moved.
\end{para}

\begin{para}
    \textsuperscript{82:3}\thinspace{}De\-fend the poor and fa\-ther\-less. Do jus\-tice to the afflicted and des\-ti\-tute.
    \textsuperscript{4}\thinspace{}Res\-cue the poor and needy. De\-liv\-er them from the hand of the wick\-ed.
\end{para}

\begin{para}
    \textsuperscript{92:12}\thinspace{}The right\-eous will flour\-ish like the palm tree. He will grow like a ce\-dar in Lebanon.
\end{para}

\begin{para}
    \textsuperscript{96:2}\thinspace{}Sing to \textsc{God}, bless his name. Pro\-claim his sal\-va\-tion from day to day.
    \textsuperscript{3}\thinspace{}De\-clare his glo\-ry a\-mong the nations, his mar\-vel\-lous deeds a\-mong all the peoples.
\end{para}

\begin{para}
    \textsuperscript{100:1}\thinspace{}Make a joy\-ful noise to \textsc{God}, all the earth.
    \textsuperscript{2}\thinspace{}Serve \textsc{God} with glad\-ness. Come be\-fore him with sing\-ing.
\end{para}

\begin{para}
    \textsuperscript{102:17}\thinspace{}\textsc{God} regards the pray\-er of the des\-ti\-tute and does not de\-spise their pray\-er.
\end{para}

\begin{para}
    \textsuperscript{103:8}\thinspace{}\textsc{God} is mer\-ci\-ful and gra\-cious, slow to an\-ger, abounding in lov\-ing-kind\-ness.
\end{para}

\begin{para}
    \textsuperscript{105:1}\thinspace{}Give thanks to \textsc{God}. Call on his name. Make known his deeds a\-mong the peoples.
    \textsuperscript{2}\thinspace{}Sing to him. Sing praises to him. Talk a\-bout all his mar\-vel\-ous deeds.
    \textsuperscript{3}\thinspace{}Glo\-ry in his ho\-ly name. Let the hearts of those who seek \textsc{God} re\-joice.
    \textsuperscript{4}\thinspace{}Seek \textsc{God} and his strength. Seek his face con\-tin\-u\-al\-ly.
    \textsuperscript{5}\thinspace{}Re\-mem\-ber the wonders that he has done, his miracles and the judgments he has pro\-nounced.
\end{para}

\begin{para}
    \textsuperscript{106:1}\thinspace{}Praise \textsc{God}. Give thanks to \textsc{God}, for he is good; for his lov\-ing-kind\-ness endures for\-ev\-er.
\end{para}

\begin{para}
    \textsuperscript{111:10}\thinspace{}The fear of \textsc{God} is the be\-gin\-ning of wis\-dom. Those who o\-bey his commandments have good un\-der\-stand\-ing.
\end{para}

\begin{para}
    \textsuperscript{112:1}\thinspace{}Bless\-ed is the man who fears \textsc{God}, who delights great\-ly in his commandments.
    \textsuperscript{4}\thinspace{}For the up\-right, light arises in the dark\-ness. He is gra\-cious, mer\-ci\-ful and right\-eous.
    \textsuperscript{5}\thinspace{}It is well with the man who deals gen\-er\-ous\-ly and lends, who conducts his af\-fairs with jus\-tice.
\end{para}

\begin{para}
    \textsuperscript{118:8}\thinspace{}It is bet\-ter to take ref\-uge in \textsc{God} than to put con\-fi\-dence in man.
    \textsuperscript{9}\thinspace{}It is bet\-ter to take ref\-uge in \textsc{God} than to put con\-fi\-dence in prin\-ces.
\end{para}

\begin{para}
    \textsuperscript{122:6}\thinspace{}Pray for the peace of Je\-ru\-sa\-lem.
\end{para}

\begin{para}
    \textsuperscript{127:1}\thinspace{}Ex\-cept \textsc{God} build the house, they la\-bor in vain who build it. Ex\-cept \textsc{God} keep the cit\-y, the watch\-man stays a\-wake in vain.
    \textsuperscript{2}\thinspace{}It is vain for you to rise up ear\-ly, to stay up late, to eat the bread of toil---for he gives sleep to his be\-lov\-ed.
\end{para}

\begin{para}
    \textsuperscript{127:3}\thinspace{}Chil\-dren are a her\-it\-age from \textsc{God} and the fruit of the womb is a re\-ward.
\end{para}

\begin{para}
    \textsuperscript{133:1}\thinspace{}How good and pleas\-ant it is for brothers to live to\-geth\-er in u\-ni\-ty.
\end{para}

\begin{para}
    \textsuperscript{146:3}\thinspace{}Do not put your trust in prin\-ces or in the son of man, who can\-not save.
\end{para}


\end{document}

% Copyright 2021 Michael Hendricks
%
% Permission is granted to use, copy, modify, and distribute
% this work for any purpose with or without fee.
%
% This work is provided as-is without any express or implied
% warranty. In no event is the author liable for any damages
% arising from this work.
