%
% see tex.stackexchange.com/a/600/151276
\def\changemargin#1#2{\list{}{\rightmargin#2\leftmargin#1}\item[]}
\let\endchangemargin=\endlist

\documentclass[openany,12pt,english]{book}
\usepackage[utf8]{inputenc}
\usepackage{babel}
\usepackage{microtype}
\usepackage{csquotes}

\usepackage[
  paperwidth=8.5in,
  paperheight=11in,
  top=1in,
  outer=1in,
  bottom=1.25in,
  inner=1.5in
]{geometry}

\title{Confucius}
\author{}
\date{}

\newenvironment{para}{\par\pretolerance=100\tolerance=200\setlength{\emergencystretch}{0.6em}\relax}{\par}

\begin{document}

\maketitle

% Copyright page
{
\thispagestyle{empty}
\vspace*{\fill}
\begin{changemargin}{.8in}{.8in}
\begin{center}
\textcopyright{}~2021 Michael Hendricks

\vspace{0.5\baselineskip}

Permission is granted to use, copy, modify, and distribute
this work for any purpose with or without fee.

\vspace{0.5\baselineskip}

This work is provided as-is without any express or implied
warranty. In no event is the author liable for any damages
arising from this work.
\end{center}
\end{changemargin}
}
\pagenumbering{arabic}
\clearpage{}

% Table of Contents
\clearpage{}

\markboth{}{}
\twocolumn[\begin{changemargin}{.6in}{.6in}
  \begin{center}
    \itshape
    An a\-bridg\-ment of the An\-a\-lects.
    Verse references such
    as~\textsuperscript{1:3}\thinspace{}indicate
    the original source of the text, but
    don't imply a full quotation.
  \end{center}
\end{changemargin}]{}


\begin{para}
    \textsuperscript{1:1}\thinspace{}Is\-n't it a pleas\-ure to learn some\-thing, then to ap\-ply it at the right time? Is\-n't it a joy to have friends come from a\-far? Is\-n't he a su\-pe\-ri\-or man who takes no of\-fense when his mer\-its are ig\-nored?
\end{para}

\begin{para}
    \textsuperscript{1:3}\thinspace{}Cun\-ning words and an in\-gra\-ti\-at\-ing ap\-pear\-ance rare\-ly ac\-com\-pa\-ny be\-nev\-o\-lence.
\end{para}

\begin{para}
    \textsuperscript{1:5}\thinspace{}The Way to guide a state of a thou\-sand chariots: be at\-ten\-tive to busi\-ness and trust\-wor\-thy; be fru\-gal in expenditures and love your fel\-low men; on\-ly mo\-bi\-lize the com\-mon peo\-ple at the prop\-er time of year.
\end{para}

\begin{para}
    \textsuperscript{1:6}\thinspace{}A young man should re\-spect his parents at home, re\-spect his elders a\-broad, say lit\-tle but be trust\-wor\-thy, love the peo\-ple at large but be in\-ti\-mate with the be\-nev\-o\-lent. If he still has en\-er\-gy to spare, let him stud\-y lit\-er\-a\-ture.
\end{para}

\begin{para}
    \textsuperscript{1:10}\thinspace{}Ziqin asked Zigong, “When Con\-fu\-cius arrives in a coun\-try, he al\-ways learns a\-bout its gov\-ern\-ment. Does he seek out this in\-for\-ma\-tion or is it giv\-en to him?” Zigong replied, “Con\-fu\-cius obtains it by be\-ing cor\-dial, up\-right, re\-spect\-ful, fru\-gal, and def\-er\-en\-tial. He seeks it dif\-fer\-ent\-ly than oth\-er men do.”
\end{para}

\begin{para}
    \textsuperscript{1:14}\thinspace{}A su\-pe\-ri\-or man seeks nei\-ther a full stom\-ach nor com\-fort\-a\-ble lodg\-ings. He is dil\-i\-gent in his actions but cau\-tious in his speech. He associates with those who know the Way in or\-der to be put right. Such a man may be said to love learn\-ing.
\end{para}

\begin{para}
    \textsuperscript{1:16}\thinspace{}Don't wor\-ry that men don't rec\-og\-nize your mer\-its; wor\-ry that you don't rec\-og\-nize theirs.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{2:1}\thinspace{}He who governs by vir\-tue is like the North Star: it stays in its place while all the oth\-er stars pay hom\-age to it.
\end{para}

\begin{para}
    \textsuperscript{2:3}\thinspace{}If you lead the peo\-ple with edicts and keep them or\-der\-ly with punishments, they will e\-vade the pun\-ish\-ment and have no sense of shame. If you lead them with vir\-tue and keep them or\-der\-ly with pro\-pri\-e\-ty, they will have a sense of shame and or\-der them\-selves.
\end{para}

\begin{para}
    \textsuperscript{2:6}\thinspace{}Meng Wubo asked a\-bout fil\-ial pi\-e\-ty. Con\-fu\-cius said, “Give your parents noth\-ing to wor\-ry a\-bout oth\-er than ill\-ness.”
\end{para}

\begin{para}
    \textsuperscript{2:7}\thinspace{}Ziyou asked a\-bout fil\-ial pi\-e\-ty. Con\-fu\-cius said, “Now\-a\-days fil\-ial pi\-e\-ty on\-ly means be\-ing a\-ble to feed your parents. But e\-ven dogs and hors\-es re\-ceive this much care. With\-out re\-spect, what's the dif\-fer\-ence?”
\end{para}

\begin{para}
    \textsuperscript{2:8}\thinspace{}Zixia asked a\-bout fil\-ial pi\-e\-ty. Con\-fu\-cius said, “It's the ex\-pres\-sion on your face that's dif\-fi\-cult to man\-age. It's not fil\-ial pi\-e\-ty mere\-ly for the young to shoul\-der the bur\-den when there is work to do, and to de\-fer to their elders when there is wine and food a\-vail\-a\-ble.”
\end{para}

\begin{para}
    \textsuperscript{2:10}\thinspace{}Look at the means a man employs. Ob\-serve his motives. Ex\-am\-ine what gives him com\-fort. How can a man con\-ceal his char\-ac\-ter?
\end{para}

\begin{para}
    \textsuperscript{2:11}\thinspace{}If a man learns the new by reviewing the old, he is wor\-thy to be a teach\-er.
\end{para}

\begin{para}
    \textsuperscript{2:13}\thinspace{}A su\-pe\-ri\-or man acts be\-fore he speaks, and then speaks ac\-cord\-ing to his actions.
    \textsuperscript{14}\thinspace{}A su\-pe\-ri\-or man is in\-clu\-sive and not par\-ti\-san, while an in\-fe\-ri\-or man is the op\-po\-site.
\end{para}

\begin{para}
    \textsuperscript{2:15}\thinspace{}Learn\-ing with\-out think\-ing is a waste. Think\-ing with\-out learn\-ing is dan\-ger\-ous.
\end{para}

\begin{para}
    \textsuperscript{2:17}\thinspace{}Knowl\-edge is to know when you know some\-thing, and to know when you don't.
\end{para}

\begin{para}
    \textsuperscript{2:18}\thinspace{}Zizhang was stud\-y\-ing in or\-der to get an of\-fi\-cial po\-si\-tion. Con\-fu\-cius said, “If you lis\-ten much, put a\-side what is doubt\-ful, and speak cau\-tious\-ly a\-bout the rest, you will make few mistakes. If you ob\-serve much, put a\-side what is per\-i\-lous, and act cau\-tious\-ly on the rest, you will have few regrets. Few mistakes in speech, few regrets in ac\-tion: an of\-fi\-cial po\-si\-tion will fol\-low.”
\end{para}

\begin{para}
    \textsuperscript{2:19}\thinspace{}If you raise up the straight and place them over the crook\-ed, the peo\-ple will fol\-low you. If you raise up the crook\-ed and place them over the straight, the peo\-ple will not fol\-low you.
\end{para}

\begin{para}
    \textsuperscript{2:20}\thinspace{}Ji Kangzi asked, “How can I en\-cour\-age the peo\-ple to be re\-spect\-ful, loy\-al, and dil\-i\-gent?” Con\-fu\-cius said, “Ap\-proach them with dig\-ni\-ty then they will be re\-spect\-ful. Be fil\-ial and kind then they will be loy\-al. Raise up the good and in\-struct the in\-com\-pe\-tent then they will be dil\-i\-gent.”
\end{para}

\begin{para}
    \textsuperscript{2:21}\thinspace{}Some\-one asked Con\-fu\-cius, “Why don't you en\-gage in gov\-ern\-ment?” Con\-fu\-cius said, “The Book of Documents says, ‘Just by be\-ing fil\-ial and be\-ing friend\-ly to your brothers you will have an ef\-fect on gov\-ern\-ment.’ Since this is al\-so en\-gag\-ing in gov\-ern\-ment, why do I need to ‘en\-gage in gov\-ern\-ment’?”
\end{para}

\begin{para}
    \textsuperscript{2:22}\thinspace{}I don't know what to do with an un\-trust\-wor\-thy man. How can you pull a wag\-on that's mis\-sing its yoke pin?
\end{para}

\begin{para}
    \textsuperscript{2:24}\thinspace{}To see what is right and not to do it is a lack of cour\-age.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{3:4}\thinspace{}In rit\-u\-al, pre\-fer fru\-gal\-i\-ty to ex\-trav\-a\-gance. In funerals, pre\-fer grief to for\-mal\-i\-ty.
\end{para}

\begin{para}
    \textsuperscript{3:12}\thinspace{}If I don't par\-tic\-i\-pate in a sac\-ri\-fice, it's as if I have not sacrificed at all.
\end{para}

\begin{para}
    \textsuperscript{3:15}\thinspace{}When Con\-fu\-cius entered the Grand Tem\-ple, he asked questions a\-bout eve\-ry\-thing. Some\-one remarked, “Who says that Con\-fu\-cius knows a\-bout rit\-u\-al? When visiting the Grand Tem\-ple, he asked questions a\-bout eve\-ry\-thing.” Con\-fu\-cius, hear\-ing this, said, “This is the rit\-u\-al.”
\end{para}

\begin{para}
    \textsuperscript{3:21}\thinspace{}Don't ex\-plain a\-way what is al\-read\-y done. Don't ar\-gue a\-gainst what is al\-read\-y ac\-com\-plished. Don't as\-sign blame for what is al\-read\-y past.
\end{para}

\begin{para}
    \textsuperscript{3:26}\thinspace{}Hold\-ing a high po\-si\-tion with\-out tol\-er\-ance, per\-form\-ing rit\-u\-al with\-out rev\-er\-ence, joining a fu\-ner\-al with\-out grief: how can I look at such things?
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{4:1}\thinspace{}It is be\-nev\-o\-lence that makes a neigh\-bor\-hood beau\-ti\-ful. How can a man be con\-sid\-ered wise if he chooses not to re\-side a\-mong be\-nev\-o\-lence?
\end{para}

\begin{para}
    \textsuperscript{4:2}\thinspace{}Those who lack be\-nev\-o\-lence can en\-dure nei\-ther ad\-ver\-si\-ty nor en\-joy\-ment for long. The be\-nev\-o\-lent are con\-tent with be\-nev\-o\-lence. The wise cul\-ti\-vate be\-nev\-o\-lence for its ad\-van\-tage.
\end{para}

\begin{para}
    \textsuperscript{4:5}\thinspace{}All men de\-sire wealth and hon\-or, but if they are obtained by not fol\-low\-ing the Way, they should not be kept. All men hate pov\-er\-ty and dis\-grace, but if they can't be avoided by fol\-low\-ing the Way, they should not be avoided. If a su\-pe\-ri\-or man abandons be\-nev\-o\-lence, how is he wor\-thy of that name? A su\-pe\-ri\-or man nev\-er leaves be\-nev\-o\-lence e\-ven for the space of a sin\-gle meal. In moments of haste, he acts ac\-cord\-ing to it. In times of dif\-fi\-cul\-ty, he acts ac\-cord\-ing to it.
\end{para}

\begin{para}
    \textsuperscript{4:6}\thinspace{}One who re\-al\-ly loves be\-nev\-o\-lence puts noth\-ing a\-bove it. One who re\-al\-ly hates what is not be\-nev\-o\-lent would not al\-low it to ap\-proach him. Has any\-one de\-vot\-ed all his strength to be\-nev\-o\-lence for a sin\-gle day? I have\-n't seen any\-one who lacked the strength to do so.
\end{para}

\begin{para}
    \textsuperscript{4:7}\thinspace{}Peo\-ple make errors ac\-cord\-ing to the type of per\-son they are. Ob\-serve their errors and you can know their de\-gree of be\-nev\-o\-lence.
\end{para}

\begin{para}
    \textsuperscript{4:9}\thinspace{}A gen\-tle\-man who sets his heart on the Way, but is a\-shamed of poor clothes and poor food, is not worth en\-gag\-ing in dis\-cus\-sion.
\end{para}

\begin{para}
    \textsuperscript{4:10}\thinspace{}When a su\-pe\-ri\-or man deals with the world he is not in\-var\-i\-a\-bly for or a\-gainst any\-thing. He sides with what is right.
\end{para}

\begin{para}
    \textsuperscript{4:11}\thinspace{}A su\-pe\-ri\-or man cherishes vir\-tue; an in\-fe\-ri\-or man cherishes land. A su\-pe\-ri\-or man cherishes jus\-tice; an in\-fe\-ri\-or man cherishes fa\-vors.
\end{para}

\begin{para}
    \textsuperscript{4:12}\thinspace{}One who acts out of per\-son\-al prof\-it incurs much re\-sent\-ment.
\end{para}

\begin{para}
    \textsuperscript{4:14}\thinspace{}Don't wor\-ry that you have no po\-si\-tion; wor\-ry a\-bout what it takes to mer\-it one. Don't wor\-ry that no one knows you; seek to be worth know\-ing.
\end{para}

\begin{para}
    \textsuperscript{4:16}\thinspace{}A su\-pe\-ri\-or man understands what is right. An in\-fe\-ri\-or man understands per\-son\-al prof\-it.
\end{para}

\begin{para}
    \textsuperscript{4:17}\thinspace{}When you see a wor\-thy man, think of be\-com\-ing his e\-qual; when you see an un\-wor\-thy man, look with\-in and ex\-am\-ine your\-self.
\end{para}

\begin{para}
    \textsuperscript{4:18}\thinspace{}In serv\-ing your parents, you may ad\-mon\-ish them gen\-tly; if you see that they don't fol\-low your ad\-vice, re\-main re\-spect\-ful, don't be\-come dis\-o\-be\-di\-ent, and work with\-out complaining.
\end{para}

\begin{para}
    \textsuperscript{4:22}\thinspace{}The ancients were re\-luc\-tant to speak be\-cause they would have been a\-shamed if their actions did\-n't live up to their words.
\end{para}

\begin{para}
    \textsuperscript{4:23}\thinspace{}One rare\-ly miss\-es the mark be\-cause of self-control.
\end{para}

\begin{para}
    \textsuperscript{4:24}\thinspace{}A su\-pe\-ri\-or man wants to be slow in speech but quick in ac\-tion.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{5:5}\thinspace{}What is the use of el\-o\-quence? Those who give a quick re\-tort of\-ten in\-cur the ha\-tred of others.
\end{para}

\begin{para}
    \textsuperscript{5:10}\thinspace{}Zai Yu was sleep\-ing dur\-ing the day. Con\-fu\-cius said, “Rot\-ten wood can\-not be carved, and a wall of ma\-nure can\-not be trow\-eled. What is the use of scolding Zai Yu? I used to lis\-ten to what peo\-ple said and trust them to act ac\-cord\-ing\-ly. Now I lis\-ten to what peo\-ple say and watch what they do. It is be\-cause of Zai Yu that I have changed.”
\end{para}

\begin{para}
    \textsuperscript{5:11}\thinspace{}I have\-n't seen a tru\-ly stead\-fast man. How can a man who is full of desires be stead\-fast?
\end{para}

\begin{para}
    \textsuperscript{5:15}\thinspace{}Kong the Re\-fined was men\-tal\-ly quick, fond of learn\-ing, and not a\-shamed to ask questions of his inferiors. That's why he was called “Re\-fined”.
\end{para}

\begin{para}
    \textsuperscript{5:16}\thinspace{}Zi Chan had four characteristics of the Way of a su\-pe\-ri\-or man: he was re\-spect\-ful in his pri\-vate con\-duct, re\-spect\-ful in serv\-ing his superiors, gen\-er\-ous in car\-ing for the peo\-ple, and just in employing the peo\-ple.
\end{para}

\begin{para}
    \textsuperscript{5:17}\thinspace{}Yang Ping Zhong ex\-celled in his re\-la\-tions with others. E\-ven af\-ter long ac\-quaint\-ance he con\-tin\-ued to treat them with re\-spect.
\end{para}

\begin{para}
    \textsuperscript{5:20}\thinspace{}Ji Wen Zi al\-ways thought three times be\-fore act\-ing. Hear\-ing this, Con\-fu\-cius said, “Twice is e\-nough.”
\end{para}

\begin{para}
    \textsuperscript{5:23}\thinspace{}Bo Yi and Shu Qi did not re\-mem\-ber old grievances, so there was lit\-tle re\-sent\-ment a\-gainst them.
\end{para}

\begin{para}
    \textsuperscript{5:25}\thinspace{}Glib talk, an in\-gra\-ti\-at\-ing ex\-pres\-sion, ob\-se\-qui\-ous con\-duct, and act\-ing friend\-ly to\-wards some\-one while concealing your re\-sent\-ment for them: I am a\-shamed of such be\-hav\-ior.
\end{para}

\begin{para}
    \textsuperscript{5:26}\thinspace{}My heart's de\-sire is to bring peace to the old, to share trust with my friends, and to cher\-ish the young.
\end{para}

\begin{para}
    \textsuperscript{5:27}\thinspace{}I have not yet met any\-one who can see his own faults and in\-ward\-ly ac\-cuse him\-self.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{6:3}\thinspace{}Yanhui loved learn\-ing. He nev\-er took his an\-ger out on others, and nev\-er made the same mis\-take twice.
    \textsuperscript{11}\thinspace{}How ad\-mi\-ra\-ble he was! He lived on a bowlful of rice to eat, a la\-dle\-ful of wa\-ter to drink, and a hov\-el for shel\-ter. Others could not have en\-dured such mis\-er\-y, yet Yanhui's joy was un\-changed.
\end{para}

\begin{para}
    \textsuperscript{6:12}\thinspace{}Ran Qiu said, “It's not that I don't en\-joy your Way, but I don't have the strength to fol\-low it.” Con\-fu\-cius said, “Those who don't have the strength col\-lapse a\-long the way, but you have drawn a line be\-fore starting.”
\end{para}

\begin{para}
    \textsuperscript{6:15}\thinspace{}Meng Zhifan was not boast\-ful. He once held the rear dur\-ing a re\-treat. As they reached the cit\-y gate, he whipped his horse, say\-ing, “It's not that I dared to be last, on\-ly that my horse would\-n't go.”
\end{para}

\begin{para}
    \textsuperscript{6:18}\thinspace{}When nat\-u\-ral dis\-po\-si\-tion prevails over re\-fine\-ment, you get a sav\-age. When re\-fine\-ment prevails over nat\-u\-ral dis\-po\-si\-tion, you get a pe\-dan\-tic clerk. When nat\-u\-ral dis\-po\-si\-tion and re\-fine\-ment are prop\-er\-ly blend\-ed, you get a su\-pe\-ri\-or man.
\end{para}

\begin{para}
    \textsuperscript{6:28}\thinspace{}If I have done wrong, may Heav\-en de\-test and a\-ban\-don me!
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{7:2}\thinspace{}It's nat\-u\-ral for me to store up knowl\-edge si\-lent\-ly, to learn with\-out sa\-ti\-e\-ty, and to teach others with\-out wea\-ri\-ness.
\end{para}

\begin{para}
    \textsuperscript{7:3}\thinspace{}My worries are fail\-ure to cul\-ti\-vate vir\-tue, fail\-ure to ex\-plore what I learn, in\-a\-bil\-i\-ty to fol\-low what is right when I hear it, and in\-a\-bil\-i\-ty to change what is not good.
\end{para}

\begin{para}
    \textsuperscript{7:6}\thinspace{}Set your heart on the Way. Base your\-self in vir\-tue. Re\-ly on be\-nev\-o\-lence. En\-joy the arts.
\end{para}

\begin{para}
    \textsuperscript{7:7}\thinspace{}I have nev\-er de\-nied in\-struc\-tion to any\-one, e\-ven if his own resources could on\-ly af\-ford a bun\-dle of dried meat for tu\-i\-tion.
\end{para}

\begin{para}
    \textsuperscript{7:8}\thinspace{}I on\-ly en\-light\-en those who are driv\-en by ea\-ger\-ness to learn and who are anx\-ious\-ly try\-ing to ex\-plain their ideas. If I lift up one cor\-ner and they can\-not come back with the oth\-er three, I don't re\-peat my\-self.
\end{para}

\begin{para}
    \textsuperscript{7:9}\thinspace{}When Con\-fu\-cius was eat\-ing in the pres\-ence of a mourn\-er, he nev\-er ate his fill.
\end{para}

\begin{para}
    \textsuperscript{7:11}\thinspace{}Few have the a\-bil\-i\-ty to act when employed and to hide when dismissed. If I were com\-mand\-ing the Three Armies, who would I take with me? A man who would wres\-tle a ti\-ger with bare hands, or cross a riv\-er on foot, dy\-ing with\-out re\-gret---I would not take that man. It must be some\-one who approaches a task with ap\-pre\-hen\-sion, likes to make plans, and can ex\-e\-cute them suc\-cess\-ful\-ly.
\end{para}

\begin{para}
    \textsuperscript{7:12}\thinspace{}If wealth can be sought de\-cent\-ly, I will do so, e\-ven if I have to have a low\-ly job. If it can\-not be sought de\-cent\-ly, I will fol\-low what I love.
\end{para}

\begin{para}
    \textsuperscript{7:15}\thinspace{}Bo Yi and Shu Qi sought be\-nev\-o\-lence and got it. Why should they have complaints?
\end{para}

\begin{para}
    \textsuperscript{7:16}\thinspace{}Coarse rice to eat, wa\-ter to drink, and your bent arm for a pil\-low---there is still joy in the midst of these things. Wealth and hon\-or obtained by un\-right\-eous means are like fleet\-ing clouds to me.
\end{para}

\begin{para}
    \textsuperscript{7:17}\thinspace{}Give me a few more years; if I can stud\-y un\-til fif\-ty, I may be free from great mistakes.
\end{para}

\begin{para}
    \textsuperscript{7:19}\thinspace{}I'm the kind of man who is so ea\-ger for knowl\-edge that he forgets to eat, who is so joy\-ful that he forgets to wor\-ry, and who is not a\-ware that old age is com\-ing.
\end{para}

\begin{para}
    \textsuperscript{7:20}\thinspace{}I was not born with knowl\-edge, but I love an\-tiq\-ui\-ty and ear\-nest\-ly seek it.
\end{para}

\begin{para}
    \textsuperscript{7:22}\thinspace{}When walk\-ing with any two peo\-ple, they must have some\-thing to teach me. I se\-lect their strengths and fol\-low them, their weaknesses and cor\-rect my\-self.
\end{para}

\begin{para}
    \textsuperscript{7:23}\thinspace{}Heav\-en gave birth to the vir\-tue in me. What can an\-gry rulers do to me?
\end{para}

\begin{para}
    \textsuperscript{7:24}\thinspace{}My friends, I hide noth\-ing from you. What\-ev\-er I do, I share with you. That is who I am.
\end{para}

\begin{para}
    \textsuperscript{7:26}\thinspace{}I can't hope to meet a sage, but I would be con\-tent to meet a su\-pe\-ri\-or man. I can't hope to meet a tru\-ly good man, but I would be con\-tent to meet a stead\-fast man. It's hard to be stead\-fast when noth\-ing pretends to be some\-thing, emp\-ti\-ness pretends to be full\-ness, and pov\-er\-ty pretends to be com\-fort.
\end{para}

\begin{para}
    \textsuperscript{7:27}\thinspace{}Con\-fu\-cius fished with a line, but did\-n't use a net. He hunt\-ed, but did\-n't shoot roosting birds.
\end{para}

\begin{para}
    \textsuperscript{7:28}\thinspace{}There may be those who can act with\-out knowl\-edge, but I am not one of them. I hear much, se\-lect what is good, and fol\-low it. I see much, and re\-mem\-ber it.
\end{para}

\begin{para}
    \textsuperscript{7:29}\thinspace{}To ap\-prove of a man's vis\-it does\-n't mean ap\-prov\-ing of what he does af\-ter he leaves. When a man purifies him\-self to ap\-proach me, I ap\-prove of his pu\-ri\-ty with\-out en\-dors\-ing his past or his fu\-ture.
\end{para}

\begin{para}
    \textsuperscript{7:30}\thinspace{}Is be\-nev\-o\-lence far a\-way? As soon as I want be\-nev\-o\-lence, it is at hand.
\end{para}

\begin{para}
    \textsuperscript{7:31}\thinspace{}I'm for\-tu\-nate. When I make a mis\-take, others al\-ways no\-tice it.
\end{para}

\begin{para}
    \textsuperscript{7:34}\thinspace{}How would I dare claim to be be\-nev\-o\-lent or a sage? Rath\-er, it may be said that I strive for them with\-out flag\-ging, and teach others with\-out wea\-ri\-ness.
\end{para}

\begin{para}
    \textsuperscript{7:36}\thinspace{}Ex\-trav\-a\-gance leads to ar\-ro\-gance; fru\-gal\-i\-ty leads to stin\-gi\-ness. It's bet\-ter to be stin\-gy than ar\-ro\-gant.
\end{para}

\begin{para}
    \textsuperscript{7:37}\thinspace{}A su\-pe\-ri\-or man is eas\-y\-go\-ing and un\-per\-turbed. An in\-fe\-ri\-or man is al\-ways tense and anx\-ious.
\end{para}

\begin{para}
    \textsuperscript{7:38}\thinspace{}Con\-fu\-cius was warm yet stern; au\-thor\-i\-ta\-tive but not fierce; re\-spect\-ful yet at ease.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{8:1}\thinspace{}Tai Bo can be said to have had the highest vir\-tue. Three times he de\-clined his claim to the king\-dom.
\end{para}

\begin{para}
    \textsuperscript{8:2}\thinspace{}Cour\-te\-sy with\-out pro\-pri\-e\-ty is tire\-some. Cau\-tion with\-out pro\-pri\-e\-ty is ti\-mid\-i\-ty. Brav\-er\-y with\-out pro\-pri\-e\-ty is cha\-os. Can\-dor with\-out pro\-pri\-e\-ty is rude\-ness.
\end{para}

\begin{para}
    \textsuperscript{8:9}\thinspace{}The peo\-ple can be made to fol\-low a path, but they can't be made to un\-der\-stand it.
\end{para}

\begin{para}
    \textsuperscript{8:10}\thinspace{}A man who loves brav\-er\-y and hates pov\-er\-ty will pro\-ceed to un\-ru\-ly be\-hav\-ior. A man with\-out be\-nev\-o\-lence who is de\-spised by others too much will pro\-ceed to un\-ru\-ly be\-hav\-ior.
\end{para}

\begin{para}
    \textsuperscript{8:11}\thinspace{}If a man had talents as ad\-mi\-ra\-ble as the Duke of Zhou, but was ar\-ro\-gant and stin\-gy, the rest would not be worth no\-tic\-ing.
\end{para}

\begin{para}
    \textsuperscript{8:13}\thinspace{}Have sin\-cere faith, love learn\-ing, and de\-fend the good Way un\-to death. Don't en\-ter a state that is in per\-il, and don't stay in one that is in cha\-os. Show your\-self when the Way prevails in the world, but hide when it does not. When the Way prevails in the state, it is shame\-ful to be poor and hum\-ble. When the Way does not pre\-vail in the state, it is shame\-ful to be rich and no\-ble.
\end{para}

\begin{para}
    \textsuperscript{8:14}\thinspace{}Don't plan the policies of a po\-si\-tion that you don't hold.
\end{para}

\begin{para}
    \textsuperscript{8:21}\thinspace{}I can find no fault in Yu. His food and drink were fru\-gal, yet his offerings to the spirits and gods showed ut\-ter de\-vo\-tion. His clothes were coarse, yet his rit\-u\-al cap and robes were mag\-nif\-i\-cent. His dwell\-ing was hum\-ble, yet he exhausted his strength on ir\-ri\-ga\-tion ditches and canals.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{9:3}\thinspace{}The rules of rit\-u\-al pre\-scribe a hemp cap. Now\-a\-days it's made of silk. That's more fru\-gal. I fol\-low the com\-mon prac\-tice. The rules of rit\-u\-al pre\-scribe bow\-ing be\-low the steps. Now\-a\-days peo\-ple bow af\-ter as\-cend\-ing. That's ar\-ro\-gant. Al\-though it goes a\-gainst the com\-mon prac\-tice, I bow be\-low the steps.
\end{para}

\begin{para}
    \textsuperscript{9:4}\thinspace{}Con\-fu\-cius for\-bade four things: don't act on spec\-u\-la\-tion, don't de\-mand cer\-tain\-ty, don't be stub\-born, and don't be e\-go\-tis\-ti\-cal.
\end{para}

\begin{para}
    \textsuperscript{9:5}\thinspace{}When Con\-fu\-cius was in dan\-ger in Kuang, he said, “Since King Wen died, does\-n't his cul\-ture re\-side in me? If Heav\-en does\-n't want this cul\-ture destroyed, what can the peo\-ple of Kuang do to me?”
\end{para}

\begin{para}
    \textsuperscript{9:6}\thinspace{}I was poor when I was young, so I learn\-ed man\-y me\-ni\-al skills. Must a su\-pe\-ri\-or man be skilled in man\-y things? No, not man\-y.
\end{para}

\begin{para}
    \textsuperscript{9:8}\thinspace{}Do I have knowl\-edge? No, I don't. A rus\-tic asked me a ques\-tion and my mind went blank. I ham\-mered at both ends of the ques\-tion un\-til I got eve\-ry\-thing out of it.
\end{para}

\begin{para}
    \textsuperscript{9:13}\thinspace{}Zigong asked, “If you had a beau\-ti\-ful piece of jade here, would you hide it safe\-ly in a box or try to get a good price and sell it?” Con\-fu\-cius said, “Sell it! All I'm waiting for is the right of\-fer.”
\end{para}

\begin{para}
    \textsuperscript{9:16}\thinspace{}I have no dif\-fi\-cul\-ty serv\-ing high officials when a\-broad, serv\-ing my elders at home, giv\-ing my best efforts in fu\-ner\-al arrangements, and avoiding trou\-ble from wine.
\end{para}

\begin{para}
    \textsuperscript{9:18}\thinspace{}I have nev\-er seen a man who loves vir\-tue as much as he loves beau\-ty in wom\-en.
\end{para}

\begin{para}
    \textsuperscript{9:19}\thinspace{}It's like build\-ing a mound: if I stop be\-fore the last bas\-ket of earth, it re\-mains un\-fin\-ished. It's like fill\-ing a ditch: al\-though I've dumped in on\-ly a sin\-gle bas\-ket of earth, if I con\-tin\-ue then I'm mak\-ing prog\-ress.
\end{para}

\begin{para}
    \textsuperscript{9:23}\thinspace{}The young\-er gen\-er\-a\-tion should be held in awe. How do we know that it will not e\-qual the pres\-ent one? But if a man reaches for\-ty or fif\-ty years old with\-out dis\-tin\-guish\-ing him\-self, he no long\-er deserves awe.
\end{para}

\begin{para}
    \textsuperscript{9:24}\thinspace{}I can do noth\-ing with the man who enjoys po\-lite words with\-out unfolding their pur\-pose, or who agrees with ex\-em\-pla\-ry words with\-out reforming him\-self.
\end{para}

\begin{para}
    \textsuperscript{9:25}\thinspace{}Put loy\-al\-ty and trust\-wor\-thi\-ness a\-bove eve\-ry\-thing else. On\-ly have friends who are your mor\-al equals. When you make a mis\-take, don't be a\-fraid to cor\-rect it.
\end{para}

\begin{para}
    \textsuperscript{9:30}\thinspace{}You may be a\-ble to stud\-y to\-geth\-er, and yet be un\-a\-ble to pur\-sue the Way to\-geth\-er. You may be a\-ble to pur\-sue the Way to\-geth\-er, and yet be un\-a\-ble to take a stand to\-geth\-er. You may be a\-ble to take a stand to\-geth\-er, and yet be un\-a\-ble to share au\-thor\-i\-ty to\-geth\-er.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{10:1}\thinspace{}In his vil\-lage, Con\-fu\-cius was most def\-er\-en\-tial, as if he were un\-a\-ble to speak. In the an\-ces\-tral tem\-ple or at court, he was el\-o\-quent but cau\-tious.
\end{para}

\begin{para}
    \textsuperscript{10:8}\thinspace{}E\-ven if there was plen\-ty of meat, he would not eat more meat than rice. On\-ly with wine did he have no set lim\-it, but he nev\-er drank to the point of con\-fu\-sion.
\end{para}

\begin{para}
    \textsuperscript{10:10}\thinspace{}He did\-n't con\-verse while eat\-ing, or talk in bed.
    \textsuperscript{11}\thinspace{}E\-ven with a sim\-ple meal of coarse rice and veg\-e\-ta\-ble broth, he made an of\-fer\-ing with so\-lem\-ni\-ty.
\end{para}

\begin{para}
    \textsuperscript{10:17}\thinspace{}When the stables burned, Con\-fu\-cius returned from court and asked, “Was any\-one hurt?” He did\-n't ask a\-bout the hors\-es.
\end{para}

\begin{para}
    \textsuperscript{10:22}\thinspace{}When a friend died who had no fam\-i\-ly to make arrangements, he said, “Let the fu\-ner\-al be at my home.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{11:4}\thinspace{}Yan Hui is no help to me. Eve\-ry\-thing I say pleases him.
\end{para}

\begin{para}
    \textsuperscript{11:12}\thinspace{}If you aren't yet a\-ble to serve peo\-ple, how would you be a\-ble to serve the spirits. If you don't yet un\-der\-stand life, how could you un\-der\-stand death?
\end{para}

\begin{para}
    \textsuperscript{11:14}\thinspace{}The peo\-ple of Lu were re\-build\-ing the Long Treas\-ur\-y. Min Ziqian said, “Why not re\-store the old struc\-ture? Why must it be changed com\-plete\-ly?” Con\-fu\-cius said, “This man ei\-ther says noth\-ing, or his words hit the mark.”
\end{para}

\begin{para}
    \textsuperscript{11:16}\thinspace{}Go\-ing too far is as bad as fall\-ing short.
\end{para}

\begin{para}
    \textsuperscript{11:21}\thinspace{}If a man's opinions are sound, is he a su\-pe\-ri\-or man or is it on\-ly a sol\-emn ap\-pear\-ance?
\end{para}

\begin{para}
    \textsuperscript{11:22}\thinspace{}Zilu asked, “Should I im\-me\-di\-ate\-ly put in\-to prac\-tice what I learn?” Con\-fu\-cius said, “You have a fa\-ther and el\-der brothers to con\-sult first.” Ranyou asked the same ques\-tion. Con\-fu\-cius said, “You should im\-me\-di\-ate\-ly put in\-to prac\-tice what you learn.” Gongxi Hua said, “I'm con\-fused. May I ask for an ex\-pla\-na\-tion?” Con\-fu\-cius said, “Ranyou holds back, so I urged him for\-ward. Zilu has en\-er\-gy for two, so I held him back.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{12:1}\thinspace{}Yan Hui asked a\-bout be\-nev\-o\-lence. Con\-fu\-cius said, “Be\-nev\-o\-lence is over\-com\-ing one\-self and returning to pro\-pri\-e\-ty. Be\-nev\-o\-lence co\-mes from one\-self. How could it come from others? Don't look at any\-thing im\-prop\-er. Don't lis\-ten to any\-thing im\-prop\-er. Don't say any\-thing im\-prop\-er. Don't do any\-thing im\-prop\-er.”
\end{para}

\begin{para}
    \textsuperscript{12:2}\thinspace{}Zhonggong asked a\-bout be\-nev\-o\-lence. Con\-fu\-cius said, “When a\-broad, be\-have as if you were re\-ceiv\-ing an im\-por\-tant guest. When employing the peo\-ple, be\-have as if you were assisting at an im\-por\-tant sac\-ri\-fice. Don't im\-pose on others what you don't want for your\-self.”
\end{para}

\begin{para}
    \textsuperscript{12:3}\thinspace{}Si\-ma Niu asked a\-bout be\-nev\-o\-lence. Con\-fu\-cius said, “A be\-nev\-o\-lent per\-son is re\-luc\-tant to speak. When do\-ing it is dif\-fi\-cult, how can you not be re\-luc\-tant to speak a\-bout it.”
\end{para}

\begin{para}
    \textsuperscript{12:4}\thinspace{}Si\-ma Niu asked a\-bout the su\-pe\-ri\-or man. Con\-fu\-cius said, “A su\-pe\-ri\-or man is free from wor\-ry and fear. If in\-tro\-spec\-tion finds noth\-ing to be a\-shamed of, why should he have wor\-ry or fear?”
\end{para}

\begin{para}
    \textsuperscript{12:6}\thinspace{}Zizhang asked a\-bout dis\-cern\-ment. Con\-fu\-cius said, “He who is not in\-flu\-enced by slan\-der that saturates and by accusations that spread like a rash, can be called dis\-cern\-ing. He can al\-so be called far\-sight\-ed.”
\end{para}

\begin{para}
    \textsuperscript{12:7}\thinspace{}Zigong asked a\-bout gov\-ern\-ment. Con\-fu\-cius said, “Suf\-fi\-cient food, suf\-fi\-cient arms, and the trust of the peo\-ple. If I had to give up one of these three, I would give up the arms. If I had to give up one of the remaining two, I would give up the food. Death has been with us since an\-cient times, but with\-out trust the peo\-ple can\-not stand.”
\end{para}

\begin{para}
    \textsuperscript{12:10}\thinspace{}Zizhang asked a\-bout ac\-cu\-mu\-lat\-ing vir\-tue. Con\-fu\-cius said, “Make these your guid\-ing principles: loy\-al\-ty, trust\-wor\-thi\-ness, and fol\-low\-ing what is right. This is the way to ac\-cu\-mu\-late vir\-tue.”
\end{para}

\begin{para}
    \textsuperscript{12:13}\thinspace{}In hear\-ing lawsuits, I am the same as any\-one. What we need is to make it so that there are no lawsuits.
\end{para}

\begin{para}
    \textsuperscript{12:14}\thinspace{}Zizhang asked a\-bout gov\-ern\-ment. Con\-fu\-cius said, “Pon\-der it with\-out tir\-ing, and car\-ry it out with loy\-al\-ty.”
\end{para}

\begin{para}
    \textsuperscript{12:15}\thinspace{}Stud\-y cul\-ture broad\-ly, and re\-strain your\-self with pro\-pri\-e\-ty. You will not stray from what is right.
\end{para}

\begin{para}
    \textsuperscript{12:16}\thinspace{}The su\-pe\-ri\-or man perfects the good in peo\-ple, he does not per\-fect the bad. The in\-fe\-ri\-or man does the op\-po\-site.
\end{para}

\begin{para}
    \textsuperscript{12:19}\thinspace{}Ji Kangzi asked Con\-fu\-cius a\-bout gov\-ern\-ment: “What if I were to kill those who don't fol\-low the Way in or\-der to help those who do?” Con\-fu\-cius said, “In administering your gov\-ern\-ment, what use is kill\-ing? If you de\-sire what is good, the peo\-ple will be good. The vir\-tue of the su\-pe\-ri\-or man is like the wind, and the vir\-tue of the in\-fe\-ri\-or man is like the grass. When the wind blows, the grass is sure to bend.”
\end{para}

\begin{para}
    \textsuperscript{12:20}\thinspace{}Zizhang asked, “When can a gen\-tle\-man be called dis\-tin\-guished? By this I mean that he is sure to be known in his state and in his clan.” Con\-fu\-cius said, “That is fame, not dis\-tinc\-tion. A dis\-tin\-guished man has up\-right char\-ac\-ter, loves what is right, examines what peo\-ple say and observes their fa\-cial expressions, and is mind\-ful of def\-er\-ence to others. Such a man will be dis\-tin\-guished in his state and in his clan. As for fame, it is e\-nough to be whol\-ly con\-fi\-dent a\-bout as\-sum\-ing the ap\-pear\-ance of be\-nev\-o\-lence while act\-ing to the con\-tra\-ry. Such a man is sure to be known in his state and in his clan.”
\end{para}

\begin{para}
    \textsuperscript{12:21}\thinspace{}Fan Chi asked, “How do I ac\-cu\-mu\-late vir\-tue, re\-form wick\-ed\-ness, and rec\-og\-nize con\-fu\-sion?” Con\-fu\-cius said, “Ex\-cel\-lent ques\-tion! Put the ef\-fort be\-fore the re\-ward: is that not ac\-cu\-mu\-lat\-ing vir\-tue? At\-tack your own evil rath\-er than the evil of others: is that not reforming wick\-ed\-ness? In a mo\-ment of an\-ger, to for\-get your own safe\-ty and that of your fam\-i\-ly: is that not con\-fu\-sion?”
\end{para}

\begin{para}
    \textsuperscript{12:23}\thinspace{}Zigong asked how to treat friends. Con\-fu\-cius said, “Loy\-al\-ly ad\-vise them, and skill\-ful\-ly show them the Way. If that does\-n't work, then stop. Don't dis\-grace your\-self.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{13:1}\thinspace{}Zilu asked a\-bout gov\-ern\-ment. Con\-fu\-cius said, “Set an ex\-am\-ple for them, and en\-cour\-age them. Be tire\-less.”
\end{para}

\begin{para}
    \textsuperscript{13:2}\thinspace{}Zhonggong asked a\-bout gov\-ern\-ment. Con\-fu\-cius said, “Set an ex\-am\-ple for your officers. Grant par\-don for mi\-nor offenses. Pro\-mote men of vir\-tue and tal\-ent.”
\end{para}

\begin{para}
    \textsuperscript{13:3}\thinspace{}A su\-pe\-ri\-or man re\-mains si\-lent a\-bout things he does\-n't un\-der\-stand. He permits no care\-less\-ness in his speech.
\end{para}

\begin{para}
    \textsuperscript{13:5}\thinspace{}Con\-sid\-er a man who can re\-cite the three hun\-dred Poems. If he fails when you en\-trust him with gov\-ern\-ance, and can\-not act on his own in\-i\-ti\-a\-tive, what use is all his learn\-ing?
\end{para}

\begin{para}
    \textsuperscript{13:6}\thinspace{}If a man is up\-right in his per\-son\-al con\-duct, others will fol\-low with\-out him giv\-ing orders. If a man is not up\-right in his per\-son\-al con\-duct, e\-ven if he gives orders, others won't o\-bey.
\end{para}

\begin{para}
    \textsuperscript{13:12}\thinspace{}E\-ven with a true king, it would still take a gen\-er\-a\-tion for be\-nev\-o\-lence to pre\-vail.
\end{para}

\begin{para}
    \textsuperscript{13:13}\thinspace{}If a man can cor\-rect him\-self, what dif\-fi\-cul\-ty will he have in governing? If he can't cor\-rect him\-self, how can he cor\-rect others?
\end{para}

\begin{para}
    \textsuperscript{13:15}\thinspace{}A sin\-gle say\-ing can\-not ru\-in a coun\-try, but there is one say\-ing: “The on\-ly pleas\-ure of rul\-ing is that no one contradicts you.” If a rul\-er is good and no one contradicts him, that's fi\-ne. But if a rul\-er is not good and no one contradicts him, then is\-n't this al\-most a say\-ing that could ru\-in a coun\-try?
\end{para}

\begin{para}
    \textsuperscript{13:16}\thinspace{}The Gov\-er\-nor of She asked a\-bout gov\-ern\-ment. Con\-fu\-cius said, “Peo\-ple near\-by are pleased, and peo\-ple far a\-way come.”
\end{para}

\begin{para}
    \textsuperscript{13:17}\thinspace{}Zixia asked a\-bout gov\-ern\-ment. Con\-fu\-cius said, “Don't try to rush things. Don't look for small advantages. If you try to rush things, you won't reach your goal. If you look for small advantages, you won't ac\-com\-plish great things.”
\end{para}

\begin{para}
    \textsuperscript{13:19}\thinspace{}Be cour\-te\-ous at home. Be re\-spect\-ful in han\-dling pub\-lic af\-fairs. Be loy\-al in deal\-ing with others. E\-ven a\-mong barbarians, you can't dis\-card these.
\end{para}

\begin{para}
    \textsuperscript{13:21}\thinspace{}If I can't find mod\-er\-ate men to as\-so\-ci\-ate with, I must find the reck\-less and the cau\-tious. The reck\-less ad\-vance to\-wards the goal. The cau\-tious have lines they won't cross.
\end{para}

\begin{para}
    \textsuperscript{13:23}\thinspace{}A su\-pe\-ri\-or man seeks har\-mo\-ny with\-out con\-form\-i\-ty. An in\-fe\-ri\-or man seeks con\-form\-i\-ty with\-out har\-mo\-ny.
\end{para}

\begin{para}
    \textsuperscript{13:24}\thinspace{}It's not e\-nough for all the peo\-ple to love you. It's not e\-nough for all the peo\-ple to hate you. It's bet\-ter if the good peo\-ple love you and the bad peo\-ple hate you.
\end{para}

\begin{para}
    \textsuperscript{13:25}\thinspace{}If you try to please a su\-pe\-ri\-or man with\-out fol\-low\-ing the Way, he will not be pleased; but when employing others, he on\-ly assigns tasks ac\-cord\-ing to their ca\-pac\-i\-ty. If you try to please an in\-fe\-ri\-or man with\-out fol\-low\-ing the Way, he will be pleased; but when employing others, he expects them to be good at eve\-ry\-thing.
\end{para}

\begin{para}
    \textsuperscript{13:26}\thinspace{}A su\-pe\-ri\-or man has a dig\-ni\-fied ease with\-out be\-ing ar\-ro\-gant. An in\-fe\-ri\-or man is ar\-ro\-gant with\-out a dig\-ni\-fied ease.
\end{para}

\begin{para}
    \textsuperscript{13:27}\thinspace{}Firm, res\-o\-lute, sim\-ple, slow to speak: these are close to be\-nev\-o\-lence.
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{14:1}\thinspace{}To serve a state in which the Way prevails, then to con\-tin\-ue serv\-ing it af\-ter it has lost the Way. That is shame\-ful.
\end{para}

\begin{para}
    \textsuperscript{14:2}\thinspace{}A gen\-tle\-man who is at\-tached to ma\-te\-ri\-al comforts is not wor\-thy of be\-ing called a gen\-tle\-man.
\end{para}

\begin{para}
    \textsuperscript{14:4}\thinspace{}A vir\-tu\-ous man will cer\-tain\-ly have some\-thing to say, but a man with some\-thing to say is not nec\-es\-sar\-i\-ly vir\-tu\-ous. A be\-nev\-o\-lent man will cer\-tain\-ly be brave, but a brave man is not nec\-es\-sar\-i\-ly be\-nev\-o\-lent.
\end{para}

\begin{para}
    \textsuperscript{14:7}\thinspace{}Can you love some\-one with\-out mak\-ing them work hard? Can you be loy\-al to some\-one with\-out instructing them?
\end{para}

\begin{para}
    \textsuperscript{14:10}\thinspace{}To be poor with\-out re\-sent\-ment is dif\-fi\-cult; to be rich with\-out ar\-ro\-gance is eas\-y.
\end{para}

\begin{para}
    \textsuperscript{14:12}\thinspace{}A man who is wise, free from desires, brave, ac\-com\-plished, and al\-so re\-fined through rit\-u\-al and mu\-sic may be called a com\-plete man. But, does a com\-plete man need all this now\-a\-days? If he sees prof\-it and thinks of what is right; if he sees dan\-ger and is read\-y to give his life; if he endures long hard\-ship and does\-n't for\-get the words he lives by: he may be called a com\-plete man.
\end{para}

\begin{para}
    \textsuperscript{14:20}\thinspace{}When your words are not hum\-ble, it's dif\-fi\-cult to live up to them.
\end{para}

\begin{para}
    \textsuperscript{14:22}\thinspace{}Zilu asked how to serve a rul\-er. Con\-fu\-cius said, “Don't use de\-ceit when you op\-pose him.”
\end{para}

\begin{para}
    \textsuperscript{14:27}\thinspace{}A su\-pe\-ri\-or man is a\-shamed if his words sur\-pass his deeds.
\end{para}

\begin{para}
    \textsuperscript{14:28}\thinspace{}The Way of a su\-pe\-ri\-or man has three things that I can't find in my\-self: the be\-nev\-o\-lent are not anx\-ious, the wise are not con\-fused, the brave are not a\-fraid.
\end{para}

\begin{para}
    \textsuperscript{14:29}\thinspace{}Zigong was judg\-ing oth\-er peo\-ple. Con\-fu\-cius said, “Zigong must be of such su\-pe\-ri\-or char\-ac\-ter him\-self that he has lei\-sure for this. I don't.”
\end{para}

\begin{para}
    \textsuperscript{14:30}\thinspace{}Don't wor\-ry that men don't rec\-og\-nize you; wor\-ry a\-bout your own lack of a\-bil\-i\-ty.
\end{para}

\begin{para}
    \textsuperscript{14:31}\thinspace{}To be the first to be a\-ware of de\-cep\-tion or bad faith, with\-out an\-tic\-i\-pat\-ing or suspecting it: is that not be\-ing wor\-thy?
\end{para}

\begin{para}
    \textsuperscript{14:34}\thinspace{}Some\-one asked, “What do you think of the say\-ing, ‘Re\-pay re\-sent\-ment with kind\-ness’?” Con\-fu\-cius said, “Then how will you re\-pay kind\-ness? Re\-pay re\-sent\-ment with uprightness. Re\-pay kind\-ness with kind\-ness.”
\end{para}

\begin{para}
    \textsuperscript{14:35}\thinspace{}No one understands me! I don't com\-plain a\-gainst Heav\-en, nor do I blame men. I stud\-y here be\-low, and I pen\-e\-trate up a\-bove. Who understands me? Heav\-en does!
\end{para}

\begin{para}
    \textsuperscript{14:36}\thinspace{}Gongbo Liao had slandered Zilu. Zifu Jingbo said, “I have e\-nough pow\-er to have Gongbo Liao's corpse dis\-played in the mar\-ket\-place.” Con\-fu\-cius said, “If circumstances al\-low, the Way will pre\-vail; if not, the Way will be a\-ban\-doned. What ef\-fect does Gongbo Liao have on these circumstances?”
\end{para}

\begin{para}
    \textsuperscript{14:37}\thinspace{}The highest vir\-tue is to a\-void the world; next, to a\-void cer\-tain places; next, to a\-void cer\-tain attitudes; next, to a\-void cer\-tain words.
\end{para}

\begin{para}
    \textsuperscript{14:41}\thinspace{}When those a\-bove love pro\-pri\-e\-ty, the com\-mon peo\-ple are eas\-y to lead.
\end{para}

\begin{para}
    \textsuperscript{14:44}\thinspace{}Con\-fu\-cius employed a boy to car\-ry messages. Some\-one asked, “Is he mak\-ing any prog\-ress?” Con\-fu\-cius said, “He's not seek\-ing prog\-ress. He wants quick results.”
\end{para}

\bigskip{}

\begin{para}
    \textsuperscript{15:2}\thinspace{}A su\-pe\-ri\-or man is stead\-fast in the face of ad\-ver\-si\-ty, while an in\-fe\-ri\-or man is engulfed by it.
\end{para}

\begin{para}
    \textsuperscript{15:5}\thinspace{}Can't Shun be said to have a\-chieved or\-der with\-out forc\-ing? What did he do? He just sat reverently on the throne, and faced due south.
\end{para}

\begin{para}
    \textsuperscript{15:6}\thinspace{}If your words are loy\-al and trust\-wor\-thy, and your con\-duct is sin\-cere and re\-spect\-ful, then your con\-duct will be prop\-er e\-ven a\-mong barbarians. If your words are not loy\-al and trust\-wor\-thy, and your con\-duct is not sin\-cere and re\-spect\-ful, how can your con\-duct be prop\-er, e\-ven in your own neigh\-bor\-hood?
\end{para}

\begin{para}
    \textsuperscript{15:8}\thinspace{}If you fail to speak to a man who is ca\-pa\-ble of un\-der\-stand\-ing, you waste the man. If you speak to a man who is in\-ca\-pa\-ble of un\-der\-stand\-ing, you waste your words. The wise waste nei\-ther men nor words.
\end{para}

\begin{para}
    \textsuperscript{15:9}\thinspace{}A be\-nev\-o\-lent man will not in\-jure his be\-nev\-o\-lence in or\-der to save his life, but he might sac\-ri\-fice his life in or\-der to a\-chieve be\-nev\-o\-lence.
\end{para}

\begin{para}
    \textsuperscript{15:10}\thinspace{}In what\-ev\-er coun\-try you're liv\-ing, serve its most vir\-tu\-ous officers, and be\-friend its most be\-nev\-o\-lent gentelmen. This is how to prac\-tice be\-nev\-o\-lence.
\end{para}

\begin{para}
    \textsuperscript{15:12}\thinspace{}If a man does\-n't con\-cern him\-self with what's far a\-way, he will have sorrows near at hand.
\end{para}

\begin{para}
    \textsuperscript{15:15}\thinspace{}If you de\-mand much from your\-self and lit\-tle from others, you will keep re\-sent\-ment at a dis\-tance.
\end{para}

\begin{para}
    \textsuperscript{15:18}\thinspace{}A su\-pe\-ri\-or man takes what is right as his ba\-sis. He practices it with pro\-pri\-e\-ty, demonstrates it with mod\-es\-ty, and completes it with good faith.
\end{para}

\begin{para}
    \textsuperscript{15:20}\thinspace{}A su\-pe\-ri\-or man hates the thought of leav\-ing the world with\-out mak\-ing a name for him\-self.
    \textsuperscript{21}\thinspace{}A su\-pe\-ri\-or man makes demands on him\-self; an in\-fe\-ri\-or man makes demands on others.
    \textsuperscript{22}\thinspace{}A su\-pe\-ri\-or man is con\-fi\-dent with\-out be\-ing con\-ten\-tious; gathers with others but does\-n't form cliques.
    \textsuperscript{23}\thinspace{}A su\-pe\-ri\-or man does\-n't pro\-mote a man be\-cause of his words, and does\-n't re\-ject the words be\-cause of the man.
\end{para}

\begin{para}
    \textsuperscript{15:27}\thinspace{}Clev\-er words dis\-rupt vir\-tue. Im\-pa\-tience in small matters disrupts great plans.
\end{para}

\begin{para}
    \textsuperscript{15:28}\thinspace{} When eve\-ry\-one loves a man, you must in\-ves\-ti\-gate. When eve\-ry\-one hates a man, you must in\-ves\-ti\-gate.
\end{para}

\begin{para}
    \textsuperscript{15:30}\thinspace{}To have a fault and not cor\-rect it is a fault in\-deed.
\end{para}

\begin{para}
    \textsuperscript{15:31}\thinspace{}I once spent a whole day with\-out eat\-ing and a whole night with\-out sleep\-ing in or\-der to think. I got noth\-ing out of it. It would have been bet\-ter to stud\-y.
\end{para}

\begin{para}
    \textsuperscript{15:32}\thinspace{}A su\-pe\-ri\-or man makes his plans a\-round the Way, not a\-round food. A su\-pe\-ri\-or man worries a\-bout the Way, not a\-bout pov\-er\-ty.
\end{para}

\begin{para}
    \textsuperscript{15:36}\thinspace{}In the pur\-suit of be\-nev\-o\-lence, don't yield e\-ven to your teach\-er.
\end{para}

\begin{para}
    \textsuperscript{15:37}\thinspace{}A su\-pe\-ri\-or man is prin\-ci\-pled, but not rig\-id.
\end{para}

\begin{para}
    \textsuperscript{15:38}\thinspace{}In serv\-ing your rul\-er, de\-vo\-tion to du\-ty co\-mes be\-fore any thought of com\-pen\-sa\-tion.
    \textsuperscript{39}\thinspace{}In teach\-ing, there is no such thing as so\-cial class.
\end{para}

\begin{para}
    \textsuperscript{15:40}\thinspace{}Those who fol\-low a dif\-fer\-ent Way can't make plans to\-geth\-er.
\end{para}

\begin{para}
    \textsuperscript{15:41}\thinspace{}In expressing your\-self, get the point a\-cross then stop.
\end{para}


\end{document}

% Copyright 2021 Michael Hendricks
%
% Permission is granted to use, copy, modify, and distribute
% this work for any purpose with or without fee.
%
% This work is provided as-is without any express or implied
% warranty. In no event is the author liable for any damages
% arising from this work.
